### 2023-08-17 Data-processing.Rmd
Initial commit

---
title: "Data-processing"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2023-08-17 Manuscript/Manuscript.qmd
Manuscript.qmd -> corresponds with Manuscript_V04.docx
Ready for review by the authorship team

---
title: "Equity conceptualizations and standards within transportation literature: a scoping review"
bibliography: All_papers_in_extracted.bib
format: docx
---

```{r, include=FALSE, message=FALSE, warning=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
```

```{r setup}
library(huxtable)
library(sf)
library(ggplot2)
library(tidyverse)
library(readxl)
library(DiagrammeR)
library(scales)
library(corrplot)
library(patchwork)
library(flextable)
library(ftExtra)
```

```{r PRISMA-flow-chart}
Pflow_chart <- grViz("digraph flowchart {
      node [fontname = Helvetica, shape = rectangle, style = filled, color = grey90]

      tab1 [label = '6832 references imported for screening']
      tab1_1 [label = '611 duplicates removed']
      tab2 [label = '6221 studies screened - title and abstract (step 1)']
      tab2_1 [label = '4511 studies excluded']
      tab3 [label = '1710 studies assessed for full-text eligibility (step 2 and 3)']
      tab3_1 [label = 
      <1545 studies excluded:<br ALIGN = 'LEFT'/> 
      &#8226; 1061 No ES (only EC)<br ALIGN = 'LEFT'/> 
      &#8226; 285 No full-text<br ALIGN = 'LEFT'/> 
      &#8226; 147 No EC and no ES<br ALIGN = 'LEFT'/> 
      &#8226; 38 No EC (only ES)<br ALIGN = 'LEFT'/> 
      &#8226; 13 Other<br ALIGN = 'LEFT'/> 
      &#8226; 1 Wrong study type<br ALIGN = 'LEFT'/> >]
      tab4 [label = 'Data from 165 studies extracted']

      blank1[label = '', width = 0.01, height = 0.01]
      tab1 -> blank1[dir=none];
      blank1 -> tab1_1[minlen=10];
      {{ rank = same; blank1 tab1_1 }}
      blank1 -> tab2

      blank2[label = '', width = 0.01, height = 0.01]
      tab2 -> blank2[dir=none];
      blank2 -> tab2_1[minlen=10];
      {{ rank = same; blank2 tab2_1 }}
      blank2 -> tab3

      blank3[label = '', width = 0.01, height = 0.01]
      tab3 -> blank3[dir=none];
      blank3 -> tab3_1[minlen=10];
      {{ rank = same; blank3 tab3_1 }}
      blank3 -> tab4
      }
      ")
```

```{r, eval=FALSE}
Pflow_chart %>%
  DiagrammeRsvg::export_svg() %>% 
  charToRaw() %>% rsvg::rsvg_png(file="Pflow_chart.png")
```

```{r import-extraction-data}
Consensus_all <- read_excel("ALL_Extraction_Results_Feb22_edited.xlsx") |> filter(`Reviewer Name` == "Consensus")

Consensus_only_analysis <- read_excel("ALL_Extraction_Results_Feb22_edited.xlsx") |> 
  filter(`Reviewer Name` == "Consensus") |> 
  filter(`Included in analysis` == "Yes")
```

```{r by-year-and-continent-plot}
year_cont_plot <- ggplot(data = Consensus_only_analysis,
                         aes(y = `_Publication year`, 
                         fill = `_Case study continent`)) +
  geom_bar(aes(x = (..count..)/sum(..count..)),col="black") + 
  scale_x_continuous(labels=scales::percent,
                     breaks = seq(0,0.2, by=0.05),
                     limits = c(0,0.2)) +
  xlab("% of papers") +
  theme_classic() + 
  theme(text=element_text(size=10),
        axis.title.y = element_blank(),
        legend.position = "bottom") + guides(fill = guide_legend(nrow = 1,
                                                                 title=NULL))
```

```{r modes-by-type-plot}
df_mode <- strsplit(Consensus_only_analysis$`_mode`, "; ", fixed = T)

df_mode <- data.frame(
  id=rep(Consensus_only_analysis$`Covidence #`, lengths(df_mode)),
  mode=unlist(df_mode))

df_mode <- df_mode |> group_by(mode) |> 
  summarize(count = n()) |> 
  mutate(count = count/sum(count))

mode_plot <- ggplot(data = df_mode, aes(x = forcats::fct_reorder(mode,count), y= count))+ 
  geom_bar(stat="identity", fill = "coral2", col = "black") + 
  scale_y_continuous(labels=scales::percent,
                     breaks = seq(0,0.4, by=0.05),
                     limits = c(0,0.4)) +
  coord_flip() +
  labs(y = "% of mentions in papers", subtitle = "Modes") + 
  theme_classic() + 
  theme(text=element_text(size=10),
        axis.title.y = element_blank())
```

```{r population-groups-plot}
df_pop <- strsplit(Consensus_only_analysis$`_population`, "; ", 
                       fixed = T) 
df_pop <- data.frame(
  id=rep(Consensus_only_analysis$`Covidence #`, lengths(df_pop)),
  pop=unlist(df_pop)) |> 
  mutate(pop = ifelse(pop == "Vulnerable populations (composite)", "Comp. vul", pop))

df_pop <- df_pop |> group_by(pop) |> summarize(count = n()) |> 
  mutate(count = count/sum(count))

#adding grouping var so they print on plot in the order we want
df_pop$group <- c(7,9,6,8,2,3,5,10,1,4)

pop_plot <- ggplot(data = df_pop, aes(x = pop, y = count)) +
  geom_bar(stat="identity", fill = c(rep("skyblue2",each=8), "grey", "skyblue2"), col = "black") + 
  scale_y_continuous(labels=scales::percent, limits = c(0,0.25), breaks = seq(0,0.25, by=0.05))+
  scale_x_discrete(limits=(df_pop$pop)[order(df_pop$group)]) +
  coord_flip() +
  labs(y = "% of mentions in papers", subtitle = "Population groups") + 
  theme_classic() + 
  theme(text=element_text(size=10),
        axis.title.y = element_blank())
```

```{r concs}
df_conc <- strsplit(Consensus_only_analysis$`_equity conceptualization`, "; ", 
                                                   fixed = T)

                            df_conc <- data.frame(
                              id=rep(Consensus_only_analysis$`Covidence #`, lengths(df_conc)),
                              conc=unlist(df_conc)) %>% 
                              mutate(conc = ifelse(conc == "Inequitable exposure to externalities", "Ineq. externalities", conc),
                                     conc = ifelse(conc == "Sufficientarism / Capabilities approach", "Emerg. theories", conc),
                                     conc = ifelse(conc == "Transport-related exclusion", "TR exclusion", conc))

df_standard_mcat <- strsplit(Consensus_only_analysis$`_type of equity standard2`, "; ", 
                       fixed = T) 
df_standard_mcat <- data.frame(
  id=rep(Consensus_only_analysis$`Covidence #`, lengths(df_standard_mcat)),
  std_mcat=unlist(df_standard_mcat))

standards_conc <- merge(df_conc, df_standard_mcat, by="id", all = T)

NUM_standards_conc <- standards_conc %>% group_by(id) %>% summarize(count = n())
NUM_ENTRIES_standards_conc <- NUM_standards_conc$count %>% sum()

standards_conc <- merge(standards_conc, NUM_standards_conc, by="id", all = T)

# standards_conc_plot <- ggplot(data = standards_conc,
#                               aes(x = `conc`, fill = `std_mcat`)) +
#   geom_bar() +
#   labs(x = "Conceptualizations (summarized)", fill = "Standards (main categories)") +
#   theme_classic() + 
#   theme(text=element_text(size=10),
#         axis.title.y = element_blank(),
#         axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
#   coord_flip()

# c("Wellbeing","Vertical equity","Utilitarian","TR exclusion", "Spatial equity", "Rights", "Ineq. externalities", "Horizontal equity", "Emerg. theories")
```

```{r, message=FALSE}
standards_conc <- standards_conc %>% 
  group_by(conc, std_mcat) %>% summarize(total = n())

standards_conc <- standards_conc %>%
  mutate(total_perc_in_conc = total/sum(total),
         total_perc_in_all = total/NUM_ENTRIES_standards_conc)
sums_concs <- standards_conc %>% group_by(conc) %>% summarize(
  total_perc_in_all = sum(total_perc_in_all),
  total_perc_in_conc = 1) %>% mutate(std_mcat = "TOTALS")

new_rows <- data.frame(conc = c("Emerg. theories", "Utilitarian", "TOTALS"),
                       total_perc_in_all = c(NA, NA, NA),
                       total_perc_in_conc = c(NA, NA, NA),
                       std_mcat = c("Environmental +","Environmental +", "TOTALS"))

sums_std_mcat <- standards_conc %>% group_by(std_mcat) %>% summarize(total_perc_in_all = sum(total_perc_in_all),
  total_perc_in_conc = 1) %>% mutate(conc = "TOTALS")

totals_standards_conc <- rbind(standards_conc, sums_concs)
totals_standards_conc <- rbind(totals_standards_conc, sums_std_mcat)
totals_standards_conc <- rbind(totals_standards_conc, new_rows)

#adding the grouping numbers, so the order that vars are printed are in geom_tile() are chronological
totals_standards_conc$group <- c(12,13,14,
                    41,42,43,44,
                    31,32,33,34,
                    21,22,23,24,
                    51,52,53,54,
                    71,72,73,74,
                    02,03,04,
                    81,82,83,84,
                    61,62,63,64,
                    15,45,35,25,55,75,05,85,65,
                    91,92,93,94,
                    11,01,95)

#adding long definitions, for table under heatmap

totals_standards_conc_table <- data.frame(
  Conceptualizations = c("The distribution of transport system benefits/costs should favourably prioritize disadvantaged groups.",
                         "Concerning the links between transport system interaction and social exclusions.",
                         "Evaluation of transport system benefits/costs through wellbeing measures (e.g., livability, quality of life, subjective well being).",
                         "Equal distribution of transport system benefits/costs across space.",
                         "Equal distribution of transport system benefits/costs across population."),
  Conceptualizations_cont = c("Concerning the equitable distribution of externalities (e.g., road safety, congestion, pollution).",
                         "Concerning legal/human rights of access (e.g., mobilty rights, rights to the city, discriminatory application of laws).",
                         "E.g., Sufficientarism: minimums informed by individual capabilities. Capabilities approach: the capability of living the life one values (rather than having the right to do so)",
                         "Maximizing benefits/welfare overall, without consideration of distribution.",
                         " "),
  Standards = c("Related to user characteristics (e.g., affordability levels, perceived equity).",
                "Access to destinations indicators (e.g., # of parks within 30-min transit trip), the use of travel impedance thresholds (e.g., 300m to bus stop).",
                "E.g., Level of service, quality, universal design standards.",
                "E.g., noise, air, pollution criteria, safety.",
                " ")
)

```

```{r}
set_flextable_defaults(font.size= 7,
                       font.family = NULL,
                       text.align = 'justify')

totals_conc_table_flex <- 
  # compose(totals_conc_table_flex,
  #         i = 1, j = c(1,2), part = "header", value = as_paragraph(c("Conceptualizations", "Conceptualizations"))) %>%
  prepend_chunks(flextable(totals_standards_conc_table %>% select(-c("Standards"))),
                 part = "body", i = c(1,2,3,4,5), j = 1, as_b(c("Vertical equity: ", "TR exclusion: ", "Wellbeing: ", "Spatial equity: ", "Horizontal equity: "))) %>% 
  prepend_chunks(part = "body", i = c(1,2,3,4), j = 2, as_b(c("Ineq. externalities: ", "Rights: ", "Emerg. theories: ", "Utilitarian: "))) %>% 
  # merge_h(part="header", i=1) %>% #header merge
  merge_at(i=5, j=1:2) %>%
  # align( align="center", part="header") %>%
  delete_part("header") %>%
  border_outer() %>%
  border_inner_h() %>%
  autofit()

totals_standards_table_flex <- 
  # compose(totals_standards_table_flex,
  #         i = 1, j = c(1), part = "header", value = as_paragraph(c("Standards"))) %>% 
  prepend_chunks(flextable(totals_standards_conc_table %>% select("Standards")),
                 part = "body", i = c(1,2,3,4), j = 1, as_b(c("Population: ", "Opportunity: ", "Infrastructure: ", "Environmental +: "))) %>%
  merge_at(i= 4:5, j= 1) %>%
  delete_part("header")%>%
  border_outer() %>%
  border_inner_h() %>%
  autofit()
```

```{r}
standards_conc_percentages_plot <- ggplot(totals_standards_conc, 
       aes(x=std_mcat, y=conc, fill= total_perc_in_all 
           #width = rep(10, each=50)
           ))+ 
  geom_tile() + 
  scale_fill_gradient2(low = "#AFAFED", mid= "#3F3FD4", high = "#191970", midpoint = 0.2, limits = c(0,.4), na.value = "#FCB5AC") +
  scale_x_discrete(limits=(totals_standards_conc$std_mcat)[order(totals_standards_conc$group)][1:5]) + 
  scale_y_discrete(limits=(totals_standards_conc$conc)[order(totals_standards_conc$group)] ) +
  geom_text(aes(label = total_perc_in_all %>% percent(accuracy = 0.1)), color = "white", size = 2.5) +
  geom_vline(xintercept=4.5, size=1.5, color = "#191970") +
  geom_hline(yintercept = 43.5, size=1.5, color="#191970")+
  theme_classic() +
  coord_flip()+
  theme(legend.position = "none",
        axis.ticks = element_blank(),
        axis.text.x=element_text(angle=20,hjust=1, colour = "black"),
        axis.text = element_text(size=7, colour = "black"),
        axis.title = element_blank())+
  ggtext::geom_richtext(aes(x=5.385, y=23.5, label="Conceptualizations"), fill="grey", size=2) +
  ggtext::geom_richtext(aes(x=3, y=49, label="Standards"), angle = 270, fill="grey", size=2) 
# standards_conc_percentages_plot
# ggsave("standards_conc_percentages_plot.png")
```

```{r, warning=FALSE}
#the heatmap on top and table of stand/concept definitions on the bottom.
standards_conc_percentages_plot_and_table <- cowplot::plot_grid(
  gen_grob(totals_standards_table_flex, scaling="fixed", fit="auto"),
  standards_conc_percentages_plot,
  NULL,
  gen_grob(totals_conc_table_flex, scaling = "fixed", fit="auto"),
  nrow = 2, ncol=2, rel_heights = c(1.5,1), rel_widths = c(1,4) )
#standards_conc_percentages_plot_and_table
# ggsave("standards_conc_percentages_plot_and_table.png")
```

```{r echo=FALSE}
#function for creating comments
word_comment <- function(comment, highlight = "") {
  if (isTRUE(knitr:::pandoc_to() == "docx")) {
    paste0('[', comment, ']{.comment-start id="0" author="Anastasia"',
           'date="2023-08-12T15:00:00Z"}', highlight, '[]{.comment-end id="0"}')
  }
}
```

# Introduction

Equitable transportation system from a planning perspective, involves fair production, distribution, and reproduction of transportation benefits and costs within a community [@pereiraDistributiveJusticeEquity2017a; @shellerMobilityJusticePolitics2018; @pereiraTransportationEquity2021].
Transportation equity is a multi-scalar and multi-dimensional topic centered on the distribution of transport-related benefits and burdens across population groups and communities within urban, suburban or rural areas.
Benefits and burdens can include, but are not limited to, accessibility, mobility, affordability, health and environmental issues that impact individuals, marginalized members of a community, and the community as a whole [@lucasMeasuringTransportEquity2019].

Transportation systems are complex: they support multi-modal uses and are home to emerging new technologies and service modes that complicate the balance of benefit and burdens [@guoSystematicOverviewTransportation2020].
Consequently, achieving an *equitable* system is not a positivist attribute; it should be normatively defined by the community  and broader society [@pereiraTransportationEquity2021; @paezMeasuringAccessibilityPositive2012] during the ongoing process of measurement, operationalization, and improvement.
Both academic literature and planners are grappling to contextualize what an equitable transportation system may look like along spatial, temporal, environmental, and socio-economical dimensions [@boisjolyHowGetThere2017; @pereiraTransportationEquity2021].
Synthesizing methods on how to identify, measure, and operationalize transportation equity is urgently needed as planners and decision-makers are undecided on how to distribute existing transportation resources and simultaneously invest in future transportation infrastructure *fairly*.

The application of equity within the realm of transportation planning has proliferated in the academic literature but concrete adoption of the concepts in planning practice has been minimal [@pereiraTransportationEquity2021; @boisjolyHowGetThere2017; @doranPursuitCyclingEquity2021; @linovskiEquityTransportationPlanning2020; @litmanEvaluatingTransportationEquity2022].
Academic literature has approached equity through measures such as accessibility and activity participation [@paezMeasuringAccessibilityPositive2012; @allenPlanningTransportSocial2020], affordability [@isalouTestingHousingTransportation2014], environmental impacts and safety [@guoSystematicOverviewTransportation2020], and health [@fransenIdentifyingPublicTransport2015; @smithSystematicLiteratureReview2017], and also conceptualizations such as frameworks for equity and social exclusion [@lucasProvidingTransportSocial2006; @lucasTransportPovertyIts2016], and distributive justice theories [@behbahaniDesigningRoadNetwork2019; @lewisExploringEquityHow2021; @pereiraDistributiveJusticeEquity2017a; @vecchioAccessibilityCapabilitiesApproach2021], among others.
However, a gap in the literature is the absence of a comprehensive review that examines how equity is conceptualizated, theorized and operationalized through measures and standards.
These inquiries are vital to planning agencies and decision-makers, and as such, this review aims to definitively benchmark the knowledge in the literature from this perspective.

Though aspects of transport equity in metropolitan planning documents such as accessibility [@boisjolyHowGetThere2017] or mode-specific equity [@doranPursuitCyclingEquity2021] have been reviewed, no review to date captures how equity has been theorized and translated into standards from the perspective of equity dimensions.
Our work contributes in this direction, reviewing which equity dimensions, standards, and conceptualizations have been applied in the academic literature and identifying where gaps exist in the literature.
Ultimately, this review aims to collate the academic knowledge and present it in an operationalizable manner to both academics and decision-makers in hopes to catalyze the uptake in equitable planning practices and guide future research programmes which develop definitions, measures, and recommendations for evaluating transportation equity.

This paper is structured as follows.
In Section 2, we set the stage by defining what do we mean by equity dimensions, conceptualizations and types of standards used in the literature search and throughout the paper.
In Section 3, we outline the methods used in the scoping review conducted.
In Section 4, we summarize the findings of the literature.
In Section 5, we discuss the methods connects to standards, conceptualizations and dimensions.
In Section 6, we summarize the findings and link them to future transport planning agendas in addition to visioning ways the findings can be used to build *just* transportation planning processes.

# Setting the stage

Transportation equity literature exists within cross-cutting categories.
We borrow a "Where?", "When?", "Who?", "What?", and "How?" framing to categorize aspects of how equity is assessed.
This categorization is outlined in @jaggarPhilosophicalChallengesGlobal2009 to describe substantive sets of answers to questions of distributive justice (and further discussed in the context of justice within the critical human geography discipline [@przybylinskiWhereJusticeGeography2022]).
For questions of *justice*, convincing answers to these questions require a rationale: a "Why?" [@jaggarPhilosophicalChallengesGlobal2009].
For the purpose of this review focused on *equity*, the focus is on studies that substantively answer "Where?", "When?", "Who?", "What?", and "How?"; the answer to "Why?" may be incidental. 
Though critical, answers to "Why?" are not overwhelming common in the transportation equity academic literature that also address  "Where?", "When?", "Who?", "What?", and "How?". This is discussed as a next step in planning for *justice* transportation literature in Section 6.

Furthermore, it is important to distinguish between equity and justice.
Our focus is on equity, which is nested within and can be interpreted as the application of justice theories.
For instance, *mobility justice* seeks to understand how broader power inequities inform the governance and control of movement are reproduced by mobility systems [@shellerMobilityJusticePolitics2018]; in understanding what is *just* and *unjust,* equity is the application of processes to move towards justice.
What is justice must be decided at a community level, a goal of the equity processes, and cannot be captured in a single academic text.

### Definitions

For all literature reviewed in the transportation domain, equity *dimensions*, *conceptualizations*, and *standards* are discussed through the "Where?", "When?", "Who?", "What?", and "How?" framings.
As follows, we define each of these categories and how they inter-connect.

**Equity dimensions**: the assessment of transportation equity encompasses various dimensions, mirroring the multifaceted nature of transportation systems. These dimensions can be distinguished as the **"What?"**, specifically: (1) mobility/accessibility, (2) traffic-related pollution, (3) traffic safety, and (4) human-health as provided in a conceptual framework by @lucasMeasuringTransportEquity2019.
All papers reviewed include dimensions that cover the **"What?"** of the transport equity issue and can be further disaggregated:

-   *When* and *Where*: includes the article's publication year, the geographic context (i.e., urban, suburban and rural areas, countries) and location.
-   *Who*: includes the type of population group, mode type, activity/opportunity type, and transport network status.
-   *How*: includes the measure used to assess equity dimensions as supported by a **equity conceptualization** and **equity standard**. **Equity conceptualizations** are some type of motivating equity principle, philosophy or theory. For example: theoretical and conceptual frameworks (e.g., transport-related social exclusion, transport disadvantage, transport poverty, horizontal equity, vertical equity) and equity principles (e.g., Utilitarism, Capabilities Approach, Sufficientarism). **Equity standards** are thresholds that when operationalized can be used to define when an aspect of the transportation system is in/equitable. They can be quantitative thresholds, qualitative descriptions, and/or mixed-method thresholds. Examples include: maximum travel distance/cost/time to or from key destinations, levels of maximum exposure to externalities (i.e., noise or air pollution), distributional assessment (e.g., composite indicator, range, maximum distance or gap, etc.), dis/satisfaction with travel, etc.

papers that include equity **dimensions**, **conceptualizations**, **standards** and substantive answers to the "When?", "Where?", and "Who?" in relation to equity in transportation systems (i.e., the users, the infrastructure, the service, the outcomes, or a combination of all) are used to answer the following research questions:

-   What are the equity dimensions, equity standards, and equity conceptualizations and how are they applied in the existing transportation equity academic literature?
-   Based on how equity has been applied in the literature, what may be paths forward in theorizing justice in the context of transportation systems?

# Literature review methods

This review examines the breadth and depth of the academic literature on transportation equity that analyses equity within at least one equity dimensions and implements equity conceptualizations and equity standards as defined in the previous section. This review follows the Joanna Briggs Institute (JBI) approach to the conduct of scoping reviews which builds upon the Arksey and O'Malley (2005) framework [@petersUpdatedMethodologicalGuidance2020].
The JBI is a global organization that creates frameworks, protocols and guidance for the synthesis of research.
This review is also guided by the Preferred Reporting Items for Systematic Reviews and Meta-Analyses extension for scoping reviews (PRISMA-ScR), consistent with the JBI approach [@triccoPRISMAExtensionScoping2018].
Using these methods allow the review to explore the broad topic of transportation equity within the literature and collate knowledge from across methodologies in the academic literature.

The primary research question and the protocol was drafted and refined from preliminary searches of related-reviews e.g., [@iglesiasHowUnevenUrban2019; @sagarisUsingPARFrame2020; @vecchioAccessibilityCapabilitiesApproach2021], consultations with the authorship team who are engaged in transportation equity related research, and a University of Toronto Research Services Librarian and Liaison Librarian in City Studies.
The methods are described in two parts: (i) the development of the search strategy and (ii) the evidence selection and data extraction.

## Search strategy

To guide the selection of search terms within the search query, **inclusion** and **exclusion** criteria were developed [@petersUpdatedMethodologicalGuidance2020].
For the inclusion criteria, the mnemonic PCC (population, concept, and context) was adopted (see Appendix @fig-A1: for details).

Next, the inclusion and exclusion criteria was applied to develop the search strategy.
The search strategy was developed iteratively, at each stage adding topic search terms (e.g., terms in the title, abstract or key words) to the set of terms tied together with logical connector terms "AND" and "OR".
These stages are summarized in what follows and the full search term queries are detailed in Appendix @fig-A1.

1.  An initial limited search of the core collections within the Web of Science (WoS) platform was undertaken to identify key papers. Terms for 'transportation' and 'equity' were generated, respectively. The text words contained in the titles and abstracts of relevant papers, the index terms used to describe the papers, and subject heading searches when available (depending on the database within WoS). This search was iterated on and took the form: ("Transport" OR "Transit" OR "Car\*" OR "Walk" OR "Bike"...**1**) AND ("Equity" OR "Justice" OR "Equity" OR "Fair"...**2**), where **1** and **2** signify additional  terms that relate to 'transportation' and 'equity', respectively.
2.  Upon inspection of the preliminary search results and team consensus, the 'equity' set of search terms was expanded into three sets of terms. The first set describes equity theories, the second describes equity dimensions, and the third describes terms referring to standards. All three sets of terms were iteratively augmented. The finalized search query took the following general form: ("Transport" OR "Transit" OR "Car\*" OR "Walk" OR "Bike"...**1**) AND ("Equity" OR "Justice" OR "Equity" OR "Fair"...**2**) AND ("Accessibility" OR "Mobility" OR ...3) AND ("Standard" OR "Threshold" OR ...4) where **1**,**2**,**3**, and **4** signify additional terms included in the sets combined with "OR" logical connectors.

The search strategy, including all identified keywords and index terms, were applied to each database and platform searched (i.e., Web of Science General Collection -Science Citation Index Expanded, Social Sciences Citation Index (Web of Science), and Transportation Research International Documentation (TRID)).
The exports of the final search query was completed by the lead researcher on March 21st 2021.

## Evidence selection and data extraction

Evidence selection consists of scanning the exported literature from the search strategy and retaining the papers that fit the inclusion and exclusion criteria.
This process was pilot-tested with a subset of papers before being implemented on the full set.
Then, *Covidence*, an online application for literature screening, was used for all steps of selection and data extraction on the full export of literature. All steps are visualized in @fig-fig1.

```{r methods_framework-fig1}
#| label: fig-fig1
#| fig-cap: "Evidence selection process framework. Step 1 (orange) is title and abstract screening, step 2 (green) is full-text review, and step 3 (purple) is data extraction."
knitr::include_graphics("Methods_framework.png")
```

The process is summarized in the following three steps:

1.  The first step (orange box in @fig-fig1) included screening all titles and abstracts of papers on whether they included transportation equity as defined by the PCC. All papers were voted on by two independent reviewers for inclusion, exclusion, or uncertain inclusion. All uncertain papers, conflicting papers, and papers missing abstracts were voted on by a third reviewer for inclusion or exclusion.
2.  The second step (green box in @fig-fig1) included reviewing all full-text papers which passed step 1. These papers were reviewed based on if their study design included an equity standard and equity conceptualization. All papers are voted on by two independent reviewers for inclusion or exclusion. If an article was voted to be excluded, it was tagged with one of five possible reasons, namely (1) no standards included, (2) no conceptualizations included, (3) no standard and no conceptualization included, (4) send back -- QA issue, (5) other. Discrepancies were resolved by a third reviewer.
3.  A data extraction template was filled out for each record: one reviewer for the third and final step (purple box in @fig-fig1). The data extraction template was created to balance complexity of categories and simplicity of summary; information related to "What?" (equity dimension), "Who?" (population group, mode type, type of destination/activity), "Where?" (region of study), and "How?" (the equity standard(s) and associated conceptualization(s)) was filled out for each study. @fig-A2 contains the template that was input into *Covidence* and used within this review.

```{r PRISMA-fig2}
#| label: fig-fig2
#| fig-cap: "PRISMA flow diagram for the evidence selection process. ES signifies equity standard and EC signifies equity conceptualization."
knitr::include_graphics("Pflow_chart.png")
```

The evidence selection process is also represented using the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) flow diagram [@pagePRISMA2020Statement2021] in @fig-fig2.
Notably, two rounds of exclusion occurred during the assessment for full-text eligibility. 1710 studies entered step 2, 1223 were excluded and the remaining 487 papers entered step 3. The data extraction template used by the reviewers (authorship team) in step 3 revealed that inclusion was initially too generous, and some papers did not have a sufficiently strong equity standard and/or equity conceptualization. As such, 322 papers were further excluded and data extraction was completed for the remaining 165 papers. A summary of the reasons for exclusion of the 1545 papers (between steps 2 and 3) are included in @fig-fig2.

# Summary of `r word_comment("NOTE: in this section, only papers that are included in the review (i.e., have a full template) are cited.", "findings")`

A synthesis of the findings from the data extraction template (shown in @fig-A2) is detailed in this section.
The presentation of findings is less granular than the template to highlight key trends in the literature.

## Temporal and spatial focus (the "When" and "Where")

```{r counts-case-study-and-urban}
case_cont_count <- strsplit(Consensus_only_analysis$`_Case study continent`, "; ",
                       fixed = T)
case_cont_count <- data.frame(
  id=rep(Consensus_only_analysis$`_Case study continent`, lengths(case_cont_count)),
  case_cont=unlist(case_cont_count))
case_cont_count <- case_cont_count |> group_by(case_cont) |> summarize(count = n())

urban_form_count <- strsplit(Consensus_only_analysis$`_Case study urban form`, "; ",
                       fixed = T)
urban_form_count <- data.frame(
  id=rep(Consensus_only_analysis$`_Case study urban form`, lengths(urban_form_count)),
  urban_form=unlist(urban_form_count))
urban_form_count <- urban_form_count |> group_by(urban_form) |> summarize(count = n()) 
```

@fig-fig3 displays the papers included in this review by year of publication and case study continent. Evidently, transportation equity related papers are increased in popularity throughout the years, particularly since 2019.
Of note is the geographic scope of the case studies present in the papers.
The majority of papers (`r ((case_cont_count$count[4] + case_cont_count$count[5]+ case_cont_count$count[6])/sum(case_cont_count$count)) |> percent()`) contain case studies based in the Global North e.g., North America (particularly USA and Canada), Europe (particularly UK, France, Spain and Scandinavia), and Oceania (Australia and New Zealand).
How *equity* is operationalized is context-specific, and with the majority of transportation systems *not* in the Global North, the literature included in this review falls short in representing the global perspective.

```{r plotting-year-cont-plot-fig3}
#| label: fig-fig3
#| fig-cap: "Papers included in the review by year of publication and case study continent."
year_cont_plot
ggsave("year_cont_plot.png")
```

The few studies in the Global South are predominately in the **Asian** continent, specifically China but also India, Thailand, South Korean, Iran, Philippines, Indonesia, Israel, and Japan.
These studies focus on a variety of modes, population groups, and equity conceptualizations and standards that cannot be succinctly summarized. <!-- develop further, if there's room?-->

The next most common focus within the literature from the Global South centers on **South America**.
Many of these studies mention a systematic absence of evidence relevant to the region [@vecchioTransportEquityLatin2020]. Despite the growing recognition in the literature of the interconnections between transport development, social exclusion, and poverty [@benevenutoExaminingTransportNeeds2020], studies underscore an ongoing neglect of the social dimension of transport during the planning stage [@benevenutoExaminingTransportNeeds2020; @boisjolyAccessibilityMeasurementsSao2020].
Many studies also indicate affordability as one of the main mobility barriers in the region [@falavignaAssessingInequalitiesPublic2016; @rivasHowAffordableTransportation2018], while some highlight multi-dimensional concerns such as public transport accessibility and quality of walking environments that contribute to mobility inequalities [@tiznadoaitkenRoleAccessibilityPublic2018].

Within the reviewed literature, studies pertaining to **Africa** are less numerous compared to the South American literature.
A shared characteristic among the studies from the two continents is a scarcity of official transport data [@friedMeasuringSustainableDevelopment2020] and reliance on external policy guidelines. These studies also incorporate the utilization of informal transportation options and pressures to development physical road network infrastructure (supporting car dependency) over meeting mobility/accessibility needs of citizens [@thondooSmallCitiesBig2020].
To address these challenges, researchers compile databases based on open and geo-referenced data, calculate objective and/or subjective measures [@berheAdaptationDissonanceQuality2014], and focus on advancing transport justice for low to medium income countries (LMIC) by aligning with external policy guidelines such as the Sustainability Development Goals (SDGs), particularly those related to universal accessibility [@friedMeasuringSustainableDevelopment2020].

From all the studies included in the review, `r (urban_form_count$count[3]/sum(urban_form_count$count)) |> percent()` focus on the urban and suburban context and are highly varied in their research aims. To give an example, from the perspective of cycling as a mode, @coxQualitativeStudyAccessibility2020 qualitatively examines cycling as a mode for people with disability in a typical mid-size town in the UK and @ampeImpactChildBike2020 identifies the lateral clearance that motorists should maintain when passing cyclists with children seats; both focus on understanding barriers to cycling from different perspectives, but within different urban environments.

The remainder of the studies focus on rural regions (`r (urban_form_count$count[2]/sum(urban_form_count$count)) |> percent()`). To illustrate examples, @caoIndicatorsSocioSpatialTransport2017 examined transportation disadvantage in remote places which rely on inter-island ferry trips in the rural Philippines.
Similarly, @parrySocialVulnerabilityClimatic2018 studied remote communities in the Amazonian and suggests that "increasing accessibility through road building would be maladaptive, exposing marginalized people to further harm and exacerbating climatic change by driving deforestation" (pp. 125).

## Populations of interest (the "Who" pt.1)

@fig-fig4 showcases the categories of population group types that are the focus of the reviewed papers.
From this tally, papers that consider income groups is the most widely represented in the literature.
Particularly, most papers that focus on income pay particular consideration to the lowest-income groups, as they experience the greatest burden of transport inequalities, be it accessibility, affordability, or other dimensions [@peungnumsaiGridBasedSpatialAnalysis2020; @zhaoUnequableSpatialAccessibility2020; @falavignaAssessingInequalitiesPublic2016].

```{r plotting-population-group-plot-fig4}
#| label: fig-fig4
#| fig-cap: "The proportion of papers that focus on each type of population group. Categories for population groups were generated upon data extraction."
pop_plot
ggsave("pop_plot.png")
```

An abundant amount of literature suggests low-levels of household income is a significant determinant of transport-related inequities (e.g., access to public transport supply in Bangkok region (Thailand) [@peungnumsaiGridBasedSpatialAnalysis2020], access to employment opportunities in various cities in Brazil [@boisjolyAccessibilityMeasurementsSao2020], and less environmental noise, air, and higher green space per resident ratios in Rijnmond region (Netherlands) [@kruizeEnvironmentalEquityRole2007].
But this should be kept in context, as low-income does not necessarily mean lower transport-related benefits.
For instance, in Sheffield (UK), @mearsUnderstandingSocioeconomicEquity2019 demonstrates that historically working-class neighbourhoods (i.e., lower income working population) have more access to green space than other neighbourhoods due to urban planning approaches during the Victorian-era. However, the quality of green spaces are less than average.
Similarly, @bertrandMeasuringMappingDisparities2008 finds that lower income groups do not always have below average accessibility depending on the granularity of analysis (i.e., the distance-to-food threshold for the cumulative opportunity measure).

Age is the second most common category of population group of focus within the reviewed literature.
Many papers focusing on this category highlight differing age-related capabilities; for instance, @martinez-jimenezAccessibilityCultureEducation2019 and @arranz-lopezSocialSpatialEquity2019 investigate travel distances and times to various opportunities based on specific age groups, acknowledging that age is an important consideration to opportunity access variability between populations.
The most commonly focused on age groups are school-aged children and older populations. School-aged children oriented papers analyze wellbeing [@laszkiewiczChildrenGreenWalk2020], exposure to green space [@corazzaMethodologyEvidenceCase2020], access to schools [@sharmaSpatialSocialInequities2022], and understanding and encouraging active travel journeys [@mackieOvercomingBarriersCycling2009; @mehdizadehWalkingTimeSchool2017]. Papers focusing on older-populations typically have similar aims as the children-focused articles e.g., understanding transport-related impacts on wellbeing e.g, [@chenSpatialAnalysisFramework2020], measuring accessibility to population-specific key destinations e.g. [@chengInvestigatingWalkingAccessibility2019], and seeking to understand how to better meet travel needs e.g., [@nordbakkeTransportUnmetActivity2015].

The third most commonly focused on population category are what we summarize as 'composite vulnerable population groups'. These papers use some sort of composite vulnerability index that captures multiple population vulnerabilities (e.g., low-income, unemployment, immigrant status, family household characteristics, etc.) typically generated from official government sources or author-informed census data creation. These indices are varied in methodology: reflecting the intersectionality of population vulnerabilities.
For instance, @awuorInfluenceEnvironmentalHealth2019 uses the Neighbourhood Equity Index (NEI) to disparities in premature death in Toronto (Canada). The NEI is a composite index that was developed by the city to capture the differences in the City's neighbourhoods by ranking them based on socio-economic characteristics (e.g., social assistance, unemployment, income) and physical environmental characteristics such as green space availability. 
Other works use national census indicators such as the social and housing deprivation index e.g., [@pucciInequalitiesJobrelatedAccessibility2019], while others using census household poverty measures and calculated transport-related indicators like accessibility to estimate transport disadvantage [@sunPublicTransportAvailability2021; @scheurerSpatialAccessibilityPublic2017], evaluation of equity in policy implementation [@aldredEquityNewActive2021], and inequitable transport-related mortality burden [@iungmanImpactUrbanTransport2021]. Similarly, Environmental Justice (EJ) indicators have been used in US literature to identify neighbourhoods that have a higher than average proportion of low-income and non-white populations (i.e., a composite vulnerable population group'). Many studies has used EJ analysis to evaluate the equity impacts of transportation projects e.g., [@rowangouldIdentifyingEnvironmentalJustice2016; @parkTransitParksEnvironmental2021; @reddySafeguardingMinorityCivil2010].

`r word_comment("Papers that focus ONLY on (Dis)ability, or ONLY on Gender, or ONLY on Race/Ethnicity, or ONLY on Employment/Education level... the only focus", ".")` Papers that exclusively focus on populations with (dis)abilities e.g., [@parkJourneyVisuallyImpaired2017; @chiscanoImprovingDesignUrban2021; @orellanaWalkRollMapping2020] are relatively common in the reviewed literature. They mainly assess universal design guidelines and the ability for people with (dis)abilities to travel. However, from another perspective papers with an exclusive focus on gender, race/ethnicity, or education level/employment are much less common in the reviewed literature. Only two papers focus on gendered differences in cycling/active transportation (e.g., @adlakhaMindGapGender2020's case study in Chennai (India) and @xieWonCycleRoute2018's case study in Cardiff (UK)). Only two papers focus on race/ethnicity exclusively focusing on how minority ethnicity communities are in proximity to green space [@silvaEnvironmentalJusticeAccessibility2018] and culturally diverse family physicians in Toronto (Canada) [@wangModelingSpatialAccessibility2011]. Furthermore, papers that focus *solely* on education/employment status are not present in the reviewed papers. This is to say, papers that feature gender, race/ethnicity, or education level/employment population groups often feature them alongside other population group characteristics (or within 'composite vulnerability measures'). This contrasts the prominence of studies that exclusively center on (dis)abilities as a population category.

The *Other* category are papers that include group population characteristics that are more difficult to classify. Examples include: veterans and access to specific-healthcare needs [@mooneyTravelDistanceBarrier2000], pregnant people and access to services [@vadrevuMeasuringSpatialEquity2016], and youth populations who live in foster care [@batscheUsingGISEnhance2012]<!-- rural populations people who use transit and their service needs relative to urban needs @sharavWhatTransitService2019, only cyclists in the context of health-implications of cycling in Delhi (India) @apparicioCyclingOneMost2021 ) -->. Overall, the diversity of the *Other* population group classification demonstrates the diversity of transportation-equity concerns across population groups in the reviewed literature and the interplay of characteristics in the literature [@vecchioFairTransportPolicies2022].

## Transport modes (the "Who" pt.2)

```{r plotting-mode-plot-fig5, fig.width=5}
#| label: fig-fig5
#| fig-cap: "The proportion of papers that investiage each type of mode. Categories for modes were generated upon data extraction."
mode_plot
ggsave("mode_plot.png")
```

Shifting to transport modes of interest, the primary emphasis within the reviewed papers centers on public transit (@fig-fig5). The variety of topics assessed from the perspective of public transit are varied. For instance, @mckeyCrowdsourcedMappingHealthy2020 identifies 'food deserts' in Dallas (USA) considering public transit accessibility. Other contributions intersect public transport and individual needs, such as, universal design and barrier-free transportation for people with disabilities e.g., [@jimenez-espadaResearchProblemUniversal2021; @liuStatusQuoChallenges2019] or how public transport can be improved/service be assessed for people with autism [@limFacilitatingIndependentCommuting2021; @feeleyValidationParatransitSkills2019]. There may be a strong focus on public passenger transport systems because they can be altered to improve equity - it is a collective social system that has the potential to be sufficiently funded in order to provide barrier-free transport for most but it requires a variety of inputs that cannot always be met (e.g., sufficient density as a result of land-use, economic development, political will).

Transit is also often central to multi-modal or holistic comparisons that may serve transport equity analysis.
As an example, @brusselAccessAccessibilityCritique2019 compares three different approaches to measure accessibility in context to the Sustainability Development Goals (11.2) for the case of Bogota (Colombia), all of which capture some/all of the public transit system while others capture road and/or pedestrian systems.
From a different perspective, @renneWhatHasAmerica2018 reviews natural disaster emergency evacuation plans from the lens of the no-car (and oftentimes vulnerable) households in regions across the USA, paying particular attention to transit and pedestrian networks.

In contrast, a few papers in the reviewed literature frame transit modes as the 'car-free' option and compare transit to car access.
This framing is notable as car travel can be seen as a direct competitor to transit or benchmark for travel times and accessibility levels [@golubUsingPrinciplesJustice2014; @martensJusticetheoreticApproachDistribution2012].
As an example, @warrenDevelopingEquitableSustainable2015 develops a goal for per capita car ownership for the developing economy of Havana (Cuba), in recognition that car mobility is needed to alleviate transportation disadvantage in the short-term where public transit is not yet sufficiently addressed.
In this context, this paper acknowledges the tension between household vulnerability and their need for mobility against sustainability goals of GHG emission reduction and car-dependency cycles.
However, not all papers see transit as a direct competitor, but as a mode that can be used to satisfy individual capabilities.
For instance, @smithAccessibilityCapabilityMinimum2012 focuses on households who live in rural areas in the UK and they synthesis their perspectives on their minimum transport needs and costs based on perceived minimum living standards for types of households (e.g., retired, no-children, with children, single, etc.).
The papers reviewed vary in the importance they place on climate urgency, with some focusing more on satisfying *all* sufficient individual needs while planning for less car-dependent cities in the future.

Following a focus on transit, a focus on pedestrian modes is the second most common mode-focus in the literature.
In the papers that focus exclusively on walking, many use or develop walkability scores to explore neighborhood perceptions [@evansAccessibilityUserNeeds2015] or pedestrian mobility focusing on middle-aged and older adults [@towneUsingWalkScore2016], gender [@kimNeighborhoodEnvironmentWalkability2016], or urban peripheries [@blecicCapabilitywiseWalkabilityEvaluation2021].
These papers use 'walkability' as a way to measure the equity in its distribution.
Other papers use walkability as an indicator for public health and urban vitality [@sungResidentialBuiltEnvironment2015; @mccormackRelationshipClusteranalysisDerived2012].

Additionally, papers that focus on the pedestrian mode also often focus on multiple modes: they often discuss 'walkability' as part of active transportation, which focuses on both walking and bicycle and/or transit.
Conceptualizations include how active transport contributes to children's physical activity levels [@mammenSchoolTravelPlanning2014], walkability as an alternative to car predominance [@bertrandMeasuringMappingDisparities2008] or tension that exists between modes, creating unsafe conditions for walking [@siuAssessmentPhysicalEnvironment2019; @ferenchakEquityAnalysisProactively2019].

In terms of car-mode focuses, cars are infrequently the only mode within a paper that is examined.
When car-mode is studied, it is often used as a comparison with transit or as the only mode of transport for areas with sub-standard transit systems e.g., [@kimmelStructuralBarriersComprehensive2018; @aljoufie2016].
Similarly, car studies focus on externalities such as air pollution and safety [@fengTradeoffsMobilityEquity2014; @houstonProximityLicensedChild2006].

```{r calculating-destinations-of-interest}
dest_count <- strsplit(Consensus_only_analysis$`_activity`, "; ",
                       fixed = T)

dest_count <- data.frame(
  id=rep(Consensus_only_analysis$`_activity`, lengths(dest_count)),
  dest_count=unlist(dest_count))

dest_count <- dest_count |> group_by(dest_count) |> summarize(count = n()) 
```

## Destinations (the "Who" pt.3)

The majority of the papers do not focus on any particular destination (e.g., `r (dest_count$count[8]/sum(dest_count$count)) |> percent()` of studies).
Within these papers, a variety of equity dimensions and modes are examined.
Typically, they are multi-modal and focus on either dimensions that impact the **trip itself** (e.g., the trip experience, the quality of infrastructure, aspects of level of service) or the **people and relevant destinations that can be accessed** (e.g., a bundle of trips made for specific population groups, enough for 'sufficient' quality of life).
For the first, the focus is on the quality of infrastructure, safety issues, perceived accessibility and dimensions of the level of service such as frequency [@zheEvaluationSharedUse2008; @prasertsubpakijEvaluatingAccessibilityBangkok2012; @furstBestBadPractices2013 ; @lattmanPerceivedAccessibilityPublic2016].
For the second, the focus is a bundle of trips made for specific population groups, such as people who are physically impaired [@wilkinson-meyersLiveOrdinaryLife2015] or women [@russellPedallingEquityExploring2021] or broadly what is enough for 'sufficient' quality of life [@churchillTransportPovertySubjective2019].
These papers further demonstrate the multi-dimensional role of transportation systems: they provide a utilitarian service that can be used to get from A to B but they too are experienced by the people that use them.
The papers that examine 'all trips' best exemplify this trend in the transportation equity literature.

In terms of the papers that study specific destinations, the most commonly studied activity types includes healthcare services (`r (dest_count$count[6]/sum(dest_count$count)) |> percent()`) followed by employment destinations (`r (dest_count$count[5]/sum(dest_count$count)) |> percent()`).
Papers that exclusively focus on healthcare typically originate from the healthcare planning literature, and look to inform planners where disparities in services exist and what can be done about it.
For instance, @wangModelingSpatialAccessibility2011 models access of the Chinese-language speaking population in Toronto (Canada) to Mandarin-speaking family physicians and suggests (inferred) that a spatial mismatch in the supply and demand is not equitable.
Papers that exclusively focus on employment typically focus on these trips as they are the most common trip purpose and are often correlated with other trip activities like shops, recreation, and other services, generally speaking.
For instance, @allenSizingTransportPoverty2019a operationalizes a low employment-based accessibility threshold and a composite population vulnerability index to identify neighborhoods in transport poverty for eight cities in Canada.
Papers focused on healthcare and employment typically source data from representative travel surveys/diaries, census data, and point-of-interest databases: they often rely on well developed and institutional data that represents 'typical trips', especially in the Global North where this data is more readily available.

But what about non-healthcare and non-employment activity types?
Papers that focus on other activities are not framed as a 'typical travel pattern' and they have different intentions.
For instance, papers that focus on places for shopping such as grocery stores or markets (`r (dest_count$count[7]/sum(dest_count$count)) |> percent()`), often aim to identifying food deserts e.g., [@choiFoodDesertsActivity2013; @jiaoHowIdentifyFood2012; @mckeyCrowdsourcedMappingHealthy2020; @kimAssessingSocialSpatial2020].
Papers that focus on educational facilities including primary school, secondary school, and post-secondary school represent `r (dest_count$count[3]/sum(dest_count$count)) |> percent()` of the studies, and examine children's active transportation to school e.g., [@laszkiewiczChildrenGreenWalk2020] and universal design e.g., [@larkinsAccessibleTransportationBuilt2011].
When green space or other places of leisure is the exclusive focus (`r (dest_count$count[6]/sum(dest_count$count)) |> percent()` of papers), studies examine different accessibility questions such as the spatial distribution of green space e.g., [@xuSocialInequalitiesPark2017], for whom its accessible to e.g., [@mavoaAreaLevelDisparitiesPublic2015], and why e.g., [@mearsUnderstandingSocioeconomicEquity2019].
Very few reviewed papers include 'community' destinations (including public service centres, places of community support, and places of worship (`r (dest_count$count[2]/sum(dest_count$count)) |> percent()` of studies) or childcare activity types (`r (dest_count$count[1]/sum(dest_count$count)) |> percent()` of studies). In the few papers that do include them, these destinations are considered in a holistic representation of activity participation [@albertsRebuildingWomenLivelihoods2016; @smithAccessibilityCapabilityMinimum2012].

Conclusions and recommendations from the papers are varied: they span across population groups, modes, and activity types -- in addition to across equity dimensions, conceptualizations, as we exemplify below.

## Methods used across equity conceptualizations and standards (the "How")

```{r calculating-concepts-and-standards}
conc_count <- strsplit(Consensus_only_analysis$`_equity conceptualization`, "; ",
                       fixed = T)
conc_count <- data.frame(
  id=rep(Consensus_only_analysis$`_equity conceptualization`, lengths(conc_count)),
  conc_count=unlist(conc_count))
conc_count <- conc_count |> group_by(conc_count) |> summarize(count = n())

stand_count <- strsplit(Consensus_only_analysis$`_type of equity standard2`, "; ",
                       fixed = T)
stand_count <- data.frame(
  id=rep(Consensus_only_analysis$`_type of equity standard2`, lengths(stand_count)),
  stand_count=unlist(stand_count))
stand_count <- stand_count |> group_by(stand_count) |> summarize(count = n()) 
```

```{r plot-standards-concs-perc, warning=FALSE}
#| label: fig-fig6
#| fig-cap: "The proportion of equity standards (vertical axis) within each type of equity conceptualization (horizontal axis) category."
standards_conc_percentages_plot_and_table
ggsave("standards_conc_percentages_plot_and_table.png")
```

`r word_comment("Answering: what conceptualizations occur in papers with opportunity standards? what types of methods?", ".")` Broadly summarizing how equity standards connect with conceptualizations, some trends emerge in methods used. **Opportunity** and **population** standards appear in the literature at similar proportions ( `r (stand_count$count[3] /sum(stand_count$count)) |> percent(accuracy=0.1)` vs. `r (stand_count$count[4] /sum(stand_count$count)) |> percent(accuracy=0.1)`), but frequently correspond to different conceptualizations. Referring to @fig-fig6, over 40% of papers that focus on **horizontal equity** and **spatial equity** conceptualizations suggest **opportunity** standards. Within these papers, travel impedance standards (a type of opportunity standard) are often suggested and accessibility indices are operationalized (i.e., a measure of the potential to interact with opportunities for populations located at each spatial unit within a region). Examples include: @chenImpactHighspeedRail2017 use a travel time threshold of 4 hours or less on high-speed rail from one municipality to another to be considered "comfortably connected", @yenisettyMeasuringAccessibilityVarious2020 assumes that populations living in areas where the travel distance to a public transit station is less than 1,200m is sufficient to interact with the transit system, and @shenMeasuringSpatialAccessibility2020 identifies regions where populations cannot access hospitals within 1 hour by car. Papers suggesting opportunity standards often employ disparity analysis through a variety of quantitative approaches such as inequality measures (e.g., Gini coefficient and Lorenz curves, poverty measures [@vanderveenOperationalizingIndicatorSufficient2020; @tiznadoaitkenRoleAccessibilityPublic2018]), spatial descriptive analysis, and the comparison to benchmarks (e.g., equal supply to demand of public transit in a spatial unit as done by @peungnumsaiGridBasedSpatialAnalysis2020) to determine which locations are spatially and horizontally (in)equitable relative to other locations in the studied region. Another branch of quantitative research conceptualize transportation system externalities as trade-offs, maximizing transport-related benefits (i.e., time savings, emissions reductions, congestion reductions, user fares) through optimization/location-allocation methodologies e.g., [@fengMulticriteriaEvaluationAccessibilitybased2014; @fakhrmoosaviIncorporatingTravelTime2021; @zhengAreabasedEquitablePricing2020; @wismadiSpatialPreferenceModelling2014]. These papers that focus primarily on **horizontal equity** and/or **spatial equity** seldom use exclusively qualitative methods.

`r word_comment("Answering: what conceptualizations occur in papers with **population** standards? what types of methods?", ".")` Whereas, papers that propose  **population** standards frequently utilize other methods.
Over 40% of papers that suggest **population** standards focus on **well-being** conceptualizations.
These papers typically ask *what is enough to lead a satisfactory life (as related to transportation)*, and the standards that are suggested include population benchmarks for comparison such as: questionnaires and relative comparisons to physical activity per week recommendations [@adlakhaMindGapGender2020; @auchinclossDesignBaselineDescription2020; @mccormackRelationshipClusteranalysisDerived2012; @kimNeighborhoodEnvironmentWalkability2016; @towneUsingWalkScore2016], summative per capita benchmarks (e.g., energy consumption for a 'decent living' is suggested in [@raoDecentLivingEmissions2012], and region-relative comparisons in health-related outcomes e.g., premature mortality rates [@awuorInfluenceEnvironmentalHealth2019], spatial access to hospitals [@pereiraGeographicAccessCOVID192021], spatial access to supermarkets, active-mode-usage, and Body Mass Index (BMI) [@murphySupermarketAccessTransport2017].
The majority of these papers use quantitative/mixed-methods to identify inequities in **wellbeing**, however, a minority do use exclusively qualitative methods to distill themes e.g. the exploration of *perceived* quality of life in [@berheAdaptationDissonanceQuality2014].

`r word_comment("Answering: what conceptualizations occur in papers that conceptualize BOTH population and opportunity standards? what types of methods?", ".")` Papers that conceptualize *both* **population** and **opportunity** standards often conceptualize **vertical equity** and **transport-related social exclusion** (note the similar proportions in these standards in @fig-fig6).
Expectedly, these papers often include a combination of methods: Questionnaires and other qualitative methods related to population standards and quantitative methods such as accessibility indices for opportunity standards are usually deployed.
For instance, census data and the estimated proportion of households within some travel distance/time/availability to/of key destinations is used to identify a variety of social exclusions e.g., [@mackettIncreasingAccessibilityCostEffectively2010; @chenExploringTravelCharacteristics2010; @danielsProposalAccessibilityPlanning2011; @sunPublicTransportAvailability2021; @sharmaPublicTransitAccessibility2021], transport-related social exclusion's link to wellbeing e.g., [@delboscTransportProblemsThat2011; @churchillTransportPovertySubjective2019], areas more likely to experience transport poverty [@allenSizingTransportPoverty2019a], food deserts [@mckeyCrowdsourcedMappingHealthy2020], or transport-related energy poverty [@robinsonDoubleEnergyVulnerability2020; @berryInvestigatingFuelPoverty2016; @berryDistributionalEffectsCarbon2019].

Similar to papers that conceptualize **wellbeing**, the majority of **social-exclusion**-conceptualizing papers use quantitative/mixed-methods to identify areas, households, and/or populations at risk. They use a variety of methods to identify _where_ populations at risk may be located, such as clustering methods [@mohriClusteringMethodMeasuring2021]. The minority of papers that employs exclusively qualitative methods use surveys to understand population travel willingness/barriers e.g., [@chenExploringTravelCharacteristics2010; @mehdizadehWalkingTimeSchool2017] and interviews/focus groups related to topics of unmet activity needs [@nordbakkeTransportUnmetActivity2015].

`r word_comment("Answering: what conceptualizations occur in papers that conceptualize infrastructure standards? what types of methods?", ".")` The largest group of papers that suggest **Infrastructure** standards conceptualize **Rights**. Recall, **Infrastructure**-standards suggesting papers represents `r (stand_count$count[3]/sum(stand_count$count)) |> percent()` of all papers and **Rights-**conceptualization is represented two times more than any other conceptualization type. These papers often focus on populations with mobility impairments and non-car users' inequities in their ability to access the transportation systems. Methods vary, but focus equally on audits of existing infrastructure relative to best-practice universal design principles [@odeckEconomicAppraisalUniversal2010; @larkinsAccessibleTransportationBuilt2011; @jimenez-espadaResearchProblemUniversal2021; @perezdelhoyoParticipatoryManagementImprove2021] and qualitative methods that interview/survey users about their perceived access to transport systems e.g., [@marquezFactorsAffectingPersonal2019; @mateo-babianoBicycleSharingAsia2017; @furstBestBadPractices2013; @velhoEffectTransportAccessibility2016; @parkJourneyVisuallyImpaired2017; @limFacilitatingIndependentCommuting2021; @stjernborgAccessibilityAllPublic2019] and experiment the suitability of existing best practice standards e.g., [@daamenAssessingGapPublic2008; @velhoEffectTransportAccessibility2016; @bharathyRevisitingClearFloor2018].

More broadly, papers that suggest **Infrastructure** standards are often multi-dimensional, and extend beyond **Rights** conceptualizations. They often suggest **Opportunity** and **Population** standards as well and often apply **Vertical**, **Horizontal** and **Spatial** equity lenses as well. These multi-dimensional papers can refer to established guidelines and suggest composite indices e.g., @racheleNeighbourhoodSocioeconomicTransport2017a combines transport network properties such as street connectivity, cul-de-sac length, street block length, traffic volume, public transport stops and service frequency inputs to define an indicator of transport design that is supportive of walkability and access to public transport. Other works include assessing the quality of infrastructure [@xuSocialInequalitiesPark2017], the severity and frequency of accidents on the system  [@benevenutoExaminingTransportNeeds2020; @appleyardTransitCorridorLivability2017], user-groups (particularly disadvantaged groups in the case of vertical equity conceptualizations) [@prasertsubpakijEvaluatingAccessibilityBangkok2012] as part of the multi-criteria indicators. Another branch of literature explicitly focuses on affordability or other barriers to the transport system, and suggests improvements to the infrastructure such that all groups (especially the most disadvantaged) can sufficiently interact with the system e.g., [@basuPracticalFrameworkBenchmarking2019; @songWomenWheelsGender2019; @welchEquityTransportDistribution2013], dabbling into conceptualizations of **transport-related social exclusion** e.g., [@kentPrioritizingLowstressEquitable2019] and **sufficientarian/capabilities** conceptualizations e.g., [@smithAccessibilityCapabilityMinimum2012].

`r word_comment("Answering: what conceptualizations occur in papers that conceptualize Environmental+ standards? what types of methods?", ".")` Though papers that conceptualize **Environmental +** standards are not common in the literature review (`r (stand_count$count[1]/sum(stand_count$count)) |> percent()` of all papers), they most frequently occur in papers focused on **Inequitable externalities**. These papers often use traffic-related air pollution, noise pollution, green-space, urban design elements, urban air temperature, health related outcome, and physical activity guidelines to quantify transport-related externalities. Methods used are almost all quantitative or mixed-methods, and the identification in inequalities is spatial clustering, the use of Gini coefficient [@fengMulticriteriaEvaluationAccessibilitybased2014], comparisons to established environmental thresholds or health guidelines [@agostfelipInclusiveModelAssessing2021; @agostfelipInclusiveModelAssessing2021; @kruizeEnvironmentalEquityRole2007; @iungmanImpactUrbanTransport2021; @apparicioCyclingOneMost2021; @khomenkoLiveableCityHealthy2020; @muellerSocioeconomicInequalitiesUrban2018], creation of composite multi-dimensional indices [@agostfelipInclusiveModelAssessing2021; @mirandaBenchmarkingSustainableUrban2012; @corazzaMethodologyEvidenceCase2020], and/or in addition to spatial analysis [@jephcoteGeospatialAnalysisNaturally2013; @carrierApplicationThreeMethods2014].

See Table XX for detailed examples of papers conceptualizations and standards across each equity dimension.

# Discussion and ways forward (identifying gaps and possible agenda for the future`r word_comment("This section is a work in progress, but we outlined some ideas. This will be completed after the team's first read using ideas generated.", ")")`

We discuss trends in how the literature has conceptualized and used standards (the "How") to evaluate the "What" of transportation equity, the equity dimensions. These trends are context-specific: they concern the "When" and "Where" (differences across publication year, across case studies continent, and between urban and rural contexts) and the "Who" (differences in the focus on the population sub-groups, transport modes, destinations of interest) as discussed in the preceding sections following the four equity dimensions that structure this review (mobility/accessibility, traffic-related pollution, human-health related, and traffic safety).

A summary of what we found in the literature:

- the "When" and "Where": 
  - most studies focus on case studies in north american and europe
  - south american/africa papers have key differences in focus compared to global north: they are more focused on affordability-as-a-barrier, tensions in infrastructure dev vs. service prioritization, data availability limitations
  - 
- the "Who":
  - low income most popular
   - (dis)abilities are exclusive focused population category in many studies; other 'vulnerabilities' are typically composite (aside from income, which is part of both exclusive-focus and composite vulnerabilities)
  - transit, with cycling and walking becoming more popular
  - cars are seldom the focus, and if they are, it is as a comparison to transit and/or focus on their negative externalities
  - destinations of interest are mostly employment and healthcare; other destinations are less studied. The many 'destination' however, is the trip itself -- the experience, the infrastrucutre, the mode-service, a bundle of trips taken a month to all locations, etc. 
- the "What" and "How":
  - accessibility/mobility -> opportunity standards are typically travel impedance standards, and affordability thresholds (e.g., more than 10% of monthly income should not be spent on transport [@rivasHowAffordableTransportation2018] to work and/or other necessary destinations). Oftentimes spatial equity, disparity analysis is operationalized. Infrastructure standards are also suggested -> oftentime Rights conceptualizations, i.e. access for all.
  - transport-related environmental externalities -> pollution/noise standards are suggested, often using WHO guidelines or local neighbourhood averages as a comparision.
  - health  -> health standards, the reliance on population transport health literature such as physical activity guideline recommendation per week or excess mortality burden. 
  - safety -> the use of engineering standards and best practice
  - cross-dimensional -> measures such as urban livability, etc. a more holistic conceptualization of transport equity

## The need for explicit conceptualizations and grounded standards

Some conceptualizations are implicit. Some standards seem arbitrary. Despite the need for and importance of setting more standards, some proposals seem arbitrary.
For example, XXXX proposes 20 ferries per day to avoid social exclusion.
Justifications for this number is lacking.
It is unclear if 10 or 30 ferries would make a difference in a specific quality-of-life outcome or if that number is tied to funding/resource constraints. We need standards linked to grounded outcomes.

Relative measures are used: context, temporal, and group-specific. E.g. a certain group should have more access in certain situations. Context-matters, but this presents a challenge in policy planning.

Equity is not justice. Some studies conflate the two. Justice should not be mixed up with equity. More justice work is needed, e.g., Restorative justice -- repairing harm is missing from the discussion

## Methods and data limitations

On the methodological side, more mixed methods are needed in transport equity research.
Conceptualizations and standards are usually discussed from purely qualitative or quantitative approaches, a missed opportunity to combine the strengths of both approaches, whether by deep diving into some particular experiences or perceptions through qualitative methods or tailoring more meaningful quantitative analysis after qualitative explorations. For instance, @xieWonCycleRoute2018 finds through interviews and go-alongs with women cyclists that the standard Cycling Level of Service (CLS) tools used by engineers to plan cycling infrastructure misses the critical gendered perspective. Further, @somenahalliAgingTransportMobility2007 surveys older adults to understand their mobility issues, revealing factors that are unseen in standard daily travel surveys.

Furthermore, this is plenty of disparity analysis, without engaging explicitly with equity conceptualizations. For example, within the mobility/accessibility dimension, metrics of accessibility (usually 15 to 60 minutes) are used to show differences among areas and groups but with scarce policy and practice implications of those results.
Aiming at specific goals and standards tied to conceptualizations is the ideal case.
When these analyses engage with metrics that may be tied to conceptualizations (like Gini coefficient or Theil index), they fall short in assessing if the result's good or bad e.g., [@mijaresEquityAnalysisUrban2013].
If a Gini coefficient of 0 means that all people have the same access to public transport stops, what does it mean a 0.2, 0.3, or 0.4? Is this good or bad news for decision-makers? Are new policies needed to reduce that number to a certain threshold, orienting future interventions? These questions usually remain unanswered despite its importance. These measurements can also bring some challenges and pitfalls, as recently summarized by @karnerAdvancesPitfallsMeasuring2023.

What are the sources of data and what are the motivations for some categories?
POI databases typically include education, health and aggregated categories for leisure and community.
Within 'community' are organizations, government services, visiting friends/family -- very broad, grandma's house is not there, social networks rarely incorporated.
Childcare, typically daycare or facilities -- domestic work, mobilities of care, mobility interdependence, are unrepresented relative to the presence of work destinations. Transport systems' focus is more than just to work or as a source of economic development, though in underdeveloped regions, transport systems as a force of economic development e.g., high-speed rail [@chenImpactHighspeedRail2017; @monzonEfficiencySpatialEquity2013; @kimImpactsHighspeedRail2015].

Data availability matters, especially when operationalizing emerging theories (i.e., sufficientariansm [@vanderveenOperationalizingIndicatorSufficient2020]):

- Leisure destinations (e.g., green space, parks, recreation) are less studied in this context.
- Some categories are missing all together -- mobilities of care.
- Issue of data availability

## Evaluate interventions and policies

There is a need to evaluate more equity interventions or policies.
In our review, only 19 out of 155 studies assess specific projects with an equity lens <!-- ignacio says CHECK THIS, make sure that this is the correct number-->.
Examples include mode-shift from driving to active school travel [@mammenSchoolTravelPlanning2014], transit fare restructures [@hickeyUsingQuantitativeMethods2010] and spatial analysis of Low Traffic Neighborhoods [@aldredEquityNewActive2021].
This is a key step towards transport justice; assessing the effects of policies on different dimensions and populations groups and evaluating if a specific context is moving towards equitable standards over time.

## Links between standards and outcomes

A more robust assessment of the implications of equity standards on life outcomes is still pending.
Estimating the benefits of increased mobility or accessibility, or reducing affordability burdens and transport externalities needs to be associated with outcomes like life and neighborhood satisfaction, subjective well-being, mental and physical health, social capital, among others.

# Concluding remark`r word_comment("This section is a work in progress. This will be completed after the team's first read using ideas generated.", "s")`

-   TBD

# Appendix

The following are additional details.

#### The search strategy:

```{r Search-query-A1}
#| label: fig-A1
#| fig-cap: "The search query. TS = topic search (keywords, abstract, title). TASCA = subject categories. Green text area transportation system related terms, blue text are equity dimension related terms, purple text are equity/justice conceptualization related terms, and orange text are standards related terms. Hits corresponds the number of papers that the search yielded and was retained into the evidence selection process."
knitr::include_graphics("Search-query.png")
```

Definitions of the population-concept context (PCC) used in the creation of the inclusion and exclusion criteria for the search strategy.

-   **Population**: the focus of the included studies should be on individuals, groups, communities, or entire regional areas that are impacted by passenger transportation infrastructure and systems (i.e., all modes and flows) from the perspective of equity (i.e., fair distribution, production, and re-production of burdens and benefits). This criteria is reflected in the creation of the first set of topic search terms that relate to transportation modes (e.g., "walking" OR "cycling" OR "transit" - see green text in @fig-A1 for the full list).
-   **Concept**: the included studies should also include equity dimensions and conceptualizes equity as discussed in the previous section. This inclusion criteria is reflected in the second and third set of topic search terms developed in the search strategy. These terms relate to types of equity dimensions (e.g., "accessibility" OR "mobility" or "transport-related air pollution" - see blue text in the @fig-A1 for the full list) and equity conceptualizations (e.g., "Justice" OR "equity" - see purple text in @fig-A1 for the full list).
-   **Context**: the included studies should also be limited to publications that include equity standards. Context can be more difficult to explicitly search for with key terms so synonyms for 'standards' were added to the query as a four set of topic search terms (e.g., threshold, indicator, criteria - see orange text in @fig-A1 for full list). Additionally, journal article and conference papers, English-language literature from any country, any study design (e.g., quantitative, qualitative, or mixed-method studies, or conceptual frameworks), and any record published within the past 30 years are included (January 1992 to March 2022). The time period is selected as the first (to the authors knowledge) peer-reviewed article which operationalized equity standards and equity conceptualization was published in 1996 [@khistyOperationalizingConceptsEquity1996]; we are broadening the search by a few years for completeness. English is selected as it is the common language spoken across the authorship team. Furthermore, papers that explicitly fall within the Transportation or related topic/category is included in the query (e.g., "Transportation", "Social Sciences", "Geography", "Civil Engineering", "Philosophy" - see the @fig-A1 for full query).

The **exclusion criteria** for the search are papers that are not within the inclusion criteria.
Specifically:

-   Literature published before January 1992.
-   Papers which do not include transportation equity dimensions.
-   Grey as concepts contained within are frequently published in a more developed form in journals.

#### Example of the data extraction template:

```{r data-extraction-template-A2}
#| label: fig-A2
#| fig-cap: "The data extraction template with associated defintions."
knitr::include_graphics("Data-extract-template.png")
```

### Sample of select papers summarized by their, equity dimension, case study continent, conceptualizations and standards

```{r}
dat <- data.frame(matrix(c("Equity dimension 1: **mobility and accessibilty**",
                           "[@rivasHowAffordableTransportation2018] - South America (select cities)",
                           "Analyses how affordable urban public transportation is in select Latin American and Caribbean countries. They look at the estimated average monthly cost of transit trips and average monthly household income and conceptualize **transportation-related** **affordability**, especially for the most economically vulnerable (**vertical equity**).",
                           "**Population standards**: The financial burden of a basket of urban public transportation trips (60 trip fares, representing 30 round-trips per month) should not exceed 10% of household monthly income.",
                           "Equity dimension 1: **mobility and accessibilty**",
                           "[@bharathyRevisitingClearFloor2018] - North America (USA - National)",
                           "This study designed a web-based tool and took a representative sample of wheeled mobility device (WhMD) users anthropometry measurements to determine if the minimum standard suggested by the ADA is sufficient. We understand this conceptualization as a type of **Rights** conceptualization that WhMD should have minimum clear floor space (as described the guidelines in line with the American Disabilities Act) to access bus shelters, bus stop pads, and transit terminals.",
                           "**Infrastructure standard**: The clear floor area for wheelchairs: 760 mm (30 in.) wide by 1220 mm (48 in.) in length as described by the ADA standards. Of note, this minimum clear floor area is insufficient for a variety of the WhMD users.",
                           "Equity dimension 1: **mobility and accessibilty**",
                           "[@ryanWhatAreWe2021] - Europe (Stockholm, Gothenburg and Malmo cities in Sweden)",
                           "Investigates what the literature and planning process is missing when we measure accessibility by comparing objective and self-reported accounts of accessibility among older people. This paper conceptualizes accessibility as from the position of the **capabilities approach** and **vertical equity** (particularly acknowledging that older people have capabilities that differ from the general population).",
                           "**Population** and **Opportunity standards**. Specifically for older populations (aged 65+), the following travel distances are suggested as equitable trip lengths to grocery stores per mode: Walking: less than or equal to 1500m, Combined public transit and walking (less than or equal to 1000m (walking element)), Combined car and walking: less than or equal to 1000m less than or equal to 1000m (walking element)), Bicycle: less than or equal to 3000m in addition to travel time threshold of less than 15 mins.",
                           "Equity dimension 1: **mobility and accessibilty**",
                           "[@wismadiSpatialPreferenceModelling2014] - Asia (Yogyakarta, Indonesia)",
                           "Explores the equitable provision of transport infrastructure provision: an application of Sen's **capability approach**. Conceptualizes equity through Sen's capability approach and spatial equity.",
                           "Opportunity standards: Areas below the relative poverty line (of its neighbours) can only be located transport resources (i.e., measure in person*kms that can be travelled at car speed, i.e., mobility) based on the following 2 benchmarks (they can be considered, together as the floor/minmum access): 1) Global: standard deviation (SD) distance to mean should be minimized. 2) Local: priority to minimise the differences with its neighbourhood",
                           "Equity dimension 1: **mobility and accessibilty**",
                           "[@zhengAreabasedEquitablePricing2020] - North America",
                           "This paper conceptualizies equity in the multimodal network (transit, car) being fair toll-pricing across differences in populatins value of time (VOT). VOT is determined based on household income, with lower income households having lower VOT and thus deserving of lower tolls (vertical equity). From this perspective, a utilitarian perspective that seeks to minimize multimodal traffic congestion through introducing toll-pricing based on VOT is implemented.",
                           "Infrastructure standard: suggest that a toll-pricing scheme based on individuals travel value-of-time (lower income people have a lower VOT) is equitable.",

                           "Equity dimension 2: **environmental pollution**",
                           "[@carrierApplicationThreeMethods2014] - North America (Montreal, Canada) ",
                           "This work examines the statistical association between different social groups and the concentration of air pollutants. They frame their work from the perspective of environmental equity. We interpret the conceptualizations to be along the lines of **inequitable externalities**, **spatial** and **vertical equity** - transport-related air pollution is a product of road transport and it impacts the air of residents in unequal spatial ways. The paper then frames this impact as unfair, particularly from the perspective of disproportionately disadvantaged residents",
                           "**Infrastructure standards**: The literature suggests that the health implications from the transport-related air pollution from major roadways is most acute at residential distance locations of 200 m or less. Residential locations should not be located within this distance threshold from the perspective of human health. **Environmental+** and **Population standards**: Uses the WHO NO² threshold as a point of comparison (annual concentrations of NO² should not exceed 40 μg/m-3). They argue that even through no neighbourhood, even those disproportionately low income, exceed the WHO limit in this case study, they still suggest that air pollution should not be disproportionately impacting disadvantaged neighbourhoods. It can be interpreted that they use the WHO threshold as a minimum threshold and suggest that air pollution levels should not be impacting disadvantaged populations disproportionately ( a relative population standard)",
                           "Equity dimension 2: **environmental pollution**",
                           "[@jephcoteGeospatialAnalysisNaturally2013] - Europe (Leicester, UK)",
                           "Geospatial analysis of naturally occurring boundaries in road-transport emissions and childrens respiratory health across a demographically diverse cityscape. Emperically identifies at what distance away from major roadways children are most impacted by transport-related pollution. This is framed in the perspective of children’s **well-being**. Children are at most risk for acute respiratory distress from elevated levels of air pollution, and as such planning should consider this point of public health.",
                           "**Infrastructure standard**: Finds that children (most vulnerable to air pollution - related to motoized traffic) are most impacted by air pollution within 283 m of a road way. This should be the distance threshold that schools and other childrens facilities are located.",
                           "Equity dimension 3: **health**",
                           "[@adlakhaMindGapGender2020] - Asia (Chennai, India)",
                           "From the perspective of disparity in gendered physical activity, this paper focuses on women's cycling as both transport and exercise. They advocate for all people achieving physical activity thresholds (**horizontal equity**) but prioritize women and especially women in neighborhoods with low-walkability and socio-economic status (**vertical equity**).",
                           "**Population standards**: All people should get 150 min of moderate activity a week or 75 min of vigorous physical activity per week.",
                           "Equity dimension 3: **health**",
                           "[@savingmothersDidSavingMothers2019] - Africa (Select urban and rural regions in Uganda)",
                           "The **well-being** of mothers, this paper examines the timely access to emergency obsteric and newborn care for child-bearing aged women in Uganda.",
                           "**Opportunity standard**: 2 hours to the nearest facility with surgical capacity with anesthesia services - this threshold is determined through the onset of bleeding to death if a women with obstetric hemorrhage does not receive adequate treatment).",
                           "Equity dimension 3: **health**",
                           "[@iungmanImpactUrbanTransport2021] - Europe (Madrid and Barcelona, Spain)",
                           "They use environmental pollution guidelines, but from the position of health. They investigate the impact of urban and transport planning on attributable mortality burden in Madrid and Barcelona and its distribution by socioeconomic status . Pre-mature mortality is linked to the exposure to pollution and motorized vehicles (**inequitable externalities**). These externalities should not be impacting people disproportionately (**vertical equity**) and should be even across space (**spatial equity**).",
                           "**Environmental+ standards**: All minimum thresholds, if exceeded this is inequitable: NO² concentration 40 μg/m³; PM 2.5 concentration 10 μg/m³; Noise 53dB for average 24 hours; Living with 300 m crow-flies distance from at least .5 hectares of greenspace; and a Change of air temperature of at least 1 ⁰C. ",
                           "Equity dimension 3: **health**",
                           "[@mehdizadehWalkingTimeSchool2017] - Asia (Rasht, Iran)",
                           "From the perspective of children’s **well-being**, assesses the walking time to school. They frame walking to school as health-related.",
                           "**Opportunity** and **Population standards**: perceived walking time to school for students aged 7-9 yrs is 10 mins, and the longer the PWTS the less likely they were to use an active mode to travel to school.",
                           "Equity dimension 3: **health**",
                           "[@murphySupermarketAccessTransport2017] - Oceania (Melbourne, Australia)",
                           "Assesses the relationship between supermarket access and transport mode used, the body mass index (BMI) of the mode-user (**wellbeing**) and the equity in access distribution by income (**vertical equity**).",
                           "**Opportunity** and **Population standards**: all households should be sufficiently active (greater than 150 min and at least 5 sessions) and households should be within 1 km euclidean distance to supermarket (80-90% of the dwellings should meet this). Planners should prioritize socially disadvantaged areas to meeting these standards first.",

                            "Equity dimension 4: **transport-related safety**",
                           "[@ferenchakEquityAnalysisProactively2019] - North America (Denver, USA)",
                           "Operationalizes and compares an equity analysis of proactively- and reactively-identified traffic safety issues from the perspective of **Spatial equity**, **Vertical equity** and **Inequitable exposure to externalities**.",
                           "**Infrastructure** and **Population standards**: standards are suggested for both reactive and proactive analysis. First, the lower the number of collisions on the road with pedestrians/cyclists (i.e., reactive safety analysis), the better. No/minimal inequalities for general population vs. equity seeking groups (high proportion of POC and/or low income in tract). Second, the lower the perceived safety, the better (i.e., if travel to school by ped. or bike is unsafe due to traffic conditions). No/minimal inequalities for general population vs. equity seeking groups (high proportion of POC and/or low income in tract).",
                           "Equity dimension 4: **transport-related safety**",
                           "[@zheEvaluationSharedUse2008] - Asia (Tokyo, Takamatsu, and Tokushima",
                           "Evaluates the observed safety of  shared use pedestrian and bicycle paths from the perspective of **well-being**.",
                           "**Infrastructure standard**: the study suggests that the safety threshold for bicycles and pedestrians to coexist on shared infrastructure is less than 0.5 pedestrians/minute per metre of sidewalk (width) and less than 3.0 cyclists/minute per metre of sidewalk (width). The standard for pedestrian/bicycle share use in terms of hourly traffic volume is less than 26 pedestrians / hour and 108 cyclists / hour for 2m wide sidewalks.",

                           "**Multi-equity dimensions**",
                           "[@aldertonWhatMeaningUrban2019] - Asia (Bangkok, Thailand) – **Mobility/ accessibility** and **health**",
                           "Establishes short-, medium-, and long-term goals for the city in collaboration with technical leaders within the municipal government for the perspective of **well-being** (urban livability): the standards included in this table relate directly to transportation systems. Indicators are inspired by the Sustainable Development Goals (SDGs) as well other global planning standards.",
                           "**Opportunity standards** are suggested: 1) Green space: % of residents living < 400 m from public open space, a large park (> 1.5ha), and/or local park, 2) Public transit access: % of residents living < 400 m of a local bus stop and <800 m of train station, 3) Facilities: % of residents living < 400 m of a community centre. The following **Infrastructure standard** is suggested: Canal water quality - dissolved oxygen content of equal to or less than 2.0 mL/L",
                           "**Multi-equity dimensions**",
                           "[@berheAdaptationDissonanceQuality2014] - Africa (Mekelle, Ethiopia) -**Mobility/ accessibility** and **health**",
                           "Examines adaption and dissonance in the quality of life (QoL) of residents. QoL is conceptualized along the lines of **well-being** and aspects of QoL directly tie into transport systems. They conduct a qualitative QoL survey of residents on the topic of three QoL domains: housing quality, access to important destinations, and affordability. They also measure quantitative indicators associated with these domains. We assume the equity goal for this paper is that subjective and objective QoL measures should not be mismatched: as discussed by the authors of this study, subjective QoL is higher than objective QoL the participant is experiencing adaption and in the reverse scenario the participant is experience dissonance.",
                           "**Opportunity standards**: 1 & 2) Access to primary or secondary education facility, percentage of households living within 1 km or 2km (walking distance), respectively from a primary school or secondary school. 3) Access to health facility, percentage of households within 40 min walking time from a health facility. 4) Access to public transport, percentage of households within a distance of 500 m from a mini-bus stop. **Population standards**: 1) Adequate family income, percentage of households earning more than the official poverty line. 2) Subjective QoL is constructed based on the households level of satisfaction for each of the eight indicators using a six point Likert-scale (1=very satisfied to 6=very dissatisfied).",
                           "**Multi-equity dimensions**",
                           "[@agostfelipInclusiveModelAssessing2021] - Europe (Castellon, Spain) - **Mobility/ accessibility**, **safety**, and **health**",
                           "Conceptualizes equity through age-friendly urban spaces that reduce (and eliminate) conditions for **transportation-related social exclusion** for older populations and prioritize those who are economically vulnerable (**vertical equity**). These guidelines are inspired by the SDGs in addition to planning guidelines used national, regional, and local guidelines used in Spain.",
                           "**Opportunity standards**: 1) Access to facilities needed for old age health. Minimum distance thresholds from the geometric center of neighborhood are suggested: at least: 1000 m from health facilities (600 m or less is preferred), elderly-specific care facilities and shops should be 600 m (300 m or less is preferred). **Population standards**: 1) Certain neighbourhoods should be prioritized above others. From this papers focus on age-friendly urban environment, they suggest that if the neighbourhood has an average old age indicator (i.e., greater than 64 years, and/or greater than 79 years, and/or aging ratio of persons aged greater than 64 relative to 15 to 64 age) should be prioritized. 2) Economic vulnerable and non-civically engaged neighbourhoods should also be prioritized. If the neighbourhood has a lower percentage of civic associations within the neighbourhood than average, and/or household income, and/or a higher than average interventions for dependency and/or social subsidies, they should be priorized. **Infrastructure standards **: 1) Green space: should be at least 10 m2 per inhabitant in the neighbourhood, greater than 15 m2 per inhab. is the goal. 2) As related to sidewalk infrastructure at least 50% of all sidewalks (preferably 75% or greater) should: have a width of 1.5m or larger, ramps should have a grade of 8% or less, be well maintained (free from deficiencies), be paved for pedestrian use, and cover public transit stops. 3) Lighting is critical for traffic-safety and a sense of safety overall. As such, at least 50% roads should: have a min. of 35 lux (road traffic) and 20 lux (pedestrian streets), and adapted traffic lights. 4) Buildings should be age-friendly. As a proxy for the quality of residential living space quality, at least 50% of residential buildings in a neighbourhood should be built within the last 50 years (preferably 75% or more). In terms of physical access into the buildings, at least 10% should have elevators and accessible entrances (preferably 25% or more). **Environment + standards **: 1) Noise at the street level should be less than 55 dB and 45 dB (but preferably less than 50 dB and 40 dB) in the daytime and nighttime, respectively.",
                           "**Multi-equity dimensions**",
                           "[@mateobabianoPedestrianNeedsMatter2016] - Asian (Manila, Philippine) - **Mobility/ accessibility** and **safety**",
                           "The perception of pedestrians’ walking environments should be sufficient across 6 themes. Equity is conceptualized around **spatial equity** (equally fair walking environments for all locations) and **rights** (the right to mobility/accessibility for pedestrians)",
                           "**Infrastructure**, **Opportunity**, and **Population standards**: percieved pedestrian perception on protection, ease, equitable access, mobility, identity, and enjoyment must be met."
                           ), ncol = 4, byrow = T))
```

```{r}
set_flextable_defaults(font.size= 7,
                       font.family = NULL,
                       text.align = 'justify')

flextable(dat) %>%
  ftExtra::colformat_md() %>%
  compose(
    i = 1, j = c(1,2,3,4), part = "header", value = as_paragraph(c("Dimension", "Continent", "Conceptualization", "Standard"))) %>%
  merge_v(j=1) %>% #merge 'dimensions' column
  rotate(part = "body", rotation = "btlr", j=1, i=c(1:5,6:7,8:12, 13:14,15:18)) %>%
  valign(valign="center", part="body", j=1) %>%
  hline( i = c(5,7,12,14), part = "body") %>%
  set_table_properties(layout = "autofit")

```

<!-- we need to sepearte by qual and quant, karen asked-->

# References

### 2023-09-28 Manuscript/Manuscript.qmd
Antonio is reviewing and editing Manuscript.qmd

@@
 # Introduction
+
+On July 20, 2021, at 9:12 a.m.
+(EDT), a rocket took off from a launching pad in West Texas.
+On board was a small group of four passengers that included Jeff Bezos, then the world's richest person [@harwoodBezos2021].
+The flight that morning was one of the first crewed private suborbital flights, and the adventure (described as "intense" by a member of the crew) lasted a total of 10 minutes and 10 seconds [@harwoodBezos2021].
+In addition to intense, it was expensive: a seat for the flight had previously been auctioned for no less than $28 million USD [@griffinPerson2021].
+It was not yet 10:00 am (EDT) after the rocket's return from its excursion to the very edge of atmosphere, when Bezos declared that this was the "[b]est day ever".
+To reporters covering the event, he said "I want to thank every Amazon employee and every Amazon customer, because you guys paid for all of this" [@johnsonRoadToSpace2021].
+
+Meanwhile, back on planet Earth, the employees that Bezos thanked for paying for his suborbital escapade were facing some very mundane problems of their own, and none as lofty as conflicting schedules that prevented them from flying in rockets [@griffinPerson2021].
+According to reports, people employed directly or indirectly by Amazon for warehouse or delivery work had, for years, been treated to "inhumane" conditions [@fungEuropean2018; @scottAmazon2019; @greeneAlgorithm2021], and subjected to surveillance on the job, crushing demands for productivity quotas that led to injury, enjoying little or no job security, and more generally treated as disposable inputs to Amazon's efficiency goals and consumers' desires [@tungAmazon2020; @reeseTeamsters2022; @middletonRobot2023].
+
+Coverage by the mainstream media of the July 20 launch was in many cases uncritical.
+"We're going to build a road to space so that our kids and their kids can build a future," Bezos declared, and added "...we need to do that to solve the problems here on Earth" [@johnsonRoadToSpace2021].
+Few reporters thought to ask questions about what problems the billionaire planned to solve on Earth, or what kind of future Bezos was trying to build, and for whose children.
+A less dispassionate observer might be excused for wondering about the basic fairness of a man amassing a nearly unimaginable fortune that allowed him to build and fly his own spaceship, while masses of his employees were treated as throwaway cogs in the vast apparatus of his logistics empire.
+
+But the question of fairness is not a simple one, and certainly not one amenable to scientific inquiry.
+What constitutes "fair" would likely look very different in democratic societies, compared to societies where no one owns anything—perhaps not even their own time or bodily autonomy—because everything belongs to some collective illusion (e.g., the state, a monarch).
+As well, what is fair would look very different in a society where very few own most, and most own very little due to a different illusion (e.g., that wealth equates merit).
+In these two cases, it is not necessarily difficult to elucidate the meaning of fair: fair is what the state, the monarch, or the extremely wealthy say it is.
+In contrast, in democratic societies (and democracy itself is a collective illusion) individuals are accorded rights, and there the relevant question is what a fair distribution of the burdens and benefits of things, from roads to space travel, would look like.
+
+Rockets are transportation technologies, tools of space-time convergence.
+By enabling movement at very high speeds they might prove important to the expansion of the human species beyond our home planet—at some indefinite time in the future.
+But in the present moment, the benefits of a private suborbital flight (e.g., the joy of movement, the sense of adventure, the awe of seeing Earth from space), accrue to just a few, whereas the burdens (e.g., the climate-altering emissions) affect all, and not even equally at that.
+In a democratic society, the members of the collective might decide that this arrangement, whereby large rewards are offered to the most qualified individuals (e.g., Bezos) to entice them to take more important positions [e.g., creating and leading Amazon; see @davisSome1945], is in fact a fair distribution of the burdens and benefits of space travel.
+Then again, the members of the collective might decide otherwise.
+
+Multiple studies in an international context suggest that most people in many societies have some tolerance for inequality [@kiatpongsanHow2014]: indeed, it appears that _some_ stratification, as suggested by @davisSome1945, is perceived as being functional.
+However, extreme inequality is often frowned upon, and can lead to social dysfunction [@acemogluWhy2000; @taydasCan2012; @duIncome2019; @houleInequality2022].
+But the perceptions of what is "fair" are malleable, and can be affected by opportunities for social mobility [@shariffIncome2016], exposure to inequality [@schroderIncome2017; @garciaChanging2023], **learned helplessness**, etc.
+
+Positive -> what is the current status of the distribution of benefits and burdens?
+Normative -> what should it be?
 

### 2023-10-03 Manuscript/Manuscript.qmd
More edits to Manuscript.qmd

--- 2023-08-17 Manuscript/Manuscript.qmd
@@
 
-Equitable transportation system from a planning perspective, involves fair production, distribution, and reproduction of transportation benefits and costs within a community [@pereiraDistributiveJusticeEquity2017a; @shellerMobilityJusticePolitics2018; @pereiraTransportationEquity2021].
-Transportation equity is a multi-scalar and multi-dimensional topic centered on the distribution of transport-related benefits and burdens across population groups and communities within urban, suburban or rural areas.
-Benefits and burdens can include, but are not limited to, accessibility, mobility, affordability, health and environmental issues that impact individuals, marginalized members of a community, and the community as a whole [@lucasMeasuringTransportEquity2019].
-
-Transportation systems are complex: they support multi-modal uses and are home to emerging new technologies and service modes that complicate the balance of benefit and burdens [@guoSystematicOverviewTransportation2020].
-Consequently, achieving an *equitable* system is not a positivist attribute; it should be normatively defined by the community  and broader society [@pereiraTransportationEquity2021; @paezMeasuringAccessibilityPositive2012] during the ongoing process of measurement, operationalization, and improvement.
-Both academic literature and planners are grappling to contextualize what an equitable transportation system may look like along spatial, temporal, environmental, and socio-economical dimensions [@boisjolyHowGetThere2017; @pereiraTransportationEquity2021].
-Synthesizing methods on how to identify, measure, and operationalize transportation equity is urgently needed as planners and decision-makers are undecided on how to distribute existing transportation resources and simultaneously invest in future transportation infrastructure *fairly*.
-
-The application of equity within the realm of transportation planning has proliferated in the academic literature but concrete adoption of the concepts in planning practice has been minimal [@pereiraTransportationEquity2021; @boisjolyHowGetThere2017; @doranPursuitCyclingEquity2021; @linovskiEquityTransportationPlanning2020; @litmanEvaluatingTransportationEquity2022].
-Academic literature has approached equity through measures such as accessibility and activity participation [@paezMeasuringAccessibilityPositive2012; @allenPlanningTransportSocial2020], affordability [@isalouTestingHousingTransportation2014], environmental impacts and safety [@guoSystematicOverviewTransportation2020], and health [@fransenIdentifyingPublicTransport2015; @smithSystematicLiteratureReview2017], and also conceptualizations such as frameworks for equity and social exclusion [@lucasProvidingTransportSocial2006; @lucasTransportPovertyIts2016], and distributive justice theories [@behbahaniDesigningRoadNetwork2019; @lewisExploringEquityHow2021; @pereiraDistributiveJusticeEquity2017a; @vecchioAccessibilityCapabilitiesApproach2021], among others.
-However, a gap in the literature is the absence of a comprehensive review that examines how equity is conceptualizated, theorized and operationalized through measures and standards.
-These inquiries are vital to planning agencies and decision-makers, and as such, this review aims to definitively benchmark the knowledge in the literature from this perspective.
-
-Though aspects of transport equity in metropolitan planning documents such as accessibility [@boisjolyHowGetThere2017] or mode-specific equity [@doranPursuitCyclingEquity2021] have been reviewed, no review to date captures how equity has been theorized and translated into standards from the perspective of equity dimensions.
-Our work contributes in this direction, reviewing which equity dimensions, standards, and conceptualizations have been applied in the academic literature and identifying where gaps exist in the literature.
-Ultimately, this review aims to collate the academic knowledge and present it in an operationalizable manner to both academics and decision-makers in hopes to catalyze the uptake in equitable planning practices and guide future research programmes which develop definitions, measures, and recommendations for evaluating transportation equity.
-
+> "_To be wealthy and honored in an unjust society is a disgrace._"
+> — Confucius, The Annalects
+
+The fiery wake of a rocket could be seen ascending, piercing the blue skies of West Texas.
+It was mid-morning on July 20, 2021, and on board was a small group of four passengers that included Jeff Bezos, the founder of Amazon and then the world's richest person [@harwoodBezos2021].
+The mission that morning was among the first ever crewed private suborbital flights, and the adventure (described as "intense" by a member of the crew) lasted a total of 10 minutes and 10 seconds [@harwoodBezos2021].
+In addition to intense, the undertaking was expensive: a seat for the flight had previously been auctioned for no less than $28 million USD [@griffinPerson2021].
+Before even 10:00 am (EDT) Bezos was back from his excursion, and right on time to declare that this was the "[b]est day ever".
+To reporters covering the event he said "I want to thank every Amazon employee and every Amazon customer, because you guys paid for all of this" [@johnsonRoadToSpace2021].
+
+Meanwhile, firmly grounded on planet Earth, the employees that Bezos thanked for his suborbital escapade were struggling with some very mundane problems of their own, none of them as lofty as conflicting schedules that prevented them from flying in rockets [@griffinPerson2021].
+According to reports, people employed directly or indirectly by Amazon for warehouse or delivery work had, for years, been treated to "inhumane" conditions [@fungEuropean2018; @scottAmazon2019; @greeneAlgorithm2021], and subjected to surveillance on the job, crushing demands for productivity quotas that led to injury, facing little or no job security, and more dealt with as disposable inputs to Amazon's targets and consumers' desires [@tungAmazon2020; @reeseTeamsters2022; @middletonRobot2023].
+
+Coverage of the July 20 launch by the mainstream media was in many cases uncritical.
+"We're going to build a road to space so that our kids and their kids can build a future" Bezos declared, before adding "...we need to do that to solve the problems here on Earth" [@johnsonRoadToSpace2021].
+Few reporters saw it fit to ask what problems the billionaire planned to solve on Earth, or what kind of future Bezos was trying to build, and for whose children.
+In other words, the billionaire was not confronted with the question of what did his trip do _for whom_ and _to whom_.
+A less dispassionate observer might have been excused for wondering (possibly aloud) about the basic _fairness_ of a man amassing a nigh unimaginable fortune that allowed him to build and fly his own spaceship, while masses of his employees were treated as throwaway cogs in the vast apparatus of his logistics empire.
+
+The question of _fairness_, however, is not a simple one, and many of us would have been stumped to explain in a clear and precise way, beyond a disquieting intuition, just why the above picture was disturbing.
+_Justice_, is a political ideal based on the principle that individuals should be treated in a fair and equitable manner [@gosslingUrban2016, p.
+2], giving and receiving whatever they are _due_ [@jaggarPhilosophicalChallengesGlobal2009, pp.
+1-2].
+The political [and contested; @vanoutriveHow2019] nature of the concept presents challenges that are only narrowly amenable to scientific inquiry.
+For starters, the notion that people are "due" something depends on the values of a society, as embodied in its systems and institutions [@karnerFrom2020].
+Values,in turn, are not subject to natural laws, but rather are the result of intersubjective realities, which is to say illusions whose legitimacy derives from a collective will to believe.
+For example, justice would likely mean something very different in a democratic society, if compared to a society where most owed their all—perhaps including their own time or bodily autonomy—to some collective illusion (e.g., the state, a monarch).
+As well, the meaning of "justice" would likely differ in yet another society where very few owned most, and most owned very little due to a different illusion (e.g., that wealth equates merit).
+In these two hypothetical cases, elucidating the meaning of "fair" would in all certainty be beyond the dreams of most, since fair would be what the state, the monarch, or the extremely wealthy said it was.
+In contrast, in democratic societies^[Paraphrasing, democracy is the worst of all collective illusions, except for all others.] individuals rights are valued above the whims of the few.
+In such a setting, the task of defining a "just" distribution of the burdens and benefits of things—from income, to roads, to space travel—quickly becomes muddled, encumbered even, by the necessity to pay attention to a multitude of voices, not all of them equally loud.
+
+The rocket that took Bezos to the edge of space is a somewhat rare example of a transportation technology, a tool of space-time convergence.
+By enabling movement at very high speeds rockets might—one day, at some indeterminate point in the future—prove essential to the expansion of the human species beyond our home planet.
+But in the present moment, the benefits of a private suborbital flight are for a few (e.g., the joy of movement, the sense of adventure, the awe of seeing Earth from space), whereas the burdens (e.g., the use of non-renewable resources, the climate-altering emissions) affect us all, and not evenly at that.
+In a plutocratic society, those with most can argue that this distribution of burdens and benefits is fair since both benefits and burdens are earned.
+The fact that the likes of Bezos can travel to space is proof that the likes of Bezos deserve those trips.
+In a democratic society, the members of the collective might actually agree that large rewards (e.g., space flight) must be offered to highly qualified individuals (e.g., Bezos) to entice them to take important responsibilities (e.g., founding and leading Amazon).
+The members of the collective might think this a fair state of affairs, having come to the conclusion of their own accord or after being persuaded by billionaires.
+But then again, the members of the collective might decide otherwise: the _values_ of those in charge of defining what is "fair" matter.
+
+Multiple national and cross-national studies suggest that people in many societies have some tolerance for inequality [@kiatpongsanHow2014]: indeed, it appears that _some_ stratification, as suggested by @davisSome1945, is thought to fulfill a valuable function.
+However, extreme inequality is often frowned upon, and can lead to social dysfunction and other ills [@acemogluWhy2000; @taydasCan2012; @duIncome2019; @houleInequality2022].
+But the perceptions of what is "fair" are neither universal or static.
+Instead, they are malleable, and can be affected by the existence of opportunities for social mobility [@shariffIncome2016; @artigeWhat2019], by exposure to inequality [@schroderIncome2017; @garciaChanging2023], by learned helplessness [@kimSocioeconomic2022], and even by how information about inequality is communicated to the public [@walkerPeople2021].
+It follows that, in general terms, there are at least three different manners of thinking about inequality: 1) in a _positive_ (or descriptive) manner, as the current state of the distribution of benefits and burdens of things; 2) also descriptively, as the _perceptions_ about the distribution of those benefits and burdens; and 3) normatively (or prescriptively), as the desired or ideal state of the same.
+Clearly, the three may coincide (e.g, if the perceived levels of inequality matched actual inequality and also how much of it the public desired).
+However, they do not necessarily have to, and in many cases will differ from one another.
+Measuring inequality as it is, in the most objective way possible is an essential task to decide whether there is a need to develop inequality-related policy to increase fairness.
+In turn, measuring the perceptions of inequality, in the most accurate way possible, may be important to achieve sufficient public buy-in in order to enhance the chances that a given policy will succeed.
+
+Transportation systems are a class of essential technologies that facilitate or impede social inclusion and activity participation by the public [@churchTransport2000; @lucasTransport2001; @socialExclusion2003; @cassSocial2005; @casasSocial2007; @prestonAccessibility2007; @paezMobility2009].
+As such, they increasingly are of interest from the perspective of equity, and a lively and rapidly growing literature has emerged in response [see _inter alia_ @martensTransport2016; @diciommoTransport2017; @guoSystematicOverviewTransportation2020; @karnerFrom2020; @vecchioTransportEquityLatin2020; @pereiraTransportationEquity2021; @vanweeEvaluating2021; @zhangLiterature2021; @desjardinsExaminingEquityAccessibility2022; @karnerAdvancesPitfallsMeasuring2023].
+A cynical rationale for this interest could be that keeping track of actual and perceived inequalities can serve _at the very least_ as a gauge of social discontent [as Chilean authorities discovered to their sorrow when unrest erupted over the increase of metro fares in 2019; @BBCNews2019; @diazpabonInequality2021].
+More optimistically, in democratic systems, there could be a genuine interest in being ready to respond to popular demands.
+Several challenges arise when approaching this endeavor, caused by the notorious complexity of transportation systems: simultaneously, they move people, goods, and information.
+As well, emerging technologies and service models can swiftly change the balance of benefits and burdens among a population [@guoSystematicOverviewTransportation2020].
+Unlike income, which is as unidimensional as it gets, the benefits and burdens of transportation systems are diffuse over space and time.
+For example, transportation systems engineered to offer higher mobility for people _somewhere_, can simultaneously cut others off essential opportunities _elsewhere_ , as @rajeEngineering2004 poignantly illustred with examples of infrastructure in the UK.
+Furthermore, the shades of policies past can continue to haunt a region and even the planet for decades or longer, as shown by the legacy of displacement and decay caused by urban highways all across the US [@archerWhite2020] and the time horizon for the impacts of climate change to be fully felt [@markolfTransportation2019].
+
+But complexity is no reason to shirk the task.
+
+The objective of this review is to investigate the state of knowledge in terms of defining and operationalizing "fairness" in the transportation domain.
+Much research has been devoted to the issues of _measuring _ equity in transportation, including @ramjerdiEquity2006, @delboscUsing2011, @welchMeasure2013, @martensMeasuring2019, and @pritchardNew2022, among many others.
+Further, there are multiple sources that discuss the conceptual and philosophical foundations of equity and fairness in transportation [e.g., @martensTransport2016; @pereiraDistributiveJusticeEquity2017a; @vanoutriveHow2019].
+Finally, previous reviews of planning documents have investigated equity from narrowly scoped perspectives, such as accessibility [@boisjolyHowGetThere2017] or a particular mode of transportation [e.g., cycling; @doranPursuitCyclingEquity2021].
+These inquiries are valuable to planning agencies and decision-makers, and the present review will tread similar, but not completely overlapping ground since, in our estimation, there is a gap in the literature in terms of understanding how standards for equity are developed and implemented in the transportation domain.
+In contrast, we do know that adoption of equity concepts in planning practice has lagged developments in academic work [@pereiraTransportationEquity2021; @boisjolyHowGetThere2017; @doranPursuitCyclingEquity2021; @linovskiEquityTransportationPlanning2020; @litmanEvaluatingTransportationEquity2022].
+
+To illustrate this gap, we note how @beilerExploring2016, in their exploration of transport equity, cite the following strategies identified by the US DOT to address justice [p. 287]:
+
+- Reduce adverse human health and environmental effects on minority and low-income populations.
+- Include all potentially affected communities in the transportation decision-making process.
+- Ensure that minority and low-income populations receive equitable benefits.
+
+While commendable, the strategies are sufficiently vague that they can be implemented in a myriad ways, either to comply with the spirit of justice, or instead performatively to deceive it [@mcculloughPerformative2023].
+Some relevant questions include: how much should the adverse effects be reduced?
+To zero?
+Or to some tolerable level of adversity?
+What should that level be?
+What are the criteria for deciding that a community is "potentially affected"?
+What benefits should be distributed, and should those be based on simple population weights?
+Or, contrariwise, should more deprived individuals be eligible for a larger share of the benefits?
+
+These questions boil down to the development and use of _standards_ for transportation justice.
+A definition of "standard" is as "something set up and established by authority as a rule for the measure of quantity, weight, extent, value, or quality"^[https://www.merriam-webster.com/dictionary/standard].
+How much pollution is allowed to be generated, and where, depends on who is affected, as well as on how much health is valued and by whom.
+For example, firms may or may not adopt lower standards for the emission of pollutants in poorer areas, or poor people may end up being relegated to areas that adopt lower standards [@gouldsonFirms2006].
+The result is the same: pollution tends to be worse were poorer people are [@delucaCode2012].
+
+Supporting the creation of (more) equitable transportation systems involves understanding the production, distribution, and management of transportation benefits and costs; how they are distributed; and what values are implemented (and by whom) in the form of standards [@pereiraDistributiveJusticeEquity2017a; @shellerMobilityJusticePolitics2018; @pereiraTransportationEquity2021].
+Thus, for this review we engage the literature with the following questions in mind: 1) what is our current understanding of the things that transportation systems do, for whom, and to whom; 2) what does the literature say about the distribution of the benefits and costs of transportation systems; 3) and what values are embodied in normative statements about said distribution.
+Ultimately, this review aims to collate the academic knowledge and present it in a manner useful to support both academics and decision-makers in the development and implementation of standards for equity in transportation planning and policy, some of which are sorely missing, even in those places where legislation explicitly calls for justice^[[As recently as -@martensFair2021, Martens and Golub note that Federal directives related to Title VI in the US "do not provide guidelines that can help agencies develop explicit standards to assess the distribution of accessibility benefits from projects or plans."]].
+
+<!--
 This paper is structured as follows.
@@
 In Section 6, we summarize the findings and link them to future transport planning agendas in addition to visioning ways the findings can be used to build *just* transportation planning processes.
+-->
 
@@
 
-Transportation equity literature exists within cross-cutting categories.
-We borrow a "Where?", "When?", "Who?", "What?", and "How?" framing to categorize aspects of how equity is assessed.
-This categorization is outlined in @jaggarPhilosophicalChallengesGlobal2009 to describe substantive sets of answers to questions of distributive justice (and further discussed in the context of justice within the critical human geography discipline [@przybylinskiWhereJusticeGeography2022]).
+> "_I should like to be able to love my country and still love justice_."
+> — Albert Camus
+
+> "_Never forget that justice is what love looks like in public_."
+> ― Cornel West
+
+<!-- From the original intro
+
+Benefits and burdens can include, but are not limited to, accessibility, mobility, affordability, health and environmental issues that impact individuals, marginalized members of a community, and the community as a whole [@lucasMeasuringTransportEquity2019].
+Academic literature has approached equity through measures such as accessibility and activity participation [@paezMeasuringAccessibilityPositive2012; @allenPlanningTransportSocial2020], affordability [@isalouTestingHousingTransportation2014], environmental impacts and safety [@guoSystematicOverviewTransportation2020], and health [@fransenIdentifyingPublicTransport2015; @smithSystematicLiteratureReview2017], and also conceptualizations such as frameworks for equity and social exclusion [@lucasProvidingTransportSocial2006; @lucasTransportPovertyIts2016], and distributive justice theories [@behbahaniDesigningRoadNetwork2019; @lewisExploringEquityHow2021; @pereiraDistributiveJusticeEquity2017a; @vecchioAccessibilityCapabilitiesApproach2021], among others.
+
+-->
+
+## What is justice?
+
+In the paragraphs above we used the terms "justice", "fairness", and "equity" in a relatively loose fashion.
+This was done purposefuly.
+As we noted, people often have a strong intution of what is "fair", "just", and "equitable".
+These conceptions may or may not match those of the authorities who set the standards of fairness.
+But in democratic societies, the authority of political leaders, bureaucrats, planners, and all those charged with the business of governing, derives from the will of the people.
+It is therefore important to make the implicit explicit, and clearly spell out those intuitions of "justice", "fairness", and "equity" to make them conceptually clear.
+Only this way can the participants of a democratic society be effective arbiters of what is just - and perhaps stand on firmer ground to experience love of their society.
+
+Let us begin by stating that justice is an end goal, that is, a desirable state of affairs that we strive to achieve.
+
+Justice is attained when people "give and receive whatever they are due" [@jaggarPhilosophicalChallengesGlobal2009, pp.
+1-2], and it ceases to exist when there are persons or groups that are denied "access to the opportunities they need to lead a meaningful and dignified life" [@karnerTransportationEquityTransportation2020, p.
+440]
+
+Several forms of justice can be distinguished, including [see @jaggarPhilosophicalChallengesGlobal2009; @karnerTransportationEquityTransportation2020]:
+
+**Retributive justice**. It is concerned with the retribution due to people who do wrongs.
+
+**Reparative justice**. This form of justice relates to the proper way to correct or rectify past wrongs even if the wrongdoers can no longer be meted retribution.
+
+**Procedural justice**.
+The main concern of this form of justice is to ensure that the views, opinions, and preferences of all stakeholders are properly accounted for when decisions that affect their lives are made.
+
+**Distributive justice**.
+Perhaps the most commonly studied form of justice [see @jaggarPhilosophicalChallengesGlobal2009, p.
+2; @pereiraDistributiveJusticeEquity2017a], and its main concern is the way the benefits and burdens of the tangible and intangible products of society.
+
+It might be argued that all of the above are particular forms of distributive justice.
+Retributive justice, for example, is often concerned with the distribution of the benefits and burdens of being a member of society: the way it is usually achieved is by distributing intangibles like "freedom" (e.g., of movement, of association) as benefits, and/or claiming tangibles like money as burdens (e.g., fines).
+Reparative justice often distributes current benefits and burdens to redress past wrongs, for example by asking those who have benefited from past wrongs, even if unwittingly, to shoulder a bigger fiscal burden in order to cover programs that mete benefits to those who are harmed by past wrongs.
+Procedural justice is the distribution of the benefits (e.g., the right to voice an opinion) and burdens (e.g., the effort required to develop an educated opinion) of the processes that lead to decisions with collective consequences.
+
+We can then speak of the purposes of distributive justice: to mete out retribution _fairly_, to repair past harms, and to ensure that procedures offer _equitable_ opportunities for engagement.
+Equity and fairness are the instruments of justice, the tools by which society advances towards the end goal of justice.
+
+<!-- Furthermore, it is important to distinguish between equity and justice. -->
+
+The term "equity" tends to encompass various tools to understand the distribution of benefits and burdens of things among a population.
+In the transportation domain, the term is somewhat loaded because it is perceived as stemming from the authority of the state, and meant to assist with decisions about regulating and financing transportation spending [@karnerTransportationEquityTransportation2020].
+Here, we are in agreement with @karnerTransportationEquityTransportation2020 that equity analysis should not be seen as an end in and of itself, but rather as a means to gather information about actual and perceived inequality.
+In this respect, the analytical tradition of equity, at least in transportation planning, means that the relevant models become embedded in the "political ecology of the estimated truth" [@kingModels1993]: in principle their assumptions and scope must be open and transparent, but they are also vulnerable to misuse.
+
+Fairness, in contrast to equity, is more complicated to define.
+The concept does not have the same history of development as an analytical tool, and can be interpreted in numerous, and possibly discordant ways.
+That this is the case is convincingly demonstrated by @martensFair2021 in their study of the application of Title VI of the Civil Rights Act of 1964 in accessibility planning practice in the US.
+Title VI explicitly talks about the distribution of benefits derived from Federal funding: "[n]o person in the United States shall, on the ground of race, color, or national origin, be excluded from participation in, be denied the benefits of, or be subjected to discrimination under any program or activity receiving Federal financial assistance." However, as @martensFair2021 show, there are several ways to comply with regulations while achieving different outcomes: from a very mild form of fairness that simply avoids intentional discrimination, to strategically overdeliver the benefits to compensate for past discrimination (reparative justice).
+What justice did fairness serve in each case?
+It depends on what was the reason for seeking justice in the first place!
+Our reading of @martensFair2021 is that fairness is a yardstick that is best used _a priori_ than _a posteriori_, for doing so risks rationalizing the outcomes rather than
+
+The focus of this report is on standards of fairness that, combined with the use of equity analysis, can help understand how better to move towards just outcomes in transportation planning.
+
+<!--- EDITED UP TO HERE --->
+
+## A framework for questions of justice
+
+For the analysis we are inspired by the framing of @jaggarPhilosophicalChallengesGlobal2009 for philosophical questions of justice.
+
+borrow a "Where?", "When?", "Who?", "What?", and "How?" framing to categorize aspects of how equity is assessed.
+
+<!-- The texture of these questions becomes more crisp when talking about transportation, which is inherently about space and time, that is to say about "where" and "when".
+Conventionally, theories of justice have been aspatial, despite an early interest of geographers on matters of justice [@]-->
+
+This categorization is outlined in  to describe substantive sets of answers to questions of distributive justice (and further discussed in the context of justice within the critical human geography discipline [@przybylinskiWhereJusticeGeography2022]).
 For questions of *justice*, convincing answers to these questions require a rationale: a "Why?" [@jaggarPhilosophicalChallengesGlobal2009].
 For the purpose of this review focused on *equity*, the focus is on studies that substantively answer "Where?", "When?", "Who?", "What?", and "How?"; the answer to "Why?" may be incidental.
+
+## Transportation and justice
+
+To understand the distributional implications of transportation systems, it is essential that we are clear about what they do. In other words, what do transportation systems _produce_?
+
+At their most fundamental, transportation systems are space-time convergence tools, tools that improve the rate at which time is traded for space.
+They usually do this by increasing the speed of movement: sidewalks facilitate walking, traffic lights facilitate the ordered flow of vehicular traffic, and a launching pad makes it possible for a rocket to takeoff.
+With complex interlocking parts (sidewalk, road, traffic sign, parking regulations), transportation systems produce _mobility_, the potential for movement.
+Combined with land use systems (the spatial distribution of opportunities on the landscape), they produce _accessibility_, the potential to reach destinations.
+The realization of those potentials happens through travel.
+But, as the adage goes, travel is derived demand, which seems to hold for most (even if not all) travel.
+
+Transportation equity literature exists within cross-cutting categories.
+
 Though critical, answers to "Why?" are not overwhelming common in the transportation equity academic literature that also address  "Where?", "When?", "Who?", "What?", and "How?".
 This is discussed as a next step in planning for *justice* transportation literature in Section 6.
-
-Furthermore, it is important to distinguish between equity and justice.
-Our focus is on equity, which is nested within and can be interpreted as the application of justice theories.
-For instance, *mobility justice* seeks to understand how broader power inequities inform the governance and control of movement are reproduced by mobility systems [@shellerMobilityJusticePolitics2018]; in understanding what is *just* and *unjust,* equity is the application of processes to move towards justice.
-What is justice must be decided at a community level, a goal of the equity processes, and cannot be captured in a single academic text.
 
@@
 -   *Who*: includes the type of population group, mode type, activity/opportunity type, and transport network status.
+
+<!-- Is it appropriate to talk about a "population group"? Think about intersectionality theory.
+
+These are some things that people _are_:
+
+- Gender identity
+- Ableness
+- Ethnicity
+- Age
+- Caste
+- Low income (in societies with low social mobility)
+- Membership in a family
+
+-->
+
 -   *How*: includes the measure used to assess equity dimensions as supported by a **equity conceptualization** and **equity standard**.

### 2023-10-05 Manuscript/Manuscript.qmd
Antonio is still editing

@@
 
-The fiery wake of a rocket could be seen ascending, piercing the blue skies of West Texas.
-It was mid-morning on July 20, 2021, and on board was a small group of four passengers that included Jeff Bezos, the founder of Amazon and then the world's richest person [@harwoodBezos2021].
+It was a common kind of morning in West Texas but for the fiery wake of a rocket that could be seen ascending, piercing the blue skies.
+The date was July 20, 2021, and on board was a small group of four passengers that included Jeff Bezos, the founder of Amazon and then the world's richest person [@harwoodBezos2021].
 The mission that morning was among the first ever crewed private suborbital flights, and the adventure (described as "intense" by a member of the crew) lasted a total of 10 minutes and 10 seconds [@harwoodBezos2021].
@@
 Instead, they are malleable, and can be affected by the existence of opportunities for social mobility [@shariffIncome2016; @artigeWhat2019], by exposure to inequality [@schroderIncome2017; @garciaChanging2023], by learned helplessness [@kimSocioeconomic2022], and even by how information about inequality is communicated to the public [@walkerPeople2021].
-It follows that, in general terms, there are at least three different manners of thinking about inequality: 1) in a _positive_ (or descriptive) manner, as the current state of the distribution of benefits and burdens of things; 2) also descriptively, as the _perceptions_ about the distribution of those benefits and burdens; and 3) normatively (or prescriptively), as the desired or ideal state of the same.
+It follows that, in general terms, there are at least three different manners of thinking about inequality^[@martensTransport2016, in the introduction of his landmark text, talks about explanatory and prescriptive theories of justice; despite the saying "the arc of history is long and bends toward justice", it is unlikely that a predictive theory of justice exists.]: 1) in a _positive_ (or descriptive) manner, as the current state of the distribution of benefits and burdens of things; 2) also descriptively, as the _perceptions_ about the distribution of those benefits and burdens; and 3) normatively (or prescriptively), as the desired or ideal state of the same.
 Clearly, the three may coincide (e.g, if the perceived levels of inequality matched actual inequality and also how much of it the public desired).
@@
 
-Supporting the creation of (more) equitable transportation systems involves understanding the production, distribution, and management of transportation benefits and costs; how they are distributed; and what values are implemented (and by whom) in the form of standards [@pereiraDistributiveJusticeEquity2017a; @shellerMobilityJusticePolitics2018; @pereiraTransportationEquity2021].
+Supporting the creation of (more) just transportation systems involves understanding the production, distribution, and management of transportation benefits and costs; how they are distributed; and what values are implemented (and by whom) in the form of standards [@pereiraDistributiveJusticeEquity2017a; @shellerMobilityJusticePolitics2018; @pereiraTransportationEquity2021].
 Thus, for this review we engage the literature with the following questions in mind: 1) what is our current understanding of the things that transportation systems do, for whom, and to whom; 2) what does the literature say about the distribution of the benefits and costs of transportation systems; 3) and what values are embodied in normative statements about said distribution.
-Ultimately, this review aims to collate the academic knowledge and present it in a manner useful to support both academics and decision-makers in the development and implementation of standards for equity in transportation planning and policy, some of which are sorely missing, even in those places where legislation explicitly calls for justice^[[As recently as -@martensFair2021, Martens and Golub note that Federal directives related to Title VI in the US "do not provide guidelines that can help agencies develop explicit standards to assess the distribution of accessibility benefits from projects or plans."]].
+Ultimately, this review aims to collate the academic knowledge and present it in a manner useful to support both academics and decision-makers in the development and implementation of standards for equity in transportation planning and policy.
+In this, we aim to provide relief even, or particularly, in those places where legislation explicitly calls for justice^[[As recently as -@martensFair2021, Martens and Golub note that Federal directives related to Title VI in the US "do not provide guidelines that can help agencies develop explicit standards to assess the distribution of accessibility benefits from projects or plans."]].
 
@@
 
-Several forms of justice can be distinguished, including [see @jaggarPhilosophicalChallengesGlobal2009; @karnerTransportationEquityTransportation2020]:
+Several forms of justice can be distinguished, including [see @jaggarPhilosophicalChallengesGlobal2009; @pereiraDistributiveJusticeEquity2017a; @karnerTransportationEquityTransportation2020]:
 
@@
 
-**Reparative justice**. This form of justice relates to the proper way to correct or rectify past wrongs even if the wrongdoers can no longer be meted retribution.
+**Reparative (or restorative) justice**. This form of justice relates to the proper way to correct or rectify past wrongs even if the wrongdoers can no longer be meted retribution.
 
@@
 That this is the case is convincingly demonstrated by @martensFair2021 in their study of the application of Title VI of the Civil Rights Act of 1964 in accessibility planning practice in the US.
-Title VI explicitly talks about the distribution of benefits derived from Federal funding: "[n]o person in the United States shall, on the ground of race, color, or national origin, be excluded from participation in, be denied the benefits of, or be subjected to discrimination under any program or activity receiving Federal financial assistance." However, as @martensFair2021 show, there are several ways to comply with regulations while achieving different outcomes: from a very mild form of fairness that simply avoids intentional discrimination, to strategically overdeliver the benefits to compensate for past discrimination (reparative justice).
-What justice did fairness serve in each case?
+Title VI explicitly talks about the distribution of benefits derived from Federal funding: "[n]o person in the United States shall, on the ground of race, color, or national origin, be excluded from participation in, be denied the benefits of, or be subjected to discrimination under any program or activity receiving Federal financial assistance." However, as @martensFair2021 show, there are several ways to comply with regulations while achieving different outcomes: from the banal (do not _knowingly_ discriminate), to the substantive (compensation for past discrimination, i.e., reparative justice).
+What justice does fairness serve in each case?
 It depends on what was the reason for seeking justice in the first place!
@@
 
-The focus of this report is on standards of fairness that, combined with the use of equity analysis, can help understand how better to move towards just outcomes in transportation planning.
+The focus of this report is on standards of fairness that, combined with the use of equity analysis, can help understand how better to move towards just transportation outcomes.
+
+## A framework for questions of justice
+
+Here we are inspired by the framing of @jaggarPhilosophicalChallengesGlobal2009 for philosophical questions of justice^[Similar questions are found peppered throughout the literature either explicity, as for example in @karnerTransportationEquityTransportation2020, who ask "of what", "for whom," and "how much" in reference to equity, or implicitly, as in @gosslingUrban2016, who asks "what?" (exposure, space, access) and "for whom?" (gender, age, ethnicity) about the outputs of transportation.].
+According to -@jaggarPhilosophicalChallengesGlobal2009, Western philosophy has approached the issue of justice by asking "Where?", "When?", "Who?", "What?", and "How?".
+Conventionally, discussions about justice have been aspatial, or rather, seen from the point of view of social space instead of geographical space, despite an early interest of geographers on the matter [@pirieSpatial1983].
+The texture of the questions becomes more immediate and crisp when talking about transportation, which is inherently about space and time.
+
+- **"Where?"** Traditionally related to the applicable domain or sphere of life relevant for justice.
+Conventionally this meant the in-group [e.g., members of the same nation state; @jaggarPhilosophicalChallengesGlobal2009, p.
+3], a notion since expanded (e.g., with the Universal Declaration of Human Rights) to include all members of the human species, but has yet to cover other living creatures.
 
@@
 
-## A framework for questions of justice
-
-For the analysis we are inspired by the framing of @jaggarPhilosophicalChallengesGlobal2009 for philosophical questions of justice.
-
-borrow a "Where?", "When?", "Who?", "What?", and "How?" framing to categorize aspects of how equity is assessed.
-
-<!-- The texture of these questions becomes more crisp when talking about transportation, which is inherently about space and time, that is to say about "where" and "when".
-Conventionally, theories of justice have been aspatial, despite an early interest of geographers on matters of justice [@]-->
+- "When?" asks what are the social circumstances within which the
+demands of justice have application.
+- "Who?" asks which entities should be regarded as subjects of justice,
+meaning who or what are entitled to make justice claims deserving of
+moral consideration.
+- "What?" asks which entities should be regarded as objects of justice,
+meaning which kinds or categories of things should be distributed in a
+just manner.
+- Finally, "How?" asks which principles are the most morally appropriate
+for guiding the allocation of various objects to various subjects in various
+circumstances.
 
 This categorization is outlined in  to describe substantive sets of answers to questions of distributive justice (and further discussed in the context of justice within the critical human geography discipline [@przybylinskiWhereJusticeGeography2022]).
+
 For questions of *justice*, convincing answers to these questions require a rationale: a "Why?" [@jaggarPhilosophicalChallengesGlobal2009].
@@
 
-To understand the distributional implications of transportation systems, it is essential that we are clear about what they do. In other words, what do transportation systems _produce_?
+To understand the distributional implications of transportation systems, it is essential that we are clear about what they do. What do transportation systems _produce_?
 
@@
 The realization of those potentials happens through travel.
-But, as the adage goes, travel is derived demand, which seems to hold for most (even if not all) travel.
+But, as the adage goes, travel is derived demand, which seems to hold for most (even if not all) situations [ref mokhtarian].
 

### 2023-10-17 Manuscript/Manuscript.qmd
Added more to the "Discussion" and my main points to the "Conclusion"

@@
 
-# Discussion and ways forward (identifying gaps and possible agenda for the future`r word_comment("This section is a work in progress, but we outlined some ideas.
-This will be completed after the team's first read using ideas generated.", ")")`
-
-We discuss trends in how the literature has conceptualized and used standards (the "How") to evaluate the "What" of transportation equity, the equity dimensions.
-These trends are context-specific: they concern the "When" and "Where" (differences across publication year, across case studies continent, and between urban and rural contexts) and the "Who" (differences in the focus on the population sub-groups, transport modes, destinations of interest) as discussed in the preceding sections following the four equity dimensions that structure this review (mobility/accessibility, traffic-related pollution, human-health related, and traffic safety).
-
-A summary of what we found in the literature:
-
+# Discussion and ways forward
+
+We discuss trends in how the literature has conceptualized and used standards (the "How") to evaluate the dimensions (the "What") of transportation equity.
+A noteable point is the trends are context-specific: they concern the "When" and "Where" (differences across publication year, across case studies continent, and between urban and rural contexts) and the "Who" (differences in the focus on the population sub-groups, transport modes, destinations of interest) as discussed in the preceding sections following the four equity dimensions that structure this review (mobility/accessibility, traffic-related pollution, human-health related, and traffic safety).
+
+To broadly summarize what we found in the literature:
+<!--
 - the "When" and "Where": 
@@
   - cross-dimensional -> measures such as urban livability, etc. a more holistic conceptualization of transport equity
-
-## The need for explicit conceptualizations and grounded standards
-
-Some conceptualizations are implicit. Some standards seem arbitrary. Despite the need for and importance of setting more standards, some proposals seem arbitrary.
-For example, XXXX proposes 20 ferries per day to avoid social exclusion.
-Justifications for this number is lacking.
-It is unclear if 10 or 30 ferries would make a difference in a specific quality-of-life outcome or if that number is tied to funding/resource constraints.
-We need standards linked to grounded outcomes.
-
-Relative measures are used: context, temporal, and group-specific.
+-->
+
+**The "When" and "Where"**: 
+The majority of papers (`r ((case_cont_count$count[4] + case_cont_count$count[5]+ case_cont_count$count[6])/sum(case_cont_count$count)) |> percent()`) focus on case studies in the Global North.
+Though their subject matter is varied, their spatial context is one situated in case studies in North American and Europe, and thus more often than not, in conversation with more developed and formal government transport planning apparatuses and technologies (e.g., planning for equitable high-speed rail [@monzonEfficiencySpatialEquity2013], autonomous vehicle technology [@eppenbergerOpportunitySharedAutonomous2021], public consultation processes [@reddySafeguardingMinorityCivil2010]).
+
+However, studies from the Global South, specifically South America and Africa, have key differences in focus compared to Global North: they are more focused on affordability-as-a-barrier, public spending tensions in developing new transport infrastructure compared to transport service prioritization and more intense data availability limitations.
+In this way, they do not engage with emerging mobility technologies with large capital costs.
+Informality in transport planning is more present.
+In the present day, these countries, compared to the Global North, are underdeveloped for reasons of past colonialism under Northern states, heavily reliant on primary sector exports (lower efficiency, lower national GDP) under growing global financial markets, and have more fragile democracies.
+Because of lower data availability and more extreme needs for 'sufficient' transport, analysis of transportation inequities are more focused along economic lines.
+For instance, public transit network being improved such that those that at least 60% below the poverty line have at least the same amount of accessibility (60 min travel time to destinations) to jobs, educational facilities and services as the total population using public transit [@basuPracticalFrameworkBenchmarking2019].
+
+Temporally, the Global North and Global South studies can be generally seen on different transport inequity continumums.
+Global North -> concerned with particularly disadvantaged groups and creating indicators that may be used to guide the remedy of formal processes that ensure access to societal benefits [@cuiSpatialAccessPublic2020].
+Global North countries often surpass international sufficiency thresholds related to transport and public health (e.g., @carrierApplicationThreeMethods2014 measure NO2 road-side air pollution and find no levels are below the WHO guideline) but advocate for the reduction in disproportional impact of air pollution on lower-income groups.
+New standards are needed that are context-specific.
+What's also missing?
+We call for: _discussions of equality_.
+The Global North inequities are rising -- the richest have the most mobility, this is missing from the discussion.
+We need new standards that reflect this.
+
+In reference to transport inequity continumus, papers on the Global South tend to operationalize international standards as guidelines.
+Though not covered in this review, Global South nations' formal processes for planning are newer (relatively), more fragile, and operating under more strict financial constraints than the Global North.
+Informal processes are thus more important to account for equity (e.g, informal transit [@friedMeasuringSustainableDevelopment2020], populations living in informal residential locations [@sharmaPublicTransitAccessibility2021]).
+Comparisons to thresholds set by governing international bodies are relevant as minimums are relatively lower (e.g., traffic related pollution, access to basic healthcare).
+Global South, along some dimensions, are lagging Global North development, but have the opportunity to plan better and not repeat Global North mistakes (e.g., car-centric development [@warrenDevelopingEquitableSustainable2015]).
+More work on this is needed.
+Additionally, what's missing?
+Similarly, _discussions of equality_.
+Inequities look different than in Global North, but systems thinking is required as global forces influence the capacity of formal transport planning processes?
+
+**The "Who"**:
+Though the literature reviewed most commonly considers low-income as a socio-demographic characteristics of populations deserving equity as either by itself or alongside other characteristics (`r df_pop$count[8]+df_pop$count[4] %>% percent()` of the papers).
+Many employ accessibility methods (e.g., walking accessibility to open space [@tangWalkingAccessibilityNeighbourhood2021]) but a others use qualitative or mixed methods to gain perspectives of inequalities that are typically cross-dimensional (e.g., @milanLiftingPeripheralFortunes2017 asks Medellin residents living in TOD areas compared to non-TOD areas about the impact of TOD on their wellbeing based on income group and gender).
+The use of qualitative methods allows for a more nuanced reflection of residents perspectives, and different inequities emerge.
+
+From a modal perspective, transit is also the primary focus of the literature; cycling and walking the next most popular focuses.
+These modes use public-space and users can interact with the public when using them (unlike car modes, which are seldom the focus of the reviewed literature).
+Upon the rise of automobility, transit was seen as a public service and something getting in the way of automobility.
+But historically, mass transit and active modes are how people get around - and automobility is in direct tension with the movement of other modes as it requires a lot of pubic land and public subsidy.
+This may be one reason why transport inequity research focuses on these modes, often car modes as driving inequalities in many dimensions (e.g., access/mobility, traffic-related pollution, traffic-safety, and human-health).
+
+In the literature reviewed, destinations of interest are mostly employment and healthcare; other destinations are less studied.
+For many papers, the 'destinations' however, are the trip themselves -- the experience, the quality of the infrastrucutre, the mode-service, a bundle of trips taken a month to all locations, etc.
+
+So what's missing from the "who"?
+More nuisances perspectives that capture the "who" of inequities.
+As shown in the variety of methods used, dimensions analysis, and conceptualizations and standards, transport inequalities are multi-dimensional.
+How decision makers define an equity-deserving community will impact results (e.g., [@rowangouldIdentifyingEnvironmentalJustice2015]).
+In this sense, standards that are defined need to be sensitive to changing community-based definitions of inequities.
+Are issues of economic-inequity at the root of transport inequities for a specific community?
+Are (dis)abilities?
+Are inequities in the service of transit the focus of a study because it can be improved to address transport inequities and are alternative modes driving these inequities?
+Access to what sorts of opportunities is driving transport inequities?
+How do populations, transit modes, and opportunities saught intersect to define the "who" of inequities?
+Community-based informed understanding of inequities are needed, and tracking how they change are needed.
+
+**The "What" and "How"**:
+Inequities that focus on accessibility/mobility -> opportunity standards are typically travel impedance standards, and affordability thresholds (e.g., more than 10% of monthly income should not be spent on transport [@rivasHowAffordableTransportation2018] to work and/or other necessary destinations).
+Oftentimes spatial equity, disparity analysis (spatial mismatch e.g., @mulleyAreNetworkPlanning2015) are operationalized.
+Infrastructure standards are also suggested -> often times access to destinations is conceptualized as a "right" as such being able to enter the transportation network and physically use it the pre-requisite that accessibility measures do not capture (e.g., (dis)abilities focused).
+
+Inequities that focus on transport-related environmental externalities -> pollution/noise standards are suggested, often using WHO guidelines or local neighbourhood averages as a comparison.
+
+Inequities that focus on human-health  -> health standards, the reliance on population transport health literature such as physical activity guideline recommendation per week or excess mortality burden.
+
+Inequities that focus on traffic-safety -> the use of engineering standards and best practice
+
+Of course, some papers focus on cross-dimensional transport inequity dimensions.
+Measures such as urban livability, etc.
+a more holistic conceptualization of transport equity.
+This is what we think is missing from the literature, and what's needed to move forward to justice.
+
+## Call 1: The need for explicit conceptualizations and grounded standards
+
+Some conceptualizations are implicit within the literature: for example, @muellerSocioeconomicInequalitiesUrban2018 relative risk of mortality as related to transport-related air pollution should not be higher in deprived groups than the general population.
+It is not evident from the paper what conceptualization drives this focus on human-health inequities but can be implicitly inferred to be general "well being".
+We believe, more explicit conceptualizations support inequities are needed in the literature.
+
+Some standards are also seemingly arbitrary.
+For example, @caoIndicatorsSocioSpatialTransport2017 proposes 20 ferries per day to avoid social exclusion for inter-island transport planning in the Philippines in their analysis, but admits that a standard should be politically determined.
+It is unclear if, for instance, 10 or 30 ferries would make a difference in a specific quality-of-life outcome or if that number is tied to funding/resource constraints.
+If literature is to recommend standards, researchers need standards linked to grounded outcomes.
+For decision-makers setting standards, measuring inequities and moving towards setting flexible guidelines for standards is the next step.
+As such, we believe, firmer justification of standards are lacking in the literature and are needed.
+
+Relative measures are often used: context, temporal, and group-specific.
 E.g.
@@
 Justice should not be mixed up with equity.
-More justice work is needed, e.g., Restorative justice -- repairing harm is missing from the discussion
-
-## Methods and data limitations
+More justice work is needed, e.g., Restorative justice -- repairing harm is missing from the discussion of addressing transport inequities.
+
+## Call 2: Need for creative methods for systems-thinking approach to inequities
 
@@
 
-## Evaluate interventions and policies
+## Call 3: More direct and explicity links between standards and experienced outcomes
+
+A more robust assessment of the implications of equity standards on life outcomes is still pending.
+Estimating the benefits of increased mobility or accessibility, or reducing affordability burdens and transport externalities needs to be associated with outcomes like life and neighborhood satisfaction, subjective well-being, mental and physical health, social capital, among others.
+
+## Call 4: Evaluations of interventions and policies
 
@@
 
-## Links between standards and outcomes
-
-A more robust assessment of the implications of equity standards on life outcomes is still pending.
-Estimating the benefits of increased mobility or accessibility, or reducing affordability burdens and transport externalities needs to be associated with outcomes like life and neighborhood satisfaction, subjective well-being, mental and physical health, social capital, among others.
-
-# Concluding remark`r word_comment("This section is a work in progress. This will be completed after the team's first read using ideas generated.", "s")`
-
--   TBD
+# Concluding remarks
+
+"Justice is a constant struggle" we need equity frameworks that support the movement for justice.
+Understanding and tracking transport inequities over time and space; specifically "who", "what", "where", "how" and "why" are integral.
+Then set policies that address these inequities but stay flexible to changing demands of justice.
+
+Flexible frameworks are important.
+As mentioned, transport is derived demand – what’s everyone’s demand look like?
+Past, current, and future?
+Creating transport equity frameworks that lend themselves to changing as the struggle for Justice moves forward and is resilient to backsliding is crucial.
+Setting standards from a systems approach: both considering the _positive_ rights (a right to have access to sufficient quality of essential services, for example) and _negative_ rights (mobility of cars must be limited as not to impact air quality, health, safety, and positive rights) must be both be conceptualized.
+Systems thinking is lacking transport equity literature.
+
+We thus set out the following calls for transport system decision makers in defining equitable transportation standards, particularly in collecting data that can be used to highlight and track inequalities such that standards can be set.
+
+1.
+**Understand what transport inequities you are measuring**.
+Common measures used in transport systems are accessibility measures.
+What does the measure mean and does potential access relate to experienced outcomes?
+Next, what does an un equitable and equitable arrangement look like ?
+This is context specific and locality sensitive.
+Global north has different standards than global south.
+
+2.
+**Understand what institutional context we’re operating in**.
+For example, what is the structure of transport planning.
+What Does WHO guidelines make sense to use as a standard for the community's context?
+
+3. **Understand what institutional context transfer looks like**. A standard set in one context, may not transfer to another community in the same region.
+
+4.
+**And be clear with our terms – what is equity and for whom?**.
+If '15 minute city' is the policy goal, what conceptualization is driving this standard?
+Is it sufficientarism?
+Egalitariansm?
+The standard can be interpreted a multitude of ways.
+The conceptualization needs to guide the standard.
+Being explicit about the conceptualization is important.
 
@@
 
-The following are additional details.
-
 #### The search strategy:
-
 ```{r Search-query-A1}

### 2023-10-18 Manuscript/Manuscript.qmd
Further edits to manuscript and renv lock updated

--- 2023-10-05 Manuscript/Manuscript.qmd
@@
 Meanwhile, firmly grounded on planet Earth, the employees that Bezos thanked for his suborbital escapade were struggling with some very mundane problems of their own, none of them as lofty as conflicting schedules that prevented them from flying in rockets [@griffinPerson2021].
-According to reports, people employed directly or indirectly by Amazon for warehouse or delivery work had, for years, been treated to "inhumane" conditions [@fungEuropean2018; @scottAmazon2019; @greeneAlgorithm2021], and subjected to surveillance on the job, crushing demands for productivity quotas that led to injury, facing little or no job security, and more dealt with as disposable inputs to Amazon's targets and consumers' desires [@tungAmazon2020; @reeseTeamsters2022; @middletonRobot2023].
+According to reports, people employed directly or indirectly by Amazon for warehouse or delivery work had, for years, been treated to "inhumane" conditions [@fungEuropean2018; @scottAmazon2019; @greeneAlgorithm2021], and subjected to surveillance on the job, crushing demands for productivity quotas that led to injury, facing little or no job security, and treated less with gratitude than as disposable inputs to Amazon's targets and consumers' desires [@tungAmazon2020; @reeseTeamsters2022; @middletonRobot2023].
 
@@
 The rocket that took Bezos to the edge of space is a somewhat rare example of a transportation technology, a tool of space-time convergence.
-By enabling movement at very high speeds rockets might—one day, at some indeterminate point in the future—prove essential to the expansion of the human species beyond our home planet.
-But in the present moment, the benefits of a private suborbital flight are for a few (e.g., the joy of movement, the sense of adventure, the awe of seeing Earth from space), whereas the burdens (e.g., the use of non-renewable resources, the climate-altering emissions) affect us all, and not evenly at that.
+By enabling movement at very high speeds, rockets might—one day, at some indeterminate point in the future—prove essential to the expansion of the human species beyond our home planet.
+But in the present moment, the benefits of a private suborbital flight (e.g., the joy of movement, the sense of adventure, the awe of seeing Earth from space) are for a few, whereas the burdens (e.g., the use of non-renewable resources, the climate-altering emissions) affect us all, and not evenly at that.
 In a plutocratic society, those with most can argue that this distribution of burdens and benefits is fair since both benefits and burdens are earned.
@@
 Unlike income, which is as unidimensional as it gets, the benefits and burdens of transportation systems are diffuse over space and time.
-For example, transportation systems engineered to offer higher mobility for people _somewhere_, can simultaneously cut others off essential opportunities _elsewhere_ , as @rajeEngineering2004 poignantly illustred with examples of infrastructure in the UK.
+For example, transportation systems engineered to offer higher mobility for people _somewhere_, can simultaneously cut others off from essential opportunities _elsewhere_, as @rajeEngineering2004 poignantly illustrated with examples of infrastructure in the UK.
 Furthermore, the shades of policies past can continue to haunt a region and even the planet for decades or longer, as shown by the legacy of displacement and decay caused by urban highways all across the US [@archerWhite2020] and the time horizon for the impacts of climate change to be fully felt [@markolfTransportation2019].
@@
 
-While commendable, the strategies are sufficiently vague that they can be implemented in a myriad ways, either to comply with the spirit of justice, or instead performatively to deceive it [@mcculloughPerformative2023].
+While commendable, the strategies are sufficiently vague that they can be implemented in a myriad ways, genuinely to comply with the spirit of justice, or else performatively to deceive it [@mcculloughPerformative2023].
 Some relevant questions include: how much should the adverse effects be reduced?
@@
 These questions boil down to the development and use of _standards_ for transportation justice.
-A definition of "standard" is as "something set up and established by authority as a rule for the measure of quantity, weight, extent, value, or quality"^[https://www.merriam-webster.com/dictionary/standard].
-How much pollution is allowed to be generated, and where, depends on who is affected, as well as on how much health is valued and by whom.
-For example, firms may or may not adopt lower standards for the emission of pollutants in poorer areas, or poor people may end up being relegated to areas that adopt lower standards [@gouldsonFirms2006].
+The term "standard" is used to mean "something set up and established by authority as a rule for the measure of quantity, weight, extent, value, or quality"^[https://www.merriam-webster.com/dictionary/standard].
+How much pollution is allowed to be generated, and where, depends on who is affected, and how much health is valued as well as by whom.
+For example, firms may or may not adopt lower standards for the emission of pollutants in poorer areas; or, possibly it is poor people who end up being relegated to areas that adopt lower standards [@gouldsonFirms2006].
 The result is the same: pollution tends to be worse were poorer people are [@delucaCode2012].
 
-Supporting the creation of (more) just transportation systems involves understanding the production, distribution, and management of transportation benefits and costs; how they are distributed; and what values are implemented (and by whom) in the form of standards [@pereiraDistributiveJusticeEquity2017a; @shellerMobilityJusticePolitics2018; @pereiraTransportationEquity2021].
+Supporting the creation of (more) just transportation systems involves understanding the production, distribution, and management of transportation benefits and costs; how they are distributed; and what values are implemented (as well as by whom) in the form of standards [@pereiraDistributiveJusticeEquity2017a; @shellerMobilityJusticePolitics2018; @pereiraTransportationEquity2021].
 Thus, for this review we engage the literature with the following questions in mind: 1) what is our current understanding of the things that transportation systems do, for whom, and to whom; 2) what does the literature say about the distribution of the benefits and costs of transportation systems; 3) and what values are embodied in normative statements about said distribution.
 Ultimately, this review aims to collate the academic knowledge and present it in a manner useful to support both academics and decision-makers in the development and implementation of standards for equity in transportation planning and policy.
-In this, we aim to provide relief even, or particularly, in those places where legislation explicitly calls for justice^[[As recently as -@martensFair2021, Martens and Golub note that Federal directives related to Title VI in the US "do not provide guidelines that can help agencies develop explicit standards to assess the distribution of accessibility benefits from projects or plans."]].
+In this, we aim to provide relief even, or particularly, to planners in those places where legislation explicitly calls for justice^[[As recently as -@martensFair2021, Martens and Golub note that Federal directives related to Title VI in the US "do not provide guidelines that can help agencies develop explicit standards to assess the distribution of accessibility benefits from projects or plans."]].
 
@@
 
-## A framework for questions of justice
+## A framework to analyze questions of justice
 

### 2023-10-19 Manuscript/Manuscript.qmd
Further edits by Anastasia and Antonio

@@
   geom_text(aes(label = total_perc_in_all %>% percent(accuracy = 0.1)), color = "white", size = 2.5) +
-  geom_vline(xintercept=4.5, size=1.5, color = "#191970") +
-  geom_hline(yintercept = 43.5, size=1.5, color="#191970")+
+  geom_vline(xintercept=4.5, linewidth=1.5, color = "#191970") +
+  geom_hline(yintercept = 43.5, linewidth=1.5, color="#191970")+
   theme_classic() +
@@
 
-It was a common kind of morning in West Texas but for the fiery wake of a rocket that could be seen ascending, piercing the blue skies.
-The date was July 20, 2021, and on board was a small group of four passengers that included Jeff Bezos, the founder of Amazon and then the world's richest person [@harwoodBezos2021].
-The mission that morning was among the first ever crewed private suborbital flights, and the adventure (described as "intense" by a member of the crew) lasted a total of 10 minutes and 10 seconds [@harwoodBezos2021].
+The fiery wake of a rocket could be seen ascending, piercing the morning skies of West Texas.
+It was July 20, 2021, and on board was a small group of four passengers that included Jeff Bezos, the founder of Amazon and then the world's richest person [@harwoodBezos2021].
+The mission that day was among the first-ever private suborbital passenger flights, and the adventure (described as "intense" by one of the passengers) lasted a total of 10 minutes and 10 seconds [@harwoodBezos2021].
 In addition to intense, the undertaking was expensive: a seat for the flight had previously been auctioned for no less than $28 million USD [@griffinPerson2021].
-Before even 10:00 am (EDT) Bezos was back from his excursion, and right on time to declare that this was the "[b]est day ever".
+Before even 10:00 am (EDT), when Bezos was back from his excursion, he took time to declare that this was the "[b]est day ever".
 To reporters covering the event he said "I want to thank every Amazon employee and every Amazon customer, because you guys paid for all of this" [@johnsonRoadToSpace2021].
 
-Meanwhile, firmly grounded on planet Earth, the employees that Bezos thanked for his suborbital escapade were struggling with some very mundane problems of their own, none of them as lofty as conflicting schedules that prevented them from flying in rockets [@griffinPerson2021].
-According to reports, people employed directly or indirectly by Amazon for warehouse or delivery work had, for years, been treated to "inhumane" conditions [@fungEuropean2018; @scottAmazon2019; @greeneAlgorithm2021], and subjected to surveillance on the job, crushing demands for productivity quotas that led to injury, facing little or no job security, and treated less with gratitude than as disposable inputs to Amazon's targets and consumers' desires [@tungAmazon2020; @reeseTeamsters2022; @middletonRobot2023].
+Meanwhile, firmly grounded on planet Earth, the employees that Bezos thanked for his suborbital escapade were struggling with some very mundane problems of their own, and none as lofty as conflicting schedules that prevented them from flying in rockets [@griffinPerson2021].
+According to reports, people employed directly or indirectly by Amazon for warehouse or delivery work had, for years, been treated to "inhumane" conditions [@fungEuropean2018; @scottAmazon2019; @greeneAlgorithm2021], and subjected to surveillance on the job, crushing demands for productivity quotas that led to injury, facing little or no job security, and treated less with gratitude than as disposable inputs to feed Amazon's earnings and consumers' desires [@tungAmazon2020; @reeseTeamsters2022; @middletonRobot2023].
 
@@
 Few reporters saw it fit to ask what problems the billionaire planned to solve on Earth, or what kind of future Bezos was trying to build, and for whose children.
-In other words, the billionaire was not confronted with the question of what did his trip do _for whom_ and _to whom_.
+In other words, the billionaire was not confronted with questions about what did his trip do _for whom_ and _to whom_.
 A less dispassionate observer might have been excused for wondering (possibly aloud) about the basic _fairness_ of a man amassing a nigh unimaginable fortune that allowed him to build and fly his own spaceship, while masses of his employees were treated as throwaway cogs in the vast apparatus of his logistics empire.
 
-The question of _fairness_, however, is not a simple one, and many of us would have been stumped to explain in a clear and precise way, beyond a disquieting intuition, just why the above picture was disturbing.
-_Justice_, is a political ideal based on the principle that individuals should be treated in a fair and equitable manner [@gosslingUrban2016, p.
+The question of _fairness_, however, is not a simple one, and most of us would probably have been stumped to explain in a clear and precise way, beyond a disquieting intuition, just why the above picture was disturbing.
+_Justice_ is a political ideal based on the principle that individuals should be treated in a fair and equitable manner [@gosslingUrban2016, p.
 2], giving and receiving whatever they are _due_ [@jaggarPhilosophicalChallengesGlobal2009, pp.
@@
 Values,in turn, are not subject to natural laws, but rather are the result of intersubjective realities, which is to say illusions whose legitimacy derives from a collective will to believe.
-For example, justice would likely mean something very different in a democratic society, if compared to a society where most owed their all—perhaps including their own time or bodily autonomy—to some collective illusion (e.g., the state, a monarch).
+For example, justice would likely mean something very different to a person in a democratic society, than to another in a society where they owed their all—perhaps including their own time or bodily autonomy—to some collective illusion (e.g., the state, a monarch).
 As well, the meaning of "justice" would likely differ in yet another society where very few owned most, and most owned very little due to a different illusion (e.g., that wealth equates merit).
@@
 But in the present moment, the benefits of a private suborbital flight (e.g., the joy of movement, the sense of adventure, the awe of seeing Earth from space) are for a few, whereas the burdens (e.g., the use of non-renewable resources, the climate-altering emissions) affect us all, and not evenly at that.
-In a plutocratic society, those with most can argue that this distribution of burdens and benefits is fair since both benefits and burdens are earned.
-The fact that the likes of Bezos can travel to space is proof that the likes of Bezos deserve those trips.
+The benefits of public transportation, a much more common transportation technology, are for most, but in many cases we have penny-pinched these systems while concentrating the costs on those who have less [@allenSuburbanization2020; @kaeorueanAnalysis2020].
+In a plutocratic society, those with most can (and often do argue^[See the so-called [Techno-optimist Manifesto](https://a16z.com/the-techno-optimist-manifesto/)]) that this distribution of burdens and benefits is fair since both benefits and burdens are earned.
+The fact that the likes of Bezos travel to space is proof that the likes of Bezos are due those trips.
 In a democratic society, the members of the collective might actually agree that large rewards (e.g., space flight) must be offered to highly qualified individuals (e.g., Bezos) to entice them to take important responsibilities (e.g., founding and leading Amazon).
-The members of the collective might think this a fair state of affairs, having come to the conclusion of their own accord or after being persuaded by billionaires.
-But then again, the members of the collective might decide otherwise: the _values_ of those in charge of defining what is "fair" matter.
-
-Multiple national and cross-national studies suggest that people in many societies have some tolerance for inequality [@kiatpongsanHow2014]: indeed, it appears that _some_ stratification, as suggested by @davisSome1945, is thought to fulfill a valuable function.
+The fact that some must give up jobs because it takes them too long to reach them by public transportation would constitute proof that those people should have studied more, been earlier risers, worked harder [@SpurrWaiting2015; @greismanLittle2017].
+Again, the members of the collective might think this a fine state of affairs, having come to this conclusion of their own accord or after being persuaded by billionaires.
+Or, contrariwise, the members of the collective might decide that this state of affairs is _unjust_: the _values_ of those in charge of defining what is "fair" matter.
+
+Multiple national and cross-national studies suggest that people in many societies do indeed have some tolerance for inequality [@kiatpongsanHow2014]: it would appear that _some_ stratification, as suggested by @davisSome1945, is perceived as fulfilling a valuable function.
 However, extreme inequality is often frowned upon, and can lead to social dysfunction and other ills [@acemogluWhy2000; @taydasCan2012; @duIncome2019; @houleInequality2022].
@@
 
-Transportation systems are a class of essential technologies that facilitate or impede social inclusion and activity participation by the public [@churchTransport2000; @lucasTransport2001; @socialExclusion2003; @cassSocial2005; @casasSocial2007; @prestonAccessibility2007; @paezMobility2009].
-As such, they increasingly are of interest from the perspective of equity, and a lively and rapidly growing literature has emerged in response [see _inter alia_ @martensTransport2016; @diciommoTransport2017; @guoSystematicOverviewTransportation2020; @karnerFrom2020; @vecchioTransportEquityLatin2020; @pereiraTransportationEquity2021; @vanweeEvaluating2021; @zhangLiterature2021; @desjardinsExaminingEquityAccessibility2022; @karnerAdvancesPitfallsMeasuring2023].
+Transportation systems, as a class of essential technologies that facilitate or impede social inclusion and activity participation [@churchTransport2000; @lucasTransport2001; @socialExclusion2003; @cassSocial2005; @casasSocial2007; @prestonAccessibility2007; @paezMobility2009], have increasingly come into focus from the perspective of equity.
+In response to this focus, a lively and rapidly growing literature has emerged on the topic [see _inter alia_ @martensTransport2016; @diciommoTransport2017; @guoSystematicOverviewTransportation2020; @karnerFrom2020; @vecchioTransportEquityLatin2020; @pereiraTransportationEquity2021; @vanweeEvaluating2021; @zhangLiterature2021; @desjardinsExaminingEquityAccessibility2022; @karnerAdvancesPitfallsMeasuring2023].
 A cynical rationale for this interest could be that keeping track of actual and perceived inequalities can serve _at the very least_ as a gauge of social discontent [as Chilean authorities discovered to their sorrow when unrest erupted over the increase of metro fares in 2019; @BBCNews2019; @diazpabonInequality2021].
-More optimistically, in democratic systems, there could be a genuine interest in being ready to respond to popular demands.
+More optimistically, in democratic systems, there could be a genuine interest in being ready to respond to popular demands for fairness.
 Several challenges arise when approaching this endeavor, caused by the notorious complexity of transportation systems: simultaneously, they move people, goods, and information.
@@
 
-But complexity is no reason to shirk the task.
-
-The objective of this review is to investigate the state of knowledge in terms of defining and operationalizing "fairness" in the transportation domain.
-Much research has been devoted to the issues of _measuring _ equity in transportation, including @ramjerdiEquity2006, @delboscUsing2011, @welchMeasure2013, @martensMeasuring2019, and @pritchardNew2022, among many others.
+But complexity is no excuse to shirk the task.
+
+The objective of this report is to scan the state of knowledge in terms of defining and operationalizing "fairness" in the transportation domain.
+Much research has been devoted to the issues of _measuring _ equity in transportation, including (among many others) @ramjerdiEquity2006, @delboscUsing2011, @welchMeasure2013, @martensMeasuring2019, and @pritchardNew2022.
 Further, there are multiple sources that discuss the conceptual and philosophical foundations of equity and fairness in transportation [e.g., @martensTransport2016; @pereiraDistributiveJusticeEquity2017a; @vanoutriveHow2019].
 Finally, previous reviews of planning documents have investigated equity from narrowly scoped perspectives, such as accessibility [@boisjolyHowGetThere2017] or a particular mode of transportation [e.g., cycling; @doranPursuitCyclingEquity2021].
-These inquiries are valuable to planning agencies and decision-makers, and the present review will tread similar, but not completely overlapping ground since, in our estimation, there is a gap in the literature in terms of understanding how standards for equity are developed and implemented in the transportation domain.
+These inquiries are valuable to scholar, planning agencies, and decision-makers alike, and the present review will tread similar, but not completely overlapping ground.
+In our estimation, there remains a gap in the literature in terms of understanding how standards for equity are developed and implemented in the transportation domain.
 In contrast, we do know that adoption of equity concepts in planning practice has lagged developments in academic work [@pereiraTransportationEquity2021; @boisjolyHowGetThere2017; @doranPursuitCyclingEquity2021; @linovskiEquityTransportationPlanning2020; @litmanEvaluatingTransportationEquity2022].
@@
 
-While commendable, the strategies are sufficiently vague that they can be implemented in a myriad ways, genuinely to comply with the spirit of justice, or else performatively to deceive it [@mcculloughPerformative2023].
+While commendable, the strategies are too vague, which makes it possible to implement them in a myriad ways, either genuinely to comply with the spirit of justice, or else performatively to deceive it [@mcculloughPerformative2023].
 Some relevant questions include: how much should the adverse effects be reduced?
@@
 What are the criteria for deciding that a community is "potentially affected"?
-What benefits should be distributed, and should those be based on simple population weights?
+What benefits should be distributed?
+Should the benefits be based on simple population weights?
 Or, contrariwise, should more deprived individuals be eligible for a larger share of the benefits?
@@
 The term "standard" is used to mean "something set up and established by authority as a rule for the measure of quantity, weight, extent, value, or quality"^[https://www.merriam-webster.com/dictionary/standard].
-How much pollution is allowed to be generated, and where, depends on who is affected, and how much health is valued as well as by whom.
-For example, firms may or may not adopt lower standards for the emission of pollutants in poorer areas; or, possibly it is poor people who end up being relegated to areas that adopt lower standards [@gouldsonFirms2006].
-The result is the same: pollution tends to be worse were poorer people are [@delucaCode2012].
-
-Supporting the creation of (more) just transportation systems involves understanding the production, distribution, and management of transportation benefits and costs; how they are distributed; and what values are implemented (as well as by whom) in the form of standards [@pereiraDistributiveJusticeEquity2017a; @shellerMobilityJusticePolitics2018; @pereiraTransportationEquity2021].
+How much pollution is allowed to be generated, and where, depends on who is affected, and how much health is valued overall, as well as by whom.
+For example, firms may or may not adopt lower standards for the emission of pollutants in poorer areas; it might be that poor people end up being relegated to areas that adopt lower standards [@gouldsonFirms2006].
+Regardless of the cause, the result is the same: pollution tends to be worse were poorer people are [@delucaCode2012].
+
+Supporting the creation of (more) just transportation systems involves understanding the production, distribution, and management of transportation benefits and costs; how they are distributed; and what values are implemented (and by whom) in the form of standards [@pereiraDistributiveJusticeEquity2017a; @shellerMobilityJusticePolitics2018; @pereiraTransportationEquity2021].
 Thus, for this review we engage the literature with the following questions in mind: 1) what is our current understanding of the things that transportation systems do, for whom, and to whom; 2) what does the literature say about the distribution of the benefits and costs of transportation systems; 3) and what values are embodied in normative statements about said distribution.
-Ultimately, this review aims to collate the academic knowledge and present it in a manner useful to support both academics and decision-makers in the development and implementation of standards for equity in transportation planning and policy.
+Ultimately, this review aims to collate the existing academic knowledge on the matter, and present it in a manner useful to support the development and implementation of standards for equity in transportation planning and policy.
 In this, we aim to provide relief even, or particularly, to planners in those places where legislation explicitly calls for justice^[[As recently as -@martensFair2021, Martens and Golub note that Federal directives related to Title VI in the US "do not provide guidelines that can help agencies develop explicit standards to assess the distribution of accessibility benefits from projects or plans."]].
 
+The rest of this report is structured as follows.
+
+After this introduction, we set the stage for our investigation by laying out some important definitions. We then describe the methods used for searching and reviewing the literature. 
 <!--
-This paper is structured as follows.
 In Section 2, we set the stage by defining what do we mean by equity dimensions, conceptualizations and types of standards used in the literature search and throughout the paper.
@@
 
-## What is justice?
+## Justice, equity, fairness, and standards
 
@@
 
+<!-- ADD: justice is also fluid, because it depends on the desirability of different states of affairs, which may change between populations or over time. -->
+
 Justice is attained when people "give and receive whatever they are due" [@jaggarPhilosophicalChallengesGlobal2009, pp.
@@
 **Distributive justice**.
-Perhaps the most commonly studied form of justice [see @jaggarPhilosophicalChallengesGlobal2009, p.
-2; @pereiraDistributiveJusticeEquity2017a], and its main concern is the way the benefits and burdens of the tangible and intangible products of society.
+This is perhaps the most commonly studied form of justice [see @jaggarPhilosophicalChallengesGlobal2009, p.
+2; @pereiraDistributiveJusticeEquity2017a], and its main concern is the way the benefits and burdens of the tangible and intangible products of society are collected by different segments of a population.
 
@@
 Retributive justice, for example, is often concerned with the distribution of the benefits and burdens of being a member of society: the way it is usually achieved is by distributing intangibles like "freedom" (e.g., of movement, of association) as benefits, and/or claiming tangibles like money as burdens (e.g., fines).
-Reparative justice often distributes current benefits and burdens to redress past wrongs, for example by asking those who have benefited from past wrongs, even if unwittingly, to shoulder a bigger fiscal burden in order to cover programs that mete benefits to those who are harmed by past wrongs.
+Reparative justice often distributes current benefits and burdens to redress past wrongs, for example by asking those who have benefited from past wrongs, even if unwittingly, to shoulder a bigger fiscal burden in order to cover programs that mete benefits to those who are still harmed by past wrongs.
 Procedural justice is the distribution of the benefits (e.g., the right to voice an opinion) and burdens (e.g., the effort required to develop an educated opinion) of the processes that lead to decisions with collective consequences.
 
-We can then speak of the purposes of distributive justice: to mete out retribution _fairly_, to repair past harms, and to ensure that procedures offer _equitable_ opportunities for engagement.
+We can then speak of the _purposes_ of distributive justice: to mete out retribution _fairly_, to repair past harms, and to ensure that procedures offer _equitable_ opportunities for engagement.
 Equity and fairness are the instruments of justice, the tools by which society advances towards the end goal of justice.
-
-<!-- Furthermore, it is important to distinguish between equity and justice. -->
 
@@
 Here, we are in agreement with @karnerTransportationEquityTransportation2020 that equity analysis should not be seen as an end in and of itself, but rather as a means to gather information about actual and perceived inequality.
-In this respect, the analytical tradition of equity, at least in transportation planning, means that the relevant models become embedded in the "political ecology of the estimated truth" [@kingModels1993]: in principle their assumptions and scope must be open and transparent, but they are also vulnerable to misuse.
-
-Fairness, in contrast to equity, is more complicated to define.
+In this respect, the analytical tradition of equity, at least in transportation planning, means that the relevant models become embedded in the "political ecology of the estimated truth" [@kingModels1993]: in principle their assumptions and scope must be open and transparent, but they are also vulnerable to misuse and even abuse, as tools of subjugation.
+
+Fairness, in contrast to equity, is somewhat more complicated to define.
 The concept does not have the same history of development as an analytical tool, and can be interpreted in numerous, and possibly discordant ways.
 That this is the case is convincingly demonstrated by @martensFair2021 in their study of the application of Title VI of the Civil Rights Act of 1964 in accessibility planning practice in the US.
-Title VI explicitly talks about the distribution of benefits derived from Federal funding: "[n]o person in the United States shall, on the ground of race, color, or national origin, be excluded from participation in, be denied the benefits of, or be subjected to discrimination under any program or activity receiving Federal financial assistance." However, as @martensFair2021 show, there are several ways to comply with regulations while achieving different outcomes: from the banal (do not _knowingly_ discriminate), to the substantive (compensation for past discrimination, i.e., reparative justice).
-What justice does fairness serve in each case?
-It depends on what was the reason for seeking justice in the first place!
-Our reading of @martensFair2021 is that fairness is a yardstick that is best used _a priori_ than _a posteriori_, for doing so risks rationalizing the outcomes rather than
-
-The focus of this report is on standards of fairness that, combined with the use of equity analysis, can help understand how better to move towards just transportation outcomes.
+Title VI explicitly talks about the distribution of benefits derived from Federal funding: "[n]o person in the United States shall, on the ground of race, color, or national origin, be excluded from participation in, be denied the benefits of, or be subjected to discrimination under any program or activity receiving Federal financial assistance." However, as @martensFair2021 show, there are several ways to comply with regulations while achieving different outcomes that range from the banal (do not _knowingly_ discriminate), to the substantive (compensation for past discrimination, i.e., reparative justice).
+What kind of justice does fairness serve in each case?
+It depends on what was the reason for seeking justice in the first place.
+Our reading of @martensFair2021 is that fairness is a yardstick that is best deployed _a priori_ than _a posteriori_, for doing the latter risks rationalizing the outcomes rather than driving them.
+
+<!--
+NEEDS MORE ON STANDARDS.
+-->
+
+The focus of this report is on standards of fairness that, combined with the use of equity analysis, can help us understand how better to move towards just transportation systems.
 
@@
 
-Here we are inspired by the framing of @jaggarPhilosophicalChallengesGlobal2009 for philosophical questions of justice^[Similar questions are found peppered throughout the literature either explicity, as for example in @karnerTransportationEquityTransportation2020, who ask "of what", "for whom," and "how much" in reference to equity, or implicitly, as in @gosslingUrban2016, who asks "what?" (exposure, space, access) and "for whom?" (gender, age, ethnicity) about the outputs of transportation.].
+For this report, we are inspired by the framing of @jaggarPhilosophicalChallengesGlobal2009 for philosophical questions of justice^[Similar questions are found peppered throughout the literature.
+This is done either explicitly, as for example in @karnerTransportationEquityTransportation2020, who ask "of what", "for whom," and "how much" in reference to equity; or implicitly, as in @gosslingUrban2016, who asks of the outputs of transportation "what?" (exposure, space, access) and "for whom?" (gender, age, ethnicity).].
 According to -@jaggarPhilosophicalChallengesGlobal2009, Western philosophy has approached the issue of justice by asking "Where?", "When?", "Who?", "What?", and "How?".
@@
 3], a notion since expanded (e.g., with the Universal Declaration of Human Rights) to include all members of the human species, but has yet to cover other living creatures.
-
-<!--- EDITED UP TO HERE --->
-
-- "When?" asks what are the social circumstances within which the
-demands of justice have application.
-- "Who?" asks which entities should be regarded as subjects of justice,
-meaning who or what are entitled to make justice claims deserving of
-moral consideration.
-- "What?" asks which entities should be regarded as objects of justice,
-meaning which kinds or categories of things should be distributed in a
-just manner.
-- Finally, "How?" asks which principles are the most morally appropriate
-for guiding the allocation of various objects to various subjects in various
-circumstances.
-
-This categorization is outlined in  to describe substantive sets of answers to questions of distributive justice (and further discussed in the context of justice within the critical human geography discipline [@przybylinskiWhereJusticeGeography2022]).
-
-For questions of *justice*, convincing answers to these questions require a rationale: a "Why?" [@jaggarPhilosophicalChallengesGlobal2009].
-For the purpose of this review focused on *equity*, the focus is on studies that substantively answer "Where?", "When?", "Who?", "What?", and "How?"; the answer to "Why?" may be incidental.
-
-## Transportation and justice
-
-To understand the distributional implications of transportation systems, it is essential that we are clear about what they do. What do transportation systems _produce_?
-
+<!-- SOUTH AMERICAN CONSTITUON THAT DOES INCLUDE IT NOTE FOR ANASTASIA--> In the case of justice in transportation, the question of "where?" is paramount; it might be argued that, by their very nature, transportation generates inequalities.
+By concentrating the effects of space-time convergence (for instance, by providing access to a transit system or a highway), an inequality is automatically generated.
+The burdens of transportation, in contrast, are often diffuse.
+They are incrementally paid, for example by a distributed population in the form of taxes by a spatially distributed population, or in poor health by people who were not party to transportation decisions made elsewhere.
+
+- Conventionally, the question of **"When?"** refers to the situation, that is, the social circumstances within which the demands of justice have application.
+In our case, we might as well ask about the temporal aspects of transportation systems, which include _when_ to invest in transportation [e.g., @rabelloEvaluation2015] (and as a result when to generate a spatial inequality), for how long the burdens and benefits can still be associated to a specific intervention, and whether claims to reparative justice expire over time.
+<!-- WHEN IN TIME SOMETHGING HAPPENS TOO - WHAT INVESTMENTS HAPPENED IN THE PAST THAT CAST SHADOWS-->
+
+- When we ask **"Who?"**, we inquire about which entities should be regarded as subjects of justice <!-- OF INEQUITIES / MEASURING THE INEQUITIES OF 'WHO'.
+LIKE INCOME, THE MOBILITY TOOL IS MODIFIABLE -- THE MOBILITY TOOL IS NOT MODIFIABLE IN MANY CASES, IT IS WHO YOU ARE .
+T IS ALMOST LIKE A CHARACTERISTICS OF THE PERSON.
+PART OF THE SOCIAL CONTRACT THAT IDENTIFIES EOPEL WITH THEIR MOBILITY TOOL-->, meaning who are entitled to make claims that deserve moral consideration from the perspective of justice.
+To make it tractable, this question is often approached through the filter of population groups, which may include the several traits, including gender identity, ableness, ethnicity, age, caste, and income.
+Often, it is appropriate to reflect on the intersections between traits, given evidence that the lived experiences of, say, a White woman and a Black women, can be markedly different.
+A possible complication in the case of transportation is that disentangling the "who" from the tools available to them is not always straightforward.
+Clearly, a person is not their mode of transportation; in practice, though, there are large segments of the population who live in situations where they cannot extricate themselves from the tools for mobility that they can use, either because they have driven themselves out of choices [see @laveryDriving2013], or have been driven out of choices by factors beyond their control, effectively ending up as captive users of a single mode [@jacquesRescuing2012; @cherancheryInvestigatingPerceptionCaptive2018].
+In societies that have grown into transportation monocultures with a predilection for automobility [@millerCollaborative2011] there may actually be less choice about mobility tools than we would like to assume.
+So, while it is important to avoid conflation of the "who" with the "what",  for analytical purposes we need to be mindful of the connection between person and their mobility tools.
+
+- **"What?"** asks which entities should be regarded as objects of justice, meaning which kinds or categories of things should be distributed in a just manner.
+To understand the distributional implications of transportation systems, it is essential that we are clear about what they do.
+In other words, what do transportation systems _produce_?
 At their most fundamental, transportation systems are space-time convergence tools, tools that improve the rate at which time is traded for space.
 They usually do this by increasing the speed of movement: sidewalks facilitate walking, traffic lights facilitate the ordered flow of vehicular traffic, and a launching pad makes it possible for a rocket to takeoff.
-With complex interlocking parts (sidewalk, road, traffic sign, parking regulations), transportation systems produce _mobility_, the potential for movement.
-Combined with land use systems (the spatial distribution of opportunities on the landscape), they produce _accessibility_, the potential to reach destinations.
-The realization of those potentials happens through travel.
-But, as the adage goes, travel is derived demand, which seems to hold for most (even if not all) situations [ref mokhtarian].
+With complex interlocking parts (sidewalk, road, traffic sign, parking regulations), transportation systems produce _mobility_, the potential for movement.The realization of this potentials happens through travel but, as the adage goes, travel is derived demand, which seems to hold for most (even if not all) situations [e.g., @mokhtarianUnderstanding2001; @redmondPositive2001; @paezEnjoyment2010; @whalenMode2013].
+For this reason, we cannot stop at considering mobility, but the ultimate goal of mobility, which is to reach destinations.
+In combination with land use systems (the spatial distribution of opportunities on the landscape), mobility produces _accessibility_, the potential to reach destinations.
+For this reason, we can think of the objects of transportation justice as being _proximate_ (the tools of mobility, mobility itself), and _ulterior_ (accessibility, opportunities for activity participation).
+The burdens of transportation are also many and varied.
+Some are direct and paid directly by the traveler (e.g., travel time, out-of-pocket costs), but many others are indirect and related to network externalities (e.g., exposure to pollution).
+
+<!-- ANASTASIA NEEDS TO REVIEW THE FOLLOWING -->
+
+- The next question is **"How?"**, and it relates to the allocation of various objects of justice ("what") to various subjects of justice ("who") in various circumstances.
+**Equity standards** are a tool for answering this distributive question: how do we allocate burdens and benefits and to whom?
+Standards are thresholds that when operationalized effectively define what is in/equitable.
+The thresholds can be quantitative, or they can be qualitative descriptions, and/or the result of mixed-method.
+Some examples include: maximum travel distance/cost/time to or from key destinations, levels of maximum exposure to externalities (i.e., noise or air pollution),un/fulfilled needs, and dis/satisfaction with travel.
+A number of theoretical and conceptual frameworks exist to support us when approaching this question, and we can draw from concepts in transport-related social exclusion, transport disadvantage, and/or transport poverty, which are typically based on equity principles, such as utilitarianism, Sen's capabilities approach, and sufficientarism.
+
+- Lastly, convincing answers to the above questions require a supporting rationale: a **"Why?"** [@jaggarPhilosophicalChallengesGlobal2009].
+This is perhaps the most slippery of all the questions posed here.
+Justice is an inherently social construct <!--, and does not exist outside of human society-->.
+Asking **why?** amounts to asking what sort of social contracts regulate human interactions.
+These contracts can be defined by constitution, but there are often unwritten and possibly contested variants.
+To give an example drawing from the Canadian Bill of Rights, a number of human rights and fundamental freedoms are recognized (including to life and liberty), "without discrimination by reason of race, national origin, colour, religion or sex".
+Notice that this declaration includes several "whos", but does not consider age as one aspect of the person.
+Does this mean that they apply universally independent of age?
+Surely not, since no reasonable person would consider a toddler's demands for freedom of association as an absolute.
+So, when do these rights fully apply in the life of a person?
+In 2017, Adrian Crook of Vancouver, B.C., was warned by the province's Ministry of Children and Family Development that his kids could not be out of home in the community, alone or with other kids the same age, without supervision.
+Crook argued that his goal was to raise self-reliant kids, partly by teaching his children (then aged 7, 8, 9 and 11) to use transit to go to school @stueckYoung2017.
+The Ministry's argument was that children riding the bus unsupervised compromised the parent's ability to provide care and was a risk.
+Statistically, though, there is evidence that travel by car is riskier than travel by bus [@morencyTraveling2018].
+If the public concern is about risk, should not children be banned from riding in cars until they are ten?
+In a functioning democracy like Canada, we have chosen to care about fundamental freedoms and rights.
+This provide an anser to the question of "why" thinking about justice.
+But, the devil is in the detail's and the other questions discussed are needed to pin the devil down by the tail.
+Are children under the age of 10 subjects of justice ("who")?
+If so, is the ability to use transit unsupervised an object of justice ("what")?
+When and where do these questions apply (e.g., riding the bus to school, or to a social event; during the day, or late at night)?
+Eventually, in 2020, a court of appeals ruled in Crook's favor, allowing his children to continue to ride the bus, thus clarifying a standard for equity with respect to the use of transit for children who desire independent mobility [@stueckCourt2020].
+
+<!-- IN THE PARAGRAPH ABOVE, ANASTASIA SUGGESTS DISCUSSING WHY DO WE THINK THAT OUTSIDE IS DANGEROUS, THAT CARS ARE SAFER, THAT CERTAIN PEOPLE ARE TO BE AVOIDED.
+
+THIS IS WHY WE ONLY FOCUS ON INEQUITIES. JUSTICE IS FLUID AND SLIPERLY. WHAT HAS BEEN DONE WHAT HAS NOT, -->
+
+<!--
+THIS BIT NOW READS LIKE A LOOSE END. DISCARD?
+
+This categorization is outlined in  to describe substantive sets of answers to questions of distributive justice (and further discussed in the context of justice within the critical human geography discipline [@przybylinskiWhereJusticeGeography2022]).
+
+-->
+
+<!--
+THESE THREE PARAGRAPHS SOUND LIKE A CONCLUSION.
+PERHAPS A CALL TO ACTINO - TO BE ABLETO MAE AN INTERNATIONAL COMPARISION ON JUSTICE.
+WE NEED TO KNOW THE WHY BUT LITERATURE SHOULD START TO OUTLINE THAT EXPLICILITY.
+-->
+
+For the purpose of this review focused on *equity*, the focus is on studies that substantively answer "Where?", "When?", "Who?", "What?", and "How?", while the answer to "Why?" may occasionally be incidental.
 
@@
 
+<!--
+MOVE TO CONCLUSIONS?
+-->
+
 ### Definitions
@@
 
+<!-- EQUITY DIMENSIONS: OBJECTS OF JUSTICE -->
+
 **Equity dimensions**: the assessment of transportation equity encompasses various dimensions, mirroring the multifaceted nature of transportation systems.
 These dimensions can be distinguished as the **"What?"**, specifically: (1) mobility/accessibility, (2) traffic-related pollution, (3) traffic safety, and (4) human-health as provided in a conceptual framework by @lucasMeasuringTransportEquity2019.
+
 All papers reviewed include dimensions that cover the **"What?"** of the transport equity issue and can be further disaggregated:
@@
 
-<!-- Is it appropriate to talk about a "population group"? Think about intersectionality theory.
-
-These are some things that people _are_:
-
-- Gender identity
-- Ableness
-- Ethnicity
-- Age
-- Caste
-- Low income (in societies with low social mobility)
-- Membership in a family
+<!--
 
 -->
-
--   *How*: includes the measure used to assess equity dimensions as supported by a **equity conceptualization** and **equity standard**.
-**Equity conceptualizations** are some type of motivating equity principle, philosophy or theory.
-For example: theoretical and conceptual frameworks (e.g., transport-related social exclusion, transport disadvantage, transport poverty, horizontal equity, vertical equity) and equity principles (e.g., Utilitarism, Capabilities Approach, Sufficientarism).
-**Equity standards** are thresholds that when operationalized can be used to define when an aspect of the transportation system is in/equitable.
-They can be quantitative thresholds, qualitative descriptions, and/or mixed-method thresholds.
-Examples include: maximum travel distance/cost/time to or from key destinations, levels of maximum exposure to externalities (i.e., noise or air pollution), distributional assessment (e.g., composite indicator, range, maximum distance or gap, etc.), dis/satisfaction with travel, etc.
 
@@
 
-# Literature review methods
+# Methods
 
@@
 **The "When" and "Where"**:
+
 The majority of papers (`r ((case_cont_count$count[4] + case_cont_count$count[5]+ case_cont_count$count[6])/sum(case_cont_count$count)) |> percent()`) focus on case studies in the Global North.
@@
 **The "Who"**:
-Though the literature reviewed most commonly considers low-income as a socio-demographic characteristics of populations deserving equity as either by itself or alongside other characteristics (`r df_pop$count[8]+df_pop$count[4] %>% percent()` of the papers).
+
+Though the literature reviewed most commonly considers low-income as a socio-demographic characteristics of populations deserving equity as either by itself or alongside other characteristics (`r prettyNum(round(df_pop$count[8]+df_pop$count[4] * 100, 2))`\% of the papers).
 Many employ accessibility methods (e.g., walking accessibility to open space [@tangWalkingAccessibilityNeighbourhood2021]) but a others use qualitative or mixed methods to gain perspectives of inequalities that are typically cross-dimensional (e.g., @milanLiftingPeripheralFortunes2017 asks Medellin residents living in TOD areas compared to non-TOD areas about the impact of TOD on their wellbeing based on income group and gender).
@@
 **The "What" and "How"**:
+
 Inequities that focus on accessibility/mobility -> opportunity standards are typically travel impedance standards, and affordability thresholds (e.g., more than 10% of monthly income should not be spent on transport [@rivasHowAffordableTransportation2018] to work and/or other necessary destinations).
@@
 
-## Call 3: More direct and explicity links between standards and experienced outcomes
+## Call 3: More direct and explicit links between standards and experienced outcomes
 

### 2023-10-20 Manuscript/Manuscript.qmd
Edits to "Setting the stage". Next up: "Methods"

@@
 
-<!-- From the original intro
-
-Benefits and burdens can include, but are not limited to, accessibility, mobility, affordability, health and environmental issues that impact individuals, marginalized members of a community, and the community as a whole [@lucasMeasuringTransportEquity2019].
-Academic literature has approached equity through measures such as accessibility and activity participation [@paezMeasuringAccessibilityPositive2012; @allenPlanningTransportSocial2020], affordability [@isalouTestingHousingTransportation2014], environmental impacts and safety [@guoSystematicOverviewTransportation2020], and health [@fransenIdentifyingPublicTransport2015; @smithSystematicLiteratureReview2017], and also conceptualizations such as frameworks for equity and social exclusion [@lucasProvidingTransportSocial2006; @lucasTransportPovertyIts2016], and distributive justice theories [@behbahaniDesigningRoadNetwork2019; @lewisExploringEquityHow2021; @pereiraDistributiveJusticeEquity2017a; @vecchioAccessibilityCapabilitiesApproach2021], among others.
-
--->
-
 ## Justice, equity, fairness, and standards
@@
 In the paragraphs above we used the terms "justice", "fairness", and "equity" in a relatively loose fashion.
-This was done purposefuly.
-As we noted, people often have a strong intution of what is "fair", "just", and "equitable".
+This was done purposefully.
+As we noted, people often have strong intuitions of what is "fair", "just", and "equitable".
 These conceptions may or may not match those of the authorities who set the standards of fairness.
 But in democratic societies, the authority of political leaders, bureaucrats, planners, and all those charged with the business of governing, derives from the will of the people.
-It is therefore important to make the implicit explicit, and clearly spell out those intuitions of "justice", "fairness", and "equity" to make them conceptually clear.
-Only this way can the participants of a democratic society be effective arbiters of what is just - and perhaps stand on firmer ground to experience love of their society.
+It is therefore important to explicitly state how we plan to use these words, to clearly spell out our intuitions of "justice", "fairness", and "equity".
+A clear mutual understanding of these concepts is essential for constructive debate, and for participants of a democratic society to be effective arbiters of what is just.
 
@@
 
-<!-- ADD: justice is also fluid, because it depends on the desirability of different states of affairs, which may change between populations or over time. -->
-
-Justice is attained when people "give and receive whatever they are due" [@jaggarPhilosophicalChallengesGlobal2009, pp.
+It is said that justice is attained when people "give and receive whatever they are due" [@jaggarPhilosophicalChallengesGlobal2009, pp.
 1-2], and it ceases to exist when there are persons or groups that are denied "access to the opportunities they need to lead a meaningful and dignified life" [@karnerTransportationEquityTransportation2020, p.
-440]
-
-Several forms of justice can be distinguished, including [see @jaggarPhilosophicalChallengesGlobal2009; @pereiraDistributiveJusticeEquity2017a; @karnerTransportationEquityTransportation2020]:
+440].
+Justice is a fluid concept, because it depends on the desirability of different states of affairs, which may change between populations or over time.
+That said, it is possible to distinguish several forms of justice, including [see @jaggarPhilosophicalChallengesGlobal2009; @pereiraDistributiveJusticeEquity2017a; @karnerTransportationEquityTransportation2020]:
 
@@
 
-We can then speak of the _purposes_ of distributive justice: to mete out retribution _fairly_, to repair past harms, and to ensure that procedures offer _equitable_ opportunities for engagement.
+We can then speak of the _purposes_ of distributive justice: to mete out retribution _fairly_, to repair past harms, and to ensure that procedures offer _equitable_ opportunities to influence outcomes.
 Equity and fairness are the instruments of justice, the tools by which society advances towards the end goal of justice.
@@
 In the transportation domain, the term is somewhat loaded because it is perceived as stemming from the authority of the state, and meant to assist with decisions about regulating and financing transportation spending [@karnerTransportationEquityTransportation2020].
-Here, we are in agreement with @karnerTransportationEquityTransportation2020 that equity analysis should not be seen as an end in and of itself, but rather as a means to gather information about actual and perceived inequality.
+Here, we are in agreement with @karnerTransportationEquityTransportation2020 that equity analysis should not be seen as an end in and of itself, but rather as a means to gather information about actual and perceived inequities.
 In this respect, the analytical tradition of equity, at least in transportation planning, means that the relevant models become embedded in the "political ecology of the estimated truth" [@kingModels1993]: in principle their assumptions and scope must be open and transparent, but they are also vulnerable to misuse and even abuse, as tools of subjugation.
@@
 That this is the case is convincingly demonstrated by @martensFair2021 in their study of the application of Title VI of the Civil Rights Act of 1964 in accessibility planning practice in the US.
-Title VI explicitly talks about the distribution of benefits derived from Federal funding: "[n]o person in the United States shall, on the ground of race, color, or national origin, be excluded from participation in, be denied the benefits of, or be subjected to discrimination under any program or activity receiving Federal financial assistance." However, as @martensFair2021 show, there are several ways to comply with regulations while achieving different outcomes that range from the banal (do not _knowingly_ discriminate), to the substantive (compensation for past discrimination, i.e., reparative justice).
+Title VI explicitly talks about the distribution of benefits derived from Federal funding: "[n]o person in the United States shall, on the ground of race, color, or national origin, be excluded from participation in, be denied the benefits of, or be subjected to discrimination under any program or activity receiving Federal financial assistance." However, as @martensFair2021 show, there are several ways to comply with regulations while achieving different outcomes, ranging from the banal (do not _knowingly_ discriminate), to the substantive (compensation for past discrimination, i.e., reparative justice).
 What kind of justice does fairness serve in each case?
@@
 
-<!--
-NEEDS MORE ON STANDARDS.
--->
-
-The focus of this report is on standards of fairness that, combined with the use of equity analysis, can help us understand how better to move towards just transportation systems.
+The last concept that we discuss in this section is that o a standard.
+Briefly, standards are a way of making concrete statements about fairness.
+Returning to the ambiguities in Title VI discussed by @martensFair2021, the attainment of justice depends on the standard used to indicate fairness.
+For example, explicit non-discrimination constitutes a very weak standard that takes aim at the actions of agencies instead of the recipients of the benefits; accordingly, any distribution of benefits would be considered fair, as long as the agency does not explicitly and knowingly targets or denies the benefits to particular groups.
+The standard provides conditions to determine whether a situation is _fair_.
+A similarly weak standard is a _Pareto improvement_, whereby it is possible to concentrate the benefits as long as no group is worse off compared to the status quo.
+A policy is fair as long it does no harm.
+A somewhat more strict standard, a Pareto-Plus improvement, stipulates that an intervention is fair when all groups receive at least _some_ (non-trivial) benefits; the size of the benefits for each group is irrelevant.
+In contrast to the notion of "do no harm", such a standard embodies the ideal that no one is denied benefits.
+An egalitarian standard would weigh the benefits or burdens by population, and fairness is achieved when each group give or receive in proportion to their size.
+An affirmative action standard is even stricter, since it requires the benefits to be distributed in a non-egalitarian way that favors those who are still harmed by past or present discriminatory practices.
+
+To quickly recap the discussion so far, justice is an end.
+But to understand what that end is, we must clearly define standards of fairness.
+Equity analysis is tool to measure where the actual or perceived distribution of the burdens and benefits of the products of transportation systems stand with respect to the standard, in other words, instruments to see how close or far we are from a just situation.
+
+In the following section we will discuss the analytical apparatus that we will use to interrogate the literature on equity standards in transportation.
 
@@
 T IS ALMOST LIKE A CHARACTERISTICS OF THE PERSON.
-PART OF THE SOCIAL CONTRACT THAT IDENTIFIES EOPEL WITH THEIR MOBILITY TOOL-->, meaning who are entitled to make claims that deserve moral consideration from the perspective of justice.
+PART OF THE SOCIAL CONTRACT THAT IDENTIFIES EOPEL WITH THEIR MOBILITY TOOL-->, meaning who are entitled to make claims of moral consideration from the perspective of justice.
 To make it tractable, this question is often approached through the filter of population groups, which may include the several traits, including gender identity, ableness, ethnicity, age, caste, and income.
@@
 So, while it is important to avoid conflation of the "who" with the "what",  for analytical purposes we need to be mindful of the connection between person and their mobility tools.
+In the case of transportatin, in addition to members of the public who use transportation systems, there is another category of **who**, that stand possibly in opposition to users, namely the entities charged with providing services, maintaining infrastructure, and so on.
+These could be ministries or departments of transportation, transit agencies, public works departments and others.
+Identifying these entities is relevant to ellucidate who is responsible for apportioning the benefits or mitigating the burdens of transportation.
 
@@
 Standards are thresholds that when operationalized effectively define what is in/equitable.
-The thresholds can be quantitative, or they can be qualitative descriptions, and/or the result of mixed-method.
+The thresholds can be quantitative (e.g., square meters of green space per capita), or they can be qualitative descriptions (e.g., do not knowingly discriminate), or a mixed of the two.
 Some examples include: maximum travel distance/cost/time to or from key destinations, levels of maximum exposure to externalities (i.e., noise or air pollution),un/fulfilled needs, and dis/satisfaction with travel.
@@
 These contracts can be defined by constitution, but there are often unwritten and possibly contested variants.
-To give an example drawing from the Canadian Bill of Rights, a number of human rights and fundamental freedoms are recognized (including to life and liberty), "without discrimination by reason of race, national origin, colour, religion or sex".
-Notice that this declaration includes several "whos", but does not consider age as one aspect of the person.
-Does this mean that they apply universally independent of age?
+To give an example drawing from the Canadian Bill of Rights, a number of rights and fundamental freedoms [including the liberty to move freely; see @chartopediaArt72023] are recognized to exist in Canada "without discrimination by reason of race, national origin, colour, religion or sex".
+Notice that this declaration includes several individual traits that define the subjects of justice, but does not consider age as one aspect of the person.
+Does this mean that they apply universally irrespective of age?
 Surely not, since no reasonable person would consider a toddler's demands for freedom of association as an absolute.
@@
 In 2017, Adrian Crook of Vancouver, B.C., was warned by the province's Ministry of Children and Family Development that his kids could not be out of home in the community, alone or with other kids the same age, without supervision.
-Crook argued that his goal was to raise self-reliant kids, partly by teaching his children (then aged 7, 8, 9 and 11) to use transit to go to school @stueckYoung2017.
-The Ministry's argument was that children riding the bus unsupervised compromised the parent's ability to provide care and was a risk.
-Statistically, though, there is evidence that travel by car is riskier than travel by bus [@morencyTraveling2018].
-If the public concern is about risk, should not children be banned from riding in cars until they are ten?
+The Ministry's argument was that children riding the bus unsupervised compromised the parent's ability to provide care and placed them at risk [@brendHere2017].
+Crook argued that the goal in teaching his kids (then aged 7, 8, 9 and 11) to use transit to go to school was to raise self-reliant children [@stueckYoung2017].
+While both sides discovered that there were not clear rules about kids riding transit alone [@brendHere2017], statistically there is evidence that travel by car is riskier than travel by bus [@morencyTraveling2018].
+If the public good is to reduce risk, should not children be banned from riding in cars until they are ten?
 In a functioning democracy like Canada, we have chosen to care about fundamental freedoms and rights.
-This provide an anser to the question of "why" thinking about justice.
-But, the devil is in the detail's and the other questions discussed are needed to pin the devil down by the tail.
-Are children under the age of 10 subjects of justice ("who")?
-If so, is the ability to use transit unsupervised an object of justice ("what")?
+This provides an answer as to "why" we would even consider children as subjects of justice.
+But the devil is in the details, and the other questions discussed above are needed to pin the devil by the tail.
+Are children under the age of 10 subjects of justice (i.e., are they a "who")?
+If so, is the ability to use transit unsupervised an object of justice (i.e., the "what")?
 When and where do these questions apply (e.g., riding the bus to school, or to a social event; during the day, or late at night)?
-Eventually, in 2020, a court of appeals ruled in Crook's favor, allowing his children to continue to ride the bus, thus clarifying a standard for equity with respect to the use of transit for children who desire independent mobility [@stueckCourt2020].
-
-<!-- IN THE PARAGRAPH ABOVE, ANASTASIA SUGGESTS DISCUSSING WHY DO WE THINK THAT OUTSIDE IS DANGEROUS, THAT CARS ARE SAFER, THAT CERTAIN PEOPLE ARE TO BE AVOIDED.
-
-THIS IS WHY WE ONLY FOCUS ON INEQUITIES. JUSTICE IS FLUID AND SLIPERLY. WHAT HAS BEEN DONE WHAT HAS NOT, -->
+Eventually, in 2020, a court of appeals ruled in Crook's favor, allowing his children to continue to ride the bus, thus providing much needed clarity with respect to the use of transit by children who desire independent mobility [@stueckCourt2020].
+In actuality, the courts provided a standard of equity: starting at a certain age, children are objects of justice from the perspective of freedom of movement, and they are due the benefits of independent mobility by transit.
+
+The focus of this report is on standards of fairness that, combined with the use of equity analysis, can help us understand how better to move towards just transportation systems.
 
 <!-- 
-THIS BIT NOW READS LIKE A LOOSE END. DISCARD?
-
-This categorization is outlined in  to describe substantive sets of answers to questions of distributive justice (and further discussed in the context of justice within the critical human geography discipline [@przybylinskiWhereJusticeGeography2022]).
-
+IN THE PARAGRAPH ABOVE, ANASTASIA SUGGESTS DISCUSSING WHY DO WE THINK THAT OUTSIDE IS DANGEROUS, THAT CARS ARE SAFER, THAT CERTAIN PEOPLE ARE TO BE AVOIDED.
+
+THIS IS WHY WE ONLY FOCUS ON INEQUITIES. JUSTICE IS FLUID AND SLIPPERY. WHAT HAS BEEN DONE WHAT HAS NOT. 
 -->
 
+# Methods
+
 <!--
-THESE THREE PARAGRAPHS SOUND LIKE A CONCLUSION.
-PERHAPS A CALL TO ACTINO - TO BE ABLETO MAE AN INTERNATIONAL COMPARISION ON JUSTICE.
-WE NEED TO KNOW THE WHY BUT LITERATURE SHOULD START TO OUTLINE THAT EXPLICILITY.
--->
-
-For the purpose of this review focused on *equity*, the focus is on studies that substantively answer "Where?", "When?", "Who?", "What?", and "How?", while the answer to "Why?" may occasionally be incidental.
-
-Transportation equity literature exists within cross-cutting categories.
-
-Though critical, answers to "Why?" are not overwhelming common in the transportation equity academic literature that also address  "Where?", "When?", "Who?", "What?", and "How?".
-This is discussed as a next step in planning for *justice* transportation literature in Section 6.
-
-<!--
-MOVE TO CONCLUSIONS?
--->
-
-### Definitions
+NOTE: This text was in the Definitions, but we decided that these definitions were a thin conceptual layer that could be removed.
+Instead of "equity dimensions", "conceptualization", and "standards" we will stick to using the where, when, who, what, how, and why conceptual framework.
+For example, the EQUITY DIMENSIONS are simply the "what" or "objects of justice".
+The "conceptualizations" are the "how".
+And so on.
 
@@
 
-<!-- EQUITY DIMENSIONS: OBJECTS OF JUSTICE -->
-
 **Equity dimensions**: the assessment of transportation equity encompasses various dimensions, mirroring the multifaceted nature of transportation systems.
@@
 -   *When* and *Where*: includes the article's publication year, the geographic context (i.e., urban, suburban and rural areas, countries) and location.
+
 -   *Who*: includes the type of population group, mode type, activity/opportunity type, and transport network status.
 
-<!--
-
+papers that include equity **dimensions**, **conceptualizations**, **standards** and substantive answers to the "When?", "Where?", and "Who?" in relation to equity in transportation systems (i.e., the users, the infrastructure, the service, the outcomes, or a combination of all) are used to answer the following research questions:
+
+-   What are the equity dimensions, equity standards, and equity conceptualizations and how are they applied in the existing transportation equity academic literature?
+
+-   Based on how equity has been applied in the literature, what may be paths forward in theorizing justice in the context of transportation systems
 -->
-
-papers that include equity **dimensions**, **conceptualizations**, **standards** and substantive answers to the "When?", "Where?", and "Who?" in relation to equity in transportation systems (i.e., the users, the infrastructure, the service, the outcomes, or a combination of all) are used to answer the following research questions:
-
--   What are the equity dimensions, equity standards, and equity conceptualizations and how are they applied in the existing transportation equity academic literature?
--   Based on how equity has been applied in the literature, what may be paths forward in theorizing justice in the context of transportation systems?
-
-# Methods
 
@@
 This is a key step towards transport justice; assessing the effects of policies on different dimensions and populations groups and evaluating if a specific context is moving towards equitable standards over time.
+
+<!--
+THESE THREE PARAGRAPHS SOUND LIKE A CONCLUSION.
+PERHAPS A CALL TO ACTINO - TO BE ABLETO MAE AN INTERNATIONAL COMPARISION ON JUSTICE.
+WE NEED TO KNOW THE WHY BUT LITERATURE SHOULD START TO OUTLINE THAT EXPLICILITY.
+-->
+
+For the purpose of this review focused on *equity*, the focus is on studies that substantively answer "Where?", "When?", "Who?", "What?", and "How?", while the answer to "Why?" may occasionally be incidental.
+
+Transportation equity literature exists within cross-cutting categories.
+
+Though critical, answers to "Why?" are not overwhelming common in the transportation equity academic literature that also address  "Where?", "When?", "Who?", "What?", and "How?".
+This is discussed as a next step in planning for *justice* transportation literature in Section 6.
+
+<!--
+MOVE TO CONCLUSIONS?
+-->
 

### 2023-10-20 Manuscript/Manuscript.qmd
Moving on to the summary of findings

@@
 
-The fiery wake of a rocket could be seen ascending, piercing the morning skies of West Texas.
+The fiery wake of a rocket could be seen ascending, piercing the morning sky above West Texas.
 It was July 20, 2021, and on board was a small group of four passengers that included Jeff Bezos, the founder of Amazon and then the world's richest person [@harwoodBezos2021].
@@
 Meanwhile, firmly grounded on planet Earth, the employees that Bezos thanked for his suborbital escapade were struggling with some very mundane problems of their own, and none as lofty as conflicting schedules that prevented them from flying in rockets [@griffinPerson2021].
-According to reports, people employed directly or indirectly by Amazon for warehouse or delivery work had, for years, been treated to "inhumane" conditions [@fungEuropean2018; @scottAmazon2019; @greeneAlgorithm2021], and subjected to surveillance on the job, crushing demands for productivity quotas that led to injury, facing little or no job security, and treated less with gratitude than as disposable inputs to feed Amazon's earnings and consumers' desires [@tungAmazon2020; @reeseTeamsters2022; @middletonRobot2023].
+According to reports, people employed directly or indirectly by Amazon for warehouse or delivery work had, for years, been treated to "inhumane" conditions [@fungEuropean2018; @scottAmazon2019; @greeneAlgorithm2021], and subjected to surveillance on the job, degrading schedules that drove drivers to urinate in bottles, crushing demands for productivity quotas that led injury, facing little or no job security, and treated less with gratitude than as disposable inputs to feed Amazon's earnings and consumers' demands [@tungAmazon2020; @BBCNews2021; @reeseTeamsters2022; @middletonRobot2023].
 
@@
 
-After this introduction, we set the stage for our investigation by laying out some important definitions. We then describe the methods used for searching and reviewing the literature. 
+After this introduction, we set the stage for our investigation by laying out some important definitions.
+We then describe the methods used for searching, selecting, and reviewing the relevant literature.
 <!--
@@
 
-In the paragraphs above we used the terms "justice", "fairness", and "equity" in a relatively loose fashion.
+In the introductory paragraphs we used the terms "justice", "fairness", and "equity" relatively loosely.
 This was done purposefully.
@@
 Some are direct and paid directly by the traveler (e.g., travel time, out-of-pocket costs), but many others are indirect and related to network externalities (e.g., exposure to pollution).
-
-<!-- ANASTASIA NEEDS TO REVIEW THE FOLLOWING -->
 
@@
 
-The focus of this report is on standards of fairness that, combined with the use of equity analysis, can help us understand how better to move towards just transportation systems.
-
 <!-- 
@@
 -->
+
+The focus of this report is on standards of fairness that, combined with the use of equity analysis, can help us understand how better to move towards just transportation systems.
 
@@
 
-This review examines the breadth and depth of the academic literature on transportation equity that analyses equity within at least one equity dimensions and implements equity conceptualizations and equity standards as defined in the previous section.
-This review follows the Joanna Briggs Institute (JBI) approach to the conduct of scoping reviews which builds upon the Arksey and O'Malley (2005) framework [@petersUpdatedMethodologicalGuidance2020].
-The JBI is a global organization that creates frameworks, protocols and guidance for the synthesis of research.
-This review is also guided by the Preferred Reporting Items for Systematic Reviews and Meta-Analyses extension for scoping reviews (PRISMA-ScR), consistent with the JBI approach [@triccoPRISMAExtensionScoping2018].
-Using these methods allow the review to explore the broad topic of transportation equity within the literature and collate knowledge from across methodologies in the academic literature.
-
-The primary research question and the protocol was drafted and refined from preliminary searches of related-reviews e.g., [@iglesiasHowUnevenUrban2019; @sagarisUsingPARFrame2020; @vecchioAccessibilityCapabilitiesApproach2021], consultations with the authorship team who are engaged in transportation equity related research, and a University of Toronto Research Services Librarian and Liaison Librarian in City Studies.
-The methods are described in two parts: (i) the development of the search strategy and (ii) the evidence selection and data extraction.
+This review examines the breadth and depth of the academic literature on transportation to identify the extent to which standards for equity are defined and employed.
+In this task, we follow the Joanna Briggs Institute (JBI) approach to the conduct of scoping reviews.
+This approach builds upon the Arksey and O'Malley (2005) framework [@petersUpdatedMethodologicalGuidance2020].
+The JBI is .
+
+Further, the review is guided too by the Preferred Reporting Items for Systematic Reviews and Meta-Analyses extension for scoping reviews (PRISMA-ScR), which is consistent with the JBI approach [@triccoPRISMAExtensionScoping2018].
+The use of these methods allows us to explore, in a consistent and organized manner, a relatively specialized topic within the broader transportation literature.
+In this way, we aim to collate the current knowledge as found in the landscape of published research.
+
+The primary research question and the protocol were initially defined in consultation among the authors of the report.
+In other words, the starting point was the level of knowledge of experts in the subject matter.
+The initial draft of the protocol was refined from preliminary searches of related-reviews [e.g., @iglesiasHowUnevenUrban2019; @sagarisUsingPARFrame2020; @vecchioAccessibilityCapabilitiesApproach2021], and in consultation with a University of Toronto Research Services Librarian and Liaison Librarian in City Studies.
+
+The methods are described in two parts: (i) development of the search strategy and (ii) selection of evidence and data extraction.
+The framework for analyzing questions of justice discussed in the preceding section was used in both stages, but was particularly valuable for selecting the evidence and for analysis for data extraction.
 
 ## Search strategy
+
+<!-- SELECTION OF DATABASES?? -->
 
@@
 
-Next, the inclusion and exclusion criteria was applied to develop the search strategy.
-The search strategy was developed iteratively, at each stage adding topic search terms (e.g., terms in the title, abstract or key words) to the set of terms tied together with logical connector terms "AND" and "OR".
-These stages are summarized in what follows and the full search term queries are detailed in Appendix @fig-A1.
+Next, the inclusion and exclusion criteria were deployed to develop the search strategy.
+The search strategy was refined iteratively, adding topic search terms by stages (e.g., terms in the title, abstract or key words).
+The terms were bundled by means of logical connector terms "AND" and "OR".
+These stages are summarized next.
+The full search term queries can be consulted in Appendix @fig-A1.
 
 1.
-An initial limited search of the core collections within the Web of Science (WoS) platform was undertaken to identify key papers.
-Terms for 'transportation' and 'equity' were generated, respectively.
-The text words contained in the titles and abstracts of relevant papers, the index terms used to describe the papers, and subject heading searches when available (depending on the database within WoS).
-This search was iterated on and took the form: ("Transport" OR "Transit" OR "Car\*" OR "Walk" OR "Bike"...**1**) AND ("Equity" OR "Justice" OR "Equity" OR "Fair"...**2**), where **1** and **2** signify additional  terms that relate to 'transportation' and 'equity', respectively.
-2.
-Upon inspection of the preliminary search results and team consensus, the 'equity' set of search terms was expanded into three sets of terms.
-The first set describes equity theories, the second describes equity dimensions, and the third describes terms referring to standards.
-All three sets of terms were iteratively augmented.
-The finalized search query took the following general form: ("Transport" OR "Transit" OR "Car\*" OR "Walk" OR "Bike"...**1**) AND ("Equity" OR "Justice" OR "Equity" OR "Fair"...**2**) AND ("Accessibility" OR "Mobility" OR ...3) AND ("Standard" OR "Threshold" OR ...4) where **1**,**2**,**3**, and **4** signify additional terms included in the sets combined with "OR" logical connectors.
-
-The search strategy, including all identified keywords and index terms, were applied to each database and platform searched (i.e., Web of Science General Collection -Science Citation Index Expanded, Social Sciences Citation Index (Web of Science), and Transportation Research International Documentation (TRID)).
-The exports of the final search query was completed by the lead researcher on March 21st 2021.
+An initial limited search of Web of Science (WoS) Core Collection was undertaken to identify key documents This collection contains documents in journals, conference proceedings, and books published all over the world.
+Separate sarches using the terms 'transportation' and 'equity' were generated.
+From these searches, we examined the text contained in the titles and abstracts of relevant papers, the index terms used to describe the papers, and subject heading searches when available.
+As we developed a clearer outline of the literature, we continued with this process by adjusting the terms used for the search.
+This took the form: ("Transport" OR "Transit" OR "Car\*" OR "Walk" OR "Bike"...**1**) AND ("Equity" OR "Justice" OR "Fair"...**2**), where **1** and **2** signify additional terms relating to 'transportation' and 'equity', respectively.
+
+1.
+Upon inspection of the preliminary search results and after achieving a consensus among the authors, the set of search terms related to 'equity' was expanded into three sets of terms.
+The first set describes theories and concepts of equity, the second describes the object of justice (i.e., the "what" in our analytical framework), and the third describes terms referring to standards (i.e., the "how").
+All three sets of terms were augmented following an iterative process of refinement.
+The final search query took the following general form: ("Transport" OR "Transit" OR "Car\*" OR "Walk" OR "Bike"...**1**) AND ("Equity" OR "Justice" OR "Equity" OR "Fair"...**2**) AND ("Accessibility" OR "Mobility" OR ...**3**) AND ("Standard" OR "Threshold" OR ...**4**) where **1**,**2**,**3**, and **4** signify additional terms included in the sets combined with "OR" logical connectors.
+
+After testing the search strategy on WoS Core Collection, we proceeded to apply to an augmented list of databases, that expanded on our search using the Core Collection of WoS.
+The databases used are: WoS General Collection-Science Citation Index Expanded, WoS Social Sciences Citation Index, and Transportation Research International Documentation (TRID).
+The final search query was completed and exported by the lead author on March 21st 2021.
 
@@
 
-Evidence selection consists of scanning the exported literature from the search strategy and retaining the papers that fit the inclusion and exclusion criteria.
+The semi-automated nature of the search strategy tends to be overly inclusive, which serves our purpose well, since we aim to begin with more documents than are strictly needed, and so reduce the risk of omitting relevant material.
+The next stage is to trim the corpus of documents, a task that can no longer be automated, and requires expert knowledge.
+Selection of evidence is where this expert knowledge really comes to bear, and it consists of scanning the literature retrieved by the search strategy to retain in the corous only those papers that fit the inclusion and exclusion criteria.
 This process was pilot-tested with a subset of papers before being implemented on the full set.
-Then, *Covidence*, an online application for literature screening, was used for all steps of selection and data extraction on the full export of literature. All steps are visualized in @fig-fig1.
+*Covidence*^[https://www.covidence.org/], an online application for literature screening, was used for all steps of selection and data extraction on the full export of literature.
+Covidence is designed for collaborative work, and helps to document the work of multiple reviewers.
+The steps of this process are shown in @fig-fig1.
 
@@
 #| fig-cap: "Evidence selection process framework. Step 1 (orange) is title and abstract screening, step 2 (green) is full-text review, and step 3 (purple) is data extraction."
-knitr::include_graphics("Methods_framework.png")
-```
-
-The process is summarized in the following three steps:
+knitr::include_graphics("figures/Methods_framework.png")
+```
+
+The following steps summarize the process:
 
@@
 The first step (orange box in @fig-fig1) included screening all titles and abstracts of papers on whether they included transportation equity as defined by the PCC.
-All papers were voted on by two independent reviewers for inclusion, exclusion, or uncertain inclusion.
-All uncertain papers, conflicting papers, and papers missing abstracts were voted on by a third reviewer for inclusion or exclusion.
-2.
-The second step (green box in @fig-fig1) included reviewing all full-text papers which passed step 1.
-These papers were reviewed based on if their study design included an equity standard and equity conceptualization.
-All papers are voted on by two independent reviewers for inclusion or exclusion.
-If an article was voted to be excluded, it was tagged with one of five possible reasons, namely (1) no standards included, (2) no conceptualizations included, (3) no standard and no conceptualization included, (4) send back -- QA issue, (5) other.
+Each paper was screened by two independent reviewers who then voted for inclusion, exclusion, or uncertain inclusion.
+All uncertain papers, conflicting papers, and papers missing abstracts were reviewed by a third person for inclusion or exclusion.
+
+1.
+The second step (green box in @fig-fig1) included scanning all full-text papers which passed step 1.
+These papers were reviewed to determine if they included a relevant "how", i.e., an standard and/or relevant theoretical or conceptual discussion.
+At this stage, papers were evaluated again by two independent reviewers who voted for inclusion or exclusion.
+If an article was voted to be excluded, it was tagged with one of five possible reasons for exclusion, namely (1) no standards included; (2) no relevant conceptual elements included; (3) no standard and no conceptual elements included; (4) send back -- QA issue; or (5) other.
 Discrepancies were resolved by a third reviewer.
-3.
-A data extraction template was filled out for each record: one reviewer for the third and final step (purple box in @fig-fig1).
-The data extraction template was created to balance complexity of categories and simplicity of summary; information related to "What?" (equity dimension), "Who?" (population group, mode type, type of destination/activity), "Where?" (region of study), and "How?" (the equity standard(s) and associated conceptualization(s)) was filled out for each study.
-@fig-A2 contains the template that was input into *Covidence* and used within this review.
+
+1.
+In the last step, a data extraction template for each record was filled by one reviewer (purple box in @fig-fig1).
+The data extraction template was created with the aim of striking a balance between the complexity of categories and the simplicity of summary; information related to "What?" (the object of justice), "Who?" (the subject of justice), "Where?" (the geographical context and sphere of life), and "How?" (equity standard(s) and concepts) was filled out for each study.
+@fig-A2 contains the template that was input into *Covidence* and used throughout.
 
@@
 #| fig-cap: "PRISMA flow diagram for the evidence selection process. ES signifies equity standard and EC signifies equity conceptualization."
-knitr::include_graphics("Pflow_chart.png")
+knitr::include_graphics("figures/Pflow_chart.png")
 ```
@@
 1710 studies entered step 2, 1223 were excluded and the remaining 487 papers entered step 3.
-The data extraction template used by the reviewers (authorship team) in step 3 revealed that inclusion was initially too generous, and some papers did not have a sufficiently strong equity standard and/or equity conceptualization.
-As such, 322 papers were further excluded and data extraction was completed for the remaining 165 papers.
+The data extraction template used by the reviewers (authorship team) in step 3 revealed that, as expected, inclusion was initially too generous, and some papers were not sufficiently relevant, because of a lack of content on standards and/or conceptual/theoretical elements.
+In this fashion, 322 papers were further excluded and data extraction was completed to give a final corpus of 165 papers.
 A summary of the reasons for exclusion of the 1545 papers (between steps 2 and 3) are included in @fig-fig2.
@@
 
-A synthesis of the findings from the data extraction template (shown in @fig-A2) is detailed in this section.
-The presentation of findings is less granular than the template to highlight key trends in the literature.
-
-## Temporal and spatial focus (the "When" and "Where")
+A synthesis of the findings that result from the data extraction process (based on the template shown in @fig-A2) is detailed in this section.
+The presentation of findings is less granular than the template to highlight what are, in our view, the key trends in the literature.
+
+## "When" and "Where" justice
 
@@
 @fig-fig3 displays the papers included in this review by year of publication and case study continent.
-Evidently, transportation equity related papers are increased in popularity throughout the years, particularly since 2019.
+Evidently, the literature related to transportation equity has grown increasingly voluminous over time, particularly since 2019.
 Of note is the geographic scope of the case studies present in the papers.
 The majority of papers (`r ((case_cont_count$count[4] + case_cont_count$count[5]+ case_cont_count$count[6])/sum(case_cont_count$count)) |> percent()`) contain case studies based in the Global North e.g., North America (particularly USA and Canada), Europe (particularly UK, France, Spain and Scandinavia), and Oceania (Australia and New Zealand).
-How *equity* is operationalized is context-specific, and with the majority of transportation systems *not* in the Global North, the literature included in this review falls short in representing the global perspective.
+For the reasons discussed previously, operationalizing *equity* is highly context-specific.
+Publication patterns, being as they are, display a disproportionately low number of items from the Global South where equity issues are perhaps as, or even more pressing, than in the Global North.
+Be it as it may, we do not lay claim to reporting from a truly global perspective.
 
@@
 
-The few studies in the Global South are predominately in the **Asian** continent, specifically China but also India, Thailand, South Korean, Iran, Philippines, Indonesia, Israel, and Japan.
-These studies focus on a variety of modes, population groups, and equity conceptualizations and standards that cannot be succinctly summarized. <!-- develop further, if there's room?-->
+The few studies in the Global South are predominately in the **Asian** continent, specifically China but also India, Thailand, Iran, Philippines, Indonesia.
+These countries add to our knowledge stemming from research reported from South Korean, Israel, and Japan.
+These studies focus on a variety of modes, population groups, and equity conceptualizations and standards that cannot be succinctly summarized.
+<!-- develop further, if there's room?-->
 
@@
 
-# Discussion and ways forward
+# Discussion and recommendations
 
@@
 
-#### The search strategy:
+## Search strategy
+
 ```{r Search-query-A1}
@@
 Hits corresponds the number of papers that the search yielded and was retained into the evidence selection process."
-knitr::include_graphics("Search-query.png")
+knitr::include_graphics("figures/Search-query.png")
 ```
@@
 
-#### Example of the data extraction template:
+## Example of the data extraction template:
 
@@
 #| fig-cap: "The data extraction template with associated defintions."
-knitr::include_graphics("Data-extract-template.png")
-```
-
-### Sample of select papers summarized by their, equity dimension, case study continent, conceptualizations and standards
+knitr::include_graphics("figures/Data-extract-template.png")
+```
+
+## Examples of papers summarized by element of the analyitical framework
 
 ```{r}
-dat <- data.frame(matrix(c("Equity dimension 1: **mobility and accessibilty**",
-                           "[@rivasHowAffordableTransportation2018] - South America (select cities)",
+dat <- data.frame(matrix(c("What?: **mobility and accessibilty**",
+                           "Where? [@rivasHowAffordableTransportation2018] - South America (select cities)",
                            "Analyses how affordable urban public transportation is in select Latin American and Caribbean countries.
 They look at the estimated average monthly cost of transit trips and average monthly household income and conceptualize **transportation-related** **affordability**, especially for the most economically vulnerable (**vertical equity**).",
-                           "**Population standards**: The financial burden of a basket of urban public transportation trips (60 trip fares, representing 30 round-trips per month) should not exceed 10% of household monthly income.",
-                           "Equity dimension 1: **mobility and accessibilty**",
-                           "[@bharathyRevisitingClearFloor2018] - North America (USA - National)",
+                           "How?: The financial burden of a basket of urban public transportation trips (60 trip fares, representing 30 round-trips per month) should not exceed 10% of household monthly income.",
+                           "What?: **mobility and accessibilty**",
+                           "Where?: [@bharathyRevisitingClearFloor2018] - North America (USA - National)",
                            "This study designed a web-based tool and took a representative sample of wheeled mobility device (WhMD) users anthropometry measurements to determine if the minimum standard suggested by the ADA is sufficient.
 We understand this conceptualization as a type of **Rights** conceptualization that WhMD should have minimum clear floor space (as described the guidelines in line with the American Disabilities Act) to access bus shelters, bus stop pads, and transit terminals.",
-                           "**Infrastructure standard**: The clear floor area for wheelchairs: 760 mm (30 in.) wide by 1220 mm (48 in.) in length as described by the ADA standards.
+                           "How?: The clear floor area for wheelchairs: 760 mm (30 in.) wide by 1220 mm (48 in.) in length as described by the ADA standards.
 Of note, this minimum clear floor area is insufficient for a variety of the WhMD users.",
-                           "Equity dimension 1: **mobility and accessibilty**",
+                           "What?: **mobility and accessibilty**",
                            "[@ryanWhatAreWe2021] - Europe (Stockholm, Gothenburg and Malmo cities in Sweden)",
@@
 This paper conceptualizes accessibility as from the position of the **capabilities approach** and **vertical equity** (particularly acknowledging that older people have capabilities that differ from the general population).",
-                           "**Population** and **Opportunity standards**.
-Specifically for older populations (aged 65+), the following travel distances are suggested as equitable trip lengths to grocery stores per mode: Walking: less than or equal to 1500m, Combined public transit and walking (less than or equal to 1000m (walking element)), Combined car and walking: less than or equal to 1000m less than or equal to 1000m (walking element)), Bicycle: less than or equal to 3000m in addition to travel time threshold of less than 15 mins.",
-                           "Equity dimension 1: **mobility and accessibilty**",
-                           "[@wismadiSpatialPreferenceModelling2014] - Asia (Yogyakarta, Indonesia)",
+                           "How?: Specifically for older populations (aged 65+), the following travel distances are suggested as equitable trip lengths to grocery stores per mode: Walking: less than or equal to 1500m, Combined public transit and walking (less than or equal to 1000m (walking element)), Combined car and walking: less than or equal to 1000m less than or equal to 1000m (walking element)), Bicycle: less than or equal to 3000m in addition to travel time threshold of less than 15 mins.",
+                           "What?: **mobility and accessibilty**",
+                           "Where?: [@wismadiSpatialPreferenceModelling2014] - Asia (Yogyakarta, Indonesia)",
                            "Explores the equitable provision of transport infrastructure provision: an application of Sen's **capability approach**.
 Conceptualizes equity through Sen's capability approach and spatial equity.",
-                           "Opportunity standards: Areas below the relative poverty line (of its neighbours) can only be located transport resources (i.e., measure in person*kms that can be travelled at car speed, i.e., mobility) based on the following 2 benchmarks (they can be considered, together as the floor/minmum access): 1) Global: standard deviation (SD) distance to mean should be minimized.
+                           "How?: Areas below the relative poverty line (of its neighbours) can only be located transport resources (i.e., measure in person*kms that can be travelled at car speed, i.e., mobility) based on the following 2 benchmarks (they can be considered, together as the floor/minmum access): 1) Global: standard deviation (SD) distance to mean should be minimized.
 2) Local: priority to minimise the differences with its neighbourhood",
-                           "Equity dimension 1: **mobility and accessibilty**",
-                           "[@zhengAreabasedEquitablePricing2020] - North America",
+                           "What?: **mobility and accessibilty**",
+                           "Where?: [@zhengAreabasedEquitablePricing2020] - North America",
                            "This paper conceptualizies equity in the multimodal network (transit, car) being fair toll-pricing across differences in populatins value of time (VOT).
@@
 From this perspective, a utilitarian perspective that seeks to minimize multimodal traffic congestion through introducing toll-pricing based on VOT is implemented.",
-                           "Infrastructure standard: suggest that a toll-pricing scheme based on individuals travel value-of-time (lower income people have a lower VOT) is equitable.",
-
-                           "Equity dimension 2: **environmental pollution**",
-                           "[@carrierApplicationThreeMethods2014] - North America (Montreal, Canada) ",
+                           "How?: suggest that a toll-pricing scheme based on individuals travel value-of-time (lower income people have a lower VOT) is equitable.",
+
+                           "What?: **environmental pollution**",
+                           "Where?: [@carrierApplicationThreeMethods2014] - North America (Montreal, Canada) ",
                            "This work examines the statistical association between different social groups and the concentration of air pollutants.
@@
 The paper then frames this impact as unfair, particularly from the perspective of disproportionately disadvantaged residents",
-                           "**Infrastructure standards**: The literature suggests that the health implications from the transport-related air pollution from major roadways is most acute at residential distance locations of 200 m or less.
+                           "How?: The literature suggests that the health implications from the transport-related air pollution from major roadways is most acute at residential distance locations of 200 m or less.
 Residential locations should not be located within this distance threshold from the perspective of human health.
@@
 It can be interpreted that they use the WHO threshold as a minimum threshold and suggest that air pollution levels should not be impacting disadvantaged populations disproportionately ( a relative population standard)",
-                           "Equity dimension 2: **environmental pollution**",
-                           "[@jephcoteGeospatialAnalysisNaturally2013] - Europe (Leicester, UK)",
+                           "What?: **environmental pollution**",
+                           "Where?: [@jephcoteGeospatialAnalysisNaturally2013] - Europe (Leicester, UK)",
                            "Geospatial analysis of naturally occurring boundaries in road-transport emissions and childrens respiratory health across a demographically diverse cityscape.
@@
 Children are at most risk for acute respiratory distress from elevated levels of air pollution, and as such planning should consider this point of public health.",
-                           "**Infrastructure standard**: Finds that children (most vulnerable to air pollution - related to motoized traffic) are most impacted by air pollution within 283 m of a road way.
+                           "How?: Finds that children (most vulnerable to air pollution - related to motoized traffic) are most impacted by air pollution within 283 m of a road way.
 This should be the distance threshold that schools and other childrens facilities are located.",
-                           "Equity dimension 3: **health**",
-                           "[@adlakhaMindGapGender2020] - Asia (Chennai, India)",
+                           "What?: **health impacts**",
+                           "Where? [@adlakhaMindGapGender2020] - Asia (Chennai, India)",
                            "From the perspective of disparity in gendered physical activity, this paper focuses on women's cycling as both transport and exercise.
 They advocate for all people achieving physical activity thresholds (**horizontal equity**) but prioritize women and especially women in neighborhoods with low-walkability and socio-economic status (**vertical equity**).",
-                           "**Population standards**: All people should get 150 min of moderate activity a week or 75 min of vigorous physical activity per week.",
-                           "Equity dimension 3: **health**",
-                           "[@savingmothersDidSavingMothers2019] - Africa (Select urban and rural regions in Uganda)",
+                           "How?: All people should get 150 min of moderate activity a week or 75 min of vigorous physical activity per week.",
+                           "What?: **health impacts**",
+                           "Where? [@savingmothersDidSavingMothers2019] - Africa (Select urban and rural regions in Uganda)",
                            "The **well-being** of mothers, this paper examines the timely access to emergency obsteric and newborn care for child-bearing aged women in Uganda.",
-                           "**Opportunity standard**: 2 hours to the nearest facility with surgical capacity with anesthesia services - this threshold is determined through the onset of bleeding to death if a women with obstetric hemorrhage does not receive adequate treatment).",
-                           "Equity dimension 3: **health**",
-                           "[@iungmanImpactUrbanTransport2021] - Europe (Madrid and Barcelona, Spain)",
+                           "How?: 2 hours to the nearest facility with surgical capacity with anesthesia services - this threshold is determined through the onset of bleeding to death if a women with obstetric hemorrhage does not receive adequate treatment).",
+                           "What?: **health impacts**",
+                           "Where?: [@iungmanImpactUrbanTransport2021] - Europe (Madrid and Barcelona, Spain)",
                            "They use environmental pollution guidelines, but from the position of health.
@@
 These externalities should not be impacting people disproportionately (**vertical equity**) and should be even across space (**spatial equity**).",
-                           "**Environmental+ standards**: All minimum thresholds, if exceeded this is inequitable: NO² concentration 40 μg/m³; PM 2.5 concentration 10 μg/m³; Noise 53dB for average 24 hours; Living with 300 m crow-flies distance from at least .5 hectares of greenspace; and a Change of air temperature of at least 1 ⁰C.
+                           "How?: All minimum thresholds, if exceeded this is inequitable: NO² concentration 40 μg/m³; PM 2.5 concentration 10 μg/m³; Noise 53dB for average 24 hours; Living with 300 m crow-flies distance from at least .5 hectares of greenspace; and a Change of air temperature of at least 1 ⁰C.
 ",
-                           "Equity dimension 3: **health**",
-                           "[@mehdizadehWalkingTimeSchool2017] - Asia (Rasht, Iran)",
+                           "What?: **health impacts**",
+                           "Where?: [@mehdizadehWalkingTimeSchool2017] - Asia (Rasht, Iran)",
                            "From the perspective of children’s **well-being**, assesses the walking time to school. They frame walking to school as health-related.",
-                           "**Opportunity** and **Population standards**: perceived walking time to school for students aged 7-9 yrs is 10 mins, and the longer the PWTS the less likely they were to use an active mode to travel to school.",
-                           "Equity dimension 3: **health**",
-                           "[@murphySupermarketAccessTransport2017] - Oceania (Melbourne, Australia)",
+                           "How?: perceived walking time to school for students aged 7-9 yrs is 10 mins, and the longer the PWTS the less likely they were to use an active mode to travel to school.",
+                           "What?: **health impacts**",
+                           "Where? [@murphySupermarketAccessTransport2017] - Oceania (Melbourne, Australia)",
                            "Assesses the relationship between supermarket access and transport mode used, the body mass index (BMI) of the mode-user (**wellbeing**) and the equity in access distribution by income (**vertical equity**).",
-                           "**Opportunity** and **Population standards**: all households should be sufficiently active (greater than 150 min and at least 5 sessions) and households should be within 1 km euclidean distance to supermarket (80-90% of the dwellings should meet this).
+                           "How?: all households should be sufficiently active (greater than 150 min and at least 5 sessions) and households should be within 1 km euclidean distance to supermarket (80-90% of the dwellings should meet this).
 Planners should prioritize socially disadvantaged areas to meeting these standards first.",
 
-                            "Equity dimension 4: **transport-related safety**",
-                           "[@ferenchakEquityAnalysisProactively2019] - North America (Denver, USA)",
+                           "What?: **transport-related safety**",
+                           "Where?: [@ferenchakEquityAnalysisProactively2019] - North America (Denver, USA)",
                            "Operationalizes and compares an equity analysis of proactively- and reactively-identified traffic safety issues from the perspective of **Spatial equity**, **Vertical equity** and **Inequitable exposure to externalities**.",
-                           "**Infrastructure** and **Population standards**: standards are suggested for both reactive and proactive analysis.
+                           "How?: standards are suggested for both reactive and proactive analysis.
 First, the lower the number of collisions on the road with pedestrians/cyclists (i.e., reactive safety analysis), the better.
@@
 equity seeking groups (high proportion of POC and/or low income in tract).",
-                           "Equity dimension 4: **transport-related safety**",
-                           "[@zheEvaluationSharedUse2008] - Asia (Tokyo, Takamatsu, and Tokushima",
+                           "What?: **transport-related safety**",
+                           "Where?: [@zheEvaluationSharedUse2008] - Asia (Tokyo, Takamatsu, and Tokushima)",
                            "Evaluates the observed safety of  shared use pedestrian and bicycle paths from the perspective of **well-being**.",
-                           "**Infrastructure standard**: the study suggests that the safety threshold for bicycles and pedestrians to coexist on shared infrastructure is less than 0.5 pedestrians/minute per metre of sidewalk (width) and less than 3.0 cyclists/minute per metre of sidewalk (width).
+                           "How?: the study suggests that the safety threshold for bicycles and pedestrians to coexist on shared infrastructure is less than 0.5 pedestrians/minute per metre of sidewalk (width) and less than 3.0 cyclists/minute per metre of sidewalk (width).
 The standard for pedestrian/bicycle share use in terms of hourly traffic volume is less than 26 pedestrians / hour and 108 cyclists / hour for 2m wide sidewalks.",
 
-                           "**Multi-equity dimensions**",
-                           "[@aldertonWhatMeaningUrban2019] - Asia (Bangkok, Thailand) – **Mobility/ accessibility** and **health**",
+                           "What?: **mobility/accessibility and health impacts**",
+                           "Where?: [@aldertonWhatMeaningUrban2019] - Asia (Bangkok, Thailand) – **Mobility/ accessibility** and **health**",
                            "Establishes short-, medium-, and long-term goals for the city in collaboration with technical leaders within the municipal government for the perspective of **well-being** (urban livability): the standards included in this table relate directly to transportation systems.
 Indicators are inspired by the Sustainable Development Goals (SDGs) as well other global planning standards.",
-                           "**Opportunity standards** are suggested: 1) Green space: % of residents living < 400 m from public open space, a large park (> 1.5ha), and/or local park, 2) Public transit access: % of residents living < 400 m of a local bus stop and <800 m of train station, 3) Facilities: % of residents living < 400 m of a community centre.
+                           "How?: 1) Green space: % of residents living < 400 m from public open space, a large park (> 1.5ha), and/or local park, 2) Public transit access: % of residents living < 400 m of a local bus stop and <800 m of train station, 3) Facilities: % of residents living < 400 m of a community centre.
 The following **Infrastructure standard** is suggested: Canal water quality - dissolved oxygen content of equal to or less than 2.0 mL/L",
-                           "**Multi-equity dimensions**",
-                           "[@berheAdaptationDissonanceQuality2014] - Africa (Mekelle, Ethiopia) -**Mobility/ accessibility** and **health**",
+                           "What?: **Mobility/ accessibility** and **health impacts**",
+                           "Where?: [@berheAdaptationDissonanceQuality2014] - Africa (Mekelle, Ethiopia)",
                          "Examines adaption and dissonance in the quality of life (QoL) of residents.
@@
 We assume the equity goal for this paper is that subjective and objective QoL measures should not be mismatched: as discussed by the authors of this study, subjective QoL is higher than objective QoL the participant is experiencing adaption and in the reverse scenario the participant is experience dissonance.",
-                           "**Opportunity standards**: 1 & 2) Access to primary or secondary education facility, percentage of households living within 1 km or 2km (walking distance), respectively from a primary school or secondary school.
+                           "How: 1 & 2) Access to primary or secondary education facility, percentage of households living within 1 km or 2km (walking distance), respectively from a primary school or secondary school.
 3) Access to health facility, percentage of households within 40 min walking time from a health facility.
@@
 2) Subjective QoL is constructed based on the households level of satisfaction for each of the eight indicators using a six point Likert-scale (1=very satisfied to 6=very dissatisfied).",
-                           "**Multi-equity dimensions**",
-                           "[@agostfelipInclusiveModelAssessing2021] - Europe (Castellon, Spain) - **Mobility/ accessibility**, **safety**, and **health**",
+                           "What?: **mobility/accessibility, health impacts, and safety**",
+                           "Where? [@agostfelipInclusiveModelAssessing2021] - Europe (Castellon, Spain)",
                            "Conceptualizes equity through age-friendly urban spaces that reduce (and eliminate) conditions for **transportation-related social exclusion** for older populations and prioritize those who are economically vulnerable (**vertical equity**).
 These guidelines are inspired by the SDGs in addition to planning guidelines used national, regional, and local guidelines used in Spain.",
-                           "**Opportunity standards**: 1) Access to facilities needed for old age health.
+                           "How: 1) Access to facilities needed for old age health.
 Minimum distance thresholds from the geometric center of neighborhood are suggested: at least: 1000 m from health facilities (600 m or less is preferred), elderly-specific care facilities and shops should be 600 m (300 m or less is preferred).
@@
 **Environment + standards **: 1) Noise at the street level should be less than 55 dB and 45 dB (but preferably less than 50 dB and 40 dB) in the daytime and nighttime, respectively.",
-                           "**Multi-equity dimensions**",
-                           "[@mateobabianoPedestrianNeedsMatter2016] - Asian (Manila, Philippine) - **Mobility/ accessibility** and **safety**",
+                           "What?: **mobility/accessibility and safety**",
+                           "Where?: [@mateobabianoPedestrianNeedsMatter2016] - Asian (Manila, Philippine)",
                            "The perception of pedestrians’ walking environments should be sufficient across 6 themes.
 Equity is conceptualized around **spatial equity** (equally fair walking environments for all locations) and **rights** (the right to mobility/accessibility for pedestrians)",
-                           "**Infrastructure**, **Opportunity**, and **Population standards**: percieved pedestrian perception on protection, ease, equitable access, mobility, identity, and enjoyment must be met."
+                           "How?: perceived pedestrian perception on protection, ease, equitable access, mobility, identity, and enjoyment must be met."
                            ), ncol = 4, byrow = T))
@@
 
-<!-- we need to sepearte by qual and quant, karen asked-->
+<!-- we need to separate by qual and quant, karen asked-->
 

### 2023-10-21 Manuscript/Manuscript.qmd
Edits to "the lay of the land"

@@
 
-> "_I should like to be able to love my country and still love justice_."
-> — Albert Camus
+> "_Man's capacity for justice makes democracy possible, but man's inclination to injustice makes democracy necessary._"
+> ― Reinhold Niebuhr
 
@@
 
-# Methods
+# Scanning the lay of the land: Methods
+
+> "_Not everything that is faced can be changed. But nothing can be changed until it is faced._"
+> — James A. Baldwin
 
@@
 The databases used are: WoS General Collection-Science Citation Index Expanded, WoS Social Sciences Citation Index, and Transportation Research International Documentation (TRID).
-The final search query was completed and exported by the lead author on March 21st 2021.
+The definitive version of the search was completed and exported by the lead author on March 21st, 2021.
 
@@
 
-# Summary of `r word_comment("NOTE: in this section, only papers that are included in the review (i.e., have a full template) are cited.", "findings")`
-
-A synthesis of the findings that result from the data extraction process (based on the template shown in @fig-A2) is detailed in this section.
-The presentation of findings is less granular than the template to highlight what are, in our view, the key trends in the literature.
-
-## "When" and "Where" justice
+# The lay of the land: a summary of findings  `r word_comment("NOTE: in this section, only papers that are included in the review (i.e., have a full template) are cited.", "findings")`
+
+> "_Give me knowledge, so that I may have kindness for all._"
+> — Native American Proverb
+
+A synthesis of the findings from the data extraction process (based on the template shown in @fig-A2) is detailed in this section.
+The presentation of findings is less granular than the template to highlight the key trends in the literature.
+
+## "When" and "Where": the context for justice
 
@@
 
-@fig-fig3 displays the papers included in this review by year of publication and case study continent.
-Evidently, the literature related to transportation equity has grown increasingly voluminous over time, particularly since 2019.
-Of note is the geographic scope of the case studies present in the papers.
-The majority of papers (`r ((case_cont_count$count[4] + case_cont_count$count[5]+ case_cont_count$count[6])/sum(case_cont_count$count)) |> percent()`) contain case studies based in the Global North e.g., North America (particularly USA and Canada), Europe (particularly UK, France, Spain and Scandinavia), and Oceania (Australia and New Zealand).
-For the reasons discussed previously, operationalizing *equity* is highly context-specific.
-Publication patterns, being as they are, display a disproportionately low number of items from the Global South where equity issues are perhaps as, or even more pressing, than in the Global North.
-Be it as it may, we do not lay claim to reporting from a truly global perspective.
+@fig-fig3 displays the papers included in this review by year of publication and geographical provenance.
+The literature related to transportation equity has grown evidently voluminous, particularly since 2019.
+Of note is the geographic scope of the relevant literature.
+The majority of papers (`r ((case_cont_count$count[4] + case_cont_count$count[5]+ case_cont_count$count[6])/sum(case_cont_count$count)) |> percent()`) contain case studies based in the Global North, with many studies from North America (particularly USA and Canada), Europe (particularly UK, France, Spain and Scandinavia), Oceania (Australia and New Zealand), and Asia (Japan and Israel).
+For the reasons previously discussed, setting standards (that is, making statements of fairness) is a highly context-specific endeavor.
+Publication patterns, being what they are, display a disproportionately low number of items from the Global South where equity issues are perhaps as, or even more pressing, than in the Global North [e.g., @BBCNews2019].
+While our search strategy and review tries to be as faithful to the literature as possible, it does not allow us to lay claim to reporting about on-the-ground transportation justice efforts from a truly global perspective.
 
@@
 
-The few studies in the Global South are predominately in the **Asian** continent, specifically China but also India, Thailand, Iran, Philippines, Indonesia.
-These countries add to our knowledge stemming from research reported from South Korean, Israel, and Japan.
+The smattering of studies from the Global South are predominately from **Asia**, notably China, but also India, Thailand, Iran, Philippines, and Indonesia.
+<!-- These countries add to our knowledge stemming from research reported from South Korean, Israel, and Japan.
 These studies focus on a variety of modes, population groups, and equity conceptualizations and standards that cannot be succinctly summarized.
-<!-- develop further, if there's room?-->
-
-The next most common focus within the literature from the Global South centers on **South America**.
-Many of these studies mention a systematic absence of evidence relevant to the region [@vecchioTransportEquityLatin2020].
-Despite the growing recognition in the literature of the interconnections between transport development, social exclusion, and poverty [@benevenutoExaminingTransportNeeds2020], studies underscore an ongoing neglect of the social dimension of transport during the planning stage [@benevenutoExaminingTransportNeeds2020; @boisjolyAccessibilityMeasurementsSao2020].
-Many studies also indicate affordability as one of the main mobility barriers in the region [@falavignaAssessingInequalitiesPublic2016; @rivasHowAffordableTransportation2018], while some highlight multi-dimensional concerns such as public transport accessibility and quality of walking environments that contribute to mobility inequalities [@tiznadoaitkenRoleAccessibilityPublic2018].
-
-Within the reviewed literature, studies pertaining to **Africa** are less numerous compared to the South American literature.
-A shared characteristic among the studies from the two continents is a scarcity of official transport data [@friedMeasuringSustainableDevelopment2020] and reliance on external policy guidelines.
-These studies also incorporate the utilization of informal transportation options and pressures to development physical road network infrastructure (supporting car dependency) over meeting mobility/accessibility needs of citizens [@thondooSmallCitiesBig2020].
-To address these challenges, researchers compile databases based on open and geo-referenced data, calculate objective and/or subjective measures [@berheAdaptationDissonanceQuality2014], and focus on advancing transport justice for low to medium income countries (LMIC) by aligning with external policy guidelines such as the Sustainability Development Goals (SDGs), particularly those related to universal accessibility [@friedMeasuringSustainableDevelopment2020].
-
-From all the studies included in the review, `r (urban_form_count$count[3]/sum(urban_form_count$count)) |> percent()` focus on the urban and suburban context and are highly varied in their research aims.
-To give an example, from the perspective of cycling as a mode, @coxQualitativeStudyAccessibility2020 qualitatively examines cycling as a mode for people with disability in a typical mid-size town in the UK and @ampeImpactChildBike2020 identifies the lateral clearance that motorists should maintain when passing cyclists with children seats; both focus on understanding barriers to cycling from different perspectives, but within different urban environments.
+<!-- develop further, if there's room?-->The next most common focus within the literature from the Global South is from **South America**.
+Many of these studies mention a systemic absence of evidence relevant to the region [@vecchioTransportEquityLatin2020].
+Despite the growing recognition in the literature of the interconnections between transport development, social exclusion, and poverty [@benevenutoExaminingTransportNeeds2020], a number of studies underscore an ongoing neglect of the social dimension of transport during the planning stage [@benevenutoExaminingTransportNeeds2020; @boisjolyAccessibilityMeasurementsSao2020].
+Many studies also point at affordability as one of the main mobility barriers in the region [@falavignaAssessingInequalitiesPublic2016; @rivasHowAffordableTransportation2018], while some highlight multi-dimensional concerns such as public transport accessibility and quality of walking environments that contribute to mobility inequalities [@tiznadoaitkenRoleAccessibilityPublic2018].
+
+Studies pertaining to **Africa** are even less numerous compared to the South American literature.
+A shared characteristic among the studies from these two continents is a scarcity of official transport data [@friedMeasuringSustainableDevelopment2020] and a reliance on policy guidelines developed by international organizations.
+These studies also incorporate the use of informal transportation options and the pressures to develop road network infrastructure (which tends to support car dependency) over meeting the mobility/accessibility needs of citizens [@thondooSmallCitiesBig2020].
+To address these challenges, researchers compile databases based on open and geo-referenced data, calculate objective and/or subjective measures [@berheAdaptationDissonanceQuality2014], and focus on advancing transport justice for low to medium income countries (LMIC) by aligning their goals with external policy guidelines such as the Sustainability Development Goals (SDGs), particularly those related to universal accessibility [@friedMeasuringSustainableDevelopment2020].
+
+Of all the studies reviewed, `r (urban_form_count$count[3]/sum(urban_form_count$count)) |> percent()` focus on urban and suburban settings and are highly varied in their research aims.
+To give an example, @coxQualitativeStudyAccessibility2020 qualitatively examine cycling as a mode of travel for people with disabilities in a typical mid-size town in the UK.
+@ampeImpactChildBike2020, on the other hand, work to identify the lateral clearance that motorists should maintain when passing cyclists with children seats.
+While both studies refer to the same objects of justice ("what", i.e., the right to the road and cycling), the subjects of justice (the "who") are different (people with disabilities and children), and the "where", that is, the geographical context for the examination of fairness, differs as well.
 
 The remainder of the studies focus on rural regions (`r (urban_form_count$count[2]/sum(urban_form_count$count)) |> percent()`).
-To illustrate examples, @caoIndicatorsSocioSpatialTransport2017 examined transportation disadvantage in remote places which rely on inter-island ferry trips in the rural Philippines.
-Similarly, @parrySocialVulnerabilityClimatic2018 studied remote communities in the Amazonian and suggests that "increasing accessibility through road building would be maladaptive, exposing marginalized people to further harm and exacerbating climatic change by driving deforestation" (pp.
+To illustrate, we highlight the work of @caoIndicatorsSocioSpatialTransport2017, who examined transportation disadvantage in remote places which rely on inter-island ferry trips in the rural Philippines.
+Similarly, @parrySocialVulnerabilityClimatic2018 studied remote communities in the Amazonian region, and suggest that "increasing accessibility through road building would be maladaptive, exposing marginalized people to further harm and exacerbating climatic change by driving deforestation" (pp.
 125).
 
-## Populations of interest (the "Who" pt.1)
-
-@fig-fig4 showcases the categories of population group types that are the focus of the reviewed papers.
+## "Who": the subjects of justice
+
+@fig-fig4 showcases the population group types that are the focus of the papers in our review.
 From this tally, papers that consider income groups is the most widely represented in the literature.
-Particularly, most papers that focus on income pay particular consideration to the lowest-income groups, as they experience the greatest burden of transport inequalities, be it accessibility, affordability, or other dimensions [@peungnumsaiGridBasedSpatialAnalysis2020; @zhaoUnequableSpatialAccessibility2020; @falavignaAssessingInequalitiesPublic2016].
+Particularly, most papers that focus on income pay usually focus on the lowest-income groups, as they experience the least benefits in terms of mobility and accessibilty, and usually the greatest burdens as well, in terms of cost, exposure, and so on [@peungnumsaiGridBasedSpatialAnalysis2020; @zhaoUnequableSpatialAccessibility2020; @falavignaAssessingInequalitiesPublic2016].
 
@@
 
-An abundant amount of literature suggests low-levels of household income is a significant determinant of transport-related inequities (e.g., access to public transport supply in Bangkok region (Thailand) [@peungnumsaiGridBasedSpatialAnalysis2020], access to employment opportunities in various cities in Brazil [@boisjolyAccessibilityMeasurementsSao2020], and less environmental noise, air, and higher green space per resident ratios in Rijnmond region (Netherlands) [@kruizeEnvironmentalEquityRole2007].
-But this should be kept in context, as low-income does not necessarily mean lower transport-related benefits.
+For example, there is an abundance of evidence to suggest that low-levels of household income is a significant determinant of transport-related inequities (e.g., access to public transport supply in Bangkok region (Thailand) [@peungnumsaiGridBasedSpatialAnalysis2020], access to employment opportunities in various cities in Brazil [@boisjolyAccessibilityMeasurementsSao2020], and unfavorable rates of environmental noise, air pollution, and green space per resident in Rijnmond region (Netherlands) [@kruizeEnvironmentalEquityRole2007].
+However, this should be kept in context, as low-income is not universally associated with lower transport-related benefits for every object of justice.
 For instance, in Sheffield (UK), @mearsUnderstandingSocioeconomicEquity2019 demonstrates that historically working-class neighbourhoods (i.e., lower income working population) have more access to green space than other neighbourhoods due to urban planning approaches during the Victorian-era.
@@
 Age is the second most common category of population group of focus within the reviewed literature.
-Many papers focusing on this category highlight differing age-related capabilities; for instance, @martinez-jimenezAccessibilityCultureEducation2019 and @arranz-lopezSocialSpatialEquity2019 investigate travel distances and times to various opportunities based on specific age groups, acknowledging that age is an important consideration to opportunity access variability between populations.
-The most commonly focused on age groups are school-aged children and older populations.
-School-aged children oriented papers analyze wellbeing [@laszkiewiczChildrenGreenWalk2020], exposure to green space [@corazzaMethodologyEvidenceCase2020], access to schools [@sharmaSpatialSocialInequities2022], and understanding and encouraging active travel journeys [@mackieOvercomingBarriersCycling2009; @mehdizadehWalkingTimeSchool2017].
-Papers focusing on older-populations typically have similar aims as the children-focused articles e.g., understanding transport-related impacts on wellbeing e.g, [@chenSpatialAnalysisFramework2020], measuring accessibility to population-specific key destinations e.g.
-[@chengInvestigatingWalkingAccessibility2019], and seeking to understand how to better meet travel needs e.g., [@nordbakkeTransportUnmetActivity2015].
-
-The third most commonly focused on population category are what we summarize as 'composite vulnerable population groups'.
-These papers use some sort of composite vulnerability index that captures multiple population vulnerabilities (e.g., low-income, unemployment, immigrant status, family household characteristics, etc.) typically generated from official government sources or author-informed census data creation.
-These indices are varied in methodology: reflecting the intersectionality of population vulnerabilities.
-For instance, @awuorInfluenceEnvironmentalHealth2019 uses the Neighbourhood Equity Index (NEI) to disparities in premature death in Toronto (Canada).
-The NEI is a composite index that was developed by the city to capture the differences in the City's neighbourhoods by ranking them based on socio-economic characteristics (e.g., social assistance, unemployment, income) and physical environmental characteristics such as green space availability.
-Other works use national census indicators such as the social and housing deprivation index e.g., [@pucciInequalitiesJobrelatedAccessibility2019], while others using census household poverty measures and calculated transport-related indicators like accessibility to estimate transport disadvantage [@sunPublicTransportAvailability2021; @scheurerSpatialAccessibilityPublic2017], evaluation of equity in policy implementation [@aldredEquityNewActive2021], and inequitable transport-related mortality burden [@iungmanImpactUrbanTransport2021].
-Similarly, Environmental Justice (EJ) indicators have been used in US literature to identify neighbourhoods that have a higher than average proportion of low-income and non-white populations (i.e., a composite vulnerable population group').
-Many studies has used EJ analysis to evaluate the equity impacts of transportation projects e.g., [@rowangouldIdentifyingEnvironmentalJustice2016; @parkTransitParksEnvironmental2021; @reddySafeguardingMinorityCivil2010].
+Many papers that focus on this category highlight differences in age-related capabilities; for instance, @martinez-jimenezAccessibilityCultureEducation2019 and @arranz-lopezSocialSpatialEquity2019 investigate travel distances and times to various opportunities based on specific age groups, acknowledging that age is an important consideration, and associated with differences in access to opportunities.
+The most commonly age groups are school-aged children and older populations.
+Research on school-aged children analyzes their wellbeing [@laszkiewiczChildrenGreenWalk2020], exposure to green space [@corazzaMethodologyEvidenceCase2020], access to schools [@sharmaSpatialSocialInequities2022], and aims at understanding and encouraging active travel journeys [@mackieOvercomingBarriersCycling2009; @mehdizadehWalkingTimeSchool2017].
+Papers that focus on older adults often have similar aims as those on children, and try to understand transport-related impacts on wellbeing [e.g, @chenSpatialAnalysisFramework2020], measuring accessibility to key destinations by population group [e.g.
+@chengInvestigatingWalkingAccessibility2019], and seeking to understand how to better meet travel needs [e.g., @nordbakkeTransportUnmetActivity2015].
+
+The third population group most commonly studied is what might be called 'composite vulnerable population groups' that exist at the intersection of several individual traits.
+These papers use some sort of composite vulnerability index that captures multiple population characteristics, including low-income, unemployment, immigrant status, family household characteristics, and so on.
+These indices are typically generated from official government sources or author-informed census data creation, and produced using a variety of methodologies.
+For instance, @awuorInfluenceEnvironmentalHealth2019 use the Neighbourhood Equity Index (NEI) to investigate disparities in premature death in Toronto (Canada).
+The NEI is a composite index that was developed by the city to capture differences in the City's neighbourhoods by ranking them based on socio-economic characteristics (e.g., social assistance, unemployment, income) and physical environmental characteristics such as green space availability.
+Other works use national census indicators such as the social and housing deprivation index [e.g., @pucciInequalitiesJobrelatedAccessibility2019 or explore transport disadvantage, equity in policy implementation, or transport-related mortality burden by means of census measures (e.g., household poverty) and transport-related indicators [e.g., accessibility; @@aldredEquityNewActive2021; @iungmanImpactUrbanTransport2021; @sunPublicTransportAvailability2021; @scheurerSpatialAccessibilityPublic2017].
+Similarly, Environmental Justice (EJ) indicators have been used in the US literature to identify neighbourhoods that have a higher than average proportion of low-income and non-white populations (i.e., a composite vulnerable population group').
+Numerous studies have used EJ analysis to evaluate the equity impacts of transportation projects [e.g., @rowangouldIdentifyingEnvironmentalJustice2016; @parkTransitParksEnvironmental2021; @reddySafeguardingMinorityCivil2010].
 
 `r word_comment("Papers that focus ONLY on (Dis)ability, or ONLY on Gender, or ONLY on Race/Ethnicity, or ONLY on Employment/Education level...
-the only focus", ".")` Papers that exclusively focus on populations with (dis)abilities e.g., [@parkJourneyVisuallyImpaired2017; @chiscanoImprovingDesignUrban2021; @orellanaWalkRollMapping2020] are relatively common in the reviewed literature.
-They mainly assess universal design guidelines and the ability for people with (dis)abilities to travel.
-However, from another perspective papers with an exclusive focus on gender, race/ethnicity, or education level/employment are much less common in the reviewed literature.
+the only focus", ".")` Papers that exclusively focus on populations with (dis)abilities e.g., [@parkJourneyVisuallyImpaired2017; @chiscanoImprovingDesignUrban2021; @orellanaWalkRollMapping2020] are relatively common in the literature.
+They often assess universal design guidelines and the capabilities of people who face disabilities to travel.
+However, from another perspective, papers with an exclusive focus on gender, race/ethnicity, or education level/employment are less common in the literature reviewed.
 Only two papers focus on gendered differences in cycling/active transportation (e.g., @adlakhaMindGapGender2020's case study in Chennai (India) and @xieWonCycleRoute2018's case study in Cardiff (UK)).
@@
 Furthermore, papers that focus *solely* on education/employment status are not present in the reviewed papers.
-This is to say, papers that feature gender, race/ethnicity, or education level/employment population groups often feature them alongside other population group characteristics (or within 'composite vulnerability measures').
-This contrasts the prominence of studies that exclusively center on (dis)abilities as a population category.
-
-The *Other* category are papers that include group population characteristics that are more difficult to classify.
-Examples include: veterans and access to specific-healthcare needs [@mooneyTravelDistanceBarrier2000], pregnant people and access to services [@vadrevuMeasuringSpatialEquity2016], and youth populations who live in foster care [@batscheUsingGISEnhance2012]<!-- rural populations people who use transit and their service needs relative to urban needs @sharavWhatTransitService2019, only cyclists in the context of health-implications of cycling in Delhi (India) @apparicioCyclingOneMost2021 ) -->.
+This is to say, papers that feature gender, race/ethnicity, or education level/employment population groups often feature them alongside other population group characteristics (as a trait in 'composite vulnerability measures').
+This contrasts the prominence of studies that exclusively center on disabilities.
+
+We include a catch-all category of papers (*Other*) that include group populations that less commonly have been the subject of research.
+Examples include: veterans and access to specific-healthcare needs [@mooneyTravelDistanceBarrier2000], pregnant people and access to services [@vadrevuMeasuringSpatialEquity2016], and youths who live in foster care [@batscheUsingGISEnhance2012]<!-- rural populations people who use transit and their service needs relative to urban needs @sharavWhatTransitService2019, only cyclists in the context of health-implications of cycling in Delhi (India) @apparicioCyclingOneMost2021 ) -->.
 Overall, the diversity of the *Other* population group classification demonstrates the diversity of transportation-equity concerns across population groups in the reviewed literature and the interplay of characteristics in the literature [@vecchioFairTransportPolicies2022].
 
-## Transport modes (the "Who" pt.2)
+## "Who" and "What": the tools of mobility
 
@@
 
-Shifting to transport modes of interest, the primary emphasis within the reviewed papers centers on public transit (@fig-fig5).
+As previously noted, mode of travel is not a trait of the individual, but in principle is a modifiable factor.
+That said, the existence of captive users by mode makes it challenging at times to extricate the mobility tool from the individual.
+In this section we choose to view the mode of travel primarily as an object of justice (i.e., a "what"), but occassionally this will be enjoined to the subject of justice (i.e., the "who").
+
+It is interesting from this perspective to notice that the primary emphasis within the literature reviewed centers on public transit (@fig-fig5)., a mode that often is perceived by users as less than ideal [@paezEnjoyment2010; @mellaDrivers2021] and that yet is often reckoned as the only realistic mobility tool for many [@jacquesRescuing2012; @welchMeasure2013; @cherancheryInvestigatingPerceptionCaptive2018].
 The variety of topics assessed from the perspective of public transit are varied.
 For instance, @mckeyCrowdsourcedMappingHealthy2020 identifies 'food deserts' in Dallas (USA) considering public transit accessibility.
-Other contributions intersect public transport and individual needs, such as, universal design and barrier-free transportation for people with disabilities e.g., [@jimenez-espadaResearchProblemUniversal2021; @liuStatusQuoChallenges2019] or how public transport can be improved/service be assessed for people with autism [@limFacilitatingIndependentCommuting2021; @feeleyValidationParatransitSkills2019].
-There may be a strong focus on public passenger transport systems because they can be altered to improve equity - it is a collective social system that has the potential to be sufficiently funded in order to provide barrier-free transport for most but it requires a variety of inputs that cannot always be met (e.g., sufficient density as a result of land-use, economic development, political will).
+Other contributions intersect public transport and individual needs.
+Examples include the study of universal design and barrier-free transportation for people who face disabilities [e.g., @jimenez-espadaResearchProblemUniversal2021; @liuStatusQuoChallenges2019] or possible improvements to public transport for people with autism [@limFacilitatingIndependentCommuting2021; @feeleyValidationParatransitSkills2019].
+The strong focus on public passenger transport systems could be due to their policy importance.
+They are a clear public good, and seen as a natural object of justice.
+Accordingly, public transportation is seen as adaptable to meet the demands of justice, for example by funding it sufficiently to provide barrier-free transport for most.
+In this respect, the literature points at a variety of factors that conspire against it, including low population and opportunity density (the outcome of decades of deliberate policy choices), fiscal prudence (the public face of the choice to lower taxes), and political will [to maintain or further unsustainable socio-technological systems, from SUVs to space tourism; @markardUnsustainabilities2023].
 
 Transit is also often central to multi-modal or holistic comparisons that may serve transport equity analysis.
-As an example, @brusselAccessAccessibilityCritique2019 compares three different approaches to measure accessibility in context to the Sustainability Development Goals (11.2) for the case of Bogota (Colombia), all of which capture some/all of the public transit system while others capture road and/or pedestrian systems.
-From a different perspective, @renneWhatHasAmerica2018 reviews natural disaster emergency evacuation plans from the lens of the no-car (and oftentimes vulnerable) households in regions across the USA, paying particular attention to transit and pedestrian networks.
-
-In contrast, a few papers in the reviewed literature frame transit modes as the 'car-free' option and compare transit to car access.
-This framing is notable as car travel can be seen as a direct competitor to transit or benchmark for travel times and accessibility levels [@golubUsingPrinciplesJustice2014; @martensJusticetheoreticApproachDistribution2012].
-As an example, @warrenDevelopingEquitableSustainable2015 develops a goal for per capita car ownership for the developing economy of Havana (Cuba), in recognition that car mobility is needed to alleviate transportation disadvantage in the short-term where public transit is not yet sufficiently addressed.
-In this context, this paper acknowledges the tension between household vulnerability and their need for mobility against sustainability goals of GHG emission reduction and car-dependency cycles.
+As an example, @brusselAccessAccessibilityCritique2019 compares three different approaches to measure accessibility in the context of the Sustainability Development Goals (11.2) for the case of Bogota (Colombia), all of which capture some/all of the public transit system while others capture road and/or pedestrian systems.
+From a different perspective, @renneWhatHasAmerica2018 reviews natural disaster emergency evacuation plans from the lens of car-less (and oftentimes vulnerable) households in regions across the USA, paying particular attention to transit and pedestrian networks.
+
+In contrast, only a few papers in the literature frame transit as a 'car-free' option and compare transit to car access.
+This framing is notable as car travel can be seen as a direct competitor to transit or as a benchmark for travel times and accessibility levels [@golubUsingPrinciplesJustice2014; @martensJusticetheoreticApproachDistribution2012].
+As an example, @warrenDevelopingEquitableSustainable2015 propose a standard for per capita car ownership for Havana (Cuba), in recognition that car mobility is needed to alleviate transportation disadvantage in the short-term where public transit has yet to be sufficiently addressed.
+In this context, @warrenDevelopingEquitableSustainable2015 acknowledge the tension between household vulnerability and their need for mobility against GHG emission reduction goals and car-dependency cycles.
 However, not all papers see transit as a direct competitor, but as a mode that can be used to satisfy individual capabilities.
-For instance, @smithAccessibilityCapabilityMinimum2012 focuses on households who live in rural areas in the UK and they synthesis their perspectives on their minimum transport needs and costs based on perceived minimum living standards for types of households (e.g., retired, no-children, with children, single, etc.).
+For instance, @smithAccessibilityCapabilityMinimum2012 focus on rural areas in the UK and various types of households (e.g., retired, no-children, with children, single, etc.).
+In this investigation, perspectives about minimum transport needs and costs for a variety of living standards are explored.
 The papers reviewed vary in the importance they place on climate urgency, with some focusing more on satisfying *all* sufficient individual needs while planning for less car-dependent cities in the future.
 
-Following a focus on transit, a focus on pedestrian modes is the second most common mode-focus in the literature.
-In the papers that focus exclusively on walking, many use or develop walkability scores to explore neighborhood perceptions [@evansAccessibilityUserNeeds2015] or pedestrian mobility focusing on middle-aged and older adults [@towneUsingWalkScore2016], gender [@kimNeighborhoodEnvironmentWalkability2016], or urban peripheries [@blecicCapabilitywiseWalkabilityEvaluation2021].
+Following a focus on transit, a focus on pedestrian travel is the second most common object of justice considered in the literature.
+Pedestrians are the most extreme case of confluence between the "what" and the "who", since the mobility tool is the person's own body, with its inherent range of abilities.
+In papers that focus exclusively on walking, many use or develop walkability scores to explore perceptions of neighborhood quality [@evansAccessibilityUserNeeds2015] or pedestrian mobility with a focus on middle-aged and older adults [@towneUsingWalkScore2016], gender [@kimNeighborhoodEnvironmentWalkability2016], or urban peripherical regions [@blecicCapabilitywiseWalkabilityEvaluation2021].
 These papers use 'walkability' as a way to measure the equity in its distribution.
@@
 
-Additionally, papers that focus on the pedestrian mode also often focus on multiple modes: they often discuss 'walkability' as part of active transportation, which focuses on both walking and bicycle and/or transit.
-Conceptualizations include how active transport contributes to children's physical activity levels [@mammenSchoolTravelPlanning2014], walkability as an alternative to car predominance [@bertrandMeasuringMappingDisparities2008] or tension that exists between modes, creating unsafe conditions for walking [@siuAssessmentPhysicalEnvironment2019; @ferenchakEquityAnalysisProactively2019].
-
-In terms of car-mode focuses, cars are infrequently the only mode within a paper that is examined.
-When car-mode is studied, it is often used as a comparison with transit or as the only mode of transport for areas with sub-standard transit systems e.g., [@kimmelStructuralBarriersComprehensive2018; @aljoufie2016].
-Similarly, car studies focus on externalities such as air pollution and safety [@fengTradeoffsMobilityEquity2014; @houstonProximityLicensedChild2006].
+Additionally, papers that focus on pedestrians also often see walking as a bridge to connect multiple modes: they often discuss 'walkability' as part of active transportation, which focuses on both walking and bicycle and/or transit.
+Concepts discussed include how active transport contributes to children's physical activity levels [@mammenSchoolTravelPlanning2014], walkability as an alternative to car predominance [@bertrandMeasuringMappingDisparities2008] or tension that exists between modes, creating unsafe conditions for walking [@siuAssessmentPhysicalEnvironment2019; @ferenchakEquityAnalysisProactively2019].
+
+In contrast to transit and walking, cars are seldom the only mode examined within a paper.
+When car as a mobility tool is studied, it is often used as a comparison with transit or as the only mode of transport for areas with sub-standard transit systems e.g., [@kimmelStructuralBarriersComprehensive2018; @aljoufie2016].
+Similarly, a number of car studies focus on externalities such as air pollution or safety [@fengTradeoffsMobilityEquity2014; @houstonProximityLicensedChild2006].
 
@@
 
-## Destinations (the "Who" pt.3)
-
-The majority of the papers do not focus on any particular destination (e.g., `r (dest_count$count[8]/sum(dest_count$count)) |> percent()` of studies).
-Within these papers, a variety of equity dimensions and modes are examined.
-Typically, they are multi-modal and focus on either dimensions that impact the **trip itself** (e.g., the trip experience, the quality of infrastructure, aspects of level of service) or the **people and relevant destinations that can be accessed** (e.g., a bundle of trips made for specific population groups, enough for 'sufficient' quality of life).
-For the first, the focus is on the quality of infrastructure, safety issues, perceived accessibility and dimensions of the level of service such as frequency [@zheEvaluationSharedUse2008; @prasertsubpakijEvaluatingAccessibilityBangkok2012; @furstBestBadPractices2013 ; @lattmanPerceivedAccessibilityPublic2016].
-For the second, the focus is a bundle of trips made for specific population groups, such as people who are physically impaired [@wilkinson-meyersLiveOrdinaryLife2015] or women [@russellPedallingEquityExploring2021] or broadly what is enough for 'sufficient' quality of life [@churchillTransportPovertySubjective2019].
+## "What": the benefits of mobility and accessibility
+
+The most fundamental benefits of transportation systems are mobility (enabling or impeding movement) and accessibility (the ease of reaching destinations).
+These benefits, however, are not necessarily valued by themselves, but rather are seen as instrumental to achieve an ulterior goal (e.g., activity participation).
+For example, although vehicle kilometers travelled (VKT) is still sometimes seen as a useful policy instrument [@zhaoRethinking2021], travelling more is not necessarily a sign of advantage when accessibility is low [@paezAccessibilityHealthCare2010], and similarly short trips may be a sign of advantage [@parkTransitParksEnvironmental2021].
+For this reason, although the right to the road (and to transportation systems more generally) is important, the literature leans heavy on the ulterior object, namely accessibility to destinations.
+
+In this respect, we find that most papers tend to be wide sighted, in that they do not focus on a particular destination (e.g., `r (dest_count$count[8]/sum(dest_count$count)) |> percent()` of studies).
+Within these papers, a variety of equity dimensions, often in combination with different modes of transportation.
+Typically, they are multi-modal and focus on factors that impact the **trip itself** (e.g., the trip experience, the quality of infrastructure, the level of service) or on the **people and relevant destinations that can be accessed** (e.g., a bundle of trips made by specific population groups, and whether they are enough for 'sufficient' quality of life).
+For the first, the focus is on the quality of infrastructure, safety issues, perceived accessibility and facets of the level of service such as frequency [@zheEvaluationSharedUse2008; @prasertsubpakijEvaluatingAccessibilityBangkok2012; @furstBestBadPractices2013 ; @lattmanPerceivedAccessibilityPublic2016].
+For the second, the focus is on bundles of trips tailored for specific population groups, such as women [@russellPedallingEquityExploring2021] or people who face physical disabilities [@wilkinson-meyersLiveOrdinaryLife2015].
+More broadly there is the question of what is enough for the quality of life to be 'sufficient' [@churchillTransportPovertySubjective2019].
 These papers further demonstrate the multi-dimensional role of transportation systems: they provide a utilitarian service that can be used to get from A to B but they too are experienced by the people that use them.
-The papers that examine 'all trips' best exemplify this trend in the transportation equity literature.
-
-In terms of the papers that study specific destinations, the most commonly studied activity types includes healthcare services (`r (dest_count$count[6]/sum(dest_count$count)) |> percent()`) followed by employment destinations (`r (dest_count$count[5]/sum(dest_count$count)) |> percent()`).
-Papers that exclusively focus on healthcare typically originate from the healthcare planning literature, and look to inform planners where disparities in services exist and what can be done about it.
-For instance, @wangModelingSpatialAccessibility2011 models access of the Chinese-language speaking population in Toronto (Canada) to Mandarin-speaking family physicians and suggests (inferred) that a spatial mismatch in the supply and demand is not equitable.
-Papers that exclusively focus on employment typically focus on these trips as they are the most common trip purpose and are often correlated with other trip activities like shops, recreation, and other services, generally speaking.
-For instance, @allenSizingTransportPoverty2019a operationalizes a low employment-based accessibility threshold and a composite population vulnerability index to identify neighborhoods in transport poverty for eight cities in Canada.
-Papers focused on healthcare and employment typically source data from representative travel surveys/diaries, census data, and point-of-interest databases: they often rely on well developed and institutional data that represents 'typical trips', especially in the Global North where this data is more readily available.
+Papers that examine 'all trips' best exemplify this trend in the transportation equity literature.
+
+In terms of papers that study specific destinations, the activities most commonly studied include healthcare services (`r (dest_count$count[6]/sum(dest_count$count)) |> percent()`), followed by employment (`r (dest_count$count[5]/sum(dest_count$count)) |> percent()`).
+Papers that exclusively focus on healthcare typically originate from the healthcare planning literature, and look to inform planners about disparities in services and corrective actions.
+For instance, @wangModelingSpatialAccessibility2011 model access of Chinese-language speakers in Toronto (Canada) to Mandarin-speaking family physicians.
+These authors infer an inequity from the spatial mismatch between language competent healthcare providers and Chinese speakers.
+Papers that exclusively focus on employment typically focus on these trips as they are the most common trip purpose and are often correlated with other trip activities like shops, recreation, and other services.
+For instance, @allenSizingTransportPoverty2019a propose a low employment-based accessibility threshold and a composite population vulnerability index to identify transportation-poor neighborhoods in urban Canada.
+Papers that focus on healthcare and employment typically source data from representative travel surveys/diaries, census data, and points-of-interest databases.
+In other words, these studies often benefit from well developed and institutional data that represents 'typical trips', especially in the Global North where these data are more readily available.
+
+<!-- DATA AVAILABILITY AS A MATTER OF JUSTICE: IS THIS AN ADDITIONAL CALL FOR THE NEXT SECTION?
+IN FACT, THE NATIONAL TRANSPORT POVERTY SURVEY EXPLICITLY AIMS TO ADDRESS THE LACK OF DATA FOR PARTICULAR POPULATIONS IN CANADA!
+The relevant issues are not new, and have at least two parts (i.e., who counts?
+Who is measured?
+Who is a subject of justice?
+And, who gets to use the data?).
+With respect to the first part, who we count and how we count them has implications for our understanding of just who is a subject of justice, as illustrated by the history of racial classifications in the US [see @leeRacial1993].
+In the case of transportation, the issue of data (collection/availability) as a matter of justice is gaining traction as digitalization of data casts a starker light on these questions...for examples, see the papers by @sourbatiSmart2021 and also, yes, by @behrendtData2023 (with Mimi Sheller).
+-->
 
 But what about non-healthcare and non-employment activity types?
-Papers that focus on other activities are not framed as a 'typical travel pattern' and they have different intentions.
-For instance, papers that focus on places for shopping such as grocery stores or markets (`r (dest_count$count[7]/sum(dest_count$count)) |> percent()`), often aim to identifying food deserts e.g., [@choiFoodDesertsActivity2013; @jiaoHowIdentifyFood2012; @mckeyCrowdsourcedMappingHealthy2020; @kimAssessingSocialSpatial2020].
-Papers that focus on educational facilities including primary school, secondary school, and post-secondary school represent `r (dest_count$count[3]/sum(dest_count$count)) |> percent()` of the studies, and examine children's active transportation to school e.g., [@laszkiewiczChildrenGreenWalk2020] and universal design e.g., [@larkinsAccessibleTransportationBuilt2011].
+Papers that focus on other activities are not framed as 'typical travel patterns' and they often have different intentions.
+For instance, papers that focus on shopping destinations, such as grocery stores or markets (`r (dest_count$count[7]/sum(dest_count$count)) |> percent()`), frequently aim to identify food deserts [e.g., @choiFoodDesertsActivity2013; @jiaoHowIdentifyFood2012; @mckeyCrowdsourcedMappingHealthy2020; @kimAssessingSocialSpatial2020].
+Papers that focus on educational facilities, including primary, secondary, and post-secondary schools, represent only `r (dest_count$count[3]/sum(dest_count$count)) |> percent()` of the studies, and examine children's active transportation to school [e.g.,@laszkiewiczChildrenGreenWalk2020] or universal design [e.g., @larkinsAccessibleTransportationBuilt2011].
 When green space or other places of leisure is the exclusive focus (`r (dest_count$count[6]/sum(dest_count$count)) |> percent()` of papers), studies examine different accessibility questions such as the spatial distribution of green space e.g., [@xuSocialInequalitiesPark2017], for whom its accessible to e.g., [@mavoaAreaLevelDisparitiesPublic2015], and why e.g., [@mearsUnderstandingSocioeconomicEquity2019].
@@
 In the few papers that do include them, these destinations are considered in a holistic representation of activity participation [@albertsRebuildingWomenLivelihoods2016; @smithAccessibilityCapabilityMinimum2012].
-
-Conclusions and recommendations from the papers are varied: they span across population groups, modes, and activity types -- in addition to across equity dimensions, conceptualizations, as we exemplify below.
-
-## Methods used across equity conceptualizations and standards (the "How")
+The lack of information about community destinations is particularly pronounced in the case of children, as per @desjardinsChildren2022.
+
+The literature offers a variety of conclusions and recommendations that span across the benefits of transportation (e.g., destinations types), often comingled with population groups and the tools of mobility.
+In addition, as we will discuss next, there is a variety of perspectives in terms of the philosophical foundations and the standards of fairness.
+
+<!--
+
+EDITED UP TO THIS POINT.
+
+-->
+
+## "How": concepts of justice and standards of fairness
 
@@
 
-# Discussion and recommendations
+# Moving forward: discussion and recommendations
+
+> "_Justice is itself the great standing policy of civil society; and any eminent departure from it, under any circumstances, lies under the suspicion of being no policy at all._"
+> — Edmund Burke
 
@@
 
-In reference to transport inequity continumus, papers on the Global South tend to operationalize international standards as guidelines.
+In reference to transport inequity continuum, papers on the Global South tend to operationalize international standards as guidelines.
 Though not covered in this review, Global South nations' formal processes for planning are newer (relatively), more fragile, and operating under more strict financial constraints than the Global North.
@@
 
+## Call 3: Data justice
+
+<!-- see my comments above -->
+
 What are the sources of data and what are the motivations for some categories?
@@
 
-## Call 3: More direct and explicit links between standards and experienced outcomes
+## Call 4: More direct and explicit links between standards and experienced outcomes
 
@@
 
-## Call 4: Evaluations of interventions and policies
+## Call 5: Evaluations of interventions and policies
 
@@
 THESE THREE PARAGRAPHS SOUND LIKE A CONCLUSION.
-PERHAPS A CALL TO ACTINO - TO BE ABLETO MAE AN INTERNATIONAL COMPARISION ON JUSTICE.
+PERHAPS A CALL TO ACTION - TO BE ABLE TO MAKE AN INTERNATIONAL COMPARISION ON JUSTICE.
 WE NEED TO KNOW THE WHY BUT LITERATURE SHOULD START TO OUTLINE THAT EXPLICILITY.
@@
 
-<!--
-MOVE TO CONCLUSIONS?
--->
-
 # Concluding remarks
+
+> "The way to right wrongs is to turn the light of truth upon them._" 
+> — Ida B. Wells
 

### 2023-10-22 Manuscript/Manuscript.qmd
Antonio completed first exhaustive set of edits. This is now going back to Anastasia.

@@
 Meanwhile, firmly grounded on planet Earth, the employees that Bezos thanked for his suborbital escapade were struggling with some very mundane problems of their own, and none as lofty as conflicting schedules that prevented them from flying in rockets [@griffinPerson2021].
-According to reports, people employed directly or indirectly by Amazon for warehouse or delivery work had, for years, been treated to "inhumane" conditions [@fungEuropean2018; @scottAmazon2019; @greeneAlgorithm2021], and subjected to surveillance on the job, degrading schedules that drove drivers to urinate in bottles, crushing demands for productivity quotas that led injury, facing little or no job security, and treated less with gratitude than as disposable inputs to feed Amazon's earnings and consumers' demands [@tungAmazon2020; @BBCNews2021; @reeseTeamsters2022; @middletonRobot2023].
+According to reports, people employed directly or indirectly by Amazon for warehouse or delivery work had, for years, been treated to "inhumane" conditions [@fungEuropean2018; @scottAmazon2019; @greeneAlgorithm2021], and subjected to surveillance on the job, degrading schedules that drove drivers to urinate in bottles, crushing demands for productivity quotas that led to injury, all the while facing little or no job security.
+From this perspective, they were treated less with gratitude and more as disposable inputs to feed Amazon's earnings and consumers' demands [@tungAmazon2020; @BBCNews2021; @reeseTeamsters2022; @middletonRobot2023].
 
@@
 The question of _fairness_, however, is not a simple one, and most of us would probably have been stumped to explain in a clear and precise way, beyond a disquieting intuition, just why the above picture was disturbing.
-_Justice_ is a political ideal based on the principle that individuals should be treated in a fair and equitable manner [@gosslingUrban2016, p.
+_Justice_ is a political ideal based on the principle that individuals should be treated in a _fair_ and _equitable_ manner [@gosslingUrban2016, p.
 2], giving and receiving whatever they are _due_ [@jaggarPhilosophicalChallengesGlobal2009, pp.
@@
 For starters, the notion that people are "due" something depends on the values of a society, as embodied in its systems and institutions [@karnerFrom2020].
-Values,in turn, are not subject to natural laws, but rather are the result of intersubjective realities, which is to say illusions whose legitimacy derives from a collective will to believe.
-For example, justice would likely mean something very different to a person in a democratic society, than to another in a society where they owed their all—perhaps including their own time or bodily autonomy—to some collective illusion (e.g., the state, a monarch).
+Values, in turn, are not subject to natural laws, but rather are the result of intersubjective realities, which is to say illusions whose legitimacy derives from a collective will to believe.
+For example, justice would likely mean something very different to a person in a democratic society, than to another in a society where they owed their all—perhaps including their own time or bodily autonomy—to some collective illusion (e.g., the state, or a monarch).
 As well, the meaning of "justice" would likely differ in yet another society where very few owned most, and most owned very little due to a different illusion (e.g., that wealth equates merit).
 In these two hypothetical cases, elucidating the meaning of "fair" would in all certainty be beyond the dreams of most, since fair would be what the state, the monarch, or the extremely wealthy said it was.
-In contrast, in democratic societies^[Paraphrasing, democracy is the worst of all collective illusions, except for all others.] individuals rights are valued above the whims of the few.
+In contrast, in democratic societies^[Paraphrasing, democracy is the worst of collective illusions, except for all others.] individuals rights are valued above the whims of the few.
 In such a setting, the task of defining a "just" distribution of the burdens and benefits of things—from income, to roads, to space travel—quickly becomes muddled, encumbered even, by the necessity to pay attention to a multitude of voices, not all of them equally loud.
@@
 Finally, previous reviews of planning documents have investigated equity from narrowly scoped perspectives, such as accessibility [@boisjolyHowGetThere2017] or a particular mode of transportation [e.g., cycling; @doranPursuitCyclingEquity2021].
-These inquiries are valuable to scholar, planning agencies, and decision-makers alike, and the present review will tread similar, but not completely overlapping ground.
+These inquiries are valuable to scholars, planning agencies, the public, and decision-makers alike, and the present review will tread similar, but not completely overlapping ground.
 In our estimation, there remains a gap in the literature in terms of understanding how standards for equity are developed and implemented in the transportation domain.
@@
 These questions boil down to the development and use of _standards_ for transportation justice.
-The term "standard" is used to mean "something set up and established by authority as a rule for the measure of quantity, weight, extent, value, or quality"^[https://www.merriam-webster.com/dictionary/standard].
+The term "standard" connotes "something set up and established by authority as a rule for the measure of quantity, weight, extent, value, or quality"^[https://www.merriam-webster.com/dictionary/standard].
 How much pollution is allowed to be generated, and where, depends on who is affected, and how much health is valued overall, as well as by whom.
-For example, firms may or may not adopt lower standards for the emission of pollutants in poorer areas; it might be that poor people end up being relegated to areas that adopt lower standards [@gouldsonFirms2006].
+For example, firms may or may not adopt lower standards for the emission of pollutants in poorer areas; it might be that poor people end up being relegated to areas that already had lower standards [@gouldsonFirms2006].
 Regardless of the cause, the result is the same: pollution tends to be worse were poorer people are [@delucaCode2012].
@@
 Ultimately, this review aims to collate the existing academic knowledge on the matter, and present it in a manner useful to support the development and implementation of standards for equity in transportation planning and policy.
-In this, we aim to provide relief even, or particularly, to planners in those places where legislation explicitly calls for justice^[[As recently as -@martensFair2021, Martens and Golub note that Federal directives related to Title VI in the US "do not provide guidelines that can help agencies develop explicit standards to assess the distribution of accessibility benefits from projects or plans."]].
+In this, we aim to provide relief even, or particularly, to planners in those places where calls for justice are explicitly made through legislation^[[As recently as -@martensFair2021, Martens and Golub note that Federal directives related to Title VI in the US "do not provide guidelines that can help agencies develop explicit standards to assess the distribution of accessibility benefits from projects or plans."]].
 
@@
 We then describe the methods used for searching, selecting, and reviewing the relevant literature.
-<!--
-In Section 2, we set the stage by defining what do we mean by equity dimensions, conceptualizations and types of standards used in the literature search and throughout the paper.
-In Section 3, we outline the methods used in the scoping review conducted.
-In Section 4, we summarize the findings of the literature.
-In Section 5, we discuss the methods connects to standards, conceptualizations and dimensions.
-In Section 6, we summarize the findings and link them to future transport planning agendas in addition to visioning ways the findings can be used to build *just* transportation planning processes.
--->
+This is followed by a description of the landscape surveyed, which is then appraised critically.
+We will conclude with some calls for action to improve the practice of setting and using standards of fairnes for transportation justice.
 
@@
 
-Let us begin by stating that justice is an end goal, that is, a desirable state of affairs that we strive to achieve.
+Let us begin by stating that justice is an end goal, that is, a desirable state of affairs that we are morally obligated to achieve.
 
@@
 It might be argued that all of the above are particular forms of distributive justice.
-Retributive justice, for example, is often concerned with the distribution of the benefits and burdens of being a member of society: the way it is usually achieved is by distributing intangibles like "freedom" (e.g., of movement, of association) as benefits, and/or claiming tangibles like money as burdens (e.g., fines).
-Reparative justice often distributes current benefits and burdens to redress past wrongs, for example by asking those who have benefited from past wrongs, even if unwittingly, to shoulder a bigger fiscal burden in order to cover programs that mete benefits to those who are still harmed by past wrongs.
+Retributive justice, for example, is often concerned with the distribution of the benefits and burdens of being a member of society; the way it is usually achieved is by distributing intangibles like "freedom" (e.g., of movement, of association) as benefits, and/or claiming tangibles like resources as burdens (e.g., fines).
+Reparative justice often distributes current benefits and burdens to redress past wrongs, for example by asking those who have benefited from said wrongs, even if unwittingly, to shoulder a bigger fiscal burden in order to cover programs that mete benefits to those who are still harmed by past wrongs.
 Procedural justice is the distribution of the benefits (e.g., the right to voice an opinion) and burdens (e.g., the effort required to develop an educated opinion) of the processes that lead to decisions with collective consequences.
@@
 We can then speak of the _purposes_ of distributive justice: to mete out retribution _fairly_, to repair past harms, and to ensure that procedures offer _equitable_ opportunities to influence outcomes.
-Equity and fairness are the instruments of justice, the tools by which society advances towards the end goal of justice.
+Equity and fairness from this perspective are the instruments of justice, the tools by which society advances towards the end goal of justice.
 
@@
 The concept does not have the same history of development as an analytical tool, and can be interpreted in numerous, and possibly discordant ways.
-That this is the case is convincingly demonstrated by @martensFair2021 in their study of the application of Title VI of the Civil Rights Act of 1964 in accessibility planning practice in the US.
+That this is the case is convincingly demonstrated by @martensFair2021 in their study of the application of Title VI of the Civil Rights Act of 1964 in accessibility planning in the US.
 Title VI explicitly talks about the distribution of benefits derived from Federal funding: "[n]o person in the United States shall, on the ground of race, color, or national origin, be excluded from participation in, be denied the benefits of, or be subjected to discrimination under any program or activity receiving Federal financial assistance." However, as @martensFair2021 show, there are several ways to comply with regulations while achieving different outcomes, ranging from the banal (do not _knowingly_ discriminate), to the substantive (compensation for past discrimination, i.e., reparative justice).
@@
 
-The last concept that we discuss in this section is that o a standard.
+The last concept that we discuss in this section is that of a standard.
 Briefly, standards are a way of making concrete statements about fairness.
 Returning to the ambiguities in Title VI discussed by @martensFair2021, the attainment of justice depends on the standard used to indicate fairness.
-For example, explicit non-discrimination constitutes a very weak standard that takes aim at the actions of agencies instead of the recipients of the benefits; accordingly, any distribution of benefits would be considered fair, as long as the agency does not explicitly and knowingly targets or denies the benefits to particular groups.
+For example, explicit non-discrimination constitutes a very weak standard that takes aim at the actions of agencies instead of the recipients of the benefits; accordingly, any distribution of benefits would be considered fair, as long as the agency does not explicitly and knowingly target or deny benefits to particular groups.
 The standard provides conditions to determine whether a situation is _fair_.
@@
 An egalitarian standard would weigh the benefits or burdens by population, and fairness is achieved when each group give or receive in proportion to their size.
-An affirmative action standard is even stricter, since it requires the benefits to be distributed in a non-egalitarian way that favors those who are still harmed by past or present discriminatory practices.
+In contrast, an affirmative action standard is even stricter, since it requires the benefits to be distributed in a non-egalitarian way that favors those who are still harmed by past or present discriminatory practices.
 
@@
 Conventionally this meant the in-group [e.g., members of the same nation state; @jaggarPhilosophicalChallengesGlobal2009, p.
-3], a notion since expanded (e.g., with the Universal Declaration of Human Rights) to include all members of the human species, but has yet to cover other living creatures.
-<!-- SOUTH AMERICAN CONSTITUON THAT DOES INCLUDE IT NOTE FOR ANASTASIA--> In the case of justice in transportation, the question of "where?" is paramount; it might be argued that, by their very nature, transportation generates inequalities.
+3], a notion since expanded (e.g., with the Universal Declaration of Human Rights) to include all members of the human species.
+In the case of justice in transportation, the question of "where?" is paramount; it might be argued that, by their very nature, transportation generates inequalities.
 By concentrating the effects of space-time convergence (for instance, by providing access to a transit system or a highway), an inequality is automatically generated.
 The burdens of transportation, in contrast, are often diffuse.
-They are incrementally paid, for example by a distributed population in the form of taxes by a spatially distributed population, or in poor health by people who were not party to transportation decisions made elsewhere.
+They are incrementally paid, for example by a distributed population in the form of taxes, or by a population with a different spatial distribution in the form of poor health.
 
@@
 
-- When we ask **"Who?"**, we inquire about which entities should be regarded as subjects of justice <!-- OF INEQUITIES / MEASURING THE INEQUITIES OF 'WHO'.
+- When we ask **"Who?"**, we inquire about which entities should be regarded as subjects of justice,<!-- OF INEQUITIES / MEASURING THE INEQUITIES OF 'WHO'.
 LIKE INCOME, THE MOBILITY TOOL IS MODIFIABLE -- THE MOBILITY TOOL IS NOT MODIFIABLE IN MANY CASES, IT IS WHO YOU ARE .
 T IS ALMOST LIKE A CHARACTERISTICS OF THE PERSON.
-PART OF THE SOCIAL CONTRACT THAT IDENTIFIES EOPEL WITH THEIR MOBILITY TOOL-->, meaning who are entitled to make claims of moral consideration from the perspective of justice.
-To make it tractable, this question is often approached through the filter of population groups, which may include the several traits, including gender identity, ableness, ethnicity, age, caste, and income.
-Often, it is appropriate to reflect on the intersections between traits, given evidence that the lived experiences of, say, a White woman and a Black women, can be markedly different.
+PART OF THE SOCIAL CONTRACT THAT IDENTIFIES EOPEL WITH THEIR MOBILITY TOOL--> meaning those entitled to make claims of moral consideration from the perspective of justice.
+To make it tractable, this question is often approached through the filter of population groups, which may include several concurrent traits, such as gender identity, ableness, ethnicity, age, caste, and income.
+Often, it is appropriate to reflect on the intersections between traits, given evidence that the lived experiences of, say, a White woman and a Black women, can be markedly different between each other, in addition to being different from those of men.
 A possible complication in the case of transportation is that disentangling the "who" from the tools available to them is not always straightforward.
@@
 In societies that have grown into transportation monocultures with a predilection for automobility [@millerCollaborative2011] there may actually be less choice about mobility tools than we would like to assume.
-So, while it is important to avoid conflation of the "who" with the "what",  for analytical purposes we need to be mindful of the connection between person and their mobility tools.
-In the case of transportatin, in addition to members of the public who use transportation systems, there is another category of **who**, that stand possibly in opposition to users, namely the entities charged with providing services, maintaining infrastructure, and so on.
+So, while it is important to avoid conflating the "who" with the "what", for analytical purposes we need to be mindful of the connection between person and their mobility tools.
+In the case of transportation, in addition to members of the public who use transportation systems, there is another category of **who**, that stands possibly in opposition to users, namely the entities charged with providing services, maintaining infrastructure, and so on.
 These could be ministries or departments of transportation, transit agencies, public works departments and others.
-Identifying these entities is relevant to ellucidate who is responsible for apportioning the benefits or mitigating the burdens of transportation.
+Identifying these entities is relevant to ellucidate who is responsible, for example for apportioning the benefits or mitigating the burdens of transportation.
 
@@
 In other words, what do transportation systems _produce_?
-At their most fundamental, transportation systems are space-time convergence tools, tools that improve the rate at which time is traded for space.
-They usually do this by increasing the speed of movement: sidewalks facilitate walking, traffic lights facilitate the ordered flow of vehicular traffic, and a launching pad makes it possible for a rocket to takeoff.
-With complex interlocking parts (sidewalk, road, traffic sign, parking regulations), transportation systems produce _mobility_, the potential for movement.The realization of this potentials happens through travel but, as the adage goes, travel is derived demand, which seems to hold for most (even if not all) situations [e.g., @mokhtarianUnderstanding2001; @redmondPositive2001; @paezEnjoyment2010; @whalenMode2013].
-For this reason, we cannot stop at considering mobility, but the ultimate goal of mobility, which is to reach destinations.
+At their most fundamental, transportation systems are space-time convergence technologies, tools that improve the rate at which time is traded for space.
+They usually do this by increasing the speed of movement: sidewalks facilitate walking, traffic lights facilitate the ordered flow of vehicular traffic, and a launching pad makes it possible for a rocket to take off.
+With complex interlocking parts (sidewalk, road, traffic sign, parking regulations), transportation systems produce _mobility_, the potential for movement.
+The realization of this potentials happens through travel.
+However, as the adage goes, travel is derived demand, which seems to hold for most (even if not all) situations [e.g., @mokhtarianUnderstanding2001; @redmondPositive2001; @paezEnjoyment2010; @whalenMode2013].
+For this reason, we cannot stop at considering mobility, but the ulterior goal of mobility, which is to reach destinations.
 In combination with land use systems (the spatial distribution of opportunities on the landscape), mobility produces _accessibility_, the potential to reach destinations.
-For this reason, we can think of the objects of transportation justice as being _proximate_ (the tools of mobility, mobility itself), and _ulterior_ (accessibility, opportunities for activity participation).
+In this manner, we can think of the objects of transportation justice as being _proximate_ (the tools of mobility, mobility itself), and _ulterior_ (accessibility, opportunities for activity participation).
 The burdens of transportation are also many and varied.
@@
 
-- The next question is **"How?"**, and it relates to the allocation of various objects of justice ("what") to various subjects of justice ("who") in various circumstances.
+- The next question is **"How?"**, and it relates to the allocation of various objects of justice ("what") to various subjects of justice ("who") in various circumstances ("when" and "where").
 **Equity standards** are a tool for answering this distributive question: how do we allocate burdens and benefits and to whom?
-Standards are thresholds that when operationalized effectively define what is in/equitable.
-The thresholds can be quantitative (e.g., square meters of green space per capita), or they can be qualitative descriptions (e.g., do not knowingly discriminate), or a mixed of the two.
+Standards are thresholds that when operationalized effectively define what is fair, that is, in/equitable.
+The thresholds can be quantitative (e.g., square meters of green space per capita), or they can be qualitative descriptions (e.g., do not knowingly discriminate), or a mix of the two.
 Some examples include: maximum travel distance/cost/time to or from key destinations, levels of maximum exposure to externalities (i.e., noise or air pollution),un/fulfilled needs, and dis/satisfaction with travel.
@@
 Justice is an inherently social construct <!--, and does not exist outside of human society-->.
-Asking **why?** amounts to asking what sort of social contracts regulate human interactions.
+Asking **why?** amounts to asking what sort of social contracts regulate human interactions, that is, what are the rules that our collective will to believe impose on each of us.
 These contracts can be defined by constitution, but there are often unwritten and possibly contested variants.
@@
 Crook argued that the goal in teaching his kids (then aged 7, 8, 9 and 11) to use transit to go to school was to raise self-reliant children [@stueckYoung2017].
-While both sides discovered that there were not clear rules about kids riding transit alone [@brendHere2017], statistically there is evidence that travel by car is riskier than travel by bus [@morencyTraveling2018].
+While both sides discovered that there no clear rules existed about kids riding transit alone [@brendHere2017], statistically there is evidence that travel by car is riskier than travel by bus [@morencyTraveling2018].
 If the public good is to reduce risk, should not children be banned from riding in cars until they are ten?
-In a functioning democracy like Canada, we have chosen to care about fundamental freedoms and rights.
+In a functioning democracy like Canada, people have chosen to care about fundamental freedoms and rights.
 This provides an answer as to "why" we would even consider children as subjects of justice.
-But the devil is in the details, and the other questions discussed above are needed to pin the devil by the tail.
+But the devil is in the details, and answering the other questions discussed above is essential to pin the devil by the tail.
 Are children under the age of 10 subjects of justice (i.e., are they a "who")?
 If so, is the ability to use transit unsupervised an object of justice (i.e., the "what")?
-When and where do these questions apply (e.g., riding the bus to school, or to a social event; during the day, or late at night)?
+When and where do these questions apply (e.g., to to to school, or to a social event; during the day, or late at night)?
 Eventually, in 2020, a court of appeals ruled in Crook's favor, allowing his children to continue to ride the bus, thus providing much needed clarity with respect to the use of transit by children who desire independent mobility [@stueckCourt2020].
-In actuality, the courts provided a standard of equity: starting at a certain age, children are objects of justice from the perspective of freedom of movement, and they are due the benefits of independent mobility by transit.
+In actuality, the courts _created_ a standard of equity: starting at a certain age, children are objects of justice from the perspective of freedom of movement, and they are due the benefits of independent mobility by transit.
 
@@
 This review examines the breadth and depth of the academic literature on transportation to identify the extent to which standards for equity are defined and employed.
-In this task, we follow the Joanna Briggs Institute (JBI) approach to the conduct of scoping reviews.
-This approach builds upon the Arksey and O'Malley (2005) framework [@petersUpdatedMethodologicalGuidance2020].
-The JBI is .
-
-Further, the review is guided too by the Preferred Reporting Items for Systematic Reviews and Meta-Analyses extension for scoping reviews (PRISMA-ScR), which is consistent with the JBI approach [@triccoPRISMAExtensionScoping2018].
+In this task, we follow the Joanna Briggs Institute (JBI) approach to the conduct of scoping reviews, an approach that builds upon the Arksey and O'Malley (2005) framework [@petersUpdatedMethodologicalGuidance2020].
+
+Further, the review is also guided by the Preferred Reporting Items for Systematic Reviews and Meta-Analyses, particularly the extension for scoping reviews (PRISMA-ScR), which is consistent with the JBI approach [@triccoPRISMAExtensionScoping2018].
 The use of these methods allows us to explore, in a consistent and organized manner, a relatively specialized topic within the broader transportation literature.
@@
 The primary research question and the protocol were initially defined in consultation among the authors of the report.
-In other words, the starting point was the level of knowledge of experts in the subject matter.
-The initial draft of the protocol was refined from preliminary searches of related-reviews [e.g., @iglesiasHowUnevenUrban2019; @sagarisUsingPARFrame2020; @vecchioAccessibilityCapabilitiesApproach2021], and in consultation with a University of Toronto Research Services Librarian and Liaison Librarian in City Studies.
+In other words, the starting point for the review was the level of knowledge of a group of experts in the field of transportation.
+The initial draft of the protocol was refined from preliminary searches of related-reviews [e.g., @iglesiasHowUnevenUrban2019; @sagarisUsingPARFrame2020; @vecchioAccessibilityCapabilitiesApproach2021], and in consultation with a University of Toronto Research Services Librarian and a Liaison Librarian in City Studies.
 
@@
 Papers that focus on educational facilities, including primary, secondary, and post-secondary schools, represent only `r (dest_count$count[3]/sum(dest_count$count)) |> percent()` of the studies, and examine children's active transportation to school [e.g.,@laszkiewiczChildrenGreenWalk2020] or universal design [e.g., @larkinsAccessibleTransportationBuilt2011].
-When green space or other places of leisure is the exclusive focus (`r (dest_count$count[6]/sum(dest_count$count)) |> percent()` of papers), studies examine different accessibility questions such as the spatial distribution of green space e.g., [@xuSocialInequalitiesPark2017], for whom its accessible to e.g., [@mavoaAreaLevelDisparitiesPublic2015], and why e.g., [@mearsUnderstandingSocioeconomicEquity2019].
-Very few reviewed papers include 'community' destinations (including public service centres, places of community support, and places of worship (`r (dest_count$count[2]/sum(dest_count$count)) |> percent()` of studies) or childcare activity types (`r (dest_count$count[1]/sum(dest_count$count)) |> percent()` of studies).
+When green space or other places of leisure is the exclusive focus (`r (dest_count$count[6]/sum(dest_count$count)) |> percent()` of papers), studies examine different accessibility questions such as the spatial distribution of green space e.g., [@xuSocialInequalitiesPark2017], for whom its accessible to [e.g., @mavoaAreaLevelDisparitiesPublic2015], and the reasons why the distribution of these spaces may be unequitable [e.g., @mearsUnderstandingSocioeconomicEquity2019].
+Very few of the papers we reviewed include 'community' destinations (e.g., public service centres, places of community support, and places of worship (`r (dest_count$count[2]/sum(dest_count$count)) |> percent()` of studies) or childcare activity types (`r (dest_count$count[1]/sum(dest_count$count)) |> percent()` of studies).
 In the few papers that do include them, these destinations are considered in a holistic representation of activity participation [@albertsRebuildingWomenLivelihoods2016; @smithAccessibilityCapabilityMinimum2012].
@@
 In addition, as we will discuss next, there is a variety of perspectives in terms of the philosophical foundations and the standards of fairness.
-
-<!--
-
-EDITED UP TO THIS POINT.
-
--->
 
@@
 `r word_comment("Answering: what conceptualizations occur in papers with opportunity standards?
-what types of methods?", ".")` Broadly summarizing how equity standards connect with conceptualizations, some trends emerge in methods used.
-**Opportunity** and **population** standards appear in the literature at similar proportions ( `r (stand_count$count[3] /sum(stand_count$count)) |> percent(accuracy=0.1)` vs.
-`r (stand_count$count[4] /sum(stand_count$count)) |> percent(accuracy=0.1)`), but frequently correspond to different conceptualizations.
-Referring to @fig-fig6, over 40% of papers that focus on **horizontal equity** and **spatial equity** conceptualizations suggest **opportunity** standards.
-Within these papers, travel impedance standards (a type of opportunity standard) are often suggested and accessibility indices are operationalized (i.e., a measure of the potential to interact with opportunities for populations located at each spatial unit within a region).
-Examples include: @chenImpactHighspeedRail2017 use a travel time threshold of 4 hours or less on high-speed rail from one municipality to another to be considered "comfortably connected", @yenisettyMeasuringAccessibilityVarious2020 assumes that populations living in areas where the travel distance to a public transit station is less than 1,200m is sufficient to interact with the transit system, and @shenMeasuringSpatialAccessibility2020 identifies regions where populations cannot access hospitals within 1 hour by car.
-Papers suggesting opportunity standards often employ disparity analysis through a variety of quantitative approaches such as inequality measures (e.g., Gini coefficient and Lorenz curves, poverty measures [@vanderveenOperationalizingIndicatorSufficient2020; @tiznadoaitkenRoleAccessibilityPublic2018]), spatial descriptive analysis, and the comparison to benchmarks (e.g., equal supply to demand of public transit in a spatial unit as done by @peungnumsaiGridBasedSpatialAnalysis2020) to determine which locations are spatially and horizontally (in)equitable relative to other locations in the studied region.
-Another branch of quantitative research conceptualize transportation system externalities as trade-offs, maximizing transport-related benefits (i.e., time savings, emissions reductions, congestion reductions, user fares) through optimization/location-allocation methodologies e.g., [@fengMulticriteriaEvaluationAccessibilitybased2014; @fakhrmoosaviIncorporatingTravelTime2021; @zhengAreabasedEquitablePricing2020; @wismadiSpatialPreferenceModelling2014].
-These papers that focus primarily on **horizontal equity** and/or **spatial equity** seldom use exclusively qualitative methods.
+what types of methods?", ".")` We begin this section with our reflections about the way equity standards connect with conceptual and theoretical frameworks of justice.
+Broadly speaking, some trends emerge in terms of the methods used.
+**Opportunity** and **population** standards appear in the literature in similar proportions ( `r (stand_count$count[3] /sum(stand_count$count)) |> percent(accuracy=0.1)` vs.
+`r (stand_count$count[4] /sum(stand_count$count)) |> percent(accuracy=0.1)`), but frequently correspond to different conceptualizations of justice.
+Referring to @fig-fig6, much attention has been given to questions of **horizontal equity** and **spatial equity**, and over 40% of these papers propose standards that relate to **opportunities**.
+Within these papers, a frequent suggestion is to use standards that relate to travel impedance, which controls how fast and/ro how costly it is to move.
+We think of this as a type of opportunity standard because travel impedance is essential to the measurement of accessibility (i.e., the ease of reaching opportunities).
+Some examples include the work of @chenImpactHighspeedRail2017, who use a standard of 4 hours or less by high-speed rail for municipalities to be considered "comfortably connected".
+In a similar manner, @yenisettyMeasuringAccessibilityVarious2020 assumes that a travel distance of less than 1,200 m is sufficient for a resident to interact with its transit system locally.
+And @shenMeasuringSpatialAccessibility2020 identify regions where populations cannot access hospitals within 1 hour by car - in other words, more than sixty minutes travel time is already seen as an unfair burden.
+Papers that suggest opportunity standards often employ disparity analysis through a variety of quantitative approaches.
+These include inequality measures such as the Gini coefficient and Lorenz curve, or poverty measures [@vanderveenOperationalizingIndicatorSufficient2020; @tiznadoaitkenRoleAccessibilityPublic2018]).
+There are also examples of spatial descriptive analysis as well as comparisons to benchmarks [e.g., equal supply to demand of public transit in a spatial unit as in @peungnumsaiGridBasedSpatialAnalysis2020] to determine which locations are spatially and horizontally (in)equitable.
+Another branch of quantitative research conceive the externalities of transportation system as trade-offs, and aim to maximize transport-related benefits (i.e., time savings, emissions reductions, congestion reductions, user fares) through optimization/location-allocation methodologies e.g., [@fengMulticriteriaEvaluationAccessibilitybased2014; @fakhrmoosaviIncorporatingTravelTime2021; @zhengAreabasedEquitablePricing2020; @wismadiSpatialPreferenceModelling2014].
+Papers that focus primarily on **horizontal equity** and/or **spatial equity** seldom use exclusively qualitative methods.
 
 `r word_comment("Answering: what conceptualizations occur in papers with **population** standards?
-what types of methods?", ".")` Whereas, papers that propose  **population** standards frequently utilize other methods.
-Over 40% of papers that suggest **population** standards focus on **well-being** conceptualizations.
-These papers typically ask *what is enough to lead a satisfactory life (as related to transportation)*, and the standards that are suggested include population benchmarks for comparison such as: questionnaires and relative comparisons to physical activity per week recommendations [@adlakhaMindGapGender2020; @auchinclossDesignBaselineDescription2020; @mccormackRelationshipClusteranalysisDerived2012; @kimNeighborhoodEnvironmentWalkability2016; @towneUsingWalkScore2016], summative per capita benchmarks (e.g., energy consumption for a 'decent living' is suggested in [@raoDecentLivingEmissions2012], and region-relative comparisons in health-related outcomes e.g., premature mortality rates [@awuorInfluenceEnvironmentalHealth2019], spatial access to hospitals [@pereiraGeographicAccessCOVID192021], spatial access to supermarkets, active-mode-usage, and Body Mass Index (BMI) [@murphySupermarketAccessTransport2017].
-The majority of these papers use quantitative/mixed-methods to identify inequities in **wellbeing**, however, a minority do use exclusively qualitative methods to distill themes e.g.
-the exploration of *perceived* quality of life in [@berheAdaptationDissonanceQuality2014].
+what types of methods?", ".")` A different way to approach standards is to see them from the lens of **population**, and when this is done other methods are commonly used.
+Over 40% of papers that suggest **population**-based standards are founded on concepts of **well-being**.
+These papers typically ask *what is enough to lead a satisfactory life (as related to transportation)*, and the standards suggested include population benchmarks for comparison such as: questionnaires and relative comparisons to recommended physical activity per week [@adlakhaMindGapGender2020; @auchinclossDesignBaselineDescription2020; @mccormackRelationshipClusteranalysisDerived2012; @kimNeighborhoodEnvironmentWalkability2016; @towneUsingWalkScore2016].
+Summative per capita benchmarks include energy consumption for a 'decent living' [as suggested in @raoDecentLivingEmissions2012], and region-relative comparisons in health-related outcomes, such as premature mortality rates [@awuorInfluenceEnvironmentalHealth2019], spatial access to hospitals [@pereiraGeographicAccessCOVID192021], as well as spatial access to supermarkets, active-mode-usage, and Body Mass Index (BMI) [@murphySupermarketAccessTransport2017].
+The majority of these papers use quantitative/mixed-methods to identify inequities in **wellbeing**, however, a minority do use exclusively qualitative methods to distill themes, as for example in the exploration of *perceived* quality of life by @berheAdaptationDissonanceQuality2014.
 
 `r word_comment("Answering: what conceptualizations occur in papers that conceptualize BOTH population and opportunity standards?
-what types of methods?", ".")` Papers that conceptualize *both* **population** and **opportunity** standards often conceptualize **vertical equity** and **transport-related social exclusion** (note the similar proportions in these standards in @fig-fig6).
-Expectedly, these papers often include a combination of methods: Questionnaires and other qualitative methods related to population standards and quantitative methods such as accessibility indices for opportunity standards are usually deployed.
-For instance, census data and the estimated proportion of households within some travel distance/time/availability to/of key destinations is used to identify a variety of social exclusions e.g., [@mackettIncreasingAccessibilityCostEffectively2010; @chenExploringTravelCharacteristics2010; @danielsProposalAccessibilityPlanning2011; @sunPublicTransportAvailability2021; @sharmaPublicTransitAccessibility2021], transport-related social exclusion's link to wellbeing e.g., [@delboscTransportProblemsThat2011; @churchillTransportPovertySubjective2019], areas more likely to experience transport poverty [@allenSizingTransportPoverty2019a], food deserts [@mckeyCrowdsourcedMappingHealthy2020], or transport-related energy poverty [@robinsonDoubleEnergyVulnerability2020; @berryInvestigatingFuelPoverty2016; @berryDistributionalEffectsCarbon2019].
-
-Similar to papers that conceptualize **wellbeing**, the majority of **social-exclusion**-conceptualizing papers use quantitative/mixed-methods to identify areas, households, and/or populations at risk.
+what types of methods?", ".")` Papers that combine **population**- and **opportunity**-based standards often rely on **vertical equity** and **transport-related social exclusion** (note the similar proportions in these standards in @fig-fig6).
+Perhaps not surprisingly, these papers often include a combination of methods: questionnaires and other qualitative methods related to population standards and quantitative methods such as accessibility indices for opportunity standards are usually deployed.
+For instance, census data and the estimated proportion of households within some travel distance/time/availability to/of key destinations is used to identify a variety of social exclusionary situations [e.g., @mackettIncreasingAccessibilityCostEffectively2010; @chenExploringTravelCharacteristics2010; @danielsProposalAccessibilityPlanning2011; @sunPublicTransportAvailability2021; @sharmaPublicTransitAccessibility2021], the linkages between well-being and transport-related social exclusion [e.g., @delboscTransportProblemsThat2011; @churchillTransportPovertySubjective2019], areas more likely to experience transport poverty [@allenSizingTransportPoverty2019a], food deserts [@mckeyCrowdsourcedMappingHealthy2020], or transport-related energy poverty [@robinsonDoubleEnergyVulnerability2020; @berryInvestigatingFuelPoverty2016; @berryDistributionalEffectsCarbon2019].
+
+Similar to papers interested in **wellbeing**, the majority of papers concerned with **social-exclusion** use quantitative/mixed-methods to identify areas, households, and/or populations at risk.
 They use a variety of methods to identify _where_ populations at risk may be located, such as clustering methods [@mohriClusteringMethodMeasuring2021].
-The minority of papers that employs exclusively qualitative methods use surveys to understand population travel willingness/barriers e.g., [@chenExploringTravelCharacteristics2010; @mehdizadehWalkingTimeSchool2017] and interviews/focus groups related to topics of unmet activity needs [@nordbakkeTransportUnmetActivity2015].
+Few of the papers that employ exclusively qualitative methods use survey data to study the willingness to travel, or the barriers to travel [e.g., @chenExploringTravelCharacteristics2010; @mehdizadehWalkingTimeSchool2017] or interviews/focus groups related to topics of unmet activity needs [@nordbakkeTransportUnmetActivity2015].
+Travel that _did not_ take place is notoriously slippery to capture in standard surveys, and understanding whether this is a preference or the consequence of constraints is not straightforward either [e.g., @paezParticipation2012].
 
 `r word_comment("Answering: what conceptualizations occur in papers that conceptualize infrastructure standards?
-what types of methods?", ".")` The largest group of papers that suggest **Infrastructure** standards conceptualize **Rights**.
-Recall, **Infrastructure**-standards suggesting papers represents `r (stand_count$count[3]/sum(stand_count$count)) |> percent()` of all papers and **Rights-**conceptualization is represented two times more than any other conceptualization type.
-These papers often focus on populations with mobility impairments and non-car users' inequities in their ability to access the transportation systems.
-Methods vary, but focus equally on audits of existing infrastructure relative to best-practice universal design principles [@odeckEconomicAppraisalUniversal2010; @larkinsAccessibleTransportationBuilt2011; @jimenez-espadaResearchProblemUniversal2021; @perezdelhoyoParticipatoryManagementImprove2021] and qualitative methods that interview/survey users about their perceived access to transport systems e.g., [@marquezFactorsAffectingPersonal2019; @mateo-babianoBicycleSharingAsia2017; @furstBestBadPractices2013; @velhoEffectTransportAccessibility2016; @parkJourneyVisuallyImpaired2017; @limFacilitatingIndependentCommuting2021; @stjernborgAccessibilityAllPublic2019] and experiment the suitability of existing best practice standards e.g., [@daamenAssessingGapPublic2008; @velhoEffectTransportAccessibility2016; @bharathyRevisitingClearFloor2018].
-
-More broadly, papers that suggest **Infrastructure** standards are often multi-dimensional, and extend beyond **Rights** conceptualizations.
-They often suggest **Opportunity** and **Population** standards as well and often apply **Vertical**, **Horizontal** and **Spatial** equity lenses as well.
-These multi-dimensional papers can refer to established guidelines and suggest composite indices e.g., @racheleNeighbourhoodSocioeconomicTransport2017a combines transport network properties such as street connectivity, cul-de-sac length, street block length, traffic volume, public transport stops and service frequency inputs to define an indicator of transport design that is supportive of walkability and access to public transport.
-Other works include assessing the quality of infrastructure [@xuSocialInequalitiesPark2017], the severity and frequency of accidents on the system  [@benevenutoExaminingTransportNeeds2020; @appleyardTransitCorridorLivability2017], user-groups (particularly disadvantaged groups in the case of vertical equity conceptualizations) [@prasertsubpakijEvaluatingAccessibilityBangkok2012] as part of the multi-criteria indicators.
-Another branch of literature explicitly focuses on affordability or other barriers to the transport system, and suggests improvements to the infrastructure such that all groups (especially the most disadvantaged) can sufficiently interact with the system e.g., [@basuPracticalFrameworkBenchmarking2019; @songWomenWheelsGender2019; @welchEquityTransportDistribution2013], dabbling into conceptualizations of **transport-related social exclusion** e.g., [@kentPrioritizingLowstressEquitable2019] and **sufficientarian/capabilities** conceptualizations e.g., [@smithAccessibilityCapabilityMinimum2012].
+what types of methods?", ".")` Besides **opportunity**- and **population**-based standards, another way to make statements about fairness is through **infrastructure**-based standards.
+Among papers that use this lens, the most common conceptual grounding used relates to **rights**.
+**Infrastructure**-standard appear in `r (stand_count$count[3]/sum(stand_count$count)) |> percent()` of all papers in this review, and within this segment of the corpus, discussions of **rights** appear two times as often than any other conceptual approach.
+These papers often focus on populations who face mobility disabilities and non-car users, and the discussion often centers about their right to partake of transportation systems.
+The methods used by this strand of the literature are varied but more or less equally split between infrastructure and environmental audits and qualitative methods.
+Audits of existing infrastructure are sometimes used in comparisons against best-practice universal design [@odeckEconomicAppraisalUniversal2010; @larkinsAccessibleTransportationBuilt2011; @jimenez-espadaResearchProblemUniversal2021; @perezdelhoyoParticipatoryManagementImprove2021] or to understand the elements of infrastructure that correlate well with use of a mode by particular populations [e.g., walking by older adults; @moniruzzamanInvestigation2016].
+Qualitative methods are used to interview/survey users about their perceived access to transport systems [e.g., @marquezFactorsAffectingPersonal2019; @mateo-babianoBicycleSharingAsia2017; @furstBestBadPractices2013; @velhoEffectTransportAccessibility2016; @parkJourneyVisuallyImpaired2017; @limFacilitatingIndependentCommuting2021; @stjernborgAccessibilityAllPublic2019; @desjardinsGoing2021] or to assess standards under best-practice considerations [e.g., @daamenAssessingGapPublic2008; @velhoEffectTransportAccessibility2016; @bharathyRevisitingClearFloor2018].
+
+Papers that suggest **infrastructure**-based standards are also sometimes multi-dimensional, and extend beyond **rights** (to the infrastructure) to suggest **opportunity**- and **population**-based standards as well.
+These papers often apply **vertical**, **horizontal** and **spatial** equity lenses.
+Papers in this slice of the literature can refer to established guidelines and suggest composite indices.
+For instance, @racheleNeighbourhoodSocioeconomicTransport2017a combine various properties of transport networks (e.g., street connectivity, cul-de-sac length, street block length, traffic volume, public transport stops and service frequency), and use these attributes as inputs to define an indicator of transport design that supports walkability and access to public transport.
+Other works assess the quality of infrastructure [@xuSocialInequalitiesPark2017], the severity and frequency of accidents on the system  [@benevenutoExaminingTransportNeeds2020; @appleyardTransitCorridorLivability2017], user-groups (particularly disadvantaged groups in the case of vertical equity) [@prasertsubpakijEvaluatingAccessibilityBangkok2012] as part of multi-criteria indicators.
+Yet other branch of literature explicitly focuses on affordability or other barriers to the transport system, and suggests improvements to the infrastructure such that all groups (especially the most disadvantaged) can sufficiently interact with the system [e.g., @basuPracticalFrameworkBenchmarking2019; @songWomenWheelsGender2019; @welchEquityTransportDistribution2013].
+These studies are grounded on concepts of **transport-related social exclusion** [e.g., @kentPrioritizingLowstressEquitable2019] or **sufficientarian/capabilities** approaches [e.g., @smithAccessibilityCapabilityMinimum2012].
 
 `r word_comment("Answering: what conceptualizations occur in papers that conceptualize Environmental+ standards?
-what types of methods?", ".")` Though papers that conceptualize **Environmental +** standards are not common in the literature review (`r (stand_count$count[1]/sum(stand_count$count)) |> percent()` of all papers), they most frequently occur in papers focused on **Inequitable externalities**.
-These papers often use traffic-related air pollution, noise pollution, green-space, urban design elements, urban air temperature, health related outcome, and physical activity guidelines to quantify transport-related externalities.
-Methods used are almost all quantitative or mixed-methods, and the identification in inequalities is spatial clustering, the use of Gini coefficient [@fengMulticriteriaEvaluationAccessibilitybased2014], comparisons to established environmental thresholds or health guidelines [@agostfelipInclusiveModelAssessing2021; @agostfelipInclusiveModelAssessing2021; @kruizeEnvironmentalEquityRole2007; @iungmanImpactUrbanTransport2021; @apparicioCyclingOneMost2021; @khomenkoLiveableCityHealthy2020; @muellerSocioeconomicInequalitiesUrban2018], creation of composite multi-dimensional indices [@agostfelipInclusiveModelAssessing2021; @mirandaBenchmarkingSustainableUrban2012; @corazzaMethodologyEvidenceCase2020], and/or in addition to spatial analysis [@jephcoteGeospatialAnalysisNaturally2013; @carrierApplicationThreeMethods2014].
-
-See Table XX for detailed examples of papers conceptualizations and standards across each equity dimension.
-
-# Moving forward: discussion and recommendations
+what types of methods?", ".")` Our corpus of literature was more or less muted in terms of studies with **environmental**-based standards (only about `r (stand_count$count[1]/sum(stand_count$count)) |> percent()` of all papers), and the most frequent focus for the within our corpus is on **inequitable externalities**.
+This paucity in our corpus could be due to the fact that the the environmental burdens of transportation are covered more broadly in other literatures (e.g., environmental justice).
+Papers that work with environmental-based standards often use traffic-related air pollution, noise pollution, green-space, urban design elements, urban air temperature, health related outcome, and physical activity guidelines to quantify externalities.
+The methods used are almost always quantitative or mixed-methods, and inequalities are identified using spatial clustering techniques, Gini coefficient [@fengMulticriteriaEvaluationAccessibilitybased2014], comparisons to established environmental thresholds or health guidelines [@agostfelipInclusiveModelAssessing2021; @agostfelipInclusiveModelAssessing2021; @kruizeEnvironmentalEquityRole2007; @iungmanImpactUrbanTransport2021; @apparicioCyclingOneMost2021; @khomenkoLiveableCityHealthy2020; @muellerSocioeconomicInequalitiesUrban2018].
+They sometimes creatE composite multi-dimensional indices [@agostfelipInclusiveModelAssessing2021; @mirandaBenchmarkingSustainableUrban2012; @corazzaMethodologyEvidenceCase2020], occasionally supported by spatial analysis [@jephcoteGeospatialAnalysisNaturally2013; @carrierApplicationThreeMethods2014].
+
+Table XX presents detailed examples of various standards as retrieved from our review.
+
+# A critical appraisal of the lay of the land
 
@@
 
-We discuss trends in how the literature has conceptualized and used standards (the "How") to evaluate the dimensions (the "What") of transportation equity.
-A noteable point is the trends are context-specific: they concern the "When" and "Where" (differences across publication year, across case studies continent, and between urban and rural contexts) and the "Who" (differences in the focus on the population sub-groups, transport modes, destinations of interest) as discussed in the preceding sections following the four equity dimensions that structure this review (mobility/accessibility, traffic-related pollution, human-health related, and traffic safety).
-
-To broadly summarize what we found in the literature:
+Applying the analytical framework to questions of justice in transportation yields a number of insights.
+Here, we critically review some of the trends that we identified from the literature in terms of concepts and standards (the "how") and how they apply to the objects (the "what") and subjects (the "who") of justice, as well as under which situations ("when" and "where" ).
+
 <!--
@@
 
-**The "When" and "Where"**:
-
-The majority of papers (`r ((case_cont_count$count[4] + case_cont_count$count[5]+ case_cont_count$count[6])/sum(case_cont_count$count)) |> percent()`) focus on case studies in the Global North.
-Though their subject matter is varied, their spatial context is one situated in case studies in North American and Europe, and thus more often than not, in conversation with more developed and formal government transport planning apparatuses and technologies (e.g., planning for equitable high-speed rail [@monzonEfficiencySpatialEquity2013], autonomous vehicle technology [@eppenbergerOpportunitySharedAutonomous2021], public consultation processes [@reddySafeguardingMinorityCivil2010]).
-
-However, studies from the Global South, specifically South America and Africa, have key differences in focus compared to Global North: they are more focused on affordability-as-a-barrier, public spending tensions in developing new transport infrastructure compared to transport service prioritization and more intense data availability limitations.
-In this way, they do not engage with emerging mobility technologies with large capital costs.
-Informality in transport planning is more present.
-In the present day, these countries, compared to the Global North, are underdeveloped for reasons of past colonialism under Northern states, heavily reliant on primary sector exports (lower efficiency, lower national GDP) under growing global financial markets, and have more fragile democracies.
-Because of lower data availability and more extreme needs for 'sufficient' transport, analysis of transportation inequities are more focused along economic lines.
-For instance, public transit network being improved such that those that at least 60% below the poverty line have at least the same amount of accessibility (60 min travel time to destinations) to jobs, educational facilities and services as the total population using public transit [@basuPracticalFrameworkBenchmarking2019].
-
-Temporally, the Global North and Global South studies can be generally seen on different transport inequity continumums.
+## "When" and "Where" is transportation considered a sphere of justice
+
+Most papers within our corpus (`r ((case_cont_count$count[4] + case_cont_count$count[5]+ case_cont_count$count[6])/sum(case_cont_count$count)) |> percent()`) focus on case studies in the Global North.
+Though their subject matter is varied, their spatial context mainly pertains to North American and Europe, and thus more often than not, in conversation with more developed and formal government transport planning apparatuses and technologies [e.g., planning for equitable high-speed rail, as in @monzonEfficiencySpatialEquity2013; or autonomous vehicle technology, as in @eppenbergerOpportunitySharedAutonomous2021; or on the public consultation processes, as in @reddySafeguardingMinorityCivil2010]).
+
+However, studies from the Global South, specifically South America and Africa, have key differences in focus compared to the Global North.
+Three emerge from our review: 1) affordability-as-a-barrier is often what motivates them; 2) the tension between investing in new transportation infrastructure (typically roads) or prioritizing other modes; and 3) more serious data availability limitations.
+In this way, work from the Global South does not engage as immediately with emerging mobility technologies that have large capital costs.
+Informal aspects present in transport planning are a more overbearing presence.
+In the present day, countries in the Global South still struggle with the consequences of past colonialism under Northern states.
+This has left them more heavily reliant on primary sector exports (lower efficiency, lower national GDP) under growing global financial markets, and with more fragile democracies.
+Because of lower data availability and more extreme needs for 'sufficient' transport, analysis of transportation inequities are often cast along economic lines.
+An example of a standard in this context involves imrpving transit to ensure that at least 60% of those below the poverty line face a similar burden travelling to work, school, and services, as the average transit user in the region [i.e., 60 min travel time to destinations; @basuPracticalFrameworkBenchmarking2019].
+
+Temporally, the Global North and Global South studies can be generally seen as laying on different transport inequity continuums.
 Global North -> concerned with particularly disadvantaged groups and creating indicators that may be used to guide the remedy of formal processes that ensure access to societal benefits [@cuiSpatialAccessPublic2020].
@@
 
-**The "Who"**:
+## "Who" are the subjects of transportation justice
 
@@
 
-**The "What" and "How"**:
-
-Inequities that focus on accessibility/mobility -> opportunity standards are typically travel impedance standards, and affordability thresholds (e.g., more than 10% of monthly income should not be spent on transport [@rivasHowAffordableTransportation2018] to work and/or other necessary destinations).
+## "What" are the objects of transportation justice
+
+<!-- 
+NOTE: THIS SECTION DISCUSSED AT THE SAME TIME "WHAT" AND "HOW", BUT I SUGGEST TO SPLIT THE DISCUSSION IN TWO PARTS, SO THAT THE "HOW" IS ON PAR WITH OTHER ELEMENTS OF THE ANALYTICAL FRAMEWORK.
+IF POSSIBLE, IT WOULD BE GOOD TO FURTHER DEVELOP THE "HOW" SECTION, AS IT IS THE MOST IMPORTANT OF THE REPORT!
+-->
+
+Inequities that focus on accessibility/mobility -> opportunity standards are typically travel impedance standards, and affordability thresholds (e.g.,more than 10% of monthly income should not be spent on transport [@rivasHowAffordableTransportation2018] to work and/or other necessary destinations).
 Oftentimes spatial equity, disparity analysis (spatial mismatch e.g., @mulleyAreNetworkPlanning2015) are operationalized.
@@
 This is what we think is missing from the literature, and what's needed to move forward to justice.
+
+## "How" is fairness determined
+
+<!-- 
+NOTE: MOST IMPORTANT!
+-->
+
+# Moving forward: concluding remarks and calls for action
+
+> "The way to right wrongs is to turn the light of truth upon them._"  
+> — Ida B. Wells
+
+<!--
+THE FOLLOWING THREE PARAGRAPHS ARE STILL A LITTLE BIT UNDERCOOKED. PLEASE REVIEW CAREFULLY AND HARMONIZE WITH THE REST OF THE EDITS.
+
+THERE IS AN OPPORTUNITY HERE TO CIRCLE BACK TO BEZOS SUBORBITAL FLIGHT, TO CLOSE THE REPORT WITH A REMINDER OF WHY IT IS IMPORTANT TO LEARN ABOUT JUSTICE.
+-->
+
+"Justice is a constant struggle" and greater knowledge is necessary to help bend the arc of history towards justice.
+Understanding and tracking transport inequities over time and space; specifically "who", "what", "where", "how" and "why" are integral.
+Then set policies that address these inequities but stay flexible to changing demands of justice.
+
+Flexible frameworks are important.
+As mentioned, transport is derived demand – what’s everyone’s demand look like?
+Past, current, and future?
+Creating transport equity frameworks that lend themselves to changing as the struggle for Justice moves forward and is resilient to backsliding is crucial.
+Setting standards from a systems approach: both considering the _positive_ rights (a right to have access to sufficient quality of essential services, for example) and _negative_ rights (mobility of cars must be limited as not to impact air quality, health, safety, and positive rights) must be both be conceptualized.
+Systems thinking is lacking transport equity literature.
+
+We thus set out the following calls for action. Some of these need to be heed by decision- and policy-makers; others by scholars; and others by members of the public.
+
+<!--
+NOTE: IS THE TEXT BELOW A LITTLE REDUNDANT AFTER MOVING THE CALLS TO THIS SECTION?
+
+for all stakeholders transport system decision makers in defining equitable transportation standards, particularly in collecting data that can be used to highlight and track inequalities such that standards can be set.
+
+1.
+**Understand what transport inequities you are measuring**.
+Common measures used in transport systems are accessibility measures.
+What does the measure mean and does potential access relate to experienced outcomes?
+Next, what does an un equitable and equitable arrangement look like ?
+This is context specific and locality sensitive.
+Global north has different standards than global south.
+
+2.
+**Understand what institutional context we’re operating in**.
+For example, what is the structure of transport planning.
+What Does WHO guidelines make sense to use as a standard for the community's context?
+
+3. **Understand what institutional context transfer looks like**. A standard set in one context, may not transfer to another community in the same region.
+
+4.
+**And be clear with our terms – what is equity and for whom?**.
+If '15 minute city' is the policy goal, what conceptualization is driving this standard?
+Is it sufficientarism?
+Egalitariansm?
+The standard can be interpreted a multitude of ways.
+The conceptualization needs to guide the standard.
+Being explicit about the conceptualization is important.
+
+-->
 
@@
 This is discussed as a next step in planning for *justice* transportation literature in Section 6.
-
-# Concluding remarks
-
-> "The way to right wrongs is to turn the light of truth upon them._" 
-> — Ida B. Wells
-
-"Justice is a constant struggle" we need equity frameworks that support the movement for justice.
-Understanding and tracking transport inequities over time and space; specifically "who", "what", "where", "how" and "why" are integral.
-Then set policies that address these inequities but stay flexible to changing demands of justice.
-
-Flexible frameworks are important.
-As mentioned, transport is derived demand – what’s everyone’s demand look like?
-Past, current, and future?
-Creating transport equity frameworks that lend themselves to changing as the struggle for Justice moves forward and is resilient to backsliding is crucial.
-Setting standards from a systems approach: both considering the _positive_ rights (a right to have access to sufficient quality of essential services, for example) and _negative_ rights (mobility of cars must be limited as not to impact air quality, health, safety, and positive rights) must be both be conceptualized.
-Systems thinking is lacking transport equity literature.
-
-We thus set out the following calls for transport system decision makers in defining equitable transportation standards, particularly in collecting data that can be used to highlight and track inequalities such that standards can be set.
-
-1.
-**Understand what transport inequities you are measuring**.
-Common measures used in transport systems are accessibility measures.
-What does the measure mean and does potential access relate to experienced outcomes?
-Next, what does an un equitable and equitable arrangement look like ?
-This is context specific and locality sensitive.
-Global north has different standards than global south.
-
-2.
-**Understand what institutional context we’re operating in**.
-For example, what is the structure of transport planning.
-What Does WHO guidelines make sense to use as a standard for the community's context?
-
-3. **Understand what institutional context transfer looks like**. A standard set in one context, may not transfer to another community in the same region.
-
-4.
-**And be clear with our terms – what is equity and for whom?**.
-If '15 minute city' is the policy goal, what conceptualization is driving this standard?
-Is it sufficientarism?
-Egalitariansm?
-The standard can be interpreted a multitude of ways.
-The conceptualization needs to guide the standard.
-Being explicit about the conceptualization is important.
 

### 2023-10-23 Manuscript/Manuscript.qmd
Almost done revision - what is remaining to be tied up is: - "What" are the objects of transportation justice - "How" is fairness determined" - Call 5 - Concluding paragraph

@@
         axis.title.y = element_blank(),
-        legend.position = "bottom") + guides(fill = guide_legend(nrow = 1,
+        legend.position = "bottom") + guides(fill = guide_legend(nrow = 2,
                                                                  title=NULL))
@@
 
-The question of _fairness_, however, is not a simple one, and most of us would probably have been stumped to explain in a clear and precise way, beyond a disquieting intuition, just why the above picture was disturbing.
+The question of _fairness_ is not a simple one.
+Most of us would probably have been stumped to explain it in a <!-- clear and-->precise way<!--, beyond a disquieting intuition,--> just why the above picture was disturbing.
 _Justice_ is a political ideal based on the principle that individuals should be treated in a _fair_ and _equitable_ manner [@gosslingUrban2016, p.
@@
 1-2].
-The political [and contested; @vanoutriveHow2019] nature of the concept presents challenges that are only narrowly amenable to scientific inquiry.
+The political (and contested; @vanoutriveHow2019) nature of the concept presents challenges that are only narrowly amenable to scientific inquiry.
 For starters, the notion that people are "due" something depends on the values of a society, as embodied in its systems and institutions [@karnerFrom2020].
 Values, in turn, are not subject to natural laws, but rather are the result of intersubjective realities, which is to say illusions whose legitimacy derives from a collective will to believe.
-For example, justice would likely mean something very different to a person in a democratic society, than to another in a society where they owed their all—perhaps including their own time or bodily autonomy—to some collective illusion (e.g., the state, or a monarch).
+For example, justice would likely mean something very different to a person in a democratic society, than to another in a society where they owed their all<!-- —perhaps including their own time or bodily autonomy—--> to some collective illusion (e.g., the state, or a monarch).
 As well, the meaning of "justice" would likely differ in yet another society where very few owned most, and most owned very little due to a different illusion (e.g., that wealth equates merit).
-In these two hypothetical cases, elucidating the meaning of "fair" would in all certainty be beyond the dreams of most, since fair would be what the state, the monarch, or the extremely wealthy said it was.
-In contrast, in democratic societies^[Paraphrasing, democracy is the worst of collective illusions, except for all others.] individuals rights are valued above the whims of the few.
-In such a setting, the task of defining a "just" distribution of the burdens and benefits of things—from income, to roads, to space travel—quickly becomes muddled, encumbered even, by the necessity to pay attention to a multitude of voices, not all of them equally loud.
+In these two hypothetical cases, elucidating the meaning of "fair" would in all certainty be beyond the dreams of most, since fair would be what organizational structures with power (e.g., the state, the monarch, or the extremely wealthy) said it was.
+In contrast, in democratic societies^[Paraphrasing, democracy is the worst of collective illusions, except for all others.] individuals rights and a collective vision are valued above the whims of the few.
+In such a setting, the task of defining a "just" distribution of the burdens and benefits of things —from income, to roads, to space travel- quickly becomes muddled, encumbered even, by the necessity to pay attention to a multitude of voices, not all of them equally loud.
 
@@
 In a plutocratic society, those with most can (and often do argue^[See the so-called [Techno-optimist Manifesto](https://a16z.com/the-techno-optimist-manifesto/)]) that this distribution of burdens and benefits is fair since both benefits and burdens are earned.
-The fact that the likes of Bezos travel to space is proof that the likes of Bezos are due those trips.
+The fact that the likes of Bezos do travel to space is proof that the likes of Bezos are due those trips.
 In a democratic society, the members of the collective might actually agree that large rewards (e.g., space flight) must be offered to highly qualified individuals (e.g., Bezos) to entice them to take important responsibilities (e.g., founding and leading Amazon).
@@
 Instead, they are malleable, and can be affected by the existence of opportunities for social mobility [@shariffIncome2016; @artigeWhat2019], by exposure to inequality [@schroderIncome2017; @garciaChanging2023], by learned helplessness [@kimSocioeconomic2022], and even by how information about inequality is communicated to the public [@walkerPeople2021].
-It follows that, in general terms, there are at least three different manners of thinking about inequality^[@martensTransport2016, in the introduction of his landmark text, talks about explanatory and prescriptive theories of justice; despite the saying "the arc of history is long and bends toward justice", it is unlikely that a predictive theory of justice exists.]: 1) in a _positive_ (or descriptive) manner, as the current state of the distribution of benefits and burdens of things; 2) also descriptively, as the _perceptions_ about the distribution of those benefits and burdens; and 3) normatively (or prescriptively), as the desired or ideal state of the same.
+It follows that, in general terms, there are at least three different manners of thinking about inequality^[@martensTransport2016, in the introduction of his landmark text, talks about explanatory and prescriptive theories of justice; despite the saying "the arc of history is long and bends toward justice", it is unlikely that a predictive theory of justice exists.]: 1) in a _positive_ (or descriptive) manner, as the current or historic state of the distribution of benefits and burdens of things; 2) also descriptively, as the _perceptions_ about the distribution of those benefits and burdens; and 3) normatively (or prescriptively), as the desired or ideal state of the same.
 Clearly, the three may coincide (e.g, if the perceived levels of inequality matched actual inequality and also how much of it the public desired).
 However, they do not necessarily have to, and in many cases will differ from one another.
-Measuring inequality as it is, in the most objective way possible is an essential task to decide whether there is a need to develop inequality-related policy to increase fairness.
-In turn, measuring the perceptions of inequality, in the most accurate way possible, may be important to achieve sufficient public buy-in in order to enhance the chances that a given policy will succeed.
+Measuring inequality as it is and in the most objective way possible, is an essential task to decide whether there is a need to develop inequality-related policies to increase fairness.
+In turn, measuring the perceptions of inequality, in the most accurate way possible, may be important to achieve sufficient public buy-in in order to enhance the chances that given policies will succeed.
 
 Transportation systems, as a class of essential technologies that facilitate or impede social inclusion and activity participation [@churchTransport2000; @lucasTransport2001; @socialExclusion2003; @cassSocial2005; @casasSocial2007; @prestonAccessibility2007; @paezMobility2009], have increasingly come into focus from the perspective of equity.
-In response to this focus, a lively and rapidly growing literature has emerged on the topic [see _inter alia_ @martensTransport2016; @diciommoTransport2017; @guoSystematicOverviewTransportation2020; @karnerFrom2020; @vecchioTransportEquityLatin2020; @pereiraTransportationEquity2021; @vanweeEvaluating2021; @zhangLiterature2021; @desjardinsExaminingEquityAccessibility2022; @karnerAdvancesPitfallsMeasuring2023].
-A cynical rationale for this interest could be that keeping track of actual and perceived inequalities can serve _at the very least_ as a gauge of social discontent [as Chilean authorities discovered to their sorrow when unrest erupted over the increase of metro fares in 2019; @BBCNews2019; @diazpabonInequality2021].
-More optimistically, in democratic systems, there could be a genuine interest in being ready to respond to popular demands for fairness.
-Several challenges arise when approaching this endeavor, caused by the notorious complexity of transportation systems: simultaneously, they move people, goods, and information.
-As well, emerging technologies and service models can swiftly change the balance of benefits and burdens among a population [@guoSystematicOverviewTransportation2020].
-Unlike income, which is as unidimensional as it gets, the benefits and burdens of transportation systems are diffuse over space and time.
+In response to this focus, a lively and rapidly growing literature has emerged on the topic (see _inter alia_ [@martensTransport2016; @diciommoTransport2017; @guoSystematicOverviewTransportation2020; @karnerFrom2020; @vecchioTransportEquityLatin2020; @pereiraTransportationEquity2021; @vanweeEvaluating2021; @zhangLiterature2021; @desjardinsExaminingEquityAccessibility2022; @karnerAdvancesPitfallsMeasuring2023]).
+A cynical rationale for this interest could be that keeping track of objective and perceived inequalities can serve _at the very least_ as a gauge of social discontent (as Chilean authorities discovered when the increase in Santiago Metro's subway fare sparked a period of massive demonstrations and unrest in October 2019 motivated by country-wide social inequalities; [@BBCNews2019; @diazpabonInequality2021]).
+More optimistically, in democratic systems, tracking objective and perceived inequalities could be of interest for governing bodies to respond to popular demands for fairness.
+Several challenges arise when approaching this endeavor.
+The notorious complexity of transportation systems is one challenge: transport systems simultaneously move people, goods, and information, and emerging technologies and service models can swiftly change the balance of benefits and burdens among a population [@guoSystematicOverviewTransportation2020].
+The benefits and burdens of transportation systems are diffuse over space and time.
 For example, transportation systems engineered to offer higher mobility for people _somewhere_, can simultaneously cut others off from essential opportunities _elsewhere_, as @rajeEngineering2004 poignantly illustrated with examples of infrastructure in the UK.
@@
 We then describe the methods used for searching, selecting, and reviewing the relevant literature.
-This is followed by a description of the landscape surveyed, which is then appraised critically.
-We will conclude with some calls for action to improve the practice of setting and using standards of fairnes for transportation justice.
+This is followed by a description of the findings from the review, which is then appraised critically.
+We conclude with some calls for action to improve the practice of setting and using standards of fairness for transportation justice.
 
@@
 440].
-Justice is a fluid concept, because it depends on the desirability of different states of affairs, which may change between populations or over time.
+Justice is a fluid concept, because it depends on the desirability of different states of affairs, which may change between populations and over time.
 That said, it is possible to distinguish several forms of justice, including [see @jaggarPhilosophicalChallengesGlobal2009; @pereiraDistributiveJusticeEquity2017a; @karnerTransportationEquityTransportation2020]:
@@
 
-The term "equity" tends to encompass various tools to understand the distribution of benefits and burdens of things among a population.
+The term "equity" tends to encompass various tools to understand the distribution of benefits and burdens of things among a population, with a particular emphasis on those with the least advantage and equality in their outcomes.
 In the transportation domain, the term is somewhat loaded because it is perceived as stemming from the authority of the state, and meant to assist with decisions about regulating and financing transportation spending [@karnerTransportationEquityTransportation2020].
 Here, we are in agreement with @karnerTransportationEquityTransportation2020 that equity analysis should not be seen as an end in and of itself, but rather as a means to gather information about actual and perceived inequities.
-In this respect, the analytical tradition of equity, at least in transportation planning, means that the relevant models become embedded in the "political ecology of the estimated truth" [@kingModels1993]: in principle their assumptions and scope must be open and transparent, but they are also vulnerable to misuse and even abuse, as tools of subjugation.
+In this respect, the analytical tradition of equity, at least in transportation planning, means that the relevant models become embedded in the "political ecology of the estimated truth" [@kingModels1993]: in principle their assumptions and scope must be open and transparent, or else they may be more vulnerable to misuse and even abuse as tools of subjugation.
 
@@
 What kind of justice does fairness serve in each case?
-It depends on what was the reason for seeking justice in the first place.
+It depends on what was the rationale for seeking justice in the first place.
 Our reading of @martensFair2021 is that fairness is a yardstick that is best deployed _a priori_ than _a posteriori_, for doing the latter risks rationalizing the outcomes rather than driving them.
@@
 
-To quickly recap the discussion so far, justice is an end.
+To recap the discussion of definitions so far, justice is an end.
 But to understand what that end is, we must clearly define standards of fairness.
@@
 
-- **"Where?"** Traditionally related to the applicable domain or sphere of life relevant for justice.
-Conventionally this meant the in-group [e.g., members of the same nation state; @jaggarPhilosophicalChallengesGlobal2009, p.
-3], a notion since expanded (e.g., with the Universal Declaration of Human Rights) to include all members of the human species.
-In the case of justice in transportation, the question of "where?" is paramount; it might be argued that, by their very nature, transportation generates inequalities.
+- **"Where?"**: Such questions traditionally relate to the applicable domain or sphere of life relevant for justice.
+Conventionally this meant the in-group e.g., members of the same nation state [@jaggarPhilosophicalChallengesGlobal2009, p.
+3]<!--, a notion that can be expanded to include all members of the human species (e.g., with the Universal Declaration of Human Rights) and nature (e.g., such as the Ecuadorian constitution, the first ever country to grant legal rights to nature [@akchurinConstructingRightsNature2015])-->.
+In the case of justice in transportation, the question of "where?" is paramount; as it might be argued that, by their very nature, transportation generates inequalities.
 By concentrating the effects of space-time convergence (for instance, by providing access to a transit system or a highway), an inequality is automatically generated.
@@
 They are incrementally paid, for example by a distributed population in the form of taxes, or by a population with a different spatial distribution in the form of poor health.
-
-- Conventionally, the question of **"When?"** refers to the situation, that is, the social circumstances within which the demands of justice have application.
-In our case, we might as well ask about the temporal aspects of transportation systems, which include _when_ to invest in transportation [e.g., @rabelloEvaluation2015] (and as a result when to generate a spatial inequality), for how long the burdens and benefits can still be associated to a specific intervention, and whether claims to reparative justice expire over time.
-<!-- WHEN IN TIME SOMETHGING HAPPENS TOO - WHAT INVESTMENTS HAPPENED IN THE PAST THAT CAST SHADOWS-->
-
-- When we ask **"Who?"**, we inquire about which entities should be regarded as subjects of justice,<!-- OF INEQUITIES / MEASURING THE INEQUITIES OF 'WHO'.
-LIKE INCOME, THE MOBILITY TOOL IS MODIFIABLE -- THE MOBILITY TOOL IS NOT MODIFIABLE IN MANY CASES, IT IS WHO YOU ARE .
-T IS ALMOST LIKE A CHARACTERISTICS OF THE PERSON.
-PART OF THE SOCIAL CONTRACT THAT IDENTIFIES EOPEL WITH THEIR MOBILITY TOOL--> meaning those entitled to make claims of moral consideration from the perspective of justice.
+As such, the definition of the spatial boundaries in the analysis is the answer to "where?".
+
+- Conventionally, the question of **"When?"** refers to the temporal circumstances within which the demands of justice have application.
+In the cause of transportation justice, we ask about the temporal aspects of transportation systems, as examples: _when_ did the equity analysis take place and under what historical policy context, _the right time_ to invest in transportation infrastructure [e.g., @rabelloEvaluation2015] (and as a result when to generate a spatial inequality), for _how long_ the burdens and benefits can still be associated to a specific transportation intervention, or even _timelines_ of reparative justice interventions that reconcile the shadows of past transportation-related injustices.
+
+- When answering **"Who?"**, we inquire about which entities should be regarded as subjects or arbiters of justice,<!-- OF INEQUITIES / MEASURING THE INEQUITIES OF 'WHO'.
+LIKE INCOME, THE MOBILITY TOOL IS MODIFIABLE -- THE MOBILITY TOOL IS NOT MODIFIABLE IN MANY CASES, IT IS WHO YOU ARE - LIKE A CHARACTERISTICS OF THE PERSON.
+PART OF THE SOCIAL CONTRACT THAT IDENTIFIES PEOPLE WITH THEIR MOBILITY TOOL--> meaning those entitled to make claims of moral consideration from the perspective of justice.
 To make it tractable, this question is often approached through the filter of population groups, which may include several concurrent traits, such as gender identity, ableness, ethnicity, age, caste, and income.
-Often, it is appropriate to reflect on the intersections between traits, given evidence that the lived experiences of, say, a White woman and a Black women, can be markedly different between each other, in addition to being different from those of men.
-A possible complication in the case of transportation is that disentangling the "who" from the tools available to them is not always straightforward.
-Clearly, a person is not their mode of transportation; in practice, though, there are large segments of the population who live in situations where they cannot extricate themselves from the tools for mobility that they can use, either because they have driven themselves out of choices [see @laveryDriving2013], or have been driven out of choices by factors beyond their control, effectively ending up as captive users of a single mode [@jacquesRescuing2012; @cherancheryInvestigatingPerceptionCaptive2018].
+Often, it is appropriate to reflect on the intersections between traits, given evidence that the lived experiences of, say, a White woman and a Black women, can be markedly different between each other, in addition to being different from those of White men.
+A possible complication in the case of transportation is that disentangling the "who" from their mobility tools is not always straightforward.
+Clearly, a person is not their mode of transportation; however in practice, there are large segments of the population who live in situations where they cannot extricate themselves from the tools for mobility that they can use, either because they have driven themselves out of choices [see @laveryDriving2013], or have been driven out of choices by factors beyond their control (e.g., captive users of a single mode [@jacquesRescuing2012; @cherancheryInvestigatingPerceptionCaptive2018]).
 In societies that have grown into transportation monocultures with a predilection for automobility [@millerCollaborative2011] there may actually be less choice about mobility tools than we would like to assume.
@@
 In the case of transportation, in addition to members of the public who use transportation systems, there is another category of **who**, that stands possibly in opposition to users, namely the entities charged with providing services, maintaining infrastructure, and so on.
-These could be ministries or departments of transportation, transit agencies, public works departments and others.
+These could be ministries or departments of transportation, transit agencies, public works departments and others having the power to act upon transportation (in)justices.
 Identifying these entities is relevant to ellucidate who is responsible, for example for apportioning the benefits or mitigating the burdens of transportation.
 
-- **"What?"** asks which entities should be regarded as objects of justice, meaning which kinds or categories of things should be distributed in a just manner.
+- **"What?"** asks which entities should be regarded as objects of inequities, meaning which kinds or categories of things should be distributed in a just manner.
 To understand the distributional implications of transportation systems, it is essential that we are clear about what they do.
@@
 **Equity standards** are a tool for answering this distributive question: how do we allocate burdens and benefits and to whom?
-Standards are thresholds that when operationalized effectively define what is fair, that is, in/equitable.
+Standards are thresholds that when operationalized effectively define what is fair, that is, (in)equitable.
 The thresholds can be quantitative (e.g., square meters of green space per capita), or they can be qualitative descriptions (e.g., do not knowingly discriminate), or a mix of the two.
@@
 This is perhaps the most slippery of all the questions posed here.
-Justice is an inherently social construct <!--, and does not exist outside of human society-->.
-Asking **why?** amounts to asking what sort of social contracts regulate human interactions, that is, what are the rules that our collective will to believe impose on each of us.
+Justice is an inherently social construct<!--, and does not exist outside of human society-->.
+Asking **why?** amounts to asking what sort of social contracts regulate human interactions, that is, what are the rules that our collective will to believe imposes on each of us.
 These contracts can be defined by constitution, but there are often unwritten and possibly contested variants.
-To give an example drawing from the Canadian Bill of Rights, a number of rights and fundamental freedoms [including the liberty to move freely; see @chartopediaArt72023] are recognized to exist in Canada "without discrimination by reason of race, national origin, colour, religion or sex".
+To give an example drawing from the Canadian Charter of Rights and Freedoms, a number of rights and fundamental freedoms [including the liberty to move freely; see @chartopediaArt72023] are recognized to exist in Canada "without discrimination by reason of race, national origin, colour, religion or sex".
 Notice that this declaration includes several individual traits that define the subjects of justice, but does not consider age as one aspect of the person.
@@
 Crook argued that the goal in teaching his kids (then aged 7, 8, 9 and 11) to use transit to go to school was to raise self-reliant children [@stueckYoung2017].
-While both sides discovered that there no clear rules existed about kids riding transit alone [@brendHere2017], statistically there is evidence that travel by car is riskier than travel by bus [@morencyTraveling2018].
-If the public good is to reduce risk, should not children be banned from riding in cars until they are ten?
-In a functioning democracy like Canada, people have chosen to care about fundamental freedoms and rights.
-This provides an answer as to "why" we would even consider children as subjects of justice.
+While both sides discovered that no clear rules existed regarding childrens' independent transit use [@brendHere2017], statistically there is evidence that travel by car is riskier than travel by bus [@morencyTraveling2018].
+If the public good is to reduce risk, shouldn't children be banned from riding in cars?
+The Canadian legal system concerns itself with fundamental freedoms and rights, one of which is the liberty to move freely.
+This provides an answer (of many) as to "why" we would even consider children as subjects of justice.
 But the devil is in the details, and answering the other questions discussed above is essential to pin the devil by the tail.
-Are children under the age of 10 subjects of justice (i.e., are they a "who")?
-If so, is the ability to use transit unsupervised an object of justice (i.e., the "what")?
-When and where do these questions apply (e.g., to to to school, or to a social event; during the day, or late at night)?
+Are children under the age of 10 subjects of justice (i.e., are they the "who") and who is deciding if they are (i.e., the courts and the parents in this case)?
+If so, is the ability to use transit unsupervised an object of justice that should be justly distributed (i.e., the "what") and under what circumstances (e.g, to school, to a social event; during the day, or late at night?).
+"When" (e.g., for all children going forward, for only 5 years) and "where" (e.g., only in the suburbs, across the country) do these questions apply?
 Eventually, in 2020, a court of appeals ruled in Crook's favor, allowing his children to continue to ride the bus, thus providing much needed clarity with respect to the use of transit by children who desire independent mobility [@stueckCourt2020].
@@
 
-<!-- 
-IN THE PARAGRAPH ABOVE, ANASTASIA SUGGESTS DISCUSSING WHY DO WE THINK THAT OUTSIDE IS DANGEROUS, THAT CARS ARE SAFER, THAT CERTAIN PEOPLE ARE TO BE AVOIDED.
-
-THIS IS WHY WE ONLY FOCUS ON INEQUITIES. JUSTICE IS FLUID AND SLIPPERY. WHAT HAS BEEN DONE WHAT HAS NOT. 
--->
-
-The focus of this report is on standards of fairness that, combined with the use of equity analysis, can help us understand how better to move towards just transportation systems.
+Another perspective on the answer to "Why?" are cultural _norms_ in providing rationale.
+In this case, the liberty of movement for all - including children - reigns supreme.
+But reigns supreme, "Why"?
+The outside is dangerous, in part due to vehicular traffic [@morencyTraveling2018], and in part because of potential violence from others.
+Parents may have over-pronounced 'stranger danger' based on neighbourhood and socio-economic characteristics of neighbourhoods [@francisLlBeDriving2017] and children themselves may fear potential violence based on real threats of victimization [@painParanoidParentingRematerializing2006].
+Who's voice, what perception of fear, and actions we take to intervene define and reproduce cultural norms, all play into the rationale behind the question "Why?".
+In this way, analyzing the "Why?" in the reviewed literature is left out of this review as answers to "Why?" are not explicitly stated and challenging to infer implicitly.
+As such, the focus of this report is on the standards of fairness that, combined with the use of equity analysis, can help us understand how better to move towards just transportation systems and better formulate answers to "Why?".
 
@@
 The initial draft of the protocol was refined from preliminary searches of related-reviews [e.g., @iglesiasHowUnevenUrban2019; @sagarisUsingPARFrame2020; @vecchioAccessibilityCapabilitiesApproach2021], and in consultation with a University of Toronto Research Services Librarian and a Liaison Librarian in City Studies.
-
 The methods are described in two parts: (i) development of the search strategy and (ii) selection of evidence and data extraction.
@@
 
-<!-- SELECTION OF DATABASES?? -->
+<!-- SELECTION OF DATABASES?? NOTE: Antonio, the final paragraph in this subsection addresses this question -->
 
@@
 In the last step, a data extraction template for each record was filled by one reviewer (purple box in @fig-fig1).
-The data extraction template was created with the aim of striking a balance between the complexity of categories and the simplicity of summary; information related to "What?" (the object of justice), "Who?" (the subject of justice), "Where?" (the geographical context and sphere of life), and "How?" (equity standard(s) and concepts) was filled out for each study.
+The data extraction template was created with the aim of striking a balance between the complexity of categories and the simplicity of summary; information related to "Where?" (the geographical context and sphere of life), "When?" (temporal circumstances for the application of justice), "Who?" (the subject of justice), "What?" (the object of justice), and "How?" (equity standard(s) and concepts) was filled out for each study.
 @fig-A2 contains the template that was input into *Covidence* and used throughout.
@@
 
-# The lay of the land: a summary of findings  `r word_comment("NOTE: in this section, only papers that are included in the review (i.e., have a full template) are cited.", "findings")`
+# The lay of the land: a summary of findings  
+<!--"NOTE: in this section, only papers that are included in the review (i.e., have a full template) are cited."-->
 
@@
 @ampeImpactChildBike2020, on the other hand, work to identify the lateral clearance that motorists should maintain when passing cyclists with children seats.
-While both studies refer to the same objects of justice ("what", i.e., the right to the road and cycling), the subjects of justice (the "who") are different (people with disabilities and children), and the "where", that is, the geographical context for the examination of fairness, differs as well.
-
-The remainder of the studies focus on rural regions (`r (urban_form_count$count[2]/sum(urban_form_count$count)) |> percent()`).
+While both studies refer to the same objects of justice ("what", i.e., the right to the road and cycling), the subjects of inequities (the "who") are different (people with disabilities and children), and the "where", that is, the geographical context for the examination of fairness, differs as well.
+
+The remainder of the studies reviewed focus on rural regions (`r (urban_form_count$count[2]/sum(urban_form_count$count)) |> percent()`).
 To illustrate, we highlight the work of @caoIndicatorsSocioSpatialTransport2017, who examined transportation disadvantage in remote places which rely on inter-island ferry trips in the rural Philippines.
@@
 @fig-fig4 showcases the population group types that are the focus of the papers in our review.
-From this tally, papers that consider income groups is the most widely represented in the literature.
-Particularly, most papers that focus on income pay usually focus on the lowest-income groups, as they experience the least benefits in terms of mobility and accessibilty, and usually the greatest burdens as well, in terms of cost, exposure, and so on [@peungnumsaiGridBasedSpatialAnalysis2020; @zhaoUnequableSpatialAccessibility2020; @falavignaAssessingInequalitiesPublic2016].
+From this tally, papers that consider population by income groups are the most widely represented in the literature reviewed.
+Particularly, most papers that center income usually focus on the lowest-income groups, as they experience the least benefits in terms of mobility and accessibility, and usually the greatest burdens as well, in terms of cost, exposure, and so on [@peungnumsaiGridBasedSpatialAnalysis2020; @zhaoUnequableSpatialAccessibility2020; @falavignaAssessingInequalitiesPublic2016].
 
@@
 Many papers that focus on this category highlight differences in age-related capabilities; for instance, @martinez-jimenezAccessibilityCultureEducation2019 and @arranz-lopezSocialSpatialEquity2019 investigate travel distances and times to various opportunities based on specific age groups, acknowledging that age is an important consideration, and associated with differences in access to opportunities.
-The most commonly age groups are school-aged children and older populations.
+The most commonly studied age groups are school-aged children and older populations.
 Research on school-aged children analyzes their wellbeing [@laszkiewiczChildrenGreenWalk2020], exposure to green space [@corazzaMethodologyEvidenceCase2020], access to schools [@sharmaSpatialSocialInequities2022], and aims at understanding and encouraging active travel journeys [@mackieOvercomingBarriersCycling2009; @mehdizadehWalkingTimeSchool2017].
@@
 
-The third population group most commonly studied is what might be called 'composite vulnerable population groups' that exist at the intersection of several individual traits.
-These papers use some sort of composite vulnerability index that captures multiple population characteristics, including low-income, unemployment, immigrant status, family household characteristics, and so on.
+The third population group most commonly studied is what might be called 'composite vulnerable population groups': the intersection of several individual traits.
+These papers use composite vulnerability indices that captures multiple population characteristics, including low-income, unemployment, immigrant status, family household characteristics, and so on.
 These indices are typically generated from official government sources or author-informed census data creation, and produced using a variety of methodologies.
@@
 
-`r word_comment("Papers that focus ONLY on (Dis)ability, or ONLY on Gender, or ONLY on Race/Ethnicity, or ONLY on Employment/Education level...
-the only focus", ".")` Papers that exclusively focus on populations with (dis)abilities e.g., [@parkJourneyVisuallyImpaired2017; @chiscanoImprovingDesignUrban2021; @orellanaWalkRollMapping2020] are relatively common in the literature.
+<!--`r word_comment("Papers that focus ONLY on (Dis)ability, or ONLY on Gender, or ONLY on Race/Ethnicity, or ONLY on Employment/Education level... the only focus", ".")`-->
+Papers that exclusively focus on populations with (dis)abilities e.g., [@parkJourneyVisuallyImpaired2017; @chiscanoImprovingDesignUrban2021; @orellanaWalkRollMapping2020] are relatively common in the literature.
 They often assess universal design guidelines and the capabilities of people who face disabilities to travel.
@@
 That said, the existence of captive users by mode makes it challenging at times to extricate the mobility tool from the individual.
-In this section we choose to view the mode of travel primarily as an object of justice (i.e., a "what"), but occassionally this will be enjoined to the subject of justice (i.e., the "who").
-
-It is interesting from this perspective to notice that the primary emphasis within the literature reviewed centers on public transit (@fig-fig5)., a mode that often is perceived by users as less than ideal [@paezEnjoyment2010; @mellaDrivers2021] and that yet is often reckoned as the only realistic mobility tool for many [@jacquesRescuing2012; @welchMeasure2013; @cherancheryInvestigatingPerceptionCaptive2018].
+In this section, we choose to view the mode of travel primarily as an object of justice (i.e., a "what"), but occasionally this will be enjoined to the subject of justice (i.e., the "who").
+
+It is interesting to note that the primary emphasis within the literature reviewed centers on public transit (@fig-fig5), a mode that often is perceived by users as less than ideal [@paezEnjoyment2010; @mellaDrivers2021] and that yet is often reckoned as the only realistic mobility tool for many [@jacquesRescuing2012; @welchMeasure2013; @cherancheryInvestigatingPerceptionCaptive2018].
 The variety of topics assessed from the perspective of public transit are varied.
@@
 
-The literature offers a variety of conclusions and recommendations that span across the benefits of transportation (e.g., destinations types), often comingled with population groups and the tools of mobility.
+The literature offers a variety of conclusions and recommendations that span across the benefits of transportation (e.g., destinations types), often co-mingled with population groups and the tools of mobility.
 In addition, as we will discuss next, there is a variety of perspectives in terms of the philosophical foundations and the standards of fairness.
@@
 ```
-
-`r word_comment("Answering: what conceptualizations occur in papers with opportunity standards?
-what types of methods?", ".")` We begin this section with our reflections about the way equity standards connect with conceptual and theoretical frameworks of justice.
+<!--`r word_comment("Answering: what conceptualizations occur in papers with opportunity standards? what types of methods?", ".")`-->
+We begin this section with our reflections about the way equity standards connect with conceptual and theoretical frameworks of justice.
 Broadly speaking, some trends emerge in terms of the methods used.
@@
 Referring to @fig-fig6, much attention has been given to questions of **horizontal equity** and **spatial equity**, and over 40% of these papers propose standards that relate to **opportunities**.
-Within these papers, a frequent suggestion is to use standards that relate to travel impedance, which controls how fast and/ro how costly it is to move.
+Within these papers, a frequent suggestion is to use standards that relate to travel impedance, which describes how fast, far, or costly it is to travel from one place to another.
 We think of this as a type of opportunity standard because travel impedance is essential to the measurement of accessibility (i.e., the ease of reaching opportunities).
@@
 
-`r word_comment("Answering: what conceptualizations occur in papers with **population** standards?
-what types of methods?", ".")` A different way to approach standards is to see them from the lens of **population**, and when this is done other methods are commonly used.
+<!--`r word_comment("Answering: what conceptualizations occur in papers with **population** standards?
+what types of methods?", ".")`--> A different way to approach standards is to see them from the lens of **population**, and when this is done, other methods are commonly used.
 Over 40% of papers that suggest **population**-based standards are founded on concepts of **well-being**.
@@
 
-`r word_comment("Answering: what conceptualizations occur in papers that conceptualize BOTH population and opportunity standards?
-what types of methods?", ".")` Papers that combine **population**- and **opportunity**-based standards often rely on **vertical equity** and **transport-related social exclusion** (note the similar proportions in these standards in @fig-fig6).
+<!--`r word_comment("Answering: what conceptualizations occur in papers that conceptualize BOTH population and opportunity standards?
+what types of methods?", ".")`--> Papers that combine **population**- and **opportunity**-based standards often rely on **vertical equity** and **transport-related social exclusion** (note the similar proportions in these standards in @fig-fig6).
 Perhaps not surprisingly, these papers often include a combination of methods: questionnaires and other qualitative methods related to population standards and quantitative methods such as accessibility indices for opportunity standards are usually deployed.
@@
 
-`r word_comment("Answering: what conceptualizations occur in papers that conceptualize infrastructure standards?
-what types of methods?", ".")` Besides **opportunity**- and **population**-based standards, another way to make statements about fairness is through **infrastructure**-based standards.
+<!--`r word_comment("Answering: what conceptualizations occur in papers that conceptualize infrastructure standards?
+what types of methods?", ".")`--> Besides **opportunity**- and **population**-based standards, another way to make statements about fairness is through **infrastructure**-based standards.
 Among papers that use this lens, the most common conceptual grounding used relates to **rights**.
@@
 Other works assess the quality of infrastructure [@xuSocialInequalitiesPark2017], the severity and frequency of accidents on the system  [@benevenutoExaminingTransportNeeds2020; @appleyardTransitCorridorLivability2017], user-groups (particularly disadvantaged groups in the case of vertical equity) [@prasertsubpakijEvaluatingAccessibilityBangkok2012] as part of multi-criteria indicators.
-Yet other branch of literature explicitly focuses on affordability or other barriers to the transport system, and suggests improvements to the infrastructure such that all groups (especially the most disadvantaged) can sufficiently interact with the system [e.g., @basuPracticalFrameworkBenchmarking2019; @songWomenWheelsGender2019; @welchEquityTransportDistribution2013].
+Yet another branch of literature explicitly focuses on affordability or other barriers to the transport system, and suggests improvements to the infrastructure such that all groups (especially the most disadvantaged) can sufficiently interact with the system [e.g., @basuPracticalFrameworkBenchmarking2019; @songWomenWheelsGender2019; @welchEquityTransportDistribution2013].
 These studies are grounded on concepts of **transport-related social exclusion** [e.g., @kentPrioritizingLowstressEquitable2019] or **sufficientarian/capabilities** approaches [e.g., @smithAccessibilityCapabilityMinimum2012].
 
-`r word_comment("Answering: what conceptualizations occur in papers that conceptualize Environmental+ standards?
-what types of methods?", ".")` Our corpus of literature was more or less muted in terms of studies with **environmental**-based standards (only about `r (stand_count$count[1]/sum(stand_count$count)) |> percent()` of all papers), and the most frequent focus for the within our corpus is on **inequitable externalities**.
+<!--`r word_comment("Answering: what conceptualizations occur in papers that conceptualize Environmental+ standards?
+what types of methods?", ".")`--> Our corpus of literature was more or less muted in terms of studies with **environmental**-based standards (only about `r (stand_count$count[1]/sum(stand_count$count)) |> percent()` of all papers).
+Of those papers, the most frequent focus is on **inequitable externalities**.
 This paucity in our corpus could be due to the fact that the the environmental burdens of transportation are covered more broadly in other literatures (e.g., environmental justice).
@@
 The methods used are almost always quantitative or mixed-methods, and inequalities are identified using spatial clustering techniques, Gini coefficient [@fengMulticriteriaEvaluationAccessibilitybased2014], comparisons to established environmental thresholds or health guidelines [@agostfelipInclusiveModelAssessing2021; @agostfelipInclusiveModelAssessing2021; @kruizeEnvironmentalEquityRole2007; @iungmanImpactUrbanTransport2021; @apparicioCyclingOneMost2021; @khomenkoLiveableCityHealthy2020; @muellerSocioeconomicInequalitiesUrban2018].
-They sometimes creatE composite multi-dimensional indices [@agostfelipInclusiveModelAssessing2021; @mirandaBenchmarkingSustainableUrban2012; @corazzaMethodologyEvidenceCase2020], occasionally supported by spatial analysis [@jephcoteGeospatialAnalysisNaturally2013; @carrierApplicationThreeMethods2014].
-
-Table XX presents detailed examples of various standards as retrieved from our review.
+They sometimes create composite multi-dimensional indices [@agostfelipInclusiveModelAssessing2021; @mirandaBenchmarkingSustainableUrban2012; @corazzaMethodologyEvidenceCase2020], occasionally supported by spatial analysis [@jephcoteGeospatialAnalysisNaturally2013; @carrierApplicationThreeMethods2014].
+
+See the final table in the appendix that presents detailed examples of various standards as retrieved from our corpus.
 
@@
 Because of lower data availability and more extreme needs for 'sufficient' transport, analysis of transportation inequities are often cast along economic lines.
-An example of a standard in this context involves imrpving transit to ensure that at least 60% of those below the poverty line face a similar burden travelling to work, school, and services, as the average transit user in the region [i.e., 60 min travel time to destinations; @basuPracticalFrameworkBenchmarking2019].
-
-Temporally, the Global North and Global South studies can be generally seen as laying on different transport inequity continuums.
-Global North -> concerned with particularly disadvantaged groups and creating indicators that may be used to guide the remedy of formal processes that ensure access to societal benefits [@cuiSpatialAccessPublic2020].
-Global North countries often surpass international sufficiency thresholds related to transport and public health (e.g., @carrierApplicationThreeMethods2014 measure NO2 road-side air pollution and find no levels are below the WHO guideline) but advocate for the reduction in disproportional impact of air pollution on lower-income groups.
-New standards are needed that are context-specific.
-What's also missing?
-We call for: _discussions of equality_.
-The Global North inequities are rising -- the richest have the most mobility, this is missing from the discussion.
-We need new standards that reflect this.
-
-In reference to transport inequity continuum, papers on the Global South tend to operationalize international standards as guidelines.
+An example of a standard in this context involves improving transit to ensure that at least 60% of those below the poverty line face a similar burden travelling to work, school, and services, as the average transit user in the region [i.e., 60 min travel time to destinations; @basuPracticalFrameworkBenchmarking2019].
+
+Temporally, we see the literature from the Global North and Global South in our corupus as laying on different transport inequity continuums.
+Literature from the Global North is concerned with particularly disadvantaged groups and creating indicators that may be used to guide the remedy of formal processes that ensure access to societal benefits e.g., [@cuiSpatialAccessPublic2020].
+Global North countries often surpass international sufficiency thresholds related to transport and public health (e.g., @carrierApplicationThreeMethods2014 measures NO2 road-side air pollution and find no levels are below the WHO guidelines) but advocate for the reduction in disproportional impact of air pollution on lower-income groups.
+These international guidelines, while perhaps being relevant at a time of lower transport-related development, are no longer relevant for some areas in the Global North.
+Along this line, inequities in the Global North are rising - high-income countries contain the wealthiest populations that have the most mobility (and produce associated carbon externalities) [@chancelCarbonInequalityKyoto2015].
+
+In reference to transport inequity continuum, literature from the Global South tend to operationalize international standards as guidelines.
 Though not covered in this review, Global South nations' formal processes for planning are newer (relatively), more fragile, and operating under more strict financial constraints than the Global North.
@@
 Global South, along some dimensions, are lagging Global North development, but have the opportunity to plan better and not repeat Global North mistakes (e.g., car-centric development [@warrenDevelopingEquitableSustainable2015]).
-More work on this is needed.
-Additionally, what's missing?
-Similarly, _discussions of equality_.
-Inequities look different than in Global North, but systems thinking is required as global forces influence the capacity of formal transport planning processes?
+
+So what's missing from the answers to "Where?" and "When?": standards that are context-specific, particularly those that frame the disproportionate impacts of inequities on both disadvantage groups and the most advantaged.
+Inequities in the Global North and Global South are broadly different as a result of globalizing forces, but systems thinking is required to advocate for equity standards that re-balance, develop, and uphold equitable formal transport planning processes.
 
@@
 Though the literature reviewed most commonly considers low-income as a socio-demographic characteristics of populations deserving equity as either by itself or alongside other characteristics (`r prettyNum(round(df_pop$count[8]+df_pop$count[4] * 100, 2))`\% of the papers).
-Many employ accessibility methods (e.g., walking accessibility to open space [@tangWalkingAccessibilityNeighbourhood2021]) but a others use qualitative or mixed methods to gain perspectives of inequalities that are typically cross-dimensional (e.g., @milanLiftingPeripheralFortunes2017 asks Medellin residents living in TOD areas compared to non-TOD areas about the impact of TOD on their wellbeing based on income group and gender).
+Many employ accessibility methods (e.g., walking accessibility to open space [@tangWalkingAccessibilityNeighbourhood2021]) but others use qualitative or mixed-methods to gain perspectives of inequalities that are typically cross-dimensional (e.g., @milanLiftingPeripheralFortunes2017 asks Medellin (Colombia) residents living in TOD areas compared to non-TOD areas about the impact of TOD on their wellbeing based on income group and gender).
 The use of qualitative methods allows for a more nuanced reflection of residents perspectives, and different inequities emerge.
 
-From a modal perspective, transit is also the primary focus of the literature; cycling and walking the next most popular focuses.
+From a modal perspective, transit is also the primary focus of the literature; cycling and walking are the next most popular focuses.
 These modes use public-space and users can interact with the public when using them (unlike car modes, which are seldom the focus of the reviewed literature).
-Upon the rise of automobility, transit was seen as a public service and something getting in the way of automobility.
+Upon the rise of automobility, transit was seen as a public service and something hindering automobility.
 But historically, mass transit and active modes are how people get around - and automobility is in direct tension with the movement of other modes as it requires a lot of pubic land and public subsidy.
@@
 
-In the literature reviewed, destinations of interest are mostly employment and healthcare; other destinations are less studied.
-For many papers, the 'destinations' however, are the trip themselves -- the experience, the quality of the infrastrucutre, the mode-service, a bundle of trips taken a month to all locations, etc.
-
-So what's missing from the "who"?
-More nuisances perspectives that capture the "who" of inequities.
+In the reviewed corpus, destinations of interest are predominantly employment and healthcare; other destination-types are less frequently studied.
+We know other destinations-type matter too, but we can only assume due to data restrictions and/or incomplete framing of "Who?" are the subjects of justice, these destinations are underrepresented in the corpus.
+In another significant branch of the corpus however, many papers do not consider specific destinations but the trip themselves -- be it the the travel experience, the quality of the infrastructure, the mode-service, a bundle of trips taken a month to all locations, etc.
+Again, the framing of "Who?" is exceptionally important in determining if the subject of justice should be a specific destination type, composite destinations, or aspects of the transport service quality.
+
+So what's missing from the "who?" question?
+We believe, more nuisances perspectives that capture the "who" of inequities is needed.
 As shown in the variety of methods used, dimensions analysis, and conceptualizations and standards, transport inequalities are multi-dimensional.
@@
 Community-based informed understanding of inequities are needed, and tracking how they change are needed.
+Different slices of analysis are important to different groups, as such, we believe all should be studied with an explicit justice framing.
 
@@
 
-<!--
-THE FOLLOWING THREE PARAGRAPHS ARE STILL A LITTLE BIT UNDERCOOKED. PLEASE REVIEW CAREFULLY AND HARMONIZE WITH THE REST OF THE EDITS.
-
-THERE IS AN OPPORTUNITY HERE TO CIRCLE BACK TO BEZOS SUBORBITAL FLIGHT, TO CLOSE THE REPORT WITH A REMINDER OF WHY IT IS IMPORTANT TO LEARN ABOUT JUSTICE.
--->
-
-"Justice is a constant struggle" and greater knowledge is necessary to help bend the arc of history towards justice.
-Understanding and tracking transport inequities over time and space; specifically "who", "what", "where", "how" and "why" are integral.
-Then set policies that address these inequities but stay flexible to changing demands of justice.
+Measuring context-specific evidence of transport inequities is necessary to build knowledge of injustices and inspire creative interventions that indeed bend the arc of history towards justice.
+Understanding and tracking transport inequities over time and space; specifically knowledge of the "who", "what", "where", "how" and especially "why" (the rationale) are integral.
+With this knowledge, we can set policies that address these inequities while importantly staying flexible to changing demands of justice.
 
 Flexible frameworks are important.
-As mentioned, transport is derived demand – what’s everyone’s demand look like?
+As mentioned, transport is derived demand – but what’s everyone’s demand look like?
 Past, current, and future?
@@
 Setting standards from a systems approach: both considering the _positive_ rights (a right to have access to sufficient quality of essential services, for example) and _negative_ rights (mobility of cars must be limited as not to impact air quality, health, safety, and positive rights) must be both be conceptualized.
-Systems thinking is lacking transport equity literature.
-
-We thus set out the following calls for action. Some of these need to be heed by decision- and policy-makers; others by scholars; and others by members of the public.
+
+In this report, we outline a conceptual framework for structuring equity analysis that may move us closer to transport justice, summarize findings from 165 relevant transportation equity papers from the academic literature, and critically appraised these findings with the conceptual transport justice framework.
+From what we saw and what we did not see in the reviewed literature, we set out the following calls for action.
+Some of these calls need to be heed by decision- and policy-makers; others by scholars; and others by members of the public.
 
 <!--
-NOTE: IS THE TEXT BELOW A LITTLE REDUNDANT AFTER MOVING THE CALLS TO THIS SECTION?
+NOTE: IS THE TEXT BELOW A LITTLE REDUNDANT AFTER MOVING THE CALLS TO THIS SECTION? AS: Agreed.
 
@@
 
-Some conceptualizations are implicit within the literature: for example, @muellerSocioeconomicInequalitiesUrban2018 relative risk of mortality as related to transport-related air pollution should not be higher in deprived groups than the general population.
+The rationale behind the conceptualizations of standards are often implicit within the literature: for example, @muellerSocioeconomicInequalitiesUrban2018 suggests the relative risk of mortality as related to transport-related air pollution should not be higher in deprived groups than the general population.
 It is not evident from the paper what conceptualization drives this focus on human-health inequities but can be implicitly inferred to be general "well being".
-We believe, more explicit conceptualizations support inequities are needed in the literature.
+To move equity analysis outputs towards just transport futures, explicit inclusion of the rationale for justice, the "Why?", should become more common place.
 
@@
 For decision-makers setting standards, measuring inequities and moving towards setting flexible guidelines for standards is the next step.
-As such, we believe, firmer justification of standards are lacking in the literature and are needed.
-
-Relative measures are often used: context, temporal, and group-specific.
-E.g.
-a certain group should have more access in certain situations.
+As such, we believe, firmer justification of standards are lacking in the reviewed literature and are needed.
+
+<!-- Further, relative measures are often used: these include context, temporal, and group-specific measures such a certain group should have more access in certain situations.
 Context-matters, but this presents a challenge in policy planning.
@@
 More justice work is needed, e.g., Restorative justice -- repairing harm is missing from the discussion of addressing transport inequities.
+-->
 
@@
 
-On the methodological side, more mixed methods are needed in transport equity research.
+On the methodological side, we believe more mixed methods are needed in transport equity research.
 Conceptualizations and standards are usually discussed from purely qualitative or quantitative approaches, a missed opportunity to combine the strengths of both approaches, whether by deep diving into some particular experiences or perceptions through qualitative methods or tailoring more meaningful quantitative analysis after qualitative explorations.
@@
 
-Furthermore, this is plenty of disparity analysis, without engaging explicitly with equity conceptualizations.
-For example, within the mobility/accessibility dimension, metrics of accessibility (usually 15 to 60 minutes) are used to show differences among areas and groups but with scarce policy and practice implications of those results.
-Aiming at specific goals and standards tied to conceptualizations is the ideal case.
-When these analyses engage with metrics that may be tied to conceptualizations (like Gini coefficient or Theil index), they fall short in assessing if the result's good or bad e.g., [@mijaresEquityAnalysisUrban2013].
-If a Gini coefficient of 0 means that all people have the same access to public transport stops, what does it mean a 0.2, 0.3, or 0.4?
+While there is plenty of disparity analysis in the reviewed literature, frequently the standards used do not explicitly reflect practical application.
+For example, metrics of accessibility (usually measured with travel impedance cut-offs of between 15 to 60 minutes depending on the destination, population group and mode) are used to show descriptive differences among areas and groups but with scarce practice implications.
+How low of an accessibility index is too low?
+Conversely, if a sufficient accessibility is met, high accessibility results in higher inequalities and are inequalities of a certain level an issue?
+This lack of discussion makes translating results from these equity analysis into policy practice that may move conditions towards justice challenging.
+Creative methods and discussion of metrics of quantitative accessibility with intention of practical intervention should be paired with results.
+
+Other times, when analysis engages with metrics that may be tied to conceptualizations (like Gini coefficient or Theil index), they fall short in assessing if the results' good or bad e.g., [@mijaresEquityAnalysisUrban2013].
+If a Gini coefficient of 0 means that all people have the same access to public transport stops, what does it mean if the result is 0.2, 0.3, or 0.4?
 Is this good or bad news for decision-makers?
@@
 These questions usually remain unanswered despite its importance.
-These measurements can also bring some challenges and pitfalls, as recently summarized by @karnerAdvancesPitfallsMeasuring2023.
-
-## Call 3: Data justice
-
-<!-- see my comments above -->
-
-What are the sources of data and what are the motivations for some categories?
-POI databases typically include education, health and aggregated categories for leisure and community.
-Within 'community' are organizations, government services, visiting friends/family -- very broad, grandma's house is not there, social networks rarely incorporated.
-Childcare, typically daycare or facilities -- domestic work, mobilities of care, mobility interdependence, are unrepresented relative to the presence of work destinations.
-Transport systems' focus is more than just to work or as a source of economic development, though in underdeveloped regions, transport systems as a force of economic development e.g., high-speed rail [@chenImpactHighspeedRail2017; @monzonEfficiencySpatialEquity2013; @kimImpactsHighspeedRail2015].
-
-Data availability matters, especially when operationalizing emerging theories (i.e., sufficientariansm [@vanderveenOperationalizingIndicatorSufficient2020]):
-
-- Leisure destinations (e.g., green space, parks, recreation) are less studied in this context.
-- Some categories are missing all together -- mobilities of care.
-- Issue of data availability
+These measurements can also bring some challenges and pitfalls, as recently summarized by @karnerAdvancesPitfallsMeasuring2023, but are necessarily to move equity analysis into application and towards transport justice.
+
+## Call 3: Data availability as a matter of justice
+
+In the review of the literature, we were left wondering specifically - what are the motivations for the use of some destination-types and the choice to note include other destinations types?
+For instance, papers including leisure destinations (e.g., green space, parks, recreation) are infrequently studied and some categories are missing all together like 'mobilities of care'.
+
+Of course, like all literature reviews, there are limitations, but we suspect that since methods used are predominately quantitative, the reliance on commonly used point of interest (POI) databases is also high.
+These databases typically include education, health and occasionally include aggregated categories for leisure and community.
+They are quiet on types of 'community' and 'leisure' (are they a community organization, government services, a visit to a family or friend, a care-giving destination).
+Further, they may be missing categories all together like childcare, typically daycare or facilities -- domestic work, mobilities of care, and mobility interdependence.
+These missing categories are unrepresented in the reviewed relative to the presence of work and healthcare destinations.
+
+We can only track what we have collected and compiled; and we know that transport systems' focus is more than just to work or as a source of economic development (though in underdeveloped regions, transport systems as a force of economic development is pronounced based on the concentration of Global South transport-induce economic development studies e.g., high-speed rail [@chenImpactHighspeedRail2017; @monzonEfficiencySpatialEquity2013; @kimImpactsHighspeedRail2015]).
+As such, data availability matters, especially when operationalizing emerging theories (i.e., sufficientariansm [@vanderveenOperationalizingIndicatorSufficient2020]) in equity analysis.
+
+The calls for and relevant issues of data availability are not new, but they have at least three parts.
+What and who is the subject of justice?
+When/Where/How is it measured?
+and who gets the data?
+Deciding who is the subject of justice frames the data collection activity; if it's the mobility of those that do domestic work the classification of who does this work and how it changes over time/space is fundamental.
+How we classify has implications for our understanding of just who is a subject of justice, as illustrated by the history of racial classifications in the US [see @leeRacial1993].
+What methods we're using, under what spatio-temporal boundaries we conduct data collection, and who has access to the collected data (and who does not) can all impact fulsome conceptualizations of justice.
+In the case of transportation, the issue of data (collection/availability) as a matter of justice is gaining traction as digitalization of data casts a starker light on these questions [e.g., see the papers by @sourbatiSmart2021 and @behrendtData2023].
 
@@
 
-A more robust assessment of the implications of equity standards on life outcomes is still pending.
-Estimating the benefits of increased mobility or accessibility, or reducing affordability burdens and transport externalities needs to be associated with outcomes like life and neighborhood satisfaction, subjective well-being, mental and physical health, social capital, among others.
-
-## Call 5: Evaluations of interventions and policies
-
-There is a need to evaluate more equity interventions or policies.
-In our review, only 19 out of 155 studies assess specific projects with an equity lens <!-- ignacio says CHECK THIS, make sure that this is the correct number-->.
+We believe that a more robust assessment of the implications of equity standards on life outcomes is lacking in the literature and is essential for equity analysis that translate into just practice.
+While the estimation of the benefits of increased mobility or accessibility, or reducing affordability burdens and transport externalities is commonplace in the literature, these estimations need to be associated with outcomes like life and neighborhood satisfaction, subjective well-being, mental and physical health, social capital, among others.
+
+## Call 5: Towards evaluations of interventions and policies in and across communities
+<!-- not done-->
+The assessment of the effects of policies contextualize in "Who" is the subject of justice, "What" the distribution of transport-related benefits and burdens looks like, "How" these inequities are measured and benchmarked, contextualized in the "Where" across moving "When"s can allow for equity analyses to assist in formulating "Why?" rationales and ultimately towards transport justice.
+
+Through explicit connections to justice
+
+ecreative methodsinequalities, for
+
+We believe there is a need to evaluate more equity interventions or policies.
+In our review, only 19 out of 165 studies assess specific transport-related policy interventions through an equity lens <!-- double check this number-->.
 Examples include mode-shift from driving to active school travel [@mammenSchoolTravelPlanning2014], transit fare restructures [@hickeyUsingQuantitativeMethods2010] and spatial analysis of Low Traffic Neighborhoods [@aldredEquityNewActive2021].
-This is a key step towards transport justice; assessing the effects of policies on different dimensions and populations groups and evaluating if a specific context is moving towards equitable standards over time.
-
-<!--
-THESE THREE PARAGRAPHS SOUND LIKE A CONCLUSION.
-PERHAPS A CALL TO ACTION - TO BE ABLE TO MAKE AN INTERNATIONAL COMPARISION ON JUSTICE.
-WE NEED TO KNOW THE WHY BUT LITERATURE SHOULD START TO OUTLINE THAT EXPLICILITY.
--->
-
-For the purpose of this review focused on *equity*, the focus is on studies that substantively answer "Where?", "When?", "Who?", "What?", and "How?", while the answer to "Why?" may occasionally be incidental.
-
-Transportation equity literature exists within cross-cutting categories.
-
-Though critical, answers to "Why?" are not overwhelming common in the transportation equity academic literature that also address  "Where?", "When?", "Who?", "What?", and "How?".
-This is discussed as a next step in planning for *justice* transportation literature in Section 6.
+The assessment of interventions and tracking associated outcomes can be thought of as a key step towards transport justice.
+Outcomes of interventions can be compared within and between communities, and cross community comparisons can be created that may expedite the adoption of effective policies that move towards just outcomes.
+
+## Concluding remarks
+<!-- not done 
+THERE IS AN OPPORTUNITY HERE TO CIRCLE BACK TO BEZOS SUBORBITAL FLIGHT, TO CLOSE THE REPORT WITH A REMINDER OF WHY IT IS IMPORTANT TO LEARN ABOUT JUSTICE.-->
+In summary, transportation equity analysis is important.
+Must be framed and tracked to inform the "Why?" rationale.
+Standards that results should be flexible to account for the fluidity of justice demands.
+As Bezo flies, we must interrogate _how_ does he do so -- at who's expense and why is that unfair.
+The rationale for Why is fundamental to fairness and justice.
+We must be explicit, intentional and clear in our work in equity analysis and towards implementations that move towards justice.
 
@@
 
-<!-- we need to separate by qual and quant, karen asked-->
+<!-- we need to separate by qual and quant maybe?, karen asked-->
 

### 2023-10-23 Manuscript/Manuscript.qmd
Remaining: - "What" are the objects of transportation justice - "How" is fairness determined

@@
 ## "What" are the objects of transportation justice
-
 <!-- 
@@
 -->
-
 Inequities that focus on accessibility/mobility -> opportunity standards are typically travel impedance standards, and affordability thresholds (e.g.,more than 10% of monthly income should not be spent on transport [@rivasHowAffordableTransportation2018] to work and/or other necessary destinations).
@@
 
-<!-- Further, relative measures are often used: these include context, temporal, and group-specific measures such a certain group should have more access in certain situations.
-Context-matters, but this presents a challenge in policy planning.
-
-Equity is not justice.
-Some studies conflate the two.
-Justice should not be mixed up with equity.
-More justice work is needed, e.g., Restorative justice -- repairing harm is missing from the discussion of addressing transport inequities.
--->
-
 ## Call 2: Need for creative methods for systems-thinking approach to inequities
@@
 ## Call 5: Towards evaluations of interventions and policies in and across communities
-<!-- not done-->
-The assessment of the effects of policies contextualize in "Who" is the subject of justice, "What" the distribution of transport-related benefits and burdens looks like, "How" these inequities are measured and benchmarked, contextualized in the "Where" across moving "When"s can allow for equity analyses to assist in formulating "Why?" rationales and ultimately towards transport justice.
-
-Through explicit connections to justice
-
-ecreative methodsinequalities, for
-
 We believe there is a need to evaluate more equity interventions or policies.
@@
 
-## Concluding remarks
-<!-- not done 
-THERE IS AN OPPORTUNITY HERE TO CIRCLE BACK TO BEZOS SUBORBITAL FLIGHT, TO CLOSE THE REPORT WITH A REMINDER OF WHY IT IS IMPORTANT TO LEARN ABOUT JUSTICE.-->
-In summary, transportation equity analysis is important.
-Must be framed and tracked to inform the "Why?" rationale.
-Standards that results should be flexible to account for the fluidity of justice demands.
-As Bezo flies, we must interrogate _how_ does he do so -- at who's expense and why is that unfair.
-The rationale for Why is fundamental to fairness and justice.
-We must be explicit, intentional and clear in our work in equity analysis and towards implementations that move towards justice.
+## Final remarks
+<!-- HERE IS AN OPPORTUNITY HERE TO CIRCLE BACK TO BEZOS SUBORBITAL FLIGHT, TO CLOSE THE REPORT WITH A REMINDER OF WHY IT IS IMPORTANT TO LEARN ABOUT JUSTICE.-->
+To conclude, transportation equity analysis is important.
+In this report, we outline a conceptual framework for structuring equity analysis that may move us closer to transport justice, summarize findings from 165 relevant transportation equity papers from the academic literature, and critically appraised these findings with the conceptual transport justice framework.
+
+This framework can be succinctly summarized as a classification exercise that explicitly supports a "Why" rationale.
+In conducting an equity analysis, the subject of justice "Who" must contextualizes the process, "What" the distribution of transport-related benefits and burdens looks like, "How" these inequities are measured and benchmarked, surrounded by the "Where" across temporally changing "When"s all should be explicitly stated.
+In this way, the rationale for "Why" the equity analysis is important and what root cause of/for (in)justice is motivating the analysis in the first place may be creitical to ultimately mobilizing equity analyses into transport justice.
+
+From this framework and the critical apraisal of the literature, we outline five call to actions for researchers and decision-makers that should be adopted to move equity analysis closer to transport justice.
+These include:
+
+- Explicit "Why?" rationale along with supporting "Where", "When", "Who", "What: and "How" framings for equity analysis. This means specific conceptualizations and grounded standards.
+- Creative methods of equity appraisal that can be paired with systems thinking; current methods are not used to support this processes.
+- Data availability as a matter of justice: who is not being counted and who should be by whom and why? 
+- More direct links between equity standards and experienced conditions and outcomes. 
+- We shouldn't shy away from appraise of policies. Work towards analysis that can be internationally compared.
+
+As mentioned in the introductory paragraphs, defining _fairness_ is complex, but complexity is no excuse to shirk the task.
+A founding rationale for the root "Why?" of injustice is needed to inform equity analysis and set standards: why is it that Bezo (and the likes) ability to fly to space while others are firmly grounded on earth on penny-pinched public transport systems to Bezo-connected job opportunities unfair?
+What aspects?
+To whom and in what ways?
+The rationale for "Why" is fundamental to fairness and justice.
+We must be explicit, intentional and clear in our work in equity analysis such that standards can be created, equity analysis be tracked, and used to inform policy intervention that move towards justice.
 

### 2023-10-24 Manuscript/Manuscript.qmd
The outstanding sub-sections are complete (they may need additional references, but read well enough).

@@
 The NEI is a composite index that was developed by the city to capture differences in the City's neighbourhoods by ranking them based on socio-economic characteristics (e.g., social assistance, unemployment, income) and physical environmental characteristics such as green space availability.
-Other works use national census indicators such as the social and housing deprivation index [e.g., @pucciInequalitiesJobrelatedAccessibility2019 or explore transport disadvantage, equity in policy implementation, or transport-related mortality burden by means of census measures (e.g., household poverty) and transport-related indicators [e.g., accessibility; @@aldredEquityNewActive2021; @iungmanImpactUrbanTransport2021; @sunPublicTransportAvailability2021; @scheurerSpatialAccessibilityPublic2017].
+Other works use national census indicators such as the social and housing deprivation index [e.g., @pucciInequalitiesJobrelatedAccessibility2019 or explore transport disadvantage, equity in policy implementation, or transport-related mortality burden by means of census measures (e.g., household poverty) and transport-related indicators [e.g., accessibility; @aldredEquityNewActive2021; @iungmanImpactUrbanTransport2021; @sunPublicTransportAvailability2021; @scheurerSpatialAccessibilityPublic2017].
 Similarly, Environmental Justice (EJ) indicators have been used in the US literature to identify neighbourhoods that have a higher than average proportion of low-income and non-white populations (i.e., a composite vulnerable population group').
@@
 IF POSSIBLE, IT WOULD BE GOOD TO FURTHER DEVELOP THE "HOW" SECTION, AS IT IS THE MOST IMPORTANT OF THE REPORT!
+
+Here, I've decided to talk in more depth about the equity conceptualization groups (rights, vertical equity, horizontal equity, ineq.
+externalities, TR exclusion, wellbeing, spatial equity, utilitarianism) across each equity dimension (access/mobility, envionrmental externality, human health, and traffic-safety).
+
+NOTE: I think we should remove 'inequitable externality" from the conceptualizations -- fold it into horizontal and/or vertical equity. Would need to update Fig 6.
 -->
-Inequities that focus on accessibility/mobility -> opportunity standards are typically travel impedance standards, and affordability thresholds (e.g.,more than 10% of monthly income should not be spent on transport [@rivasHowAffordableTransportation2018] to work and/or other necessary destinations).
-Oftentimes spatial equity, disparity analysis (spatial mismatch e.g., @mulleyAreNetworkPlanning2015) are operationalized.
-Infrastructure standards are also suggested -> often times access to destinations is conceptualized as a "right" as such being able to enter the transportation network and physically use it the pre-requisite that accessibility measures do not capture (e.g., (dis)abilities focused).
-
-Inequities that focus on transport-related environmental externalities -> pollution/noise standards are suggested, often using WHO guidelines or local neighbourhood averages as a comparison.
-
-Inequities that focus on human-health  -> health standards, the reliance on population transport health literature such as physical activity guideline recommendation per week or excess mortality burden.
-
-Inequities that focus on traffic-safety -> the use of engineering standards and best practice
-
-Of course, some papers focus on cross-dimensional transport inequity dimensions.
-Measures such as urban livability, etc.
-a more holistic conceptualization of transport equity.
-This is what we think is missing from the literature, and what's needed to move forward to justice.
+Defining "What" are the things measured and the inequities that characterise their distribution is fundamental to tracking the progress towards transport justice.
+In the corpus reviewed, we characterize the 'thing' of inequities (the "What") in the following dimensions: 1) a focus on accessibility and mobility, 2) transport-related environmental externalities, 3) centering human-health as related to transport systems, 4) human safety related to vehicular-traffic, and 5) cross-dimensional classifications.
+
+Depending on the characterisation of "What", the thing of inequities and the object of justice, the focus varies.
+The literature offers a variety of conclusions and recommendations that span across the benefits of transportation (e.g., destinations types) that are often co-mingled with population groups and the tools of mobility.
+For instance, in a slice of the literature centered on the distributions of accessibility/mobility, some literature centers their equity analysis on populations with disabilities (the "Who") and their limitations in accessing transportation infrastructure (their mobility tools) e.g., [@jimenez-espadaResearchProblemUniversal2021; @daamenAssessingGapPublic2008; @bharathyRevisitingClearFloor2018]; the "What?" are these inequities of accessing existing transportation infrastructure (relative to other populations) and can be extended to describe the unequal application of _Rights_ in a legal sense and/or beyond as the _right_ to participate in the production of urban space [@lefebvreDroitVille1967].
+Inequities in accessing infrastructure, the "What", can be seen as supported by a rationale for justice (the "Why").
+In a lot of the reviewed corpus, the rationale often stopped short in advocating for the _right_ to the city, and instead frames the rational as a _legal or envisioned legal right_ (e.g., ADA regulation [@bharathyRevisitingClearFloor2018] or the goal of "access for all" in a land-use transportation master plan [@limFacilitatingIndependentCommuting2021]).
+
+Another slice of the objects (the "What") of justice in the corpus see their distributions assessed through conceptualizations of _horizontal_ and/or _spatial_ equity.
+These papers typically apply quantitative methods and see _inequity_ in the distribution as the issue at hand; they are quiet in their rationales of justice.
+As mentioned in the summary of findings, examples often include literature that has selected some sort of travel impedance threshold (e.g, populations who cannot access hospitals within 1 hour by car [@shenMeasuringSpatialAccessibility2020]) and discusses the distribution of the calculated accessibility index spatially across the population e.g., [@monzonEfficiencySpatialEquity2013] or population groups (the "Who") e.g., [@sharmaPublicTransitAccessibility2021].
+These papers may also conceptualize traffic-related air pollution, noise pollution, urban air temperature, among other transport-related externalities.
+Inequities in this "What" can be theorized to be equal if similar levels are attainable by all populations (horizontal equity) or all spatial areas (spatial equity).
+<!-- maybe change this theorization to 'equality' not horizontal/spatial equity?--> However, discussion of both minimums and maximums (those contributing the most to inequities) is seldom explicitly discussed through the lens of justice in the reviewed corpus.
+
+<!-- maybe change the "What" to be an object of INEQUITIES? not just justice? -->
+Another branch of papers theorize the object of justice to be population _well-being_ and they typically ask _what is enough to lead a satisfactory life (as related to transportation)_.
+These papers are typically quantitative or mixed-methods, and typically more explicitly identify the object of (in)justice, relative to papers concerned with _spatial_ and _horizontal_ inequities.
+For instance, as previously discussed, papers using minimum physical activity guidelines and questionnaires that seek to understand population's activity related to active transportation infrastructure.
+We can assume the use of mixed/qualitative methods in a study's design as well as the use of standards linked to health outcomes (e.g., minimum travel times to get emergency treatment, premature mortality) leads to papers with a firmer identification of objects of justice.
+
+Finally, there is another branch of research that is predominately quantitative and incorporates qualitative results or uses mixed-methods.
+These papers conceptualize the object of justice to be _transport related social exclusion_,  _vertical equity_, and/or _sufficientarian/capabilities_.
+What is characteristic of these papers is the identification of standards that are linked to often linked to outcomes like papers that conceptualize _well being_, however outcomes are often more tangible or welfare informed.
+As discussed, disadvantaged groups are the focus of analysis and the linkages between well-being and transport-related social exclusion [e.g., @delboscTransportProblemsThat2011; @churchillTransportPovertySubjective2019], areas more likely to experience transport poverty [@allenSizingTransportPoverty2019a], food deserts [@mckeyCrowdsourcedMappingHealthy2020], or transport-related energy poverty [@robinsonDoubleEnergyVulnerability2020; @berryInvestigatingFuelPoverty2016; @berryDistributionalEffectsCarbon2019] are central to the object of justice.
+
+<!-- In what's described, what's missing from equity analysis in their conceptualization of the "What?" is more explicit rationale for the object of justice.
+A strong conceptualization of the "What" can make theorizing the "Why" rationale of justice easier.
+-->
 
 ## "How" is fairness determined
-
 <!-- 
 NOTE: MOST IMPORTANT!
+
+I need to add in more references from the corpus here... ran out of time.
 -->
+How we define fairness should be fundamental to equity analysis.
+In the reviewed literature, we summarize the philosophical perspectives and "how" (by what methods and measures) they are operationalized into standards that we group as: 1) opportunity standards, 2) infrastructure standards, 3) environmental+ standards, and 4) population standards.
+
+Papers that put forward opportunity standards typically use methods related to travel impedance (e.g., more than 10% of monthly income should not be spent on transport [@rivasHowAffordableTransportation2018] to work and/or other necessary destinations).
+Alternatively, they operationalize spatial mismatch (disparity analysis) [e.g., @mulleyAreNetworkPlanning2015] and identify areas of _relative_ (regionally) inequities that should be addressed base on impedance thresholds to certain destination types.
+Others apply these same types of methods but tend to focus of human health, traffic-related safety, or cross-dimensional "What"s.
+Methods can include inequality measures, poverty measures, and optimization methods to create 'optimal' landscapes, and once again vary depending on "Who" and "What" is centered.
+
+Infrastructure standards are also suggested and used implicitly or explicitly as standards for fairness.
+Often times, access to destinations is conceptualized as a _Right_ or a _right_, as such being able to enter the transportation network and physically use it the pre-requisite that accessibility measures do not capture (e.g., (dis)abilities focused).
+But reflected in the selection of a wide variety of approaches, the way the equity analysis is approached (who is centered and what is counted) informs _how_ fairness is assessed; be it an established engineering standard, a critic of an engineering standard, or a community-informed infrastructure need.
+
+Papers that propose transport-related environmental externalities standards are typically informed by international or regional health- or environmental- standards.
+A commonly used guidance document are WHO standards, be it traffic-related pollution, or noise standards and SGS goals on urban sustainability.
+The advantage of the use of these international standards is that they can facilitate cross community and temporal comparison to track progress towards equity targets; though they have limitations, as they have been critiques for being too general for practical application.
+Overall, the ability to measure externalities tangibly (i.e., the road-side noise pollution produced by vehicles, the reduction of emissions associated with public transit modal shift) create interpretable numbers that can be measured and compared against to track progress towards these goals.
+These tangible metrics may be easier to interpret and compare across communities than using accessibility/mobility measures (being their methods are abundant, and contain embedded differing assumptions depending on the application).
+
+In terms of population standards, papers that operationalize them in their analysis typically focus on health-informed standards (e.g., physical activity guidelines [@timperioSafetyNumbersDoes2015]) or identify a significant population-based threshold (e.g., an EJ community as defined by U.S.
+EJ analysis [@rowangouldIdentifyingEnvironmentalJustice2016]).
+
+We believe that all forms of these standards, when linked to tangible outcomes can be used to track "fairness" in a region and can be used to track multi-dimensional progress towards justice depending on how robustly these standards are justified.
+Justification is exceptionally important, and we believe that a lot of the literature reviewed may miss this mark.
+However, literature that does not is, more often than not, connects to qualitative results and/or operationalizes mixed-methods.
 
@@
 
-<!--
-NOTE: IS THE TEXT BELOW A LITTLE REDUNDANT AFTER MOVING THE CALLS TO THIS SECTION? AS: Agreed.
-
-for all stakeholders transport system decision makers in defining equitable transportation standards, particularly in collecting data that can be used to highlight and track inequalities such that standards can be set.
-
-1.
-**Understand what transport inequities you are measuring**.
-Common measures used in transport systems are accessibility measures.
-What does the measure mean and does potential access relate to experienced outcomes?
-Next, what does an un equitable and equitable arrangement look like ?
-This is context specific and locality sensitive.
-Global north has different standards than global south.
-
-2.
-**Understand what institutional context we’re operating in**.
-For example, what is the structure of transport planning.
-What Does WHO guidelines make sense to use as a standard for the community's context?
-
-3. **Understand what institutional context transfer looks like**. A standard set in one context, may not transfer to another community in the same region.
-
-4.
-**And be clear with our terms – what is equity and for whom?**.
-If '15 minute city' is the policy goal, what conceptualization is driving this standard?
+## Call 1: The need for explicit conceptualizations and grounded standards
+<!--*And be clear with our terms – what is equity and for whom?**-->
+The rationale behind the conceptualizations of standards are often implicit within the literature: for example, @muellerSocioeconomicInequalitiesUrban2018 suggests the relative risk of mortality as related to transport-related air pollution should not be higher in deprived groups than the general population.
+It is not evident from the paper what conceptualization drives this focus on human-health inequities but can be implicitly inferred to be general "well being".
+To move equity analysis outputs towards just transport futures, explicit inclusion of the rationale for justice, the "Why?", should become more common place.
+
+Some standards are also seemingly arbitrary.
+For example, @caoIndicatorsSocioSpatialTransport2017 proposes 20 ferries per day to avoid social exclusion for inter-island transport planning in the Philippines in their analysis, but admits that a standard should be politically determined.
+It is unclear if, for instance, 10 or 30 ferries would make a difference in a specific quality-of-life outcome or if that number is tied to funding/resource constraints.
+As another example, if '15 minute city' is the policy goal, what conceptualization is driving this standard?
 Is it sufficientarism?
@@
 The standard can be interpreted a multitude of ways.
-The conceptualization needs to guide the standard.
-Being explicit about the conceptualization is important.
-
--->
-
-## Call 1: The need for explicit conceptualizations and grounded standards
-
-The rationale behind the conceptualizations of standards are often implicit within the literature: for example, @muellerSocioeconomicInequalitiesUrban2018 suggests the relative risk of mortality as related to transport-related air pollution should not be higher in deprived groups than the general population.
-It is not evident from the paper what conceptualization drives this focus on human-health inequities but can be implicitly inferred to be general "well being".
-To move equity analysis outputs towards just transport futures, explicit inclusion of the rationale for justice, the "Why?", should become more common place.
-
-Some standards are also seemingly arbitrary.
-For example, @caoIndicatorsSocioSpatialTransport2017 proposes 20 ferries per day to avoid social exclusion for inter-island transport planning in the Philippines in their analysis, but admits that a standard should be politically determined.
-It is unclear if, for instance, 10 or 30 ferries would make a difference in a specific quality-of-life outcome or if that number is tied to funding/resource constraints.
+The framing used to conceptualize justice must guide the standard.
+Being explicit about justifications for the framing is important.
+
 If literature is to recommend standards, researchers need standards linked to grounded outcomes.
 For decision-makers setting standards, measuring inequities and moving towards setting flexible guidelines for standards is the next step.
-As such, we believe, firmer justification of standards are lacking in the reviewed literature and are needed.
+As such, we believe, firmer justification of standards are lacking in the reviewed literature but necessary.
 
 ## Call 2: Need for creative methods for systems-thinking approach to inequities
-
+<!--**Understand what transport inequities you are measuring**-->
 On the methodological side, we believe more mixed methods are needed in transport equity research.
@@
 This lack of discussion makes translating results from these equity analysis into policy practice that may move conditions towards justice challenging.
-Creative methods and discussion of metrics of quantitative accessibility with intention of practical intervention should be paired with results.
+Creative methods and discussion of metrics of quantitative accessibility with intention of practical intervention should be paired with results; these methods should yield interpretable results and the explicit discussion of minimums _and_ maximums values in the distribution of the object of justice (the "What"), as applicable, is critical.
 
@@
 ## Call 4: More direct and explicit links between standards and experienced outcomes 
-
+<!--*Understand what institutional context we’re operating in**-->
 We believe that a more robust assessment of the implications of equity standards on life outcomes is lacking in the literature and is essential for equity analysis that translate into just practice.
 While the estimation of the benefits of increased mobility or accessibility, or reducing affordability burdens and transport externalities is commonplace in the literature, these estimations need to be associated with outcomes like life and neighborhood satisfaction, subjective well-being, mental and physical health, social capital, among others.
+Further, the use of standards that are tangibly linked to outcomes can more concretely define the objects of justice (the "What").
+They may also be used to track progress towards goals across time and space.
 
 ## Call 5: Towards evaluations of interventions and policies in and across communities
+<!--**Understand what institutional context transfer looks like**-->
 We believe there is a need to evaluate more equity interventions or policies.
@@
 ## Final remarks
-<!-- HERE IS AN OPPORTUNITY HERE TO CIRCLE BACK TO BEZOS SUBORBITAL FLIGHT, TO CLOSE THE REPORT WITH A REMINDER OF WHY IT IS IMPORTANT TO LEARN ABOUT JUSTICE.-->
-To conclude, transportation equity analysis is important.
-In this report, we outline a conceptual framework for structuring equity analysis that may move us closer to transport justice, summarize findings from 165 relevant transportation equity papers from the academic literature, and critically appraised these findings with the conceptual transport justice framework.
-
-This framework can be succinctly summarized as a classification exercise that explicitly supports a "Why" rationale.
-In conducting an equity analysis, the subject of justice "Who" must contextualizes the process, "What" the distribution of transport-related benefits and burdens looks like, "How" these inequities are measured and benchmarked, surrounded by the "Where" across temporally changing "When"s all should be explicitly stated.
-In this way, the rationale for "Why" the equity analysis is important and what root cause of/for (in)justice is motivating the analysis in the first place may be creitical to ultimately mobilizing equity analyses into transport justice.
-
-From this framework and the critical apraisal of the literature, we outline five call to actions for researchers and decision-makers that should be adopted to move equity analysis closer to transport justice.
+<!-- HERE IS AN OPPORTUNITY HERE TO CIRCLE BACK TO BEZOS SUBORBITAL FLIGHT, TO CLOSE THE REPORT WITH A REMINDER OF WHY IT IS IMPORTANT TO LEARN ABOUT JUSTICE.
+
+And this is my attempt - more time is needed but this first attempt is okay! -->
+To conclude, the analysis of transportation equity is vitally important for planning: as transportation systems and technologies converge space-time, they by definition leave certain populations, depending on the area and time scale, more or less burdened or benefited.
+The analysis and tracking of equity implications is fundamental in formulating standards that are flexible to the fluid demands of transport justice.
+In this spirit, this report outlines a conceptual framework for structuring equity analysis that may move us closer to transport justice, summarize findings from 165 relevant transportation equity papers from the academic literature, and critically appraised these findings with the conceptual transport justice framework.
+
+The transportation justice framework can be seen as a flexible classification exercise that is suppored by a "Why" rationale.
+In conducting an equity analysis, the subject of justice (the "Who?") must contextualize the process, the distribution of the object of justice (the "what?") must be measured, be it transport-related benefits, burdens or both, "How?" these inequities are measured and benchmarked must be described, and these questions must be contextualized by the "Where?" across temporally changing "When?"s.
+The explicit acknowledge and answers to these questions clarify the boundaries of the equity analysis.
+In this way, the rationale for "Why?" the equity analysis is important and what root cause of/for (in)justice is motivating the analysis in the first place may be critical to ultimately mobilizing equity analyses into experienced transport justice.
+
+From this framework and the critical appraisal of the literature, we outline five call to actions for researchers and decision-makers that should be adopted to move equity analysis closer to transport justice.
 These include:
@@
 - Data availability as a matter of justice: who is not being counted and who should be by whom and why? 
-- More direct links between equity standards and experienced conditions and outcomes. 
-- We shouldn't shy away from appraise of policies. Work towards analysis that can be internationally compared.
-
-As mentioned in the introductory paragraphs, defining _fairness_ is complex, but complexity is no excuse to shirk the task.
-A founding rationale for the root "Why?" of injustice is needed to inform equity analysis and set standards: why is it that Bezo (and the likes) ability to fly to space while others are firmly grounded on earth on penny-pinched public transport systems to Bezo-connected job opportunities unfair?
-What aspects?
+- More explicit links between equity standards and experienced conditions and outcomes. 
+- Encouragement to continue work towards analysis and comparison of applied interventions, across communities and internationally.
+
+As mentioned in the introductory paragraphs, defining _fairness_ is complex, but complexity is no excuse to shirk away from the task.
+A founding rationale for the root "Why?" of in/justice is needed to inform equity analysis and set standards: why is it that Bezo (and the likes) ability to fly to space compared to the majority of others who are firmly grounded on earth on penny-pinched public transport systems traveling to Bezo-connected job opportunities felt as _unfair_?
+What aspects are _unfair_?
 To whom and in what ways?
-The rationale for "Why" is fundamental to fairness and justice.
-We must be explicit, intentional and clear in our work in equity analysis such that standards can be created, equity analysis be tracked, and used to inform policy intervention that move towards justice.
+The answer to "Why?" is fundamental in pinning down fairness and justice.
+We must be explicit, intentional and clear in our work in transport equity analysis such that standards can be created, equity analysis be tracked and used to inform policy intervention that move towards justice.
 

### 2023-10-24 Manuscript/Manuscript.qmd
With the template "executive summary", "reading guide" and "glossary" sections.

@@
 ```
+# Executive Summary
+
+_If we plan to write an executive summary, it should ease readers into the report. The full report is long so this may be a useful 2-3 page summary addition._
+
+# Reading guide
+
+_This portion may be helpful to guide readers to desired sections, if they are not interested in reading the full report. For instance:_
+
+_- If you're interested in why we wrote this report read the "Introduction."_
+_- Curious about the conceptual framework for transport justice? Read the framework in "A framework to analyze questions of justice"_
+_- Want to know about our calls for action to move transport equity analysis and standard-setting substantively towards questions of transport justice?
+Read the "Moving forward: concluding remarks and calls for action".
+This section is geared to both researchers and decision-makers._
+
+# Glossary
+_This section may be helpful for readers to clarify our definitions of these terms so they can more easily follow the text and as a quick reference:_
+
+- Fair: 
+- Justice: 
+- Equity: 
+- Equality:
+- Dimensions:
+- Conceptualizations:
+
+Etc.,
 

### 2023-10-24 Manuscript/Manuscript.qmd
Edits before sharing with M. Palm and S. Farber

@@
 ---
-title: "Equity conceptualizations and standards within transportation literature: a scoping review"
+title: "Searching for standards of fairness in the transportation justice literature: A scoping review"
 bibliography: All_papers_in_extracted.bib
@@
 ```
+
+We need a super nice cover.
+
+\newpage
+
 # Executive Summary
 
+**NOTE:** After brainstorming, Anastasia, Ignacio, and Antonio suggest that an executive summary could be a good strategy for readers who are too busy to read, and to provide a good idea of what readers who do plan to read will find in this document.
+
 _If we plan to write an executive summary, it should ease readers into the report. The full report is long so this may be a useful 2-3 page summary addition._
 
+\newpage
+
 # Reading guide
 
-_This portion may be helpful to guide readers to desired sections, if they are not interested in reading the full report. For instance:_
+_This portion may be helpful to guide readers to desired sections, if they are pressed for time or interested in only particular parts of the report. For instance:_
 
@@
 
+\newpage
+
 # Glossary
@@
 
-- Fair: 
+- Accessibility:  
+- Capabilities approach:  
+- Equality:  
+- Equity:  
+- Fairness:  
 - Justice:  
-- Equity: 
-- Equality:
-- Dimensions:
-- Conceptualizations:
+- Mobility:  
+- Object of justice:  
+- Policy:  
+- Social exclusion:  
+- Space-time convergence:  
+- Subject of justice:   
+- Sufficientarianism:  
+- Theories of justice:  
+- Utilitarianism:
 
@@
 
-# Introduction
+\newpage
+
+# Justice in transportation?
 
@@
 The question of _fairness_ is not a simple one.
-Most of us would probably have been stumped to explain it in a <!-- clear and-->precise way<!--, beyond a disquieting intuition,--> just why the above picture was disturbing.
+Most of us would probably have been stumped to explain it in a precise way just why the above picture was disturbing.
 _Justice_ is a political ideal based on the principle that individuals should be treated in a _fair_ and _equitable_ manner [@gosslingUrban2016, p.
@@
 Values, in turn, are not subject to natural laws, but rather are the result of intersubjective realities, which is to say illusions whose legitimacy derives from a collective will to believe.
-For example, justice would likely mean something very different to a person in a democratic society, than to another in a society where they owed their all<!-- —perhaps including their own time or bodily autonomy—--> to some collective illusion (e.g., the state, or a monarch).
+For example, justice would likely mean something very different to a person in a democratic society, than to another in a society where they owed their all to some collective illusion (e.g., the state, or a monarch).
 As well, the meaning of "justice" would likely differ in yet another society where very few owned most, and most owned very little due to a different illusion (e.g., that wealth equates merit).
@@
 
-To illustrate this gap, we note how @beilerExploring2016, in their exploration of transport equity, cite the following strategies identified by the US DOT to address justice [p. 287]:
+To illustrate this gap, we note how @beilerExploring2016, in their exploration of transport equity, cite the following strategies identified by the US DOT to address matters of justice [p. 287]:
 
@@
 How much pollution is allowed to be generated, and where, depends on who is affected, and how much health is valued overall, as well as by whom.
-For example, firms may or may not adopt lower standards for the emission of pollutants in poorer areas; it might be that poor people end up being relegated to areas that already had lower standards [@gouldsonFirms2006].
+For example, firms may or may not adopt lower standards for the emission of pollutants in poorer areas; it might be that poor people end up being relegated to areas that already had lower emission standards [@gouldsonFirms2006].
 Regardless of the cause, the result is the same: pollution tends to be worse were poorer people are [@delucaCode2012].
@@
 This is perhaps the most slippery of all the questions posed here.
-Justice is an inherently social construct<!--, and does not exist outside of human society-->.
-Asking **why?** amounts to asking what sort of social contracts regulate human interactions, that is, what are the rules that our collective will to believe imposes on each of us.
+Justice is an inherently social construct.
+Asking **why?** amounts to asking what sort of social contracts regulate human interactions, or in other words, what are the rules that our collective will to believe imposes on each of us.
 These contracts can be defined by constitution, but there are often unwritten and possibly contested variants.
@@
 
-<!--
-NOTE: This text was in the Definitions, but we decided that these definitions were a thin conceptual layer that could be removed.
-Instead of "equity dimensions", "conceptualization", and "standards" we will stick to using the where, when, who, what, how, and why conceptual framework.
-For example, the EQUITY DIMENSIONS are simply the "what" or "objects of justice".
-The "conceptualizations" are the "how".
-And so on.
-
-For all literature reviewed in the transportation domain, equity *dimensions*, *conceptualizations*, and *standards* are discussed through the "Where?", "When?", "Who?", "What?", and "How?" framings.
-As follows, we define each of these categories and how they inter-connect.
-
-**Equity dimensions**: the assessment of transportation equity encompasses various dimensions, mirroring the multifaceted nature of transportation systems.
-These dimensions can be distinguished as the **"What?"**, specifically: (1) mobility/accessibility, (2) traffic-related pollution, (3) traffic safety, and (4) human-health as provided in a conceptual framework by @lucasMeasuringTransportEquity2019.
-
-All papers reviewed include dimensions that cover the **"What?"** of the transport equity issue and can be further disaggregated:
-
--   *When* and *Where*: includes the article's publication year, the geographic context (i.e., urban, suburban and rural areas, countries) and location.
-
--   *Who*: includes the type of population group, mode type, activity/opportunity type, and transport network status.
-
-papers that include equity **dimensions**, **conceptualizations**, **standards** and substantive answers to the "When?", "Where?", and "Who?" in relation to equity in transportation systems (i.e., the users, the infrastructure, the service, the outcomes, or a combination of all) are used to answer the following research questions:
-
--   What are the equity dimensions, equity standards, and equity conceptualizations and how are they applied in the existing transportation equity academic literature?
-
--   Based on how equity has been applied in the literature, what may be paths forward in theorizing justice in the context of transportation systems
--->
-
 This review examines the breadth and depth of the academic literature on transportation to identify the extent to which standards for equity are defined and employed.
@@
 ## Search strategy
-
-<!-- SELECTION OF DATABASES?? NOTE: Antonio, the final paragraph in this subsection addresses this question -->
 
@@
 # The lay of the land: a summary of findings
+
 <!--"NOTE: in this section, only papers that are included in the review (i.e., have a full template) are cited."-->
@@
 The smattering of studies from the Global South are predominately from **Asia**, notably China, but also India, Thailand, Iran, Philippines, and Indonesia.
-<!-- These countries add to our knowledge stemming from research reported from South Korean, Israel, and Japan.
-These studies focus on a variety of modes, population groups, and equity conceptualizations and standards that cannot be succinctly summarized.
-<!-- develop further, if there's room?-->The next most common focus within the literature from the Global South is from **South America**.
+The next most common focus within the literature from the Global South is from **South America**.
 Many of these studies mention a systemic absence of evidence relevant to the region [@vecchioTransportEquityLatin2020].
@@
 
-<!--`r word_comment("Papers that focus ONLY on (Dis)ability, or ONLY on Gender, or ONLY on Race/Ethnicity, or ONLY on Employment/Education level... the only focus", ".")`-->
 Papers that exclusively focus on populations with (dis)abilities e.g., [@parkJourneyVisuallyImpaired2017; @chiscanoImprovingDesignUrban2021; @orellanaWalkRollMapping2020] are relatively common in the literature.
@@
 We include a catch-all category of papers (*Other*) that include group populations that less commonly have been the subject of research.
-Examples include: veterans and access to specific-healthcare needs [@mooneyTravelDistanceBarrier2000], pregnant people and access to services [@vadrevuMeasuringSpatialEquity2016], and youths who live in foster care [@batscheUsingGISEnhance2012]<!-- rural populations people who use transit and their service needs relative to urban needs @sharavWhatTransitService2019, only cyclists in the context of health-implications of cycling in Delhi (India) @apparicioCyclingOneMost2021 ) -->.
+Examples include: veterans and access to specific-healthcare needs [@mooneyTravelDistanceBarrier2000], pregnant people and access to services [@vadrevuMeasuringSpatialEquity2016], and youths who live in foster care [@batscheUsingGISEnhance2012].
 Overall, the diversity of the *Other* population group classification demonstrates the diversity of transportation-equity concerns across population groups in the reviewed literature and the interplay of characteristics in the literature [@vecchioFairTransportPolicies2022].
@@
 In other words, these studies often benefit from well developed and institutional data that represents 'typical trips', especially in the Global North where these data are more readily available.
-
-<!-- DATA AVAILABILITY AS A MATTER OF JUSTICE: IS THIS AN ADDITIONAL CALL FOR THE NEXT SECTION?
-IN FACT, THE NATIONAL TRANSPORT POVERTY SURVEY EXPLICITLY AIMS TO ADDRESS THE LACK OF DATA FOR PARTICULAR POPULATIONS IN CANADA!
-The relevant issues are not new, and have at least two parts (i.e., who counts?
-Who is measured?
-Who is a subject of justice?
-And, who gets to use the data?).
-With respect to the first part, who we count and how we count them has implications for our understanding of just who is a subject of justice, as illustrated by the history of racial classifications in the US [see @leeRacial1993].
-In the case of transportation, the issue of data (collection/availability) as a matter of justice is gaining traction as digitalization of data casts a starker light on these questions...for examples, see the papers by @sourbatiSmart2021 and also, yes, by @behrendtData2023 (with Mimi Sheller).
--->
 
@@
 
-<!--`r word_comment("Answering: what conceptualizations occur in papers with **population** standards?
-what types of methods?", ".")`--> A different way to approach standards is to see them from the lens of **population**, and when this is done, other methods are commonly used.
+A different way to approach standards is to see them from the lens of **population**, and when this is done, other methods are commonly used.
 Over 40% of papers that suggest **population**-based standards are founded on concepts of **well-being**.
@@
 
-<!--`r word_comment("Answering: what conceptualizations occur in papers that conceptualize BOTH population and opportunity standards?
-what types of methods?", ".")`--> Papers that combine **population**- and **opportunity**-based standards often rely on **vertical equity** and **transport-related social exclusion** (note the similar proportions in these standards in @fig-fig6).
+Papers that combine **population**- and **opportunity**-based standards often rely on **vertical equity** and **transport-related social exclusion** (note the similar proportions in these standards in @fig-fig6).
 Perhaps not surprisingly, these papers often include a combination of methods: questionnaires and other qualitative methods related to population standards and quantitative methods such as accessibility indices for opportunity standards are usually deployed.
@@
 
-<!--`r word_comment("Answering: what conceptualizations occur in papers that conceptualize infrastructure standards?
-what types of methods?", ".")`--> Besides **opportunity**- and **population**-based standards, another way to make statements about fairness is through **infrastructure**-based standards.
+Besides **opportunity**- and **population**-based standards, another way to make statements about fairness is through **infrastructure**-based standards.
 Among papers that use this lens, the most common conceptual grounding used relates to **rights**.
@@
 
-<!--`r word_comment("Answering: what conceptualizations occur in papers that conceptualize Environmental+ standards?
-what types of methods?", ".")`--> Our corpus of literature was more or less muted in terms of studies with **environmental**-based standards (only about `r (stand_count$count[1]/sum(stand_count$count)) |> percent()` of all papers).
+Our corpus of literature was more or less muted in terms of studies with **environmental**-based standards (only about `r (stand_count$count[1]/sum(stand_count$count)) |> percent()` of all papers).
 Of those papers, the most frequent focus is on **inequitable externalities**.
@@
 Here, we critically review some of the trends that we identified from the literature in terms of concepts and standards (the "how") and how they apply to the objects (the "what") and subjects (the "who") of justice, as well as under which situations ("when" and "where" ).
-
-<!--
-- the "When" and "Where": 
-  - most studies focus on case studies in north american and europe
-  - south american/africa papers have key differences in focus compared to global north: they are more focused on affordability-as-a-barrier, tensions in infrastructure dev vs.
-service prioritization, data availability limitations
-  - 
-- the "Who":
-  - low income most popular
-   - (dis)abilities are exclusive focused population category in many studies; other 'vulnerabilities' are typically composite (aside from income, which is part of both exclusive-focus and composite vulnerabilities)
-  - transit, with cycling and walking becoming more popular
-  - cars are seldom the focus, and if they are, it is as a comparison to transit and/or focus on their negative externalities
-  - destinations of interest are mostly employment and healthcare; other destinations are less studied.
-The many 'destination' however, is the trip itself -- the experience, the infrastrucutre, the mode-service, a bundle of trips taken a month to all locations, etc.
-- the "What" and "How":
-  - accessibility/mobility -> opportunity standards are typically travel impedance standards, and affordability thresholds (e.g., more than 10% of monthly income should not be spent on transport [@rivasHowAffordableTransportation2018] to work and/or other necessary destinations).
-Oftentimes spatial equity, disparity analysis is operationalized.
-Infrastructure standards are also suggested -> oftentime Rights conceptualizations, i.e.
-access for all.
-  - transport-related environmental externalities -> pollution/noise standards are suggested, often using WHO guidelines or local neighbourhood averages as a comparision.
-  - health  -> health standards, the reliance on population transport health literature such as physical activity guideline recommendation per week or excess mortality burden. 
-  - safety -> the use of engineering standards and best practice
-  - cross-dimensional -> measures such as urban livability, etc. a more holistic conceptualization of transport equity
--->
 
@@
 
-From a modal perspective, transit is also the primary focus of the literature; cycling and walking are the next most popular focuses.
-These modes use public-space and users can interact with the public when using them (unlike car modes, which are seldom the focus of the reviewed literature).
-Upon the rise of automobility, transit was seen as a public service and something hindering automobility.
-But historically, mass transit and active modes are how people get around - and automobility is in direct tension with the movement of other modes as it requires a lot of pubic land and public subsidy.
-This may be one reason why transport inequity research focuses on these modes, often car modes as driving inequalities in many dimensions (e.g., access/mobility, traffic-related pollution, traffic-safety, and human-health).
-
-In the reviewed corpus, destinations of interest are predominantly employment and healthcare; other destination-types are less frequently studied.
-We know other destinations-type matter too, but we can only assume due to data restrictions and/or incomplete framing of "Who?" are the subjects of justice, these destinations are underrepresented in the corpus.
-In another significant branch of the corpus however, many papers do not consider specific destinations but the trip themselves -- be it the the travel experience, the quality of the infrastructure, the mode-service, a bundle of trips taken a month to all locations, etc.
-Again, the framing of "Who?" is exceptionally important in determining if the subject of justice should be a specific destination type, composite destinations, or aspects of the transport service quality.
+From a modal perspective, transit users are a primary focus of the literature.
+Upon the rise of automobility, transit came to be seen as a social service and sometimes as a hindrance to the full realization of automobility.
+After transit users, cyclists and pedestrians are the next most common populations of interest.
+Historically, walking is how people get around [@robertsShort1998], and cycling and transit use space more efficiently and have relatively low impact, but automobility is in direct tension with other modes as it requires massive amounts of space and receives extraordinary public subsidies, from the way its true cost is systematically underestimated [@gosslingSocial2019] to the way governments support the industries that feed it [@timperleyFight2021].
+This may be one reason why transport inequity research focuses on these modes, although often car is the main driver of inequalities in other dimensions (e.g., access/mobility, traffic-related pollution, traffic-safety, and human-health).
+The focus on active travel and transit is to some extent understandable, given the common but perhaps misleading notion, that travel by car equates privilege.
+This is, we would argue, a somewhat mypoic omission in the literature reviewed, since it is possible that there are obligated car users who lack other privileges including, for example, people with mobility disabilities [@darcyRoad2018] or the working poor [@blumenbergAutomobile2012].
 
@@
 
+<!-- Age, gender, sex, ethnicity, etc.??-->
+
 ## "What" are the objects of transportation justice
+
 <!-- 
@@
 Depending on the characterisation of "What", the thing of inequities and the object of justice, the focus varies.
+In the corpus reviewed, destinations of interest are predominantly to employment and healthcare; other destination types are less frequently studied.
+We know other destinations-type matter too, but we can only assume due to data restrictions and/or incomplete framing of "Who?" are the subjects of justice, these destinations are underrepresented in the corpus.
+In another significant branch of the corpus however, many papers do not consider specific destinations but the trip themselves -- be it the the travel experience, the quality of the infrastructure, the mode or service, a bundle of trips taken a month to all locations, etc.
+Again, the framing of "Who?" is exceptionally important in determining what objects of justice are relevant to whom.
+
 The literature offers a variety of conclusions and recommendations that span across the benefits of transportation (e.g., destinations types) that are often co-mingled with population groups and the tools of mobility.
@@
 
-# Moving forward: concluding remarks and calls for action
+# Moving forward: calls for action
 
@@
 
-Measuring context-specific evidence of transport inequities is necessary to build knowledge of injustices and inspire creative interventions that indeed bend the arc of history towards justice.
-Understanding and tracking transport inequities over time and space; specifically knowledge of the "who", "what", "where", "how" and especially "why" (the rationale) are integral.
-With this knowledge, we can set policies that address these inequities while importantly staying flexible to changing demands of justice.
+Measuring context-specific evidence of transport inequities is necessary to build knowledge of injustices and inspire creative interventions that advance the cause of justice.
+Understanding and tracking transport inequities over time and space, specifically knowledge of the "who", "what", "where", "how" and especially "why" (the rationale) is essential.
+Armed with this knowledge, the public, policy- and decision-makers, and scholars can help design policies to address these inequities, while remaining adaptable to face the changing demands of justice.
 
 Flexible frameworks are important.
-As mentioned, transport is derived demand – but what’s everyone’s demand look like?
-Past, current, and future?
-Creating transport equity frameworks that lend themselves to changing as the struggle for Justice moves forward and is resilient to backsliding is crucial.
-Setting standards from a systems approach: both considering the _positive_ rights (a right to have access to sufficient quality of essential services, for example) and _negative_ rights (mobility of cars must be limited as not to impact air quality, health, safety, and positive rights) must be both be conceptualized.
-
-In this report, we outline a conceptual framework for structuring equity analysis that may move us closer to transport justice, summarize findings from 165 relevant transportation equity papers from the academic literature, and critically appraised these findings with the conceptual transport justice framework.
+As mentioned, transport is derived demand.
+But what is considered (in)essential today may change during the next pandemic or technological revolution.
+Who is to say what demand for travel will look like tomorrow and beyond?
+Creating adaptable transport equity frameworks that lend themselves to change as the demands of justice change is as important as ensuring that they are resilient to backsliding.
+We argue for the necessity of setting standards from a systems approach, consiering both _positive_ rights (a right to have access to sufficient quality of essential services, for example) as well as _negative_ rights (mobility of cars must be limited to reduce their impact on air quality, health, safety, and positive rights).
+
+In this report, we outlined a conceptual framework for structuring equity analysis that aims to satisfy the above needs.
+We also summarize the findings from 165 relevant transportation equity papers from the academic literature, and critically appraise them.
 From what we saw and what we did not see in the reviewed literature, we set out the following calls for action.
-Some of these calls need to be heed by decision- and policy-makers; others by scholars; and others by members of the public.
-
-## Call 1: The need for explicit conceptualizations and grounded standards
+Some of these calls target one or more of different audiences, including decision- and policy-makers, planners, scholars, and members of the public.
+
+## Call 1: The need for explicitly state the conceptual grounding for standards
 <!--*And be clear with our terms – what is equity and for whom?**-->
-The rationale behind the conceptualizations of standards are often implicit within the literature: for example, @muellerSocioeconomicInequalitiesUrban2018 suggests the relative risk of mortality as related to transport-related air pollution should not be higher in deprived groups than the general population.
-It is not evident from the paper what conceptualization drives this focus on human-health inequities but can be implicitly inferred to be general "well being".
+The conceptual grounding for standards is often left implicit within the literature: for example, @muellerSocioeconomicInequalitiesUrban2018 suggests the relative risk of mortality as related to transport-related air pollution should not be higher in deprived groups than the general population.
+It is not evident from the paper what are the conceptual foundations for this analysis, but we can infer with with some confidence that this is a standard in the service of reparative justice.
 To move equity analysis outputs towards just transport futures, explicit inclusion of the rationale for justice, the "Why?", should become more common place.
@@
 Some standards are also seemingly arbitrary.
-For example, @caoIndicatorsSocioSpatialTransport2017 proposes 20 ferries per day to avoid social exclusion for inter-island transport planning in the Philippines in their analysis, but admits that a standard should be politically determined.
-It is unclear if, for instance, 10 or 30 ferries would make a difference in a specific quality-of-life outcome or if that number is tied to funding/resource constraints.
-As another example, if '15 minute city' is the policy goal, what conceptualization is driving this standard?
+For example, @caoIndicatorsSocioSpatialTransport2017 proposes 20 ferries per day to avoid social exclusion for inter-island transport planning in the Philippines.
+However, these authors admit that a standard should be politically determined.
+It is unclear if, for instance, 10 or 30 ferries would make a difference in any specific object of justice (e.g., accessibility to particular destinations) or if that number is tied to funding or resource constraints.
+As another example, what is the conceptual underpinning of the 15-minute city, where 15 minutes is the standard for travel times?
 Is it sufficientarism?
-Egalitariansm?
-The standard can be interpreted a multitude of ways.
+Or is it egalitariansm?
+The standard can be interpreted in a multitude of ways.
 The framing used to conceptualize justice must guide the standard.
-Being explicit about justifications for the framing is important.
-
-If literature is to recommend standards, researchers need standards linked to grounded outcomes.
-For decision-makers setting standards, measuring inequities and moving towards setting flexible guidelines for standards is the next step.
-As such, we believe, firmer justification of standards are lacking in the reviewed literature but necessary.
+As @martensFair2021 convincingly show, being explicit about "why" is important.
+
+Furtheromre, if one function of the literature is to recommend standards, researchers need to link them to grounded outcomes.
+For decision-makers, setting standards, measuring inequities and moving towards setting flexible guidelines for standards is the next step.
+As such, we believe, firmer justification of standards are often missing in the reviewed literature, but they are sorely needed.
 
@@
 On the methodological side, we believe more mixed methods are needed in transport equity research.
-Conceptualizations and standards are usually discussed from purely qualitative or quantitative approaches, a missed opportunity to combine the strengths of both approaches, whether by deep diving into some particular experiences or perceptions through qualitative methods or tailoring more meaningful quantitative analysis after qualitative explorations.
-For instance, @xieWonCycleRoute2018 finds through interviews and go-alongs with women cyclists that the standard Cycling Level of Service (CLS) tools used by engineers to plan cycling infrastructure misses the critical gendered perspective.
-Further, @somenahalliAgingTransportMobility2007 surveys older adults to understand their mobility issues, revealing factors that are unseen in standard daily travel surveys.
+Concepts and standards are usually discussed from purely qualitative or quantitative approaches.
+We believe this is a missed opportunity to combine the strengths of both approaches, whether by deep diving into some particular experiences or perceptions through qualitative methods or tailoring more meaningful quantitative analysis after qualitative explorations.
+For instance, @xieWonCycleRoute2018 find through interviews and go-alongs with women cyclists that the standard Cycling Level of Service (CLS) tools used by engineers to plan cycling infrastructure misses the critical gendered perspective.
+Further, @somenahalliAgingTransportMobility2007 survey older adults to understand their mobility issues, revealing factors that are unseen in standard daily travel surveys.
 
 While there is plenty of disparity analysis in the reviewed literature, frequently the standards used do not explicitly reflect practical application.
-For example, metrics of accessibility (usually measured with travel impedance cut-offs of between 15 to 60 minutes depending on the destination, population group and mode) are used to show descriptive differences among areas and groups but with scarce practice implications.
+For example, metrics of accessibility (usually measured with travel impedance cut-offs of between 15 to 60 minutes depending on the destination, population group and mode) are used to show descriptive differences among areas and groups but with scarce implications as to the experience of travellers.
 How low of an accessibility index is too low?
@@
 This lack of discussion makes translating results from these equity analysis into policy practice that may move conditions towards justice challenging.
-Creative methods and discussion of metrics of quantitative accessibility with intention of practical intervention should be paired with results; these methods should yield interpretable results and the explicit discussion of minimums _and_ maximums values in the distribution of the object of justice (the "What"), as applicable, is critical.
-
-Other times, when analysis engages with metrics that may be tied to conceptualizations (like Gini coefficient or Theil index), they fall short in assessing if the results' good or bad e.g., [@mijaresEquityAnalysisUrban2013].
+Creative methods and discussion of metrics of quantitative accessibility with intention of practical intervention should be paired with results; these methods should yield interpretable results and the explicit discussion of minimum _and_ maximum values in the distribution of the object of justice (the "What"), as applicable, is critical.
+
+Other times, when analysis engages with metrics that may be tied to particular concepts of equity (like Gini coefficient or Theil index), they fall short in assessing whether the results are good or bad [e.g., @mijaresEquityAnalysisUrban2013].
 If a Gini coefficient of 0 means that all people have the same access to public transport stops, what does it mean if the result is 0.2, 0.3, or 0.4?
-Is this good or bad news for decision-makers?
+Is this good or bad news?
+How should a decision-maker interpret this?
 Are new policies needed to reduce that number to a certain threshold, orienting future interventions?
 These questions usually remain unanswered despite its importance.
-These measurements can also bring some challenges and pitfalls, as recently summarized by @karnerAdvancesPitfallsMeasuring2023, but are necessarily to move equity analysis into application and towards transport justice.
+These measurements can also bring some challenges and pitfalls, as recently summarized by @karnerAdvancesPitfallsMeasuring2023, but are necessary to move equity analysis into application and towards transport justice.
 
@@
 When/Where/How is it measured?
-and who gets the data?
+And, who gets the consult and use the data?
 Deciding who is the subject of justice frames the data collection activity; if it's the mobility of those that do domestic work the classification of who does this work and how it changes over time/space is fundamental.
@@
 
+<!-- DATA AVAILABILITY AS A MATTER OF JUSTICE: IS THIS AN ADDITIONAL CALL FOR THE NEXT SECTION?
+IN FACT, THE NATIONAL TRANSPORT POVERTY SURVEY EXPLICITLY AIMS TO ADDRESS THE LACK OF DATA FOR PARTICULAR POPULATIONS IN CANADA!
+The relevant issues are not new, and have at least two parts (i.e., who counts?
+Who is measured?
+Who is a subject of justice?
+And, who gets to use the data?).
+With respect to the first part, who we count and how we count them has implications for our understanding of just who is a subject of justice, as illustrated by the history of racial classifications in the US [see @leeRacial1993].
+In the case of transportation, the issue of data (collection/availability) as a matter of justice is gaining traction as digitalization of data casts a starker light on these questions...for examples, see the papers by @sourbatiSmart2021 and also, yes, by @behrendtData2023 (with Mimi Sheller).
+-->
+
 ## Call 4: More direct and explicit links between standards and experienced outcomes
+
 <!--*Understand what institutional context we’re operating in**-->
-We believe that a more robust assessment of the implications of equity standards on life outcomes is lacking in the literature and is essential for equity analysis that translate into just practice.
+We believe that a more robust assessment of the implications of equity standards on the experience of the public is still lacking in the literature and is essential for equity analysis that translate into just practice.
 While the estimation of the benefits of increased mobility or accessibility, or reducing affordability burdens and transport externalities is commonplace in the literature, these estimations need to be associated with outcomes like life and neighborhood satisfaction, subjective well-being, mental and physical health, social capital, among others.
 Further, the use of standards that are tangibly linked to outcomes can more concretely define the objects of justice (the "What").
-They may also be used to track progress towards goals across time and space.
+They may also be used to track progress towards justice across time and space.
 
@@
 <!--**Understand what institutional context transfer looks like**-->
-We believe there is a need to evaluate more equity interventions or policies.
+We believe that there is a need to evaluate more equity interventions or policies.
 In our review, only 19 out of 165 studies assess specific transport-related policy interventions through an equity lens <!-- double check this number-->.
@@
 
-## Final remarks
-<!-- HERE IS AN OPPORTUNITY HERE TO CIRCLE BACK TO BEZOS SUBORBITAL FLIGHT, TO CLOSE THE REPORT WITH A REMINDER OF WHY IT IS IMPORTANT TO LEARN ABOUT JUSTICE.
-
-And this is my attempt - more time is needed but this first attempt is okay! -->
-To conclude, the analysis of transportation equity is vitally important for planning: as transportation systems and technologies converge space-time, they by definition leave certain populations, depending on the area and time scale, more or less burdened or benefited.
+# Final remarks
+<!-- HERE IS AN OPPORTUNITY HERE TO CIRCLE BACK TO BEZOS SUBORBITAL FLIGHT, TO CLOSE THE REPORT WITH A REMINDER OF WHY IT IS IMPORTANT TO LEARN ABOUT JUSTICE.-->
+
+We began this report with the history-making suborbital flight of Jeff Bezos and his fellow passengers, all of them very wealthy people.
+We wondered why coverage of the event, in particular Bezos declarations planetside, were not met with more critical questions: who are the subjects of the marvelous future that Bezos fancies his trip will contribute to?
+All of the human species or only those who can afford space tourism?
+What are the benefits of space tourism and how are they distributed?
+Who are the recipients of the impacts?
+
+As a point for contrast, we noted the treatment that Amazon workers receive, including the degrading need of drivers to urinate in bottles while in their deliver trucks, in between dropping off Amazon parcels, in order to meet their productivity quotas.
+Who benefits from this indignity?
+Who pays for it?
+
+These are questions that intuitively seem easy: as we noted, humans have a keen sense of what is fair.
+But to move the political needle, it is critical that we have the concepts, the vocabulary, and the knowledge to express our intuitions of justice.
+The objective of this report is to provide a measure of clarity, with a particular focus on the definition and use of standards of fairness in transportation.
+
+We argue that transportation, as an essential technology that can facilitate or hinder participation in society, is a very relevant sphere for matters of justice.
+Tthe analysis of fairness and equity in transportation is vitally important for planning: as transportation systems and technologies converge space-time, they by definition leave certain populations, depending on the area and time scale, more or less burdened or benefited.
 The analysis and tracking of equity implications is fundamental in formulating standards that are flexible to the fluid demands of transport justice.
@@
 
-The transportation justice framework can be seen as a flexible classification exercise that is suppored by a "Why" rationale.
+The transportation justice framework can be seen as a flexible classification exercise that is supported by a "Why" rationale.
 In conducting an equity analysis, the subject of justice (the "Who?") must contextualize the process, the distribution of the object of justice (the "what?") must be measured, be it transport-related benefits, burdens or both, "How?" these inequities are measured and benchmarked must be described, and these questions must be contextualized by the "Where?" across temporally changing "When?"s.
@@
 As mentioned in the introductory paragraphs, defining _fairness_ is complex, but complexity is no excuse to shirk away from the task.
-A founding rationale for the root "Why?" of in/justice is needed to inform equity analysis and set standards: why is it that Bezo (and the likes) ability to fly to space compared to the majority of others who are firmly grounded on earth on penny-pinched public transport systems traveling to Bezo-connected job opportunities felt as _unfair_?
-What aspects are _unfair_?
-To whom and in what ways?
-The answer to "Why?" is fundamental in pinning down fairness and justice.
-We must be explicit, intentional and clear in our work in transport equity analysis such that standards can be created, equity analysis be tracked and used to inform policy intervention that move towards justice.
+We hope that this report will prove useful to promote the idea that an anonymous Amazon driver is no less deserving of "access to the opportunities they need to lead a meaningful and dignified life" [@karnerTransportationEquityTransportation2020, p.
+440] than Bezos is for all his wealth.
+More importantly, we hope that in reviewing the literature, this report will contribute to advance the development and implementation of standards that help bend the arc of history towards more just transportation systems, and ultimately better societies.
+
+\newpage
 

### 2023-11-08 Manuscript/Manuscript.qmd
Incorporation of Ignacio comments into manuscript.

@@
         axis.text = element_text(size=7, colour = "black"),
-        axis.title = element_blank())+
-  ggtext::geom_richtext(aes(x=5.385, y=23.5, label="Conceptualizations"), fill="grey", size=2) +
-  ggtext::geom_richtext(aes(x=3, y=49, label="Standards"), angle = 270, fill="grey", size=2) 
+        axis.title = element_blank())
+#+ ggtext::geom_richtext(aes(x=5.385, y=23.5, label="Conceptualizations"), fill="grey", size=2) +
+#  ggtext::geom_richtext(aes(x=3, y=49, label="Standards"), angle = 270, fill="grey", size=2) 
 # standards_conc_percentages_plot
@@
   gen_grob(totals_conc_table_flex, scaling = "fixed", fit="auto"),
-  nrow = 2, ncol=2, rel_heights = c(1.5,1), rel_widths = c(1,4) )
+  nrow = 2, ncol=2, rel_heights = c(0.8,0.7), rel_widths = c(0.8,4) )
 #standards_conc_percentages_plot_and_table
@@
 
-- Accessibility:  
-- Capabilities approach:  
-- Equality:  
-- Equity:  
-- Fairness:  
-- Justice:  
-- Mobility:  
-- Object of justice:  
-- Policy:  
-- Social exclusion:  
-- Space-time convergence:  
-- Subject of justice:   
-- Sufficientarianism:  
-- Theories of justice:  
-- Utilitarianism:
-
+- **Fairness**: is achieved when a collective and individuals are treated equally and without bias as defined by their morals and values. (*maybe add this into the intro??*)
+- **Equity**: the distribution of benefits and burdens of things among the population, with a particular emphasis on those with the least advantage and quality in their outcomes.
+- **Justice**: mutually reinforced by the tools of fairness and equity, justice is a multi-dimensional political ideal attained when people "give and receive whatever they are due" [@jaggarPhilosophicalChallengesGlobal2009, pp.
+1-2].
+It ceases to exist when there are persons or groups that are denied "access to the opportunities they need to lead a meaningful and dignified life" [@karnerTransportationEquityTransportation2020, p.
+440].
+The process towards justice is fluid and can be identified through questions that answer what is the context, subject and object of justice, and what is the rationale for justice.
+- **The temporal-spatial context of justice**: the answers to "When?" and "Where?"
+- **The subjects of justice**: the answer to "Who?" is the focus of justice.
+- **The objects of justice**: the answer to "What?" is the focus of justice; what are the things that characterise (in)justice. 
+- **The methods for justice determination**: the answer to "How?" is justice defined and markers towards it tracked.
+- **The rationale to justice**: the "Why?". Connects with the context, subject, object and methods used for justice determination.
+- **Equity standards**: are a tool for answering this distributive question: how do we allocate burdens and benefits and to whom?
+Standards are s that when operationalized effectively define what is fair.
+- **Accessibility**:  
+- **Capabilities approach**:  
+- **Equality**:  
+- **Mobility**:  
+- **Policy**:  
+- **Social exclusion**:  
+- **Space-time convergence**:  
+- **Sufficientarianism**:  
+- **Theories of justice**:  
+- **Utilitarianism**:  
 Etc.,
@@
 Few reporters saw it fit to ask what problems the billionaire planned to solve on Earth, or what kind of future Bezos was trying to build, and for whose children.
-In other words, the billionaire was not confronted with questions about what did his trip do _for whom_ and _to whom_.
+In other words, the billionaire was not confronted with questions about what his trip did _for whom_ and _to whom_.
 A less dispassionate observer might have been excused for wondering (possibly aloud) about the basic _fairness_ of a man amassing a nigh unimaginable fortune that allowed him to build and fly his own spaceship, while masses of his employees were treated as throwaway cogs in the vast apparatus of his logistics empire.
@@
 The fact that some must give up jobs because it takes them too long to reach them by public transportation would constitute proof that those people should have studied more, been earlier risers, worked harder [@SpurrWaiting2015; @greismanLittle2017].
-Again, the members of the collective might think this a fine state of affairs, having come to this conclusion of their own accord or after being persuaded by billionaires.
+Again, the members of the collective might this is a fine state of affairs, having come to this conclusion of their own accord or after being persuaded by billionaires.
 Or, contrariwise, the members of the collective might decide that this state of affairs is _unjust_: the _values_ of those in charge of defining what is "fair" matter.
@@
 The objective of this report is to scan the state of knowledge in terms of defining and operationalizing "fairness" in the transportation domain.
-Much research has been devoted to the issues of _measuring _ equity in transportation, including (among many others) @ramjerdiEquity2006, @delboscUsing2011, @welchMeasure2013, @martensMeasuring2019, and @pritchardNew2022.
+Much research has been devoted to the issues of _measuring_ equity in transportation, including (among many others) @ramjerdiEquity2006, @delboscUsing2011, @welchMeasure2013, @martensMeasuring2019, and @pritchardNew2022.
 Further, there are multiple sources that discuss the conceptual and philosophical foundations of equity and fairness in transportation [e.g., @martensTransport2016; @pereiraDistributiveJusticeEquity2017a; @vanoutriveHow2019].
@@
 Supporting the creation of (more) just transportation systems involves understanding the production, distribution, and management of transportation benefits and costs; how they are distributed; and what values are implemented (and by whom) in the form of standards [@pereiraDistributiveJusticeEquity2017a; @shellerMobilityJusticePolitics2018; @pereiraTransportationEquity2021].
-Thus, for this review we engage the literature with the following questions in mind: 1) what is our current understanding of the things that transportation systems do, for whom, and to whom; 2) what does the literature say about the distribution of the benefits and costs of transportation systems; 3) and what values are embodied in normative statements about said distribution.
+Thus, for this review we engage the literature with the following questions in mind: 1) what is our current understanding of the things that transportation systems do, for whom, and to whom; 2) what does the literature say about the distribution of the benefits and costs of transportation systems; and 3) what values are embodied in normative statements about said distribution.
 Ultimately, this review aims to collate the existing academic knowledge on the matter, and present it in a manner useful to support the development and implementation of standards for equity in transportation planning and policy.
@@
 
-**Retributive justice**. It is concerned with the retribution due to people who do wrongs.
-
-**Reparative (or restorative) justice**. This form of justice relates to the proper way to correct or rectify past wrongs even if the wrongdoers can no longer be meted retribution.
-
-**Procedural justice**.
-The main concern of this form of justice is to ensure that the views, opinions, and preferences of all stakeholders are properly accounted for when decisions that affect their lives are made.
-
-**Distributive justice**.
-This is perhaps the most commonly studied form of justice [see @jaggarPhilosophicalChallengesGlobal2009, p.
+**Retributive justice** is concerned with the proportional retribution of wrongdoers with relation to legitimate punishers and the innocent [@sep-justice-retributive].
+
+**Reparative (or restorative) justice** focuses on the reparation of caused harm; it centers 'reintegrative shaming' to restore victims, wrongdoers and the community according to moral values [@tylerRestorativeJusticeProcedural2006].
+
+**Procedural justice** strives to ensure that the views and preferences of all stakeholders are fairly accounted for in the decision-making and inter-personal procedures affecting their lives and communities [@tylerRestorativeJusticeProcedural2006].
+
+**Distributive justice** is perhaps the most commonly studied form of justice [see @jaggarPhilosophicalChallengesGlobal2009, p.
 2; @pereiraDistributiveJusticeEquity2017a], and its main concern is the way the benefits and burdens of the tangible and intangible products of society are collected by different segments of a population.
 
-It might be argued that all of the above are particular forms of distributive justice.
-Retributive justice, for example, is often concerned with the distribution of the benefits and burdens of being a member of society; the way it is usually achieved is by distributing intangibles like "freedom" (e.g., of movement, of association) as benefits, and/or claiming tangibles like resources as burdens (e.g., fines).
-Reparative justice often distributes current benefits and burdens to redress past wrongs, for example by asking those who have benefited from said wrongs, even if unwittingly, to shoulder a bigger fiscal burden in order to cover programs that mete benefits to those who are still harmed by past wrongs.
-Procedural justice is the distribution of the benefits (e.g., the right to voice an opinion) and burdens (e.g., the effort required to develop an educated opinion) of the processes that lead to decisions with collective consequences.
+It might be argued that all of the above touch on forms of distributive justice.
+Retributive justice, for example, could be framed as being concerned with the distribution of the benefits and burdens of being a member of society; the way it is usually achieved is by distributing intangibles of a society's moral values like "freedom" (e.g., of movement, of association) as benefits, and the distribution of tangible resources as burdens (e.g., fines as a punishment).
+Reparative justice could look like the distribution of benefits and burdens to redress past wrongs, for example by asking those who have benefited from said wrongs, even if unwittingly, to shoulder a bigger fiscal burden in order to cover programs that mete benefits to those who are still harmed by past wrongs.
+Procedural justice could be the distribution of the benefits (e.g., the right to voice an opinion as a recognized stakeholder in the process) and burdens (e.g., the effort required to develop an educated opinion) of the processes that lead to decisions with collective consequences.
 
@@
 
-The term "equity" tends to encompass various tools to understand the distribution of benefits and burdens of things among a population, with a particular emphasis on those with the least advantage and equality in their outcomes.
+The term "equity", as conceptualized alongside distributive justice, tends to encompass various tools to understand the distribution of benefits and burdens of things among a population; there is a particular emphasis on those with the least advantage and equality in their outcomes.
 In the transportation domain, the term is somewhat loaded because it is perceived as stemming from the authority of the state, and meant to assist with decisions about regulating and financing transportation spending [@karnerTransportationEquityTransportation2020].
@@
 That this is the case is convincingly demonstrated by @martensFair2021 in their study of the application of Title VI of the Civil Rights Act of 1964 in accessibility planning in the US.
-Title VI explicitly talks about the distribution of benefits derived from Federal funding: "[n]o person in the United States shall, on the ground of race, color, or national origin, be excluded from participation in, be denied the benefits of, or be subjected to discrimination under any program or activity receiving Federal financial assistance." However, as @martensFair2021 show, there are several ways to comply with regulations while achieving different outcomes, ranging from the banal (do not _knowingly_ discriminate), to the substantive (compensation for past discrimination, i.e., reparative justice).
+Title VI explicitly talks about the distribution of benefits derived from Federal funding: "[n]o person in the United States shall, on the ground of race, color, or national origin, be excluded from participation in, be denied the benefits of, or be subjected to discrimination under any program or activity receiving Federal financial assistance." However, as @martensFair2021 show, there are several ways to comply with regulations while achieving different outcomes, ranging from the banal (do not _knowingly_ discriminate), to the substantive (compensation for past discrimination within a societal context that recognizes harm was done, i.e., reparative justice).
 What kind of justice does fairness serve in each case?
@@
 
-To recap the discussion of definitions so far, justice is an end.
-But to understand what that end is, we must clearly define standards of fairness.
+To recap the discussion of definitions so far, justice is a goal of social progress.
+But to understand what that goal is, we must clearly define standards of fairness.
 Equity analysis is tool to measure where the actual or perceived distribution of the burdens and benefits of the products of transportation systems stand with respect to the standard, in other words, instruments to see how close or far we are from a just situation.
@@
 We begin this section with our reflections about the way equity standards connect with conceptual and theoretical frameworks of justice.
-Broadly speaking, some trends emerge in terms of the methods used.
-**Opportunity** and **population** standards appear in the literature in similar proportions ( `r (stand_count$count[3] /sum(stand_count$count)) |> percent(accuracy=0.1)` vs.
-`r (stand_count$count[4] /sum(stand_count$count)) |> percent(accuracy=0.1)`), but frequently correspond to different conceptualizations of justice.
-Referring to @fig-fig6, much attention has been given to questions of **horizontal equity** and **spatial equity**, and over 40% of these papers propose standards that relate to **opportunities**.
-Within these papers, a frequent suggestion is to use standards that relate to travel impedance, which describes how fast, far, or costly it is to travel from one place to another.
+Broadly speaking, some trends emerge in terms of the methods used depending on the way equity is conceptualized and the type of standard suggested; See @fig-fig6 for definitions of of conceptualizations along the horizontal axis and types of standards along the vertical axis.
+These categories emerged as we reviewed the literature and we use them to discuss trends; they are by no means an exhaustive list of equity conceptualizations and types of standards.
+
+Referring to @fig-fig6, **Opportunity** and **population** are the most commonly suggested standards in the literature reviewed; they appear in the literature in similar proportions ( `r (stand_count$count[3] /sum(stand_count$count)) |> percent(accuracy=0.1)` vs.
+`r (stand_count$count[4] /sum(stand_count$count)) |> percent(accuracy=0.1)`), but frequently correspond to different conceptualizations of equity.
+Papers suggesting **opportunity standards** often address questions of **horizontal equity** and **spatial equity** and suggest standards that relate to travel impedance; these standards describes how fast, far, or costly it is to travel from one place to another.
 We think of this as a type of opportunity standard because travel impedance is essential to the measurement of accessibility (i.e., the ease of reaching opportunities).
-Some examples include the work of @chenImpactHighspeedRail2017, who use a standard of 4 hours or less by high-speed rail for municipalities to be considered "comfortably connected".
+Some examples include the work of @chenImpactHighspeedRail2017, who uses a standard of 4 hours or less by high-speed rail for municipalities to be considered "comfortably connected".
 In a similar manner, @yenisettyMeasuringAccessibilityVarious2020 assumes that a travel distance of less than 1,200 m is sufficient for a resident to interact with its transit system locally.
 And @shenMeasuringSpatialAccessibility2020 identify regions where populations cannot access hospitals within 1 hour by car - in other words, more than sixty minutes travel time is already seen as an unfair burden.
-Papers that suggest opportunity standards often employ disparity analysis through a variety of quantitative approaches.
-These include inequality measures such as the Gini coefficient and Lorenz curve, or poverty measures [@vanderveenOperationalizingIndicatorSufficient2020; @tiznadoaitkenRoleAccessibilityPublic2018]).
-There are also examples of spatial descriptive analysis as well as comparisons to benchmarks [e.g., equal supply to demand of public transit in a spatial unit as in @peungnumsaiGridBasedSpatialAnalysis2020] to determine which locations are spatially and horizontally (in)equitable.
-Another branch of quantitative research conceive the externalities of transportation system as trade-offs, and aim to maximize transport-related benefits (i.e., time savings, emissions reductions, congestion reductions, user fares) through optimization/location-allocation methodologies e.g., [@fengMulticriteriaEvaluationAccessibilitybased2014; @fakhrmoosaviIncorporatingTravelTime2021; @zhengAreabasedEquitablePricing2020; @wismadiSpatialPreferenceModelling2014].
-Papers that focus primarily on **horizontal equity** and/or **spatial equity** seldom use exclusively qualitative methods.
-
-A different way to approach standards is to see them from the lens of **population**, and when this is done, other methods are commonly used.
-Over 40% of papers that suggest **population**-based standards are founded on concepts of **well-being**.
-These papers typically ask *what is enough to lead a satisfactory life (as related to transportation)*, and the standards suggested include population benchmarks for comparison such as: questionnaires and relative comparisons to recommended physical activity per week [@adlakhaMindGapGender2020; @auchinclossDesignBaselineDescription2020; @mccormackRelationshipClusteranalysisDerived2012; @kimNeighborhoodEnvironmentWalkability2016; @towneUsingWalkScore2016].
-Summative per capita benchmarks include energy consumption for a 'decent living' [as suggested in @raoDecentLivingEmissions2012], and region-relative comparisons in health-related outcomes, such as premature mortality rates [@awuorInfluenceEnvironmentalHealth2019], spatial access to hospitals [@pereiraGeographicAccessCOVID192021], as well as spatial access to supermarkets, active-mode-usage, and Body Mass Index (BMI) [@murphySupermarketAccessTransport2017].
-The majority of these papers use quantitative/mixed-methods to identify inequities in **wellbeing**, however, a minority do use exclusively qualitative methods to distill themes, as for example in the exploration of *perceived* quality of life by @berheAdaptationDissonanceQuality2014.
+These papers uses these standards as a benchmark to indicate what areas are better or worse-off in terms of travel impedance; we can interpret them to be a type of _difference_ threshold.
+
+Papers that suggest these **opportunity** standards sometimes also employ disparity analysis through a variety of quantitative approaches to discuss the fairness in distribution.
+These include inequality measures such as the Gini coefficient and Lorenz curve, or poverty measures e.g., [@vanderveenOperationalizingIndicatorSufficient2020; @tiznadoaitkenRoleAccessibilityPublic2018]).
+They can also include spatial descriptive analysis as well as comparisons to service benchmarks [e.g., equal supply to demand of public transit in a spatial unit as in @peungnumsaiGridBasedSpatialAnalysis2020] to determine which locations are spatially and horizontally (in)equitable.
+Another branch of opportunity-standard-suggesting papers conceive the externalities of transportation system as trade-offs, and aim to maximize transport-related benefits (i.e., time savings, emissions reductions, congestion reductions, user fares) through optimization/location-allocation methodologies e.g., [@fengMulticriteriaEvaluationAccessibilitybased2014; @fakhrmoosaviIncorporatingTravelTime2021; @zhengAreabasedEquitablePricing2020; @wismadiSpatialPreferenceModelling2014].
+In the reviewed papers that focus primarily on **horizontal equity** and/or **spatial equity** seldom use exclusively qualitative methods and operationalise predominately quantitative methods.
+
+A different way to approach standards is to see them from the lens of **population**, and when this is done, other methods are commonly used, often depending on how equity is conceptualised.
+Papers that suggest **population**-based standards are often founded on concepts of **well-being**: they typically ask *what is enough to lead a satisfactory life (as related to transportation)?*.
+Standards suggested include: questionnaires and relative comparisons to recommended physical activity per week e.g., [@adlakhaMindGapGender2020; @auchinclossDesignBaselineDescription2020; @mccormackRelationshipClusteranalysisDerived2012; @kimNeighborhoodEnvironmentWalkability2016; @towneUsingWalkScore2016], summative per capita benchmarks (e.g., such as energy consumption for a 'decent living' as suggested in @raoDecentLivingEmissions2012), region-relative comparisons in health-related outcomes (e.g., premature mortality rates [@awuorInfluenceEnvironmentalHealth2019]), spatial access benchmarks to hospitals e.g., [@pereiraGeographicAccessCOVID192021], as well as spatial access to supermarkets, active-mode-usage, and Body Mass Index (BMI) e.g., [@murphySupermarketAccessTransport2017].
+The majority of these papers use quantitative/mixed-methods to identify inequities in **wellbeing**.
+However, a minority do use exclusively qualitative methods to distill themes, as for example in the exploration of perceived quality of life by @berheAdaptationDissonanceQuality2014.
 
 Papers that combine **population**- and **opportunity**-based standards often rely on **vertical equity** and **transport-related social exclusion** (note the similar proportions in these standards in @fig-fig6).
-Perhaps not surprisingly, these papers often include a combination of methods: questionnaires and other qualitative methods related to population standards and quantitative methods such as accessibility indices for opportunity standards are usually deployed.
-For instance, census data and the estimated proportion of households within some travel distance/time/availability to/of key destinations is used to identify a variety of social exclusionary situations [e.g., @mackettIncreasingAccessibilityCostEffectively2010; @chenExploringTravelCharacteristics2010; @danielsProposalAccessibilityPlanning2011; @sunPublicTransportAvailability2021; @sharmaPublicTransitAccessibility2021], the linkages between well-being and transport-related social exclusion [e.g., @delboscTransportProblemsThat2011; @churchillTransportPovertySubjective2019], areas more likely to experience transport poverty [@allenSizingTransportPoverty2019a], food deserts [@mckeyCrowdsourcedMappingHealthy2020], or transport-related energy poverty [@robinsonDoubleEnergyVulnerability2020; @berryInvestigatingFuelPoverty2016; @berryDistributionalEffectsCarbon2019].
+Perhaps not surprisingly, these papers commonly use a combination of methods: questionnaires and other qualitative methods related to population standards and quantitative methods such as accessibility indices for opportunity standards are usually deployed.
+For instance, census data and the estimated proportion of households within some travel distance/time/availability to/of key destinations is used to identify a variety of social exclusionary situations e.g., [@mackettIncreasingAccessibilityCostEffectively2010; @chenExploringTravelCharacteristics2010; @danielsProposalAccessibilityPlanning2011; @sunPublicTransportAvailability2021; @sharmaPublicTransitAccessibility2021], the linkages between well-being and transport-related social exclusion e.g.,[@delboscTransportProblemsThat2011; @churchillTransportPovertySubjective2019], areas more likely to experience transport poverty [@allenSizingTransportPoverty2019a], food deserts [@mckeyCrowdsourcedMappingHealthy2020], or transport-related energy poverty e.g., [@robinsonDoubleEnergyVulnerability2020; @berryInvestigatingFuelPoverty2016; @berryDistributionalEffectsCarbon2019].
 
@@
 They use a variety of methods to identify _where_ populations at risk may be located, such as clustering methods [@mohriClusteringMethodMeasuring2021].
-Few of the papers that employ exclusively qualitative methods use survey data to study the willingness to travel, or the barriers to travel [e.g., @chenExploringTravelCharacteristics2010; @mehdizadehWalkingTimeSchool2017] or interviews/focus groups related to topics of unmet activity needs [@nordbakkeTransportUnmetActivity2015].
-Travel that _did not_ take place is notoriously slippery to capture in standard surveys, and understanding whether this is a preference or the consequence of constraints is not straightforward either [e.g., @paezParticipation2012].
-
-Besides **opportunity**- and **population**-based standards, another way to make statements about fairness is through **infrastructure**-based standards.
+Some papers that exclusively use qualitative methods operationalise survey data to study the willingness/barriers to travel e.g., [@chenExploringTravelCharacteristics2010; @mehdizadehWalkingTimeSchool2017] or conducts interviews on topics of unmet activity needs [@nordbakkeTransportUnmetActivity2015].
+Travel that _did not_ take place is notoriously slippery to capture in standard surveys, and understanding whether this is a preference or the consequence of constraints is not straightforward [e.g., @paezParticipation2012].
+
+Besides **opportunity**- and **population**-based standards, another way to create statements about fairness is through **infrastructure**-based standards.
 Among papers that use this lens, the most common conceptual grounding used relates to **rights**.
-**Infrastructure**-standard appear in `r (stand_count$count[3]/sum(stand_count$count)) |> percent()` of all papers in this review, and within this segment of the corpus, discussions of **rights** appear two times as often than any other conceptual approach.
-These papers often focus on populations who face mobility disabilities and non-car users, and the discussion often centers about their right to partake of transportation systems.
-The methods used by this strand of the literature are varied but more or less equally split between infrastructure and environmental audits and qualitative methods.
-Audits of existing infrastructure are sometimes used in comparisons against best-practice universal design [@odeckEconomicAppraisalUniversal2010; @larkinsAccessibleTransportationBuilt2011; @jimenez-espadaResearchProblemUniversal2021; @perezdelhoyoParticipatoryManagementImprove2021] or to understand the elements of infrastructure that correlate well with use of a mode by particular populations [e.g., walking by older adults; @moniruzzamanInvestigation2016].
+**Infrastructure**-standards appear in `r (stand_count$count[3]/sum(stand_count$count)) |> percent()` of all papers in this review, and within this segment of the corpus, discussions of **rights** appear two times as often than any other conceptual approach.
+These papers often focus on populations who face mobility disabilities and non-car users, and the discussion often centers on their right to use existing transportation systems.
+The methods operationalised by this strand of the literature are varied but are more or less split between infrastructure and environmental audits, and qualitative methods.
+Audits of existing infrastructure are sometimes used in comparisons against best-practice universal design e.g., [@odeckEconomicAppraisalUniversal2010; @larkinsAccessibleTransportationBuilt2011; @jimenez-espadaResearchProblemUniversal2021; @perezdelhoyoParticipatoryManagementImprove2021] or to understand the elements of infrastructure that correlate well with use of a mode by particular populations [e.g., walking by older adults @moniruzzamanInvestigation2016].
 Qualitative methods are used to interview/survey users about their perceived access to transport systems [e.g., @marquezFactorsAffectingPersonal2019; @mateo-babianoBicycleSharingAsia2017; @furstBestBadPractices2013; @velhoEffectTransportAccessibility2016; @parkJourneyVisuallyImpaired2017; @limFacilitatingIndependentCommuting2021; @stjernborgAccessibilityAllPublic2019; @desjardinsGoing2021] or to assess standards under best-practice considerations [e.g., @daamenAssessingGapPublic2008; @velhoEffectTransportAccessibility2016; @bharathyRevisitingClearFloor2018].
@@
 For instance, @racheleNeighbourhoodSocioeconomicTransport2017a combine various properties of transport networks (e.g., street connectivity, cul-de-sac length, street block length, traffic volume, public transport stops and service frequency), and use these attributes as inputs to define an indicator of transport design that supports walkability and access to public transport.
-Other works assess the quality of infrastructure [@xuSocialInequalitiesPark2017], the severity and frequency of accidents on the system  [@benevenutoExaminingTransportNeeds2020; @appleyardTransitCorridorLivability2017], user-groups (particularly disadvantaged groups in the case of vertical equity) [@prasertsubpakijEvaluatingAccessibilityBangkok2012] as part of multi-criteria indicators.
-Yet another branch of literature explicitly focuses on affordability or other barriers to the transport system, and suggests improvements to the infrastructure such that all groups (especially the most disadvantaged) can sufficiently interact with the system [e.g., @basuPracticalFrameworkBenchmarking2019; @songWomenWheelsGender2019; @welchEquityTransportDistribution2013].
+Other works assess the quality of infrastructure [@xuSocialInequalitiesPark2017], the severity and frequency of accidents on the system [@benevenutoExaminingTransportNeeds2020; @appleyardTransitCorridorLivability2017], user-groups (particularly disadvantaged groups in the case of vertical equity conceptualisations) [@prasertsubpakijEvaluatingAccessibilityBangkok2012] as part of multi-criteria indicators.
+Yet another branch of infrastructure-based standard-suggesting literature explicitly focuses on affordability or other barriers to the transport system, and suggests improvements to the infrastructure such that all groups (especially the most disadvantaged) can sufficiently interact with the system [e.g., @basuPracticalFrameworkBenchmarking2019; @songWomenWheelsGender2019; @welchEquityTransportDistribution2013].
 These studies are grounded on concepts of **transport-related social exclusion** [e.g., @kentPrioritizingLowstressEquitable2019] or **sufficientarian/capabilities** approaches [e.g., @smithAccessibilityCapabilityMinimum2012].
@@
 Papers that work with environmental-based standards often use traffic-related air pollution, noise pollution, green-space, urban design elements, urban air temperature, health related outcome, and physical activity guidelines to quantify externalities.
-The methods used are almost always quantitative or mixed-methods, and inequalities are identified using spatial clustering techniques, Gini coefficient [@fengMulticriteriaEvaluationAccessibilitybased2014], comparisons to established environmental thresholds or health guidelines [@agostfelipInclusiveModelAssessing2021; @agostfelipInclusiveModelAssessing2021; @kruizeEnvironmentalEquityRole2007; @iungmanImpactUrbanTransport2021; @apparicioCyclingOneMost2021; @khomenkoLiveableCityHealthy2020; @muellerSocioeconomicInequalitiesUrban2018].
+The methods used are almost always quantitative or mixed-methods, and inequalities are identified using spatial clustering techniques, Gini coefficients [@fengMulticriteriaEvaluationAccessibilitybased2014], comparisons to established environmental thresholds or health guidelines [@agostfelipInclusiveModelAssessing2021; @agostfelipInclusiveModelAssessing2021; @kruizeEnvironmentalEquityRole2007; @iungmanImpactUrbanTransport2021; @apparicioCyclingOneMost2021; @khomenkoLiveableCityHealthy2020; @muellerSocioeconomicInequalitiesUrban2018].
 They sometimes create composite multi-dimensional indices [@agostfelipInclusiveModelAssessing2021; @mirandaBenchmarkingSustainableUrban2012; @corazzaMethodologyEvidenceCase2020], occasionally supported by spatial analysis [@jephcoteGeospatialAnalysisNaturally2013; @carrierApplicationThreeMethods2014].
@@
 
-Applying the analytical framework to questions of justice in transportation yields a number of insights.
-Here, we critically review some of the trends that we identified from the literature in terms of concepts and standards (the "how") and how they apply to the objects (the "what") and subjects (the "who") of justice, as well as under which situations ("when" and "where" ).
+Here, we critically review some of the trends identified from the literature in terms of concepts and standards (the "how") and how they apply to the objects (the "what") and subjects (the "who") of justice, as well as under which situations ("when" and "where").
 
@@
 Because of lower data availability and more extreme needs for 'sufficient' transport, analysis of transportation inequities are often cast along economic lines.
-An example of a standard in this context involves improving transit to ensure that at least 60% of those below the poverty line face a similar burden travelling to work, school, and services, as the average transit user in the region [i.e., 60 min travel time to destinations; @basuPracticalFrameworkBenchmarking2019].
+An example of a standard in this context involves improving transit to ensure that at least 60% of those below the poverty line face a similar burden travelling to work, school, and services, as the average transit user in the region [i.e., 60 min travel time to destinations @basuPracticalFrameworkBenchmarking2019].
 
@@
 
-From a modal perspective, transit users are a primary focus of the literature.
+From a modal perspective, people who use a certain mode can be seen as the subjects of transportation equity in the literature.
+In this way, transit users are the primary focus.
 Upon the rise of automobility, transit came to be seen as a social service and sometimes as a hindrance to the full realization of automobility.
 After transit users, cyclists and pedestrians are the next most common populations of interest.
-Historically, walking is how people get around [@robertsShort1998], and cycling and transit use space more efficiently and have relatively low impact, but automobility is in direct tension with other modes as it requires massive amounts of space and receives extraordinary public subsidies, from the way its true cost is systematically underestimated [@gosslingSocial2019] to the way governments support the industries that feed it [@timperleyFight2021].
+Historically, walking is how people get around [@robertsShort1998], and cycling and transit use space more efficiently and have relatively low impact.
+Automobility is in direct tension with other modes as it requires massive amounts of space and receives extraordinary public subsidies, from the way its true cost is systematically underestimated [@gosslingSocial2019] to the way governments support the industries that feed it [@timperleyFight2021].
 This may be one reason why transport inequity research focuses on these modes, although often car is the main driver of inequalities in other dimensions (e.g., access/mobility, traffic-related pollution, traffic-safety, and human-health).
 The focus on active travel and transit is to some extent understandable, given the common but perhaps misleading notion, that travel by car equates privilege.
-This is, we would argue, a somewhat mypoic omission in the literature reviewed, since it is possible that there are obligated car users who lack other privileges including, for example, people with mobility disabilities [@darcyRoad2018] or the working poor [@blumenbergAutomobile2012].
+We argue that this is a somewhat mypoic omission in the literature reviewed, since it is possible that there are obligated car users who lack other privileges including, for example, people with mobility disabilities [@darcyRoad2018] or the working poor [@blumenbergAutomobile2012].
 
@@
 We believe, more nuisances perspectives that capture the "who" of inequities is needed.
-As shown in the variety of methods used, dimensions analysis, and conceptualizations and standards, transport inequalities are multi-dimensional.
+We must go beyond the focus on the low-income transit rider; positioning intersectionality of socio-economic and mode-use to create community-based definitions of equity and tailored standards are critical.
+
+As shown in the variety of methods, dimensions, conceptualizations and standards used, transport inequalities are multi-dimensional.
 How decision makers define an equity-deserving community will impact results (e.g., [@rowangouldIdentifyingEnvironmentalJustice2015]).
@@
 Are inequities in the service of transit the focus of a study because it can be improved to address transport inequities and are alternative modes driving these inequities?
-Access to what sorts of opportunities is driving transport inequities?
+Access to what sorts of opportunities are driving transport inequities?
 How do populations, transit modes, and opportunities saught intersect to define the "who" of inequities?
-Community-based informed understanding of inequities are needed, and tracking how they change are needed.
+A community- informed understanding of inequities and tracking how it changes are needed.
 Different slices of analysis are important to different groups, as such, we believe all should be studied with an explicit justice framing.
 
-<!-- Age, gender, sex, ethnicity, etc.??-->
-
 ## "What" are the objects of transportation justice
 
-<!-- 
-NOTE: THIS SECTION DISCUSSED AT THE SAME TIME "WHAT" AND "HOW", BUT I SUGGEST TO SPLIT THE DISCUSSION IN TWO PARTS, SO THAT THE "HOW" IS ON PAR WITH OTHER ELEMENTS OF THE ANALYTICAL FRAMEWORK.
-IF POSSIBLE, IT WOULD BE GOOD TO FURTHER DEVELOP THE "HOW" SECTION, AS IT IS THE MOST IMPORTANT OF THE REPORT!
-
-Here, I've decided to talk in more depth about the equity conceptualization groups (rights, vertical equity, horizontal equity, ineq.
-externalities, TR exclusion, wellbeing, spatial equity, utilitarianism) across each equity dimension (access/mobility, envionrmental externality, human health, and traffic-safety).
-
-NOTE: I think we should remove 'inequitable externality" from the conceptualizations -- fold it into horizontal and/or vertical equity. Would need to update Fig 6.
--->
 Defining "What" are the things measured and the inequities that characterise their distribution is fundamental to tracking the progress towards transport justice.
@@
 Inequities in this "What" can be theorized to be equal if similar levels are attainable by all populations (horizontal equity) or all spatial areas (spatial equity).
-<!-- maybe change this theorization to 'equality' not horizontal/spatial equity?--> However, discussion of both minimums and maximums (those contributing the most to inequities) is seldom explicitly discussed through the lens of justice in the reviewed corpus.
-
-<!-- maybe change the "What" to be an object of INEQUITIES? not just justice? -->
+However, discussion of both minimums and maximums (those contributing the most to inequities) is seldom explicitly discussed through the lens of justice in the reviewed corpus.
+
 Another branch of papers theorize the object of justice to be population _well-being_ and they typically ask _what is enough to lead a satisfactory life (as related to transportation)_.
@@
 These papers conceptualize the object of justice to be _transport related social exclusion_,  _vertical equity_, and/or _sufficientarian/capabilities_.
-What is characteristic of these papers is the identification of standards that are linked to often linked to outcomes like papers that conceptualize _well being_, however outcomes are often more tangible or welfare informed.
-As discussed, disadvantaged groups are the focus of analysis and the linkages between well-being and transport-related social exclusion [e.g., @delboscTransportProblemsThat2011; @churchillTransportPovertySubjective2019], areas more likely to experience transport poverty [@allenSizingTransportPoverty2019a], food deserts [@mckeyCrowdsourcedMappingHealthy2020], or transport-related energy poverty [@robinsonDoubleEnergyVulnerability2020; @berryInvestigatingFuelPoverty2016; @berryDistributionalEffectsCarbon2019] are central to the object of justice.
-
-<!-- In what's described, what's missing from equity analysis in their conceptualization of the "What?" is more explicit rationale for the object of justice.
-A strong conceptualization of the "What" can make theorizing the "Why" rationale of justice easier.
--->
+What is characteristic of these papers is the identification of standards that are often linked to outcomes like papers that conceptualize _well being_, however outcomes are often more tangible or welfare informed.
+As discussed, disadvantaged groups are the focus of analysis and the linkages between well-being and transport-related social exclusion [e.g., @delboscTransportProblemsThat2011; @churchillTransportPovertySubjective2019], areas more likely to experience transport poverty [@allenSizingTransportPoverty2019a], food deserts [@mckeyCrowdsourcedMappingHealthy2020], or transport-related energy poverty [@robinsonDoubleEnergyVulnerability2020; @berryInvestigatingFuelPoverty2016; @berryDistributionalEffectsCarbon2019] are centered as the objects of justice.
+
+What's missing from the "What?" in the equity analyses of the literature reviewed is the consistently explicit rationale for the object of justice.
+Specifically, "What?" are the benefits and burdens of a transportation system in terms of its: accessibility and mobility, or environmental externalities, or human-health externalities, or traffic related safety, or multi-criteria mix.
+In the inter-disciplinary analysis of transport systems, crystalising "What?" is the object of justice can create a stronger foundation to select appropriate methods to determine fairness and support the "Why?", the rationale of justice.
+In the equity literature reviewed, strong conceptualizations of the object of justice often falls short.
 
 ## "How" is fairness determined
-<!-- 
-NOTE: MOST IMPORTANT!
-
-I need to add in more references from the corpus here... ran out of time.
--->
+
 How we define fairness should be fundamental to equity analysis.
@@
 Alternatively, they operationalize spatial mismatch (disparity analysis) [e.g., @mulleyAreNetworkPlanning2015] and identify areas of _relative_ (regionally) inequities that should be addressed base on impedance thresholds to certain destination types.
-Others apply these same types of methods but tend to focus of human health, traffic-related safety, or cross-dimensional "What"s.
-Methods can include inequality measures, poverty measures, and optimization methods to create 'optimal' landscapes, and once again vary depending on "Who" and "What" is centered.
+Others apply these same types of methods but tend to focus of human health, traffic-related safety, or cross-dimensional "What?"s.
+Methods can include inequality measures, poverty measures, and optimization methods to create 'optimal' landscapes, and once again vary depending on "Who?" and "What>" is centered.
 
@@
 Justification is exceptionally important, and we believe that a lot of the literature reviewed may miss this mark.
-However, literature that does not is, more often than not, connects to qualitative results and/or operationalizes mixed-methods.
+However, literature that does not, usually connects to qualitative results and/or operationalizes mixed-methods.
 
@@
 
-## Call 1: The need for explicitly state the conceptual grounding for standards
+## Call 1: The need for explicit conceptuations of equity to ground standards
 <!--*And be clear with our terms – what is equity and for whom?**-->
 The conceptual grounding for standards is often left implicit within the literature: for example, @muellerSocioeconomicInequalitiesUrban2018 suggests the relative risk of mortality as related to transport-related air pollution should not be higher in deprived groups than the general population.
-It is not evident from the paper what are the conceptual foundations for this analysis, but we can infer with with some confidence that this is a standard in the service of reparative justice.
+It is not evident from the paper what are the conceptual foundations for this analysis, but we can infer with some confidence that this is a standard that conceptualizes _wellbeing_ and could be in service of reparative justice.
 To move equity analysis outputs towards just transport futures, explicit inclusion of the rationale for justice, the "Why?", should become more common place.
-
-Some standards are also seemingly arbitrary.
+We must be clear with our terms – what is equity and for whom?
+
+Standards are used as a benchmark for fairness, but some standards are also seemingly arbitrary in the reviewed literature.
 For example, @caoIndicatorsSocioSpatialTransport2017 proposes 20 ferries per day to avoid social exclusion for inter-island transport planning in the Philippines.
@@
 
-Furtheromre, if one function of the literature is to recommend standards, researchers need to link them to grounded outcomes.
-For decision-makers, setting standards, measuring inequities and moving towards setting flexible guidelines for standards is the next step.
-As such, we believe, firmer justification of standards are often missing in the reviewed literature, but they are sorely needed.
-
-## Call 2: Need for creative methods for systems-thinking approach to inequities
+Furthermore, if one function of the literature is to recommend standards, researchers need to link them to conceptualisations of equity that are compatible with justice frameworks.
+For decision-makers, setting standards, measuring inequities and moving towards setting flexible guidelines for standards that are also compatible with community-informed calls for justice is the next step.
+As such, we believe, firmer justification of standards using justice frameworks are often missing in the reviewed literature, but they are sorely needed.
+
+## Call 2: The need for creative systems-thinking approaches to measuring (in)equities
 <!--**Understand what transport inequities you are measuring**-->
-On the methodological side, we believe more mixed methods are needed in transport equity research.
+On the methodological side, we believe the wider application of mixed-methods are needed in transport equity research.
 Concepts and standards are usually discussed from purely qualitative or quantitative approaches.
@@
 How low of an accessibility index is too low?
-Conversely, if a sufficient accessibility is met, high accessibility results in higher inequalities and are inequalities of a certain level an issue?
+Conversely, if a sufficient accessibility is met, high accessibility results in higher inequalities, but are inequalities of a certain level an issue?
 This lack of discussion makes translating results from these equity analysis into policy practice that may move conditions towards justice challenging.
-Creative methods and discussion of metrics of quantitative accessibility with intention of practical intervention should be paired with results; these methods should yield interpretable results and the explicit discussion of minimum _and_ maximum values in the distribution of the object of justice (the "What"), as applicable, is critical.
+Creative methods and discussion of metrics of quantitative accessibility with intention of practical intervention should be paired with results; they should yield interpretable results and the explicit discussion of minimum _and_ maximum values in the distribution of the object of justice (the "What?"), as applicable, is critical.
 
@@
 These questions usually remain unanswered despite its importance.
-These measurements can also bring some challenges and pitfalls, as recently summarized by @karnerAdvancesPitfallsMeasuring2023, but are necessary to move equity analysis into application and towards transport justice.
+These measurements can also bring some challenges and pitfalls (as recently summarized by @karnerAdvancesPitfallsMeasuring2023) but are necessary to move equity analysis into application and towards transport justice.
 
@@
 
-In the review of the literature, we were left wondering specifically - what are the motivations for the use of some destination-types and the choice to note include other destinations types?
+In the review of the literature, we were left wondering specifically - what are the motivations for the use of some destination-types and the choice to not include other destinations types?
 For instance, papers including leisure destinations (e.g., green space, parks, recreation) are infrequently studied and some categories are missing all together like 'mobilities of care'.
@@
 We can only track what we have collected and compiled; and we know that transport systems' focus is more than just to work or as a source of economic development (though in underdeveloped regions, transport systems as a force of economic development is pronounced based on the concentration of Global South transport-induce economic development studies e.g., high-speed rail [@chenImpactHighspeedRail2017; @monzonEfficiencySpatialEquity2013; @kimImpactsHighspeedRail2015]).
-As such, data availability matters, especially when operationalizing emerging theories (i.e., sufficientariansm [@vanderveenOperationalizingIndicatorSufficient2020]) in equity analysis.
+As such, data availability matters, especially in the context of the Global South and rural geographies as fewer official data sources and public research resources exist relative to urban communities and areas in the Global North.
+As well as in the operationalization of emerging conceptual theories (i.e., sufficientariansm [@vanderveenOperationalizingIndicatorSufficient2020]) in equity analysis.
 
@@
 IN FACT, THE NATIONAL TRANSPORT POVERTY  SURVEY EXPLICITLY AIMS TO ADDRESS THE LACK OF DATA FOR PARTICULAR POPULATIONS IN CANADA!
-The relevant issues are not new, and have at least two parts (i.e., who counts?
-Who is measured?
-Who is a subject of justice?
-And, who gets to use the data?).
-With respect to the first part, who we count and how we count them has implications for our understanding of just who is a subject of justice, as illustrated by the history of racial classifications in the US [see @leeRacial1993].
-In the case of transportation, the issue of data (collection/availability) as a matter of justice is gaining traction as digitalization of data casts a starker light on these questions...for examples, see the papers by @sourbatiSmart2021 and also, yes, by @behrendtData2023 (with Mimi Sheller).
 -->
 
-## Call 4: More direct and explicit links between standards and experienced outcomes
-
+## Call 4: Direct and explicit links between standards and experienced outcomes
 <!--*Understand what institutional context we’re operating in**-->
-We believe that a more robust assessment of the implications of equity standards on the experience of the public is still lacking in the literature and is essential for equity analysis that translate into just practice.
+We believe that robust assessments of the implications of equity standards on the experience of the public is still lacking in the literature, but it is essential for equity analysis that translates into just practice.
 While the estimation of the benefits of increased mobility or accessibility, or reducing affordability burdens and transport externalities is commonplace in the literature, these estimations need to be associated with outcomes like life and neighborhood satisfaction, subjective well-being, mental and physical health, social capital, among others.
-Further, the use of standards that are tangibly linked to outcomes can more concretely define the objects of justice (the "What").
-They may also be used to track progress towards justice across time and space.
+
+Composite measures such as a transport-land-use index proposed by @appleyardTransitCorridorLivability2017 is an example of systems-thinking approach that links findings to quality-of-life proxies; they ground their measure in the principle of _livability_ along corridors of varying levels of estimated transport-land-use integration.
+However, the methods used demonstrate areas of relative high and low livability could go further; they could be tied to absolute goals of integration or livability.
+Relatably, @higgsUrbanLiveabilityIndex2019 develops an urban livability index and demonstrates the relationship between the population's use of a certain mode in a neighbourhood and a one unit increase in the index.
+However, are there absolute minimums or maximums for the index or the mode-choice goals that should not be crossed?
+
+We need more explicit discussion of the boundaries in the distribution of the object of justice (the "What?") alongside these creative methods.
+These links may be used to track progress towards justice across time and space, a critical point for practitioners.
 
@@
 <!--**Understand what institutional context transfer looks like**-->
-We believe that there is a need to evaluate more equity interventions or policies.
-In our review, only 19 out of 165 studies assess specific transport-related policy interventions through an equity lens <!-- double check this number-->.
+We believe there is a need to evaluate equity interventions and policies: track their before, during, and after impacts.
+In this review, only approximately 10% of studies assess specific transport-related policy interventions through an equity lens <!-- double check this number-->.
 Examples include mode-shift from driving to active school travel [@mammenSchoolTravelPlanning2014], transit fare restructures [@hickeyUsingQuantitativeMethods2010] and spatial analysis of Low Traffic Neighborhoods [@aldredEquityNewActive2021].
 The assessment of interventions and tracking associated outcomes can be thought of as a key step towards transport justice.
-Outcomes of interventions can be compared within and between communities, and cross community comparisons can be created that may expedite the adoption of effective policies that move towards just outcomes.
+
+Outcomes of interventions can be compared within and between communities, and cross community comparisons can be created that may expedite the adoption of effective policies that move towards just outcomes; the presence of these synthesis and comparative studies could support brave decision-makers in the application of research into practice.
 
@@
 We argue that transportation, as an essential technology that can facilitate or hinder participation in society, is a very relevant sphere for matters of justice.
-Tthe analysis of fairness and equity in transportation is vitally important for planning: as transportation systems and technologies converge space-time, they by definition leave certain populations, depending on the area and time scale, more or less burdened or benefited.
+The analysis of fairness and equity in transportation is vitally important for planning: as transportation systems and technologies converge space-time, they by definition leave certain populations, depending on the area and time scale, more or less burdened or benefited.
 The analysis and tracking of equity implications is fundamental in formulating standards that are flexible to the fluid demands of transport justice.
@@
 
-## Examples of papers summarized by element of the analyitical framework
+## Examples of papers summarized by element of the analytical framework
 

### 2023-11-15 Manuscript/Manuscript.qmd
Incorporated Matt Palm's executive summary (EC), revised and inserted comments in EC, made minor edits to body and to Glossary, and added acknowledgements section

@@
 title: "Searching for standards of fairness in the transportation justice literature: A scoping review"
+subtitle: "Searching for standards of fairness"
+author:
+  - name:
+      given: Anastasia
+      family: Soukhov
+    affiliations:
+      - name: McMaster University
+        department: School of Earth, Environment and Society
+    email: soukhoa@mcmaster.ca
+  - name:
+      given: Ignacio
+      family: Tiznado-Aitken
+    affiliations:
+      - name: University of Toronto Scarborough
+        department: Department of Geography and Planning
+    email: i.tiznadoaitken@utoronto.ca
+  - name:
+      given: Matthew
+      family: Palm
+    affiliations:
+      - name: University of North Carolina Chapel Hill
+        department: Department of City and Regional Planning
+    email: palmmatt@unc.edu
+  - name:
+      given: Steven
+      family: Farber
+    affiliations:
+      - name: University of Toronto Scarborough
+        department: Department of Geography and Planning
+    email: steven.farber@utoronto.ca
+  - name:
+      given: Antonio
+      family: Paez
+    affiliations:
+      - name: McMaster University
+        department: School of Earth, Environment and Society
+    email: paezha@mcmaster.ca
 bibliography: All_papers_in_extracted.bib
@@
   DiagrammeRsvg::export_svg() %>% 
-  charToRaw() %>% rsvg::rsvg_png(file="Pflow_chart.png")
+  charToRaw() %>% rsvg::rsvg_png(file="figures/Pflow_chart.png")
 ```
@@
 # standards_conc_percentages_plot
-# ggsave("standards_conc_percentages_plot.png")
+# ggsave("figures/standards_conc_percentages_plot.png")
 ```
@@
 #standards_conc_percentages_plot_and_table
-# ggsave("standards_conc_percentages_plot_and_table.png")
+# ggsave("figures/standards_conc_percentages_plot_and_table.png")
 ```
@@
 
-**NOTE:** After brainstorming, Anastasia, Ignacio, and Antonio suggest that an executive summary could be a good strategy for readers who are too busy to read, and to provide a good idea of what readers who do plan to read will find in this document.
-
-_If we plan to write an executive summary, it should ease readers into the report. The full report is long so this may be a useful 2-3 page summary addition._
+Modern societies provide people with varying degrees of mobility and accessibility, from low wage Amazon workers’ depending on infrequent bus service, to Jeff Bezos taking private flights into space.
+Societies implicitly tolerate different levels of inequality in access, though these standards are occasionally made legible through policy standards.
+For example, U.S.
+transit agencies are required to set and adhere to maximum levels of disparity in service quality between minority and non-minority routes as defined by the U.S.
+Civil Rights Act.
+In contrast to prior reviews on how to measure inequality in transportation systems, this report is concerned with the implied or explicit standards that are used to judge whether the measured inequalities are fundamentally “fair” or unacceptable.
+The present study achieves this objective by scanning the state of academic knowledge in defining “fairness” in the transport domain.
+
+Borrowing from philosophers of justice [@jaggarPhilosophicalChallengesGlobal2009], the report’s authors ask the following questions about existing standards of fairness in practice:
+
+-	**Where and when** are equity standards applied within the research depending on location and time of publication? How are context-specific planning processes changing?
+-	**Who** benefits and is burdened by transportation systems? Along which lines of identity are transportation inequities examined?
+-	**What** are those benefits and burdens of transportation systems? Are they qualified in sidewalk widths, bus frequency, travel times, accessibility, or some other consideration?
+-	**How**: are equity standards and inequities measured?
+Are they infrastructure provision thresholds, acceptable pollution levels, or some other kind of cut off?
+How are these studies born out of broader notions of justice, i.e.
+disability rights, environmental justice, egalitarianism, or something else?
+
+**Methods**
+The authors conduct a scoping review of the literature.
+The review involves entering combinations of equity-related keywords into academic databases to identify relevant studies.
+The team initially identified 6,832 references from three diverse databases that they shrunk to a final sample using a four-step process:
+
+1.	Removing 611 duplicate studies.
+2.	Having a team of trained students review the remaining titles and abstracts for relevance, with 4,511 studies excluded for not sufficiently addressing transportation equity. 
+3.	Reviewing the full-texts of the remaining 1,711 studies.
+At least two trained students reviewed each document, with authors breaking any ties.
+This resulted in 1,545 additional studies being excluded.
+At this stage, studies could be excluded if they did not include any standard for, or conceptualization of, equity.
+4.	Extracting data on the "When", "Where", "Who", "What", and "How" of transport equity in each study.
+This included noting methods used, standards suggested, and inferring conceptualizations of equity that may connect to broader notions of justice from the remaining 165 studies.
+
+**Results**
+**When and Where: the contexts of justice**.
+Most (60%) studies deploying standards for equity are published on the Global North, including cases from Europe, the U.S., Canada, Australia, New Zealand, Japan, and Israel.
+These studies focus on groups seen as disadvantaged within national contexts, with standards related to national policies or issues.
+In contrast, research on the Global South tends to adopt international standards in analyses, such as the U.N.’s Sustainable Development Goals (SDGs).
+Within the Global South, studies most commonly cover Asia, followed by Latin America.
+Studies in the latter region focus disproportionately on mobility barriers and the financial affordability of transportation.
+Studies pertaining to Africa are numbered, though they cover a broader scope of issues including informal transportation, citizens’ accessibility needs, and infrastructure development needs.
+Across the globe, the reviewed studies apply equity standards overwhelmingly to urban contexts (85%) and are concerned with a focused set of issues around.
+Studies on rural contexts are broader in the topics examined, from the equity of ferry service in the Philippines to the impacts of road construction on remote Amazonian communities.
+
+**Who: the subjects of justice.** Income groups are studied most, followed by age groups, and people with disabilities.
+People with low incomes generally experience the least benefits and greatest burdens of transportation systems, including regarding costs and pollution exposure.
+Studies examining equity with respect to age demonstrate how older adults and children face different levels of access to key destinations due to differences in their capabilities.
+Many studies apply universal design guidelines and laws as equity standards to evaluate the equity of transportation systems for people with disabilities.
+Finally, researchers also frequently develop composite indices of vulnerability to identify neighbourhoods with high concentrations of multiple vulnerable populations.
+This approach is common when examining environmental justice impacts.
+
+<!--We suggest: Who and What: the tools of mobility. I like our more-->
+**What: the tools of mobility.** Most studies deploying equity standards apply them to either transit (37%) or pedestrian modes (25%).
+Many transit analyses are multimodal in nature, comparing the access transit provides against the access automobiles provided as an implicit standard for comparison.
+Others combine transit with non-auto modes to examine access for households without vehicles.
+These studies focus on delivering equitable mobility, sometimes in context of also meeting climate goals.
+However, they often fail to include explicit minimums of service that should be provided to different populations.
+In contrast, rights-based, infrastructure-focused studies apply principles of universal design in the study of transport systems and related infrastructure.
+
+<!--we suggest: What: the benefits of mobility and accessibility. I like ours more; there is a distinction. -->
+**What: the benefits of mobility.** Studies measure either movement (i.e., trips taken, trip quality, etc.) or the potential for movement (i.e.
+accessibility to specific destinations, or generally), with most studies focused on the latter.
+Over a quarter of the accessibility focused papers do not focus on access to specific destinations, but instead on the accessibility of the trip (i.e., level of service, barriers of infrastructure) or bundles of destinations.
+Studies that center particular destinations that should be reached/accessed include: employment, healthcare, grocery stores, and greenspace.
+Access to community support or places of worship, as well as childcare activities are less common in the reviewed literature.
+
+<!--we suggest: "How": concepts of justice and standards of fairness. I like ours more - there is a distinction. -->
+**How: concepts and standards of justice.** The most common standards are defined using access to opportunities cut offs (37.2%), opportunity standards, or thresholds related to the perceptions or activities of specific populations (35.8%), population standards.
+The latter often includes standards informed by public health, such as the number of physical activities undertaken in a week and perceived life satisfaction.
+Opportunity standards are evaluated through horizontal equity and spatial equity lenses, while population measures are evaluated through well-being frameworks.
+Another category that emerged in the literature are infrastructure-based standards, which include adherence to universal design guidelines or specific benchmarks for road network characteristics; they are less common in the literature comparatively and are often framed in rights-based terms, particularly as they relate to achieving universal design.
+But infrastructure standards around sustainable infrastructure (i.e.
+block lengths or distance between transit stops) have also been evaluated through vertical, horizontal and spatial lenses.
+The distribution of studies with respect to the kinds of standards deployed, and the broader concepts of justice used, are presented in the Figure below:
+
+![The proportion of equity standards (vertical axis) within each type of equity conceptualization (horizontal axis) category.](figures/standards_conc_percentages_plot_and_table.png)
+
+**Calls to Action**
+
+From what does and does not exist in the state of knowledge today, the authors challenge practitioners, researchers, advocates, and government to commit to five calls to action:
+
+<!--we suggest:
+Call 1: The need for explicit conceptualizations of equity to ground standards. I like Matts!-->
+Call 1: Underpin standards in rigorous concepts of justice.
+
+The standards of equity used in research are often weakly linked to any legally, intellectually, or publicly accepted conceptualizations of justice.
+Many standards are arbitrary.
+For example, the 15-minute city sets a threshold that is easy to remember, but is this really what is needed to achieve sufficient sustainable mobility for urban residents, as opposed to a 10 or 20 minute city?
+And what conceptualization of human need should determine what services are essential enough to need to be within this threshold, and for whom?
+
+<!--we suggest:
+Call 2: The need for creative systems-thinking approaches to measuring (in)equities I like Matts!-->
+Call 2: Develop creative methods for systems-thinking approaches to equity.
+
+Mixed-methods research can help government set meaningful standards by better linking quantitative thresholds with qualitative assessments of well-being, freedom of movement and perceived access, among other dimensions.
+Yet most of the reviewed research is purely quantitative or, to a much lesser degree, purely qualitative.
+The constant deployment of metrics Lorenz Curves of Gini coefficients without research on where such measures need to be to improve quality of life (i.e., a standard), will not help governments with limited resources achieve the best possible equity outcomes within their respective contexts.
+Mobilizing Justice’s Priority Populations working group was inaugurated to advance new research in this area.
+
+<!-- We suggested:
+Call 3: Data availability as a matter of justice-->
+Call 3: Data availability as a matter of justice.
+
+Several areas of critical importance, such as access to care and community, are under-researched, and the authors suspect this relates to lack of data on these kinds of destinations.
+Furthermore, data issues prevent the application of many methods of equity analyses in the Global South.
+The production, availability, transparency, and ownership of data is a matter of justice on which governments and NGOs should act.
+In the Canadian context, Mobilizing Justice’s National Survey working group will advance data practices by conducting the first national survey on transportation poverty.
+
+<!-- We suggested:
+Call 4: Direct and explicit links between standards and experienced outcomes. Sure I like Matt's more!-->
+Call 4: Develop more direct and explicit links between standards and lived experiences.
+
+Linking standards to lived experiences and individual outcomes can help increase their likelihood of adoption by governments.
+Further, they can help decision makers understand the role of mobility in improving policy outcomes in other domains like nutrition, physical and mental health, education, and human development.
+These links can help make the standards used in research and policy more comprehensible and relatable to community members and people with lived experience navigating unequal mobility systems.
+Mobilizing Justice launched it’s Data Driven Standards working group to advance research in this area.
+
+<!-- We suggested:
+Towards evaluations of interventions and policies in and across communities-->
+Call 5: Rigorously evaluate interventions and policies.
+
+Approximately 10% of the reviewed studies assessed interventions.
+Delivering new equity-focused policies and programs that do not actually advance equity harms communities, metastasizes mistrust, and delegitimates future efforts to promote transportation justice.
+Furthermore, policies that benefit one group of people in one context may not benefit other groups in other situations.
+A deep breadth of equity evaluations is needed to avoid failure and guide policymakers towards interventions that will work for the communities they serve.
+Mobilizing Justice launched its Innovative Pilots working group to begin filling this gap, through further work is needed.
 
@@
 
-_This portion may be helpful to guide readers to desired sections, if they are pressed for time or interested in only particular parts of the report. For instance:_
-
-_- If you're interested in why we wrote this report read the "Introduction."_  
-_- Curious about the conceptual framework for transport justice? Read the framework in "A framework to analyze questions of justice"_  
-_- Want to know about our calls for action to move transport equity analysis and standard-setting substantively towards questions of transport justice?
-Read the "Moving forward: concluding remarks and calls for action".
-This section is geared to both researchers and decision-makers._
-
-\newpage
+If you're interested in the motivation behind this report, read the first section, [Justice in transportation?](#sect1)
+
+Keen to jump straight to a description of the conceptual framework for transport justice? See the second section, particularly the subsection: [A framework to analyze questions of justice](#sect2.2).
+
+Interested in the methodological details of the literature review? See [section three](#sect3) and extended information in the [Appendix](#sect8)
+
+Care more about the results and the critical appraisal of the results from the review? See sections [four](#sect4) and [five](#sect5), respectively.
+
+Want to know about our calls for action to move transport equity analysis and standard-setting substantively towards questions of transport justice?
+Read the sixth section [Moving forward: calls for action](#sect6) and the final concluding section [Final remarks](#sect7).
+These sections are geared to both researchers and decision-makers.
 
 # Glossary
-_This section may be helpful for readers to clarify our definitions of these terms so they can more easily follow the text and as a quick reference:_
-
-- **Fairness**: is achieved when a collective and individuals are treated equally and without bias as defined by their morals and values. (*maybe add this into the intro??*)
+
+- **Fairness**: is achieved when a collective and individuals are treated equally and without bias as defined by their morals and values.
 - **Equity**: the distribution of benefits and burdens of things among the population, with a particular emphasis on those with the least advantage and quality in their outcomes.
@@
 The process towards justice is fluid and can be identified through questions that answer what is the context, subject and object of justice, and what is the rationale for justice.
-- **The temporal-spatial context of justice**: the answers to "When?" and "Where?"
+- **The temporal-spatial context of justice**: the answers to "Where?" and "When?".
+What regions, within what countries, spatial contexts, point of time, and processes of planning are impacting transport systems?
 - **The subjects of justice**: the answer to "Who?" is the focus of justice.
@@
 - **The methods for justice determination**: the answer to "How?" is justice defined and markers towards it tracked.
-- **The rationale to justice**: the "Why?". Connects with the context, subject, object and methods used for justice determination.
+- **The rationale to justice**: the "Why?". The rationale connects with the context, subject, object and methods used for justice determination.
 - **Equity standards**: are a tool for answering this distributive question: how do we allocate burdens and benefits and to whom?
-Standards are s that when operationalized effectively define what is fair.
+Standards are something that when operationalized effectively define what is fair.
 - **Accessibility**:  
@@
 Etc.,
+<!--NOTE: not too sure how many of these terms we need.. or if we need a glossary at all. Perhaps we can ask peer-reviewers about this need.-->
 
@@
 
-# Justice in transportation?
+# Justice in transportation? {#sect1}
 
@@
 
-# Setting the stage
+# Setting the stage {#sect2}
 
@@
 
-## Justice, equity, fairness, and standards
+## Justice, equity, fairness, and standards {#sect2.1}
 
@@
 
-## A framework to analyze questions of justice
+## A framework to analyze questions of justice {#sect2.2}
 
@@
 
-# Scanning the lay of the land: Methods
+# Scanning the lay of the land: Methods {#sect3}
 
@@
 An initial limited search of Web of Science (WoS) Core Collection was undertaken to identify key documents This collection contains documents in journals, conference proceedings, and books published all over the world.
-Separate sarches using the terms 'transportation' and 'equity' were generated.
+Separate searches using the terms 'transportation' and 'equity' were generated.
 From these searches, we examined the text contained in the titles and abstracts of relevant papers, the index terms used to describe the papers, and subject heading searches when available.
@@
 
-1.
+2.
 The second step (green box in @fig-fig1) included scanning all full-text papers which passed step 1.
@@
 
-1.
+3.
 In the last step, a data extraction template for each record was filled by one reviewer (purple box in @fig-fig1).
@@
 
-# The lay of the land: a summary of findings
+# The lay of the land: a summary of findings {#sect4}
 
@@
 year_cont_plot
-ggsave("year_cont_plot.png")
+ggsave("figures/year_cont_plot.png")
 ```
@@
 pop_plot
-ggsave("pop_plot.png")
+ggsave("figures/pop_plot.png")
 ```
@@
 mode_plot
-ggsave("mode_plot.png")
+ggsave("figures/mode_plot.png")
 ```
@@
 
-In terms of papers that study specific destinations, the activities most commonly studied include healthcare services (`r (dest_count$count[6]/sum(dest_count$count)) |> percent()`), followed by employment (`r (dest_count$count[5]/sum(dest_count$count)) |> percent()`).
+In terms of papers that study specific destinations, healthcare services (`r (dest_count$count[6]/sum(dest_count$count)) |> percent()`) and employment (`r (dest_count$count[5]/sum(dest_count$count)) |> percent()`) are activities commonly studied.
 Papers that exclusively focus on healthcare typically originate from the healthcare planning literature, and look to inform planners about disparities in services and corrective actions.
@@
 standards_conc_percentages_plot_and_table
-ggsave("standards_conc_percentages_plot_and_table.png")
+ggsave("figures/standards_conc_percentages_plot_and_table.png")
 ```
@@
 
-# A critical appraisal of the lay of the land
+# A critical appraisal of the lay of the land {#sect5}
 
@@
 Most papers within our corpus (`r ((case_cont_count$count[4] + case_cont_count$count[5]+ case_cont_count$count[6])/sum(case_cont_count$count)) |> percent()`) focus on case studies in the Global North.
-Though their subject matter is varied, their spatial context mainly pertains to North American and Europe, and thus more often than not, in conversation with more developed and formal government transport planning apparatuses and technologies [e.g., planning for equitable high-speed rail, as in @monzonEfficiencySpatialEquity2013; or autonomous vehicle technology, as in @eppenbergerOpportunitySharedAutonomous2021; or on the public consultation processes, as in @reddySafeguardingMinorityCivil2010]).
+Though their subject matter is varied, their spatial context mainly pertains to North American and Europe, and thus more often than not, in conversation with more developed and formal government transport planning apparatuses and technologies [e.g., planning for equitable high-speed rail, as in @monzonEfficiencySpatialEquity2013; autonomous vehicle technology, as in @eppenbergerOpportunitySharedAutonomous2021; or on the public consultation processes, as in @reddySafeguardingMinorityCivil2010]).
 
@@
 Literature from the Global North is concerned with particularly disadvantaged groups and creating indicators that may be used to guide the remedy of formal processes that ensure access to societal benefits e.g., [@cuiSpatialAccessPublic2020].
-Global North countries often surpass international sufficiency thresholds related to transport and public health (e.g., @carrierApplicationThreeMethods2014 measures NO2 road-side air pollution and find no levels are below the WHO guidelines) but advocate for the reduction in disproportional impact of air pollution on lower-income groups.
+Global North countries often surpass international sufficiency thresholds related to transport and public health (e.g., @carrierApplicationThreeMethods2014 measures NO_{2} road-side air pollution and find no levels are below the WHO guidelines) but advocate for the reduction in disproportional impact of air pollution on lower-income groups.
 These international guidelines, while perhaps being relevant at a time of lower transport-related development, are no longer relevant for some areas in the Global North.
@@
 
-So what's missing from the "who?" question?
-We believe, more nuisances perspectives that capture the "who" of inequities is needed.
+So what's missing from the "Who" question?
+We believe, more nuisances perspectives of inequities are needed.
 We must go beyond the focus on the low-income transit rider; positioning intersectionality of socio-economic and mode-use to create community-based definitions of equity and tailored standards are critical.
@@
 
-# Moving forward: calls for action
+# Moving forward: calls for action {#sect6}
 
@@
 
-# Final remarks
-<!-- HERE IS AN OPPORTUNITY HERE TO CIRCLE BACK TO BEZOS SUBORBITAL FLIGHT, TO CLOSE THE REPORT WITH A REMINDER OF WHY IT IS IMPORTANT TO LEARN ABOUT JUSTICE.-->
-
+# Final remarks {#sect7}
 We began this report with the history-making suborbital flight of Jeff Bezos and his fellow passengers, all of them very wealthy people.
@@
 
+# Acknowledgements
+This report would not have been possible without the help and inspiration of countless others not let listed.
+In the early stages of the methodological design preparation of this report, the support from a team of literature review specialists at University of Toronto was invaluable; particularly the Research Services and Liaison Librarians.
+These experts provided guidance in cornering down the topic, identifying guidance material on literature synthesis and relevant databases, and in introducing us to the world of knowledge synthesis.
+Though our path diverged from a traditional scoping review, choosing to place and discuss the reviewed equity literature in connection to notions of justice, the protocol published on September 2022 (found [here](https://osf.io/rsb92)) was followed with only minor changes -- a reduction in the databases searched due to resource constraints.
+
+Furthermore, thank you for the hours of support from undergraduate and graduate research assistants who poured through title and abstract screening and helped determine initial stages of full-text eligibility.
+Reviewing such a quantity of literature was new to many of us, we are deeply thankful for your help: Jan Domalaon, Dancel Gayle, Eryn Maloney, Mahdis Moghadasi, Anika Munir, and Subaita Refaaf.
+
 \newpage
 
-# Appendix
+# Appendix {#sect8}
 

### 2023-11-15 just-transportation/just-transportation.qmd
First polished version: PRE-RELEASE VERSION

--- 2023-11-15 Manuscript/Manuscript.qmd
@@
 ---
-title: "Searching for standards of fairness in the transportation justice literature: A scoping review"
+title: "Searching for standards of fairness in the transportation justice literature"
 subtitle: "Searching for standards of fairness"
+format:
+  mj-report-pdf: 
+    keep-tex: true
+doctype: "Report (Draft)"
+
+# Author information. You can add or delete authors here
 author:
@@
     affiliations: 
-      - name: University of Toronto Scarborough
+      - name: University of Toronto
         department: Department of Geography and Planning
@@
     affiliations: 
-      - name: University of North Carolina Chapel Hill
+      - name: The University of North Carolina at Chapel Hill
         department: Department of City and Regional Planning
@@
       given: Antonio
-      family: Paez
+      family: Páez
     affiliations: 
@@
     email: paezha@mcmaster.ca
-bibliography: All_papers_in_extracted.bib
-format: docx
+
+# Date: default is last modified but could be a string.  
+date: last-modified
+
+# Write the acknowledgments; remember to use a tab at the beginning of each paragraph
+acknowledgements: | 
+  This report would not have been possible without the help and inspiration of many individuals not listed as authors.
+In the early stages of this report we received invaluable help from specialists at the University of Toronto as we learned about various approaches to literature reviews, particularly the Research Services and Liaison Librarians.
+These experts provided guidance in cornering down the topic, identifying guidance material on literature synthesis and relevant databases, and eased us into the world of knowledge synthesis.
+Though our path diverged from a traditional scoping review, the protocol published on September 2022 (found [here](https://osf.io/rsb92)) was followed with only minor changes, with only a reduction in the databases searched due to resource constraints.
+We also wish to acknowledge the many hours of work by undergraduate and graduate research assistants who poured through title and abstract screening to help determine full-text eligibility at the earliest stages of our search.
+The volume of literature initially reviewed was greater than any of us had experienced before, but important to be as inclusive as possible.
+For this support, we are deeply thankful to Jan Domalaon, Dancel Gayle, Eryn Maloney, Mahdis Moghadasi, Anika Munir, and Subaita Refaaf.
+  The views and opinions expressed here, as well as any equivocations and/or awkwardness are the authors' and the authors' alone.
+
+# Place the cover image and the toc image in folder `images`
+cover_image: cover-picture.png
+cover_image_caption: "Source: Marco Verch (https://www.flickr.com/photos/160866001@N07/50597690137/)"
+toc_image: toc-picture.png
+toc_image_caption: "Source: Progressive Charlestown (https://www.progressive-charlestown.com/2020/01/amazon-drivers-die-for-you.html)"
+# Information for generating citation
+citation:
+  type: report
+  doi: "10.somenumber/document.00010"
+  issued: last-modified
+  url: https://example.com/summarizing-output
+  number: "MJ-0001"
+bibliography: bibliography.bib
 ---
@@
 ```{r, eval=FALSE}
-Pflow_chart %>%
-  DiagrammeRsvg::export_svg() %>% 
-  charToRaw() %>% rsvg::rsvg_png(file="figures/Pflow_chart.png")
+Pflow_chart |>
+  DiagrammeRsvg::export_svg() |> 
+  charToRaw() |> rsvg::rsvg_png(file="figures/Pflow_chart.png")
 ```
@@
                               id=rep(Consensus_only_analysis$`Covidence #`, lengths(df_conc)),
-                              conc=unlist(df_conc)) %>% 
+                              conc=unlist(df_conc)) |> 
                               mutate(conc = ifelse(conc == "Inequitable exposure to externalities", "Ineq. externalities", conc),
@@
 
-NUM_standards_conc <- standards_conc %>% group_by(id) %>% summarize(count = n())
-NUM_ENTRIES_standards_conc <- NUM_standards_conc$count %>% sum()
+NUM_standards_conc <- standards_conc |> group_by(id) |> summarize(count = n())
+NUM_ENTRIES_standards_conc <- NUM_standards_conc$count |> sum()
 
@@
 ```{r, message=FALSE}
-standards_conc <- standards_conc %>% 
-  group_by(conc, std_mcat) %>% summarize(total = n())
-
-standards_conc <- standards_conc %>%
+standards_conc <- standards_conc |> 
+  group_by(conc, std_mcat) |> summarize(total = n())
+
+standards_conc <- standards_conc |>
   mutate(total_perc_in_conc = total/sum(total),
          total_perc_in_all = total/NUM_ENTRIES_standards_conc)
-sums_concs <- standards_conc %>% group_by(conc) %>% summarize(
+sums_concs <- standards_conc |> group_by(conc) |> summarize(
   total_perc_in_all = sum(total_perc_in_all),
-  total_perc_in_conc = 1) %>% mutate(std_mcat = "TOTALS")
+  total_perc_in_conc = 1) |> mutate(std_mcat = "TOTALS")
 
@@
 
-sums_std_mcat <- standards_conc %>% group_by(std_mcat) %>% summarize(total_perc_in_all = sum(total_perc_in_all),
-  total_perc_in_conc = 1) %>% mutate(conc = "TOTALS")
+sums_std_mcat <- standards_conc |> group_by(std_mcat) |> summarize(total_perc_in_all = sum(total_perc_in_all),
+  total_perc_in_conc = 1) |> mutate(conc = "TOTALS")
 
@@
 ```{r}
-set_flextable_defaults(font.size= 7,
+set_flextable_defaults(font.size= 6,
                        font.family = NULL,
@@
   # compose(totals_conc_table_flex,
-  #         i = 1, j = c(1,2), part = "header", value = as_paragraph(c("Conceptualizations", "Conceptualizations"))) %>%
-  prepend_chunks(flextable(totals_standards_conc_table %>% select(-c("Standards"))),
-                 part = "body", i = c(1,2,3,4,5), j = 1, as_b(c("Vertical equity: ", "TR exclusion: ", "Wellbeing: ", "Spatial equity: ", "Horizontal equity: "))) %>% 
-  prepend_chunks(part = "body", i = c(1,2,3,4), j = 2, as_b(c("Ineq. externalities: ", "Rights: ", "Emerg. theories: ", "Utilitarian: "))) %>% 
-  # merge_h(part="header", i=1) %>% #header merge
-  merge_at(i=5, j=1:2) %>%
-  # align( align="center", part="header") %>%
-  delete_part("header") %>%
-  border_outer() %>%
-  border_inner_h() %>%
+  #         i = 1, j = c(1,2), part = "header", value = as_paragraph(c("Conceptualizations", "Conceptualizations"))) |>
+  prepend_chunks(flextable(totals_standards_conc_table |> select(-c("Standards"))),
+                 part = "body", i = c(1,2,3,4,5), j = 1, as_b(c("Vertical equity: ", "TR exclusion: ", "Wellbeing: ", "Spatial equity: ", "Horizontal equity: "))) |> 
+  prepend_chunks(part = "body", i = c(1,2,3,4), j = 2, as_b(c("Ineq. externalities: ", "Rights: ", "Emerg. theories: ", "Utilitarian: "))) |> 
+  # merge_h(part="header", i=1) |> #header merge
+  merge_at(i=5, j=1:2) |>
+  # align( align="center", part="header") |>
+  delete_part("header") |>
+  border_outer() |>
+  border_inner_h() |>
   autofit()
@@
   # compose(totals_standards_table_flex,
-  #         i = 1, j = c(1), part = "header", value = as_paragraph(c("Standards"))) %>% 
-  prepend_chunks(flextable(totals_standards_conc_table %>% select("Standards")),
-                 part = "body", i = c(1,2,3,4), j = 1, as_b(c("Population: ", "Opportunity: ", "Infrastructure: ", "Environmental +: "))) %>%
-  merge_at(i= 4:5, j= 1) %>%
-  delete_part("header")%>%
-  border_outer() %>%
-  border_inner_h() %>%
+  #         i = 1, j = c(1), part = "header", value = as_paragraph(c("Standards"))) |> 
+  prepend_chunks(flextable(totals_standards_conc_table |> select("Standards")),
+                 part = "body", i = c(1,2,3,4), j = 1, as_b(c("Population: ", "Opportunity: ", "Infrastructure: ", "Environmental +: "))) |>
+  merge_at(i= 4:5, j= 1) |>
+  delete_part("header")|>
+  border_outer() |>
+  border_inner_h() |>
   autofit()
@@
   scale_y_discrete(limits=(totals_standards_conc$conc)[order(totals_standards_conc$group)] ) +
-  geom_text(aes(label = total_perc_in_all %>% percent(accuracy = 0.1)), color = "white", size = 2.5) +
+  geom_text(aes(label = total_perc_in_all |> percent(accuracy = 0.1)), color = "white", size = 2.5) +
   geom_vline(xintercept=4.5, linewidth=1.5, color = "#191970") +
@@
 
-We need a super nice cover.
-
-\newpage
-
 # Executive Summary
@@
 Modern societies provide people with varying degrees of mobility and accessibility, from low wage Amazon workers’ depending on infrequent bus service, to Jeff Bezos taking private flights into space.
-Societies implicitly tolerate different levels of inequality in access, though these standards are occasionally made legible through policy standards.
+Societies implicitly tolerate different levels of inequality in access, though these levels are occasionally made legible through policy standards.
 For example, U.S.
@@
 **Methods**
-The authors conduct a scoping review of the literature.
+
+The authors conducted a scoping review of the literature.
 The review involves entering combinations of equity-related keywords into academic databases to identify relevant studies.
-The team initially identified 6,832 references from three diverse databases that they shrunk to a final sample using a four-step process:
-
-1.	Removing 611 duplicate studies.
-2.	Having a team of trained students review the remaining titles and abstracts for relevance, with 4,511 studies excluded for not sufficiently addressing transportation equity. 
-3.	Reviewing the full-texts of the remaining 1,711 studies.
-At least two trained students reviewed each document, with authors breaking any ties.
+The team initially discovered 6,832 references from three diverse databases that they shrunk to a final corpus of studies using a four-step process:
+
+1.	A total of 611 duplicate studies were removed.  
+2.	The researchers, supported by a team of trained research assistants, reviewed the remaining titles and abstracts for relevance, with 4,511 studies excluded for not sufficiently addressing transportation equity.
+3.	The full-texts of the remaining 1,711 studies were reviewed.
+At least two trained research assistants reviewed each document, with one of the the main researchers (listed as authors) breaking any ties.
 This resulted in 1,545 additional studies being excluded.
-At this stage, studies could be excluded if they did not include any standard for, or conceptualization of, equity.
-4.	Extracting data on the "When", "Where", "Who", "What", and "How" of transport equity in each study.
-This included noting methods used, standards suggested, and inferring conceptualizations of equity that may connect to broader notions of justice from the remaining 165 studies.
+At this stage, studies could be excluded if they did not include any standard for, or conceptual grounding for, equity.
+4.	The researchers examined the final corpus of 165 studies to extract data on the "When", "Where", "Who", "What", and "How" of transport equity in each study.
+This included annotating the methods used, standards suggested, and inferring conceptualizations of equity that may connect to broader notions of justice.
 
 **Results**
+
 **When and Where: the contexts of justice**.
-Most (60%) studies deploying standards for equity are published on the Global North, including cases from Europe, the U.S., Canada, Australia, New Zealand, Japan, and Israel.
-These studies focus on groups seen as disadvantaged within national contexts, with standards related to national policies or issues.
-In contrast, research on the Global South tends to adopt international standards in analyses, such as the U.N.’s Sustainable Development Goals (SDGs).
+Most (60%) studies that deployed standards for equity stem from the Global North, including cases from Europe, the U.S., Canada, Australia, New Zealand, Japan, and Israel.
+The focus of these studies is on groups seen as disadvantaged within their respective national contexts, and strive to relate standards to national policies or issues.
+In contrast, research stemming from the Global South tends to adopt international standards in analyses, such as the U.N.’s Sustainable Development Goals (SDGs).
 Within the Global South, studies most commonly cover Asia, followed by Latin America.
@@
 Studies pertaining to Africa are numbered, though they cover a broader scope of issues including informal transportation, citizens’ accessibility needs, and infrastructure development needs.
-Across the globe, the reviewed studies apply equity standards overwhelmingly to urban contexts (85%) and are concerned with a focused set of issues around.
+Across the globe, the studies reviewed apply equity standards overwhelmingly to urban contexts (85%) and are concerned with a focused set of issues around the need for and the impacts of daily mobility.
 Studies on rural contexts are broader in the topics examined, from the equity of ferry service in the Philippines to the impacts of road construction on remote Amazonian communities.
@@
 <!--We suggest: Who and What: the tools of mobility. I like our more-->
-**What: the tools of mobility.** Most studies deploying equity standards apply them to either transit (37%) or pedestrian modes (25%).
-Many transit analyses are multimodal in nature, comparing the access transit provides against the access automobiles provided as an implicit standard for comparison.
+**What: the tools of mobility.** Most studies that deploy equity standards apply them to either transit (37%) or pedestrian modes (25%).
+Many transit analyses are multimodal in nature, comparing accessibility by transit to accessibility by automobiles as an implicit standard.
 Others combine transit with non-auto modes to examine access for households without vehicles.
 These studies focus on delivering equitable mobility, sometimes in context of also meeting climate goals.
-However, they often fail to include explicit minimums of service that should be provided to different populations.
+However, they often fail to include explicit minima of service that should be provided to different populations.
 In contrast, rights-based, infrastructure-focused studies apply principles of universal design in the study of transport systems and related infrastructure.
@@
 accessibility to specific destinations, or generally), with most studies focused on the latter.
-Over a quarter of the accessibility focused papers do not focus on access to specific destinations, but instead on the accessibility of the trip (i.e., level of service, barriers of infrastructure) or bundles of destinations.
-Studies that center particular destinations that should be reached/accessed include: employment, healthcare, grocery stores, and greenspace.
-Access to community support or places of worship, as well as childcare activities are less common in the reviewed literature.
+Over a quarter of the papers that focus on accessibility do not consider specific destinations, but instead concentrate on the accessibility of the trip (i.e., level of service, barriers of infrastructure) or bundles of destinations.
+Other studies are centred on particular destinations considered as essential, including employment, healthcare, grocery stores, and greenspace.
+Access to community support or places of worship, as well as childcare activities are far less common in the literature reviewed.
 
 <!--we suggest: "How": concepts of justice and standards of fairness. I like ours more - there is a distinction. -->
-**How: concepts and standards of justice.** The most common standards are defined using access to opportunities cut offs (37.2%), opportunity standards, or thresholds related to the perceptions or activities of specific populations (35.8%), population standards.
+**How: concepts and standards of justice.** The most common standards found in the literature, which we term opportunity standards, are defined using cut offs in access to opportunities (37.2%).
+Almost as common is the formulation of standards that use thresholds related to the perceptions or activities of specific populations (35.8%); we call these population standards.
 The latter often includes standards informed by public health, such as the number of physical activities undertaken in a week and perceived life satisfaction.
-Opportunity standards are evaluated through horizontal equity and spatial equity lenses, while population measures are evaluated through well-being frameworks.
-Another category that emerged in the literature are infrastructure-based standards, which include adherence to universal design guidelines or specific benchmarks for road network characteristics; they are less common in the literature comparatively and are often framed in rights-based terms, particularly as they relate to achieving universal design.
-But infrastructure standards around sustainable infrastructure (i.e.
-block lengths or distance between transit stops) have also been evaluated through vertical, horizontal and spatial lenses.
+Opportunity standards are evaluated through horizontal equity and spatial equity lenses, while population ones are evaluated through well-being frameworks.
+Another category of standards that emerged from the literature are based on infrastructure, and these include adherence to universal design guidelines or specific benchmarks for road network characteristics.
+Standards of thsi kind are comparatively less common in the literature and are often framed in terms of rights, particularly as they relate to achieving universal design.
+Infrastructure standards around sustainable infrastructure (i.e.
+block lengths or distance between transit stops) are also evaluated through vertical, horizontal and spatial lenses.
 The distribution of studies with respect to the kinds of standards deployed, and the broader concepts of justice used, are presented in the Figure below:
@@
 
-From what does and does not exist in the state of knowledge today, the authors challenge practitioners, researchers, advocates, and government to commit to five calls to action:
+The review conducted for this report provides a snapshot of the state of knowledge today in terms of how standards of fairness are developed and deployed.
+The review is telling both in terms of what it reveals about the literature, what is covered and what is not.
+Based on these, the authors challenge practitioners, researchers, advocates, and governments to commit to five calls to action:
 
@@
 
-The standards of equity used in research are often weakly linked to any legally, intellectually, or publicly accepted conceptualizations of justice.
+The standards of equity used in research are often weakly linked to any legally, intellectually, or publicly accepted concepts of justice.
 Many standards are arbitrary.
 For example, the 15-minute city sets a threshold that is easy to remember, but is this really what is needed to achieve sufficient sustainable mobility for urban residents, as opposed to a 10 or 20 minute city?
-And what conceptualization of human need should determine what services are essential enough to need to be within this threshold, and for whom?
+And what conceptualization of human need should determine what services are essential enough to be included within this threshold, and for whom?
 
@@
 
-Mixed-methods research can help government set meaningful standards by better linking quantitative thresholds with qualitative assessments of well-being, freedom of movement and perceived access, among other dimensions.
-Yet most of the reviewed research is purely quantitative or, to a much lesser degree, purely qualitative.
+Mixed-methods research can help governments set meaningful standards by better linking quantitative thresholds with qualitative assessments of well-being, as well as freedom of movement and perceived access, among other facets of transportation.
+Yet most of the research reviewed is purely quantitative or, to a much lesser degree, purely qualitative.
 The constant deployment of metrics Lorenz Curves of Gini coefficients without research on where such measures need to be to improve quality of life (i.e., a standard), will not help governments with limited resources achieve the best possible equity outcomes within their respective contexts.
@@
 Call 3: Data availability as a matter of justice-->
-Call 3: Data availability as a matter of justice.
+Call 3: Making data available is a matter of justice.
 
@@
 The production, availability, transparency, and ownership of data is a matter of justice on which governments and NGOs should act.
-In the Canadian context, Mobilizing Justice’s National Survey working group will advance data practices by conducting the first national survey on transportation poverty.
+In the Canadian context, Mobilizing Justice’s National Survey working group is working to advance data practices by conducting the first national survey on transportation poverty and striving to make data available in the form of Open Data Products.
 
@@
 
-Linking standards to lived experiences and individual outcomes can help increase their likelihood of adoption by governments.
-Further, they can help decision makers understand the role of mobility in improving policy outcomes in other domains like nutrition, physical and mental health, education, and human development.
+Linking standards to lived experiences and individual outcomes can help increase the likelihood of their adoption by governments.
+Further, they can help decision makers understand the role of transportation in improving policy outcomes in other domains like nutrition, physical and mental health, education, and human development.
 These links can help make the standards used in research and policy more comprehensible and relatable to community members and people with lived experience navigating unequal mobility systems.
@@
 
-Approximately 10% of the reviewed studies assessed interventions.
+Approximately 10% of the studies reviewed aimed to assess interventions.
 Delivering new equity-focused policies and programs that do not actually advance equity harms communities, metastasizes mistrust, and delegitimates future efforts to promote transportation justice.
@@
 
-If you're interested in the motivation behind this report, read the first section, [Justice in transportation?](#sect1)
+This is your guide to scanning this document.
+
+If you are primarily interested in the motivation behind this report, read the first section, [Just transportation?](#sect1)
 
@@
 
+\newpage
+
 # Glossary
 
-- **Fairness**: is achieved when a collective and individuals are treated equally and without bias as defined by their morals and values.
+- **Accessibility**: the potential to reach destinations.
+
+<!-- 
+- **Capabilities approach**:  
+-->
+
+- **Equality**: a state where every group is treated identically.
+
 - **Equity**: the distribution of benefits and burdens of things among the population, with a particular emphasis on those with the least advantage and quality in their outcomes.
-- **Justice**: mutually reinforced by the tools of fairness and equity, justice is a multi-dimensional political ideal attained when people "give and receive whatever they are due" [@jaggarPhilosophicalChallengesGlobal2009, pp.
-1-2].
-It ceases to exist when there are persons or groups that are denied "access to the opportunities they need to lead a meaningful and dignified life" [@karnerTransportationEquityTransportation2020, p.
-440].
-The process towards justice is fluid and can be identified through questions that answer what is the context, subject and object of justice, and what is the rationale for justice.
+
+- **Equity standards**: tools for answering the question of how to distribute the burdens and benefits and to whom? When operationalized standards effectively define what is fair.
+
+- **Fairness**: a state of being that is achieved when a collective and individuals are treated equally and without bias as defined by their morals and values.
+
+- **Justice**: A political ideal whereby people give and receive what they are due.
+
 - **The temporal-spatial context of justice**: the answers to "Where?" and "When?".
 What regions, within what countries, spatial contexts, point of time, and processes of planning are impacting transport systems?
+
 - **The subjects of justice**: the answer to "Who?" is the focus of justice.
+
 - **The objects of justice**: the answer to "What?" is the focus of justice; what are the things that characterise (in)justice.
+
 - **The methods for justice determination**: the answer to "How?" is justice defined and markers towards it tracked.
+
 - **The rationale to justice**: the "Why?". The rationale connects with the context, subject, object and methods used for justice determination.
-- **Equity standards**: are a tool for answering this distributive question: how do we allocate burdens and benefits and to whom?
-Standards are something that when operationalized effectively define what is fair.
-- **Accessibility**:  
-- **Capabilities approach**:  
-- **Equality**:  
-- **Mobility**:  
-- **Policy**:  
+
+- **Mobility**: The potential to move.
+
+- **Policy**: A deliberate system of principles to guide decisions and achieve desired outcomes.
+
+<!--
 - **Social exclusion**:  
-- **Space-time convergence**:  
+-->
+
+- **Space-time convergence**: the main product of a transportation system is to reduce the time needed to cover a unit of distance; this effect makes space and time converge.
+
+<!--
 - **Sufficientarianism**:
+
 - **Theories of justice**:
+
 - **Utilitarianism**:  
-Etc.,
-<!--NOTE: not too sure how many of these terms we need.. or if we need a glossary at all. Perhaps we can ask peer-reviewers about this need.-->
+-->
 
@@
 
-# Justice in transportation? {#sect1}
+# Just transportation? {#sect1}
 
@@
 
-The fiery wake of a rocket could be seen ascending, piercing the morning sky above West Texas.
-It was July 20, 2021, and on board was a small group of four passengers that included Jeff Bezos, the founder of Amazon and then the world's richest person [@harwoodBezos2021].
+The fiery wake of a rocket could be seen ascending, piercing the sky above West Texas.
+It was the morning of July 20, 2021, and on board of the rocket was a small group of four passengers that included Jeff Bezos, the founder of Amazon and then the world's richest person [@harwoodBezos2021].
 The mission that day was among the first-ever private suborbital passenger flights, and the adventure (described as "intense" by one of the passengers) lasted a total of 10 minutes and 10 seconds [@harwoodBezos2021].
 In addition to intense, the undertaking was expensive: a seat for the flight had previously been auctioned for no less than $28 million USD [@griffinPerson2021].
-Before even 10:00 am (EDT), when Bezos was back from his excursion, he took time to declare that this was the "[b]est day ever".
+Before it was even 10:00 am (EDT), Bezos was back from his excursion, and he took time to declare that this was the "[b]est day ever".
 To reporters covering the event he said "I want to thank every Amazon employee and every Amazon customer, because you guys paid for all of this" [@johnsonRoadToSpace2021].
 
-Meanwhile, firmly grounded on planet Earth, the employees that Bezos thanked for his suborbital escapade were struggling with some very mundane problems of their own, and none as lofty as conflicting schedules that prevented them from flying in rockets [@griffinPerson2021].
+Meanwhile, firmly grounded on planet Earth, the employees whom Bezos thanked for his suborbital jaunt were struggling with some very mundane problems of their own, and none as lofty as conflicting schedules that prevented them from flying in rockets [@griffinPerson2021].
 According to reports, people employed directly or indirectly by Amazon for warehouse or delivery work had, for years, been treated to "inhumane" conditions [@fungEuropean2018; @scottAmazon2019; @greeneAlgorithm2021], and subjected to surveillance on the job, degrading schedules that drove drivers to urinate in bottles, crushing demands for productivity quotas that led to injury, all the while facing little or no job security.
@@
 In other words, the billionaire was not confronted with questions about what his trip did _for whom_ and _to whom_.
-A less dispassionate observer might have been excused for wondering (possibly aloud) about the basic _fairness_ of a man amassing a nigh unimaginable fortune that allowed him to build and fly his own spaceship, while masses of his employees were treated as throwaway cogs in the vast apparatus of his logistics empire.
+A less dispassionate observer might have been excused for wondering (possibly aloud) about the basic _fairness_ of a man amassing a nigh unimaginable fortune that allowed him to build and fly his own spaceship, while masses of his employees were treated as throwaway cogs in the vast apparatus of his emporium^[Not unlike in the empire of another space billionaire, where things feel like the future except for the lives of workers [@wongTesla2017], and everything soars---including worker injuries [@taylorSpaceX2023].].
 
 The question of _fairness_ is not a simple one.
-Most of us would probably have been stumped to explain it in a precise way just why the above picture was disturbing.
+Most of us would probably have been stumped to explain in a precise way just why the above picture was disturbing.
 _Justice_ is a political ideal based on the principle that individuals should be treated in a _fair_ and _equitable_ manner [@gosslingUrban2016, p.
@@
 As well, the meaning of "justice" would likely differ in yet another society where very few owned most, and most owned very little due to a different illusion (e.g., that wealth equates merit).
-In these two hypothetical cases, elucidating the meaning of "fair" would in all certainty be beyond the dreams of most, since fair would be what organizational structures with power (e.g., the state, the monarch, or the extremely wealthy) said it was.
-In contrast, in democratic societies^[Paraphrasing, democracy is the worst of collective illusions, except for all others.] individuals rights and a collective vision are valued above the whims of the few.
-In such a setting, the task of defining a "just" distribution of the burdens and benefits of things —from income, to roads, to space travel- quickly becomes muddled, encumbered even, by the necessity to pay attention to a multitude of voices, not all of them equally loud.
+In these two hypothetical cases, elucidating the meaning of "fair" would in all certainty be beyond the dreams of most, since fair would be whatever organizational structures with power (e.g., the state, the monarch, or the extremely wealthy) said it was.
+In contrast, in democratic societies^[Paraphrasing, democracy is the worst of collective illusions, except for all others.] individual rights tempered by a collective vision are valued above the whims of the few.
+In such a setting, the task of defining a "just" distribution of the burdens and benefits of things---from income, to roads, to space travel---quickly becomes muddled, encumbered even, by the necessity to pay attention to a multitude of voices, not all of them equally loud.
 
 The rocket that took Bezos to the edge of space is a somewhat rare example of a transportation technology, a tool of space-time convergence.
-By enabling movement at very high speeds, rockets might—one day, at some indeterminate point in the future—prove essential to the expansion of the human species beyond our home planet.
+By enabling movement at very high speeds, rockets might---one day, at some indeterminate point in the future---prove essential to the expansion of the human species beyond our home planet.
 But in the present moment, the benefits of a private suborbital flight (e.g., the joy of movement, the sense of adventure, the awe of seeing Earth from space) are for a few, whereas the burdens (e.g., the use of non-renewable resources, the climate-altering emissions) affect us all, and not evenly at that.
-The benefits of public transportation, a much more common transportation technology, are for most, but in many cases we have penny-pinched these systems while concentrating the costs on those who have less [@allenSuburbanization2020; @kaeorueanAnalysis2020].
-In a plutocratic society, those with most can (and often do argue^[See the so-called [Techno-optimist Manifesto](https://a16z.com/the-techno-optimist-manifesto/)]) that this distribution of burdens and benefits is fair since both benefits and burdens are earned.
+The benefits of public transportation, a much more common transportation technology, are for most, but in many cases we have penny-pinched these systems, leaving them moribund, while concentrating the costs on those who have less [@allenSuburbanization2020; @kaeorueanAnalysis2020].
+In a plutocratic society, those with most can (and often do argue^[Either directly, as in the so-called Techno-optimist Manifesto [@andreessenTechno2023], or through their mouthpieces [@mcardleBillionaires2021]]) that this distribution of burdens and benefits is fair since both benefits and burdens are earned.
 The fact that the likes of Bezos do travel to space is proof that the likes of Bezos are due those trips.
 In a democratic society, the members of the collective might actually agree that large rewards (e.g., space flight) must be offered to highly qualified individuals (e.g., Bezos) to entice them to take important responsibilities (e.g., founding and leading Amazon).
-The fact that some must give up jobs because it takes them too long to reach them by public transportation would constitute proof that those people should have studied more, been earlier risers, worked harder [@SpurrWaiting2015; @greismanLittle2017].
-Again, the members of the collective might this is a fine state of affairs, having come to this conclusion of their own accord or after being persuaded by billionaires.
+The fact that some must give up jobs because it is too time consuming to reach them by public transportation would constitute proof that those people should have studied more, been earlier risers, worked harder [@SpurrWaiting2015; @greismanLittle2017].
+Again, the members of the collective might believe that this is a fine state of affairs, having come to this conclusion of their own accord or after being persuaded by billionaires.
 Or, contrariwise, the members of the collective might decide that this state of affairs is _unjust_: the _values_ of those in charge of defining what is "fair" matter.
@@
 In response to this focus, a lively and rapidly growing literature has emerged on the topic (see _inter alia_ [@martensTransport2016; @diciommoTransport2017; @guoSystematicOverviewTransportation2020; @karnerFrom2020; @vecchioTransportEquityLatin2020; @pereiraTransportationEquity2021; @vanweeEvaluating2021; @zhangLiterature2021; @desjardinsExaminingEquityAccessibility2022; @karnerAdvancesPitfallsMeasuring2023]).
-A cynical rationale for this interest could be that keeping track of objective and perceived inequalities can serve _at the very least_ as a gauge of social discontent (as Chilean authorities discovered when the increase in Santiago Metro's subway fare sparked a period of massive demonstrations and unrest in October 2019 motivated by country-wide social inequalities; [@BBCNews2019; @diazpabonInequality2021]).
+A cynical rationale for this interest could be that keeping track of objective and perceived inequalities can serve _at the very least_ as a gauge of social discontent (as Chilean authorities discovered to their woe when an increase in Santiago Metro's fares became a flashpoint for social inequalities, and sparked a period of massive demonstrations and unrest in October 2019; [@BBCNews2019; @diazpabonInequality2021]).
 More optimistically, in democratic systems, tracking objective and perceived inequalities could be of interest for governing bodies to respond to popular demands for fairness.
 Several challenges arise when approaching this endeavor.
-The notorious complexity of transportation systems is one challenge: transport systems simultaneously move people, goods, and information, and emerging technologies and service models can swiftly change the balance of benefits and burdens among a population [@guoSystematicOverviewTransportation2020].
+The notorious complexity of transportation systems is one of them: transport systems simultaneously move people, goods, and information.
+Emerging technologies and service models can swiftly change the balance of benefits and burdens among a population [@guoSystematicOverviewTransportation2020], turning both users and service providers into digital rentiers [@birchBig2022].
 The benefits and burdens of transportation systems are diffuse over space and time.
@@
 
-While commendable, the strategies are too vague, which makes it possible to implement them in a myriad ways, either genuinely to comply with the spirit of justice, or else performatively to deceive it [@mcculloughPerformative2023].
+While commendable, the strategies are too vague, which means it is possible to implement them in a myriad ways, either genuinely to comply with the spirit of justice, or else performatively to deceive it [@mcculloughPerformative2023].
 Some relevant questions include: how much should the adverse effects be reduced?
 To zero?
-Or to some tolerable level of adversity?
+Or to some tolerable level of adversity greater than zero?
 What should that level be?
@@
 
-Supporting the creation of (more) just transportation systems involves understanding the production, distribution, and management of transportation benefits and costs; how they are distributed; and what values are implemented (and by whom) in the form of standards [@pereiraDistributiveJusticeEquity2017a; @shellerMobilityJusticePolitics2018; @pereiraTransportationEquity2021].
+Supporting the creation of (more) just transportation systems involves understanding the production and management of transportation benefits and costs; how they are distributed; and what values are implemented (and by whom) in the form of standards [@pereiraDistributiveJusticeEquity2017a; @shellerMobilityJusticePolitics2018; @pereiraTransportationEquity2021].
 Thus, for this review we engage the literature with the following questions in mind: 1) what is our current understanding of the things that transportation systems do, for whom, and to whom; 2) what does the literature say about the distribution of the benefits and costs of transportation systems; and 3) what values are embodied in normative statements about said distribution.
 Ultimately, this review aims to collate the existing academic knowledge on the matter, and present it in a manner useful to support the development and implementation of standards for equity in transportation planning and policy.
-In this, we aim to provide relief even, or particularly, to planners in those places where calls for justice are explicitly made through legislation^[[As recently as -@martensFair2021, Martens and Golub note that Federal directives related to Title VI in the US "do not provide guidelines that can help agencies develop explicit standards to assess the distribution of accessibility benefits from projects or plans."]].
+In this, we aim to provide relief to planners, especially in those places where calls for justice are explicitly made through legislation^[[As recently as -@martensFair2021, Martens and Golub note that Federal directives related to Title VI in the US "do not provide guidelines that can help agencies develop explicit standards to assess the distribution of accessibility benefits from projects or plans."]].
 
@@
 Justice is a fluid concept, because it depends on the desirability of different states of affairs, which may change between populations and over time.
-That said, it is possible to distinguish several forms of justice, including [see @jaggarPhilosophicalChallengesGlobal2009; @pereiraDistributiveJusticeEquity2017a; @karnerTransportationEquityTransportation2020]:
+That said, it is possible to distinguish several forms of justice, including those listed below [see @jaggarPhilosophicalChallengesGlobal2009; @pereiraDistributiveJusticeEquity2017a; @karnerTransportationEquityTransportation2020].
 
@@
 
-**Reparative (or restorative) justice** focuses on the reparation of caused harm; it centers 'reintegrative shaming' to restore victims, wrongdoers and the community according to moral values [@tylerRestorativeJusticeProcedural2006].
+**Reparative (or restorative) justice** focuses on the reparation of caused harm; it centers 'reintegrative shaming' to restore victims, wrongdoers, and the community according to moral values [@tylerRestorativeJusticeProcedural2006].
 
@@
 It might be argued that all of the above touch on forms of distributive justice.
-Retributive justice, for example, could be framed as being concerned with the distribution of the benefits and burdens of being a member of society; the way it is usually achieved is by distributing intangibles of a society's moral values like "freedom" (e.g., of movement, of association) as benefits, and the distribution of tangible resources as burdens (e.g., fines as a punishment).
+Retributive justice, for example, could be framed as being concerned with the distribution of the benefits and burdens of being a member of society; the way it is usually achieved is by distributing intangibles of a society's moral values like "freedom" (e.g., of movement, of association) as benefits, and/or the distribution of tangible resources as burdens (e.g., fines as a punishment).
 Reparative justice could look like the distribution of benefits and burdens to redress past wrongs, for example by asking those who have benefited from said wrongs, even if unwittingly, to shoulder a bigger fiscal burden in order to cover programs that mete benefits to those who are still harmed by past wrongs.
@@
 In contrast to the notion of "do no harm", such a standard embodies the ideal that no one is denied benefits.
-An egalitarian standard would weigh the benefits or burdens by population, and fairness is achieved when each group give or receive in proportion to their size.
+An egalitarian standard would weigh the benefits or burdens by population, and fairness is achieved when each group gives or receives in proportion to their size.
 In contrast, an affirmative action standard is even stricter, since it requires the benefits to be distributed in a non-egalitarian way that favors those who are still harmed by past or present discriminatory practices.
@@
 
-In the following section we will discuss the analytical apparatus that we will use to interrogate the literature on equity standards in transportation.
+In the following section we discuss the analytical apparatus that we use to interrogate the literature on equity standards in transportation.
 
@@
 This is done either explicitly, as for example in @karnerTransportationEquityTransportation2020, who ask "of what", "for whom," and "how much" in reference to equity; or implicitly, as in @gosslingUrban2016, who asks of the outputs of transportation "what?" (exposure, space, access) and "for whom?" (gender, age, ethnicity).].
-According to -@jaggarPhilosophicalChallengesGlobal2009, Western philosophy has approached the issue of justice by asking "Where?", "When?", "Who?", "What?", and "How?".
+According to @jaggarPhilosophicalChallengesGlobal2009, Western philosophy has approached the issue of justice by asking "Where?", "When?", "Who?", "What?", and "How?".
 Conventionally, discussions about justice have been aspatial, or rather, seen from the point of view of social space instead of geographical space, despite an early interest of geographers on the matter [@pirieSpatial1983].
@@
 3]<!--, a notion that can be expanded to include all members of the human species (e.g., with the Universal Declaration of Human Rights) and nature (e.g., such as the Ecuadorian constitution, the first ever country to grant legal rights to nature [@akchurinConstructingRightsNature2015])-->.
-In the case of justice in transportation, the question of "where?" is paramount; as it might be argued that, by their very nature, transportation generates inequalities.
+In the case of justice in transportation, the question of "where?" is paramount, as it might be argued that, by their very nature, transportation generates inequalities.
 By concentrating the effects of space-time convergence (for instance, by providing access to a transit system or a highway), an inequality is automatically generated.
@@
 A possible complication in the case of transportation is that disentangling the "who" from their mobility tools is not always straightforward.
-Clearly, a person is not their mode of transportation; however in practice, there are large segments of the population who live in situations where they cannot extricate themselves from the tools for mobility that they can use, either because they have driven themselves out of choices [see @laveryDriving2013], or have been driven out of choices by factors beyond their control (e.g., captive users of a single mode [@jacquesRescuing2012; @cherancheryInvestigatingPerceptionCaptive2018]).
+Clearly, a person is not their mode of transportation; however in practice, there are large segments of the population who live in situations where they cannot extricate themselves from the mobility tools mobility that they can use, either because they have driven themselves out of choices [see @laveryDriving2013], or have been driven out of choices by factors beyond their control [e.g., captive users of a single mode @jacquesRescuing2012; @cherancheryInvestigatingPerceptionCaptive2018].
 In societies that have grown into transportation monocultures with a predilection for automobility [@millerCollaborative2011] there may actually be less choice about mobility tools than we would like to assume.
@@
 For this reason, we cannot stop at considering mobility, but the ulterior goal of mobility, which is to reach destinations.
-In combination with land use systems (the spatial distribution of opportunities on the landscape), mobility produces _accessibility_, the potential to reach destinations.
+In combination with land use systems (the spatial distribution of opportunities on the landscape), mobility produces _accessibility_, the potential to reach places where activities that the traveler values take place.
 In this manner, we can think of the objects of transportation justice as being _proximate_ (the tools of mobility, mobility itself), and _ulterior_ (accessibility, opportunities for activity participation).
@@
 Does this mean that they apply universally irrespective of age?
-Surely not, since no reasonable person would consider a toddler's demands for freedom of association as an absolute.
+Surely not, since no reasonable person would consider a toddler's demands for freedom of association as absolute.
 So, when do these rights fully apply in the life of a person?
@@
 The Ministry's argument was that children riding the bus unsupervised compromised the parent's ability to provide care and placed them at risk [@brendHere2017].
-Crook argued that the goal in teaching his kids (then aged 7, 8, 9 and 11) to use transit to go to school was to raise self-reliant children [@stueckYoung2017].
-While both sides discovered that no clear rules existed regarding childrens' independent transit use [@brendHere2017], statistically there is evidence that travel by car is riskier than travel by bus [@morencyTraveling2018].
+Crook argued that the goal of teaching his kids (then aged 7, 8, 9 and 11) to ride transit to school was to raise self-reliant children [@stueckYoung2017].
+While both sides discovered to everyone's surprise that no clear rules existed regarding children's independent transit use [@brendHere2017], statistically there is evidence that travel by car is riskier than travel by bus [@morencyTraveling2018].
 If the public good is to reduce risk, shouldn't children be banned from riding in cars?
@@
 But the devil is in the details, and answering the other questions discussed above is essential to pin the devil by the tail.
-Are children under the age of 10 subjects of justice (i.e., are they the "who") and who is deciding if they are (i.e., the courts and the parents in this case)?
+Are children under the age of 10 subjects of justice (i.e., are they a "who") and who decides if they are (i.e., the courts or the parents in this case)?
 If so, is the ability to use transit unsupervised an object of justice that should be justly distributed (i.e., the "what") and under what circumstances (e.g, to school, to a social event; during the day, or late at night?).
 "When" (e.g., for all children going forward, for only 5 years) and "where" (e.g., only in the suburbs, across the country) do these questions apply?
-Eventually, in 2020, a court of appeals ruled in Crook's favor, allowing his children to continue to ride the bus, thus providing much needed clarity with respect to the use of transit by children who desire independent mobility [@stueckCourt2020].
+Eventually, in 2020, a court of appeals ruled in Crook's favor, allowing his children to continue to ride the bus, thus providing much needed clarity with respect to the use of transit by children who desire and are encouraged to exercise independent mobility [@stueckCourt2020].
 In actuality, the courts _created_ a standard of equity: starting at a certain age, children are objects of justice from the perspective of freedom of movement, and they are due the benefits of independent mobility by transit.
 
-Another perspective on the answer to "Why?" are cultural _norms_ in providing rationale.
-In this case, the liberty of movement for all - including children - reigns supreme.
-But reigns supreme, "Why"?
-The outside is dangerous, in part due to vehicular traffic [@morencyTraveling2018], and in part because of potential violence from others.
+Another perspective on the answer to "Why?" are cultural _norms_ that provide a rationale.
+In this case, the liberty of movement for all---including children---reigns supreme.
+But a competing concern is that the outside is dangerous, and it is so in part due to vehicular traffic [@morencyTraveling2018], and in part because of potential violence from others.
 Parents may have over-pronounced 'stranger danger' based on neighbourhood and socio-economic characteristics of neighbourhoods [@francisLlBeDriving2017] and children themselves may fear potential violence based on real threats of victimization [@painParanoidParentingRematerializing2006].
-Who's voice, what perception of fear, and actions we take to intervene define and reproduce cultural norms, all play into the rationale behind the question "Why?".
-In this way, analyzing the "Why?" in the reviewed literature is left out of this review as answers to "Why?" are not explicitly stated and challenging to infer implicitly.
+Whose voice, what perception of fear, and actions we take to intervene define and reproduce cultural norms, all play into the rationale behind the question "Why?".
+In this way, analyzing the "Why?" in the reviewed literature is not the focuse of this review, partly because answers to "Why?" are not explicitly stated and challenge inference.
 As such, the focus of this report is on the standards of fairness that, combined with the use of equity analysis, can help us understand how better to move towards just transportation systems and better formulate answers to "Why?".
@@
 Pedestrians are the most extreme case of confluence between the "what" and the "who", since the mobility tool is the person's own body, with its inherent range of abilities.
-In papers that focus exclusively on walking, many use or develop walkability scores to explore perceptions of neighborhood quality [@evansAccessibilityUserNeeds2015] or pedestrian mobility with a focus on middle-aged and older adults [@towneUsingWalkScore2016], gender [@kimNeighborhoodEnvironmentWalkability2016], or urban peripherical regions [@blecicCapabilitywiseWalkabilityEvaluation2021].
+In papers that focus exclusively on walking, many use or develop walkability scores to explore perceptions of neighbourhood quality [@evansAccessibilityUserNeeds2015] or pedestrian mobility with a focus on middle-aged and older adults [@towneUsingWalkScore2016], gender [@kimNeighborhoodEnvironmentWalkability2016], or urban peripherical regions [@blecicCapabilitywiseWalkabilityEvaluation2021].
 These papers use 'walkability' as a way to measure the equity in its distribution.
@@
 Papers that exclusively focus on employment typically focus on these trips as they are the most common trip purpose and are often correlated with other trip activities like shops, recreation, and other services.
-For instance, @allenSizingTransportPoverty2019a propose a low employment-based accessibility threshold and a composite population vulnerability index to identify transportation-poor neighborhoods in urban Canada.
+For instance, @allenSizingTransportPoverty2019a propose a low employment-based accessibility threshold and a composite population vulnerability index to identify transportation-poor neighbourhoods in urban Canada.
 Papers that focus on healthcare and employment typically source data from representative travel surveys/diaries, census data, and points-of-interest databases.
@@
 #| fig-cap: "The proportion of equity standards (vertical axis) within each type of equity conceptualization (horizontal axis) category."
-standards_conc_percentages_plot_and_table
-ggsave("figures/standards_conc_percentages_plot_and_table.png")
+
+knitr::include_graphics("figures/standards_conc_percentages_plot_and_table.png")
 ```
@@
 
-## Call 1: The need for explicit conceptuations of equity to ground standards
-<!--*And be clear with our terms – what is equity and for whom?**-->
+## Call 1: Underpin standards in rigorous concepts of justice
+
 The conceptual grounding for standards is often left implicit within the literature: for example, @muellerSocioeconomicInequalitiesUrban2018 suggests the relative risk of mortality as related to transport-related air pollution should not be higher in deprived groups than the general population.
@@
 
-## Call 2: The need for creative systems-thinking approaches to measuring (in)equities
-<!--**Understand what transport inequities you are measuring**-->
+## Call 2: Develop creative methods for systems-thinking approaches to equity
+
 On the methodological side, we believe the wider application of mixed-methods are needed in transport equity research.
@@
 
-## Call 3: Data availability as a matter of justice
+## Call 3: Making data available is a matter of justice
 
@@
 
-## Call 4: Direct and explicit links between standards and experienced outcomes
-<!--*Understand what institutional context we’re operating in**-->
+## Call 4: Develop more direct and explicit links between standards and lived experiences
+
 We believe that robust assessments of the implications of equity standards on the experience of the public is still lacking in the literature, but it is essential for equity analysis that translates into just practice.
-While the estimation of the benefits of increased mobility or accessibility, or reducing affordability burdens and transport externalities is commonplace in the literature, these estimations need to be associated with outcomes like life and neighborhood satisfaction, subjective well-being, mental and physical health, social capital, among others.
+While the estimation of the benefits of increased mobility or accessibility, or reducing affordability burdens and transport externalities is commonplace in the literature, these estimations need to be associated with outcomes like life and neighbourhood satisfaction, subjective well-being, mental and physical health, social capital, among others.
 
@@
 
-## Call 5: Towards evaluations of interventions and policies in and across communities
-<!--**Understand what institutional context transfer looks like**-->
+## Call 5: Rigorously evaluate interventions and policies
+
 We believe there is a need to evaluate equity interventions and policies: track their before, during, and after impacts.
 In this review, only approximately 10% of studies assess specific transport-related policy interventions through an equity lens <!-- double check this number-->.
-Examples include mode-shift from driving to active school travel [@mammenSchoolTravelPlanning2014], transit fare restructures [@hickeyUsingQuantitativeMethods2010] and spatial analysis of Low Traffic Neighborhoods [@aldredEquityNewActive2021].
+Examples include mode-shift from driving to active school travel [@mammenSchoolTravelPlanning2014], transit fare restructures [@hickeyUsingQuantitativeMethods2010] and spatial analysis of Low Traffic neighbourhoods [@aldredEquityNewActive2021].
 The assessment of interventions and tracking associated outcomes can be thought of as a key step towards transport justice.
@@
 # Final remarks {#sect7}
+
 We began this report with the history-making suborbital flight of Jeff Bezos and his fellow passengers, all of them very wealthy people.
-We wondered why coverage of the event, in particular Bezos declarations planetside, were not met with more critical questions: who are the subjects of the marvelous future that Bezos fancies his trip will contribute to?
+We wondered why coverage of the event, in particular Bezos' declarations planetside, were not met with more critical questions: who are the subjects of the marvelous future that Bezos fancies his trip will contribute to?
 All of the human species or only those who can afford space tourism?
@@
 
-As a point for contrast, we noted the treatment that Amazon workers receive, including the degrading need of drivers to urinate in bottles while in their deliver trucks, in between dropping off Amazon parcels, in order to meet their productivity quotas.
+As a point for contrast, we noted the treatment that Amazon workers receive, including the degrading need of drivers to urinate in bottles while in their delivery trucks, in between dropping off Amazon parcels, in order to meet their productivity quotas.
 Who benefits from this indignity?
@@
 We argue that transportation, as an essential technology that can facilitate or hinder participation in society, is a very relevant sphere for matters of justice.
-The analysis of fairness and equity in transportation is vitally important for planning: as transportation systems and technologies converge space-time, they by definition leave certain populations, depending on the area and time scale, more or less burdened or benefited.
+The analysis of fairness and equity in transportation is vitally important for planning, Transportation systems and technologies converge space-time, and by definition they _always_ leave certain populations, depending on the area and time scale, more or less burdened or benefited.
 The analysis and tracking of equity implications is fundamental in formulating standards that are flexible to the fluid demands of transport justice.
@@
 
-- Explicit "Why?" rationale along with supporting "Where", "When", "Who", "What: and "How" framings for equity analysis. This means specific conceptualizations and grounded standards.
+- Explicitly explain "Why?", the rationale for justice, along with supporting "Where", "When", "Who", "What, and "How" framings for equity analysis.
+This means specific conceptualizations and grounded standards.
 - Creative methods of equity appraisal that can be paired with systems thinking; current methods are not used to support this processes.
@@
 440] than Bezos is for all his wealth.
-More importantly, we hope that in reviewing the literature, this report will contribute to advance the development and implementation of standards that help bend the arc of history towards more just transportation systems, and ultimately better societies.
-
-# Acknowledgements
-This report would not have been possible without the help and inspiration of countless others not let listed.
-In the early stages of the methodological design preparation of this report, the support from a team of literature review specialists at University of Toronto was invaluable; particularly the Research Services and Liaison Librarians.
-These experts provided guidance in cornering down the topic, identifying guidance material on literature synthesis and relevant databases, and in introducing us to the world of knowledge synthesis.
-Though our path diverged from a traditional scoping review, choosing to place and discuss the reviewed equity literature in connection to notions of justice, the protocol published on September 2022 (found [here](https://osf.io/rsb92)) was followed with only minor changes -- a reduction in the databases searched due to resource constraints.
-
-Furthermore, thank you for the hours of support from undergraduate and graduate research assistants who poured through title and abstract screening and helped determine initial stages of full-text eligibility.
-Reviewing such a quantity of literature was new to many of us, we are deeply thankful for your help: Jan Domalaon, Dancel Gayle, Eryn Maloney, Mahdis Moghadasi, Anika Munir, and Subaita Refaaf.
+Likewise, we hope that in reviewing the literature, this report will contribute to advance the development and implementation of standards that help bend the arc of history towards more just transportation systems, and ultimately better societies.
+
+\newpage
+
+# References
+
+<!-- 
+\bibsplit[2]
+
+Use \bibsplit to split the references from the body of the text. Value "[2]" represents the number of reference in the left column (Note: Please avoid single column figures & tables on this page.) -->
+
+:::{#refs}
+
+:::
 
@@
                            "From the perspective of disparity in gendered physical activity, this paper focuses on women's cycling as both transport and exercise.
-They advocate for all people achieving physical activity thresholds (**horizontal equity**) but prioritize women and especially women in neighborhoods with low-walkability and socio-economic status (**vertical equity**).",
+They advocate for all people achieving physical activity thresholds (**horizontal equity**) but prioritize women and especially women in neighbourhoods with low-walkability and socio-economic status (**vertical equity**).",
                            "How?: All people should get 150 min of moderate activity a week or 75 min of vigorous physical activity per week.",
@@
                            "How: 1) Access to facilities needed for old age health.
-Minimum distance thresholds from the geometric center of neighborhood are suggested: at least: 1000 m from health facilities (600 m or less is preferred), elderly-specific care facilities and shops should be 600 m (300 m or less is preferred).
+Minimum distance thresholds from the geometric center of neighbourhood are suggested: at least: 1000 m from health facilities (600 m or less is preferred), elderly-specific care facilities and shops should be 600 m (300 m or less is preferred).
 **Population standards**: 1) Certain neighbourhoods should be prioritized above others.
@@
 
-flextable(dat) %>%
-  ftExtra::colformat_md() %>%
+flextable(dat) |>
+  ftExtra::colformat_md() |>
   compose(
-    i = 1, j = c(1,2,3,4), part = "header", value = as_paragraph(c("Dimension", "Continent", "Conceptualization", "Standard"))) %>%
-  merge_v(j=1) %>% #merge 'dimensions' column
-  rotate(part = "body", rotation = "btlr", j=1, i=c(1:5,6:7,8:12, 13:14,15:18)) %>%
-  valign(valign="center", part="body", j=1) %>%
-  hline( i = c(5,7,12,14), part = "body") %>%
+    i = 1, j = c(1,2,3,4), part = "header", value = as_paragraph(c("Dimension", "Continent", "Conceptualization", "Standard"))) |>
+  merge_v(j=1) |> #merge 'dimensions' column
+  rotate(part = "body", rotation = "btlr", j=1, i=c(1:5,6:7,8:12, 13:14,15:18)) |>
+  valign(valign="center", part="body", j=1) |>
+  hline( i = c(5,7,12,14), part = "body") |>
   set_table_properties(layout = "autofit")
@@
 
-# References
+<!-- 
+DO NOT EDIT BELOW THIS COMMENT
+-->
+
+\newpage
+
+```{=tex}
+% LAST PAGE
+% This page is left blank on purpose
+
+% Set style for this page
+\thispagestyle{empty}
+
+% Council logo at bottom
+\begin{center}
+    \begin{figure}[b]
+        \includegraphics[width=\textwidth]{images/sshrc-canada-banner.png}
+    \end{figure}    
+\end{center}
+```

### 2023-11-16 README.Rmd
README in repo now offers a web-version of the report

--- 2023-11-15 just-transportation/just-transportation.qmd
@@
 ---
-title: "Searching for standards of fairness in the transportation justice literature"
-subtitle: "Searching for standards of fairness"
-format:
-  mj-report-pdf: 
-    keep-tex: true
-doctype: "Report (Draft)"
-
-# Author information. You can add or delete authors here
-author:
-  - name:
-      given: Anastasia
-      family: Soukhov
-    affiliations: 
-      - name: McMaster University
-        department: School of Earth, Environment and Society
-    email: soukhoa@mcmaster.ca
-  - name:
-      given: Ignacio
-      family: Tiznado-Aitken
-    affiliations: 
-      - name: University of Toronto
-        department: Department of Geography and Planning
-    email: i.tiznadoaitken@utoronto.ca  
-  - name:
-      given: Matthew
-      family: Palm
-    affiliations: 
-      - name: The University of North Carolina at Chapel Hill
-        department: Department of City and Regional Planning
-    email: palmmatt@unc.edu  
-  - name:
-      given: Steven
-      family: Farber
-    affiliations: 
-      - name: University of Toronto
-        department: Department of Geography and Planning
-    email: steven.farber@utoronto.ca  
-  - name:
-      given: Antonio
-      family: Páez
-    affiliations: 
-      - name: McMaster University
-        department: School of Earth, Environment and Society
-    email: paezha@mcmaster.ca
-
-# Date: default is last modified but could be a string.  
-date: last-modified
-
-# Write the acknowledgments; remember to use a tab at the beginning of each paragraph
-acknowledgements: | 
-  This report would not have been possible without the help and inspiration of many individuals not listed as authors.
-In the early stages of this report we received invaluable help from specialists at the University of Toronto as we learned about various approaches to literature reviews, particularly the Research Services and Liaison Librarians.
-These experts provided guidance in cornering down the topic, identifying guidance material on literature synthesis and relevant databases, and eased us into the world of knowledge synthesis.
-Though our path diverged from a traditional scoping review, the protocol published on September 2022 (found [here](https://osf.io/rsb92)) was followed with only minor changes, with only a reduction in the databases searched due to resource constraints.
-We also wish to acknowledge the many hours of work by undergraduate and graduate research assistants who poured through title and abstract screening to help determine full-text eligibility at the earliest stages of our search.
-The volume of literature initially reviewed was greater than any of us had experienced before, but important to be as inclusive as possible.
-For this support, we are deeply thankful to Jan Domalaon, Dancel Gayle, Eryn Maloney, Mahdis Moghadasi, Anika Munir, and Subaita Refaaf.
-  The views and opinions expressed here, as well as any equivocations and/or awkwardness are the authors' and the authors' alone.
-
-# Place the cover image and the toc image in folder `images`
-cover_image: cover-picture.png
-cover_image_caption: "Source: Marco Verch (https://www.flickr.com/photos/160866001@N07/50597690137/)"
-toc_image: toc-picture.png
-toc_image_caption: "Source: Progressive Charlestown (https://www.progressive-charlestown.com/2020/01/amazon-drivers-die-for-you.html)"
-# Information for generating citation
-citation:
-  type: report
-  doi: "10.somenumber/document.00010"
-  issued: last-modified
-  url: https://example.com/summarizing-output
-  number: "MJ-0001"
-bibliography: bibliography.bib
+output: github_document
+bibliography: just-transportation/bibliography.bib
 ---
+
+<!-- README.md is generated from README.Rmd. Please edit that file -->
+
+```{r, include = FALSE}
+knitr::opts_chunk$set(
+  collapse = TRUE,
+  comment = "#>"
+)
+```
+
+# Searching for standards of fairness in the transportation justice literature
+
+<!-- badges: start -->
+
+<!-- badges: end -->
 
@@
   DiagrammeRsvg::export_svg() |> 
-  charToRaw() |> rsvg::rsvg_png(file="figures/Pflow_chart.png")
+  charToRaw() |> rsvg::rsvg_png(file="just-transportation/figures/Pflow_chart.png")
 ```
@@
 ```{r import-extraction-data}
-Consensus_all <- read_excel("ALL_Extraction_Results_Feb22_edited.xlsx") |> filter(`Reviewer Name` == "Consensus")
-
-Consensus_only_analysis <- read_excel("ALL_Extraction_Results_Feb22_edited.xlsx") |> 
+Consensus_all <- read_excel("just-transportation/ALL_Extraction_Results_Feb22_edited.xlsx") |> filter(`Reviewer Name` == "Consensus")
+
+Consensus_only_analysis <- read_excel("just-transportation/ALL_Extraction_Results_Feb22_edited.xlsx") |> 
   filter(`Reviewer Name` == "Consensus") |> 
@@
 # standards_conc_percentages_plot
-# ggsave("figures/standards_conc_percentages_plot.png")
+# ggsave("just-transportation/figures/standards_conc_percentages_plot.png")
 ```
@@
 #standards_conc_percentages_plot_and_table
-# ggsave("figures/standards_conc_percentages_plot_and_table.png")
+# ggsave("just-transportation/figures/standards_conc_percentages_plot_and_table.png")
 ```
@@
 
+```{r include=FALSE}
+pdftools::pdf_convert('just-transportation/just-transportation.pdf',pages = 1)
+```
+
+```{r cover, echo=FALSE}
+#| label: cover
+knitr::include_graphics("just-transportation_1.png")
+```
+
 # Executive Summary
+
+----
 
@@
 
-![The proportion of equity standards (vertical axis) within each type of equity conceptualization (horizontal axis) category.](figures/standards_conc_percentages_plot_and_table.png)
+![The proportion of equity standards (vertical axis) within each type of equity conceptualization (horizontal axis) category.](just-transportation/figures/standards_conc_percentages_plot_and_table.png)
 
@@
 
+----
+
+**CONTENTS**
+
+1. [Just transportation?](#sect1)
+
+Just transportation? Or justice in transportation?
+
+2. [Setting the stage](#sect2)
+
+The analytical framework for the report.
+
+3. [Scanning the lay of the land: Methods](#sect3)
+
+The methods used for the literature review.
+
+4. [The lay of the land: a summary of findings](#sect4)
+
+A summary of the findings.
+
+5. [A critical appraisal of the lay of the land](#sect5)
+
+Critically appraising the findings.
+
+6. [Moving forward: calls for action](#sect6)
+
+What next?
+
+7. [Final remarks](#sect#7)
+
 This is your guide to scanning this document.
@@
 
+----
+
 - **Accessibility**: the potential to reach destinations.
@@
 # Just transportation? {#sect1}
+
+----
 
@@
 # Setting the stage {#sect2}
+
+----
 
@@
 
+----
+
 > "_Not everything that is faced can be changed. But nothing can be changed until it is faced._"  
@@
 #| fig-cap: "Evidence selection process framework. Step 1 (orange) is title and abstract screening, step 2 (green) is full-text review, and step 3 (purple) is data extraction."
-knitr::include_graphics("figures/Methods_framework.png")
+knitr::include_graphics("just-transportation/figures/Methods_framework.png")
 ```
@@
 #| fig-cap: "PRISMA flow diagram for the evidence selection process. ES signifies equity standard and EC signifies equity conceptualization."
-knitr::include_graphics("figures/Pflow_chart.png")
+knitr::include_graphics("just-transportation/figures/Pflow_chart.png")
 ```
@@
 year_cont_plot
-ggsave("figures/year_cont_plot.png")
+ggsave("just-transportation/figures/year_cont_plot.png")
 ```
@@
 pop_plot
-ggsave("figures/pop_plot.png")
+ggsave("just-transportation/figures/pop_plot.png")
 ```
@@
 mode_plot
-ggsave("figures/mode_plot.png")
+ggsave("just-transportation/figures/mode_plot.png")
 ```
@@
 
-knitr::include_graphics("figures/standards_conc_percentages_plot_and_table.png")
+knitr::include_graphics("just-transportation/figures/standards_conc_percentages_plot_and_table.png")
 ```
@@
 # A critical appraisal of the lay of the land {#sect5}
+
+----
 
@@
 
+----
+
 > "The way to right wrongs is to turn the light of truth upon them._"  
@@
 
+----
+
 We began this report with the history-making suborbital flight of Jeff Bezos and his fellow passengers, all of them very wealthy people.
@@
 
+----
+
 <!-- 
@@
 :::
-
-\newpage
-
-# Appendix {#sect8}
-
-## Search strategy
-
-```{r Search-query-A1}
-#| label: fig-A1
-#| fig-cap: "The search query.
-TS = topic search (keywords, abstract, title).
-TASCA = subject categories.
-Green text area transportation system related terms, blue text are equity dimension related terms, purple text are equity/justice conceptualization related terms, and orange text are standards related terms.
-Hits corresponds the number of papers that the search yielded and was retained into the evidence selection process."
-knitr::include_graphics("figures/Search-query.png")
-```
-
-Definitions of the population-concept context (PCC) used in the creation of the inclusion and exclusion criteria for the search strategy.
-
--   **Population**: the focus of the included studies should be on individuals, groups, communities, or entire regional areas that are impacted by passenger transportation infrastructure and systems (i.e., all modes and flows) from the perspective of equity (i.e., fair distribution, production, and re-production of burdens and benefits).
-This criteria is reflected in the creation of the first set of topic search terms that relate to transportation modes (e.g., "walking" OR "cycling" OR "transit" - see green text in @fig-A1 for the full list).
--   **Concept**: the included studies should also include equity dimensions and conceptualizes equity as discussed in the previous section.
-This inclusion criteria is reflected in the second and third set of topic search terms developed in the search strategy.
-These terms relate to types of equity dimensions (e.g., "accessibility" OR "mobility" or "transport-related air pollution" - see blue text in the @fig-A1 for the full list) and equity conceptualizations (e.g., "Justice" OR "equity" - see purple text in @fig-A1 for the full list).
--   **Context**: the included studies should also be limited to publications that include equity standards.
-Context can be more difficult to explicitly search for with key terms so synonyms for 'standards' were added to the query as a four set of topic search terms (e.g., threshold, indicator, criteria - see orange text in @fig-A1 for full list).
-Additionally, journal article and conference papers, English-language literature from any country, any study design (e.g., quantitative, qualitative, or mixed-method studies, or conceptual frameworks), and any record published within the past 30 years are included (January 1992 to March 2022).
-The time period is selected as the first (to the authors knowledge) peer-reviewed article which operationalized equity standards and equity conceptualization was published in 1996 [@khistyOperationalizingConceptsEquity1996]; we are broadening the search by a few years for completeness.
-English is selected as it is the common language spoken across the authorship team.
-Furthermore, papers that explicitly fall within the Transportation or related topic/category is included in the query (e.g., "Transportation", "Social Sciences", "Geography", "Civil Engineering", "Philosophy" - see the @fig-A1 for full query).
-
-The **exclusion criteria** for the search are papers that are not within the inclusion criteria.
-Specifically:
-
--   Literature published before January 1992.
--   Papers which do not include transportation equity dimensions.
--   Grey as concepts contained within are frequently published in a more developed form in journals.
-
-## Example of the data extraction template:
-
-```{r data-extraction-template-A2}
-#| label: fig-A2
-#| fig-cap: "The data extraction template with associated defintions."
-knitr::include_graphics("figures/Data-extract-template.png")
-```
-
-## Examples of papers summarized by element of the analytical framework
-
-```{r}
-dat <- data.frame(matrix(c("What?: **mobility and accessibilty**",
-                           "Where? [@rivasHowAffordableTransportation2018] - South America (select cities)",
-                           "Analyses how affordable urban public transportation is in select Latin American and Caribbean countries.
-They look at the estimated average monthly cost of transit trips and average monthly household income and conceptualize **transportation-related** **affordability**, especially for the most economically vulnerable (**vertical equity**).",
-                           "How?: The financial burden of a basket of urban public transportation trips (60 trip fares, representing 30 round-trips per month) should not exceed 10% of household monthly income.",
-                           "What?: **mobility and accessibilty**",
-                           "Where?: [@bharathyRevisitingClearFloor2018] - North America (USA - National)",
-                           "This study designed a web-based tool and took a representative sample of wheeled mobility device (WhMD) users anthropometry measurements to determine if the minimum standard suggested by the ADA is sufficient.
-We understand this conceptualization as a type of **Rights** conceptualization that WhMD should have minimum clear floor space (as described the guidelines in line with the American Disabilities Act) to access bus shelters, bus stop pads, and transit terminals.",
-                           "How?: The clear floor area for wheelchairs: 760 mm (30 in.) wide by 1220 mm (48 in.) in length as described by the ADA standards.
-Of note, this minimum clear floor area is insufficient for a variety of the WhMD users.",
-                           "What?: **mobility and accessibilty**",
-                           "[@ryanWhatAreWe2021] - Europe (Stockholm, Gothenburg and Malmo cities in Sweden)",
-                           "Investigates what the literature and planning process is missing when we measure accessibility by comparing objective and self-reported accounts of accessibility among older people.
-This paper conceptualizes accessibility as from the position of the **capabilities approach** and **vertical equity** (particularly acknowledging that older people have capabilities that differ from the general population).",
-                           "How?: Specifically for older populations (aged 65+), the following travel distances are suggested as equitable trip lengths to grocery stores per mode: Walking: less than or equal to 1500m, Combined public transit and walking (less than or equal to 1000m (walking element)), Combined car and walking: less than or equal to 1000m less than or equal to 1000m (walking element)), Bicycle: less than or equal to 3000m in addition to travel time threshold of less than 15 mins.",
-                           "What?: **mobility and accessibilty**",
-                           "Where?: [@wismadiSpatialPreferenceModelling2014] - Asia (Yogyakarta, Indonesia)",
-                           "Explores the equitable provision of transport infrastructure provision: an application of Sen's **capability approach**.
-Conceptualizes equity through Sen's capability approach and spatial equity.",
-                           "How?: Areas below the relative poverty line (of its neighbours) can only be located transport resources (i.e., measure in person*kms that can be travelled at car speed, i.e., mobility) based on the following 2 benchmarks (they can be considered, together as the floor/minmum access): 1) Global: standard deviation (SD) distance to mean should be minimized.
-2) Local: priority to minimise the differences with its neighbourhood",
-                           "What?: **mobility and accessibilty**",
-                           "Where?: [@zhengAreabasedEquitablePricing2020] - North America",
-                           "This paper conceptualizies equity in the multimodal network (transit, car) being fair toll-pricing across differences in populatins value of time (VOT).
-VOT is determined based on household income, with lower income households having lower VOT and thus deserving of lower tolls (vertical equity).
-From this perspective, a utilitarian perspective that seeks to minimize multimodal traffic congestion through introducing toll-pricing based on VOT is implemented.",
-                           "How?: suggest that a toll-pricing scheme based on individuals travel value-of-time (lower income people have a lower VOT) is equitable.",
-
-                           "What?: **environmental pollution**",
-                           "Where?: [@carrierApplicationThreeMethods2014] - North America (Montreal, Canada) ",
-                           "This work examines the statistical association between different social groups and the concentration of air pollutants.
-They frame their work from the perspective of environmental equity.
-We interpret the conceptualizations to be along the lines of **inequitable externalities**, **spatial** and **vertical equity** - transport-related air pollution is a product of road transport and it impacts the air of residents in unequal spatial ways.
-The paper then frames this impact as unfair, particularly from the perspective of disproportionately disadvantaged residents",
-                           "How?: The literature suggests that the health implications from the transport-related air pollution from major roadways is most acute at residential distance locations of 200 m or less.
-Residential locations should not be located within this distance threshold from the perspective of human health.
-**Environmental+** and **Population standards**: Uses the WHO NO² threshold as a point of comparison (annual concentrations of NO² should not exceed 40 μg/m-3).
-They argue that even through no neighbourhood, even those disproportionately low income, exceed the WHO limit in this case study, they still suggest that air pollution should not be disproportionately impacting disadvantaged neighbourhoods.
-It can be interpreted that they use the WHO threshold as a minimum threshold and suggest that air pollution levels should not be impacting disadvantaged populations disproportionately ( a relative population standard)",
-                           "What?: **environmental pollution**",
-                           "Where?: [@jephcoteGeospatialAnalysisNaturally2013] - Europe (Leicester, UK)",
-                           "Geospatial analysis of naturally occurring boundaries in road-transport emissions and childrens respiratory health across a demographically diverse cityscape.
-Emperically identifies at what distance away from major roadways children are most impacted by transport-related pollution.
-This is framed in the perspective of children’s **well-being**.
-Children are at most risk for acute respiratory distress from elevated levels of air pollution, and as such planning should consider this point of public health.",
-                           "How?: Finds that children (most vulnerable to air pollution - related to motoized traffic) are most impacted by air pollution within 283 m of a road way.
-This should be the distance threshold that schools and other childrens facilities are located.",
-                           "What?: **health impacts**",
-                           "Where? [@adlakhaMindGapGender2020] - Asia (Chennai, India)",
-                           "From the perspective of disparity in gendered physical activity, this paper focuses on women's cycling as both transport and exercise.
-They advocate for all people achieving physical activity thresholds (**horizontal equity**) but prioritize women and especially women in neighbourhoods with low-walkability and socio-economic status (**vertical equity**).",
-                           "How?: All people should get 150 min of moderate activity a week or 75 min of vigorous physical activity per week.",
-                           "What?: **health impacts**",
-                           "Where? [@savingmothersDidSavingMothers2019] - Africa (Select urban and rural regions in Uganda)",
-                           "The **well-being** of mothers, this paper examines the timely access to emergency obsteric and newborn care for child-bearing aged women in Uganda.",
-                           "How?: 2 hours to the nearest facility with surgical capacity with anesthesia services - this threshold is determined through the onset of bleeding to death if a women with obstetric hemorrhage does not receive adequate treatment).",
-                           "What?: **health impacts**",
-                           "Where?: [@iungmanImpactUrbanTransport2021] - Europe (Madrid and Barcelona, Spain)",
-                           "They use environmental pollution guidelines, but from the position of health.
-They investigate the impact of urban and transport planning on attributable mortality burden in Madrid and Barcelona and its distribution by socioeconomic status .
-Pre-mature mortality is linked to the exposure to pollution and motorized vehicles (**inequitable externalities**).
-These externalities should not be impacting people disproportionately (**vertical equity**) and should be even across space (**spatial equity**).",
-                           "How?: All minimum thresholds, if exceeded this is inequitable: NO² concentration 40 μg/m³; PM 2.5 concentration 10 μg/m³; Noise 53dB for average 24 hours; Living with 300 m crow-flies distance from at least .5 hectares of greenspace; and a Change of air temperature of at least 1 ⁰C.
-",
-                           "What?: **health impacts**",
-                           "Where?: [@mehdizadehWalkingTimeSchool2017] - Asia (Rasht, Iran)",
-                           "From the perspective of children’s **well-being**, assesses the walking time to school. They frame walking to school as health-related.",
-                           "How?: perceived walking time to school for students aged 7-9 yrs is 10 mins, and the longer the PWTS the less likely they were to use an active mode to travel to school.",
-                           "What?: **health impacts**",
-                           "Where? [@murphySupermarketAccessTransport2017] - Oceania (Melbourne, Australia)",
-                           "Assesses the relationship between supermarket access and transport mode used, the body mass index (BMI) of the mode-user (**wellbeing**) and the equity in access distribution by income (**vertical equity**).",
-                           "How?: all households should be sufficiently active (greater than 150 min and at least 5 sessions) and households should be within 1 km euclidean distance to supermarket (80-90% of the dwellings should meet this).
-Planners should prioritize socially disadvantaged areas to meeting these standards first.",
-
-                           "What?: **transport-related safety**",
-                           "Where?: [@ferenchakEquityAnalysisProactively2019] - North America (Denver, USA)",
-                           "Operationalizes and compares an equity analysis of proactively- and reactively-identified traffic safety issues from the perspective of **Spatial equity**, **Vertical equity** and **Inequitable exposure to externalities**.",
-                           "How?: standards are suggested for both reactive and proactive analysis.
-First, the lower the number of collisions on the road with pedestrians/cyclists (i.e., reactive safety analysis), the better.
-No/minimal inequalities for general population vs.
-equity seeking groups (high proportion of POC and/or low income in tract).
-Second, the lower the perceived safety, the better (i.e., if travel to school by ped.
-or bike is unsafe due to traffic conditions).
-No/minimal inequalities for general population vs.
-equity seeking groups (high proportion of POC and/or low income in tract).",
-                           "What?: **transport-related safety**",
-                           "Where?: [@zheEvaluationSharedUse2008] - Asia (Tokyo, Takamatsu, and Tokushima)",
-                           "Evaluates the observed safety of  shared use pedestrian and bicycle paths from the perspective of **well-being**.",
-                           "How?: the study suggests that the safety threshold for bicycles and pedestrians to coexist on shared infrastructure is less than 0.5 pedestrians/minute per metre of sidewalk (width) and less than 3.0 cyclists/minute per metre of sidewalk (width).
-The standard for pedestrian/bicycle share use in terms of hourly traffic volume is less than 26 pedestrians / hour and 108 cyclists / hour for 2m wide sidewalks.",
-
-                           "What?: **mobility/accessibility and health impacts**",
-                           "Where?: [@aldertonWhatMeaningUrban2019] - Asia (Bangkok, Thailand) – **Mobility/ accessibility** and **health**",
-                           "Establishes short-, medium-, and long-term goals for the city in collaboration with technical leaders within the municipal government for the perspective of **well-being** (urban livability): the standards included in this table relate directly to transportation systems.
-Indicators are inspired by the Sustainable Development Goals (SDGs) as well other global planning standards.",
-                           "How?: 1) Green space: % of residents living < 400 m from public open space, a large park (> 1.5ha), and/or local park, 2) Public transit access: % of residents living < 400 m of a local bus stop and <800 m of train station, 3) Facilities: % of residents living < 400 m of a community centre.
-The following **Infrastructure standard** is suggested: Canal water quality - dissolved oxygen content of equal to or less than 2.0 mL/L",
-                           "What?: **Mobility/ accessibility** and **health impacts**",
-                           "Where?: [@berheAdaptationDissonanceQuality2014] - Africa (Mekelle, Ethiopia)",
-                         "Examines adaption and dissonance in the quality of life (QoL) of residents.
-QoL is conceptualized along the lines of **well-being** and aspects of QoL directly tie into transport systems.
-They conduct a qualitative QoL survey of residents on the topic of three QoL domains: housing quality, access to important destinations, and affordability.
-They also measure quantitative indicators associated with these domains.
-We assume the equity goal for this paper is that subjective and objective QoL measures should not be mismatched: as discussed by the authors of this study, subjective QoL is higher than objective QoL the participant is experiencing adaption and in the reverse scenario the participant is experience dissonance.",
-                           "How: 1 & 2) Access to primary or secondary education facility, percentage of households living within 1 km or 2km (walking distance), respectively from a primary school or secondary school.
-3) Access to health facility, percentage of households within 40 min walking time from a health facility.
-4) Access to public transport, percentage of households within a distance of 500 m from a mini-bus stop.
-**Population standards**: 1) Adequate family income, percentage of households earning more than the official poverty line.
-2) Subjective QoL is constructed based on the households level of satisfaction for each of the eight indicators using a six point Likert-scale (1=very satisfied to 6=very dissatisfied).",
-                           "What?: **mobility/accessibility, health impacts, and safety**",
-                           "Where? [@agostfelipInclusiveModelAssessing2021] - Europe (Castellon, Spain)",
-                           "Conceptualizes equity through age-friendly urban spaces that reduce (and eliminate) conditions for **transportation-related social exclusion** for older populations and prioritize those who are economically vulnerable (**vertical equity**).
-These guidelines are inspired by the SDGs in addition to planning guidelines used national, regional, and local guidelines used in Spain.",
-                           "How: 1) Access to facilities needed for old age health.
-Minimum distance thresholds from the geometric center of neighbourhood are suggested: at least: 1000 m from health facilities (600 m or less is preferred), elderly-specific care facilities and shops should be 600 m (300 m or less is preferred).
-**Population standards**: 1) Certain neighbourhoods should be prioritized above others.
-From this papers focus on age-friendly urban environment, they suggest that if the neighbourhood has an average old age indicator (i.e., greater than 64 years, and/or greater than 79 years, and/or aging ratio of persons aged greater than 64 relative to 15 to 64 age) should be prioritized.
-2) Economic vulnerable and non-civically engaged neighbourhoods should also be prioritized.
-If the neighbourhood has a lower percentage of civic associations within the neighbourhood than average, and/or household income, and/or a higher than average interventions for dependency and/or social subsidies, they should be priorized.
-**Infrastructure standards **: 1) Green space: should be at least 10 m2 per inhabitant in the neighbourhood, greater than 15 m2 per inhab.
-is the goal.
-2) As related to sidewalk infrastructure at least 50% of all sidewalks (preferably 75% or greater) should: have a width of 1.5m or larger, ramps should have a grade of 8% or less, be well maintained (free from deficiencies), be paved for pedestrian use, and cover public transit stops.
-3) Lighting is critical for traffic-safety and a sense of safety overall.
-As such, at least 50% roads should: have a min.
-of 35 lux (road traffic) and 20 lux (pedestrian streets), and adapted traffic lights.
-4) Buildings should be age-friendly.
-As a proxy for the quality of residential living space quality, at least 50% of residential buildings in a neighbourhood should be built within the last 50 years (preferably 75% or more).
-In terms of physical access into the buildings, at least 10% should have elevators and accessible entrances (preferably 25% or more).
-**Environment + standards **: 1) Noise at the street level should be less than 55 dB and 45 dB (but preferably less than 50 dB and 40 dB) in the daytime and nighttime, respectively.",
-                           "What?: **mobility/accessibility and safety**",
-                           "Where?: [@mateobabianoPedestrianNeedsMatter2016] - Asian (Manila, Philippine)",
-                           "The perception of pedestrians’ walking environments should be sufficient across 6 themes.
-Equity is conceptualized around **spatial equity** (equally fair walking environments for all locations) and **rights** (the right to mobility/accessibility for pedestrians)",
-                           "How?: perceived pedestrian perception on protection, ease, equitable access, mobility, identity, and enjoyment must be met."
-                           ), ncol = 4, byrow = T))
-```
-
-```{r}
-set_flextable_defaults(font.size= 7,
-                       font.family = NULL,
-                       text.align = 'justify')
-
-flextable(dat) |>
-  ftExtra::colformat_md() |>
-  compose(
-    i = 1, j = c(1,2,3,4), part = "header", value = as_paragraph(c("Dimension", "Continent", "Conceptualization", "Standard"))) |>
-  merge_v(j=1) |> #merge 'dimensions' column
-  rotate(part = "body", rotation = "btlr", j=1, i=c(1:5,6:7,8:12, 13:14,15:18)) |>
-  valign(valign="center", part="body", j=1) |>
-  hline( i = c(5,7,12,14), part = "body") |>
-  set_table_properties(layout = "autofit")
-
-```
-
-<!-- we need to separate by qual and quant maybe?, karen asked-->
-
-<!-- 
-DO NOT EDIT BELOW THIS COMMENT
--->
-
-\newpage
-
-```{=tex}
-% LAST PAGE
-% This page is left blank on purpose
-
-% Set style for this page
-\thispagestyle{empty}
-
-% Council logo at bottom
-\begin{center}
-    \begin{figure}[b]
-        \includegraphics[width=\textwidth]{images/sshrc-canada-banner.png}
-    \end{figure}    
-\end{center}
-```

### 2024-02-23 just-transportation/just-transportation.qmd
Incorporation of some practioner comments into just-transportation.qmd (new .pdf not produced, some comments outstanding)

@@
 Another category of standards that emerged from the literature are based on infrastructure, and these include adherence to universal design guidelines or specific benchmarks for road network characteristics.
-Standards of thsi kind are comparatively less common in the literature and are often framed in terms of rights, particularly as they relate to achieving universal design.
+Standards of this kind are comparatively less common in the literature and are often framed in terms of rights, particularly as they relate to achieving universal design.
 Infrastructure standards around sustainable infrastructure (i.e.
@@
 Yet most of the research reviewed is purely quantitative or, to a much lesser degree, purely qualitative.
-The constant deployment of metrics Lorenz Curves of Gini coefficients without research on where such measures need to be to improve quality of life (i.e., a standard), will not help governments with limited resources achieve the best possible equity outcomes within their respective contexts.
+The frequent deployment of metrics Lorenz Curves of Gini coefficients without research on where such measures need to be to improve quality of life (i.e., a standard), will not help governments with limited resources achieve the best possible equity outcomes within their respective contexts.
 Mobilizing Justice’s Priority Populations working group was inaugurated to advance new research in this area.
@@
 Approximately 10% of the studies reviewed aimed to assess interventions.
-Delivering new equity-focused policies and programs that do not actually advance equity harms communities, metastasizes mistrust, and delegitimates future efforts to promote transportation justice.
+Delivering new equity-focused policies and programs that do not actually advance equity harms communities, generates mistrust, and delegitimates future efforts to promote transportation justice.
 Furthermore, policies that benefit one group of people in one context may not benefit other groups in other situations.
@@
 For example, firms may or may not adopt lower standards for the emission of pollutants in poorer areas; it might be that poor people end up being relegated to areas that already had lower emission standards [@gouldsonFirms2006].
-Regardless of the cause, the result is the same: pollution tends to be worse were poorer people are [@delucaCode2012].
+Regardless of the cause, the result is the same: pollution tends to be worse where poorer people are [@delucaCode2012].
 
@@
 But in democratic societies, the authority of political leaders, bureaucrats, planners, and all those charged with the business of governing, derives from the will of the people.
-It is therefore important to explicitly state how we plan to use these words, to clearly spell out our intuitions of "justice", "fairness", and "equity".
+It is therefore important to explicitly state how we plan to use these words, to clearly spell out our intuitions (what we mean by) of "justice", "fairness", and "equity".
 A clear mutual understanding of these concepts is essential for constructive debate, and for participants of a democratic society to be effective arbiters of what is just.
@@
 A possible complication in the case of transportation is that disentangling the "who" from their mobility tools is not always straightforward.
-Clearly, a person is not their mode of transportation; however in practice, there are large segments of the population who live in situations where they cannot extricate themselves from the mobility tools mobility that they can use, either because they have driven themselves out of choices [see @laveryDriving2013], or have been driven out of choices by factors beyond their control [e.g., captive users of a single mode @jacquesRescuing2012; @cherancheryInvestigatingPerceptionCaptive2018].
+Clearly, a person is not their mode of transportation; however in practice, there are large segments of the population who live in situations where they cannot extricate themselves from the mobility tools that they can use, either because they have driven themselves out of choices [see @laveryDriving2013], or have been driven out of choices by factors beyond their control [e.g., captive users of a single mode @jacquesRescuing2012; @cherancheryInvestigatingPerceptionCaptive2018].
 In societies that have grown into transportation monocultures with a predilection for automobility [@millerCollaborative2011] there may actually be less choice about mobility tools than we would like to assume.
@@
 These could be ministries or departments of transportation, transit agencies, public works departments and others having the power to act upon transportation (in)justices.
-Identifying these entities is relevant to ellucidate who is responsible, for example for apportioning the benefits or mitigating the burdens of transportation.
+Identifying these entities is relevant to elucidate who is responsible, for example for apportioning the benefits or mitigating the burdens of transportation.
 
@@
 With complex interlocking parts (sidewalk, road, traffic sign, parking regulations), transportation systems produce _mobility_, the potential for movement.
-The realization of this potentials happens through travel.
+The realization of these potentials happen through travel.
 However, as the adage goes, travel is derived demand, which seems to hold for most (even if not all) situations [e.g., @mokhtarianUnderstanding2001; @redmondPositive2001; @paezEnjoyment2010; @whalenMode2013].
@@
 Some examples include: maximum travel distance/cost/time to or from key destinations, levels of maximum exposure to externalities (i.e., noise or air pollution),un/fulfilled needs, and dis/satisfaction with travel.
-A number of theoretical and conceptual frameworks exist to support us when approaching this question, and we can draw from concepts in transport-related social exclusion, transport disadvantage, and/or transport poverty, which are typically based on equity principles, such as utilitarianism, Sen's capabilities approach, and sufficientarism.
+A number of theoretical and conceptual frameworks exist to support us when approaching this question, and we can draw from concepts in transport-related social exclusion, transport disadvantage, and/or transport poverty, which are typically based on equity principles, such as utilitarianism, Sen's capabilities approach, or Sufficientarianism.
 
@@
 Whose voice, what perception of fear, and actions we take to intervene define and reproduce cultural norms, all play into the rationale behind the question "Why?".
-In this way, analyzing the "Why?" in the reviewed literature is not the focuse of this review, partly because answers to "Why?" are not explicitly stated and challenge inference.
+In this way, analyzing the "Why?" in the reviewed literature is not the focus of this review, partly because answers to "Why?" are not explicitly stated and challenge inference.
 As such, the focus of this report is on the standards of fairness that, combined with the use of equity analysis, can help us understand how better to move towards just transportation systems and better formulate answers to "Why?".
@@
 The next stage is to trim the corpus of documents, a task that can no longer be automated, and requires expert knowledge.
-Selection of evidence is where this expert knowledge really comes to bear, and it consists of scanning the literature retrieved by the search strategy to retain in the corous only those papers that fit the inclusion and exclusion criteria.
+Selection of evidence is where this expert knowledge really comes to bear, and it consists of scanning the literature retrieved by the search strategy to retain in the corpus only those papers that fit the inclusion and exclusion criteria.
 This process was pilot-tested with a subset of papers before being implemented on the full set.
@@
 For instance, @awuorInfluenceEnvironmentalHealth2019 use the Neighbourhood Equity Index (NEI) to investigate disparities in premature death in Toronto (Canada).
-The NEI is a composite index that was developed by the city to capture differences in the City's neighbourhoods by ranking them based on socio-economic characteristics (e.g., social assistance, unemployment, income) and physical environmental characteristics such as green space availability.
+The NEI is a composite index that was developed to capture differences in the city's neighbourhoods by ranking them based on socio-economic characteristics (e.g., social assistance, unemployment, income) and physical environmental characteristics such as green space availability.
 Other works use national census indicators such as the social and housing deprivation index [e.g., @pucciInequalitiesJobrelatedAccessibility2019 or explore transport disadvantage, equity in policy implementation, or transport-related mortality burden by means of census measures (e.g., household poverty) and transport-related indicators [e.g., accessibility; @aldredEquityNewActive2021; @iungmanImpactUrbanTransport2021; @sunPublicTransportAvailability2021; @scheurerSpatialAccessibilityPublic2017].
@@
 They often assess universal design guidelines and the capabilities of people who face disabilities to travel.
-However, from another perspective, papers with an exclusive focus on gender, race/ethnicity, or education level/employment are less common in the literature reviewed.
+Papers with an exclusive focus on gender, race/ethnicity, or education level/employment are less common in the literature reviewed.
 Only two papers focus on gendered differences in cycling/active transportation (e.g., @adlakhaMindGapGender2020's case study in Chennai (India) and @xieWonCycleRoute2018's case study in Cardiff (UK)).
@@
 #| label: fig-fig5
-#| fig-cap: "The proportion of papers that investiage each type of mode. Categories for modes were generated upon data extraction."
+#| fig-cap: "The proportion of papers that investigate each type of mode. Categories for modes were generated upon data extraction."
 mode_plot
@@
 Pedestrians are the most extreme case of confluence between the "what" and the "who", since the mobility tool is the person's own body, with its inherent range of abilities.
-In papers that focus exclusively on walking, many use or develop walkability scores to explore perceptions of neighbourhood quality [@evansAccessibilityUserNeeds2015] or pedestrian mobility with a focus on middle-aged and older adults [@towneUsingWalkScore2016], gender [@kimNeighborhoodEnvironmentWalkability2016], or urban peripherical regions [@blecicCapabilitywiseWalkabilityEvaluation2021].
+In papers that focus exclusively on walking, many use or develop walkability scores to explore perceptions of neighbourhood quality [@evansAccessibilityUserNeeds2015] or pedestrian mobility with a focus on middle-aged and older adults [@towneUsingWalkScore2016], gender [@kimNeighborhoodEnvironmentWalkability2016], or urban peripheral regions [@blecicCapabilitywiseWalkabilityEvaluation2021].
 These papers use 'walkability' as a way to measure the equity in its distribution.
@@
 These benefits, however, are not necessarily valued by themselves, but rather are seen as instrumental to achieve an ulterior goal (e.g., activity participation).
-For example, although vehicle kilometers travelled (VKT) is still sometimes seen as a useful policy instrument [@zhaoRethinking2021], travelling more is not necessarily a sign of advantage when accessibility is low [@paezAccessibilityHealthCare2010], and similarly short trips may be a sign of advantage [@parkTransitParksEnvironmental2021].
+For example, although vehicle kilometers traveled (VKT) is still sometimes seen as a useful policy instrument [@zhaoRethinking2021], travelling more is not necessarily a sign of advantage when accessibility is low [@paezAccessibilityHealthCare2010], and similarly short trips may be a sign of advantage [@parkTransitParksEnvironmental2021].
 For this reason, although the right to the road (and to transportation systems more generally) is important, the literature leans heavy on the ulterior object, namely accessibility to destinations.
@@
 For instance, papers that focus on shopping destinations, such as grocery stores or markets (`r (dest_count$count[7]/sum(dest_count$count)) |> percent()`), frequently aim to identify food deserts [e.g., @choiFoodDesertsActivity2013; @jiaoHowIdentifyFood2012; @mckeyCrowdsourcedMappingHealthy2020; @kimAssessingSocialSpatial2020].
-Papers that focus on educational facilities, including primary, secondary, and post-secondary schools, represent only `r (dest_count$count[3]/sum(dest_count$count)) |> percent()` of the studies, and examine children's active transportation to school [e.g.,@laszkiewiczChildrenGreenWalk2020] or universal design [e.g., @larkinsAccessibleTransportationBuilt2011].
-When green space or other places of leisure is the exclusive focus (`r (dest_count$count[6]/sum(dest_count$count)) |> percent()` of papers), studies examine different accessibility questions such as the spatial distribution of green space e.g., [@xuSocialInequalitiesPark2017], for whom its accessible to [e.g., @mavoaAreaLevelDisparitiesPublic2015], and the reasons why the distribution of these spaces may be unequitable [e.g., @mearsUnderstandingSocioeconomicEquity2019].
+Papers that focus on educational facilities, including primary, secondary, and post-secondary schools, represent `r (dest_count$count[3]/sum(dest_count$count)) |> percent()` of the studies, and examine children's active transportation to school [e.g.,@laszkiewiczChildrenGreenWalk2020] or universal design [e.g., @larkinsAccessibleTransportationBuilt2011].
+When green space or other places of leisure is the exclusive focus (`r (dest_count$count[6]/sum(dest_count$count)) |> percent()` of papers), studies examine different accessibility questions such as the spatial distribution of green space e.g., [@xuSocialInequalitiesPark2017], for whom its accessible to [e.g., @mavoaAreaLevelDisparitiesPublic2015], and the reasons why the distribution of these spaces may be inequitable [e.g., @mearsUnderstandingSocioeconomicEquity2019].
 Very few of the papers we reviewed include 'community' destinations (e.g., public service centres, places of community support, and places of worship (`r (dest_count$count[2]/sum(dest_count$count)) |> percent()` of studies) or childcare activity types (`r (dest_count$count[1]/sum(dest_count$count)) |> percent()` of studies).
@@
 Of those papers, the most frequent focus is on **inequitable externalities**.
-This paucity in our corpus could be due to the fact that the the environmental burdens of transportation are covered more broadly in other literatures (e.g., environmental justice).
+This paucity in our corpus could be due to the fact that the the environmental burdens of transportation are covered more broadly in other literature (e.g., environmental justice).
 Papers that work with environmental-based standards often use traffic-related air pollution, noise pollution, green-space, urban design elements, urban air temperature, health related outcome, and physical activity guidelines to quantify externalities.
@@
 Most papers within our corpus (`r ((case_cont_count$count[4] + case_cont_count$count[5]+ case_cont_count$count[6])/sum(case_cont_count$count)) |> percent()`) focus on case studies in the Global North.
-Though their subject matter is varied, their spatial context mainly pertains to North American and Europe, and thus more often than not, in conversation with more developed and formal government transport planning apparatuses and technologies [e.g., planning for equitable high-speed rail, as in @monzonEfficiencySpatialEquity2013; autonomous vehicle technology, as in @eppenbergerOpportunitySharedAutonomous2021; or on the public consultation processes, as in @reddySafeguardingMinorityCivil2010]).
+Though their subject matter is varied, their spatial context mainly pertains to North America and Europe, and thus more often than not, in conversation with more developed and formal government transport planning apparatuses and technologies [e.g., planning for equitable high-speed rail, as in @monzonEfficiencySpatialEquity2013; autonomous vehicle technology, as in @eppenbergerOpportunitySharedAutonomous2021; or on the public consultation processes, as in @reddySafeguardingMinorityCivil2010]).
 
@@
 
-Temporally, we see the literature from the Global North and Global South in our corupus as laying on different transport inequity continuums.
+Temporally, we see the literature from the Global North and Global South in our corpus as laying on different transport inequity continuum.
 Literature from the Global North is concerned with particularly disadvantaged groups and creating indicators that may be used to guide the remedy of formal processes that ensure access to societal benefits e.g., [@cuiSpatialAccessPublic2020].
-Global North countries often surpass international sufficiency thresholds related to transport and public health (e.g., @carrierApplicationThreeMethods2014 measures NO_{2} road-side air pollution and find no levels are below the WHO guidelines) but advocate for the reduction in disproportional impact of air pollution on lower-income groups.
+Global North countries often surpass international sufficiency thresholds related to transport and public health (e.g., @carrierApplicationThreeMethods2014 measures $\text{NO}_2$ road-side air pollution and find levels are below the WHO guidelines) but advocate for the reduction in disproportional impact of air pollution on lower-income groups.
 These international guidelines, while perhaps being relevant at a time of lower transport-related development, are no longer relevant for some areas in the Global North.
@@
 
-Though the literature reviewed most commonly considers low-income as a socio-demographic characteristics of populations deserving equity as either by itself or alongside other characteristics (`r prettyNum(round(df_pop$count[8]+df_pop$count[4] * 100, 2))`\% of the papers).
+Though the literature reviewed most commonly considers low-income as a socio-demographic characteristics of populations deserving equity as either by itself or alongside other characteristics (`r prettyNum(round(df_pop$count[8]+df_pop$count[4] * 100, 1))`\% of the papers).
 Many employ accessibility methods (e.g., walking accessibility to open space [@tangWalkingAccessibilityNeighbourhood2021]) but others use qualitative or mixed-methods to gain perspectives of inequalities that are typically cross-dimensional (e.g., @milanLiftingPeripheralFortunes2017 asks Medellin (Colombia) residents living in TOD areas compared to non-TOD areas about the impact of TOD on their wellbeing based on income group and gender).
@@
 After transit users, cyclists and pedestrians are the next most common populations of interest.
-Historically, walking is how people get around [@robertsShort1998], and cycling and transit use space more efficiently and have relatively low impact.
+Historically, walking is how people got around [@robertsShort1998], and cycling and transit use space more efficiently and have relatively low impact <!-- ADD REFERENCE-->.
 Automobility is in direct tension with other modes as it requires massive amounts of space and receives extraordinary public subsidies, from the way its true cost is systematically underestimated [@gosslingSocial2019] to the way governments support the industries that feed it [@timperleyFight2021].
@@
 The focus on active travel and transit is to some extent understandable, given the common but perhaps misleading notion, that travel by car equates privilege.
-We argue that this is a somewhat mypoic omission in the literature reviewed, since it is possible that there are obligated car users who lack other privileges including, for example, people with mobility disabilities [@darcyRoad2018] or the working poor [@blumenbergAutomobile2012].
+We argue that this is a somewhat myopic omission in the literature reviewed, since it is possible that there are obligated car users who lack other privileges including, for example, people with mobility disabilities [@darcyRoad2018] or the working poor [@blumenbergAutomobile2012].
 
@@
 As shown in the variety of methods, dimensions, conceptualizations and standards used, transport inequalities are multi-dimensional.
-How decision makers define an equity-deserving community will impact results (e.g., [@rowangouldIdentifyingEnvironmentalJustice2015]).
+How decision makers define an equity-deserving community will impact results e.g., [@rowangouldIdentifyingEnvironmentalJustice2015].
 In this sense, standards that are defined need to be sensitive to changing community-based definitions of inequities.
@@
 Access to what sorts of opportunities are driving transport inequities?
-How do populations, transit modes, and opportunities saught intersect to define the "who" of inequities?
+How do populations, transit modes, and opportunities sought intersect to define the "who" of inequities?
 A community- informed understanding of inequities and tracking how it changes are needed.
@@
 As discussed, disadvantaged groups are the focus of analysis and the linkages between well-being and transport-related social exclusion [e.g., @delboscTransportProblemsThat2011; @churchillTransportPovertySubjective2019], areas more likely to experience transport poverty [@allenSizingTransportPoverty2019a], food deserts [@mckeyCrowdsourcedMappingHealthy2020], or transport-related energy poverty [@robinsonDoubleEnergyVulnerability2020; @berryInvestigatingFuelPoverty2016; @berryDistributionalEffectsCarbon2019] are centered as the objects of justice.
+<!--rephrase this paragraph, and think about how its different than the last section..?-->
 
@@
 Papers that put forward opportunity standards typically use methods related to travel impedance (e.g., more than 10% of monthly income should not be spent on transport [@rivasHowAffordableTransportation2018] to work and/or other necessary destinations).
-Alternatively, they operationalize spatial mismatch (disparity analysis) [e.g., @mulleyAreNetworkPlanning2015] and identify areas of _relative_ (regionally) inequities that should be addressed base on impedance thresholds to certain destination types.
+Alternatively, they operationalize spatial mismatch (disparity analysis) [e.g., @mulleyAreNetworkPlanning2015] and identify areas of _relative_ (regionally) inequities that should be addressed based on impedance thresholds to certain destination types.
 Others apply these same types of methods but tend to focus of human health, traffic-related safety, or cross-dimensional "What?"s.
-Methods can include inequality measures, poverty measures, and optimization methods to create 'optimal' landscapes, and once again vary depending on "Who?" and "What>" is centered.
+Methods can include inequality measures, poverty measures, and optimization methods to create 'optimal' landscapes, and once again vary depending on "Who?" and "What?" is centered.
 
@@
 A commonly used guidance document are WHO standards, be it traffic-related pollution, or noise standards and SGS goals on urban sustainability.
-The advantage of the use of these international standards is that they can facilitate cross community and temporal comparison to track progress towards equity targets; though they have limitations, as they have been critiques for being too general for practical application.
+The advantage of the use of these international standards is that they can facilitate cross community and temporal comparison to track progress towards equity targets; though they have limitations, as they have been critiqued for being too general for practical application.
 Overall, the ability to measure externalities tangibly (i.e., the road-side noise pollution produced by vehicles, the reduction of emissions associated with public transit modal shift) create interpretable numbers that can be measured and compared against to track progress towards these goals.
-These tangible metrics may be easier to interpret and compare across communities than using accessibility/mobility measures (being their methods are abundant, and contain embedded differing assumptions depending on the application).
+These tangible metrics may be easier to interpret and compare across communities than using accessibility/mobility measures (being their methods are abundant, and contain differing embedded assumptions depending on the application).
 
@@
 
+<!--TO ADD: I get the sense that going through the 5 whys would be helpful in this space. -->
 Measuring context-specific evidence of transport inequities is necessary to build knowledge of injustices and inspire creative interventions that advance the cause of justice.
@@
 Creating adaptable transport equity frameworks that lend themselves to change as the demands of justice change is as important as ensuring that they are resilient to backsliding.
-We argue for the necessity of setting standards from a systems approach, consiering both _positive_ rights (a right to have access to sufficient quality of essential services, for example) as well as _negative_ rights (mobility of cars must be limited to reduce their impact on air quality, health, safety, and positive rights).
+We argue for the necessity of setting standards from a systems approach, considering both _positive_ rights (a right to have access to sufficient quality of essential services, for example) as well as _negative_ rights (mobility of cars must be limited to reduce their impact on air quality, health, safety, and positive rights).
 
@@
 Is it sufficientarism?
-Or is it egalitariansm?
+Or is it egalitarianism?
 The standard can be interpreted in a multitude of ways.
@@
 While there is plenty of disparity analysis in the reviewed literature, frequently the standards used do not explicitly reflect practical application.
-For example, metrics of accessibility (usually measured with travel impedance cut-offs of between 15 to 60 minutes depending on the destination, population group and mode) are used to show descriptive differences among areas and groups but with scarce implications as to the experience of travellers.
+For example, metrics of accessibility (usually measured with travel impedance cut-offs of between 15 to 60 minutes depending on the destination, population group and mode) are used to show descriptive differences among areas and groups but with scarce implications as to the experience of travelers.
 How low of an accessibility index is too low?
@@
 For instance, papers including leisure destinations (e.g., green space, parks, recreation) are infrequently studied and some categories are missing all together like 'mobilities of care'.
+<!--It would be helpful to explain this term.-->
 
@@
 When/Where/How is it measured?
-And, who gets the consult and use the data?
+And, who gets to consult and use the data?
 Deciding who is the subject of justice frames the data collection activity; if it's the mobility of those that do domestic work the classification of who does this work and how it changes over time/space is fundamental.
@@
 What methods we're using, under what spatio-temporal boundaries we conduct data collection, and who has access to the collected data (and who does not) can all impact fulsome conceptualizations of justice.
-In the case of transportation, the issue of data (collection/availability) as a matter of justice is gaining traction as digitalization of data casts a starker light on these questions [e.g., see the papers by @sourbatiSmart2021 and @behrendtData2023].
+In the case of transportation, the issue of data (collection/availability) as a matter of justice is gaining traction as digitization of data casts a starker light on these questions [e.g., see the papers by @sourbatiSmart2021 and @behrendtData2023].
 
@@
 We began this report with the history-making suborbital flight of Jeff Bezos and his fellow passengers, all of them very wealthy people.
-We wondered why coverage of the event, in particular Bezos' declarations planetside, were not met with more critical questions: who are the subjects of the marvelous future that Bezos fancies his trip will contribute to?
+We wondered why coverage of the event, in particular Bezos' declarations planet-side, were not met with more critical questions: who are the subjects of the marvelous future that Bezos fancies his trip will contribute to?
 All of the human species or only those who can afford space tourism?
@@
 These are questions that intuitively seem easy: as we noted, humans have a keen sense of what is fair.
-But to move the political needle, it is critical that we have the concepts, the vocabulary, and the knowledge to express our intuitions of justice.
+But to move the political needle and establish a common understanding of justice, it is critical that we have the concepts, the vocabulary, and the knowledge to express our intuitions.
 The objective of this report is to provide a measure of clarity, with a particular focus on the definition and use of standards of fairness in transportation.
@@
 The transportation justice framework can be seen as a flexible classification exercise that is supported by a "Why" rationale.
-In conducting an equity analysis, the subject of justice (the "Who?") must contextualize the process, the distribution of the object of justice (the "what?") must be measured, be it transport-related benefits, burdens or both, "How?" these inequities are measured and benchmarked must be described, and these questions must be contextualized by the "Where?" across temporally changing "When?"s.
+In conducting an equity analysis, the subject of justice (the "Who?") must contextualize the process, the distribution of the object of justice (the "what?") must be measured, be it transport-related benefits, burdens or both, "How?" these inequities are measured and bench marked must be described, and these questions must be contextualized by the "Where?" across temporally changing "When?"s.
 The explicit acknowledge and answers to these questions clarify the boundaries of the equity analysis.

### 2024-04-14 Journal-manuscript/Journal-manuscript.qmd
Draft of journal-manuscript

---
title: "Searching for standards of fairness in the transportation justice literature"
format: docx

#   tandf-pdf:
#     keep-tex: true
#   tandf-html: default
# author:
#   - name: Anastasia Soukhov
#     affiliations:
#       - ref: MAC
#     orcid: 0000-0003-2402-304X
#     email: soukho@mcmaster.ca
#   - name: Ignacio Tiznado-Aitken
#     affiliations:
#       - ref: TORONTO
#     orcid: 0000-0003-2402-304X
#     email: i.tiznadoaitken@utoronto.ca
#   - name: Matthew Palm
#     affiliations:
#       - ref: UNC
#     orcid: 0000-0003-2402-304X
#     email: palmmatt@unc.edu
#   - name: Steven Farber
#     affiliations:
#       - ref: TORONTO
#     orcid: 0000-0003-2402-304X
#     email: steven.farber@utoronto.ca
#   - name: Antonio Páez
#     affiliations:
#       - ref: MAC
#     orcid: 0000-0003-2402-304X
#     email: paezha@mcmaster.ca
# 
# affiliations:
#   - id: MAC
#     name: McMaster University
#     department: School of Earth, Environment and Society
#     city: Hamilton, ON
#     country: Canada
#   - id: TORONTO
#     name: University of Toronto Scarborough
#     department: Department of Geography and Planning
#     city: Toronto, ON
#     country: Canada
#   - id: UNC
#     name: The University of North Carolina at Chapel Hill
#     department: Department of City and Regional Planning
#     city: Chapel Hill, NC
#     country: USA
# keywords:
  # - transportation
  # - equity
  # - literature review

bibliography: bibliography.bib  
---

```{r, include=FALSE, message=FALSE, warning=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
```

```{r setup}
library(huxtable)
library(sf)
library(ggplot2)
library(tidyverse)
library(readxl)
library(DiagrammeR)
library(scales)
library(corrplot)
library(patchwork)
library(flextable)
library(ftExtra)
```

```{r PRISMA-flow-chart}
Pflow_chart <- grViz("digraph flowchart {
      node [fontname = Helvetica, shape = rectangle, style = filled, color = grey90]

      tab1 [label = '6832 references imported for screening']
      tab1_1 [label = '611 duplicates removed']
      tab2 [label = '6221 studies screened - title and abstract (step 1)']
      tab2_1 [label = '4511 studies excluded']
      tab3 [label = '1710 studies assessed for full-text eligibility (step 2 and 3)']
      tab3_1 [label = 
      <1545 studies excluded:<br ALIGN = 'LEFT'/> 
      &#8226; 1061 No ES (only EC)<br ALIGN = 'LEFT'/> 
      &#8226; 285 No full-text<br ALIGN = 'LEFT'/> 
      &#8226; 147 No EC and no ES<br ALIGN = 'LEFT'/> 
      &#8226; 38 No EC (only ES)<br ALIGN = 'LEFT'/> 
      &#8226; 13 Other<br ALIGN = 'LEFT'/> 
      &#8226; 1 Wrong study type<br ALIGN = 'LEFT'/> >]
      tab4 [label = 'Data from 165 studies extracted']

      blank1[label = '', width = 0.01, height = 0.01]
      tab1 -> blank1[dir=none];
      blank1 -> tab1_1[minlen=10];
      {{ rank = same; blank1 tab1_1 }}
      blank1 -> tab2

      blank2[label = '', width = 0.01, height = 0.01]
      tab2 -> blank2[dir=none];
      blank2 -> tab2_1[minlen=10];
      {{ rank = same; blank2 tab2_1 }}
      blank2 -> tab3

      blank3[label = '', width = 0.01, height = 0.01]
      tab3 -> blank3[dir=none];
      blank3 -> tab3_1[minlen=10];
      {{ rank = same; blank3 tab3_1 }}
      blank3 -> tab4
      }
      ")
```

```{r, eval=FALSE}
Pflow_chart |>
  DiagrammeRsvg::export_svg() |> 
  charToRaw() |> rsvg::rsvg_png(file="figures/Pflow_chart.png")
```

```{r import-extraction-data}
Consensus_all <- read_excel("ALL_Extraction_Results_Feb22_edited.xlsx") |> filter(`Reviewer Name` == "Consensus")

Consensus_only_analysis <- read_excel("ALL_Extraction_Results_Feb22_edited.xlsx") |> 
  filter(`Reviewer Name` == "Consensus") |> 
  filter(`Included in analysis` == "Yes")
```

```{r by-year-and-continent-plot}
year_cont_plot <- ggplot(data = Consensus_only_analysis,
                         aes(y = `_Publication year`, 
                         fill = `_Case study continent`)) +
  geom_bar(aes(x = (..count..)/sum(..count..)),col="black") + 
  scale_x_continuous(labels=scales::percent,
                     breaks = seq(0,0.2, by=0.05),
                     limits = c(0,0.2)) +
  xlab("% of papers") +
  theme_classic() + 
  theme(text=element_text(size=10),
        axis.title.y = element_blank(),
        legend.position = "bottom") + guides(fill = guide_legend(nrow = 2,
                                                                 title=NULL))
```

```{r modes-by-type-plot}
df_mode <- strsplit(Consensus_only_analysis$`_mode`, "; ", fixed = T)

df_mode <- data.frame(
  id=rep(Consensus_only_analysis$`Covidence #`, lengths(df_mode)),
  mode=unlist(df_mode))

df_mode <- df_mode |> group_by(mode) |> 
  summarize(count = n()) |> 
  mutate(count = count/165,
         sub_ob_j = "What mode?")

mode_plot <- ggplot(data = df_mode, aes(x = forcats::fct_reorder(mode,count), y= count))+ 
  geom_bar(stat="identity", fill = "coral2", col = "black") + 
  scale_y_continuous(labels=scales::percent,
                     breaks = seq(0,0.4, by=0.05),
                     limits = c(0,0.4)) +
  coord_flip() +
  labs(y = "% of mentions in papers", subtitle = "Modes") + 
  theme_classic() + 
  theme(text=element_text(size=10),
        axis.title.y = element_blank())
```

```{r population-groups-plot}
df_pop <- strsplit(Consensus_only_analysis$`_population`, "; ", 
                       fixed = T) 
df_pop <- data.frame(
  id=rep(Consensus_only_analysis$`Covidence #`, lengths(df_pop)),
  pop=unlist(df_pop)) |> 
  mutate(pop = ifelse(pop == "Vulnerable populations (composite)", "Comp. vul", pop))

df_pop <- df_pop |> group_by(pop) |> summarize(count = n()) |> 
  mutate(count = count/165)

#adding grouping var so they print on plot in the order we want
df_pop$group <- c(7,9,6,8,2,3,5,10,1,4)

pop_plot <- ggplot(data = df_pop, aes(x = pop, y = count)) +
  geom_bar(stat="identity", fill = c(rep("skyblue2",each=8), "grey", "skyblue2"), col = "black") + 
  scale_y_continuous(labels=scales::percent, limits = c(0,0.25), breaks = seq(0,0.25, by=0.05))+
  scale_x_discrete(limits=(df_pop$pop)[order(df_pop$group)]) +
  coord_flip() +
  labs(y = "% of mentions in papers", subtitle = "Population groups") + 
  theme_classic() + 
  theme(text=element_text(size=10),
        axis.title.y = element_blank())
```

```{r concs}
df_conc <- strsplit(Consensus_only_analysis$`_equity conceptualization`, "; ", 
                                                   fixed = T)

df_conc <- data.frame(
                              id=rep(Consensus_only_analysis$`Covidence #`, lengths(df_conc)),
                              conc=unlist(df_conc)) |> 
                              mutate(conc = ifelse(conc == "Inequitable exposure to externalities", "Ineq. externalities", conc),
                                     conc = ifelse(conc == "Sufficientarism / Capabilities approach", "Emerg. theories", conc),
                                     conc = ifelse(conc == "Transport-related exclusion", "TR exclusion", conc))

df_standard_mcat <- strsplit(Consensus_only_analysis$`_type of equity standard2`, "; ", 
                       fixed = T) 
df_standard_mcat <- data.frame(
  id=rep(Consensus_only_analysis$`Covidence #`, lengths(df_standard_mcat)),
  std_mcat=unlist(df_standard_mcat))

standards_conc <- merge(df_conc, df_standard_mcat, by="id", all = T)

NUM_standards_conc <- standards_conc |> group_by(id) |> summarize(count = n())
NUM_ENTRIES_standards_conc <- NUM_standards_conc$count |> sum()

standards_conc <- merge(standards_conc, NUM_standards_conc, by="id", all = T)

# standards_conc_plot <- ggplot(data = standards_conc,
#                               aes(x = `conc`, fill = `std_mcat`)) +
#   geom_bar() +
#   labs(x = "Conceptualizations (summarized)", fill = "Standards (main categories)") +
#   theme_classic() + 
#   theme(text=element_text(size=10),
#         axis.title.y = element_blank(),
#         axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
#   coord_flip()

# c("Wellbeing","Vertical equity","Utilitarian","TR exclusion", "Spatial equity", "Rights", "Ineq. externalities", "Horizontal equity", "Emerg. theories")
```

```{r}
case_cont_count <- strsplit(Consensus_only_analysis$`_Case study continent`, "; ",
                       fixed = T)
case_cont_count <- data.frame(
  id=rep(Consensus_only_analysis$`_Case study continent`, lengths(case_cont_count)),
  case_cont=unlist(case_cont_count))
case_cont_count <- case_cont_count |> group_by(case_cont) |> summarize(count = n())

urban_form_count <- strsplit(Consensus_only_analysis$`_Case study urban form`, "; ",
                       fixed = T)
urban_form_count <- data.frame(
  id=rep(Consensus_only_analysis$`_Case study urban form`, lengths(urban_form_count)),
  urban_form=unlist(urban_form_count))
urban_form_count <- urban_form_count |> group_by(urban_form) |> summarize(count = n())

```
```{r calculating-destinations-of-interest}
dest_count <- strsplit(Consensus_only_analysis$`_activity`, "; ",
                       fixed = T)

dest_count <- data.frame(
  id=rep(Consensus_only_analysis$`_activity`, lengths(dest_count)),
  dest_count=unlist(dest_count))

dest_count <- dest_count |> group_by(dest_count) |> summarize(count = n()) 
```

```{r}
df_when <- Consensus_only_analysis |> 
  mutate(pub_year_group = case_when(`Publication year` >= 1992 & `Publication year` <= 2007 ~ "1992 to 2007",
                                    `Publication year` >= 2008 & `Publication year` <= 2012 ~ "2008 to 2012",
                                    `Publication year` >= 2013 & `Publication year` <= 2017 ~ "2013 to 2017",
                                    `Publication year` >= 2018 & `Publication year` <= 2022 ~ "2017 to Mar. 2021")) |>
  group_by(pub_year_group) |>
  summarise(count = n())|>
  mutate(sub_ob_j = "1.\nWhen?",
         count = count/165) |>
  rename("sub_ob_j_type" = "pub_year_group")

df_where <- data.frame(case_cont = c("Global North", "Global South"),
                                count = c(sum(case_cont_count[2:3,2],case_cont_count[4,2], 2), 
                                          sum(case_cont_count[4:6,2], 2))) |>
  mutate(sub_ob_j = "2.\nWhere?",
         count = count/165) |>
  rename("sub_ob_j_type" = "case_cont")

# THE WHO'S ARE MODE AND POPULATION GROUP
df_mode_2 <- df_mode |> mutate(sub_ob_j = "3.b Who?\n(mode)") |> 
  rename("sub_ob_j_type" = "mode")

df_pop_2 <- df_pop |> mutate(sub_ob_j = "3.a Who?\n(pop group)")  |>
  rename("sub_ob_j_type" = "pop") |> select(-c("group"))

# THE WHAT IS ARE DESTINATION group
df_dest <- dest_count |> mutate(sub_ob_j = "4.\nWhat?",
         count = count/165) |> 
  rename("sub_ob_j_type" = "dest_count")

df_all_sub_ob_just <- rbind(df_when, df_where, df_mode_2, df_pop_2, df_dest) #|>
  # mutate(sub_ob_j_type = as.factor(sub_ob_j_type),
  #        sub_ob_j = as.factor(sub_ob_j))
```
```{r}
# all_sub_ob_just_plot_stacked <- ggplot(data = df_all_sub_ob_just, 
#                                        aes(x = sub_ob_j, 
#                                            y = count,
#                                            fill = sub_ob_j_type,
#                                            fill = forcats::fct_reorder(sub_ob_j_type, count),
#                                            label = sub_ob_j_type
#                                            ))+ 
#   geom_bar(stat="identity", col = "black") + 
#   geom_text(size = 3, position = position_stack(vjust = 0.5)) +
#   scale_fill_manual(values = rep(RColorBrewer::brewer.pal(6,"Pastel2"),10))+
#   scale_y_continuous(labels=scales::percent,
#                      breaks = seq(0,1, by=0.1),
#                      limits = c(0,1)) +
#   coord_flip() +
#   labs(y = "% mentions in papers", x = element_blank(), subtitle = "Corpus Breakdown: % mention of topics in When, Where, Who and What") + 
#   theme_classic() +
#   theme(text=element_text(size=10),
#         axis.text.x  = element_text(size = 15),
#         legend.position="none")
# 
# all_sub_ob_just_plot_stacked
```

```{r}
df_all_sub_ob_just_reordered <- df_all_sub_ob_just |>
  mutate(
    sub_ob_j_type = ifelse(sub_ob_j_type=="Unspecified" & sub_ob_j == "3.b Who?\n(mode)", "Unspecified 3.b",sub_ob_j_type),
    sub_ob_j_type = ifelse(sub_ob_j_type=="Unspecified" & sub_ob_j == "4.\nWhat?", "Unspecified 4.",sub_ob_j_type),
    sub_ob_j_type = ifelse(sub_ob_j_type=="Employment" & sub_ob_j == "4.\nWhat?", "Employment 4.",sub_ob_j_type),
    sub_ob_j_type = ifelse(sub_ob_j_type=="Employment" & sub_ob_j == "3.a Who?\n(pop group)", "Employment 3.a",sub_ob_j_type),
    sub_ob_j_type = ifelse(sub_ob_j_type=="Other" & sub_ob_j == "3.b Who?\n(mode)", "Other 3.b",sub_ob_j_type),
        sub_ob_j_type = ifelse(sub_ob_j_type=="Other" & sub_ob_j == "3.a Who?\n(pop group)", "Other 3.a",sub_ob_j_type),
    sub_ob_j_type = fct_reorder(sub_ob_j_type, count, .fun='median'))

df_all_sub_ob_just_reordered$ID = seq.int(nrow(df_all_sub_ob_just))
```

```{r}
df_all_sub_ob_just_reordered_plot <- ggplot(data = df_all_sub_ob_just_reordered, 
       aes(y = reorder(sub_ob_j_type, count),
           x = count, 
           label = sub_ob_j_type,
           fill = sub_ob_j_type))+ 
  geom_bar(stat="identity",  col = "grey", alpha=0.8) +
  facet_grid(rows=vars(sub_ob_j),
             scales="free_y",
             space="free_y",
             switch = "y") +
   scale_x_continuous(labels=scales::percent,
                      breaks = seq(0,0.8, by=0.2),
                      limits = c(0,0.65),
                      expand = c(0,0))+
  scale_fill_manual(values=c(rep(RColorBrewer::brewer.pal(9,"Pastel1"),4))) +
  geom_text(size = 2.5, position = position_stack(vjust = 1), col = "Black") +
  theme_classic() +
  theme(strip.text.y = element_text(size = 5.5),
        strip.background.y = element_rect(fill = "lightgrey"),
        strip.placement = "inside",
        axis.title = element_blank(),
        axis.ticks.y=element_blank(),
        axis.text.y=element_blank(),
        axis.text.x=element_text(size=5),
        legend.position = "none",
        panel.grid.major.x = element_line(colour="grey", linewidth=0.1))
```

```{r, message=FALSE}
standards_conc <- standards_conc |> 
  group_by(conc, std_mcat) |> summarize(total = n())

standards_conc <- standards_conc |>
  mutate(total_perc_in_all = total/165)

empty_standards_conc <- data.frame(conc = c("Emerg. theories", "Utilitarian"),
                                   std_mcat = c("Environmental +", "Environmental +"), 
                                   total = c(NA,NA),
                                   total_perc_in_all = c(0,0))

totals_standards_conc <- rbind(standards_conc,empty_standards_conc)

#adding the grouping numbers, so the order that vars are printed are in geom_tile() are chronological
totals_standards_conc$group <- c(13,12,11,
                    44,43,42,41,
                    34,33,32,31,
                    24,23,22,21,
                    54,53,52,51,
                    74,73,72,71,
                    03,02,01,
                    84,83,82,81,
                    64,63,62,61,
                    14,
                    04)

#adding long definitions, for table under heatmap

totals_standards_conc_table <- data.frame(
  Conceptualizations = c("The distribution of transport system benefits/costs should favourably prioritize disadvantaged groups.",
                         "Concerning the links between transport system interaction and social exclusions.",
                         "Evaluation of transport system benefits/costs through wellbeing measures (e.g., livability, quality of life, subjective well being).",
                         "Equal distribution of transport system benefits/costs across space.",
                         "Equal distribution of transport system benefits/costs across population."),
  Conceptualizations_cont = c("Concerning the equitable distribution of externalities (e.g., road safety, congestion, pollution).",
                         "Concerning legal/human rights of access (e.g., mobilty rights, rights to the city, discriminatory application of laws).",
                         "E.g., Sufficientarism: minimums informed by individual capabilities. Capabilities approach: the capability of living the life one values (rather than having the right to do so)",
                         "Maximizing benefits/welfare overall, without consideration of distribution.",
                         " "),
  Standards = c("Related to user characteristics (e.g., affordability levels, perceived equity).",
                "Access to destinations indicators (e.g., # of parks within 30-min transit trip), the use of travel impedance thresholds (e.g., 300m to bus stop).",
                "E.g., Level of service, quality, universal design standards.",
                "E.g., noise, air, pollution criteria, safety.",
                " ")
)

```

```{r}
set_flextable_defaults(font.size= 6,
                       font.family = NULL,
                       text.align = 'justify')

totals_conc_table_flex <- 
  prepend_chunks(flextable(totals_standards_conc_table |> select(-c("Standards"))),
                 part = "body", i = c(1,2,3,4,5), j = 1, as_b(c("Vertical equity: ", "TR exclusion: ", "Wellbeing: ", "Spatial equity: ", "Horizontal equity: "))) |> 
  prepend_chunks(part = "body", i = c(1,2,3,4), j = 2, as_b(c("Ineq. externalities: ", "Rights: ", "Emerg. theories: ", "Utilitarian: "))) |> 
  # merge_h(part="header", i=1) |> #header merge
  merge_at(i=5, j=1:2) |>
  # align( align="center", part="header") |>
  delete_part("header") |>
  border_outer() |>
  border_inner_h() |>
  autofit()

totals_standards_table_flex <- 
  # compose(totals_standards_table_flex,
  #         i = 1, j = c(1), part = "header", value = as_paragraph(c("Standards"))) |> 
  prepend_chunks(flextable(totals_standards_conc_table |> select("Standards")),
                 part = "body", i = c(1,2,3,4), j = 1, as_b(c("Population: ", "Opportunity: ", "Infrastructure: ", "Environmental +: "))) |>
  merge_at(i= 4:5, j= 1) |>
  delete_part("header")|>
  border_outer() |>
  border_inner_h() |>
  autofit()
```

```{r}
standards_conc_percentages_plot <- ggplot(totals_standards_conc, 
       aes(x=std_mcat, y=conc, fill= total_perc_in_all ))+ 
  geom_tile() + 
  scale_fill_gradient2(low = "#AFAFED", mid= "#3F3FD4", high = "#191970", midpoint = 0.1, limits = c(0,.21), na.value = "grey40") +
  scale_x_discrete(limits=(totals_standards_conc$std_mcat)[order(totals_standards_conc$group)][c(4,3,2,1)]) + 
  scale_y_discrete(limits=(totals_standards_conc$conc)[order(totals_standards_conc$group)] ) +
  geom_text(aes(label = total_perc_in_all |> percent(accuracy = 1)), color = "white", size = 2.5) +
  theme_classic() +
  coord_flip()+
  theme(legend.position = "none",
        axis.ticks = element_blank(),
        axis.text.x=element_text(angle=20,hjust=1, colour = "black"),
        axis.text = element_text(size=7, colour = "black"),
        axis.title = element_blank())

#standards_conc_percentages_plot
# ggsave("figures/standards_conc_percentages_plot.png")
```

```{r calculating-concepts-and-standards}
conc_count <- strsplit(Consensus_only_analysis$`_equity conceptualization`, "; ",
                       fixed = T)
conc_count <- data.frame(
  id=rep(Consensus_only_analysis$`_equity conceptualization`, lengths(conc_count)),
  conc_count=unlist(conc_count))
conc_count <- conc_count |> group_by(conc_count) |> summarize(count = n())

stand_count <- strsplit(Consensus_only_analysis$`_type of equity standard2`, "; ",
                       fixed = T)
stand_count <- data.frame(
  id=rep(Consensus_only_analysis$`_type of equity standard2`, lengths(stand_count)),
  stand_count=unlist(stand_count))
stand_count <- stand_count |> group_by(stand_count) |> summarize(count = n()) 
```

```{r, warning=FALSE}
#the heatmap on top and table of stand/concept definitions on the bottom.
standards_conc_percentages_plot_and_table <- cowplot::plot_grid(
  gen_grob(totals_standards_table_flex, scaling="fixed", fit="auto"),
  standards_conc_percentages_plot,
  NULL,
  gen_grob(totals_conc_table_flex, scaling = "fixed", fit="auto"),
  nrow = 2, ncol=2, rel_heights = c(0.8,0.7), rel_widths = c(0.8,4) )
# standards_conc_percentages_plot_and_table
# ggsave("figures/standards_conc_percentages_plot_and_table_new.png")
```

```{r echo=FALSE}
#function for creating comments
word_comment <- function(comment, highlight = "") {
  if (isTRUE(knitr:::pandoc_to() == "docx")) {
    paste0('[', comment, ']{.comment-start id="0" author="Anastasia"',
           'date="2023-08-12T15:00:00Z"}', highlight, '[]{.comment-end id="0"}')
  }
}
```
# Introduction {#sect1}

Transportation systems, as a class of essential technologies that facilitate or impede social inclusion and activity participation <!---[@churchTransport2000; @lucasTransport2001; @socialExclusion2003; @cassSocial2005; @casasSocial2007; @prestonAccessibility2007; @paezMobility2009]-->, have increasingly come into focus from the perspective of equity; a lively and rapidly growing literature has emerged on the topic [@martensTransport2016; @karnerAdvancesPitfallsMeasuring2023; @vecchioTransportEquityLatin2020]. <!--@diciommoTransport2017; @guoSystematicOverviewTransportation2020; @karnerFrom2020; @pereiraTransportationEquity2021; @vanweeEvaluating2021; @zhangLiterature2021; @desjardinsExaminingEquityAccessibility2022; --> Aside from an ethical motivation, keeping track of objective and perceived inequalities could be of interest for governing bodies to respond to popular demands for fairness. However, transportation systems are notoriously complex: they simultaneously move people, goods, and information. Emerging technologies and service models can swiftly change the balance of benefits and burdens among a population [@guoSystematicOverviewTransportation2020], turning both users and service providers into digital rentiers [@birchBig2022]. The benefits and burdens of transportation systems are diffuse over space and time posing a challenge in evaluating and tracking inequities. Over space transportation systems engineered to offer higher mobility for people _somewhere_, can simultaneously cut others off from essential opportunities _elsewhere_, as @rajeEngineering2004 poignantly illustrated with examples of infrastructure in the UK. Over time, the shades of policies past can continue to haunt a region and the planet as shown by the legacy of displacement and decay caused by urban highways all across the US [@archerWhite2020] and the time horizon for the impacts of climate change [@markolfTransportation2019].

<!--But complexity is no excuse to shirk away from the task.--> The objective of this work is to scan the state of knowledge in terms of defining and operationalizing "fairness" in the transportation domain. Much research has been devoted to the issues of _measuring_ equity in transportation, including (among many others) [@ramjerdiEquity2006; @delboscUsing2011; @welchMeasure2013; @martensMeasuring2019; @pritchardNew2022]. Further, there are multiple works that discuss the conceptual and philosophical foundations of equity and fairness in transportation [@martensTransport2016; @pereiraDistributiveJusticeEquity2017a; @vanoutriveHow2019]. There have also been, previous reviews of planning documents have investigated equity from narrowly scoped perspectives, such as accessibility [@boisjolyHowGetThere2017] or a particular mode of transportation e.g., cycling [@doranPursuitCyclingEquity2021]. These efforts are valuable to scholars, the public, and decision-makers alike. In our estimation however, there remains a gap in the literature in terms of understanding how standards for equity are developed and implemented within the transportation domain.

To illustrate this gap, we note how @beilerExploring2016 in their exploration of transport equity, cite the following strategies identified by the US DOT to address matters of justice [p. 287]:

- Reduce adverse human health and environmental effects on minority and low-income populations.
- Include all potentially affected communities in the transportation decision-making process.
- Ensure that minority and low-income populations receive equitable benefits.

While commendable, the strategies are too vague, which means it is possible to implement them in a myriad ways, either genuinely to comply with the spirit of justice, or else performatively to deceive it [@mcculloughPerformative2023]. Some relevant questions include: how much should the adverse effects be reduced? To zero? Or to some tolerable level of adversity greater than zero? What should that level be and by what criteria? What benefits should be distributed? Should the benefits be based on simple population weights or, contrariwise, advantage more deprived individuals? These questions boil down to the development and use of _standards_ for transportation justice. In this work, we choose to focus exclusively on the academic domain as we know that adoption of equity concepts in planning practice has lagged developments in academic work [@pereiraTransportationEquity2021; @boisjolyHowGetThere2017; @linovskiEquityTransportationPlanning2020; @doranPursuitCyclingEquity2021; @litmanEvaluatingTransportationEquity2022]

Supporting the creation of (more) just transportation systems involves understanding the production and management of transportation benefits and costs; how they are distributed and what values are implemented (and by whom) in the form of standards [@pereiraDistributiveJusticeEquity2017a; @shellerMobilityJusticePolitics2018; @pereiraTransportationEquity2021]. In this spirit, our work seeks to make two significant contributions. First, it outlines a conceptual and flexible framework with supporting definitions for structuring transportation fairness and equity analysis. Second, it applies the framework to collate the existing academic knowledge that undertakes transportation equity analysis supported by equity standards and conceptualisations. We hope the findings and the presentation of academic knowledge will be useful in supporting the development and implementation of equity standards for transportation planners, policy and decision-makers.

In contrast to previous reviews on how to measure inequality in transportation systems, this work is concerned with the implied or explicit standards that are used to judge whether the measured inequalities are fundamentally “fair” or unacceptable. The present study achieves this objective by scanning the state of academic knowledge in defining “fairness” in the transport domain. Borrowing from philosophers of justice [@jaggarPhilosophicalChallengesGlobal2009], we ask the following questions to the literature and use them as a framework for presentation:

-	**What** are those benefits and burdens of transportation systems? Are they qualified in sidewalk widths, bus frequency, travel times, accessibility, or some other consideration?  
-	**Who** benefits and is burdened by transportation systems? Along which lines of identity are transportation inequities examined?
-	**Where and when** are equity standards applied? What are the differences between the research depending on location, context-specific planning processes and time of publication? 
-	**How** are equity standards and inequities measured? Are they infrastructure provision thresholds, acceptable pollution levels, or some other kind of cut off? How are these studies born out of broader notions of justice, i.e. disability rights, environmental justice, egalitarianism, or something else?

<!-- The structure of this review is as follows. After this introduction, the stage for our investigation is set by laying out some important definitions and describe the methods used for searching, selecting, and reviewing the relevant literature. This is followed by a summary of the findings and their critical appraisal. We conclude with some calls for action to improve the practice of setting and using standards of fairness for transportation justice. -->

# Background {#sect2}

In the introductory section the terms "justice", "fairness", and "equity" are used relatively loosely. This was done purposefully. As noted, people often have strong intuitions of what is "fair", "just", and "equitable" which may or may not match those of the authorities who set the standards of fairness. But in democratic societies, the authority of all those charged with the business of governing should derive from the will of the people. It is therefore important to explicitly state how we plan to use these words - what we mean by "justice", "fairness", and "equity".

**Justice** is an end goal, a desirable state of affairs that we are morally obligated to achieve. Justice is attained when people "give and receive whatever they are due" [@jaggarPhilosophicalChallengesGlobal2009, pp. 1-2], and it ceases to exist when there are persons or groups that are denied "access to the opportunities they need to lead a meaningful and dignified life" [@karnerTransportationEquityTransportation2020, p. 440].  Justice is a fluid concept, because it depends on the desirability of different states of affairs, which may change between populations, over space and time. That said, it is possible to distinguish several forms of justice  [@jaggarPhilosophicalChallengesGlobal2009; @pereiraDistributiveJusticeEquity2017a; @karnerTransportationEquityTransportation2020].

- **Retributive justice** is concerned with the proportional retribution of wrongdoers with relation to legitimate punishers and the innocent [@sep-justice-retributive].
- **Reparative (or restorative) justice** focuses on the reparation of caused harm; it centers 'reintegrative shaming' to restore victims, wrongdoers, and the community according to moral values [@tylerRestorativeJusticeProcedural2006].
- **Procedural justice** strives to ensure that the views and preferences of all stakeholders are fairly accounted for in the decision-making and inter-personal procedures affecting their lives and communities [@tylerRestorativeJusticeProcedural2006].
- **Distributive justice** is perhaps the most commonly studied form of justice [see @jaggarPhilosophicalChallengesGlobal2009, p. 2; @pereiraDistributiveJusticeEquity2017a], and its main concern is the way the benefits and burdens of the tangible and intangible products of society are collected by different segments of a population.

It might be argued that all of the above touch on forms of distributive justice. To illustrate, retributive justice, is usually achieved by distributing intangibles of a society's moral values like "freedom" (e.g., of movement, of association) as benefits, and/or the distribution of tangible resources as burdens (e.g., fines as a punishment). Reparative justice could look like the distribution of benefits and burdens to redress past wrongs, for example by asking those who have benefited from said wrongs, even if unwittingly, to shoulder a bigger fiscal burden in order to cover programs that mete benefits to those who are still harmed by past wrongs. Procedural justice could be the distribution of the benefits (e.g., the right to voice an opinion as a recognized stakeholder in the process) and burdens (e.g., the effort required to develop an educated opinion) of the processes that lead to decisions with collective consequences.

From this perspective, we can then speak of the _purposes_ of distributive justice: to mete out retribution _fairly_, to repair past harms, and to ensure that procedures offer _equitable_ opportunities to influence outcomes. From this perspective, equity and fairness are the "instruments" of justice, the tools by which society advances towards the end goal of justice.

**Equity**, as conceptualized alongside distributive justice, tends to encompass tools to understand the distribution of benefits and burdens of things among a population; often emphasizing those with the least advantage and equality in their outcomes. In the transportation domain, the term is somewhat loaded because it is perceived as stemming from the authority of the state and is meant to assist with decisions about regulating and financing transportation spending [@karnerTransportationEquityTransportation2020]. Here, we are in agreement with @karnerTransportationEquityTransportation2020 that equity analysis should not be seen as an end in and of itself, but rather as a means to gather information about actual and perceived inequities. In this respect, the analytical tradition of equity, at least in transportation planning, means that the relevant models become embedded in the "political ecology of the estimated truth" [@kingModels1993]: in principle their assumptions and scope must be open and transparent, or else they may be more vulnerable to misuse and even abuse as tools of subjugation.

**Fairness**, in contrast to equity, is somewhat more complicated to define. The concept does not have the same history of development as an analytical tool, and can be interpreted in numerous, and possibly discordant ways. This is convincingly demonstrated by @martensFair2021 in their study of the application of Title VI of the Civil Rights Act of 1964 in accessibility planning in the US. Title VI explicitly talks about the distribution of benefits derived from Federal funding: "[n]o person in the United States shall, on the ground of race, color, or national origin, be excluded from participation in, be denied the benefits of, or be subjected to discrimination under any program or activity receiving Federal financial assistance." However, as @martensFair2021 show, there are several ways to comply with regulations while achieving different outcomes, ranging from the banal (do not _knowingly_ discriminate), to the substantive (compensation for past discrimination within a societal context that recognizes harm was done, i.e., reparative justice). What kind of justice does fairness serve in each case? It depends on the rationale for seeking justice in the first place. Our reading of @martensFair2021 is that fairness is a yardstick that is best deployed _a priori_ than _a posteriori_, for doing the latter risks rationalizing the outcomes rather than driving them.

**Standards** are a way of making concrete statements about fairness. Returning to the ambiguities in Title VI discussed by @martensFair2021, the attainment of justice depends on the standard used to indicate fairness. For example, explicit non-discrimination constitutes a very weak standard that takes aim at the actions of agencies instead of the recipients of the benefits. Accordingly, any distribution of benefits would be considered fair, as long as the agency does not explicitly and knowingly target or deny benefits to particular groups. The standard provides conditions to determine whether a situation is _fair_. A similarly weak standard is a _Pareto improvement_, whereby it is possible to concentrate the benefits as long as no group is worse off compared to the status quo. A policy is fair as long it does no harm. A somewhat more strict standard, a Pareto-Plus improvement, stipulates that an intervention is fair when all groups receive at least _some_ (non-trivial) benefits; the size of the benefits for each group is irrelevant. In contrast to the notion of "do no harm", such a standard embodies the ideal that no one is denied benefits. An egalitarian standard would weigh the benefits or burdens by population, and fairness is achieved when each group gives or receives in proportion to their size. In contrast, an affirmative action standard is even stricter, since it requires the benefits to be distributed in a non-egalitarian way that favors those who are still harmed by past or present discriminatory practices.

## A framework to analyze questions of justice {#sect2.2}

<!-- To summarise the discussion of definitions, justice is a goal of social progress. But to understand what that goal is, we must clearly define standards of fairness. Equity analysis is tool to measure where the actual or perceived distribution of the burdens and benefits of the products of transportation systems stand with respect to the standard, in other words, instruments to see how close or far we are from a just situation. -->

Give the discussed definitions of "justice", "fairness", "equity" and "standards", we approach the literature with an analytical apparatus inspired by the framing of @jaggarPhilosophicalChallengesGlobal2009 for philosophical questions of justice.^[Similar questions are found peppered throughout the literature. This is done either explicitly, as for example in @karnerTransportationEquityTransportation2020, who ask "of what", "for whom," and "how much" in reference to equity; or implicitly, as in @gosslingUrban2016, who asks of the outputs ("what?") of transportation (exposure, space, access) and "for whom?" (gender, age, ethnicity).]. According to @jaggarPhilosophicalChallengesGlobal2009, Western philosophy has approached the issue of justice by asking "Where?", "When?", "Who?", "What?", and "How?", applying them to a particular domain or sphere of life relevant to justice.

In the case of justice in transportation, the question of **"Where?"** is paramount, as it might be argued that transportation generates inequalities by their very nature. By concentrating the effects of space-time convergence (for instance, by providing access to a transit system or a highway), an inequality is automatically generated. The burdens of transportation, in contrast, are often diffuse. They are incrementally paid, for example by a distributed population in the form of taxes, or by a population with a different spatial distribution in the form of poor health. As such, the answer to "Where?" is the definition of the spatial boundaries.

Conventionally, the question of **"When?"** refers to the temporal circumstances within which the demands of justice have application. In the cause of transportation justice, we ask about the temporal aspects of transportation systems, as examples: _when_ did the equity analysis take place and under what historical policy context; _the right time_ to invest in transportation infrastructure and as a result when to generate a spatial inequality [e.g., @rabelloEvaluation2015]; for _how long_ the burdens and benefits can still be associated to a specific transportation intervention; or even _timelines_ of reparative justice interventions that reconcile the shadows of past transportation-related injustices.

When answering **"Who?"**, we inquire about which entities should be regarded as subjects/arbiters of justice, meaning those entitled to make claims of moral consideration from the perspective of justice. To make it tractable, this question is often approached through the filter of population groups, which may include several concurrent traits, such as gender identity, ableness, ethnicity, age, caste, and income. Often, it is appropriate to reflect on the intersections between traits, given differences in a person's lived experiences. A possible complication in the case of transportation is that disentangling the "Who" from their mobility tools is not always straightforward. Although a person is not their mode of transportation, there are large segments of the population who live in situations where they cannot extricate themselves from the mobility tools they can use, either because they have driven themselves out of choices [@laveryDriving2013], or have been driven out of choices by factors beyond their control [e.g., captive users of a single mode @jacquesRescuing2012; @cherancheryInvestigatingPerceptionCaptive2018]. <!-- In societies that have grown into transportation monocultures with a predilection for automobility [@millerCollaborative2011] there may actually be less choice about mobility tools than we would like to assume.-->While it is important to avoid conflating the "Who" with the "What", for analytical purposes we need to be mindful of the connection between aperson and their mobility tools. <!-- In the case of transportation, in addition to members of the public who use transportation systems, there is another category of **Who**, that stands possibly in opposition to users, namely the entities charged with providing services, maintaining infrastructure, and so on. These could be ministries or departments of transportation, transit agencies, public works departments and others having the power to act upon transportation (in)justices. Identifying these entities is relevant to elucidate who is responsible, for example for apportioning the benefits or mitigating the burdens of transportation.-->

**"What?"** asks which entities should be regarded as objects of inequities, meaning which categories of things should be distributed in a just manner. To understand the distributional implications of transportation systems, it is essential to understand what do they _produce_? Transportation systems can be seen as space-time convergence technologies, mobility tools that improve the rate at which time is traded for space. They can increase the speed of movement: for instance, sidewalks and traffic lights enable potential movement of pedestrian and vehicles (i.e. mobility) and the realization of mobility through travel to needed/desired destinations. But as the adage goes, travel is derived demand [@mokhtarianUnderstanding2001; @whalenMode2013; @redmondPositive2001; @paezEnjoyment2010]. For this reason, we cannot stop at only considering mobility but must consider its ulterior goal-- reaching destinations. Coupled with land use, mobility creates accessibility to needed or desired activities. Transportation justice, thus, involves proximate (mobility tools and mobility) and ulterior (accessibility and activity opportunities) objectives. These considers all fold under objects of transportation justice, the "What?". 
<!-- At their most fundamental, transportation systems are space-time convergence technologies, tools that improve the rate at which time is traded for space. They usually do this by increasing the speed of movement: sidewalks facilitate walking, traffic lights facilitate the ordered flow of vehicular traffic, and a launching pad makes it possible for a rocket to take off. With complex interlocking parts (sidewalk, road, traffic sign, parking regulations), transportation systems produce _mobility_, the potential for movement. The realization of these potentials happen through travel. However, as the adage goes, travel is derived demand, which seems to hold for most situations [e.g., @mokhtarianUnderstanding2001; @whalenMode2013<!-- @redmondPositive2001; @paezEnjoyment2010; --><!--]. For this reason, we cannot stop at considering mobility but must consider its ulterior goal-- reaching destinations. In combination with land use systems (the spatial distribution of destinations/opportunities), mobility produces _accessibility_-- the potential to reach places where activities that are needed or desired. We can think of the objects of transportation justice as being _proximate_ (the tools of mobility, mobility itself), and _ulterior_ (accessibility, opportunities for activity participation). The burdens of transportation are also many and varied. Some are direct and paid directly by the traveler (e.g., travel time, out-of-pocket costs), but many others are indirect and related to network externalities (e.g., exposure to pollution, road accidents). -->

The next question is **"How?"**, and it relates to the allocation of various objects of justice ("what") to various subjects of justice ("who") in various circumstances ("when" and "where"). Equity standards are a tool for answering this distributive question: how do we allocate burdens and benefits? Standards are thresholds that, when operationalized effectively define what is fair. The thresholds can be quantitative (e.g., square meters of green space per capita), or they can be qualitative descriptions (e.g., do not knowingly discriminate), or a mix of the two. Some examples include: maximum travel distance/cost/time to or from key destinations, levels of maximum exposure to externalities (i.e., noise or air pollution),un/fulfilled needs, and dis/satisfaction with travel. A number of conceptualisations exist to support us when approaching this question, and we can draw from concepts in transport-related social exclusion, transport disadvantage, and/or transport poverty, which are typically based on equity principles, such as utilitarianism, Sen's capabilities approach, or Sufficientarianism.

Lastly, convincing answers to the above questions require a supporting rationale: a **"Why?"** [@jaggarPhilosophicalChallengesGlobal2009]. This is perhaps the most slippery of all the questions posed here. Justice is an inherently social construct. Asking **Why?** amounts to asking what sort of social contracts regulate human interactions, or in other words, what are the rules that our collective will to believe imposes on each of us. These contracts can be defined by constitution, but there are often unwritten and possibly contested variants. In this way, analyzing the "Why?" in the reviewed literature is not the focus of this review, partly because answers to "Why?" are not explicitly stated and challenged inference. As such, our focus is on the standards of fairness that, combined with the use of equity analysis, can help us understand how better to move towards just transportation systems and better formulate answers to "Why?".<!-- To see an example of "Why?" drawing from the Canadian Charter of Rights and Freedoms, see the Appendix.-->

## Methods {#sect2.3}

This review examines the breadth and depth of the academic literature on transportation to identify the extent to which standards for equity are defined and employed. In this task, we follow the Joanna Briggs Institute (JBI) approach to the conduct of scoping reviews, an approach that builds upon the Arksey and O'Malley (2005) framework [@petersUpdatedMethodologicalGuidance2020]. The review is also guided by the Preferred Reporting Items for Systematic Reviews and Meta-Analyses, particularly the extension for scoping reviews (PRISMA-ScR) which is consistent with the JBI approach [@triccoPRISMAExtensionScoping2018]. The use of these methods allows us to explore, in a consistent and organized manner, a relatively specialized topic within the broader transportation literature.

The primary research question and the protocol were initially defined by the authors, a group of experts in the field of transportation. The initial draft of the protocol was refined from preliminary searches of related-reviews [e.g., @zhangLiterature2021; @pereiraDistributiveJusticeEquity2017a; @vecchioAccessibilityCapabilitiesApproach2021], and in consultation with a University of Toronto Research Services Librarian and a Liaison Librarian in City Studies.

The search strategy was first developed iteratively using **inclusion** and **exclusion** criteria [@petersUpdatedMethodologicalGuidance2020]. For the inclusion criteria, the mnemonic PCC (population, concept, and context) was adopted. The search strategy was refined iteratively, adding topic search terms by stages (e.g., terms in the title, abstract or key words). The terms were bundled by means of logical connector terms "AND" and "OR". These stages are summarized as follows (see Appendix section @sec-sect61 for details):

1. An initial limited search of Web of Science (WoS) Core Collection (containing journals, conference proceedings, and books published all over the world) was undertaken to identify key documents. Separate searches using the terms 'transportation' and 'equity' were generated. From these searches, we examined the text contained in the titles and abstracts, the index terms, and subject heading searches when available. As we developed a clearer outline of the literature, we refined the terms used for the search. This took the form: ("Transport" OR "Transit" OR "Car\*" OR "Walk" OR "Bike"...**1**) AND ("Equity" OR "Justice" OR "Fair"...**2**), where **1** and **2** signify additional terms relating to 'transportation' and 'equity', respectively.

2. Upon inspection of the preliminary search results and after achieving a consensus among the authors, the set of search terms related to 'equity' was expanded into three sets of terms. The first describes theories and concepts of equity, the second describes the object of justice (i.e., the "what" in our analytical framework), and the third describes terms referring to standards (i.e., the "how"). All three sets of terms were augmented following an iterative process of refinement. The final search query took the following general form: ("Transport" OR "Transit" OR "Car\*" OR "Walk" OR "Bike"...**1**) AND ("Equity" OR "Justice" OR "Equity" OR "Fair"...**2**) AND ("Accessibility" OR "Mobility" OR ...**3**) AND ("Standard" OR "Threshold" OR ...**4**) where **1**,**2**,**3**, and **4** signify additional terms included in the sets combined with "OR" logical connectors.

After testing the search strategy on WoS Core Collection, we proceeded to apply to an augmented list of databases. The databases used were: WoS General Collection-Science Citation Index Expanded, WoS Social Sciences Citation Index, and Transportation Research International Documentation (TRID). The definitive version of the search was completed and exported by the lead author on March 21st, 2021.

The nature of the search strategy tended to be overly inclusive since we aimed to reduce the risk of omitting relevant material. The authors and a team of research assistants trimmed the initial corpus of documents from 6832 down to 165. All documents were first excluded based on title/abstract relevance where two research assistants voted, and a third from the authorship team broke ties. Then, 1710 documents were assessed based on full-text eligibility, again with each being voted on by two research assistants and an authorship team member tie-breaker. Next, using a data extraction template and workflow that was pilot-tested with a subset of papers, the authorship team extracted data from the eligible documents. *Covidence* [@covidenceCovidence2023], an online application for literature screening, was used for all steps of selection and data extraction on the full export of literature. Covidence is designed for collaborative work, and helps to document the work of multiple reviewers. The evidence selection workflow, (PRISMA) flow diagram [@pagePRISMA2020Statement2021], data extraction template, and some sample data extractions can be consulted in Appendix @fig-figA2.

To summarize, 6832 documents entered screening and ultimately 165 were retained for data extraction as part of the evidence corpus. These 165 pieces of evidence are relevant to the transportation domain and contain an explicit or inferred equity conceptualisation and standard.

# A critical appraisal of the lay of the land {#sect3}

This section critically threads together the trends identified in the corpus through the data extraction process (@fig-figA2)<!-- update this cross reference!-->. Specifically, conceptualisations and standards (the "How") and their applications to the objects of justice (the "What"), the subjects of justice (the "Who"), and under which situations ("When" and "Where"). As an overview, @fig-fig1 visualises the prominence of each category under "When?", "Where?", "Who?", and "What?". "How?" with its conceptualisations and standards is visualised in the following @fig-fig2.<!-- For the reasons reviewed, setting standards (that is, making statements of fairness) is a highly context-specific endeavor. While our search strategy and review tries to be as faithful to the literature as possible, it does not allow us to lay claim to reporting about on-the-ground transportation justice efforts from a truly global perspective.-->

```{r plotting-who-what-where-when-plot-fig1}
#| label: fig-fig1
#| fig-cap: "The proportion of category for each When, Where, Who and What; papers do often contain more than one topic and those papers are represented multiple times. Topic categories were generated upon data extraction."

df_all_sub_ob_just_reordered_plot
ggsave("figures/df_all_sub_ob_just_reordered_plot_new.png")
```

```{r plot-standards-concs-perc, warning=FALSE}
#| label: fig-fig2
#| fig-cap: "The prominence of Hows: the proportion of equity standards (vertical axis) within each type of equity conceptualisation (horizontal axis) category."

knitr::include_graphics("figures/standards_conc_percentages_plot_and_table_new.png")
```

## "When" and "Where" is transportation considered a sphere of justice {#sect3.1}

Within our corpus most papers (`r ((case_cont_count$count[4] + case_cont_count$count[5]+ case_cont_count$count[6])/165) |> percent()`) focus on case studies in the Global North, with many studies from North America (particularly USA and Canada), Europe (particularly France, Spain and Northern Europe), Oceania (Australia and New Zealand), and Asia (Japan and Israel).  Though their subject matter is varied, their spatial context mainly pertains to North America and Europe, and thus more often than not, in conversation with more developed and formal government transport planning apparatuses and technologies [e.g., planning for equitable high-speed rail, as in @monzonEfficiencySpatialEquity2013; autonomous vehicle technology, as in @eppenbergerOpportunitySharedAutonomous2021; or on the public consultation processes, as in @reddySafeguardingMinorityCivil2010]).

The smattering of studies from the Global South are predominately from Asia, notably China, but also India, Thailand, Iran, Philippines, and Indonesia. The next most common focus within the literature from the Global South is from South America. Many of these studies mention a systemic absence of evidence relevant to the region [@vecchioTransportEquityLatin2020]. Despite the growing recognition in the literature of the interconnections between transport development, social exclusion, and poverty [@benevenutoExaminingTransportNeeds2020], a number of studies underscore ongoing neglect of the social dimension of transport during the planning stage [@benevenutoExaminingTransportNeeds2020; @boisjolyAccessibilityMeasurementsSao2020]. Many studies also point at affordability as one of the main mobility barriers in the region [@falavignaAssessingInequalitiesPublic2016; @rivasHowAffordableTransportation2018], while some highlight multi-dimensional concerns such as public transport accessibility and quality of walking environments that contribute to mobility inequalities [@tiznadoaitkenRoleAccessibilityPublic2018].

Studies pertaining to Africa are even less numerous compared to the South American literature.
A shared characteristic among the studies from these two continents is a scarcity of official transport data [@friedMeasuringSustainableDevelopment2020] and a reliance on policy guidelines developed by international organizations. These studies also incorporate the use of informal transportation options and the pressures to develop road network infrastructure (which tends to support car dependency) over meeting the mobility/accessibility needs of citizens [@thondooSmallCitiesBig2020]. To address these challenges, researchers compile databases based on open and geo-referenced data, calculate objective and/or subjective measures [@berheAdaptationDissonanceQuality2014], and focus on advancing transport justice for low to medium income countries (LMIC) by aligning their goals with external policy guidelines such as the Sustainability Development Goals (SDG), particularly those related to universal accessibility [@friedMeasuringSustainableDevelopment2020].

Of all the studies reviewed, `r (urban_form_count$count[3]/sum(urban_form_count$count)) |> percent()` focus on urban and suburban settings and are highly varied in their research aims. To give an example, @coxQualitativeStudyAccessibility2020 qualitatively examine cycling as a mode of travel for people with disabilities in a typical mid-size town in the UK. @ampeImpactChildBike2020, on the other hand, work to identify the lateral clearance that motorists should maintain when passing cyclists with children seats. <!-- While both studies refer to the same objects of justice ("what", i.e., the right to the road and cycling), the subjects of inequities (the "who") are different (people with disabilities and children), and the "where", that is, the geographical context for the examination of fairness, differs as well.--> The remainder of the studies reviewed focus on rural regions (`r (urban_form_count$count[2]/sum(urban_form_count$count)) |> percent()`). We highlight the work of @caoIndicatorsSocioSpatialTransport2017, who examined transportation disadvantage in remote places which rely on inter-island ferry trips in the rural Philippines. Similarly, @parrySocialVulnerabilityClimatic2018 studied remote communities in the Amazonian region, and suggest that "increasing accessibility through road building would be maladaptive, exposing marginalized people to further harm and exacerbating climatic change by driving deforestation" (pp. 125).

Overall, studies from the Global South often have some key differences in focus compared to the Global North:

- affordability-as-a-barrier at the user- or policy maker- level is more often the motivation in work from the Global South;
- the expression of greater tensions in investing in new transportation infrastructure, such as roads in rural/under-developed areas, compared to prioritization of non-car modes. Studies centered in the Global North often focus on the later; and 
- more significant data availability limitations and reliance on 'informally' collected data.

We surmise these differences center on the direct economic outcomes tied to transport infrastructure. Often, work from the Global South does not engage as intimately with emerging mobility technologies that have large capital costs. Informal aspects present in transport planning also have a more overbearing presence. In the present day, countries in the Global South still struggle with the consequences of past colonialism under Northern states. This has left them more reliant on primary sector exports (lower efficiency, lower national GDP) under growing global financial markets, and with more fragile democracies. We infer, because of lower data availability, reliance on crowd-sourced or ‘informally’ collected data, and more extreme needs for ‘sufficient’ transport, analysis of transportation inequities is often cast along economic lines.

We characterise the Global North and South literature as laying at different points temporally on the transport inequity continuum. As Global South' formal planning processes operate under stricter financial constraints, they rely on more informal processes to address unmet needs - a significant source of equity concerns [e.g, informal transit in Nairobi @friedMeasuringSustainableDevelopment2020; populations living in informal residents in Mumbai @sharmaPublicTransitAccessibility2021]. Operationalization of international standards are also more relevant [e.g., WHO noise and air pollution in Delhi @apparicioCyclingOneMost2021]. In contrast, these standards are no longer as relevant in the Global North [e.g., $NO_2$ pollution near roadways in Montreal are under the WHO guideline @carrierApplicationThreeMethods2014], though perhaps being more relevant at a time of lower transport-related development. In these ways, the Global South temporally lags Global North development. However, this lag presents opportunities for all: for the Global South, to adopt relevant enhancements to Global North formal equity planning processes [e.g., indicator creation for disadvantages groups @cuiSpatialAccessPublic2020] and not repeat their mistakes [e.g., from the entrenched car-centric development @warrenDevelopingEquitableSustainable2015, to the disproportionate contribution to carbon-intense mobility @chancelCarbonInequalityKyoto2015]. For the Global North, reckon with their contributions to uneven development globally and environmental impacts as well as adopt relevant aspects from informal planning processes. <!--`r word_comment("I condensed this paragraph. The idea here is the global south temporally lags the global north in how equity is assessed/planned for, but both can learn lessons from these differences. Both corpuses can be seen at different points of equity assessment developlment. Why is this the focus? Because, I can't clearly see differences in --one year or multiple time frames, if they do before and after evaluations or just dianoses, etc.-- in the GS vs GN as you suggest. If you don't like what's written, or if it's not clearly a temporal angle, we can maybe talk about how the focus of 'who' changes depending on the time published (i.e., it used to be car vs. transit, but now it's car vs. all other modes -- especially cycling and walking). However, you can also see how 'time' here is just another way to examine the who/what/where/when/how." , ".")` -->

## "Who" are the subjects of transportation justice and their mobility tools {#sect3.2}

### Population groups
Turning to the "Who?" of population group characteristics in @fig-fig1, the majority focus on "Income", especially the lowest-income groups who face lower mobility and accessibility and higher costs and exposure [@peungnumsaiGridBasedSpatialAnalysis2020; @zhaoUnequableSpatialAccessibility2020; @falavignaAssessingInequalitiesPublic2016]. Evidence suggests that low household income is a significant determinant of transport-related inequities [e.g., urban access to public transport in Thailand [@peungnumsaiGridBasedSpatialAnalysis2020], access to urban employment opportunities in Brazil [@boisjolyAccessibilityMeasurementsSao2020], and unfavorable rates of environmental noise, air pollution, and green space in Netherlands [@kruizeEnvironmentalEquityRole2007]. Yet, low-income is not universally associated with lower transport-related benefits for every object of justice - context matters. For instance, in Sheffield, UK @mearsUnderstandingSocioeconomicEquity2019 demonstrate that historically working-class (lower-income) neighbourhoods have _more_ access to green space than other neighbourhoods. This is potentially due to historic urban planning approaches, though the quality is less than average. Similarly, @bertrandMeasuringMappingDisparities2008 find that lower income groups do not always have below average accessibility depending on the granularity of analysis.

"Age" is the second most common category within population group: these papers highlight differences in age-related capabilities. For instance, @martinez-jimenezAccessibilityCultureEducation2019 and @arranz-lopezSocialSpatialEquity2019 investigate travel distances/times to various opportunities, demonstrating how age is associated with differences in opportunity access. The most extreme age groups are commonly of focus: namely, school-aged children and older populations. For children, analysis of wellbeing [@laszkiewiczChildrenGreenWalk2020], safety and access to schools [@sharmaSpatialSocialInequities2022, @corazzaMethodologyEvidenceCase2020], and promoting active travel [@mackieOvercomingBarriersCycling2009; @mehdizadehWalkingTimeSchool2017] are common-place. Papers that focus on older adults similarly investigate transport-related wellbeing [e.g, @chenSpatialAnalysisFramework2020], access to age-specific destinations [e.g. @chengInvestigatingWalkingAccessibility2019], and improved options to meet unmet travel needs [e.g., @nordbakkeTransportUnmetActivity2015].

Typically, the corpus features papers that focus on intersecting (multiple) characteristics. However, the exception are what we classify as studies that focus on  "(Dis)abilities" or "All Populations". These categories are often of exclusive focus. Studies with a "(Dis)abilities" focus assess travel capabilities, namely through physical accessibility and universal design guidelines [e.g., @parkJourneyVisuallyImpaired2017; @chiscanoImprovingDesignUrban2021; @orellanaWalkRollMapping2020]. "All Populations" papers make no distinction in population. This is done, for instance in @kitaDoesLevelTransportation2020, who investigates disparities in accessibility to food-stores and their  self-reported capability/frequency of going outdoors. Often in both "(Dis)abilities" and "All Populations" focused papers, an im/explicit motivation is access (to necessities and desired destinations) for all.

As mentioned, the corpus is largely intersectional. For instance, a large proportion of papers apply a "Composite Vulnerability Index"; an _informed_ combination of several individual traits like low income, unemployment and immigrant status. These indices are generated from official government sources or author-informed census data creation methods. As an example, the Neighbourhood Equity Index (NEI) is a neighbourhood-ranked measure of vulnerability created by the City of Toronto and used in @awuorInfluenceEnvironmentalHealth2019 to examine disparities in premature death. Other works use national census indicators such as the social and housing deprivation index [e.g., @pucciInequalitiesJobrelatedAccessibility2019] or explore transport disadvantage, equity in policy implementation, or transport-related mortality burden by means of census measures (e.g., household poverty) and transport-related accessibility indicators [e.g.,  @aldredEquityNewActive2021; @iungmanImpactUrbanTransport2021; @sunPublicTransportAvailability2021; @scheurerSpatialAccessibilityPublic2017]. Similarly, Environmental Justice (EJ) indicators have been used in the US literature to identify neighbourhoods that have a higher than average proportion of low-income and non-white populations and evaluate the equity impacts of transportation projects [e.g., @rowangouldIdentifyingEnvironmentalJustice2016; @parkTransitParksEnvironmental2021; @reddySafeguardingMinorityCivil2010].

Intersectional considerations are so prevalent in the corpus that papers studying "Gender", "Race/ethnicity", "Education", or "Employment" level are infrequently studied exclusively. Only a few papers focus exclusively on gendered differences in active transportation [e.g., @adlakhaMindGapGender2020, @xieWonCycleRoute2018], race/ethnicity's relationship to green space proximity [@silvaEnvironmentalJusticeAccessibility2018], and culturally-appropriate opportunities [e.g., @wangModelingSpatialAccessibility2011]. Papers that focus *solely* on "Employment" or "Education" level are absent all together. Furthermore, "Other" population groups are also frequently considered in the corpus: this is a catch-all category that includes group populations less commonly the subject of research e.g.,  veterans and access to specific-healthcare needs [@mooneyTravelDistanceBarrier2000], pregnant people and access to services [@vadrevuMeasuringSpatialEquity2016], and youths who live in foster care [@batscheUsingGISEnhance2012]. Overall, the diversity of population groups considered in the corpus, often intersectionally, demonstrates the variety of transportation-equity concerns addressed in the corpus.

### Modes (mobility tools)
Travel mode, though modifiable, is intertwined with individual identity, posing challenges for equity analysis. In this sub-section, we choose to view the mode of travel primarily as those who use the mode, hence the subject of justice (i.e., the “who”), but occasionally, this will be enjoined to an object of justice (i.e., a “what”).

Public transit characterises the majority of the corpus (@fig-fig1). Despite transit being perceived unfavourably by some [@paezEnjoyment2010; @mellaDrivers2021], it is often viewed as the only and primary mobility option mode for many [@jacquesRescuing2012; @welchMeasure2013; @cherancheryInvestigatingPerceptionCaptive2018]. The public transit focused papers assess a variety of topics such as: food desert identification [@mckeyCrowdsourcedMappingHealthy2020], barrier-free transportation for people who face disabilities [e.g., @jimenez-espadaResearchProblemUniversal2021; @liuStatusQuoChallenges2019] and people with autism [@limFacilitatingIndependentCommuting2021; @feeleyValidationParatransitSkills2019]. We infer from the corpus that public transit is seen as a public good and hence a natural object of justice. Accordingly, public transportation is adaptable to meet the demands of justice, for example, by funding it sufficiently to provide barrier-free transport for most, despite challenges like low densities, fiscal constraints and political will [@markardUnsustainabilities2023].

Transit also plays a central role in multimodal comparisons for transport equity analysis: these papers feature "Walk" or "Car" or some other category in @fig-fig1. For instance, @brusselAccessAccessibilityCritique2019 compares public transit, pedestrian and road network accessibility measurements in the context of SDG 11.2 in Bogota. @renneWhatHasAmerica2018 reviews natural disaster evacuation plans, focusing on car-less households in the USA and emphasizing transit and pedestrian networks. A few papers also frame transit as being in direct competition with car travel or use their comparison as a benchmark [@golubUsingPrinciplesJustice2014; @martensJusticetheoreticApproachDistribution2012]. For instance, @warrenDevelopingEquitableSustainable2015 propose car ownership standards for Havana, Cuba, acknowledging the tension between mobility needs in transit under-served areas and emission reduction goals. However, this framing is not universal, transit is sometimes seen by the corpus as a mode to fulfill individual capabilities. As an example, @smithAccessibilityCapabilityMinimum2012 explores perspectives about transport needs and costs to achieve perceived sufficient living standards for those living in rural areas in the UK. Notably, papers vary in the importance they place on climate urgency, with some focusing more on satisfying *all* sufficient individual needs while planning for less car-dependent cities in the future.

Following transit, pedestrian travel ("Walk") is the second most studied object of justice. Pedestrians represent a unique convergence of "what" and "who," utilizing their own bodies for mobility. Papers focusing solely on walking often use walkability scores to assess neighbourhood quality [@evansAccessibilityUserNeeds2015] pedestrian mobility in different demographics [@towneUsingWalkScore2016; @kimNeighborhoodEnvironmentWalkability2016], or urban peripheral regions [@blecicCapabilitywiseWalkabilityEvaluation2021], measuring equity distribution. Some use walkability to gauge public health and urban vitality [@sungResidentialBuiltEnvironment2015; @mccormackRelationshipClusteranalysisDerived2012]. Furthermore, pedestrian-focused papers often discuss walking as a bridge to connect multiple modes as part of active transportation systems: linking walking, with cycling ("Bicycle"), and "Transit". Topics include how active transport impacts children's physical activity [@mammenSchoolTravelPlanning2014], walkability as a car alternative [@bertrandMeasuringMappingDisparities2008], and tensions between modes leading to unsafe walking conditions [@siuAssessmentPhysicalEnvironment2019; @ferenchakEquityAnalysisProactively2019].

Papers with a pedestrian focus also often see walking as a bridge to connect multiple modes: they discuss 'walkability' as part of active transportation, which focuses on both walking, bicycle and/or transit. Concepts discussed include how active transport contributes to children's physical activity levels [@mammenSchoolTravelPlanning2014], walkability as an alternative to car predominance [@bertrandMeasuringMappingDisparities2008] or tension that exists between modes, creating unsafe conditions for walking [@siuAssessmentPhysicalEnvironment2019; @ferenchakEquityAnalysisProactively2019].

The third most studied category is "Car". Unlike transit and walking, cars are rarely the sole focus. When studied in the corpus, they're critically compared with transit or considered in areas with inadequate transit [@kimmelStructuralBarriersComprehensive2018; @aljoufie2016]. Upon the rise of automobility, walking and cycling were in the way; <!-- though being the way people got around historically [@robertsShort1998] and cycling's efficiency at a lower impact-->Transit came to be seen as a social service and sometimes as a hindrance to the full realization of automobility. We surmise that the corpus' focus on non-car modes is a reckoning with automobility's legacy and ongoing requirements of disproportionate amounts of space, public subsidies and government support that underestimates its true cost [@gosslingSocial2019; @timperleyFight2021] including externalities like air pollution or safety [@fengTradeoffsMobilityEquity2014; @houstonProximityLicensedChild2006]. Other than "Car" being compared to walk, cycle or transit, some papers reflect the existing ways of travel and alongside "Car" including "Other" (@fig-fig1) options namely wheelchair accessible taxi and/or paratransit [@marquezFactorsAffectingPersonal2019; @wilkinsonmeyersLiveOrdinaryLife2015], travel on waterways [@vadrevuMeasuringSpatialEquity2016; @caoIndicatorsSocioSpatialTransport2017; @parrySocialVulnerabilityClimatic2018], motorcycles or other micro-mobility [@berryInvestigatingFuelPoverty2016; @savingmothersDidSavingMothers2019; @tiwariDevelopmentPublicTransport2021], or by emergency vehicle [@pedigoInvestigationDisparitiesGeographic2010; @patelDeterminingGeographicAreas2007].

Papers that pay no particular attention to any mode are "Unspecified": as examples, a focus on road infrastructure or road network distances [@wismadiSpatialPreferenceModelling2014; @mishraEIABasedComparative2014], travel needs generally [@titheridgeSocialExclusionAccessibility2008; @benevenutoExaminingTransportNeeds2020], realized travel [@abasoloEquityUtilizationAccess2001a], or externalities of realized travel [@iungmanImpactUrbanTransport2021].

<!-- 
### Who's missing?
As shown, "who" is the subject of inequities has been well studied. So what's missing from the "Who" question? We believe, more nuisances perspectives of inequities are needed. We must go beyond the focus on the low-income transit rider; positioning intersectionality of socio-economic and mode-use to create community-based definitions of equity and tailored standards are critical.

As shown in the variety of methods, dimensions, conceptualisations and standards used transport inequalities are multi-dimensional. How decision makers define an equity-deserving community will impact results e.g., [@rowangouldIdentifyingEnvironmentalJustice2015]. In this sense, standards that are defined need to be sensitive to changing community-based definitions of inequities. Are issues of economic-inequity at the root of transport inequities for a specific community? Are (dis)abilities? Are inequities in the service of transit the focus of a study because it can be improved to address transport inequities and are alternative modes driving these inequities? Access to what sorts of opportunities are driving transport inequities? How do populations, transit modes, and opportunities sought intersect to define the "who" of inequities? A community- informed understanding of inequities and tracking how it changes are needed. Different slices of analysis are important to different groups, as such, we believe all should be studied with an explicit justice framing. -->

## "What" are the objects of transportation justice {#sect3.3}
<!-- Though the literature reviewed most commonly considers low-income as a socio-demographic characteristics of populations deserving equity as either by itself or alongside other characteristics (`r prettyNum(round(df_pop$count[8]+df_pop$count[4] * 100, 1))`\% of the papers). Many employ accessibility methods (e.g., walking accessibility to open space [@tangWalkingAccessibilityNeighbourhood2021]) but others use qualitative or mixed-methods to gain perspectives of inequalities that are typically cross-dimensional (e.g., @milanLiftingPeripheralFortunes2017 asks Medellin, Colombia residents living in TOD areas compared to non-TOD areas about the impact of TOD on their wellbeing based on income group and gender). The use of qualitative methods allows for a more nuanced reflection of residents perspectives, and different inequities emerge.  -->
From the corpus reviewed, we infer that one of the most fundamental benefits of transportation systems are mobility (enabling or impeding movement) and accessibility (the ease of reaching destinations). These benefits are sometimes valued by themselves but are often seen by the corpus as instrumental to achieve an ulterior goal (e.g., activity participation and associated wellbeing benefits). For example, although vehicle kilometers traveled (VKT) is sometimes seen as a useful policy instrument [@zhaoRethinking2021], travelling more is not necessarily a sign of advantage when accessibility is low [@paezAccessibilityHealthCare2010], and short trips may be a sign of advantage [@parkTransitParksEnvironmental2021]. For this reason, although the right to the road (and transportation systems more generally) is important, the literature leans heavily on the ulterior object, namely accessibility to destinations.

### Impartial to destination interaction
Most papers take a broad approach, with `r (dest_count$count[8]/165) |> percent()` focusing on "Unspecified" destinations (@fig-fig1). They examine various equity dimensions across different transportation modes. Some papers focus on factors that impact the trip itself like infrastructure quality, safety, and service level [@zheEvaluationSharedUse2008; @prasertsubpakijEvaluatingAccessibilityBangkok2012; @furstBestBadPractices2013 ; @lattmanPerceivedAccessibilityPublic2016]. Others analyze trips tailored for specific groups like women or people with disabilities [@russellPedallingEquityExploring2021; @wilkinsonmeyersLiveOrdinaryLife2015], often with a consideration of what constitutes 'sufficient' quality of life [@churchillTransportPovertySubjective2019]. In sum, these papers reveal the multifaceted nature of transportation systems: they serve utilitarian purposes while also shaping user experiences.

### Partial to destination interaction
Among studied destinations, "Healthcare" services (`r (dest_count$count[6]/165) |> percent()`) and "Employment" (`r (dest_count$count[5]/165) |> percent()`) receive the most attention (@fig-fig1). Papers on healthcare often highlight disparities in services, like @wangModelingSpatialAccessibility2011, who assess access to Mandarin-speaking family physicians for Chinese speakers in Toronto. Similarly, papers focusing on employment [e.g., @allenSizingTransportPoverty2019a] may do so since it is the most common trip purpose and is often correlated with other trip activities like shops, recreation, and other services, often identifying transportation-poor neighbourhoods. These studies typically use travel surveys, census data, and point-of-interest databases, and benefit from well-developed and institutional data that represents ‘typical trips,’ especially in the Global North where this data is more readily available.

Other activity types receive less focus but serve distinct purposes beyond typically studied travel patterns. "Shopping" destinations (`r (dest_count$count[7]/165) |> percent()`) often aim to identify food deserts [@choiFoodDesertsActivity2013; @jiaoHowIdentifyFood2012; @mckeyCrowdsourcedMappingHealthy2020; @kimAssessingSocialSpatial2020]. "Education"-related papers (`r (dest_count$count[3]/165) |> percent()`) explore children's active transportation to school [@larkinsAccessibleTransportationBuilt2011] or universal design [@larkinsAccessibleTransportationBuilt2011]. Places of "Leisure" (`r (dest_count$count[6]/165) |> percent()`) prompt accessibility questions like spatial distribution [@xuSocialInequalitiesPark2017], accessibility [@mavoaAreaLevelDisparitiesPublic2015], and explores of inequity [@mearsUnderstandingSocioeconomicEquity2019]. Fewer papers cover "Community" destinations (e.g., public service centres, places of community support or worship) (`r (dest_count$count[2]/165) |> percent()`) or "Childcare" (`r (dest_count$count[1]/165) |> percent()`), but they're integral to study holistic activity participation [@albertsRebuildingWomenLivelihoods2016; @smithAccessibilityCapabilityMinimum2012]. The lack of information about community destinations, especially for children, is notable in the literature broadly [@desjardinsChildren2022].

### Dimensions of measurement

Defining the object of justice (the "What"), whether specific destinations or the trip itself, shapes measurements of inequality crucial for tracking progress towards transport justice. In the reviewed corpus, we characterize these dimensions as a focus on:

- accessibility and mobility; 
- transport-related environmental externalities; 
- centering human-health as related to transport systems; 
- human safety related to vehicular-traffic; and 
- cross-dimensional classifications.

As discussed next, these intersect with conceptualizations of fairness and applied standards, forming the philosophical foundations for assessing justice in transportation.

<!-- ### What's missing?  -->

<!-- What's missing from the "What?" in the equity analyses of the literature reviewed is the consistently explicit rationale for the object of justice. Specifically, "What?" are the benefits and burdens of a transportation system in terms of its: accessibility and mobility, or environmental externalities, or human-health externalities, or traffic related safety, or multi-criteria mix. In the inter-disciplinary analysis of transport systems, crystalising "What?" is the object of justice can create a stronger foundation to select appropriate methods to determine fairness and support the "Why?", the rationale of justice. In the corpus reviewed, strong conceptualizations of the object of justice often falls short. -->

## "How" is fairness determined {#sect3.4}

### Conceptualisations of fairness 
The philosophical foundations that are explicit or most often implicitly inferred from the corpus are broadly characterized as follows (@fig-fig2 details definitions). These categories emerged as we reviewed the corpus and are used to discuss trends, but are not an exhaustive list of conceptualizations.

- Vertical equity (`r df_conc|> group_by(conc) |> summarise(total = n()) |> subset(conc == "Vertical equity", select = "total") |> mutate(total = total/165) |> pull(total)  |> percent()`)
- Transportation related social exclusion (`r df_conc|> group_by(conc) |> summarise(total = n()) |> subset(conc == "TR exclusion", select = "total") |> mutate(total = total/165) |> pull(total)  |> percent()`)
- Well-being (`r df_conc|> group_by(conc) |> summarise(total = n()) |> subset(conc == "Wellbeing", select = "total") |> mutate(total = total/165) |> pull(total)  |> percent()`)
- Spatial equity (`r df_conc|> group_by(conc) |> summarise(total = n()) |> subset(conc == "Spatial equity", select = "total") |> mutate(total = total/165) |> pull(total)  |> percent()`)
- Horizontal equity (`r df_conc|> group_by(conc) |> summarise(total = n()) |> subset(conc == "Horizontal equity", select = "total") |> mutate(total = total/165) |> pull(total)  |> percent()`)
- Inequitable externalities (`r df_conc|> group_by(conc) |> summarise(total = n()) |> subset(conc == "Ineq. externalities", select = "total") |> mutate(total = total/165) |> pull(total)  |> percent()`)
- Rights (`r df_conc|> group_by(conc) |> summarise(total = n()) |> subset(conc == "Rights", select = "total") |> mutate(total = total/165) |> pull(total)  |> percent()`)
- Emerging theories (`r df_conc|> group_by(conc) |> summarise(total = n()) |> subset(conc == "Emerg. theories", select = "total") |> mutate(total = total/165) |> pull(total)  |> percent()`)
- Utilitarian (`r df_conc|> group_by(conc) |> summarise(total = n()) |> subset(conc == "Utilitarian", select = "total") |> mutate(total = total/165) |> pull(total)  |> percent()`)

A few slices of the literature and how they are characterized by their foundational conceptualisations of fairness are summarized. For example, a slice of the corpus is supported by broader "Rights" conceptualisations: these papers often focus on equity for people with disabilities or non-car users and associated challenges accessing transport infrastructure [@jimenez-espadaResearchProblemUniversal2021; @daamenAssessingGapPublic2008; @bharathyRevisitingClearFloor2018]. While many papers are underpinned with the right to the city (the _right_ to participate in the production of urban space [@lefebvreDroitVille1967]) [@adliRightCityApplying2018], others emphasize legal _Rights_ like ADA regulations [@bharathyRevisitingClearFloor2018] or the goal of _access for all_ in land-use transportation master plans [@limFacilitatingIndependentCommuting2021].

In another set of the corpus, distributions are examined through concepts of "Horizontal" and "Spatial equity", often using quantitative methods to assess distributional disparities without explicit justice rationales. Examples include setting travel impedance thresholds [@shenMeasuringSpatialAccessibility2020] and mapping accessibility indices spatially across populations [@monzonEfficiencySpatialEquity2013] or population-groups [@sharmaPublicTransitAccessibility2021]. These papers may also address traffic-related air and noise pollution or urban temperatures. In these papers, equity is theoretically achieved if similar levels are attainable for all populations (horizontal equity) or spatial areas (spatial equity), discussions rarely explore both minimum and maximum levels contributing to inequities from a justice perspective.

Papers that center "Well-being" assess what constitutes a satisfactory life in relation to transportation; this set of literature uses mixed-methods and, in contrast to the last set often explicitly identifies objects of injustice. Some use physical activity guidelines and surveys to understand activity related to active transportation infrastructure [@adlakhaMindGapGender2020; @auchinclossDesignBaselineDescription2020; @mccormackRelationshipClusteranalysisDerived2012]. Mixed or qualitative methods combined with health-related outcome standards (e.g., travel times for emergency treatment [@savingmothersDidSavingMothers2019], or premature mortality [@awuorInfluenceEnvironmentalHealth2019]) lead to firmer justice identifications.

Another research branch, often quantitative with some qualitative or mixed-methods studies, focuses on "Transport-related social exclusion", "Vertical equity", and/or "Sufficientarian/capabilities". They also often explicitly identify objects of justice and link standards to tangible and welfare-informed outcomes. They focus on groups from perspectives of disadvantage such as: social exclusion and transport poverty [@allenSizingTransportPoverty2019a; @delboscTransportProblemsThat2011; @churchillTransportPovertySubjective2019]], food deserts [@mckeyCrowdsourcedMappingHealthy2020], and energy poverty [@robinsonDoubleEnergyVulnerability2020; @berryInvestigatingFuelPoverty2016; @berryDistributionalEffectsCarbon2019].

### Standards and methods of measuring fairness
With the frequent way conceptualisations are applied within the corpus detailed, the following categories of standards that overlap with at least one conceptualisation are included (see @fig-fig2 for definitions):

- Opportunity standards (`r df_standard_mcat |> group_by(std_mcat) |> summarise(total = n()) |> subset(std_mcat == "Opportunity", select = "total") |> mutate(total = total/165) |> pull(total)  |> percent()`)
- Population standards  (`r df_standard_mcat |> group_by(std_mcat) |> summarise(total = n()) |> subset(std_mcat == "Population", select = "total") |> mutate(total = total/165) |> pull(total)  |> percent()`)
- Infrastructure standards (`r df_standard_mcat |> group_by(std_mcat) |> summarise(total = n()) |> subset(std_mcat == "Infrastructure", select = "total") |> mutate(total = total/165) |> pull(total)  |> percent()`)
- Environment+ standards (`r df_standard_mcat |> group_by(std_mcat) |> summarise(total = n()) |> subset(std_mcat == "Environmental +", select = "total") |> mutate(total = total/165) |> pull(total)  |> percent()`)

<!-- Opportunity standards often address horizontal and spatial equity, focusing on travel impedance metrics like speed, distance, or cost. For instance, @chenImpactHighspeedRail2017 set a 4-hour high-speed rail travel limit for municipalities to be considered "comfortably connected." Similarly, @yenisettyMeasuringAccessibilityVarious2020 deem a travel distance of less than 1,200 meters sufficient for local transit access. @shenMeasuringSpatialAccessibility2020 highlight regions where hospital access exceeds a 1-hour car journey, indicating inequity. These standards serve as benchmarks to assess disparities in travel impedance, acting as difference thresholds. -->
Papers suggesting "Opportunity" standards often employ quantitative methods to analyze disparities and assess distributional fairness. Many stem from travel impedance thresholds based on speed, distance, or cost [@chenImpactHighspeedRail2017; @yenisettyMeasuringAccessibilityVarious2020; @shenMeasuringSpatialAccessibility2020]. Inequality measures like the Gini coefficient and poverty measures are used to empirically define travel impedance thresholds [@vanderveenOperationalizingIndicatorSufficient2020; @tiznadoaitkenRoleAccessibilityPublic2018]. Further, methods tangential to travel impedance, like limiting transport expenditure to 10% of monthly income [@rivasHowAffordableTransportation2018], or addressing spatial mismatch [@mulleyAreNetworkPlanning2015], pinpointing areas of relative regional inequities are also used. Notably, many papers suggesting "Opportunity" standards include cross-dimensional considerations, employing similar methods but tailored to different focal points. For instance @peungnumsaiGridBasedSpatialAnalysis2020 suggest service benchmarks of equal supply and demand of public transit, revealing "Horizontal equities" as well. Others conceive the externalities of transportation system as trade-offs and aim to maximize transport-related benefits (i.e., time savings, emissions reductions, congestion reductions, user fares) through optimization/location-allocation methodologies e.g., [@fakhrmoosaviIncorporatingTravelTime2021; @zhengAreabasedEquitablePricing2020; @wismadiSpatialPreferenceModelling2014].

"Population" standards are often founded on "Well-being" conceptualisations, assessing what constitutes a satisfactory life related to transportation from a variety of socio-demographic and spatial angles. Methods include establishing threshold based on questionnaires and comparisons to recommended physical activity levels e.g. [@auchinclossDesignBaselineDescription2020; @mccormackRelationshipClusteranalysisDerived2012; @kimNeighborhoodEnvironmentWalkability2016; @timperioSafetyNumbersDoes2015], region-relative comparisons in health outcomes in a spatial unit e.g. [premature mortality rates @awuorInfluenceEnvironmentalHealth2019], spatial access benchmarks based on population-related characteristics e.g. [supermarket access @murphySupermarketAccessTransport2017; hospital access @pereiraGeographicAccessCOVID192021], summative per capita benchmarks e.g., [decent living energy consumption levels @raoDecentLivingEmissions2012], and community-informed spatial boundaries e.g., [EJ defined communities @rowangouldIdentifyingEnvironmentalJustice2016]. While most of these papers use quantitative or mixed-methods approaches, some employ exclusively qualitative methods to explore themes [@berheAdaptationDissonanceQuality2014].

Papers that feature both "Population" and "Opportunity" standards are often founded on "Vertical equity", "Well-being", and/or "Transport-related social exclusion" conceptualisations. They feature mix-methods, with questionnaires and qualitative approaches for "population" standards and quantitative methods like accessibility indices for "opportunity" standards. Census data and household estimates within specific travel distances or times to key destinations identify social exclusionary situations [@chenExploringTravelCharacteristics2010; @danielsProposalAccessibilityPlanning2011; @sunPublicTransportAvailability2021; @sharmaPublicTransitAccessibility2021], linkages between transport disadvantages [@delboscTransportProblemsThat2011], areas experiencing transport poverty [@allenSizingTransportPoverty2019a; @churchillTransportPovertySubjective2019], food deserts [@mckeyCrowdsourcedMappingHealthy2020], or transport-related energy poverty [@robinsonDoubleEnergyVulnerability2020; @berryInvestigatingFuelPoverty2016; @berryDistributionalEffectsCarbon2019]. They employ various methods, such as clustering techniques [@mohriClusteringMethodMeasuring2021]. Some exclusively use qualitative methods to analyze survey data on travel willingness/barriers or conduct interviews on unmet activity needs [@chenExploringTravelCharacteristics2010; @mehdizadehWalkingTimeSchool2017; @nordbakkeTransportUnmetActivity2015]. <!-- Understanding the reasons for travel that did not occur, whether due to preference or constraints, remains challenging (A. Paez and Farber, 2012).-->

"Infrastructure" standards offer another perspective on fairness, commonly grounded in "Rights" conceptualisations. Discussions of rights are twice as frequent in this segment of the corpus compared to other conceptualisations. These papers most frequently address the rights of non-car users and populations with mobility disabilities. Though various methods are applied, infrastructure and environmental audits as well as qualitative approaches are most prominent. Infrastructure audits compare existing infrastructure against universal design best practices [@odeckEconomicAppraisalUniversal2010; @larkinsAccessibleTransportationBuilt2011; @jimenez-espadaResearchProblemUniversal2021; @perezdelhoyoParticipatoryManagementImprove2021] or investigate elements correlating with mode use by specific population groups [@moniruzzamanInvestigation2016]. Qualitative methods include interviews/surveys on perceived access [@marquezFactorsAffectingPersonal2019; @mateo-babianoBicycleSharingAsia2017; @furstBestBadPractices2013; @velhoEffectTransportAccessibility2016; @parkJourneyVisuallyImpaired2017; @limFacilitatingIndependentCommuting2021; @stjernborgAccessibilityAllPublic2019; @desjardinsGoing2021] and assessment of standards under best-practice criteria [@daamenAssessingGapPublic2008; @velhoEffectTransportAccessibility2016; @bharathyRevisitingClearFloor2018].

Additionally, "Infrastructure" standards papers sometimes encompass multiple dimensions, moving beyond rights (to the infrastructure) and suggest "Opportunity" and/or "Population" standards. These papers often employ "Vertical", "Horizontal", and "Spatial equity" lenses. These multi-dimensional papers often refer to guidelines and propose composite indices. For example, [@racheleNeighbourhoodSocioeconomicTransport2017a] integrate various transport network properties to define an indicator supporting walkability and public transport access. Others evaluate infrastructure quality [@xuSocialInequalitiesPark2017], accident severity [@benevenutoExaminingTransportNeeds2020; @appleyardTransitCorridorLivability2017], and user-groups, especially disadvantaged ones with respect to vertical equity [@prasertsubpakijEvaluatingAccessibilityBangkok2012] of multi-criteria indicators. Some focus explicitly on affordability and barriers, proposing infrastructure enhancements for better inclusivity, especially for the most disadvantaged [@basuPracticalFrameworkBenchmarking2019; @songWomenWheelsGender2019; @welchEquityTransportDistribution2013], grounded in transport-related social exclusion [@kentPrioritizingLowstressEquitable2019] or capabilities approaches [@smithAccessibilityCapabilityMinimum2012].

Our corpus minimally focuses on "Environmental+" standards. These studies predominantly examine "inequitable externalities", possibly because environmental burdens of transportation are addressed more broadly in other literature (e.g., environmental justice). However, these papers included in our corpus present some interesting insights. They often use traffic-related air and noise pollution, green space, urban design, urban air temperature, health outcomes, and physical activity guidelines to assess transport-related externalities. Methods employed are primarily quantitative or mixed-methods, identifying inequalities through Gini coefficients [@fengTradeoffsMobilityEquity2014] and some develop composite indices [@agostfelipInclusiveModelAssessing2021; @mirandaBenchmarkingSustainableUrban2012; @corazzaMethodologyEvidenceCase2020], occasionally incorporating spatial analysis [@jephcoteGeospatialAnalysisNaturally2013; @carrierApplicationThreeMethods2014]. Commonly, many use established thresholds or health guidelines e.g., WHO guidelines for Active Aging and targets included in the United Nations’ SDG [@agostfelipInclusiveModelAssessing2021], OECD or WHO thresholds for traffic noise level, $NO_2$ levels, available green space and $PM_{2.5}$ level [@kruizeEnvironmentalEquityRole2007; @iungmanImpactUrbanTransport2021; @apparicioCyclingOneMost2021; @khomenkoLiveableCityHealthy2020; @muellerSocioeconomicInequalitiesUrban2018]. These metrics are sometimes criticized for their general applicability. Nonetheless, their use provides interpretable values for tracking progress, offering comparability across communities, unlike accessibility measures, which vary in methods and assumptions.

# Moving forward: calls for action {#sect4}

Justice is a goal of social progress. To understand what that goal is, clearly defining standards of fairness is essential. Measuring context-specific evidence of transport inequities is necessary to build knowledge of injustices and inspire creative interventions that advance the cause of justice. Understanding and tracking transport inequities over time and space, specifically knowledge of "Where" (spatial context), "When" (temporal context), "Who" (the subject of justice, i.e., population groups and mobility tools they use), "What" (the object of justice, i.e., what benefit/burden is produced by the system and its distribution), "How" (the conceptualisation and standards used) and especially "Why" (the rationale and not included in the focus of this review) is essential. Armed with this knowledge, policy can be designed to address these inequities while remaining adaptable to evolving justice demands.

Flexible frameworks are important as transport is derived demand. What is (in)formally considered (in)essential today, could change during the next pandemic or technological revolution. Creating adaptable transport equity frameworks that lend themselves to change with the demands of justice is as important to ensure resiliency to backsliding. From the diverse corpus reviewed, we argue for the necessity of setting standards from a systems approach, considering both _positive_ rights (a right to have access to sufficient quality of essential services, for example) and _negative_ rights (mobility of cars must be limited to reduce their impact on air quality, health, safety, and positive rights).

In this spirit, this work makes two significant contributions. It outlines a conceptual and flexible "Where", "When", "Who", "What" and "How" framework with supporting definitions for structuring transportation fairness and equity analysis. It also applies the framework in the synthesis and critical appraisal of 165 academic articles that undertake standard- and conceptualisation- supported equity analysis; these studies use equity analysis as instruments to gauge proximity to their context's just situation. Following this framework and the reviewed literature it answers:

-	**What** are those benefits and burdens of transportation systems? 
-	**Who** benefits and is burdened by transportation systems? Along which lines of identity are transportation inequities examined?
-	**Where and when** are equity standards applied?
-	**How** are equity standards and inequities measured? 
<!--Maybe we should answer these questions explicitly here? Maybe we should also urge practitioners to adopt this flexible framework in their equity synthesis?-->
From what's reviewed and using the framework we recommend five calls to action.

## Call 1: Underpin standards with rigorous concepts of justice {#sect4.1}

The conceptual grounding for standards is often left implicit within the literature: for example, @muellerSocioeconomicInequalitiesUrban2018 suggests the relative risk of mortality from transport-related air pollution should not be higher in deprived groups than the general population. While the coonceptual justice foundations are not explicitly declared, we infer the standard conceptualizes _wellbeing_ and could be in service of reparative justice. To move equity analysis outputs towards just transport futures, explicit inclusion of justice rationales (the "Why?") should become more common place. We must be clear with our terms – what is equity and for whom?

Standards serve as benchmark for fairness, yet some standards are seemingly arbitrary in the reviewed literature. For example, @caoIndicatorsSocioSpatialTransport2017 proposes 20 ferries per day to avoid social exclusion for inter-island transport planning in the Philippines, though acknowledging the standard should be politically determined. It is unclear from the paper however, if selecting a different benchmark such as 10 or 30 ferries would make a difference in any specific object of justice (e.g., accessibility to particular destinations) or if that number is tied to funding or resource constraints. As another example, what is the conceptual underpinning of the 15-minute city, where 15 minutes is the standard for travel times. Is it sufficientarism? Or is it egalitarianism? The standard can be interpreted in a multitude of ways. We argue that the framing used to conceptualize justice must guide the selection of the standard. As @martensFair2021 convincingly show, being explicit about "Why" is important.

For decision-makers, setting standards, measuring inequities and moving towards setting flexible guidelines for standards that are also compatible with community-informed calls for justice is the next step for transportation equity planning. If we assume that one function of the academic literature is to recommend standards, researchers must connect compatible equity conceptualisations and standards to justice frameworks. Currently, these fundamental connections are sorely needed.

## Call 2: Develop creative methods for systems-thinking approaches to equity {#sect4.2}

On the methodological side, transportation equity research would benefit from wider use of mixed-methods. Concepts and standards are often discussed from purely qualitative or quantitative approaches. This is a missed opportunity to combine the strengths of both approaches, whether by deep diving into some particular experiences or perceptions through qualitative methods or tailoring more meaningful quantitative analysis after qualitative explorations. As mixed-method examples, @xieWonCycleRoute2018 find through interviews and go-alongs with women cyclists that the standard Cycling Level of Service (CLS) tools used by engineers to plan cycling infrastructure misses the critical gendered perspective. Further, @somenahalliAgingTransportMobility2007 survey older adults to understand their mobility issues, revealing factors that are unseen in standard daily travel surveys.

While disparity analysis is frequently used in the reviewed literature, the standards suggested do not often align with practical application. For example, metrics of accessibility (usually measured with travel impedance cut-offs of between 15 to 60 minutes depending on the destination, population group and mode) are used to show descriptive differences among areas and groups but with scarce implications as to the experience of travelers. How low of an accessibility index is too low? Relatedly, given sufficient accessibility levels are reached, excessively high accessibility values will result in high inequalities. Are inequalities of a certain level an issue? This lack of discussion makes translating results from these equity analysis into policy practice that may move conditions towards justice challenging. Creative methods and discussion of metrics of quantitative accessibility with intention of practical intervention should be paired with results; they should yield interpretable results and the explicit discussion of minimum _and_ maximum values in the distribution of the object of justice (the "What?"), as applicable, is critical.

Other times, when analysis engages with metrics that may be tied to particular concepts of equity (like Gini coefficient or Theil index), they fall short in assessing whether the results are good or bad [e.g., @mijaresEquityAnalysisUrban2013]. If a Gini coefficient of 0 means that all people have the same access to public transport stops, what does it mean if the result is 0.2, 0.3, or 0.4? Is this good or bad news? How should a decision-maker interpret this? Are new policies needed to reduce that number to a certain threshold, orienting future interventions? These questions usually remain unanswered in the literature despite its importance. These measurements can also bring some challenges and pitfalls (as recently summarized by @karnerAdvancesPitfallsMeasuring2023) but are necessary to move equity analysis into application and towards transport justice.

## Call 3: Making data available is a matter of justice {#sect4.3}

In the review of the literature, we were left wondering specifically - what are the motivations for the use of some destination-types and the choice to not include other destinations types? For instance, papers including leisure destinations (e.g., green space, parks, recreation) are infrequently studied and some categories are missing all together. We suspect that since methods used are predominately quantitative, the reliance on commonly used point of interest databases is also high. These databases typically include education, health and occasionally include aggregated categories for leisure and community. They are quiet on types of 'community' and 'leisure' (are they a community organization, government services, a visit to a family or friend, a care-giving destination). Further, they may be missing categories all together like childcare, typically daycare or facilities -- domestic work, mobilities of care, and mobility interdependence. These missing categories are unrepresented in the reviewed relative to the presence of work and healthcare destinations.

We can only track what we have collected and compiled; and we know that transport systems' focus is more than just to work or as a source of economic development (though in underdeveloped regions, transport systems as a force of economic development is pronounced based on the concentration of Global South transport-induce economic development studies e.g., high-speed rail [@chenImpactHighspeedRail2017; @monzonEfficiencySpatialEquity2013; @kimImpactsHighspeedRail2015]). As such, data availability matters, especially in the context of the Global South and rural geographies as fewer official data sources and public research resources exist relative to urban communities and areas in the Global North. As well as in the operationalization of emerging conceptual theories (i.e., sufficientariansm [@vanderveenOperationalizingIndicatorSufficient2020]) in equity analysis.

The calls for and relevant issues of data availability are not new, but they have at least three parts. What and who is the subject of justice? When/Where/How is it measured? And, who gets to consult and use the data? Deciding who is the subject of justice frames the data collection activity; if it's the mobility of those that do domestic work, the classification of who does this work and how it changes over time/space is fundamental. How we classify has implications for our understanding of just who is a subject of justice, as illustrated by the history of racial classifications in the US [see @leeRacial1993]. What methods we're using, under what spatio-temporal boundaries we conduct data collection, and who has (and doesn't have) access to the collected data can all impact fulsome conceptualizations of justice. In the case of transportation, the issue of data collection/availability as a matter of justice is gaining traction as digitization of data casts a starker light on these questions [@sourbatiSmart2021; @behrendtData2023].

## Call 4: Develop more direct and explicit links between standards and lived experiences {#sect4.4}

Robust assessments of the implications of equity standards on the experience of the public is still lacking in the literature, but it is essential for equity analysis that translates into just practice. While the estimation of the benefits of increased mobility or accessibility, or reducing affordability burdens and transport externalities is commonplace in the literature, these estimations need to be associated with outcomes like life and neighbourhood satisfaction, subjective well-being, mental and physical health, social capital, among others.

Composite measures such as a transport-land-use index proposed by @appleyardTransitCorridorLivability2017 is an example of systems-thinking approach that links findings to quality-of-life proxies; they ground their measure in the principle of _livability_ along corridors of varying levels of estimated transport-land-use integration. However, the methods used demonstrate areas of relative high and low livability could go further; they could be tied to absolute goals of integration or livability. Relatably, @higgsUrbanLiveabilityIndex2019 develops an urban livability index and demonstrates the relationship between the population's use of a certain mode in a neighbourhood and a one unit increase in the index. However, are there absolute minimums or maximums for the index or the mode-choice goals that should not be crossed?

More explicit discussion of the boundaries in the distribution of the object of justice (the "What?") alongside these creative methods are needed. These links may be used to track progress towards justice across time and space, a critical point for practitioners.

## Call 5: Rigorously evaluate interventions and policies {#sect4.5}

We believe there is a need to evaluate equity interventions and policies: track their before, during, and after impacts. In this review, only approximately 10% of studies assess specific transport-related policy interventions through an equity lens <!-- double check this number-->. Examples include mode-shift from driving to active school travel [@mammenSchoolTravelPlanning2014], transit fare restructures [@hickeyUsingQuantitativeMethods2010] and spatial analysis of Low Traffic neighbourhoods [@aldredEquityNewActive2021]. The assessment of interventions and tracking associated outcomes can be thought of as a key step towards transport justice.

Outcomes of interventions can be compared within and between communities, and cross community comparisons can be created that may expedite the adoption of effective policies that move towards just outcomes; the presence of these synthesis and comparative studies could support brave decision-makers in the application of research into practice.

<!-- # Final remarks {#sect5} -->

<!-- To move the political needle and establish a common understanding of justice, it is critical that we have the concepts, the vocabulary, and the knowledge to express our intuitions. We argue that transportation, as an essential technology that can facilitate or hinder participation in society, is a very relevant sphere for matters of justice. The analysis of fairness and equity in transportation is vitally important for planning. Transportation systems and technologies converge space-time, and by definition they _always_ leave certain populations, depending on the area and time scale, more or less burdened or benefited. The analysis and tracking of equity implications is fundamental in formulating standards that are flexible to the fluid demands of transport justice. -->

<!-- In this spirit, this work makes two significant contributions. It outlines a conceptual framework with supporting definitions for structuring fairness and equity analysis that may move us closer to transport justice. Then using this framework, we critically appraises findings from 165 relevant studies from the academic literature and recommend five calls to action. -->

<!-- Defining _fairness_ is complex, but complexity is no excuse to shirk from the task. We hope this work contributes to the promotion of the idea that all are deserving of "access to the opportunities they need to lead a meaningful and dignified life" [@karnerTransportationEquityTransportation2020, p. 440]. This idea should guide our justice conceptualisation. Likewise, we hope that in reviewing the literature, this work will contribute to advance the development and implementation of equity standards that help bend the arc of history towards more just transportation systems, and ultimately better societies. -->

<!-- `r word_comment(" - A CONSIDERATION FOR THE FUTURE - What do you think - do we provide a 'framework' for practioners/researchers? We perhaps provide a theoretical framework to think through these questions that they could follow + definitions. However, in our work we use who/what/where/when/how to organise our review. As it stands, we don't demonstrate this frameing could be used to organize a study/research project - like a worked example of how 1 specific paper fits into each category. This could be solved with a sub-section that demonstrates how one paper fits into each category (I know we have a few examples in the table in the appendix but there's no room in the paper and explain it).", ".")` -->

<!-- The transportation justice framework can be seen as a flexible classification exercise that is supported by a "Why" rationale. In conducting an equity analysis, the subject of justice (the "Who?") must contextualize the process, the distribution of the object of justice (the "What?") must be measured, be it transport-related benefits, burdens or both, "How?" these inequities are measured and benchmarked must be described, and these questions must be contextualized by the "Where?" across temporally changing "When?"s. The explicit acknowledge and answers to these questions clarify the boundaries of the equity analysis. In this way, the rationale for "Why?" the equity analysis is important and what root cause of/for (in)justice is motivating the analysis in the first place may be critical to ultimately mobilizing equity analyses into experienced transport justice. -->

<!-- From this framework and the critical appraisal of the literature, we outline five call to actions for researchers and decision-makers that should be adopted to move equity analysis closer to transport justice. These include: -->

<!-- - Explicitly explain "Why?", the rationale for justice, along with supporting "Where", "When", "Who", "What, and "How" framings for equity analysis. This means specific conceptualizations and grounded standards. -->
<!-- - Creative methods of equity appraisal that can be paired with systems thinking; current methods are not used to support this processes. -->
<!-- - Data availability as a matter of justice: who is not being counted and who should be by whom and why?  -->
<!-- - More explicit links between equity standards and experienced conditions and outcomes.  -->
<!-- - Encouragement to continue work towards analysis and comparison of applied interventions, across communities and internationally.  -->

**Declarations:**

During the preparation of this work the authors used ChatGPT exclusively for copyediting: aiming to reduce word count, keep original style and correct typos. After using this tool, the authors reviewed and edited the content as needed and take full responsibility for the content of the publication.

\newpage

# References

::: {#refs}
:::

# Appendix {#sec-sect6}

## Why? as part of the framework to analyse questions of justice {#sec-sect61}

To give an example drawing from the Canadian Charter of Rights and Freedoms, a number of rights and fundamental freedoms [including the liberty to move freely; see @chartopediaArt72023] are recognized to exist in Canada "without discrimination by reason of race, national origin, colour, religion or sex". Notice that this declaration includes several individual traits that define the subjects of justice, but does not consider age as one aspect of the person. Does this mean that they apply universally irrespective of age? Surely not, since no reasonable person would consider a toddler's demands for freedom of association as absolute. So, when do these rights fully apply in the life of a person? In 2017, Adrian Crook of Vancouver, B.C., was warned by the province's Ministry of Children and Family Development that his kids could not be out of home in the community, alone or with other kids the same age, without supervision. The Ministry's argument was that children riding the bus unsupervised compromised the parent's ability to provide care and placed them at risk [@brendHere2017]. Crook argued that the goal of teaching his kids (then aged 7, 8, 9 and 11) to ride transit to school was to raise self-reliant children [@stueckYoung2017]. While both sides discovered to everyone's surprise that no clear rules existed regarding children's independent transit use [@brendHere2017], statistically there is evidence that travel by car is riskier than travel by bus [@morencyTraveling2018]. If the public good is to reduce risk, shouldn't children be banned from riding in cars? The Canadian legal system concerns itself with fundamental freedoms and rights, one of which is the liberty to move freely. This provides an answer (of many) as to "why" we would even consider children as subjects of justice. But the devil is in the details, and answering the other questions discussed above is essential to pin the devil by the tail. Are children under the age of 10 subjects of justice (i.e., are they a "who") and who decides if they are (i.e., the courts or the parents in this case)? If so, is the ability to use transit unsupervised an object of justice that should be justly distributed (i.e., the "what") and under what circumstances (e.g, to school, to a social event; during the day, or late at night?). "When" (e.g., for all children going forward, for only 5 years) and "where" (e.g., only in the suburbs, across the country) do these questions apply? Eventually, in 2020, a court of appeals ruled in Crook's favor, allowing his children to continue to ride the bus, thus providing much needed clarity with respect to the use of transit by children who desire and are encouraged to exercise independent mobility [@stueckCourt2020]. In actuality, the courts _created_ a standard of equity: starting at a certain age, children are objects of justice from the perspective of freedom of movement, and they are due the benefits of independent mobility by transit.

Another perspective on the answer to "Why?" are cultural _norms_ that provide a rationale. In this case, the liberty of movement for all---including children---reigns supreme. But a competing concern is that the outside is dangerous, and it is so in part due to vehicular traffic [@morencyTraveling2018], and in part because of potential violence from others. Parents may have over-pronounced 'stranger danger' based on neighbourhood and socio-economic characteristics of neighbourhoods [@francisLlBeDriving2017] and children themselves may fear potential violence based on real threats of victimization [@painParanoidParentingRematerializing2006]. Whose voice, what perception of fear, and actions we take to intervene define and reproduce cultural norms, all play into the rationale behind the question "Why?".

## Evidence search strategy {#sect5-2}

```{r PRISMA-figA1}
#| label: fig-figA1
#| fig-cap: "PRISMA flow diagram for the evidence selection process. ES signifies equity standard and EC signifies equity conceptualization."
knitr::include_graphics("figures/Pflow_chart.png")
```

The evidence selection process is also represented using the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) flow diagram [@pagePRISMA2020Statement2021] in @fig-figA1. Notably, two rounds of exclusion occurred during the assessment for full-text eligibility. 1710 studies entered step 2, 1223 were excluded and the remaining 487 papers entered step 3. The data extraction template used by the reviewers (authorship team) in step 3 revealed that, as expected, inclusion was initially too generous, and some papers were not sufficiently relevant, because of a lack of content on standards and/or conceptual/theoretical elements. In this fashion, 322 papers were further excluded and data extraction was completed to give a final corpus of 165 papers. A summary of the reasons for exclusion of the 1545 papers (between steps 2 and 3) are included in @fig-figA1.

```{r Search-query-A2}
#| label: fig-figA2
#| fig-cap: "The search query. TS = topic search (keywords, abstract, title). TASCA = subject categories. Green text area transportation system related terms, blue text are equity dimension related terms, purple text are equity/justice conceptualization related terms, and orange text are standards related terms. Hits corresponds the number of papers that the search yielded and was retained into the evidence selection process."
knitr::include_graphics("figures/Search-query.png")
```

Definitions of the population-concept context (PCC) used in the creation of the inclusion and exclusion criteria for the search strategy.

-   **Population**: the focus of the included studies should be on individuals, groups, communities, or entire regional areas that are impacted by passenger transportation infrastructure and systems (i.e., all modes and flows) from the perspective of equity (i.e., fair distribution, production, and re-production of burdens and benefits). This criteria is reflected in the creation of the first set of topic search terms that relate to transportation modes (e.g., "walking" OR "cycling" OR "transit" - see green text in @fig-figA2 for the full list).
-   **Concept**: the included studies should also include equity dimensions and conceptualizes equity as discussed in the previous section. This inclusion criteria is reflected in the second and third set of topic search terms developed in the search strategy. These terms relate to types of equity dimensions (e.g., "accessibility" OR "mobility" or "transport-related air pollution" - see blue text in the @fig-figA2 for the full list) and equity conceptualizations (e.g., "Justice" OR "equity" - see purple text in @fig-figA2 for the full list).
-   **Context**: the included studies should also be limited to publications that include equity standards. Context can be more difficult to explicitly search for with key terms so synonyms for 'standards' were added to the query as a four set of topic search terms (e.g., threshold, indicator, criteria - see orange text in @fig-figA1 for full list). Additionally, journal article and conference papers, English-language literature from any country, any study design (e.g., quantitative, qualitative, or mixed-method studies, or conceptual frameworks), and any record published within the past 30 years are included (January 1992 to March 2022). The time period is selected as the first (to the authors knowledge) peer-reviewed article which operationalized equity standards and equity conceptualization was published in 1996 [@khistyOperationalizingConceptsEquity1996]; we are broadening the search by a few years for completeness. English is selected as it is the common language spoken across the authorship team. Furthermore, papers that explicitly fall within the Transportation or related topic/category is included in the query (e.g., "Transportation", "Social Sciences", "Geography", "Civil Engineering", "Philosophy" - see the @fig-figA2 for full query).

The **exclusion criteria** for the search are papers that are not within the inclusion criteria.
Specifically:

-   Literature published before January 1992.
-   Papers which do not include transportation equity dimensions.
-   Grey literature, as concepts contained within are frequently published in a more developed form in journals.

## Evidence selection and data extraction {#sect5-3}

```{r methods_framework-fig1}
#| label: fig-figA3
#| fig-cap: "Evidence selection workflow. Step 1 (orange) is title and abstract screening, step 2 (green) is full-text review, and step 3 (purple) is data extraction."
knitr::include_graphics("figures/Methods_framework.png")
```

The following steps summarize the evidence selection process:

1. The first step (orange box in @fig-figA3) included screening all titles and abstracts of papers on whether they included transportation equity as defined by the PCC. Each paper was screened by two independent reviewers who then voted for inclusion, exclusion, or uncertain inclusion. All uncertain papers, conflicting papers, and papers missing abstracts were reviewed by a third person for inclusion or exclusion.

2. The second step (green box in @fig-figA3) included scanning all full-text papers which passed step 1. These papers were reviewed to determine if they included a relevant "how", i.e., an standard and/or relevant theoretical or conceptual discussion. At this stage, papers were evaluated again by two independent reviewers who voted for inclusion or exclusion. If an article was voted to be excluded, it was tagged with one of five possible reasons for exclusion, namely (1) no standards included; (2) no relevant conceptual elements included; (3) no standard and no conceptual elements included; (4) send back -- QA issue; or (5) other. Discrepancies were resolved by a third reviewer.

3. In the last step, a data extraction template for each record was filled by one reviewer (purple box in @fig-figA3). The data extraction template was created with the aim of striking a balance between the complexity of categories and the simplicity of summary; information related to "Where?" (the geographical context and sphere of life), "When?" (temporal circumstances for the application of justice), "Who?" (the subject of justice), "What?" (the object of justice), and "How?" (equity standard(s) and concepts) was filled out for each study. @fig-figA4 contains the template that was input into *Covidence* and used throughout.

Data extraction for each document that passed through all three steps was then extracted using this template:
```{r data-extraction-template-A4}
#| label: fig-figA4
#| fig-cap: "The data extraction template with associated defintions."
knitr::include_graphics("figures/Data-extract-template.png")
```

To provide further example of data extraction, a table of the resulting literature looked like:
```{r}
dat <- data.frame(matrix(c("What?: **mobility and accessibilty**",
                           "Where? [@rivasHowAffordableTransportation2018] - South America (select cities)",
                           "Analyses how affordable urban public transportation is in select Latin American and Caribbean countries. They look at the estimated average monthly cost of transit trips and average monthly household income and conceptualize **transportation-related** **affordability**, especially for the most economically vulnerable (**vertical equity**).",
                           "How?: The financial burden of a basket of urban public transportation trips (60 trip fares, representing 30 round-trips per month) should not exceed 10% of household monthly income.",
                           "What?: **mobility and accessibilty**",
                           "Where?: [@bharathyRevisitingClearFloor2018] - North America (USA - National)",
                           "This study designed a web-based tool and took a representative sample of wheeled mobility device (WhMD) users anthropometry measurements to determine if the minimum standard suggested by the ADA is sufficient. We understand this conceptualization as a type of **Rights** conceptualization that WhMD should have minimum clear floor space (as described the guidelines in line with the American Disabilities Act) to access bus shelters, bus stop pads, and transit terminals.",
                           "How?: The clear floor area for wheelchairs: 760 mm (30 in.) wide by 1220 mm (48 in.) in length as described by the ADA standards. Of note, this minimum clear floor area is insufficient for a variety of the WhMD users.",
                           "What?: **mobility and accessibilty**",
                           "[@ryanWhatAreWe2021] - Europe (Stockholm, Gothenburg and Malmo cities in Sweden)",
                           "Investigates what the literature and planning process is missing when we measure accessibility by comparing objective and self-reported accounts of accessibility among older people. This paper conceptualizes accessibility as from the position of the **capabilities approach** and **vertical equity** (particularly acknowledging that older people have capabilities that differ from the general population).",
                           "How?: Specifically for older populations (aged 65+), the following travel distances are suggested as equitable trip lengths to grocery stores per mode: Walking: less than or equal to 1500m, Combined public transit and walking (less than or equal to 1000m (walking element)), Combined car and walking: less than or equal to 1000m less than or equal to 1000m (walking element)), Bicycle: less than or equal to 3000m in addition to travel time threshold of less than 15 mins.",
                           "What?: **mobility and accessibilty**",
                           "Where?: [@wismadiSpatialPreferenceModelling2014] - Asia (Yogyakarta, Indonesia)",
                           "Explores the equitable provision of transport infrastructure provision: an application of Sen's **capability approach**. Conceptualizes equity through Sen's capability approach and spatial equity.",
                           "How?: Areas below the relative poverty line (of its neighbours) can only be located transport resources (i.e., measure in person*kms that can be travelled at car speed, i.e., mobility) based on the following 2 benchmarks (they can be considered, together as the floor/minmum access): 1) Global: standard deviation (SD) distance to mean should be minimized. 2) Local: priority to minimise the differences with its neighbourhood",
                           "What?: **mobility and accessibilty**",
                           "Where?: [@zhengAreabasedEquitablePricing2020] - North America",
                           "This paper conceptualizies equity in the multimodal network (transit, car) being fair toll-pricing across differences in populatins value of time (VOT). VOT is determined based on household income, with lower income households having lower VOT and thus deserving of lower tolls (vertical equity). From this perspective, a utilitarian perspective that seeks to minimize multimodal traffic congestion through introducing toll-pricing based on VOT is implemented.",
                           "How?: suggest that a toll-pricing scheme based on individuals travel value-of-time (lower income people have a lower VOT) is equitable.",

                           "What?: **environmental pollution**",
                           "Where?: [@carrierApplicationThreeMethods2014] - North America (Montreal, Canada) ",
                           "This work examines the statistical association between different social groups and the concentration of air pollutants. They frame their work from the perspective of environmental equity. We interpret the conceptualizations to be along the lines of **inequitable externalities**, **spatial** and **vertical equity** - transport-related air pollution is a product of road transport and it impacts the air of residents in unequal spatial ways. The paper then frames this impact as unfair, particularly from the perspective of disproportionately disadvantaged residents",
                           "How?: The literature suggests that the health implications from the transport-related air pollution from major roadways is most acute at residential distance locations of 200 m or less. Residential locations should not be located within this distance threshold from the perspective of human health. **Environmental+** and **Population standards**: Uses the WHO NO² threshold as a point of comparison (annual concentrations of NO² should not exceed 40 μg/m-3). They argue that even through no neighbourhood, even those disproportionately low income, exceed the WHO limit in this case study, they still suggest that air pollution should not be disproportionately impacting disadvantaged neighbourhoods. It can be interpreted that they use the WHO threshold as a minimum threshold and suggest that air pollution levels should not be impacting disadvantaged populations disproportionately ( a relative population standard)",
                           "What?: **environmental pollution**",
                           "Where?: [@jephcoteGeospatialAnalysisNaturally2013] - Europe (Leicester, UK)",
                           "Geospatial analysis of naturally occurring boundaries in road-transport emissions and childrens respiratory health across a demographically diverse cityscape. Emperically identifies at what distance away from major roadways children are most impacted by transport-related pollution. This is framed in the perspective of children’s **well-being**. Children are at most risk for acute respiratory distress from elevated levels of air pollution, and as such planning should consider this point of public health.",
                           "How?: Finds that children (most vulnerable to air pollution - related to motoized traffic) are most impacted by air pollution within 283 m of a road way. This should be the distance threshold that schools and other childrens facilities are located.",
                           "What?: **health impacts**",
                           "Where? [@adlakhaMindGapGender2020] - Asia (Chennai, India)",
                           "From the perspective of disparity in gendered physical activity, this paper focuses on women's cycling as both transport and exercise. They advocate for all people achieving physical activity thresholds (**horizontal equity**) but prioritize women and especially women in neighbourhoods with low-walkability and socio-economic status (**vertical equity**).",
                           "How?: All people should get 150 min of moderate activity a week or 75 min of vigorous physical activity per week.",
                           "What?: **health impacts**",
                           "Where? [@savingmothersDidSavingMothers2019] - Africa (Select urban and rural regions in Uganda)",
                           "The **well-being** of mothers, this paper examines the timely access to emergency obsteric and newborn care for child-bearing aged women in Uganda.",
                           "How?: 2 hours to the nearest facility with surgical capacity with anesthesia services - this threshold is determined through the onset of bleeding to death if a women with obstetric hemorrhage does not receive adequate treatment).",
                           "What?: **health impacts**",
                           "Where?: [@iungmanImpactUrbanTransport2021] - Europe (Madrid and Barcelona, Spain)",
                           "They use environmental pollution guidelines, but from the position of health. They investigate the impact of urban and transport planning on attributable mortality burden in Madrid and Barcelona and its distribution by socioeconomic status . Pre-mature mortality is linked to the exposure to pollution and motorized vehicles (**inequitable externalities**). These externalities should not be impacting people disproportionately (**vertical equity**) and should be even across space (**spatial equity**).",
                           "How?: All minimum thresholds, if exceeded this is inequitable: NO² concentration 40 μg/m³; PM 2.5 concentration 10 μg/m³; Noise 53dB for average 24 hours; Living with 300 m crow-flies distance from at least .5 hectares of greenspace; and a Change of air temperature of at least 1 ⁰C. ",
                           "What?: **health impacts**",
                           "Where?: [@mehdizadehWalkingTimeSchool2017] - Asia (Rasht, Iran)",
                           "From the perspective of children’s **well-being**, assesses the walking time to school. They frame walking to school as health-related.",
                           "How?: perceived walking time to school for students aged 7-9 yrs is 10 mins, and the longer the PWTS the less likely they were to use an active mode to travel to school.",
                           "What?: **health impacts**",
                           "Where? [@murphySupermarketAccessTransport2017] - Oceania (Melbourne, Australia)",
                           "Assesses the relationship between supermarket access and transport mode used, the body mass index (BMI) of the mode-user (**wellbeing**) and the equity in access distribution by income (**vertical equity**).",
                           "How?: all households should be sufficiently active (greater than 150 min and at least 5 sessions) and households should be within 1 km euclidean distance to supermarket (80-90% of the dwellings should meet this). Planners should prioritize socially disadvantaged areas to meeting these standards first.",

                           "What?: **transport-related safety**",
                           "Where?: [@ferenchakEquityAnalysisProactively2019] - North America (Denver, USA)",
                           "Operationalizes and compares an equity analysis of proactively- and reactively-identified traffic safety issues from the perspective of **Spatial equity**, **Vertical equity** and **Inequitable exposure to externalities**.",
                           "How?: standards are suggested for both reactive and proactive analysis. First, the lower the number of collisions on the road with pedestrians/cyclists (i.e., reactive safety analysis), the better. No/minimal inequalities for general population vs. equity seeking groups (high proportion of POC and/or low income in tract). Second, the lower the perceived safety, the better (i.e., if travel to school by ped. or bike is unsafe due to traffic conditions). No/minimal inequalities for general population vs. equity seeking groups (high proportion of POC and/or low income in tract).",
                           "What?: **transport-related safety**",
                           "Where?: [@zheEvaluationSharedUse2008] - Asia (Tokyo, Takamatsu, and Tokushima)",
                           "Evaluates the observed safety of  shared use pedestrian and bicycle paths from the perspective of **well-being**.",
                           "How?: the study suggests that the safety threshold for bicycles and pedestrians to coexist on shared infrastructure is less than 0.5 pedestrians/minute per metre of sidewalk (width) and less than 3.0 cyclists/minute per metre of sidewalk (width). The standard for pedestrian/bicycle share use in terms of hourly traffic volume is less than 26 pedestrians / hour and 108 cyclists / hour for 2m wide sidewalks.",

                           "What?: **mobility/accessibility and health impacts**",
                           "Where?: [@aldertonWhatMeaningUrban2019] - Asia (Bangkok, Thailand) – **Mobility/ accessibility** and **health**",
                           "Establishes short-, medium-, and long-term goals for the city in collaboration with technical leaders within the municipal government for the perspective of **well-being** (urban livability): the standards included in this table relate directly to transportation systems. Indicators are inspired by the Sustainable Development Goals (SDGs) as well other global planning standards.",
                           "How?: 1) Green space: % of residents living < 400 m from public open space, a large park (> 1.5ha), and/or local park, 2) Public transit access: % of residents living < 400 m of a local bus stop and <800 m of train station, 3) Facilities: % of residents living < 400 m of a community centre. The following **Infrastructure standard** is suggested: Canal water quality - dissolved oxygen content of equal to or less than 2.0 mL/L",
                           "What?: **Mobility/ accessibility** and **health impacts**",
                           "Where?: [@berheAdaptationDissonanceQuality2014] - Africa (Mekelle, Ethiopia)",
                         "Examines adaption and dissonance in the quality of life (QoL) of residents. QoL is conceptualized along the lines of **well-being** and aspects of QoL directly tie into transport systems. They conduct a qualitative QoL survey of residents on the topic of three QoL domains: housing quality, access to important destinations, and affordability. They also measure quantitative indicators associated with these domains. We assume the equity goal for this paper is that subjective and objective QoL measures should not be mismatched: as discussed by the authors of this study, subjective QoL is higher than objective QoL the participant is experiencing adaption and in the reverse scenario the participant is experience dissonance.",
                           "How: 1 & 2) Access to primary or secondary education facility, percentage of households living within 1 km or 2km (walking distance), respectively from a primary school or secondary school. 3) Access to health facility, percentage of households within 40 min walking time from a health facility. 4) Access to public transport, percentage of households within a distance of 500 m from a mini-bus stop. **Population standards**: 1) Adequate family income, percentage of households earning more than the official poverty line. 2) Subjective QoL is constructed based on the households level of satisfaction for each of the eight indicators using a six point Likert-scale (1=very satisfied to 6=very dissatisfied).",
                           "What?: **mobility/accessibility, health impacts, and safety**",
                           "Where? [@agostfelipInclusiveModelAssessing2021] - Europe (Castellon, Spain)",
                           "Conceptualizes equity through age-friendly urban spaces that reduce (and eliminate) conditions for **transportation-related social exclusion** for older populations and prioritize those who are economically vulnerable (**vertical equity**). These guidelines are inspired by the SDGs in addition to planning guidelines used national, regional, and local guidelines used in Spain.",
                           "How: 1) Access to facilities needed for old age health. Minimum distance thresholds from the geometric center of neighbourhood are suggested: at least: 1000 m from health facilities (600 m or less is preferred), elderly-specific care facilities and shops should be 600 m (300 m or less is preferred). **Population standards**: 1) Certain neighbourhoods should be prioritized above others. From this papers focus on age-friendly urban environment, they suggest that if the neighbourhood has an average old age indicator (i.e., greater than 64 years, and/or greater than 79 years, and/or aging ratio of persons aged greater than 64 relative to 15 to 64 age) should be prioritized. 2) Economic vulnerable and non-civically engaged neighbourhoods should also be prioritized. If the neighbourhood has a lower percentage of civic associations within the neighbourhood than average, and/or household income, and/or a higher than average interventions for dependency and/or social subsidies, they should be priorized. **Infrastructure standards **: 1) Green space: should be at least 10 m2 per inhabitant in the neighbourhood, greater than 15 m2 per inhab. is the goal. 2) As related to sidewalk infrastructure at least 50% of all sidewalks (preferably 75% or greater) should: have a width of 1.5m or larger, ramps should have a grade of 8% or less, be well maintained (free from deficiencies), be paved for pedestrian use, and cover public transit stops. 3) Lighting is critical for traffic-safety and a sense of safety overall. As such, at least 50% roads should: have a min. of 35 lux (road traffic) and 20 lux (pedestrian streets), and adapted traffic lights. 4) Buildings should be age-friendly. As a proxy for the quality of residential living space quality, at least 50% of residential buildings in a neighbourhood should be built within the last 50 years (preferably 75% or more). In terms of physical access into the buildings, at least 10% should have elevators and accessible entrances (preferably 25% or more). **Environment + standards **: 1) Noise at the street level should be less than 55 dB and 45 dB (but preferably less than 50 dB and 40 dB) in the daytime and nighttime, respectively.",
                           "What?: **mobility/accessibility and safety**",
                           "Where?: [@mateobabianoPedestrianNeedsMatter2016] - Asian (Manila, Philippine)",
                           "The perception of pedestrians’ walking environments should be sufficient across 6 themes. Equity is conceptualized around **spatial equity** (equally fair walking environments for all locations) and **rights** (the right to mobility/accessibility for pedestrians)",
                           "How?: perceived pedestrian perception on protection, ease, equitable access, mobility, identity, and enjoyment must be met."
                           ), ncol = 4, byrow = T))
```

```{r}
set_flextable_defaults(font.size= 7,
                       font.family = NULL,
                       text.align = 'justify')

flextable(dat) |>
  ftExtra::colformat_md() |>
  compose(
    i = 1, j = c(1,2,3,4), part = "header", value = as_paragraph(c("Dimension", "Continent", "Conceptualization", "Standard"))) |>
  merge_v(j=1) |> #merge 'dimensions' column
  rotate(part = "body", rotation = "btlr", j=1, i=c(1:5,6:7,8:12, 13:14,15:18)) |>
  valign(valign="center", part="body", j=1) |>
  hline( i = c(5,7,12,14), part = "body") |>
  set_table_properties(layout = "autofit")

```

### 2024-04-14 Manuscript/Manuscript.qmd
Draft of journal-manuscript

@@
 ---
-title: "Searching for standards of fairness in the transportation justice literature: A scoping review"
-subtitle: "Searching for standards of fairness"
+title: "Searching for standards of fairness in the transportation justice literature"
+format:
+  pdf: default
+  tandf-pdf:
+    keep-tex: true
+
 author:
@@
     email: paezha@mcmaster.ca
-bibliography: All_papers_in_extracted.bib
-format: docx
+bibliography: bibliography.bib
 ---
@@
 accessibility to specific destinations, or generally), with most studies focused on the latter.
-Over a quarter of the accessibility focused papers do not focus on access to specific destinations, but instead on the accessibility of the trip (i.e., level of service, barriers of infrastructure) or bundles of destinations.
-Studies that center particular destinations that should be reached/accessed include: employment, healthcare, grocery stores, and greenspace.
-Access to community support or places of worship, as well as childcare activities are less common in the reviewed literature.
+Over a quarter of the papers that focus on accessibility do not consider specific destinations, but instead concentrate on the accessibility of the trip (i.e., level of service, barriers of infrastructure) or bundles of destinations.
+Other studies are centred on particular destinations considered as essential, including employment, healthcare, grocery stores, and greenspace.
+Access to community support or places of worship, as well as childcare activities are far less common in the literature reviewed.
 
 <!--we suggest: "How": concepts of justice and standards of fairness. I like ours more - there is a distinction. -->
-**How: concepts and standards of justice.** The most common standards are defined using access to opportunities cut offs (37.2%), opportunity standards, or thresholds related to the perceptions or activities of specific populations (35.8%), population standards.
+**How: concepts and standards of justice.** The most common standards found in the literature, which we term opportunity standards, are defined using cut offs in access to opportunities (37.2%).
+Almost as common is the formulation of standards that use thresholds related to the perceptions or activities of specific populations (35.8%); we call these population standards.
 The latter often includes standards informed by public health, such as the number of physical activities undertaken in a week and perceived life satisfaction.
-Opportunity standards are evaluated through horizontal equity and spatial equity lenses, while population measures are evaluated through well-being frameworks.
-Another category that emerged in the literature are infrastructure-based standards, which include adherence to universal design guidelines or specific benchmarks for road network characteristics; they are less common in the literature comparatively and are often framed in rights-based terms, particularly as they relate to achieving universal design.
-But infrastructure standards around sustainable infrastructure (i.e.
-block lengths or distance between transit stops) have also been evaluated through vertical, horizontal and spatial lenses.
+Opportunity standards are evaluated through horizontal equity and spatial equity lenses, while population ones are evaluated through well-being frameworks.
+Another category of standards that emerged from the literature are based on infrastructure, and these include adherence to universal design guidelines or specific benchmarks for road network characteristics.
+Standards of this kind are comparatively less common in the literature and are often framed in terms of rights, particularly as they relate to achieving universal design.
+Infrastructure standards around sustainable infrastructure (i.e.
+block lengths or distance between transit stops) are also evaluated through vertical, horizontal and spatial lenses.
 The distribution of studies with respect to the kinds of standards deployed, and the broader concepts of justice used, are presented in the Figure below:
@@
 
-Mixed-methods research can help government set meaningful standards by better linking quantitative thresholds with qualitative assessments of well-being, freedom of movement and perceived access, among other dimensions.
-Yet most of the reviewed research is purely quantitative or, to a much lesser degree, purely qualitative.
-The constant deployment of metrics Lorenz Curves of Gini coefficients without research on where such measures need to be to improve quality of life (i.e., a standard), will not help governments with limited resources achieve the best possible equity outcomes within their respective contexts.
+Mixed-methods research can help governments set meaningful standards by better linking quantitative thresholds with qualitative assessments of well-being, as well as freedom of movement and perceived access, among other facets of transportation.
+Yet most of the research reviewed is purely quantitative or, to a much lesser degree, purely qualitative.
+The frequent deployment of metrics Lorenz Curves of Gini coefficients without research on where such measures need to be to improve quality of life (i.e., a standard), will not help governments with limited resources achieve the best possible equity outcomes within their respective contexts.
 Mobilizing Justice’s Priority Populations working group was inaugurated to advance new research in this area.
@@
 
-Approximately 10% of the reviewed studies assessed interventions.
-Delivering new equity-focused policies and programs that do not actually advance equity harms communities, metastasizes mistrust, and delegitimates future efforts to promote transportation justice.
+Approximately 10% of the studies reviewed aimed to assess interventions.
+Delivering new equity-focused policies and programs that do not actually advance equity harms communities, generates mistrust, and delegitimates future efforts to promote transportation justice.
 Furthermore, policies that benefit one group of people in one context may not benefit other groups in other situations.
@@
 For example, firms may or may not adopt lower standards for the emission of pollutants in poorer areas; it might be that poor people end up being relegated to areas that already had lower emission standards [@gouldsonFirms2006].
-Regardless of the cause, the result is the same: pollution tends to be worse were poorer people are [@delucaCode2012].
-
-Supporting the creation of (more) just transportation systems involves understanding the production, distribution, and management of transportation benefits and costs; how they are distributed; and what values are implemented (and by whom) in the form of standards [@pereiraDistributiveJusticeEquity2017a; @shellerMobilityJusticePolitics2018; @pereiraTransportationEquity2021].
+Regardless of the cause, the result is the same: pollution tends to be worse where poorer people are [@delucaCode2012].
+
+Supporting the creation of (more) just transportation systems involves understanding the production and management of transportation benefits and costs; how they are distributed; and what values are implemented (and by whom) in the form of standards [@pereiraDistributiveJusticeEquity2017a; @shellerMobilityJusticePolitics2018; @pereiraTransportationEquity2021].
 Thus, for this review we engage the literature with the following questions in mind: 1) what is our current understanding of the things that transportation systems do, for whom, and to whom; 2) what does the literature say about the distribution of the benefits and costs of transportation systems; and 3) what values are embodied in normative statements about said distribution.
 Ultimately, this review aims to collate the existing academic knowledge on the matter, and present it in a manner useful to support the development and implementation of standards for equity in transportation planning and policy.
-In this, we aim to provide relief even, or particularly, to planners in those places where calls for justice are explicitly made through legislation^[[As recently as -@martensFair2021, Martens and Golub note that Federal directives related to Title VI in the US "do not provide guidelines that can help agencies develop explicit standards to assess the distribution of accessibility benefits from projects or plans."]].
+In this, we aim to provide relief to planners, especially in those places where calls for justice are explicitly made through legislation^[[As recently as -@martensFair2021, Martens and Golub note that Federal directives related to Title VI in the US "do not provide guidelines that can help agencies develop explicit standards to assess the distribution of accessibility benefits from projects or plans."]].
 
@@
 # Setting the stage {#sect2}
-
-> "_Man's capacity for justice makes democracy possible, but man's inclination to injustice makes democracy necessary._"
-> ― Reinhold Niebuhr
-
-> "_Never forget that justice is what love looks like in public_."
-> ― Cornel West
 
@@
 A possible complication in the case of transportation is that disentangling the "who" from their mobility tools is not always straightforward.
-Clearly, a person is not their mode of transportation; however in practice, there are large segments of the population who live in situations where they cannot extricate themselves from the tools for mobility that they can use, either because they have driven themselves out of choices [see @laveryDriving2013], or have been driven out of choices by factors beyond their control (e.g., captive users of a single mode [@jacquesRescuing2012; @cherancheryInvestigatingPerceptionCaptive2018]).
+Clearly, a person is not their mode of transportation; however in practice, there are large segments of the population who live in situations where they cannot extricate themselves from the mobility tools that they can use, either because they have driven themselves out of choices [see @laveryDriving2013], or have been driven out of choices by factors beyond their control [e.g., captive users of a single mode @jacquesRescuing2012; @cherancheryInvestigatingPerceptionCaptive2018].
 In societies that have grown into transportation monocultures with a predilection for automobility [@millerCollaborative2011] there may actually be less choice about mobility tools than we would like to assume.
@@
 With complex interlocking parts (sidewalk, road, traffic sign, parking regulations), transportation systems produce _mobility_, the potential for movement.
-The realization of this potentials happens through travel.
+The realization of these potentials happen through travel.
 However, as the adage goes, travel is derived demand, which seems to hold for most (even if not all) situations [e.g., @mokhtarianUnderstanding2001; @redmondPositive2001; @paezEnjoyment2010; @whalenMode2013].
 For this reason, we cannot stop at considering mobility, but the ulterior goal of mobility, which is to reach destinations.
-In combination with land use systems (the spatial distribution of opportunities on the landscape), mobility produces _accessibility_, the potential to reach destinations.
+In combination with land use systems (the spatial distribution of opportunities on the landscape), mobility produces _accessibility_, the potential to reach places where activities that the traveler values take place.
 In this manner, we can think of the objects of transportation justice as being _proximate_ (the tools of mobility, mobility itself), and _ulterior_ (accessibility, opportunities for activity participation).
@@
 But the devil is in the details, and answering the other questions discussed above is essential to pin the devil by the tail.
-Are children under the age of 10 subjects of justice (i.e., are they the "who") and who is deciding if they are (i.e., the courts and the parents in this case)?
+Are children under the age of 10 subjects of justice (i.e., are they a "who") and who decides if they are (i.e., the courts or the parents in this case)?
 If so, is the ability to use transit unsupervised an object of justice that should be justly distributed (i.e., the "what") and under what circumstances (e.g, to school, to a social event; during the day, or late at night?).
 "When" (e.g., for all children going forward, for only 5 years) and "where" (e.g., only in the suburbs, across the country) do these questions apply?
-Eventually, in 2020, a court of appeals ruled in Crook's favor, allowing his children to continue to ride the bus, thus providing much needed clarity with respect to the use of transit by children who desire independent mobility [@stueckCourt2020].
+Eventually, in 2020, a court of appeals ruled in Crook's favor, allowing his children to continue to ride the bus, thus providing much needed clarity with respect to the use of transit by children who desire and are encouraged to exercise independent mobility [@stueckCourt2020].
 In actuality, the courts _created_ a standard of equity: starting at a certain age, children are objects of justice from the perspective of freedom of movement, and they are due the benefits of independent mobility by transit.
 
-Another perspective on the answer to "Why?" are cultural _norms_ in providing rationale.
-In this case, the liberty of movement for all - including children - reigns supreme.
-But reigns supreme, "Why"?
-The outside is dangerous, in part due to vehicular traffic [@morencyTraveling2018], and in part because of potential violence from others.
+Another perspective on the answer to "Why?" are cultural _norms_ that provide a rationale.
+In this case, the liberty of movement for all---including children---reigns supreme.
+But a competing concern is that the outside is dangerous, and it is so in part due to vehicular traffic [@morencyTraveling2018], and in part because of potential violence from others.
 Parents may have over-pronounced 'stranger danger' based on neighbourhood and socio-economic characteristics of neighbourhoods [@francisLlBeDriving2017] and children themselves may fear potential violence based on real threats of victimization [@painParanoidParentingRematerializing2006].
-Who's voice, what perception of fear, and actions we take to intervene define and reproduce cultural norms, all play into the rationale behind the question "Why?".
-In this way, analyzing the "Why?" in the reviewed literature is left out of this review as answers to "Why?" are not explicitly stated and challenging to infer implicitly.
+Whose voice, what perception of fear, and actions we take to intervene define and reproduce cultural norms, all play into the rationale behind the question "Why?".
+In this way, analyzing the "Why?" in the reviewed literature is not the focus of this review, partly because answers to "Why?" are not explicitly stated and challenge inference.
 As such, the focus of this report is on the standards of fairness that, combined with the use of equity analysis, can help us understand how better to move towards just transportation systems and better formulate answers to "Why?".
@@
 Pedestrians are the most extreme case of confluence between the "what" and the "who", since the mobility tool is the person's own body, with its inherent range of abilities.
-In papers that focus exclusively on walking, many use or develop walkability scores to explore perceptions of neighborhood quality [@evansAccessibilityUserNeeds2015] or pedestrian mobility with a focus on middle-aged and older adults [@towneUsingWalkScore2016], gender [@kimNeighborhoodEnvironmentWalkability2016], or urban peripherical regions [@blecicCapabilitywiseWalkabilityEvaluation2021].
+In papers that focus exclusively on walking, many use or develop walkability scores to explore perceptions of neighbourhood quality [@evansAccessibilityUserNeeds2015] or pedestrian mobility with a focus on middle-aged and older adults [@towneUsingWalkScore2016], gender [@kimNeighborhoodEnvironmentWalkability2016], or urban peripheral regions [@blecicCapabilitywiseWalkabilityEvaluation2021].
 These papers use 'walkability' as a way to measure the equity in its distribution.
@@
 # Final remarks {#sect7}
+
 We began this report with the history-making suborbital flight of Jeff Bezos and his fellow passengers, all of them very wealthy people.
-We wondered why coverage of the event, in particular Bezos declarations planetside, were not met with more critical questions: who are the subjects of the marvelous future that Bezos fancies his trip will contribute to?
+We wondered why coverage of the event, in particular Bezos' declarations planet-side, were not met with more critical questions: who are the subjects of the marvelous future that Bezos fancies his trip will contribute to?
 All of the human species or only those who can afford space tourism?
@@
 -   Papers which do not include transportation equity dimensions.
--   Grey as concepts contained within are frequently published in a more developed form in journals.
+-   Grey literature, as concepts contained within are frequently published in a more developed form in journals.
 

### 2024-05-10 Journal-manuscript/Journal-manuscript.qmd
Verision for AP review

@@
 ---
-title: "Searching for standards of fairness in the transportation justice literature"
+title: "Is it fair? Searching for standards of fairness in the transportation justice literature"
 format: docx
-
+csl: apa7.csl
 #   tandf-pdf:
@@
   DiagrammeRsvg::export_svg() |> 
-  charToRaw() |> rsvg::rsvg_png(file="figures/Pflow_chart.png")
+  charToRaw() |> rsvg::rsvg_png(file="figures/Pflow_chart.tiff")
 ```
@@
 df_all_sub_ob_just_reordered$ID = seq.int(nrow(df_all_sub_ob_just))
-```
-
-```{r}
+
+#adding color column 
+df_all_sub_ob_just_reordered <- df_all_sub_ob_just_reordered |>
+  mutate(colour = c("#FCBBA1", "#FC9272", "#FB6A4A", "#EF3B2C",
+                     "#D9D9D9", "#BDBDBD",
+                    "#E5F5E0", "#C7E9C0", "#A1D99B", "#74C476", "#41AB5D", "#238B45",
+                    "#F7FBFF", "#DEEBF7", "#C6DBEF", "#9ECAE1", "#6BAED6", "#4292C6", "#2171B5", "#08519C", "#08306B","#08519C",
+                    "#FCFBFD", "#EFEDF5", "#DADAEB", "#BCBDDC", "#9E9AC8", "#807DBA", "#6A51A3", "#54278F"))
+#c(RColorBrewer::brewer.pal(9,"Blues"))
+```
+
+```{r, echo=FALSE}
 df_all_sub_ob_just_reordered_plot <- ggplot(data = df_all_sub_ob_just_reordered, 
@@
            fill = sub_ob_j_type))+ 
-  geom_bar(stat="identity",  col = "grey", alpha=0.8) +
+  geom_bar(stat="identity",  col = "grey", alpha=0.8, size=0.05) +
   facet_grid(rows=vars(sub_ob_j),
@@
                       expand = c(0,0))+
-  scale_fill_manual(values=c(rep(RColorBrewer::brewer.pal(9,"Pastel1"),4))) +
+  scale_fill_manual(values= c("1992 to 2007" = "#FCBBA1",
+                              "2008 to 2012" =  "#FC9272", 
+                              "2013 to 2017" =  "#FB6A4A",
+                              "2017 to Mar. 2021" =  "#EF3B2C",
+                              "Global North" =  "#D9D9D9",
+                              "Global South" =  "#BDBDBD",
+                              "(Dis)Abilities" =  "#DEEBF7", 
+                              "Age" =  "#6BAED6", 
+                              "All populations" =  "#F7FBFF",
+                              "Comp. vul" =  "#C6DBEF", 
+                              "Education level" =  "#6BAED6", 
+                              "Employment 3.a" =  "#9ECAE1", 
+                              "Gender" =  "#DEEBF7", 
+                              "Income" =  "#4292C6", 
+                              "Other 3.a" =  "#9ECAE1",
+                              "Race/Ethnicity" =  "#C6DBEF",
+                              "Bicycle" =  "#A1D99B", 
+                              "Car" =  "#74C476", 
+                              "Other 3.b" =  "#E5F5E0", 
+                              "Pedestrian" =  "#41AB5D",
+                              "Transit" =  "#238B45",
+                              "Unspecified 3.b" =  "#C7E9C0",
+                              "Childcare" =  "#BCBDDC", 
+                              "Community" =  "#DADAEB", 
+                              "Education" =  "#EFEDF5", 
+                              "Employment 4." =  "#9E9AC8", 
+                              "Healthcare" =  "#807DBA", 
+                              "Leisure" =  "#DADAEB", 
+                              "Shops" =  "#BCBDDC", 
+                              "Unspecified 4." =  "#6A51A3")) + 
+                      #c(rep(RColorBrewer::brewer.pal(9,"Pastel2"),4))) +
   geom_text(size = 2.5, position = position_stack(vjust = 1), col = "Black") +
   theme_classic() +
-  theme(strip.text.y = element_text(size = 5.5),
-        strip.background.y = element_rect(fill = "lightgrey"),
+  theme(strip.text.y = element_text(size = 5),
+        strip.background.y = element_rect(fill = "lightgrey", linewidth=0.3),
         strip.placement = "inside",
@@
         axis.text.y=element_blank(),
-        axis.text.x=element_text(size=5),
+        axis.text.x=element_text(size=7),
+        axis.line = element_line(linewidth=0.3),
         legend.position = "none",
-        panel.grid.major.x = element_line(colour="grey", linewidth=0.1))
+        panel.grid.major.x = element_line(colour="grey", linewidth=0.3))
+```
+
+```{r, eval=FALSE}
+df_all_sub_ob_just_reordered_plot
+ggsave("figures/df_all_sub_ob_just_reordered_plot.tiff", dpi=300) 
+#manually remove the 3.a, 3.b. and 4. from the figure.
 ```
@@
         axis.title = element_blank()) 
-
-#standards_conc_percentages_plot
-# ggsave("figures/standards_conc_percentages_plot.png")
 ```
@@
   nrow = 2, ncol=2, rel_heights = c(0.8,0.7), rel_widths = c(0.8,4) )
-# standards_conc_percentages_plot_and_table
-# ggsave("figures/standards_conc_percentages_plot_and_table_new.png")
+```
+
+```{r, eval=FALSE}
+standards_conc_percentages_plot_and_table
+ggsave("figures/standards_conc_percentages_plot_and_table.tiff", dpi=300)
 ```
@@
 
-Transportation systems, as a class of essential technologies that facilitate or impede social inclusion and activity participation <!---[@churchTransport2000; @lucasTransport2001; @socialExclusion2003; @cassSocial2005; @casasSocial2007; @prestonAccessibility2007; @paezMobility2009]-->, have increasingly come into focus from the perspective of equity; a lively and rapidly growing literature has emerged on the topic [@martensTransport2016; @karnerAdvancesPitfallsMeasuring2023; @vecchioTransportEquityLatin2020].
+Transportation systems, as a class of essential technologies that facilitate or impede social inclusion and activity participation<!---[@churchTransport2000; @lucasTransport2001; @socialExclusion2003; @cassSocial2005; @casasSocial2007; @prestonAccessibility2007; @paezMobility2009]-->, have increasingly come into focus from the perspective of equity; a lively and rapidly growing literature has emerged on the topic [@martensTransport2016; @karnerAdvancesPitfallsMeasuring2024; @vecchioTransportEquityLatin2020].
 <!--@diciommoTransport2017; @guoSystematicOverviewTransportation2020; @karnerFrom2020; @pereiraTransportationEquity2021; @vanweeEvaluating2021; @zhangLiterature2021; @desjardinsExaminingEquityAccessibility2022; --> Aside from an ethical motivation, keeping track of objective and perceived inequalities could be of interest for governing bodies to respond to popular demands for fairness.
@@
 The benefits and burdens of transportation systems are diffuse over space and time posing a challenge in evaluating and tracking inequities.
-Over space transportation systems engineered to offer higher mobility for people _somewhere_, can simultaneously cut others off from essential opportunities _elsewhere_, as @rajeEngineering2004 poignantly illustrated with examples of infrastructure in the UK.
+Transportation systems that are engineered to offer higher mobility for people _somewhere_ can simultaneously cut others off from essential opportunities _elsewhere_, as @rajeEngineering2004 poignantly illustrated with examples of infrastructure in the UK.
 Over time, the shades of policies past can continue to haunt a region and the planet as shown by the legacy of displacement and decay caused by urban highways all across the US [@archerWhite2020] and the time horizon for the impacts of climate change [@markolfTransportation2019].
@@
 Further, there are multiple works that discuss the conceptual and philosophical foundations of equity and fairness in transportation [@martensTransport2016; @pereiraDistributiveJusticeEquity2017a; @vanoutriveHow2019].
-There have also been, previous reviews of planning documents have investigated equity from narrowly scoped perspectives, such as accessibility [@boisjolyHowGetThere2017] or a particular mode of transportation e.g., cycling [@doranPursuitCyclingEquity2021].
+Previous reviews of planning documents have investigated equity from narrowly scoped perspectives, such as accessibility [@boisjolyHowGetThere2017] or a particular mode of transportation e.g., cycling [@doranPursuitCyclingEquity2021].
 These efforts are valuable to scholars, the public, and decision-makers alike.
@@
 
-To illustrate this gap, we note how @beilerExploring2016 in their exploration of transport equity, cite the following strategies identified by the US DOT to address matters of justice [p. 287]:
+To illustrate this gap, we note how @beilerExploring2016 in their exploration of transport equity, cite the following strategies identified by the U.S.
+Department of Transportation to address matters of justice [p.
+287]:
 
@@
 
-While commendable, the strategies are too vague, which means it is possible to implement them in a myriad ways, either genuinely to comply with the spirit of justice, or else performatively to deceive it [@mcculloughPerformative2023].
+While commendable, the strategies are too vague, which means it is possible to implement them in a myriad of ways, either genuinely to comply with the spirit of justice, or else performatively to deceive it [@mcculloughPerformative2023].
 Some relevant questions include: how much should the adverse effects be reduced?
@@
 What benefits should be distributed?
-Should the benefits be based on simple population weights or, contrariwise, advantage more deprived individuals?
-These questions boil down to the development and use of _standards_ for transportation justice.
+How much of the benefits should minority and low-income residents receive for policymakers to declare that a project is sufficiently ‘equitable'?
+These questions touch on the development and use of _standards_ for transportation justice.
 In this work, we choose to focus exclusively on the academic domain as we know that adoption of equity concepts in planning practice has lagged developments in academic work [@pereiraTransportationEquity2021; @boisjolyHowGetThere2017; @linovskiEquityTransportationPlanning2020; @doranPursuitCyclingEquity2021; @litmanEvaluatingTransportationEquity2022]
@@
 Second, it applies the framework to collate the existing academic knowledge that undertakes transportation equity analysis supported by equity standards and conceptualisations.
-We hope the findings and the presentation of academic knowledge will be useful in supporting the development and implementation of equity standards for transportation planners, policy and decision-makers.
-
+We hope the findings and the presentation of academic knowledge will be useful in supporting the development and implementation of equity standards for transportation planners, policy, and decision makers.
 In contrast to previous reviews on how to measure inequality in transportation systems, this work is concerned with the implied or explicit standards that are used to judge whether the measured inequalities are fundamentally “fair” or unacceptable.
@@
 -	**What** are those benefits and burdens of transportation systems? Are they qualified in sidewalk widths, bus frequency, travel times, accessibility, or some other consideration?  
--	**Who** benefits and is burdened by transportation systems? Along which lines of identity are transportation inequities examined?
--	**Where and when** are equity standards applied? What are the differences between the research depending on location, context-specific planning processes and time of publication? 
+-	**Who** benefits and who is burdened by transportation systems? Along which lines of identity are transportation inequities examined?
+-	**Where and when** are equity standards applied in the literature? What are the differences between the research depending on location, context-specific planning processes and time of publication? 
 -	**How** are equity standards and inequities measured?
@@
 
-<!-- The structure of this review is as follows.
-After this introduction, the stage for our investigation is set by laying out some important definitions and describe the methods used for searching, selecting, and reviewing the relevant literature.
-This is followed by a summary of the findings and their critical appraisal.
-We conclude with some calls for action to improve the practice of setting and using standards of fairness for transportation justice.
--->
-
 # Background {#sect2}
 
-In the introductory section the terms "justice", "fairness", and "equity" are used relatively loosely.
+In the introductory section the terms "justice", "equity", and "fairness" are used relatively loosely.
 This was done purposefully.
-As noted, people often have strong intuitions of what is "fair", "just", and "equitable" which may or may not match those of the authorities who set the standards of fairness.
+As noted, people often have strong intuitions of what is "just", "equitable", and "fair" which may or may not match those of the authorities who set the "standards" of fairness.
 But in democratic societies, the authority of all those charged with the business of governing should derive from the will of the people.
-It is therefore important to explicitly state how we plan to use these words - what we mean by "justice", "fairness", and "equity".
+It is therefore important to explicitly state what we mean by these terms.
+In the following, we describe justice as embracing moral motivation, equity as measurement tools, fairness as a yardstick of justice, and standards as statements of fairness.
 
@@
 - **Retributive justice** is concerned with the proportional retribution of wrongdoers with relation to legitimate punishers and the innocent [@sep-justice-retributive].
-- **Reparative (or restorative) justice** focuses on the reparation of caused harm; it centers 'reintegrative shaming' to restore victims, wrongdoers, and the community according to moral values [@tylerRestorativeJusticeProcedural2006].
+- **Reparative (or restorative) justice** focuses on the reparation of caused harm; it centers the needs and voices of victims to restore wrongdoers and the community according to moral values [@tylerRestorativeJusticeProcedural2006; @braithwaite2003principles].
+In planning and policy contexts, reparative justice is often conceived of as accountability mechanisms in which institutions that caused harm financially compensate victims [@safransky2022grammars].
 - **Procedural justice** strives to ensure that the views and preferences of all stakeholders are fairly accounted for in the decision-making and inter-personal procedures affecting their lives and communities [@tylerRestorativeJusticeProcedural2006].
-- **Distributive justice** is perhaps the most commonly studied form of justice [see @jaggarPhilosophicalChallengesGlobal2009, p.
-2; @pereiraDistributiveJusticeEquity2017a], and its main concern is the way the benefits and burdens of the tangible and intangible products of society are collected by different segments of a population.
+- **Distributive justice** is perhaps the most studied form of justice [see @jaggarPhilosophicalChallengesGlobal2009, p.
+2] as well in the transportation domain [@pereiraDistributiveJusticeEquity2017a].
+Its main concern is the way the benefits and burdens of the tangible and intangible products of society are collected by different segments of a population.
 
@@
 
-**Equity**, as conceptualized alongside distributive justice, tends to encompass tools to understand the distribution of benefits and burdens of things among a population; often emphasizing those with the least advantage and equality in their outcomes.
+**Equity**, as conceptualized alongside distributive justice, tends to encompass tools to understand the distribution of benefits and burdens of things among a population, often emphasizing those with the least advantage and their outcomes.
 In the transportation domain, the term is somewhat loaded because it is perceived as stemming from the authority of the state and is meant to assist with decisions about regulating and financing transportation spending [@karnerTransportationEquityTransportation2020].
 Here, we are in agreement with @karnerTransportationEquityTransportation2020 that equity analysis should not be seen as an end in and of itself, but rather as a means to gather information about actual and perceived inequities.
-In this respect, the analytical tradition of equity, at least in transportation planning, means that the relevant models become embedded in the "political ecology of the estimated truth" [@kingModels1993]: in principle their assumptions and scope must be open and transparent, or else they may be more vulnerable to misuse and even abuse as tools of subjugation.
+In this respect, the analytical tradition of equity, at least in transportation planning, means that the relevant models become embedded in the "political ecology of the estimated truth" [@kingModels1993]: in principle, their assumptions and scope must be open and transparent, or else they may be more vulnerable to misuse and even abuse as tools of subjugation.
 
@@
 The standard provides conditions to determine whether a situation is _fair_.
-A similarly weak standard is a _Pareto improvement_, whereby it is possible to concentrate the benefits as long as no group is worse off compared to the status quo.
-A policy is fair as long it does no harm.
-A somewhat more strict standard, a Pareto-Plus improvement, stipulates that an intervention is fair when all groups receive at least _some_ (non-trivial) benefits; the size of the benefits for each group is irrelevant.
+A similarly weak standard is a Pareto improvement, whereby it is possible to concentrate the benefits as long as no group is worse off compared to the status quo e.g., [@tan2016pareto; @xuParetoimprovingPoliciesIdealized2018].
+From this standard, a policy could be considered fair as long it does no harm.
+<!--A somewhat more strict standard, a Pareto-Plus improvement, stipulates that an intervention is fair when all groups receive at least _some_ (non-trivial) benefits; the size of the benefits for each group is irrelevant.
 In contrast to the notion of "do no harm", such a standard embodies the ideal that no one is denied benefits.
-An egalitarian standard would weigh the benefits or burdens by population, and fairness is achieved when each group gives or receives in proportion to their size.
-In contrast, an affirmative action standard is even stricter, since it requires the benefits to be distributed in a non-egalitarian way that favors those who are still harmed by past or present discriminatory practices.
+--> A standard based on egalitarian theories such as proportional equity would weigh the benefits or burdens by population, and fairness is achieved when each group gives or receives in proportion to their size [@billsLookingMeanEquity2017;@martensJusticetheoreticApproachDistribution2012] In contrast, an affirmative action standard based in restorative justice is even stricter.
+It requires the benefits to be distributed in a non-egalitarian way that favors those who are still harmed by past or present discriminatory practices [@bierbaumMobilityJusticeLinking2021].
 
@@
 As such, the answer to "Where?" is the definition of the spatial boundaries.
+For example, we can distinguish literature from the global South and North, allowing us to examine how they approach, deal with, and position themselves in different spatial contexts in addressing transport inequalities.
 
 Conventionally, the question of **"When?"** refers to the temporal circumstances within which the demands of justice have application.
-In the cause of transportation justice, we ask about the temporal aspects of transportation systems, as examples: _when_ did the equity analysis take place and under what historical policy context; _the right time_ to invest in transportation infrastructure and as a result when to generate a spatial inequality [e.g., @rabelloEvaluation2015]; for _how long_ the burdens and benefits can still be associated to a specific transportation intervention; or even _timelines_ of reparative justice interventions that reconcile the shadows of past transportation-related injustices.
+In the cause of transportation justice, one could ask about the temporal aspects of transportation systems, as examples: _when_ did the equity analysis take place and under what historical policy context; _the right time_ to invest in transportation infrastructure and as a result when to generate a spatial inequality [@rabelloEvaluation2015]; for _how long_ the burdens and benefits can still be associated to a specific transportation intervention; or even _timelines_ of reparative justice interventions that reconcile the shadows of past transportation-related injustices.
+In this literature review, given the amount of papers reviewed and the inconsistencies in how temporal aspects were addressed, we exclusively focus on the paper's publication year.
 
@@
 A possible complication in the case of transportation is that disentangling the "Who" from their mobility tools is not always straightforward.
-Although a person is not their mode of transportation, there are large segments of the population who live in situations where they cannot extricate themselves from the mobility tools they can use, either because they have driven themselves out of choices [@laveryDriving2013], or have been driven out of choices by factors beyond their control [e.g., captive users of a single mode @jacquesRescuing2012; @cherancheryInvestigatingPerceptionCaptive2018].
-<!-- In societies that have grown into transportation monocultures with a predilection for automobility [@millerCollaborative2011] there may actually be less choice about mobility tools than we would like to assume.-->While it is important to avoid conflating the "Who" with the "What", for analytical purposes we need to be mindful of the connection between aperson and their mobility tools.
+Although a person is not their mode of transportation, there are large segments of the population who live in situations where they cannot extricate themselves from the mobility tools they can use, either because they have driven themselves out of choices [@laveryDriving2013], or have been driven out of choices by factors beyond their control such as captive users of a single mode [@jacquesRescuing2012; @cherancheryInvestigatingPerceptionCaptive2018].
+<!-- In societies that have grown into transportation monocultures with a predilection for automobility [@millerCollaborative2011] there may actually be less choice about mobility tools than we would like to assume.-->While it is important to avoid conflating the "Who" with the "What", for analytical purposes we need to be mindful of the connection between a person and their mobility tools.
 <!-- In the case of transportation, in addition to members of the public who use transportation systems, there is another category of **Who**, that stands possibly in opposition to users, namely the entities charged with providing services, maintaining infrastructure, and so on.
@@
 To understand the distributional implications of transportation systems, it is essential to understand what do they _produce_?
-Transportation systems can be seen as space-time convergence technologies, mobility tools that improve the rate at which time is traded for space.
+Transportation systems are space-time convergence technologies, mobility tools that improve the rate at which time is traded for space.
 They can increase the speed of movement: for instance, sidewalks and traffic lights enable potential movement of pedestrian and vehicles (i.e.
@@
 For this reason, we cannot stop at only considering mobility but must consider its ulterior goal-- reaching destinations.
-Coupled with land use, mobility creates accessibility to needed or desired activities.
+Coupled with land use, mobility creates accessibility to needed or desired activities as well as varied burdens.
 Transportation justice, thus, involves proximate (mobility tools and mobility) and ulterior (accessibility and activity opportunities) objectives.
-These considers all fold under objects of transportation justice, the "What?".
-<!-- At their most fundamental, transportation systems are space-time convergence technologies, tools that improve the rate at which time is traded for space.
-They usually do this by increasing the speed of movement: sidewalks facilitate walking, traffic lights facilitate the ordered flow of vehicular traffic, and a launching pad makes it possible for a rocket to take off.
-With complex interlocking parts (sidewalk, road, traffic sign, parking regulations), transportation systems produce _mobility_, the potential for movement.
-The realization of these potentials happen through travel.
-However, as the adage goes, travel is derived demand, which seems to hold for most situations [e.g., @mokhtarianUnderstanding2001; @whalenMode2013<!-- @redmondPositive2001; @paezEnjoyment2010; --><!--].
-For this reason, we cannot stop at considering mobility but must consider its ulterior goal-- reaching destinations.
-In combination with land use systems (the spatial distribution of destinations/opportunities), mobility produces _accessibility_-- the potential to reach places where activities that are needed or desired.
-We can think of the objects of transportation justice as being _proximate_ (the tools of mobility, mobility itself), and _ulterior_ (accessibility, opportunities for activity participation).
-The burdens of transportation are also many and varied.
-Some are direct and paid directly by the traveler (e.g., travel time, out-of-pocket costs), but many others are indirect and related to network externalities (e.g., exposure to pollution, road accidents).
--->
+<!-- As well, the burdens of transportation are also varied.
+Some are paid directly by the traveler (e.g., travel time, out-of-pocket costs), but many others are indirect and related to network externalities (e.g., exposure to pollution, road accidents).-->
 
@@
 Justice is an inherently social construct.
-Asking **Why?** amounts to asking what sort of social contracts regulate human interactions, or in other words, what are the rules that our collective will to believe imposes on each of us.
+Asking **Why?** amounts to asking what sort of social contracts regulate human interactions.
+In other words, what are the rules that our collective will to believe imposes on each of us.
 These contracts can be defined by constitution, but there are often unwritten and possibly contested variants.
@@
 The primary research question and the protocol were initially defined by the authors, a group of experts in the field of transportation.
-The initial draft of the protocol was refined from preliminary searches of related-reviews [e.g., @zhangLiterature2021; @pereiraDistributiveJusticeEquity2017a; @vecchioAccessibilityCapabilitiesApproach2021], and in consultation with a University of Toronto Research Services Librarian and a Liaison Librarian in City Studies.
+The initial draft of the protocol was refined from preliminary searches of related reviews e.g., [@zhangLiterature2021; @pereiraDistributiveJusticeEquity2017a; @vecchioAccessibilityCapabilitiesApproach2021], and in consultation with a University of Toronto Research Services Librarian and a Liaison Librarian in City Studies.
 
@@
 
-# A critical appraisal of the lay of the land {#sect3}
-
-This section critically threads together the trends identified in the corpus through the data extraction process (@fig-figA2)<!-- update this cross reference!-->.
+# An appraisal of the lay of the land {#sect3}
+
+This section threads together the trends identified in the corpus through the data extraction process (@fig-figA2)<!-- update this cross reference!-->.
 Specifically, conceptualisations and standards (the "How") and their applications to the objects of justice (the "What"), the subjects of justice (the "Who"), and under which situations ("When" and "Where").
 As an overview, @fig-fig1 visualises the prominence of each category under "When?", "Where?", "Who?", and "What?".
-"How?" with its conceptualisations and standards is visualised in the following @fig-fig2.<!-- For the reasons reviewed, setting standards (that is, making statements of fairness) is a highly context-specific endeavor.
+The "How?" category, with its conceptualisations and standards, is visualised in the following @fig-fig2.<!-- For the reasons reviewed, setting standards (that is, making statements of fairness) is a highly context-specific endeavor.
 While our search strategy and review tries to be as faithful to the literature as possible, it does not allow us to lay claim to reporting about on-the-ground transportation justice efforts from a truly global perspective.-->
@@
 
-df_all_sub_ob_just_reordered_plot
-ggsave("figures/df_all_sub_ob_just_reordered_plot_new.png")
+knitr::include_graphics("figures/df_all_sub_ob_just_reordered_plot.tiff")
 ```
@@
 
-knitr::include_graphics("figures/standards_conc_percentages_plot_and_table_new.png")
+knitr::include_graphics("figures/standards_conc_percentages_plot_and_table.tiff")
 ```
@@
 Within our corpus most papers (`r ((case_cont_count$count[4] + case_cont_count$count[5]+ case_cont_count$count[6])/165) |> percent()`) focus on case studies in the Global North, with many studies from North America (particularly USA and Canada), Europe (particularly France, Spain and Northern Europe), Oceania (Australia and New Zealand), and Asia (Japan and Israel).
-Though their subject matter is varied, their spatial context mainly pertains to North America and Europe, and thus more often than not, in conversation with more developed and formal government transport planning apparatuses and technologies [e.g., planning for equitable high-speed rail, as in @monzonEfficiencySpatialEquity2013; autonomous vehicle technology, as in @eppenbergerOpportunitySharedAutonomous2021; or on the public consultation processes, as in @reddySafeguardingMinorityCivil2010]).
+Though their subject matter is varied, their spatial context mainly pertains to North America and Europe, and thus more often than not, in conversation with more developed and formal government transport planning apparatuses and technologies e.g., planning for equitable high-speed rail [@monzonEfficiencySpatialEquity2013], autonomous vehicle technology [@eppenbergerOpportunitySharedAutonomous2021], or on the public consultation processes [@reddySafeguardingMinorityCivil2010].
 
@@
 Studies pertaining to Africa are even less numerous compared to the South American literature.
-A shared characteristic among the studies from these two continents is a scarcity of official transport data [@friedMeasuringSustainableDevelopment2020] and a reliance on policy guidelines developed by international organizations.
-These studies also incorporate the use of informal transportation options and the pressures to develop road network infrastructure (which tends to support car dependency) over meeting the mobility/accessibility needs of citizens [@thondooSmallCitiesBig2020].
-To address these challenges, researchers compile databases based on open and geo-referenced data, calculate objective and/or subjective measures [@berheAdaptationDissonanceQuality2014], and focus on advancing transport justice for low to medium income countries (LMIC) by aligning their goals with external policy guidelines such as the Sustainability Development Goals (SDG), particularly those related to universal accessibility [@friedMeasuringSustainableDevelopment2020].
+A shared characteristic among the studies from these two continents is a scarcity of official transport data [@friedMeasuringSustainableDevelopment2020].
+These studies also incorporate the use of informal transportation options and tensions in developing road network infrastructure (which tends to support car dependency) over meeting the mobility/accessibility needs of citizens more equitably and sustainably [@thondooSmallCitiesBig2020].
+To address these challenges, researchers compile databases based on open and geo-referenced data, calculate objective and/or subjective measures [@berheAdaptationDissonanceQuality2014], and focus on advancing transport justice for low to medium income countries by aligning their goals with external policy guidelines such as the Sustainability Development Goals (SDG), particularly those related to universal accessibility [@friedMeasuringSustainableDevelopment2020].
 
@@
 
-- affordability-as-a-barrier at the user- or policy maker- level is more often the motivation in work from the Global South;
+- affordability as a barrier at the user- or policy maker- level is more often the motivation in work from the Global South;
 - the expression of greater tensions in investing in new transportation infrastructure, such as roads in rural/under-developed areas, compared to prioritization of non-car modes.
@@
 
-We characterise the Global North and South literature as laying at different points temporally on the transport inequity continuum.
-As Global South' formal planning processes operate under stricter financial constraints, they rely on more informal processes to address unmet needs - a significant source of equity concerns [e.g, informal transit in Nairobi @friedMeasuringSustainableDevelopment2020; populations living in informal residents in Mumbai @sharmaPublicTransitAccessibility2021].
-Operationalization of international standards are also more relevant [e.g., WHO noise and air pollution in Delhi @apparicioCyclingOneMost2021].
-In contrast, these standards are no longer as relevant in the Global North [e.g., $NO_2$ pollution near roadways in Montreal are under the WHO guideline @carrierApplicationThreeMethods2014], though perhaps being more relevant at a time of lower transport-related development.
+We characterise the Global North and South literature as addressing and being situated at different points in the process of addressing transport inequities.
+As Global South' formal planning processes operate under stricter financial constraints, they rely on more informal processes to address unmet needs - a significant source of equity concerns e.g, informal transit in Nairobi [@friedMeasuringSustainableDevelopment2020], populations living in informal residents in Mumbai [@sharmaPublicTransitAccessibility2021].
+Operationalization of international standards are also more relevant e.g., using the WHO noise and air pollution standard in Delhi case study [@apparicioCyclingOneMost2021].
+In contrast, these minimum standards may not be relevant in some areas of the Global North e.g., $NO_2$ pollution near roadways in Montreal are consistently below the WHO standard though populations are disproportionately impacted [@carrierApplicationThreeMethods2014].
 In these ways, the Global South temporally lags Global North development.
-However, this lag presents opportunities for all: for the Global South, to adopt relevant enhancements to Global North formal equity planning processes [e.g., indicator creation for disadvantages groups @cuiSpatialAccessPublic2020] and not repeat their mistakes [e.g., from the entrenched car-centric development @warrenDevelopingEquitableSustainable2015, to the disproportionate contribution to carbon-intense mobility @chancelCarbonInequalityKyoto2015].
-For the Global North, reckon with their contributions to uneven development globally and environmental impacts as well as adopt relevant aspects from informal planning processes.
+However, this lag presents opportunities for all: For the Global South, to adopt potentially successful enhancements from Global North formal equity planning processes (e.g., indicator creation for disadvantages groups [@cuiSpatialAccessPublic2020]) and not repeat past mistakes (e.g., from entrenched car-centric development [@warrenDevelopingEquitableSustainable2015] to the disproportionate contribution to carbon-intense mobility [@chancelCarbonInequalityKyoto2015].
+For the Global North, there is an opportunity to reckon with their own contributions to uneven development globally and environmental impacts as well as adopt relevant aspects from informal planning processes.
 <!--`r word_comment("I condensed this paragraph.
@@
 Turning to the "Who?" of population group characteristics in @fig-fig1, the majority focus on "Income", especially the lowest-income groups who face lower mobility and accessibility and higher costs and exposure [@peungnumsaiGridBasedSpatialAnalysis2020; @zhaoUnequableSpatialAccessibility2020; @falavignaAssessingInequalitiesPublic2016].
-Evidence suggests that low household income is a significant determinant of transport-related inequities [e.g., urban access to public transport in Thailand [@peungnumsaiGridBasedSpatialAnalysis2020], access to urban employment opportunities in Brazil [@boisjolyAccessibilityMeasurementsSao2020], and unfavorable rates of environmental noise, air pollution, and green space in Netherlands [@kruizeEnvironmentalEquityRole2007].
+Evidence suggests that low household income is a significant determinant of transport-related inequities e.g., urban access to public transport in Thailand [@peungnumsaiGridBasedSpatialAnalysis2020], access to urban employment opportunities in Brazil [@boisjolyAccessibilityMeasurementsSao2020], and unfavorable rates of environmental noise, air pollution, and green space in Netherlands [@kruizeEnvironmentalEquityRole2007].
 Yet, low-income is not universally associated with lower transport-related benefits for every object of justice - context matters.
-For instance, in Sheffield, UK @mearsUnderstandingSocioeconomicEquity2019 demonstrate that historically working-class (lower-income) neighbourhoods have _more_ access to green space than other neighbourhoods.
+For instance, in Sheffield, UK, @mearsUnderstandingSocioeconomicEquity2019 demonstrate that historically working-class (lower-income) neighbourhoods have _more_ access to green space than other neighbourhoods.
 This is potentially due to historic urban planning approaches, though the quality is less than average.
-Similarly, @bertrandMeasuringMappingDisparities2008 find that lower income groups do not always have below average accessibility depending on the granularity of analysis.
-
-"Age" is the second most common category within population group: these papers highlight differences in age-related capabilities.
+Similarly, @bertrandMeasuringMappingDisparities2008 find that lower income groups do not always have below average accessibility depending on the granularity of analysis; similar findings are uncovered in other spatial and temporal contexts [@fothEquitableTransitExamining2013a; @allenSizingTransportPoverty2019a]
+
+"Age" is the second most common population group category: these papers highlight differences in age-related capabilities.
 For instance, @martinez-jimenezAccessibilityCultureEducation2019 and @arranz-lopezSocialSpatialEquity2019 investigate travel distances/times to various opportunities, demonstrating how age is associated with differences in opportunity access.
 The most extreme age groups are commonly of focus: namely, school-aged children and older populations.
-For children, analysis of wellbeing [@laszkiewiczChildrenGreenWalk2020], safety and access to schools [@sharmaSpatialSocialInequities2022, @corazzaMethodologyEvidenceCase2020], and promoting active travel [@mackieOvercomingBarriersCycling2009; @mehdizadehWalkingTimeSchool2017] are common-place.
-Papers that focus on older adults similarly investigate transport-related wellbeing [e.g, @chenSpatialAnalysisFramework2020], access to age-specific destinations [e.g.
-@chengInvestigatingWalkingAccessibility2019], and improved options to meet unmet travel needs [e.g., @nordbakkeTransportUnmetActivity2015].
+For children, analysis of wellbeing [@laszkiewiczChildrenGreenWalk2020], safety and access to schools [@sharmaSpatialSocialInequities2022; @corazzaMethodologyEvidenceCase2020], and promoting active travel [@mackieOvercomingBarriersCycling2009; @mehdizadehWalkingTimeSchool2017] are common-place.
+Papers that focus on older adults similarly investigate transport-related wellbeing [@chenSpatialAnalysisFramework2020], access to age-specific destinations [@chengInvestigatingWalkingAccessibility2019], and improved options to meet unmet travel needs [@nordbakkeTransportUnmetActivity2015].
 
@@
 These categories are often of exclusive focus.
-Studies with a "(Dis)abilities" focus assess travel capabilities, namely through physical accessibility and universal design guidelines [e.g., @parkJourneyVisuallyImpaired2017; @chiscanoImprovingDesignUrban2021; @orellanaWalkRollMapping2020].
+Studies with a "(Dis)abilities" focus assess travel capabilities, namely through physical accessibility and universal design guidelines [@parkJourneyVisuallyImpaired2017; @chiscanoImprovingDesignUrban2021; @orellanaWalkRollMapping2020].
 "All Populations" papers make no distinction in population.
 This is done, for instance in @kitaDoesLevelTransportation2020, who investigates disparities in accessibility to food-stores and their self-reported capability/frequency of going outdoors.
-Often in both "(Dis)abilities" and "All Populations" focused papers, an im/explicit motivation is access (to necessities and desired destinations) for all.
-
-As mentioned, the corpus is largely intersectional.
-For instance, a large proportion of papers apply a "Composite Vulnerability Index"; an _informed_ combination of several individual traits like low income, unemployment and immigrant status.
+Often in both "(Dis)abilities" and "All Populations" focused papers, an implicit or explicit motivation is access (to necessities and desired destinations) for all.
+
+A large proportion of papers apply a "Composite Vulnerability Index"; an _informed_ combination of several individual traits like low income, unemployment and immigrant status.
 These indices are generated from official government sources or author-informed census data creation methods.
 As an example, the Neighbourhood Equity Index (NEI) is a neighbourhood-ranked measure of vulnerability created by the City of Toronto and used in @awuorInfluenceEnvironmentalHealth2019 to examine disparities in premature death.
-Other works use national census indicators such as the social and housing deprivation index [e.g., @pucciInequalitiesJobrelatedAccessibility2019] or explore transport disadvantage, equity in policy implementation, or transport-related mortality burden by means of census measures (e.g., household poverty) and transport-related accessibility indicators [e.g.,  @aldredEquityNewActive2021; @iungmanImpactUrbanTransport2021; @sunPublicTransportAvailability2021; @scheurerSpatialAccessibilityPublic2017].
-Similarly, Environmental Justice (EJ) indicators have been used in the US literature to identify neighbourhoods that have a higher than average proportion of low-income and non-white populations and evaluate the equity impacts of transportation projects [e.g., @rowangouldIdentifyingEnvironmentalJustice2016; @parkTransitParksEnvironmental2021; @reddySafeguardingMinorityCivil2010].
-
-Intersectional considerations are so prevalent in the corpus that papers studying "Gender", "Race/ethnicity", "Education", or "Employment" level are infrequently studied exclusively.
-Only a few papers focus exclusively on gendered differences in active transportation [e.g., @adlakhaMindGapGender2020, @xieWonCycleRoute2018], race/ethnicity's relationship to green space proximity [@silvaEnvironmentalJusticeAccessibility2018], and culturally-appropriate opportunities [e.g., @wangModelingSpatialAccessibility2011].
-Papers that focus *solely* on "Employment" or "Education" level are absent all together.
-Furthermore, "Other" population groups are also frequently considered in the corpus: this is a catch-all category that includes group populations less commonly the subject of research e.g.,  veterans and access to specific-healthcare needs [@mooneyTravelDistanceBarrier2000], pregnant people and access to services [@vadrevuMeasuringSpatialEquity2016], and youths who live in foster care [@batscheUsingGISEnhance2012].
-Overall, the diversity of population groups considered in the corpus, often intersectionally, demonstrates the variety of transportation-equity concerns addressed in the corpus.
+Other works use national census indicators such as the social and housing deprivation index [@pucciInequalitiesJobrelatedAccessibility2019] or explore transport disadvantage, equity in policy implementation, or transport-related mortality burden by means of census measures (e.g., household poverty) and transport-related accessibility indicators [@aldredEquityNewActive2021; @iungmanImpactUrbanTransport2021; @sunPublicTransportAvailability2021; @scheurerSpatialAccessibilityPublic2017].
+Similarly, Environmental Justice (EJ) indicators have been used in the US literature to identify neighbourhoods that have a higher than average proportion of low-income and non-white populations and evaluate the equity impacts of transportation projects [@rowangouldIdentifyingEnvironmentalJustice2016; @parkTransitParksEnvironmental2021; @reddySafeguardingMinorityCivil2010].
+
+Multi-dimensional considerations are so prevalent in the corpus that papers studying "Gender", "Race/ethnicity", "Education", or "Employment" level characteristics are infrequently studied exclusively.
+Only a few papers focus exclusively on gendered differences in active transportation [@adlakhaMindGapGender2020, @xieWonCycleRoute2018], race/ethnicity's relationship to green space proximity [@silvaEnvironmentalJusticeAccessibility2018], and culturally-appropriate opportunities [@wangModelingSpatialAccessibility2011].
+Papers that focus *solely* on "Employment" status or "Education" level are absent all together.
+Furthermore, "Other" population groups are also frequently considered in the corpus: this is a catch-all category that includes group populations less commonly the subject of research e.g.,  veterans and access to specific-healthcare needs [@mooneyTravelDistanceBarrier2000], pregnant people and access to maternity services [@vadrevuMeasuringSpatialEquity2016], and youths who live in foster care [@batscheUsingGISEnhance2012].
+Overall, the diversity of population groups considered in the corpus demonstrates the variety of transportation-equity concerns addressed in the corpus.
 
@@
 Travel mode, though modifiable, is intertwined with individual identity, posing challenges for equity analysis.
-In this sub-section, we choose to view the mode of travel primarily as those who use the mode, hence the subject of justice (i.e., the “who”), but occasionally, this will be enjoined to an object of justice (i.e., a “what”).
+In this sub-section, we choose to view the mode of travel primarily as those who use the mode, hence the subject of justice (i.e., the “who”), but occasionally, this will be enjoined to an object of justice (i.e., a “what” such as the inequities in rural transit 'mobility' relative to cars [@giuffridaAddressingPublicTransport2021]).
 
@@
 Despite transit being perceived unfavourably by some [@paezEnjoyment2010; @mellaDrivers2021], it is often viewed as the only and primary mobility option mode for many [@jacquesRescuing2012; @welchMeasure2013; @cherancheryInvestigatingPerceptionCaptive2018].
-The public transit focused papers assess a variety of topics such as: food desert identification [@mckeyCrowdsourcedMappingHealthy2020], barrier-free transportation for people who face disabilities [e.g., @jimenez-espadaResearchProblemUniversal2021; @liuStatusQuoChallenges2019] and people with autism [@limFacilitatingIndependentCommuting2021; @feeleyValidationParatransitSkills2019].
+The public transit focused papers assess a variety of topics such as: food desert identification [@mckeyCrowdsourcedMappingHealthy2020], barrier-free transportation for people who face disabilities [@jimenez-espadaResearchProblemUniversal2021; @liuStatusQuoChallenges2019] and people with autism [@limFacilitatingIndependentCommuting2021; @feeleyValidationParatransitSkills2019].
 We infer from the corpus that public transit is seen as a public good and hence a natural object of justice.
@@
 Pedestrians represent a unique convergence of "what" and "who," utilizing their own bodies for mobility.
-Papers focusing solely on walking often use walkability scores to assess neighbourhood quality [@evansAccessibilityUserNeeds2015] pedestrian mobility in different demographics [@towneUsingWalkScore2016; @kimNeighborhoodEnvironmentWalkability2016], or urban peripheral regions [@blecicCapabilitywiseWalkabilityEvaluation2021], measuring equity distribution.
+Papers focusing solely on walking often use walkability scores to assess neighbourhood quality [@evansAccessibilityUserNeeds2015], pedestrian mobility in different demographics [@towneUsingWalkScore2016; @kimNeighborhoodEnvironmentWalkability2016], or urban peripheral regions [@blecicCapabilitywiseWalkabilityEvaluation2021], measuring equity distribution.
 Some use walkability to gauge public health and urban vitality [@sungResidentialBuiltEnvironment2015; @mccormackRelationshipClusteranalysisDerived2012].
@@
 When studied in the corpus, they're critically compared with transit or considered in areas with inadequate transit [@kimmelStructuralBarriersComprehensive2018; @aljoufie2016].
-Upon the rise of automobility, walking and cycling were in the way; <!-- though being the way people got around historically [@robertsShort1998] and cycling's efficiency at a lower impact-->Transit came to be seen as a social service and sometimes as a hindrance to the full realization of automobility.
-We surmise that the corpus' focus on non-car modes is a reckoning with automobility's legacy and ongoing requirements of disproportionate amounts of space, public subsidies and government support that underestimates its true cost [@gosslingSocial2019; @timperleyFight2021] including externalities like air pollution or safety [@fengTradeoffsMobilityEquity2014; @houstonProximityLicensedChild2006].
-Other than "Car" being compared to walk, cycle or transit, some papers reflect the existing ways of travel and alongside "Car" including "Other" (@fig-fig1) options namely wheelchair accessible taxi and/or paratransit [@marquezFactorsAffectingPersonal2019; @wilkinsonmeyersLiveOrdinaryLife2015], travel on waterways [@vadrevuMeasuringSpatialEquity2016; @caoIndicatorsSocioSpatialTransport2017; @parrySocialVulnerabilityClimatic2018], motorcycles or other micro-mobility [@berryInvestigatingFuelPoverty2016; @savingmothersDidSavingMothers2019; @tiwariDevelopmentPublicTransport2021], or by emergency vehicle [@pedigoInvestigationDisparitiesGeographic2010; @patelDeterminingGeographicAreas2007].
+Upon the rise of automobility, walking and cycling were in the way; <!-- though being the way people got around historically [@robertsShort1998] and cycling's efficiency at a lower impact--> transit came to be seen as a social service and sometimes as a hindrance to the full realization of automobility.
+We surmise that the corpus' focus on non-car modes is a reckoning with automobility's legacy and ongoing requirements of disproportionate amounts of space, public subsidies and government support that underestimates its true cost [@gosslingSocial2019; @timperleyFight2021] including externalities like congestion, air pollution or safety [@zhengAreabasedEquitablePricing2020; @fengTradeoffsMobilityEquity2014; @houstonProximityLicensedChild2006].
+Other than "Car" being compared to walk, cycle or transit, some papers reflect the existing ways of travel and alongside "Car" including "Other" (@fig-fig1) options namely wheelchair accessible taxi and/or paratransit [@marquezFactorsAffectingPersonal2019; @wilkinsonmeyersLiveOrdinaryLife2015], travel on waterways [@vadrevuMeasuringSpatialEquity2016; @caoIndicatorsSocioSpatialTransport2017; @parrySocialVulnerabilityClimatic2018], motorcycle or other micro-mobility [@berryInvestigatingFuelPoverty2016; @savingmothersDidSavingMothers2019; @tiwariDevelopmentPublicTransport2021], or by emergency vehicle [@pedigoInvestigationDisparitiesGeographic2010; @patelDeterminingGeographicAreas2007].
 
@@
 -->
-From the corpus reviewed, we infer that one of the most fundamental benefits of transportation systems are mobility (enabling or impeding movement) and accessibility (the ease of reaching destinations).
+From the corpus reviewed, we infer that the most fundamental benefits of transportation systems are mobility (enabling or impeding movement) and accessibility (the ease of reaching destinations).
 These benefits are sometimes valued by themselves but are often seen by the corpus as instrumental to achieve an ulterior goal (e.g., activity participation and associated wellbeing benefits).
@@
 Papers on healthcare often highlight disparities in services, like @wangModelingSpatialAccessibility2011, who assess access to Mandarin-speaking family physicians for Chinese speakers in Toronto.
-Similarly, papers focusing on employment [e.g., @allenSizingTransportPoverty2019a] may do so since it is the most common trip purpose and is often correlated with other trip activities like shops, recreation, and other services, often identifying transportation-poor neighbourhoods.
-These studies typically use travel surveys, census data, and point-of-interest databases, and benefit from well-developed and institutional data that represents ‘typical trips,’ especially in the Global North where this data is more readily available.
+Similarly, papers focusing on employment are often aimed at identifying transportation-poor neighbourhoods [@allenSizingTransportPoverty2019a; @churchillTransportPovertySubjective2019].
+Employment is frequently used as a proxy for overall accessibility since it is the most common trip purpose and employment is usually collocated with other destinations like shops, recreation, and other services.
+These studies typically use travel surveys, census data and point-of-interest databases, and benefit from well-developed and institutional data.
+This especially holds in the Global North where this data is more readily available.
 
@@
 Places of "Leisure" (`r (dest_count$count[6]/165) |> percent()`) prompt accessibility questions like spatial distribution [@xuSocialInequalitiesPark2017], accessibility [@mavoaAreaLevelDisparitiesPublic2015], and explores of inequity [@mearsUnderstandingSocioeconomicEquity2019].
-Fewer papers cover "Community" destinations (e.g., public service centres, places of community support or worship) (`r (dest_count$count[2]/165) |> percent()`) or "Childcare" (`r (dest_count$count[1]/165) |> percent()`), but they're integral to study holistic activity participation [@albertsRebuildingWomenLivelihoods2016; @smithAccessibilityCapabilityMinimum2012].
+Fewer papers cover "Community" destinations (e.g., public service centres, places of community support or worship) (`r (dest_count$count[2]/165) |> percent()`) or "Childcare" (`r (dest_count$count[1]/165) |> percent()`), but they are integral in holistically studying activity participation [@albertsRebuildingWomenLivelihoods2016; @smithAccessibilityCapabilityMinimum2012].
 The lack of information about community destinations, especially for children, is notable in the literature broadly [@desjardinsChildren2022].
 
-### Dimensions of measurement
-
-Defining the object of justice (the "What"), whether specific destinations or the trip itself, shapes measurements of inequality crucial for tracking progress towards transport justice.
-In the reviewed corpus, we characterize these dimensions as a focus on:
-
-- accessibility and mobility; 
-- transport-related environmental externalities; 
-- centering human-health as related to transport systems; 
-- human safety related to vehicular-traffic; and 
-- cross-dimensional classifications.
-
-As discussed next, these intersect with conceptualizations of fairness and applied standards, forming the philosophical foundations for assessing justice in transportation.
+<!-- ### Dimensions of measurement -->
+
+<!-- Defining the object of justice (the "What"), whether specific destinations or the trip itself, shapes measurements of inequality crucial for tracking progress towards transport justice.
+In the reviewed corpus, we characterize these dimensions as a focus on:  -->
+
+<!-- - accessibility and mobility;  -->
+<!-- - transport-related environmental externalities;  -->
+<!-- - centering human-health as related to transport systems;  -->
+<!-- - human safety related to vehicular-traffic; and  -->
+<!-- - cross-dimensional classifications. -->
+
+<!-- As discussed next, these intersect with conceptualizations of fairness and applied standards, forming the philosophical foundations for assessing justice in transportation. -->
 
@@
 
-A few slices of the literature and how they are characterized by their foundational conceptualisations of fairness are summarized.
+A few segments of the literature and how they are characterized by their foundational conceptualisations of fairness are summarized.
 For example, a slice of the corpus is supported by broader "Rights" conceptualisations: these papers often focus on equity for people with disabilities or non-car users and associated challenges accessing transport infrastructure [@jimenez-espadaResearchProblemUniversal2021; @daamenAssessingGapPublic2008; @bharathyRevisitingClearFloor2018].
@@
 
-In another set of the corpus, distributions are examined through concepts of "Horizontal" and "Spatial equity", often using quantitative methods to assess distributional disparities without explicit justice rationales.
+In another subset of the corpus, distributions are examined through concepts of "Horizontal" and "Spatial equity", often using quantitative methods to assess distributional disparities without explicit justice rationales.
 Examples include setting travel impedance thresholds [@shenMeasuringSpatialAccessibility2020] and mapping accessibility indices spatially across populations [@monzonEfficiencySpatialEquity2013] or population-groups [@sharmaPublicTransitAccessibility2021].
@@
 Some use physical activity guidelines and surveys to understand activity related to active transportation infrastructure [@adlakhaMindGapGender2020; @auchinclossDesignBaselineDescription2020; @mccormackRelationshipClusteranalysisDerived2012].
-Mixed or qualitative methods combined with health-related outcome standards (e.g., travel times for emergency treatment [@savingmothersDidSavingMothers2019], or premature mortality [@awuorInfluenceEnvironmentalHealth2019]) lead to firmer justice identifications.
+Mixed or qualitative methods combined with health-related outcome standards (e.g., travel times for emergency treatment [@savingmothersDidSavingMothers2019], premature mortality [@awuorInfluenceEnvironmentalHealth2019]) leads to firmer justice identifications.
 
@@
 ### Standards and methods of measuring fairness
-With the frequent way conceptualisations are applied within the corpus detailed, the following categories of standards that overlap with at least one conceptualisation are included (see @fig-fig2 for definitions):
+With the ways that conceptualisations are frequently applied within the corpus detailed, the following are categories of standards that overlap with at least one conceptualisation (see @fig-fig2 for definitions):
 
@@
 For instance @peungnumsaiGridBasedSpatialAnalysis2020 suggest service benchmarks of equal supply and demand of public transit, revealing "Horizontal equities" as well.
-Others conceive the externalities of transportation system as trade-offs and aim to maximize transport-related benefits (i.e., time savings, emissions reductions, congestion reductions, user fares) through optimization/location-allocation methodologies e.g., [@fakhrmoosaviIncorporatingTravelTime2021; @zhengAreabasedEquitablePricing2020; @wismadiSpatialPreferenceModelling2014].
+Others conceive the externalities of transportation system as trade-offs and aim to maximize transport-related benefits (i.e., time savings, emissions reductions, congestion reductions, user fares) through optimization/location-allocation methodologies [@fakhrmoosaviIncorporatingTravelTime2021; @zhengAreabasedEquitablePricing2020; @wismadiSpatialPreferenceModelling2014].
 
 "Population" standards are often founded on "Well-being" conceptualisations, assessing what constitutes a satisfactory life related to transportation from a variety of socio-demographic and spatial angles.
-Methods include establishing threshold based on questionnaires and comparisons to recommended physical activity levels e.g.
-[@auchinclossDesignBaselineDescription2020; @mccormackRelationshipClusteranalysisDerived2012; @kimNeighborhoodEnvironmentWalkability2016; @timperioSafetyNumbersDoes2015], region-relative comparisons in health outcomes in a spatial unit e.g.
-[premature mortality rates @awuorInfluenceEnvironmentalHealth2019], spatial access benchmarks based on population-related characteristics e.g.
-[supermarket access @murphySupermarketAccessTransport2017; hospital access @pereiraGeographicAccessCOVID192021], summative per capita benchmarks e.g., [decent living energy consumption levels @raoDecentLivingEmissions2012], and community-informed spatial boundaries e.g., [EJ defined communities @rowangouldIdentifyingEnvironmentalJustice2016].
+Methods include: establishing thresholds based on questionnaires and comparisons to recommended physical activity levels [@auchinclossDesignBaselineDescription2020; @mccormackRelationshipClusteranalysisDerived2012; @kimNeighborhoodEnvironmentWalkability2016; @timperioSafetyNumbersDoes2015], region-relative comparisons in health outcomes in a spatial unit such as premature mortality rates [@awuorInfluenceEnvironmentalHealth2019], spatial access benchmarks based on population-related characteristics like supermarket access [@murphySupermarketAccessTransport2017] and hospital access @pereiraGeographicAccessCOVID192021], summative per capita benchmarks such as decent living energy consumption levels [@raoDecentLivingEmissions2012], and community-informed spatial boundaries like EJ defined communities [@rowangouldIdentifyingEnvironmentalJustice2016].
 While most of these papers use quantitative or mixed-methods approaches, some employ exclusively qualitative methods to explore themes [@berheAdaptationDissonanceQuality2014].
@@
 Papers that feature both "Population" and "Opportunity" standards are often founded on "Vertical equity", "Well-being", and/or "Transport-related social exclusion" conceptualisations.
-They feature mix-methods, with questionnaires and qualitative approaches for "population" standards and quantitative methods like accessibility indices for "opportunity" standards.
+They feature mixed-methods, with questionnaires and qualitative approaches for "population" standards and quantitative methods like accessibility indices for "opportunity" standards.
 Census data and household estimates within specific travel distances or times to key destinations identify social exclusionary situations [@chenExploringTravelCharacteristics2010; @danielsProposalAccessibilityPlanning2011; @sunPublicTransportAvailability2021; @sharmaPublicTransitAccessibility2021], linkages between transport disadvantages [@delboscTransportProblemsThat2011], areas experiencing transport poverty [@allenSizingTransportPoverty2019a; @churchillTransportPovertySubjective2019], food deserts [@mckeyCrowdsourcedMappingHealthy2020], or transport-related energy poverty [@robinsonDoubleEnergyVulnerability2020; @berryInvestigatingFuelPoverty2016; @berryDistributionalEffectsCarbon2019].
@@
 For example, [@racheleNeighbourhoodSocioeconomicTransport2017a] integrate various transport network properties to define an indicator supporting walkability and public transport access.
-Others evaluate infrastructure quality [@xuSocialInequalitiesPark2017], accident severity [@benevenutoExaminingTransportNeeds2020; @appleyardTransitCorridorLivability2017], and user-groups, especially disadvantaged ones with respect to vertical equity [@prasertsubpakijEvaluatingAccessibilityBangkok2012] of multi-criteria indicators.
+Others evaluate infrastructure quality [@xuSocialInequalitiesPark2017], accident severity [@benevenutoExaminingTransportNeeds2020; @appleyardTransitCorridorLivability2017], and user-groups, especially disadvantaged ones with respect to vertical equity of multi-criteria indicators [@prasertsubpakijEvaluatingAccessibilityBangkok2012].
 Some focus explicitly on affordability and barriers, proposing infrastructure enhancements for better inclusivity, especially for the most disadvantaged [@basuPracticalFrameworkBenchmarking2019; @songWomenWheelsGender2019; @welchEquityTransportDistribution2013], grounded in transport-related social exclusion [@kentPrioritizingLowstressEquitable2019] or capabilities approaches [@smithAccessibilityCapabilityMinimum2012].
@@
 Methods employed are primarily quantitative or mixed-methods, identifying inequalities through Gini coefficients [@fengTradeoffsMobilityEquity2014] and some develop composite indices [@agostfelipInclusiveModelAssessing2021; @mirandaBenchmarkingSustainableUrban2012; @corazzaMethodologyEvidenceCase2020], occasionally incorporating spatial analysis [@jephcoteGeospatialAnalysisNaturally2013; @carrierApplicationThreeMethods2014].
-Commonly, many use established thresholds or health guidelines e.g., WHO guidelines for Active Aging and targets included in the United Nations’ SDG [@agostfelipInclusiveModelAssessing2021], OECD or WHO thresholds for traffic noise level, $NO_2$ levels, available green space and $PM_{2.5}$ level [@kruizeEnvironmentalEquityRole2007; @iungmanImpactUrbanTransport2021; @apparicioCyclingOneMost2021; @khomenkoLiveableCityHealthy2020; @muellerSocioeconomicInequalitiesUrban2018].
+Commonly, many use established thresholds or health guidelines e.g., WHO guidelines for Active Aging and targets included in the United Nations’ SDG [@agostfelipInclusiveModelAssessing2021] and OECD or WHO standards for traffic noise level, $NO_2$ levels, available green space and $PM_{2.5}$ levels [@kruizeEnvironmentalEquityRole2007; @iungmanImpactUrbanTransport2021; @apparicioCyclingOneMost2021; @khomenkoLiveableCityHealthy2020; @muellerSocioeconomicInequalitiesUrban2018].
 These metrics are sometimes criticized for their general applicability.
@@
 
-Flexible frameworks are important as transport is derived demand.
+As transport is a derived demand, flexible frameworks are crucial.
 What is (in)formally considered (in)essential today, could change during the next pandemic or technological revolution.
-Creating adaptable transport equity frameworks that lend themselves to change with the demands of justice is as important to ensure resiliency to backsliding.
+Creating adaptable transport equity frameworks that lend themselves to change with the demands of justice is required to ensure resiliency to backsliding.
 From the diverse corpus reviewed, we argue for the necessity of setting standards from a systems approach, considering both _positive_ rights (a right to have access to sufficient quality of essential services, for example) and _negative_ rights (mobility of cars must be limited to reduce their impact on air quality, health, safety, and positive rights).
@@
 It outlines a conceptual and flexible "Where", "When", "Who", "What" and "How" framework with supporting definitions for structuring transportation fairness and equity analysis.
-It also applies the framework in the synthesis and critical appraisal of 165 academic articles that undertake standard- and conceptualisation- supported equity analysis; these studies use equity analysis as instruments to gauge proximity to their context's just situation.
+It also applies the framework in the synthesis and appraisal of 165 academic articles that undertake standard- and conceptualisation- supported equity analysis; these studies use equity analysis as instruments to gauge proximity to their context's just situation.
 Following this framework and the reviewed literature it answers:
@@
 -	**How** are equity standards and inequities measured?
+
 <!--Maybe we should answer these questions explicitly here? Maybe we should also urge practitioners to adopt this flexible framework in their equity synthesis?-->
-From what's reviewed and using the framework we recommend five calls to action.
+From the outlined framework and the reviewed corpus, we recommend the following five calls to action.
 
@@
 The conceptual grounding for standards is often left implicit within the literature: for example, @muellerSocioeconomicInequalitiesUrban2018 suggests the relative risk of mortality from transport-related air pollution should not be higher in deprived groups than the general population.
-While the coonceptual justice foundations are not explicitly declared, we infer the standard conceptualizes _wellbeing_ and could be in service of reparative justice.
+While the conceptual justice foundations are not explicitly declared, we infer the standard conceptualizes _wellbeing_ and could be in service of reparative justice.
 To move equity analysis outputs towards just transport futures, explicit inclusion of justice rationales (the "Why?") should become more common place.
@@
 
-Standards serve as benchmark for fairness, yet some standards are seemingly arbitrary in the reviewed literature.
+Standards serve as benchmarks for fairness, yet some standards are seemingly arbitrary in the reviewed literature.
 For example, @caoIndicatorsSocioSpatialTransport2017 proposes 20 ferries per day to avoid social exclusion for inter-island transport planning in the Philippines, though acknowledging the standard should be politically determined.
@@
 As @martensFair2021 convincingly show, being explicit about "Why" is important.
+Recent work by @karnerAdvancesPitfallsMeasuring2024 provides several concrete examples for developing standards that follow directly from different justice conceptualizations.
 
@@
 This lack of discussion makes translating results from these equity analysis into policy practice that may move conditions towards justice challenging.
-Creative methods and discussion of metrics of quantitative accessibility with intention of practical intervention should be paired with results; they should yield interpretable results and the explicit discussion of minimum _and_ maximum values in the distribution of the object of justice (the "What?"), as applicable, is critical.
-
-Other times, when analysis engages with metrics that may be tied to particular concepts of equity (like Gini coefficient or Theil index), they fall short in assessing whether the results are good or bad [e.g., @mijaresEquityAnalysisUrban2013].
+Creative methods and discussions of quantitative accessibility metrics should be paired with results; they should yield interpretable results.
+The explicit discussion of minimum _and_ maximum values in the distribution of the object of justice (the "What?"), as applicable, is critical.
+
+Other times, when analysis engages with metrics that may be tied to particular concepts of equity (like the Gini coefficient or Theil index), they fall short in assessing whether the results are good or bad [@mijaresEquityAnalysisUrban2013].
 If a Gini coefficient of 0 means that all people have the same access to public transport stops, what does it mean if the result is 0.2, 0.3, or 0.4?
@@
 These questions usually remain unanswered in the literature despite its importance.
-These measurements can also bring some challenges and pitfalls (as recently summarized by @karnerAdvancesPitfallsMeasuring2023) but are necessary to move equity analysis into application and towards transport justice.
+These measurements can also bring some challenges and pitfalls (as recently summarized by @karnerAdvancesPitfallsMeasuring2024) but are necessary to move equity analysis into application and towards transport justice.
 
@@
 Further, they may be missing categories all together like childcare, typically daycare or facilities -- domestic work, mobilities of care, and mobility interdependence.
-These missing categories are unrepresented in the reviewed relative to the presence of work and healthcare destinations.
+These missing categories are underrepresented in the reviewed corpus relative to the presence of work and healthcare destinations.
 
@@
 As such, data availability matters, especially in the context of the Global South and rural geographies as fewer official data sources and public research resources exist relative to urban communities and areas in the Global North.
-As well as in the operationalization of emerging conceptual theories (i.e., sufficientariansm [@vanderveenOperationalizingIndicatorSufficient2020]) in equity analysis.
+As well as in the operationalization of emerging conceptual theories in equity analysis, such as sufficientariansm [@vanderveenOperationalizingIndicatorSufficient2020].
 
@@
 Deciding who is the subject of justice frames the data collection activity; if it's the mobility of those that do domestic work, the classification of who does this work and how it changes over time/space is fundamental.
-How we classify has implications for our understanding of just who is a subject of justice, as illustrated by the history of racial classifications in the US [see @leeRacial1993].
+How we classify has implications for our understanding of just who is a subject of justice, as illustrated by the history of racial classifications in the US [@leeRacial1993].
 What methods we're using, under what spatio-temporal boundaries we conduct data collection, and who has (and doesn't have) access to the collected data can all impact fulsome conceptualizations of justice.
@@
 We believe there is a need to evaluate equity interventions and policies: track their before, during, and after impacts.
-In this review, only approximately 10% of studies assess specific transport-related policy interventions through an equity lens <!-- double check this number-->.
+In this review, only approximately 10% of studies assess specific transport-related policy interventions through an equity lens<!-- double check this number-->.
 Examples include mode-shift from driving to active school travel [@mammenSchoolTravelPlanning2014], transit fare restructures [@hickeyUsingQuantitativeMethods2010] and spatial analysis of Low Traffic neighbourhoods [@aldredEquityNewActive2021].
 The assessment of interventions and tracking associated outcomes can be thought of as a key step towards transport justice.
+Similarly, for how long the burdens and benefits can still be associated to a specific transportation intervention remains a critical avenue to explore.
 
@@
 
-## Why? as part of the framework to analyse questions of justice {#sec-sect61}
-
-To give an example drawing from the Canadian Charter of Rights and Freedoms, a number of rights and fundamental freedoms [including the liberty to move freely; see @chartopediaArt72023] are recognized to exist in Canada "without discrimination by reason of race, national origin, colour, religion or sex".
+<!-- ## Why? A framework extension to analyse questions of justice {#sec-sect61} -->
+
+<!-- To give an example drawing from the Canadian Charter of Rights and Freedoms, a number of rights and fundamental freedoms [including the liberty to move freely; see @chartopediaArt72023] are recognized to exist in Canada "without discrimination by reason of race, national origin, colour, religion or sex".
 Notice that this declaration includes several individual traits that define the subjects of justice, but does not consider age as one aspect of the person.
@@
 In actuality, the courts _created_ a standard of equity: starting at a certain age, children are objects of justice from the perspective of freedom of movement, and they are due the benefits of independent mobility by transit.
-
-Another perspective on the answer to "Why?" are cultural _norms_ that provide a rationale.
+-->
+
+<!-- Another perspective on the answer to "Why?" are cultural _norms_ that provide a rationale.
 In this case, the liberty of movement for all---including children---reigns supreme.
@@
 Whose voice, what perception of fear, and actions we take to intervene define and reproduce cultural norms, all play into the rationale behind the question "Why?".
+-->
 
@@
 #| fig-cap: "PRISMA flow diagram for the evidence selection process. ES signifies equity standard and EC signifies equity conceptualization."
-knitr::include_graphics("figures/Pflow_chart.png")
+knitr::include_graphics("figures/Pflow_chart.tiff")
 ```
@@
 The data extraction template was created with the aim of striking a balance between the complexity of categories and the simplicity of summary; information related to "Where?" (the geographical context and sphere of life), "When?" (temporal circumstances for the application of justice), "Who?" (the subject of justice), "What?" (the object of justice), and "How?" (equity standard(s) and concepts) was filled out for each study.
-@fig-figA4 contains the template that was input into *Covidence* and used throughout.
+The following table contains the template that was input into *Covidence* and used throughout.
 

### 2024-05-14 Journal-manuscript/Journal-manuscript.qmd
Update to declaration

@@
 
-During the preparation of this work the authors used ChatGPT exclusively for copyediting: aiming to reduce word count, keep original style and correct typos.
-After using this tool, the authors reviewed and edited the content as needed and take full responsibility for the content of the publication.
+During the preparation of this work, the first and second authors used ChatGPT exclusively for copyediting in the "An appraisal of the lay of the land" section, aiming to reduce word count while keeping the original writing style and correct typos.
+After using this tool, all authors reviewed the manuscript and take full responsibility for the content of the publication.
 

### 2024-05-15 Journal-manuscript/Journal-manuscript.qmd
Antonio's initial edits before submission

@@
 ---
-title: "Is it fair? Searching for standards of fairness in the transportation justice literature"
+title: "Searching for standards of fairness in the transportation justice literature"
 format: docx
@@
 ```{r setup}
-library(huxtable)
-library(sf)
-library(ggplot2)
-library(tidyverse)
-library(readxl)
-library(DiagrammeR)
-library(scales)
-library(corrplot)
-library(patchwork)
-library(flextable)
-library(ftExtra)
+library(huxtable) # Easily Create and Style Tables for LaTeX, HTML and Other Formats
+library(sf) # Simple Features for R
+library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
+library(tidyverse) # Easily Install and Load the 'Tidyverse'
+library(readxl) # Read Excel Files
+library(DiagrammeR) # Graph/Network Visualization
+library(scales) # Scale Functions for Visualization
+library(corrplot) # Visualization of a Correlation Matrix
+library(patchwork) # The Composer of Plots
+library(flextable) # Functions for Tabular Reporting
+library(ftExtra) # Extensions for 'Flextable'
 ```
@@
 ```
+
 # Introduction {#sect1}
 
-Transportation systems, as a class of essential technologies that facilitate or impede social inclusion and activity participation<!---[@churchTransport2000; @lucasTransport2001; @socialExclusion2003; @cassSocial2005; @casasSocial2007; @prestonAccessibility2007; @paezMobility2009]-->, have increasingly come into focus from the perspective of equity; a lively and rapidly growing literature has emerged on the topic [@martensTransport2016; @karnerAdvancesPitfallsMeasuring2024; @vecchioTransportEquityLatin2020].
-<!--@diciommoTransport2017; @guoSystematicOverviewTransportation2020; @karnerFrom2020; @pereiraTransportationEquity2021; @vanweeEvaluating2021; @zhangLiterature2021; @desjardinsExaminingEquityAccessibility2022; --> Aside from an ethical motivation, keeping track of objective and perceived inequalities could be of interest for governing bodies to respond to popular demands for fairness.
-However, transportation systems are notoriously complex: they simultaneously move people, goods, and information.
-Emerging technologies and service models can swiftly change the balance of benefits and burdens among a population [@guoSystematicOverviewTransportation2020], turning both users and service providers into digital rentiers [@birchBig2022].
-The benefits and burdens of transportation systems are diffuse over space and time posing a challenge in evaluating and tracking inequities.
-Transportation systems that are engineered to offer higher mobility for people _somewhere_ can simultaneously cut others off from essential opportunities _elsewhere_, as @rajeEngineering2004 poignantly illustrated with examples of infrastructure in the UK.
-Over time, the shades of policies past can continue to haunt a region and the planet as shown by the legacy of displacement and decay caused by urban highways all across the US [@archerWhite2020] and the time horizon for the impacts of climate change [@markolfTransportation2019].
-
-<!--But complexity is no excuse to shirk away from the task.--> The objective of this work is to scan the state of knowledge in terms of defining and operationalizing "fairness" in the transportation domain.
-Much research has been devoted to the issues of _measuring_ equity in transportation, including (among many others) [@ramjerdiEquity2006; @delboscUsing2011; @welchMeasure2013; @martensMeasuring2019; @pritchardNew2022].
-Further, there are multiple works that discuss the conceptual and philosophical foundations of equity and fairness in transportation [@martensTransport2016; @pereiraDistributiveJusticeEquity2017a; @vanoutriveHow2019].
-Previous reviews of planning documents have investigated equity from narrowly scoped perspectives, such as accessibility [@boisjolyHowGetThere2017] or a particular mode of transportation e.g., cycling [@doranPursuitCyclingEquity2021].
-These efforts are valuable to scholars, the public, and decision-makers alike.
-In our estimation however, there remains a gap in the literature in terms of understanding how standards for equity are developed and implemented within the transportation domain.
+Transportation systems are technologies essential for social inclusion and activity participation, and therefore important from an equity perspective [@martensTransport2016; @karnerAdvancesPitfallsMeasuring2024; @vecchioTransportEquityLatin2020].
+Beyond ethical motivations, tracking objective and perceived inequalities is of interest for governing bodies to respond to popular demands for fairness.
+This is a challenging task: transportation systems are notoriously complex, with benefits and burdens that are diffuse over space and time.
+To compound matters, emerging technologies and service models can swiftly change the balance of benefits and burdens among a population [@guoSystematicOverviewTransportation2020].
+Transportation systems that are engineered to offer higher mobility for people _somewhere_ can simultaneously cut others off from essential opportunities _elsewhere_, as @rajeEngineering2004 and others have poignantly illustrated.
+Over time, the shades of policies past long haunt a region and the planet, as shown by the legacy of urban highways all across the US [@archerWhite2020] and the time horizon for the impacts of climate change [@markolfTransportation2019].
+
+The objective of this work is to scan the state of knowledge in terms of defining and operationalizing fairness in transportation.
+Much research has been devoted to the issues of _measuring_ equity in transportation, including (among many others) [@delboscUsing2011; @martensMeasuring2019; @pritchardNew2022].
+Further, there are multiple works that discuss the conceptual and philosophical foundations of equity and fairness in transportation [e.g., @martensTransport2016; @pereiraDistributiveJusticeEquity2017a; @vanoutriveHow2019].
+Previous reviews of equity in planning documents have been tightly scoped to cover accessibility [e.g., @boisjolyHowGetThere2017] or a particular mode of transportation [e.g., cycling, @doranPursuitCyclingEquity2021].
+These efforts are valuable, but in our estimation there remains a gap in terms of understanding how standards for equity are developed and implemented for transportation systems.
 
@@
 
-While commendable, the strategies are too vague, which means it is possible to implement them in a myriad of ways, either genuinely to comply with the spirit of justice, or else performatively to deceive it [@mcculloughPerformative2023].
-Some relevant questions include: how much should the adverse effects be reduced?
+While commendable, the strategies are too vague, which means it is possible to implement them in multiple ways, to genuinely comply with the spirit of justice, or performatively to deceive it [@mcculloughPerformative2023].
+Relevant questions include: how much should adverse effects be reduced?
 To zero?
 Or to some tolerable level of adversity greater than zero?
-What should that level be and by what criteria?
+What should that level be?
 What benefits should be distributed?
-How much of the benefits should minority and low-income residents receive for policymakers to declare that a project is sufficiently ‘equitable'?
-These questions touch on the development and use of _standards_ for transportation justice.
-In this work, we choose to focus exclusively on the academic domain as we know that adoption of equity concepts in planning practice has lagged developments in academic work [@pereiraTransportationEquity2021; @boisjolyHowGetThere2017; @linovskiEquityTransportationPlanning2020; @doranPursuitCyclingEquity2021; @litmanEvaluatingTransportationEquity2022]
-
-Supporting the creation of (more) just transportation systems involves understanding the production and management of transportation benefits and costs; how they are distributed and what values are implemented (and by whom) in the form of standards [@pereiraDistributiveJusticeEquity2017a; @shellerMobilityJusticePolitics2018; @pereiraTransportationEquity2021].
-In this spirit, our work seeks to make two significant contributions.
-First, it outlines a conceptual and flexible framework with supporting definitions for structuring transportation fairness and equity analysis.
-Second, it applies the framework to collate the existing academic knowledge that undertakes transportation equity analysis supported by equity standards and conceptualisations.
-We hope the findings and the presentation of academic knowledge will be useful in supporting the development and implementation of equity standards for transportation planners, policy, and decision makers.
-In contrast to previous reviews on how to measure inequality in transportation systems, this work is concerned with the implied or explicit standards that are used to judge whether the measured inequalities are fundamentally “fair” or unacceptable.
-The present study achieves this objective by scanning the state of academic knowledge in defining “fairness” in the transport domain.
-Borrowing from philosophers of justice [@jaggarPhilosophicalChallengesGlobal2009], we ask the following questions to the literature and use them as a framework for presentation:
-
--	**What** are those benefits and burdens of transportation systems? Are they qualified in sidewalk widths, bus frequency, travel times, accessibility, or some other consideration?  
--	**Who** benefits and who is burdened by transportation systems? Along which lines of identity are transportation inequities examined?
--	**Where and when** are equity standards applied in the literature? What are the differences between the research depending on location, context-specific planning processes and time of publication? 
--	**How** are equity standards and inequities measured?
-Are they infrastructure provision thresholds, acceptable pollution levels, or some other kind of cut off?
-How are these studies born out of broader notions of justice, i.e.
-disability rights, environmental justice, egalitarianism, or something else?
+What share of the benefits should minority and low-income residents receive for a project to be sufficiently ‘equitable'?
+These questions touch on the development and use of _standards_.
+In this work, we choose to focus exclusively on the academic domain as we know that adoption of equity concepts in planning practice has lagged developments in academic work [@pereiraTransportationEquity2021; @boisjolyHowGetThere2017; @linovskiEquityTransportationPlanning2020; @doranPursuitCyclingEquity2021].
+
+Supporting the creation of (more) just transportation systems involves developing standards to track the production and management of benefits and costs; how they are distributed; what values they implement, as well as by whom [@pereiraDistributiveJusticeEquity2017a; @pereiraTransportationEquity2021].
+In this spirit, our work seeks to make two contributions.
+First, it outlines a conceptual and flexible framework for engaging questions of transportation fairness and equity analysis.
+Second, it applies the framework to collate the existing knowledge about equity standards.
+In contrast to previous reviews on measuring inequality in transportation systems, we are concerned with the implicit or explicit standards used to judge whether inequalities are fundamentally “fair” or unacceptable.
+To achieve this we scan the state of academic knowledge in defining standards of fairness in transportation.
 
@@
 
-In the introductory section the terms "justice", "equity", and "fairness" are used relatively loosely.
-This was done purposefully.
-As noted, people often have strong intuitions of what is "just", "equitable", and "fair" which may or may not match those of the authorities who set the "standards" of fairness.
-But in democratic societies, the authority of all those charged with the business of governing should derive from the will of the people.
-It is therefore important to explicitly state what we mean by these terms.
-In the following, we describe justice as embracing moral motivation, equity as measurement tools, fairness as a yardstick of justice, and standards as statements of fairness.
+In the introduction the terms "justice", "equity", and "fairness" were used relatively loosely.
+In the following, we describe justice as a target, equity as a set of measurement tools, fairness as a yardstick of justice, and standards as statements of fairness.
 
@@
 440].
-Justice is a fluid concept, because it depends on the desirability of different states of affairs, which may change between populations, over space and time.
-That said, it is possible to distinguish several forms of justice [@jaggarPhilosophicalChallengesGlobal2009; @pereiraDistributiveJusticeEquity2017a; @karnerTransportationEquityTransportation2020].
-
-- **Retributive justice** is concerned with the proportional retribution of wrongdoers with relation to legitimate punishers and the innocent [@sep-justice-retributive].
+Justice is fluid because it depends on the desirability of different states of affairs, which may change between populations, places, and times.
+That said, several forms of justice can be distinguished [@jaggarPhilosophicalChallengesGlobal2009; @pereiraDistributiveJusticeEquity2017a; @karnerTransportationEquityTransportation2020].
+
+- **Retributive justice** is concerned with the proportional retribution of wrongdoers relative to legitimate punishers and the innocent [@sep-justice-retributive].
+
 - **Reparative (or restorative) justice** focuses on the reparation of caused harm; it centers the needs and voices of victims to restore wrongdoers and the community according to moral values [@tylerRestorativeJusticeProcedural2006; @braithwaite2003principles].
-In planning and policy contexts, reparative justice is often conceived of as accountability mechanisms in which institutions that caused harm financially compensate victims [@safransky2022grammars].
+In planning and policy contexts, reparative justice often involves accountability mechanisms to compensate victims [@safransky2022grammars].
+
 - **Procedural justice** strives to ensure that the views and preferences of all stakeholders are fairly accounted for in the decision-making and inter-personal procedures affecting their lives and communities [@tylerRestorativeJusticeProcedural2006].
-- **Distributive justice** is perhaps the most studied form of justice [see @jaggarPhilosophicalChallengesGlobal2009, p.
-2] as well in the transportation domain [@pereiraDistributiveJusticeEquity2017a].
-Its main concern is the way the benefits and burdens of the tangible and intangible products of society are collected by different segments of a population.
-
-It might be argued that all of the above touch on forms of distributive justice.
-To illustrate, retributive justice, is usually achieved by distributing intangibles of a society's moral values like "freedom" (e.g., of movement, of association) as benefits, and/or the distribution of tangible resources as burdens (e.g., fines as a punishment).
-Reparative justice could look like the distribution of benefits and burdens to redress past wrongs, for example by asking those who have benefited from said wrongs, even if unwittingly, to shoulder a bigger fiscal burden in order to cover programs that mete benefits to those who are still harmed by past wrongs.
-Procedural justice could be the distribution of the benefits (e.g., the right to voice an opinion as a recognized stakeholder in the process) and burdens (e.g., the effort required to develop an educated opinion) of the processes that lead to decisions with collective consequences.
-
-From this perspective, we can then speak of the _purposes_ of distributive justice: to mete out retribution _fairly_, to repair past harms, and to ensure that procedures offer _equitable_ opportunities to influence outcomes.
-From this perspective, equity and fairness are the "instruments" of justice, the tools by which society advances towards the end goal of justice.
-
-**Equity**, as conceptualized alongside distributive justice, tends to encompass tools to understand the distribution of benefits and burdens of things among a population, often emphasizing those with the least advantage and their outcomes.
-In the transportation domain, the term is somewhat loaded because it is perceived as stemming from the authority of the state and is meant to assist with decisions about regulating and financing transportation spending [@karnerTransportationEquityTransportation2020].
-Here, we are in agreement with @karnerTransportationEquityTransportation2020 that equity analysis should not be seen as an end in and of itself, but rather as a means to gather information about actual and perceived inequities.
-In this respect, the analytical tradition of equity, at least in transportation planning, means that the relevant models become embedded in the "political ecology of the estimated truth" [@kingModels1993]: in principle, their assumptions and scope must be open and transparent, or else they may be more vulnerable to misuse and even abuse as tools of subjugation.
-
-**Fairness**, in contrast to equity, is somewhat more complicated to define.
+
+- **Distributive justice** is perhaps the most studied form of justice in transportation [@pereiraDistributiveJusticeEquity2017a] and elsewhere [see @jaggarPhilosophicalChallengesGlobal2009, p.
+2].
+Its main concern is the collection of benefits and burdens of the tangible and intangible products of society by different segments of a population.
+
+It might be argued that all forms of justice have a distributive facet.
+To illustrate, retributive justice is achieved by distributing societal intangibles like "freedom" (e.g., of movement, of association) as benefits, and/or tangible resources as burdens (e.g., fines as a punishment).
+Reparative justice deals with the distribution of benefits and burdens to redress past wrongs, for example by asking those who have benefited from said wrongs, even if unwittingly, to shoulder a bigger fiscal burden to benefit those who are still harmed by them.
+Procedural justice is the distribution of societal benefits (e.g., the right to voice an opinion as a recognized stakeholder in a process) and burdens (e.g., the effort required to develop an educated opinion) when making decisions with collective consequences.
+
+From this perspective, we can speak of the _purposes_ of distributive justice: to mete out retribution _fairly_, to repair past harms, and to ensure that procedures offer _equitable_ opportunities to influence outcomes.
+Accordingly, equity and fairness are the "instruments" of justice, the tools by which society moves towards it.
+
+**Equity**, as conceptualized alongside distributive justice, tends to encompass tools to understand the distribution of benefits and burdens of things among a population, often emphasizing those with the least advantage.
+In the transportation domain, the term is somewhat loaded because it is perceived as stemming from the authority of the state, and meant to assist with decisions about regulating and financing spending [@karnerTransportationEquityTransportation2020].
+We are in agreement with @karnerTransportationEquityTransportation2020 that equity analysis should not be seen as an end in and of itself, but rather as a means to gather information about actual and perceived inequities.
+In this respect, the analytical tradition of equity, at least in transportation planning, means that the relevant models become embedded in a political ecology of estimated truth, and their assumptions and scope must be open and transparent to avoid misuse and even abuse as tools of subjugation.
+
+**Fairness** is somewhat more complicated to define.
 The concept does not have the same history of development as an analytical tool, and can be interpreted in numerous, and possibly discordant ways.
-This is convincingly demonstrated by @martensFair2021 in their study of the application of Title VI of the Civil Rights Act of 1964 in accessibility planning in the US.
-Title VI explicitly talks about the distribution of benefits derived from Federal funding: "[n]o person in the United States shall, on the ground of race, color, or national origin, be excluded from participation in, be denied the benefits of, or be subjected to discrimination under any program or activity receiving Federal financial assistance." However, as @martensFair2021 show, there are several ways to comply with regulations while achieving different outcomes, ranging from the banal (do not _knowingly_ discriminate), to the substantive (compensation for past discrimination within a societal context that recognizes harm was done, i.e., reparative justice).
+This is demonstrated by @martensFair2021 in their study of the application of Title VI of the US Civil Rights Act of 1964 in accessibility planning.
+Title VI explicitly talks about the distribution of benefits derived from Federal funding: "[n]o person in the United States shall, on the ground of race, color, or national origin, be excluded from participation in, be denied the benefits of, or be subjected to discrimination under any program or activity receiving Federal financial assistance." However, as @martensFair2021 show, there are wildly different paths to compliance, ranging from the banal (do not _knowingly_ discriminate), to the substantive (compensation for past discrimination within a societal context that recognizes harm was done, i.e., reparative justice).
 What kind of justice does fairness serve in each case?
@@
 
-**Standards** are a way of making concrete statements about fairness.
-Returning to the ambiguities in Title VI discussed by @martensFair2021, the attainment of justice depends on the standard used to indicate fairness.
+A **standard** is a way of making concrete statements about fairness.
+Returning to the ambiguities in Title VI, the attainment of justice _depends on the standard_.
 For example, explicit non-discrimination constitutes a very weak standard that takes aim at the actions of agencies instead of the recipients of the benefits.
@@
 The standard provides conditions to determine whether a situation is _fair_.
-A similarly weak standard is a Pareto improvement, whereby it is possible to concentrate the benefits as long as no group is worse off compared to the status quo e.g., [@tan2016pareto; @xuParetoimprovingPoliciesIdealized2018].
-From this standard, a policy could be considered fair as long it does no harm.
+A similarly weak standard is a Pareto improvement, whereby it is possible to concentrate the benefits as long as no group is worse off compared to the status quo [@tan2016pareto; @xuParetoimprovingPoliciesIdealized2018].
+A policy based on this standard would be fair as long it does no harm.
 <!--A somewhat more strict standard, a Pareto-Plus improvement, stipulates that an intervention is fair when all groups receive at least _some_ (non-trivial) benefits; the size of the benefits for each group is irrelevant.
 In contrast to the notion of "do no harm", such a standard embodies the ideal that no one is denied benefits.
---> A standard based on egalitarian theories such as proportional equity would weigh the benefits or burdens by population, and fairness is achieved when each group gives or receives in proportion to their size [@billsLookingMeanEquity2017;@martensJusticetheoreticApproachDistribution2012] In contrast, an affirmative action standard based in restorative justice is even stricter.
-It requires the benefits to be distributed in a non-egalitarian way that favors those who are still harmed by past or present discriminatory practices [@bierbaumMobilityJusticeLinking2021].
+--> A standard based on egalitarian principles (e.g., proportional equity) would weigh the benefits or burdens by population; in fairness, each group would give or receive in proportion to their size [@billsLookingMeanEquity2017;@martensJusticetheoreticApproachDistribution2012].
+In contrast, an affirmative action standard (restorative justice) would be even stricter, requiring the benefits to be distributed in a non-egalitarian way to favor people still harmed by past or present discriminatory practices [@bierbaumMobilityJusticeLinking2021].
 
@@
 
-<!-- To summarise the discussion of definitions, justice is a goal of social progress.
-But to understand what that goal is, we must clearly define standards of fairness.
-Equity analysis is tool to measure where the actual or perceived distribution of the burdens and benefits of the products of transportation systems stand with respect to the standard, in other words, instruments to see how close or far we are from a just situation.
--->
-
-Give the discussed definitions of "justice", "fairness", "equity" and "standards", we approach the literature with an analytical apparatus inspired by the framing of @jaggarPhilosophicalChallengesGlobal2009 for philosophical questions of justice.^[Similar questions are found peppered throughout the literature.
+Having defined "justice", "fairness", "equity" and "standards", we approach the literature with an analytical apparatus inspired by the framing of @jaggarPhilosophicalChallengesGlobal2009 for philosophical questions of justice.^[Similar questions are found peppered throughout the literature.
 This is done either explicitly, as for example in @karnerTransportationEquityTransportation2020, who ask "of what", "for whom," and "how much" in reference to equity; or implicitly, as in @gosslingUrban2016, who asks of the outputs ("what?") of transportation (exposure, space, access) and "for whom?" (gender, age, ethnicity).].
@@
 
-In the case of justice in transportation, the question of **"Where?"** is paramount, as it might be argued that transportation generates inequalities by their very nature.
-By concentrating the effects of space-time convergence (for instance, by providing access to a transit system or a highway), an inequality is automatically generated.
-The burdens of transportation, in contrast, are often diffuse.
-They are incrementally paid, for example by a distributed population in the form of taxes, or by a population with a different spatial distribution in the form of poor health.
+In the case of transportation the question of **"Where?"** is paramount, as transportation by its very nature concentrates the benefits (e.g., access points to the system are not ubiquitous).
+The burdens, in contrast, are often diffuse: they are incrementally paid, for example by a distributed population in the form of taxes, or by a (possibly different) population in the form of poor health.
 As such, the answer to "Where?" is the definition of the spatial boundaries.
-For example, we can distinguish literature from the global South and North, allowing us to examine how they approach, deal with, and position themselves in different spatial contexts in addressing transport inequalities.
-
-Conventionally, the question of **"When?"** refers to the temporal circumstances within which the demands of justice have application.
-In the cause of transportation justice, one could ask about the temporal aspects of transportation systems, as examples: _when_ did the equity analysis take place and under what historical policy context; _the right time_ to invest in transportation infrastructure and as a result when to generate a spatial inequality [@rabelloEvaluation2015]; for _how long_ the burdens and benefits can still be associated to a specific transportation intervention; or even _timelines_ of reparative justice interventions that reconcile the shadows of past transportation-related injustices.
-In this literature review, given the amount of papers reviewed and the inconsistencies in how temporal aspects were addressed, we exclusively focus on the paper's publication year.
-
-When answering **"Who?"**, we inquire about which entities should be regarded as subjects/arbiters of justice, meaning those entitled to make claims of moral consideration from the perspective of justice.
-To make it tractable, this question is often approached through the filter of population groups, which may include several concurrent traits, such as gender identity, ableness, ethnicity, age, caste, and income.
-Often, it is appropriate to reflect on the intersections between traits, given differences in a person's lived experiences.
-A possible complication in the case of transportation is that disentangling the "Who" from their mobility tools is not always straightforward.
-Although a person is not their mode of transportation, there are large segments of the population who live in situations where they cannot extricate themselves from the mobility tools they can use, either because they have driven themselves out of choices [@laveryDriving2013], or have been driven out of choices by factors beyond their control such as captive users of a single mode [@jacquesRescuing2012; @cherancheryInvestigatingPerceptionCaptive2018].
-<!-- In societies that have grown into transportation monocultures with a predilection for automobility [@millerCollaborative2011] there may actually be less choice about mobility tools than we would like to assume.-->While it is important to avoid conflating the "Who" with the "What", for analytical purposes we need to be mindful of the connection between a person and their mobility tools.
-<!-- In the case of transportation, in addition to members of the public who use transportation systems, there is another category of **Who**, that stands possibly in opposition to users, namely the entities charged with providing services, maintaining infrastructure, and so on.
-These could be ministries or departments of transportation, transit agencies, public works departments and others having the power to act upon transportation (in)justices.
-Identifying these entities is relevant to elucidate who is responsible, for example for apportioning the benefits or mitigating the burdens of transportation.-->
-
-**"What?"** asks which entities should be regarded as objects of inequities, meaning which categories of things should be distributed in a just manner.
-To understand the distributional implications of transportation systems, it is essential to understand what do they _produce_?
-Transportation systems are space-time convergence technologies, mobility tools that improve the rate at which time is traded for space.
-They can increase the speed of movement: for instance, sidewalks and traffic lights enable potential movement of pedestrian and vehicles (i.e.
-mobility) and the realization of mobility through travel to needed/desired destinations.
-But as the adage goes, travel is derived demand [@mokhtarianUnderstanding2001; @whalenMode2013; @redmondPositive2001; @paezEnjoyment2010].
-For this reason, we cannot stop at only considering mobility but must consider its ulterior goal-- reaching destinations.
-Coupled with land use, mobility creates accessibility to needed or desired activities as well as varied burdens.
+
+Conventionally, the question of **"When?"** refers to the temporal circumstances within which the demands of justice apply.
+When it comes to transportation, questions regarding temporailty are important, as for example: _when_ did the equity analysis take place and under what historical policy context; when is _the right time_ to invest in infrastructure and as a result when to generate a spatial inequality [@rabelloEvaluation2015]; for _how long_ the burdens and benefits can be associated to a specific intervention; or even the _timeline_ of reparative justice to reconcile the shadows of past transportation-related injustices.
+
+When asking **"Who?"** we think about the entities that should be regarded as subjects/arbiters of justice.
+For tractability, this question is often approached through the filter of population groups, which may include several concurrent traits, including gender identity, ableness, ethnicity, age, caste, and income.
+Often, it is appropriate to consider the intersections between traits, given differences in a person's lived experiences.
+A complication in the case of transportation is that disentangling the "Who" from their mobility tools is not always straightforward.
+Although a person is not their mode of transportation, there are large segments of the population who cannot extricate themselves from the mobility tools they use, either because they have driven themselves out of choices [@laveryDriving2013], or have been driven out of choices by factors beyond their control [@jacquesRescuing2012].
+While it is important to avoid conflating the "Who" with the "What", for analytical purposes we need to be mindful of the connection between a person and their mobility tools.
+
+**"What?"** refers to which entities should be regarded as objects of inequities, meaning which categories of things should be distributed in a just manner.
+To understand the distributional implications of transportation systems, it is essential to understand what they _produce_.
+Transportation systems are technologies to improve the rate at which space is traded for time by increasing the speed of movement.
+But as the adage goes, travel is derived demand [@mokhtarianUnderstanding2001].
+For this reason, we cannot stop at considering only mobility but must consider its ulterior goal--reaching destinations.
+Coupled with land use, mobility creates accessibility to opportunities as well as varied burdens.
 Transportation justice, thus, involves proximate (mobility tools and mobility) and ulterior (accessibility and activity opportunities) objectives.
-<!-- As well, the burdens of transportation are also varied.
-Some are paid directly by the traveler (e.g., travel time, out-of-pocket costs), but many others are indirect and related to network externalities (e.g., exposure to pollution, road accidents).-->
 
 The next question is **"How?"**, and it relates to the allocation of various objects of justice ("what") to various subjects of justice ("who") in various circumstances ("when" and "where").
-Equity standards are a tool for answering this distributive question: how do we allocate burdens and benefits?
-Standards are thresholds that, when operationalized effectively define what is fair.
+Equity standards are a tool for answering this distributive question, by defining what is fair.
 The thresholds can be quantitative (e.g., square meters of green space per capita), or they can be qualitative descriptions (e.g., do not knowingly discriminate), or a mix of the two.
-Some examples include: maximum travel distance/cost/time to or from key destinations, levels of maximum exposure to externalities (i.e., noise or air pollution),un/fulfilled needs, and dis/satisfaction with travel.
-A number of conceptualisations exist to support us when approaching this question, and we can draw from concepts in transport-related social exclusion, transport disadvantage, and/or transport poverty, which are typically based on equity principles, such as utilitarianism, Sen's capabilities approach, or Sufficientarianism.
+Some examples include: maximum travel distance/cost/time to or from key destinations, levels of maximum exposure to externalities (i.e., noise or air pollution), un/fulfilled needs, and dis/satisfaction with travel.
+To support us when approaching this question we draw from concepts in transport-related social exclusion, transport disadvantage, and/or transport poverty, which are typically based on utilitarian or sufficientarian philosophies.
 
@@
 This is perhaps the most slippery of all the questions posed here.
-Justice is an inherently social construct.
-Asking **Why?** amounts to asking what sort of social contracts regulate human interactions.
-In other words, what are the rules that our collective will to believe imposes on each of us.
+Justice is inherently a social construct.
+Asking **Why?** amounts to asking what sort of social contracts regulate human interactions, that is, the self-imposed rules that result from our collective will to believe.
 These contracts can be defined by constitution, but there are often unwritten and possibly contested variants.
-In this way, analyzing the "Why?" in the reviewed literature is not the focus of this review, partly because answers to "Why?" are not explicitly stated and challenged inference.
-As such, our focus is on the standards of fairness that, combined with the use of equity analysis, can help us understand how better to move towards just transportation systems and better formulate answers to "Why?".<!-- To see an example of "Why?" drawing from the Canadian Charter of Rights and Freedoms, see the Appendix.-->
+In this way, analyzing the "Why?" in the reviewed literature is not the focus of this review, partly because answers to "Why?" are seldom explicitly stated.
+Instead, our focus is on the standards of fairness that, combined with the use of equity analysis, can help us understand how better to move towards just transportation systems and better formulate answers to "Why?".
 
@@
 
-This review examines the breadth and depth of the academic literature on transportation to identify the extent to which standards for equity are defined and employed.
+This review examines the academic literature on transportation to identify the extent to which standards for equity are defined and employed.
 In this task, we follow the Joanna Briggs Institute (JBI) approach to the conduct of scoping reviews, an approach that builds upon the Arksey and O'Malley (2005) framework [@petersUpdatedMethodologicalGuidance2020].
 The review is also guided by the Preferred Reporting Items for Systematic Reviews and Meta-Analyses, particularly the extension for scoping reviews (PRISMA-ScR) which is consistent with the JBI approach [@triccoPRISMAExtensionScoping2018].
-The use of these methods allows us to explore, in a consistent and organized manner, a relatively specialized topic within the broader transportation literature.
 
 The primary research question and the protocol were initially defined by the authors, a group of experts in the field of transportation.
-The initial draft of the protocol was refined from preliminary searches of related reviews e.g., [@zhangLiterature2021; @pereiraDistributiveJusticeEquity2017a; @vecchioAccessibilityCapabilitiesApproach2021], and in consultation with a University of Toronto Research Services Librarian and a Liaison Librarian in City Studies.
-
-The search strategy was first developed iteratively using **inclusion** and **exclusion** criteria [@petersUpdatedMethodologicalGuidance2020].
+The initial draft of the protocol was refined from preliminary searches of related reviews [e.g., @zhangLiterature2021; @pereiraDistributiveJusticeEquity2017a; @vecchioAccessibilityCapabilitiesApproach2021], and in consultation with research services librarians.
+
+The search strategy was developed iteratively using **inclusion** and **exclusion** criteria [@petersUpdatedMethodologicalGuidance2020].
 For the inclusion criteria, the mnemonic PCC (population, concept, and context) was adopted.
-The search strategy was refined iteratively, adding topic search terms by stages (e.g., terms in the title, abstract or key words).
-The terms were bundled by means of logical connector terms "AND" and "OR".
+The search strategy was refined by adding search terms.
+The terms were bundled by means of Boolean operators.
 These stages are summarized as follows (see Appendix section @sec-sect61 for details):
@@
 The first describes theories and concepts of equity, the second describes the object of justice (i.e., the "what" in our analytical framework), and the third describes terms referring to standards (i.e., the "how").
-All three sets of terms were augmented following an iterative process of refinement.
-The final search query took the following general form: ("Transport" OR "Transit" OR "Car\*" OR "Walk" OR "Bike"...**1**) AND ("Equity" OR "Justice" OR "Equity" OR "Fair"...**2**) AND ("Accessibility" OR "Mobility" OR ...**3**) AND ("Standard" OR "Threshold" OR ...**4**) where **1**,**2**,**3**, and **4** signify additional terms included in the sets combined with "OR" logical connectors.
-
-After testing the search strategy on WoS Core Collection, we proceeded to apply to an augmented list of databases.
+These sets of terms were augmented iteratively.
+The final search query took the following general form: ("Transport" OR "Transit" OR "Car\*" OR "Walk" OR "Bike"...**1**) AND ("Equity" OR "Justice" OR "Equity" OR "Fair"...**2**) AND ("Accessibility" OR "Mobility" OR ...**3**) AND ("Standard" OR "Threshold" OR ...**4**) where **1**,**2**,**3**, and **4** signify additional terms included in the sets combined with "OR" operators.
+
+After testing the search strategy on WoS Core Collection, we applied it to an augmented list of databases.
 The databases used were: WoS General Collection-Science Citation Index Expanded, WoS Social Sciences Citation Index, and Transportation Research International Documentation (TRID).
 The definitive version of the search was completed and exported by the lead author on March 21st, 2021.
-
-The nature of the search strategy tended to be overly inclusive since we aimed to reduce the risk of omitting relevant material.
-The authors and a team of research assistants trimmed the initial corpus of documents from 6832 down to 165.
-All documents were first excluded based on title/abstract relevance where two research assistants voted, and a third from the authorship team broke ties.
-Then, 1710 documents were assessed based on full-text eligibility, again with each being voted on by two research assistants and an authorship team member tie-breaker.
-Next, using a data extraction template and workflow that was pilot-tested with a subset of papers, the authorship team extracted data from the eligible documents.
-*Covidence* [@covidenceCovidence2023], an online application for literature screening, was used for all steps of selection and data extraction on the full export of literature.
-Covidence is designed for collaborative work, and helps to document the work of multiple reviewers.
+The number of documents identified in this was was 6,382.
+
+The semi-automated nature of the search strategy was overly inclusive, and reduced the risk of omitting relevant material.
+The next stage was to trim the corpus of documents, a task that could no longer be automated.
+The authors working with a group of undergraduate research assistants scanned the documents to assess their relevance based on titles and abstracts.
+Two research assistants voted on each document, and a third vote from the authorship team broke ties.
+After this step, 1,710 documents were assessed based on full-text, again with each being voted on by two research assistants and an authorship team member tie-breaker.
+Next, using a data extraction template and workflow that was pilot-tested with a subset of papers, the authorship team extracted data from the eligible documents using *Covidence* [@covidenceCovidence2023], an online application for literature screening.
 The evidence selection workflow, (PRISMA) flow diagram [@pagePRISMA2020Statement2021], data extraction template, and some sample data extractions can be consulted in Appendix @fig-figA2.
-
-To summarize, 6832 documents entered screening and ultimately 165 were retained for data extraction as part of the evidence corpus.
-These 165 pieces of evidence are relevant to the transportation domain and contain an explicit or inferred equity conceptualisation and standard.
+The end result of this scan was a corpus of 165 documents retained for data extraction.
 
@@
 
-This section threads together the trends identified in the corpus through the data extraction process (@fig-figA2)<!-- update this cross reference!-->.
-Specifically, conceptualisations and standards (the "How") and their applications to the objects of justice (the "What"), the subjects of justice (the "Who"), and under which situations ("When" and "Where").
-As an overview, @fig-fig1 visualises the prominence of each category under "When?", "Where?", "Who?", and "What?".
-The "How?" category, with its conceptualisations and standards, is visualised in the following @fig-fig2.<!-- For the reasons reviewed, setting standards (that is, making statements of fairness) is a highly context-specific endeavor.
-While our search strategy and review tries to be as faithful to the literature as possible, it does not allow us to lay claim to reporting about on-the-ground transportation justice efforts from a truly global perspective.-->
+This section threads together the trends identified from the data extracted from the corpus (@fig-figA2)<!-- update this cross reference!-->.
+Specifically, the "How" and its application to the objects of justice (the "What"), the subjects of justice (the "Who"), and under which situations ("When" and "Where").
+As an overview, @fig-fig1 shows the prominence of each category under "When?", "Where?", "Who?", and "What?".
+The "How?" category is shown in @fig-fig2.
 
@@
 
-Within our corpus most papers (`r ((case_cont_count$count[4] + case_cont_count$count[5]+ case_cont_count$count[6])/165) |> percent()`) focus on case studies in the Global North, with many studies from North America (particularly USA and Canada), Europe (particularly France, Spain and Northern Europe), Oceania (Australia and New Zealand), and Asia (Japan and Israel).
-Though their subject matter is varied, their spatial context mainly pertains to North America and Europe, and thus more often than not, in conversation with more developed and formal government transport planning apparatuses and technologies e.g., planning for equitable high-speed rail [@monzonEfficiencySpatialEquity2013], autonomous vehicle technology [@eppenbergerOpportunitySharedAutonomous2021], or on the public consultation processes [@reddySafeguardingMinorityCivil2010].
+Within our corpus most papers (`r ((case_cont_count$count[4] + case_cont_count$count[5]+ case_cont_count$count[6])/165) |> percent()`) focus on studies in the Global North, with many studies from North America (particularly USA and Canada), Europe (particularly France, Spain and Northern Europe), Oceania (Australia and New Zealand), and Asia (Japan and Israel).
+Though their subject matter is varied, their spatial context mainly pertains to North America and Europe, and thus more often than not deal with more developed and formal government transport planning apparatuses and technologies e.g., planning for equitable high-speed rail [@monzonEfficiencySpatialEquity2013], autonomous vehicle technology [@eppenbergerOpportunitySharedAutonomous2021], or on the public consultation processes [@reddySafeguardingMinorityCivil2010].
 
@@
 Many of these studies mention a systemic absence of evidence relevant to the region [@vecchioTransportEquityLatin2020].
-Despite the growing recognition in the literature of the interconnections between transport development, social exclusion, and poverty [@benevenutoExaminingTransportNeeds2020], a number of studies underscore ongoing neglect of the social dimension of transport during the planning stage [@benevenutoExaminingTransportNeeds2020; @boisjolyAccessibilityMeasurementsSao2020].
+Despite a growing recognition of the interconnections between transport development, social exclusion, and poverty [@benevenutoExaminingTransportNeeds2020], a number of studies underscore ongoing neglect of the social dimension of transport during the planning stage [@benevenutoExaminingTransportNeeds2020; @boisjolyAccessibilityMeasurementsSao2020].
 Many studies also point at affordability as one of the main mobility barriers in the region [@falavignaAssessingInequalitiesPublic2016; @rivasHowAffordableTransportation2018], while some highlight multi-dimensional concerns such as public transport accessibility and quality of walking environments that contribute to mobility inequalities [@tiznadoaitkenRoleAccessibilityPublic2018].
-
-Studies pertaining to Africa are even less numerous compared to the South American literature.
-A shared characteristic among the studies from these two continents is a scarcity of official transport data [@friedMeasuringSustainableDevelopment2020].
+Studies pertaining to Africa are even less numerous.
+
+A shared characteristic among studies from Africa and South America is a scarcity of official transport data [@friedMeasuringSustainableDevelopment2020].
 These studies also incorporate the use of informal transportation options and tensions in developing road network infrastructure (which tends to support car dependency) over meeting the mobility/accessibility needs of citizens more equitably and sustainably [@thondooSmallCitiesBig2020].
@@
 @ampeImpactChildBike2020, on the other hand, work to identify the lateral clearance that motorists should maintain when passing cyclists with children seats.
-<!-- While both studies refer to the same objects of justice ("what", i.e., the right to the road and cycling), the subjects of inequities (the "who") are different (people with disabilities and children), and the "where", that is, the geographical context for the examination of fairness, differs as well.--> The remainder of the studies reviewed focus on rural regions (`r (urban_form_count$count[2]/sum(urban_form_count$count)) |> percent()`).
-We highlight the work of @caoIndicatorsSocioSpatialTransport2017, who examined transportation disadvantage in remote places which rely on inter-island ferry trips in the rural Philippines.
+The remainder of the studies focus on rural regions (`r (urban_form_count$count[2]/sum(urban_form_count$count)) |> percent()`), or remote regions, such as those that rely on inter-island ferry trips in the Philippines [@caoIndicatorsSocioSpatialTransport2017].
 Similarly, @parrySocialVulnerabilityClimatic2018 studied remote communities in the Amazonian region, and suggest that "increasing accessibility through road building would be maladaptive, exposing marginalized people to further harm and exacerbating climatic change by driving deforestation" (pp.
@@
 - affordability as a barrier at the user- or policy maker- level is more often the motivation in work from the Global South;
+
 - the expression of greater tensions in investing in new transportation infrastructure, such as roads in rural/under-developed areas, compared to prioritization of non-car modes.
 Studies centered in the Global North often focus on the later; and
+
 - more significant data availability limitations and reliance on 'informally' collected data.
 
-We surmise these differences center on the direct economic outcomes tied to transport infrastructure.
-Often, work from the Global South does not engage as intimately with emerging mobility technologies that have large capital costs.
-Informal aspects present in transport planning also have a more overbearing presence.
-In the present day, countries in the Global South still struggle with the consequences of past colonialism under Northern states.
-This has left them more reliant on primary sector exports (lower efficiency, lower national GDP) under growing global financial markets, and with more fragile democracies.
-We infer, because of lower data availability, reliance on crowd-sourced or ‘informally’ collected data, and more extreme needs for ‘sufficient’ transport, analysis of transportation inequities is often cast along economic lines.
-
-We characterise the Global North and South literature as addressing and being situated at different points in the process of addressing transport inequities.
-As Global South' formal planning processes operate under stricter financial constraints, they rely on more informal processes to address unmet needs - a significant source of equity concerns e.g, informal transit in Nairobi [@friedMeasuringSustainableDevelopment2020], populations living in informal residents in Mumbai [@sharmaPublicTransitAccessibility2021].
-Operationalization of international standards are also more relevant e.g., using the WHO noise and air pollution standard in Delhi case study [@apparicioCyclingOneMost2021].
+These differences center on the direct economic outcomes tied to transport infrastructure.
+Often, work from the Global South does not engage as intimately with emerging mobility technologies.
+As well, the informal aspects of transport planning are more overbearing.
+Clearly, countries in the Global South still struggle with the consequences of past colonialism, which has left them more reliant on primary sector exports (lower efficiency, lower national GDP) under growing global financial markets, and with more fragile democracies.
+Due to lower data availability, reliance on crowd-sourced or ‘informally’ collected data, and more extreme needs for ‘sufficient’ transport, analysis of transportation inequities is often cast along economic lines in the Global South.
+
+The literatures on the Global North and South literature seem to be situated at different stages in the process of addressing transport inequities.
+As formal planning processes in the South operate under greater financial precariety, they rely on more informal processes to address unmet needs.
+This turns out to be a significant source of equity concerns e.g, informal transit [@friedMeasuringSustainableDevelopment2020], populations living in informal settlements [@sharmaPublicTransitAccessibility2021].
+Lack of resources makes international standards highly relevant e.g., using the WHO noise and air pollution standard [@apparicioCyclingOneMost2021].
 In contrast, these minimum standards may not be relevant in some areas of the Global North e.g., $NO_2$ pollution near roadways in Montreal are consistently below the WHO standard though populations are disproportionately impacted [@carrierApplicationThreeMethods2014].
-In these ways, the Global South temporally lags Global North development.
-However, this lag presents opportunities for all: For the Global South, to adopt potentially successful enhancements from Global North formal equity planning processes (e.g., indicator creation for disadvantages groups [@cuiSpatialAccessPublic2020]) and not repeat past mistakes (e.g., from entrenched car-centric development [@warrenDevelopingEquitableSustainable2015] to the disproportionate contribution to carbon-intense mobility [@chancelCarbonInequalityKyoto2015].
-For the Global North, there is an opportunity to reckon with their own contributions to uneven development globally and environmental impacts as well as adopt relevant aspects from informal planning processes.
-<!--`r word_comment("I condensed this paragraph.
-The idea here is the global south temporally lags the global north in how equity is assessed/planned for, but both can learn lessons from these differences.
-Both corpuses can be seen at different points of equity assessment developlment.
-Why is this the focus?
-Because, I can't clearly see differences in --one year or multiple time frames, if they do before and after evaluations or just dianoses, etc.-- in the GS vs GN as you suggest.
-If you don't like what's written, or if it's not clearly a temporal angle, we can maybe talk about how the focus of 'who' changes depending on the time published (i.e., it used to be car vs.
-transit, but now it's car vs.
-all other modes -- especially cycling and walking).
-However, you can also see how 'time' here is just another way to examine the who/what/where/when/how." , ".")` -->
+Overall, the Global South lags behind the Global North, but this situation presents surprising opportunities for the Global South, to adopt potentially successful enhancements from Global North formal equity planning processes (e.g., indicator creation for disadvantages groups [@cuiSpatialAccessPublic2020]) and to avoid past mistakes (e.g., from entrenched car-centric development [@warrenDevelopingEquitableSustainable2015] to the disproportionate contribution to carbon-intense mobility [@chancelCarbonInequalityKyoto2015].
+For the Global North, there is an opportunity to reckon with its own contributions to uneven development globally and environmental impacts as well as adopt relevant aspects from informal planning processes.
 
@@
 ### Population groups
-Turning to the "Who?" of population group characteristics in @fig-fig1, the majority focus on "Income", especially the lowest-income groups who face lower mobility and accessibility and higher costs and exposure [@peungnumsaiGridBasedSpatialAnalysis2020; @zhaoUnequableSpatialAccessibility2020; @falavignaAssessingInequalitiesPublic2016].
-Evidence suggests that low household income is a significant determinant of transport-related inequities e.g., urban access to public transport in Thailand [@peungnumsaiGridBasedSpatialAnalysis2020], access to urban employment opportunities in Brazil [@boisjolyAccessibilityMeasurementsSao2020], and unfavorable rates of environmental noise, air pollution, and green space in Netherlands [@kruizeEnvironmentalEquityRole2007].
-Yet, low-income is not universally associated with lower transport-related benefits for every object of justice - context matters.
-For instance, in Sheffield, UK, @mearsUnderstandingSocioeconomicEquity2019 demonstrate that historically working-class (lower-income) neighbourhoods have _more_ access to green space than other neighbourhoods.
-This is potentially due to historic urban planning approaches, though the quality is less than average.
-Similarly, @bertrandMeasuringMappingDisparities2008 find that lower income groups do not always have below average accessibility depending on the granularity of analysis; similar findings are uncovered in other spatial and temporal contexts [@fothEquitableTransitExamining2013a; @allenSizingTransportPoverty2019a]
-
-"Age" is the second most common population group category: these papers highlight differences in age-related capabilities.
-For instance, @martinez-jimenezAccessibilityCultureEducation2019 and @arranz-lopezSocialSpatialEquity2019 investigate travel distances/times to various opportunities, demonstrating how age is associated with differences in opportunity access.
-The most extreme age groups are commonly of focus: namely, school-aged children and older populations.
+
+Turning to the question of "who?" in @fig-fig1, the focus of the literature tends to be on "Income", especially the lowest-income groups who simultaneously face lower mobility and accessibility _and_ higher costs and exposure [@peungnumsaiGridBasedSpatialAnalysis2020; @zhaoUnequableSpatialAccessibility2020; @falavignaAssessingInequalitiesPublic2016].
+Evidence suggests that low household income is a significant determinant of transport-related inequities e.g., urban access to public transport [@peungnumsaiGridBasedSpatialAnalysis2020], access to urban employment opportunities [@boisjolyAccessibilityMeasurementsSao2020], and unfavorable rates of environmental noise, air pollution, and green space [@kruizeEnvironmentalEquityRole2007].
+Yet, low-income is not universally associated with lower transport-related benefits for every object of justice.
+For instance, in Sheffield, UK, @mearsUnderstandingSocioeconomicEquity2019 demonstrate that historically working-class (lower-income) neighbourhoods have _more_ access to green space than other neighbourhoods, but of lower quality, likely due to historic urban planning approaches.
+Similarly, @bertrandMeasuringMappingDisparities2008 find that the granularity of the analysis matters, and lower income groups do not always have lower accessibility , something echoed in other spatial and temporal contexts [@fothEquitableTransitExamining2013a; @allenSizingTransportPoverty2019a]
+
+"Age" is the second most common personal trait.
+In this regards, @martinez-jimenezAccessibilityCultureEducation2019 and @arranz-lopezSocialSpatialEquity2019 investigate travel distances/times to various opportunities, demonstrating how age is associated with differences in opportunity access.
+School-aged children and older populations are a common focus.
 For children, analysis of wellbeing [@laszkiewiczChildrenGreenWalk2020], safety and access to schools [@sharmaSpatialSocialInequities2022; @corazzaMethodologyEvidenceCase2020], and promoting active travel [@mackieOvercomingBarriersCycling2009; @mehdizadehWalkingTimeSchool2017] are common-place.
-Papers that focus on older adults similarly investigate transport-related wellbeing [@chenSpatialAnalysisFramework2020], access to age-specific destinations [@chengInvestigatingWalkingAccessibility2019], and improved options to meet unmet travel needs [@nordbakkeTransportUnmetActivity2015].
-
-Typically, the corpus features papers that focus on intersecting (multiple) characteristics.
-However, the exception are what we classify as studies that focus on  "(Dis)abilities" or "All Populations".
-These categories are often of exclusive focus.
+Papers that focus on older adults similarly investigate transport-related wellbeing [@chenSpatialAnalysisFramework2020], access to age-specific destinations [@chengInvestigatingWalkingAccessibility2019], and options to reduce unmet travel needs [@nordbakkeTransportUnmetActivity2015].
+
+Many papers focus on intersecting characteristics.
+As an exception, we classify some studies as focusing on  "(Dis)abilities" or "All Populations".
 Studies with a "(Dis)abilities" focus assess travel capabilities, namely through physical accessibility and universal design guidelines [@parkJourneyVisuallyImpaired2017; @chiscanoImprovingDesignUrban2021; @orellanaWalkRollMapping2020].
 "All Populations" papers make no distinction in population.
-This is done, for instance in @kitaDoesLevelTransportation2020, who investigates disparities in accessibility to food-stores and their self-reported capability/frequency of going outdoors.
-Often in both "(Dis)abilities" and "All Populations" focused papers, an implicit or explicit motivation is access (to necessities and desired destinations) for all.
-
-A large proportion of papers apply a "Composite Vulnerability Index"; an _informed_ combination of several individual traits like low income, unemployment and immigrant status.
-These indices are generated from official government sources or author-informed census data creation methods.
-As an example, the Neighbourhood Equity Index (NEI) is a neighbourhood-ranked measure of vulnerability created by the City of Toronto and used in @awuorInfluenceEnvironmentalHealth2019 to examine disparities in premature death.
+This is done, for instance in @kitaDoesLevelTransportation2020, who investigates disparities in accessibility to food-stores and self-reported capability/frequency of going outdoors.
+Often, the implicit or explicit motivation of these papers is access (to necessities and desired destinations) for all.
+
+A large proportion of papers apply composite vulnerability indices that combine several individual traits like low income, unemployment, and/or immigrant status.
+These indices are generated from government sources or author-informed census data creation methods.
+As an example, the Neighbourhood Equity Index (NEI) is a measure of vulnerability created by the City of Toronto and used in @awuorInfluenceEnvironmentalHealth2019 to examine disparities in premature death.
 Other works use national census indicators such as the social and housing deprivation index [@pucciInequalitiesJobrelatedAccessibility2019] or explore transport disadvantage, equity in policy implementation, or transport-related mortality burden by means of census measures (e.g., household poverty) and transport-related accessibility indicators [@aldredEquityNewActive2021; @iungmanImpactUrbanTransport2021; @sunPublicTransportAvailability2021; @scheurerSpatialAccessibilityPublic2017].
@@
 
-Multi-dimensional considerations are so prevalent in the corpus that papers studying "Gender", "Race/ethnicity", "Education", or "Employment" level characteristics are infrequently studied exclusively.
+Multi-dimensional considerations are so prevalent in the corpus that "Gender", "Race/ethnicity", "Education", or "Employment" are infrequently studied in isolation.
 Only a few papers focus exclusively on gendered differences in active transportation [@adlakhaMindGapGender2020, @xieWonCycleRoute2018], race/ethnicity's relationship to green space proximity [@silvaEnvironmentalJusticeAccessibility2018], and culturally-appropriate opportunities [@wangModelingSpatialAccessibility2011].
-Papers that focus *solely* on "Employment" status or "Education" level are absent all together.
-Furthermore, "Other" population groups are also frequently considered in the corpus: this is a catch-all category that includes group populations less commonly the subject of research e.g.,  veterans and access to specific-healthcare needs [@mooneyTravelDistanceBarrier2000], pregnant people and access to maternity services [@vadrevuMeasuringSpatialEquity2016], and youths who live in foster care [@batscheUsingGISEnhance2012].
-Overall, the diversity of population groups considered in the corpus demonstrates the variety of transportation-equity concerns addressed in the corpus.
+Papers that focus *solely* on "Employment" status or "Education" level are completely absent.
+Furthermore, "Other" population groups are also frequently considered in the corpus: this is a catch-all category that includes populations less commonly targeted by research e.g., veterans and access to specific-healthcare needs [@mooneyTravelDistanceBarrier2000], pregnant people and access to maternity services [@vadrevuMeasuringSpatialEquity2016], and youth who live in foster care [@batscheUsingGISEnhance2012].
+Overall, the diversity of population groups considered in the corpus demonstrates the variety of transportation-equity concerns addressed in the literature.
 
 ### Modes (mobility tools)
+
 Travel mode, though modifiable, is intertwined with individual identity, posing challenges for equity analysis.
-In this sub-section, we choose to view the mode of travel primarily as those who use the mode, hence the subject of justice (i.e., the “who”), but occasionally, this will be enjoined to an object of justice (i.e., a “what” such as the inequities in rural transit 'mobility' relative to cars [@giuffridaAddressingPublicTransport2021]).
-
-Public transit characterises the majority of the corpus (@fig-fig1).
-Despite transit being perceived unfavourably by some [@paezEnjoyment2010; @mellaDrivers2021], it is often viewed as the only and primary mobility option mode for many [@jacquesRescuing2012; @welchMeasure2013; @cherancheryInvestigatingPerceptionCaptive2018].
-The public transit focused papers assess a variety of topics such as: food desert identification [@mckeyCrowdsourcedMappingHealthy2020], barrier-free transportation for people who face disabilities [@jimenez-espadaResearchProblemUniversal2021; @liuStatusQuoChallenges2019] and people with autism [@limFacilitatingIndependentCommuting2021; @feeleyValidationParatransitSkills2019].
-We infer from the corpus that public transit is seen as a public good and hence a natural object of justice.
-Accordingly, public transportation is adaptable to meet the demands of justice, for example, by funding it sufficiently to provide barrier-free transport for most, despite challenges like low densities, fiscal constraints and political will [@markardUnsustainabilities2023].
-
-Transit also plays a central role in multimodal comparisons for transport equity analysis: these papers feature "Walk" or "Car" or some other category in @fig-fig1.
-For instance, @brusselAccessAccessibilityCritique2019 compares public transit, pedestrian and road network accessibility measurements in the context of SDG 11.2 in Bogota.
-@renneWhatHasAmerica2018 reviews natural disaster evacuation plans, focusing on car-less households in the USA and emphasizing transit and pedestrian networks.
-A few papers also frame transit as being in direct competition with car travel or use their comparison as a benchmark [@golubUsingPrinciplesJustice2014; @martensJusticetheoreticApproachDistribution2012].
-For instance, @warrenDevelopingEquitableSustainable2015 propose car ownership standards for Havana, Cuba, acknowledging the tension between mobility needs in transit under-served areas and emission reduction goals.
-However, this framing is not universal, transit is sometimes seen by the corpus as a mode to fulfill individual capabilities.
-As an example, @smithAccessibilityCapabilityMinimum2012 explores perspectives about transport needs and costs to achieve perceived sufficient living standards for those living in rural areas in the UK.
+Here we choose to view the mode of travel primarily as those who use the mode, hence as subjects of justice (i.e., the “who”), but occasionally, this will be enjoined to an object of justice (i.e., a “what” such as inequities in rural transit [@giuffridaAddressingPublicTransport2021]).
+
+Transit characterises the majority of the corpus (@fig-fig1).
+Despite transit being perceived unfavourably by some [e.g., @mellaDrivers2021], it is often viewed as the only and primary mobility option mode for many [@jacquesRescuing2012].
+Papers that focus on transit assess a variety of topics, including food deserts [@mckeyCrowdsourcedMappingHealthy2020] and barrier-free transportation for people who face disabilities [@jimenez-espadaResearchProblemUniversal2021; @liuStatusQuoChallenges2019; @limFacilitatingIndependentCommuting2021; @feeleyValidationParatransitSkills2019].
+We infer that transit is reckoned as a public good and hence a natural object of justice.
+Accordingly, public transportation can be modified to meet the demands of justice, by funding it sufficiently to provide barrier-free transport for most, despite challenges like low densities, fiscal constraints and political will [@markardUnsustainabilities2023].
+
+Transit also plays a central role in multimodal comparisons for transport equity analysis, and often feature in papers dealing with "Walk", "Car", or some other category in @fig-fig1.
+For instance, @brusselAccessAccessibilityCritique2019 compares transit, pedestrian and road network accessibility measurements in the context of SDG 11.2.
+@renneWhatHasAmerica2018 reviews natural disaster evacuation plans, focusing on car-less households and emphasizing transit and pedestrian networks.
+A few papers also frame transit as a direct competitor of car travel or use it as a benchmark [@golubUsingPrinciplesJustice2014; @martensJusticetheoreticApproachDistribution2012].
+For instance, @warrenDevelopingEquitableSustainable2015 propose car ownership standards while acknowledging the tension between mobility needs in transit under-served areas and emission reduction goals.
+However, this framing is not universal, and transit is sometimes seen as a mode to fulfill individual capabilities.
+As an example, @smithAccessibilityCapabilityMinimum2012 explores perspectives about transport needs and costs to achieve perceived sufficient living standards for those living in rural areas.
 Notably, papers vary in the importance they place on climate urgency, with some focusing more on satisfying *all* sufficient individual needs while planning for less car-dependent cities in the future.
 
-Following transit, pedestrian travel ("Walk") is the second most studied object of justice.
+After transit, pedestrian travel ("Walk") is the second most studied object of justice.
 Pedestrians represent a unique convergence of "what" and "who," utilizing their own bodies for mobility.
-Papers focusing solely on walking often use walkability scores to assess neighbourhood quality [@evansAccessibilityUserNeeds2015], pedestrian mobility in different demographics [@towneUsingWalkScore2016; @kimNeighborhoodEnvironmentWalkability2016], or urban peripheral regions [@blecicCapabilitywiseWalkabilityEvaluation2021], measuring equity distribution.
+Papers focusing solely on walking often use walkability scores to assess neighbourhood quality [@evansAccessibilityUserNeeds2015], mobility by different demographics [@towneUsingWalkScore2016; @kimNeighborhoodEnvironmentWalkability2016], or urban peripheral regions [@blecicCapabilitywiseWalkabilityEvaluation2021].
 Some use walkability to gauge public health and urban vitality [@sungResidentialBuiltEnvironment2015; @mccormackRelationshipClusteranalysisDerived2012].
-Furthermore, pedestrian-focused papers often discuss walking as a bridge to connect multiple modes as part of active transportation systems: linking walking, with cycling ("Bicycle"), and "Transit".
-Topics include how active transport impacts children's physical activity [@mammenSchoolTravelPlanning2014], walkability as a car alternative [@bertrandMeasuringMappingDisparities2008], and tensions between modes leading to unsafe walking conditions [@siuAssessmentPhysicalEnvironment2019; @ferenchakEquityAnalysisProactively2019].
-
 Papers with a pedestrian focus also often see walking as a bridge to connect multiple modes: they discuss 'walkability' as part of active transportation, which focuses on both walking, bicycle and/or transit.
@@
 
-The third most studied category is "Car".
-Unlike transit and walking, cars are rarely the sole focus.
-When studied in the corpus, they're critically compared with transit or considered in areas with inadequate transit [@kimmelStructuralBarriersComprehensive2018; @aljoufie2016].
-Upon the rise of automobility, walking and cycling were in the way; <!-- though being the way people got around historically [@robertsShort1998] and cycling's efficiency at a lower impact--> transit came to be seen as a social service and sometimes as a hindrance to the full realization of automobility.
-We surmise that the corpus' focus on non-car modes is a reckoning with automobility's legacy and ongoing requirements of disproportionate amounts of space, public subsidies and government support that underestimates its true cost [@gosslingSocial2019; @timperleyFight2021] including externalities like congestion, air pollution or safety [@zhengAreabasedEquitablePricing2020; @fengTradeoffsMobilityEquity2014; @houstonProximityLicensedChild2006].
-Other than "Car" being compared to walk, cycle or transit, some papers reflect the existing ways of travel and alongside "Car" including "Other" (@fig-fig1) options namely wheelchair accessible taxi and/or paratransit [@marquezFactorsAffectingPersonal2019; @wilkinsonmeyersLiveOrdinaryLife2015], travel on waterways [@vadrevuMeasuringSpatialEquity2016; @caoIndicatorsSocioSpatialTransport2017; @parrySocialVulnerabilityClimatic2018], motorcycle or other micro-mobility [@berryInvestigatingFuelPoverty2016; @savingmothersDidSavingMothers2019; @tiwariDevelopmentPublicTransport2021], or by emergency vehicle [@pedigoInvestigationDisparitiesGeographic2010; @patelDeterminingGeographicAreas2007].
+The third most studied mode is "Car".
+Unlike transit and walking, cars are rarely the sole focus of equity research.
+In our corpus they are often critically compared with transit or considered in areas with inadequate transit [@kimmelStructuralBarriersComprehensive2018; @aljoufie2016].
+Upon the rise of automobility, when walking and cycling got in the way of faster travel, transit came to be seen as a social service and sometimes as a hindrance to the full realization of automobility.
+Our assessement of the focus on non-car modes in our corpus is that it represents a reckoning with automobility's legacy and ongoing demands for space, public subsidies, and government supports that underestimate the true cost of cars [@gosslingSocial2019; @timperleyFight2021].
+In addition to comparing "Car" to other modes, some papers reflect on "Other" forms of mobility (@fig-fig1), including wheelchair accessible taxi and/or paratransit services [@marquezFactorsAffectingPersonal2019; @wilkinsonmeyersLiveOrdinaryLife2015], travel on waterways [@vadrevuMeasuringSpatialEquity2016; @caoIndicatorsSocioSpatialTransport2017; @parrySocialVulnerabilityClimatic2018], motorcycle or other micro-mobility [@berryInvestigatingFuelPoverty2016; @savingmothersDidSavingMothers2019; @tiwariDevelopmentPublicTransport2021], or emergency vehicles [@pedigoInvestigationDisparitiesGeographic2010; @patelDeterminingGeographicAreas2007].
 
@@
 
-<!-- 
-### Who's missing?
-As shown, "who" is the subject of inequities has been well studied.
-So what's missing from the "Who" question?
-We believe, more nuisances perspectives of inequities are needed.
-We must go beyond the focus on the low-income transit rider; positioning intersectionality of socio-economic and mode-use to create community-based definitions of equity and tailored standards are critical.
-
-As shown in the variety of methods, dimensions, conceptualisations and standards used transport inequalities are multi-dimensional.
+### Who is missing in the "who"?
+
+As shown, "who" is the subject of inequities provides ample but not exhaustive coverage of populations.
+But who is missing from the "Who" question?
+There is a need to go beyond the focus on low-income transit riders; positioning intersectionality of socio-economic and mode-use to create community-based definitions of equity and tailored standards are critical.
 How decision makers define an equity-deserving community will impact results e.g., [@rowangouldIdentifyingEnvironmentalJustice2015].
-In this sense, standards that are defined need to be sensitive to changing community-based definitions of inequities.
+In this sense, standards need to be sensitive to changing community-based definitions of inequities.
 Are issues of economic-inequity at the root of transport inequities for a specific community?
@@
 A community- informed understanding of inequities and tracking how it changes are needed.
-Different slices of analysis are important to different groups, as such, we believe all should be studied with an explicit justice framing.
--->
 
 ## "What" are the objects of transportation justice {#sect3.3}
-<!-- Though the literature reviewed most commonly considers low-income as a socio-demographic characteristics of populations deserving equity as either by itself or alongside other characteristics (`r prettyNum(round(df_pop$count[8]+df_pop$count[4] * 100, 1))`\% of the papers).
-Many employ accessibility methods (e.g., walking accessibility to open space [@tangWalkingAccessibilityNeighbourhood2021]) but others use qualitative or mixed-methods to gain perspectives of inequalities that are typically cross-dimensional (e.g., @milanLiftingPeripheralFortunes2017 asks Medellin, Colombia residents living in TOD areas compared to non-TOD areas about the impact of TOD on their wellbeing based on income group and gender).
-The use of qualitative methods allows for a more nuanced reflection of residents perspectives, and different inequities emerge.
--->
-From the corpus reviewed, we infer that the most fundamental benefits of transportation systems are mobility (enabling or impeding movement) and accessibility (the ease of reaching destinations).
-These benefits are sometimes valued by themselves but are often seen by the corpus as instrumental to achieve an ulterior goal (e.g., activity participation and associated wellbeing benefits).
-For example, although vehicle kilometers traveled (VKT) is sometimes seen as a useful policy instrument [@zhaoRethinking2021], travelling more is not necessarily a sign of advantage when accessibility is low [@paezAccessibilityHealthCare2010], and short trips may be a sign of advantage [@parkTransitParksEnvironmental2021].
+
+Our analysis confirms that the most fundamental benefits of transportation systems are mobility (enabling or impeding movement) and accessibility (the ease of reaching destinations).
+These benefits are sometimes valued by themselves but are often seen by the literature as instrumental to achieve an ulterior goal (e.g., activity participation and associated wellbeing benefits).
+For example, although vehicle kilometers traveled (VKT) is sometimes seen as a useful policy instrument [@zhaoRethinking2021], travelling more is not necessarily a sign of advantage when accessibility is low, and short trips may actually be a sign of advantage [@parkTransitParksEnvironmental2021].
 For this reason, although the right to the road (and transportation systems more generally) is important, the literature leans heavily on the ulterior object, namely accessibility to destinations.
 
-### Impartial to destination interaction
+### Mobility
+
 Most papers take a broad approach, with `r (dest_count$count[8]/165) |> percent()` focusing on "Unspecified" destinations (@fig-fig1).
@@
 
-### Partial to destination interaction
-Among studied destinations, "Healthcare" services (`r (dest_count$count[6]/165) |> percent()`) and "Employment" (`r (dest_count$count[5]/165) |> percent()`) receive the most attention (@fig-fig1).
+### Accessibility
+
+"Healthcare" services (`r (dest_count$count[6]/165) |> percent()`) and "Employment" (`r (dest_count$count[5]/165) |> percent()`) are the most common destinations in our corpus (@fig-fig1).
 Papers on healthcare often highlight disparities in services, like @wangModelingSpatialAccessibility2011, who assess access to Mandarin-speaking family physicians for Chinese speakers in Toronto.
@@
 
-Other activity types receive less focus but serve distinct purposes beyond typically studied travel patterns.
+Other destinations have received less attention despite serving essential needs.
 "Shopping" destinations (`r (dest_count$count[7]/165) |> percent()`) often aim to identify food deserts [@choiFoodDesertsActivity2013; @jiaoHowIdentifyFood2012; @mckeyCrowdsourcedMappingHealthy2020; @kimAssessingSocialSpatial2020].
 "Education"-related papers (`r (dest_count$count[3]/165) |> percent()`) explore children's active transportation to school [@larkinsAccessibleTransportationBuilt2011] or universal design [@larkinsAccessibleTransportationBuilt2011].
-Places of "Leisure" (`r (dest_count$count[6]/165) |> percent()`) prompt accessibility questions like spatial distribution [@xuSocialInequalitiesPark2017], accessibility [@mavoaAreaLevelDisparitiesPublic2015], and explores of inequity [@mearsUnderstandingSocioeconomicEquity2019].
-Fewer papers cover "Community" destinations (e.g., public service centres, places of community support or worship) (`r (dest_count$count[2]/165) |> percent()`) or "Childcare" (`r (dest_count$count[1]/165) |> percent()`), but they are integral in holistically studying activity participation [@albertsRebuildingWomenLivelihoods2016; @smithAccessibilityCapabilityMinimum2012].
-The lack of information about community destinations, especially for children, is notable in the literature broadly [@desjardinsChildren2022].
-
-<!-- ### Dimensions of measurement -->
-
-<!-- Defining the object of justice (the "What"), whether specific destinations or the trip itself, shapes measurements of inequality crucial for tracking progress towards transport justice.
-In the reviewed corpus, we characterize these dimensions as a focus on:  -->
-
-<!-- - accessibility and mobility;  -->
-<!-- - transport-related environmental externalities;  -->
-<!-- - centering human-health as related to transport systems;  -->
-<!-- - human safety related to vehicular-traffic; and  -->
-<!-- - cross-dimensional classifications. -->
-
-<!-- As discussed next, these intersect with conceptualizations of fairness and applied standards, forming the philosophical foundations for assessing justice in transportation. -->
-
-<!-- ### What's missing?  -->
-
-<!-- What's missing from the "What?" in the equity analyses of the literature reviewed is the consistently explicit rationale for the object of justice.
-Specifically, "What?" are the benefits and burdens of a transportation system in terms of its: accessibility and mobility, or environmental externalities, or human-health externalities, or traffic related safety, or multi-criteria mix.
-In the inter-disciplinary analysis of transport systems, crystalising "What?" is the object of justice can create a stronger foundation to select appropriate methods to determine fairness and support the "Why?", the rationale of justice.
-In the corpus reviewed, strong conceptualizations of the object of justice often falls short.
--->
+Places of "Leisure" (`r (dest_count$count[6]/165) |> percent()`) prompt questions about the spatial distribution of services [@xuSocialInequalitiesPark2017] and accessibility [@mavoaAreaLevelDisparitiesPublic2015].
+Fewer papers cover "Community" destinations (e.g., public service centres, places of community support or worship) (`r (dest_count$count[2]/165) |> percent()`) or "Childcare" (`r (dest_count$count[1]/165) |> percent()`), but they are integral to holistically study activity participation [@albertsRebuildingWomenLivelihoods2016; @smithAccessibilityCapabilityMinimum2012].
+The lack of information about community destinations, especially for children, is noticeable [@desjardinsChildren2022].
+
+### What is missing from the "what"?
+
+Missing from the "What?" are explicit rationales for the object of justice.
+Specifically, "What?" are the benefits and burdens of a transportation system in terms of its accessibility and mobility, environmental or health-related externalities, or traffic safety.
+In the analysis of transport systems, clearly stating "What?" is the object of justice can create a stronger foundation to select appropriate methods to determine fairness and support the "Why?", the rationale of justice.
+In the corpus reviewed, discussions of the object of justice often fall short of achieving this.
 
@@
 ### Conceptualisations of fairness
+
 The philosophical foundations that are explicit or most often implicitly inferred from the corpus are broadly characterized as follows (@fig-fig2 details definitions).
@@
 Notably, many papers suggesting "Opportunity" standards include cross-dimensional considerations, employing similar methods but tailored to different focal points.
-For instance @peungnumsaiGridBasedSpatialAnalysis2020 suggest service benchmarks of equal supply and demand of public transit, revealing "Horizontal equities" as well.
+For instance @peungnumsaiGridBasedSpatialAnalysis2020 suggest service benchmarks of equal supply and demand of transit, revealing "Horizontal equities" as well.
 Others conceive the externalities of transportation system as trade-offs and aim to maximize transport-related benefits (i.e., time savings, emissions reductions, congestion reductions, user fares) through optimization/location-allocation methodologies [@fakhrmoosaviIncorporatingTravelTime2021; @zhengAreabasedEquitablePricing2020; @wismadiSpatialPreferenceModelling2014].
@@
 
-Justice is a goal of social progress.
-To understand what that goal is, clearly defining standards of fairness is essential.
-Measuring context-specific evidence of transport inequities is necessary to build knowledge of injustices and inspire creative interventions that advance the cause of justice.
-Understanding and tracking transport inequities over time and space, specifically knowledge of "Where" (spatial context), "When" (temporal context), "Who" (the subject of justice, i.e., population groups and mobility tools they use), "What" (the object of justice, i.e., what benefit/burden is produced by the system and its distribution), "How" (the conceptualisation and standards used) and especially "Why" (the rationale and not included in the focus of this review) is essential.
-Armed with this knowledge, policy can be designed to address these inequities while remaining adaptable to evolving justice demands.
-
-As transport is a derived demand, flexible frameworks are crucial.
+This paper makes two contributions to the literature.
+It outlines a conceptual and flexible "Where", "When", "Who", "What" and "How" framework with supporting definitions for approaching questions about transportation justice.
+It also applies the framework in the synthesis and appraisal of 165 academic articles drawn from the transportation justice literature; these studies use equity analysis as instruments to gauge proximity to their context's just situation.
+
+In brief, justice is an end objective and standards are definitive statements of fairness.
+As such, they are essential to assess proximity to the objective.
+Context-specific evidence of transport inequities is necessary to acknowledge injustices and inspire creative interventions to advance the cause of justice.
+Armed with appropriate knowledge, policy can be designed to address inequities while remaining adaptable to the evolving demands of justice.
+Flexible frameworks are crucial.
 What is (in)formally considered (in)essential today, could change during the next pandemic or technological revolution.
-Creating adaptable transport equity frameworks that lend themselves to change with the demands of justice is required to ensure resiliency to backsliding.
-From the diverse corpus reviewed, we argue for the necessity of setting standards from a systems approach, considering both _positive_ rights (a right to have access to sufficient quality of essential services, for example) and _negative_ rights (mobility of cars must be limited to reduce their impact on air quality, health, safety, and positive rights).
-
-In this spirit, this work makes two significant contributions.
-It outlines a conceptual and flexible "Where", "When", "Who", "What" and "How" framework with supporting definitions for structuring transportation fairness and equity analysis.
-It also applies the framework in the synthesis and appraisal of 165 academic articles that undertake standard- and conceptualisation- supported equity analysis; these studies use equity analysis as instruments to gauge proximity to their context's just situation.
-Following this framework and the reviewed literature it answers:
-
--	**What** are those benefits and burdens of transportation systems? 
--	**Who** benefits and is burdened by transportation systems? Along which lines of identity are transportation inequities examined?
--	**Where and when** are equity standards applied?
--	**How** are equity standards and inequities measured?
-
-<!--Maybe we should answer these questions explicitly here? Maybe we should also urge practitioners to adopt this flexible framework in their equity synthesis?-->
-From the outlined framework and the reviewed corpus, we recommend the following five calls to action.
+Creating adaptable transport justice frameworks that lend themselves to change is required to ensure resiliency to backsliding.
+From the diverse corpus reviewed in this paper, we argue for the necessity of setting standards from a systems approach, considering both _positive_ rights (e.g., a right to have access to sufficient quality of essential services) and _negative_ rights (e.g., mobility of cars must be limited to reduce their impact on air quality, health, safety, and other positive rights).
+
+We conclude with the following five calls to action.
 
@@
 
-<!-- # Final remarks {#sect5} -->
-
-<!-- To move the political needle and establish a common understanding of justice, it is critical that we have the concepts, the vocabulary, and the knowledge to express our intuitions.
-We argue that transportation, as an essential technology that can facilitate or hinder participation in society, is a very relevant sphere for matters of justice.
-The analysis of fairness and equity in transportation is vitally important for planning.
-Transportation systems and technologies converge space-time, and by definition they _always_ leave certain populations, depending on the area and time scale, more or less burdened or benefited.
-The analysis and tracking of equity implications is fundamental in formulating standards that are flexible to the fluid demands of transport justice.
--->
-
-<!-- In this spirit, this work makes two significant contributions.
-It outlines a conceptual framework with supporting definitions for structuring fairness and equity analysis that may move us closer to transport justice.
-Then using this framework, we critically appraises findings from 165 relevant studies from the academic literature and recommend five calls to action.
--->
-
-<!-- Defining _fairness_ is complex, but complexity is no excuse to shirk from the task.
-We hope this work contributes to the promotion of the idea that all are deserving of "access to the opportunities they need to lead a meaningful and dignified life" [@karnerTransportationEquityTransportation2020, p.
-440].
-This idea should guide our justice conceptualisation.
-Likewise, we hope that in reviewing the literature, this work will contribute to advance the development and implementation of equity standards that help bend the arc of history towards more just transportation systems, and ultimately better societies.
--->
-
-<!-- `r word_comment(" - A CONSIDERATION FOR THE FUTURE - What do you think - do we provide a 'framework' for practioners/researchers?
-We perhaps provide a theoretical framework to think through these questions that they could follow + definitions.
-However, in our work we use who/what/where/when/how to organise our review.
-As it stands, we don't demonstrate this frameing could be used to organize a study/research project - like a worked example of how 1 specific paper fits into each category.
-This could be solved with a sub-section that demonstrates how one paper fits into each category (I know we have a few examples in the table in the appendix but there's no room in the paper and explain it).", ".")` -->
-
-<!-- The transportation justice framework can be seen as a flexible classification exercise that is supported by a "Why" rationale.
-In conducting an equity analysis, the subject of justice (the "Who?") must contextualize the process, the distribution of the object of justice (the "What?") must be measured, be it transport-related benefits, burdens or both, "How?" these inequities are measured and benchmarked must be described, and these questions must be contextualized by the "Where?" across temporally changing "When?"s.
-The explicit acknowledge and answers to these questions clarify the boundaries of the equity analysis.
-In this way, the rationale for "Why?" the equity analysis is important and what root cause of/for (in)justice is motivating the analysis in the first place may be critical to ultimately mobilizing equity analyses into experienced transport justice.
--->
-
-<!-- From this framework and the critical appraisal of the literature, we outline five call to actions for researchers and decision-makers that should be adopted to move equity analysis closer to transport justice.
-These include: -->
-
-<!-- - Explicitly explain "Why?", the rationale for justice, along with supporting "Where", "When", "Who", "What, and "How" framings for equity analysis.
-This means specific conceptualizations and grounded standards.
--->
-<!-- - Creative methods of equity appraisal that can be paired with systems thinking; current methods are not used to support this processes. -->
-<!-- - Data availability as a matter of justice: who is not being counted and who should be by whom and why?  -->
-<!-- - More explicit links between equity standards and experienced conditions and outcomes.  -->
-<!-- - Encouragement to continue work towards analysis and comparison of applied interventions, across communities and internationally.  -->
-
-**Declarations:**
-
-During the preparation of this work, the first and second authors used ChatGPT exclusively for copyediting in the "An appraisal of the lay of the land" section, aiming to reduce word count while keeping the original writing style and correct typos.
+# Declarations
+
+During the preparation of this work, the first and second authors used ChatGPT exclusively for copy-editing in the section "An appraisal of the lay of the land", aiming to reduce word count while keeping the original writing style and to correct typos.
 After using this tool, all authors reviewed the manuscript and take full responsibility for the content of the publication.
@@
 This paper conceptualizes accessibility as from the position of the **capabilities approach** and **vertical equity** (particularly acknowledging that older people have capabilities that differ from the general population).",
-                           "How?: Specifically for older populations (aged 65+), the following travel distances are suggested as equitable trip lengths to grocery stores per mode: Walking: less than or equal to 1500m, Combined public transit and walking (less than or equal to 1000m (walking element)), Combined car and walking: less than or equal to 1000m less than or equal to 1000m (walking element)), Bicycle: less than or equal to 3000m in addition to travel time threshold of less than 15 mins.",
+                           "How?: Specifically for older populations (aged 65+), the following travel distances are suggested as equitable trip lengths to grocery stores per mode: Walking: less than or equal to 1500m, Combined transit and walking (less than or equal to 1000m (walking element)), Combined car and walking: less than or equal to 1000m less than or equal to 1000m (walking element)), Bicycle: less than or equal to 3000m in addition to travel time threshold of less than 15 mins.",
                            "What?: **mobility and accessibilty**",
@@
 Indicators are inspired by the Sustainable Development Goals (SDGs) as well other global planning standards.",
-                           "How?: 1) Green space: % of residents living < 400 m from public open space, a large park (> 1.5ha), and/or local park, 2) Public transit access: % of residents living < 400 m of a local bus stop and <800 m of train station, 3) Facilities: % of residents living < 400 m of a community centre.
+                           "How?: 1) Green space: % of residents living < 400 m from public open space, a large park (> 1.5ha), and/or local park, 2) transit access: % of residents living < 400 m of a local bus stop and <800 m of train station, 3) Facilities: % of residents living < 400 m of a community centre.
 The following **Infrastructure standard** is suggested: Canal water quality - dissolved oxygen content of equal to or less than 2.0 mL/L",
@@
 is the goal.
-2) As related to sidewalk infrastructure at least 50% of all sidewalks (preferably 75% or greater) should: have a width of 1.5m or larger, ramps should have a grade of 8% or less, be well maintained (free from deficiencies), be paved for pedestrian use, and cover public transit stops.
+2) As related to sidewalk infrastructure at least 50% of all sidewalks (preferably 75% or greater) should: have a width of 1.5m or larger, ramps should have a grade of 8% or less, be well maintained (free from deficiencies), be paved for pedestrian use, and cover transit stops.
 3) Lighting is critical for traffic-safety and a sense of safety overall.

### 2024-05-15 README.Rmd
README includes link to pdf version of report
Antonio's initial edits before submission

@@
 )
+library(rticles)
 ```
@@
 
-```{r cover, echo=FALSE}
-#| label: cover
-knitr::include_graphics("just-transportation_1.png")
-```
+![The pdf version of this report is available [here](https://github.com/soukhova/Transportation-lit-review/blob/main/just-transportation/just-transportation.pdf)](just-transportation_1.png)
 

### 2024-05-15 Journal-manuscript/Journal-manuscript.qmd
Antonio's edits before submission

@@
 bibliography: bibliography.bib 
+filters:
+  - docx-landscape.lua
 ---
@@
 library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
+library(gt) # Easily Create Presentation-Ready Display Tables
 library(tidyverse) # Easily Install and Load the 'Tidyverse'
@@
 ```
+
 ```{r calculating-destinations-of-interest}
@@
 In this spirit, our work seeks to make two contributions.
-First, it outlines a conceptual and flexible framework for engaging questions of transportation fairness and equity analysis.
+First, it outlines a conceptual and flexible framework for engaging questions of transportation fairness and equity analysis, based on the questions "Why?", "Where?", "When?", "Who?", "What?", and "How?" (5WH).
 Second, it applies the framework to collate the existing knowledge about equity standards.
@@
 
-## A framework to analyze questions of justice {#sect2.2}
+## A framework to analyze questions of justice: 5WH {#sect2.2}
 
@@
 This is done either explicitly, as for example in @karnerTransportationEquityTransportation2020, who ask "of what", "for whom," and "how much" in reference to equity; or implicitly, as in @gosslingUrban2016, who asks of the outputs ("what?") of transportation (exposure, space, access) and "for whom?" (gender, age, ethnicity).].
-According to @jaggarPhilosophicalChallengesGlobal2009, Western philosophy has approached the issue of justice by asking "Where?", "When?", "Who?", "What?", and "How?", applying them to a particular domain or sphere of life relevant to justice.
+According to @jaggarPhilosophicalChallengesGlobal2009, Western philosophy has approached the issue of justice by asking "Why?", "Where?", "When?", "Who?", "What?", and "How?" (5WH), applying them to a particular domain or sphere of life relevant to justice.
 
@@
 
-Lastly, convincing answers to the above questions require a supporting rationale: a **"Why?"** [@jaggarPhilosophicalChallengesGlobal2009].
-This is perhaps the most slippery of all the questions posed here.
+Above all, convincing answers to the above questions require a supporting rationale: a **"Why?"** [@jaggarPhilosophicalChallengesGlobal2009].
+This is perhaps the most slippery of the questions posed here.
 Justice is inherently a social construct.
@@
 
-Our analysis confirms that the most fundamental benefits of transportation systems are mobility (enabling or impeding movement) and accessibility (the ease of reaching destinations).
-These benefits are sometimes valued by themselves but are often seen by the literature as instrumental to achieve an ulterior goal (e.g., activity participation and associated wellbeing benefits).
+The most fundamental benefit of transportation systems is mobility (enabling or impeding movement).
+This benefit is sometimes valued by itself but is often seen by the literature as instrumental to achieve an ulterior goal (e.g., activity participation and associated benefits).
 For example, although vehicle kilometers traveled (VKT) is sometimes seen as a useful policy instrument [@zhaoRethinking2021], travelling more is not necessarily a sign of advantage when accessibility is low, and short trips may actually be a sign of advantage [@parkTransitParksEnvironmental2021].
-For this reason, although the right to the road (and transportation systems more generally) is important, the literature leans heavily on the ulterior object, namely accessibility to destinations.
+For this reason, although the right to the road (and transportation systems more generally) is important, the literature leans heavily on the ulterior object, namely accessibility to (the ease of reaching) destinations.
 
@@
 
-Most papers take a broad approach, with `r (dest_count$count[8]/165) |> percent()` focusing on "Unspecified" destinations (@fig-fig1).
+Most papers in our corpus take a broad approach, with `r (dest_count$count[8]/165) |> percent()` focusing on "Unspecified" destinations (@fig-fig1).
 They examine various equity dimensions across different transportation modes.
@@
 
-"Healthcare" services (`r (dest_count$count[6]/165) |> percent()`) and "Employment" (`r (dest_count$count[5]/165) |> percent()`) are the most common destinations in our corpus (@fig-fig1).
+In terms of destinations, "Healthcare" services (`r (dest_count$count[6]/165) |> percent()`) and "Employment" (`r (dest_count$count[5]/165) |> percent()`) are the most common in our corpus (@fig-fig1).
 Papers on healthcare often highlight disparities in services, like @wangModelingSpatialAccessibility2011, who assess access to Mandarin-speaking family physicians for Chinese speakers in Toronto.
 Similarly, papers focusing on employment are often aimed at identifying transportation-poor neighbourhoods [@allenSizingTransportPoverty2019a; @churchillTransportPovertySubjective2019].
-Employment is frequently used as a proxy for overall accessibility since it is the most common trip purpose and employment is usually collocated with other destinations like shops, recreation, and other services.
-These studies typically use travel surveys, census data and point-of-interest databases, and benefit from well-developed and institutional data.
-This especially holds in the Global North where this data is more readily available.
+Employment is frequently used as a proxy for overall accessibility since it is the most common trip purpose and employment is usually co-located with destinations like shops, recreation, and other services.
+These studies typically use travel surveys, census data, and point-of-interest databases, and benefit from well-developed institutional data.
+This especially holds in the Global North where these data are more readily available.
 
@@
 "Education"-related papers (`r (dest_count$count[3]/165) |> percent()`) explore children's active transportation to school [@larkinsAccessibleTransportationBuilt2011] or universal design [@larkinsAccessibleTransportationBuilt2011].
-Places of "Leisure" (`r (dest_count$count[6]/165) |> percent()`) prompt questions about the spatial distribution of services [@xuSocialInequalitiesPark2017] and accessibility [@mavoaAreaLevelDisparitiesPublic2015].
+Places of "Leisure" (`r (dest_count$count[6]/165) |> percent()`) prompt questions about their spatial distribution [@xuSocialInequalitiesPark2017] and accessibility [@mavoaAreaLevelDisparitiesPublic2015].
 Fewer papers cover "Community" destinations (e.g., public service centres, places of community support or worship) (`r (dest_count$count[2]/165) |> percent()`) or "Childcare" (`r (dest_count$count[1]/165) |> percent()`), but they are integral to holistically study activity participation [@albertsRebuildingWomenLivelihoods2016; @smithAccessibilityCapabilityMinimum2012].
@@
 
-### What is missing from the "what"?
-
-Missing from the "What?" are explicit rationales for the object of justice.
-Specifically, "What?" are the benefits and burdens of a transportation system in terms of its accessibility and mobility, environmental or health-related externalities, or traffic safety.
-In the analysis of transport systems, clearly stating "What?" is the object of justice can create a stronger foundation to select appropriate methods to determine fairness and support the "Why?", the rationale of justice.
-In the corpus reviewed, discussions of the object of justice often fall short of achieving this.
-
 ## "How" is fairness determined {#sect3.4}
@@
 
-The philosophical foundations that are explicit or most often implicitly inferred from the corpus are broadly characterized as follows (@fig-fig2 details definitions).
-These categories emerged as we reviewed the corpus and are used to discuss trends, but are not an exhaustive list of conceptualizations.
+The conceptual foundations of fairness are not always made explicit by researchers and often had to be inferred from the corpus.
+Some categories emerged from our review and we must note that they are not exhaustive.
+In broad terms, they are (@fig-fig2 details definitions):
 
 - Vertical equity (`r df_conc|> group_by(conc) |> summarise(total = n()) |> subset(conc == "Vertical equity", select = "total") |> mutate(total = total/165) |> pull(total)  |> percent()`)
+
 - Transportation related social exclusion (`r df_conc|> group_by(conc) |> summarise(total = n()) |> subset(conc == "TR exclusion", select = "total") |> mutate(total = total/165) |> pull(total)  |> percent()`)
+
 - Well-being (`r df_conc|> group_by(conc) |> summarise(total = n()) |> subset(conc == "Wellbeing", select = "total") |> mutate(total = total/165) |> pull(total)  |> percent()`)
+
 - Spatial equity (`r df_conc|> group_by(conc) |> summarise(total = n()) |> subset(conc == "Spatial equity", select = "total") |> mutate(total = total/165) |> pull(total)  |> percent()`)
+
 - Horizontal equity (`r df_conc|> group_by(conc) |> summarise(total = n()) |> subset(conc == "Horizontal equity", select = "total") |> mutate(total = total/165) |> pull(total)  |> percent()`)
+
 - Inequitable externalities (`r df_conc|> group_by(conc) |> summarise(total = n()) |> subset(conc == "Ineq.
 externalities", select = "total") |> mutate(total = total/165) |> pull(total)  |> percent()`)
+
 - Rights (`r df_conc|> group_by(conc) |> summarise(total = n()) |> subset(conc == "Rights", select = "total") |> mutate(total = total/165) |> pull(total)  |> percent()`)
+
 - Emerging theories (`r df_conc|> group_by(conc) |> summarise(total = n()) |> subset(conc == "Emerg. theories", select = "total") |> mutate(total = total/165) |> pull(total)  |> percent()`)
+
 - Utilitarian (`r df_conc|> group_by(conc) |> summarise(total = n()) |> subset(conc == "Utilitarian", select = "total") |> mutate(total = total/165) |> pull(total)  |> percent()`)
 
-A few segments of the literature and how they are characterized by their foundational conceptualisations of fairness are summarized.
-For example, a slice of the corpus is supported by broader "Rights" conceptualisations: these papers often focus on equity for people with disabilities or non-car users and associated challenges accessing transport infrastructure [@jimenez-espadaResearchProblemUniversal2021; @daamenAssessingGapPublic2008; @bharathyRevisitingClearFloor2018].
+A slice of the corpus is supported by broader "Rights" conceptualisations of fairness: these papers often focus on equity for people with disabilities or non-car users and associated challenges accessing transport infrastructure [@jimenez-espadaResearchProblemUniversal2021; @daamenAssessingGapPublic2008; @bharathyRevisitingClearFloor2018].
 While many papers are underpinned with the right to the city (the _right_ to participate in the production of urban space [@lefebvreDroitVille1967]) [@adliRightCityApplying2018], others emphasize legal _Rights_ like ADA regulations [@bharathyRevisitingClearFloor2018] or the goal of _access for all_ in land-use transportation master plans [@limFacilitatingIndependentCommuting2021].
@@
 These papers may also address traffic-related air and noise pollution or urban temperatures.
-In these papers, equity is theoretically achieved if similar levels are attainable for all populations (horizontal equity) or spatial areas (spatial equity), discussions rarely explore both minimum and maximum levels contributing to inequities from a justice perspective.
-
-Papers that center "Well-being" assess what constitutes a satisfactory life in relation to transportation; this set of literature uses mixed-methods and, in contrast to the last set often explicitly identifies objects of injustice.
-Some use physical activity guidelines and surveys to understand activity related to active transportation infrastructure [@adlakhaMindGapGender2020; @auchinclossDesignBaselineDescription2020; @mccormackRelationshipClusteranalysisDerived2012].
-Mixed or qualitative methods combined with health-related outcome standards (e.g., travel times for emergency treatment [@savingmothersDidSavingMothers2019], premature mortality [@awuorInfluenceEnvironmentalHealth2019]) leads to firmer justice identifications.
+In these papers, equity is theoretically achieved if similar levels are attained for all populations (horizontal equity) or spatial areas (spatial equity).
+This egalitarian perspective rarely delves into minimum or maximum levels associated with harm or need satisfaction.
+
+Papers that center "Well-being" assess what constitutes a satisfactory life in relation to transportation; mixed-methods are often used and, in contrast to the last set, the objects of injustice are often identified.
+Some papers use physical activity guidelines and surveys to understand the effect of active transportation infrastructure [@adlakhaMindGapGender2020; @auchinclossDesignBaselineDescription2020; @mccormackRelationshipClusteranalysisDerived2012].
+Mixed or qualitative methods combined with health-related outcome standards (e.g., travel times for emergency treatment [@savingmothersDidSavingMothers2019], premature mortality [@awuorInfluenceEnvironmentalHealth2019]) lead to more concrete statements of fairness.
 
 Another research branch, often quantitative with some qualitative or mixed-methods studies, focuses on "Transport-related social exclusion", "Vertical equity", and/or "Sufficientarian/capabilities".
-They also often explicitly identify objects of justice and link standards to tangible and welfare-informed outcomes.
+Here, the objects of justice are often clearly identified and link standards to tangible welfare-informed outcomes.
 They focus on groups from perspectives of disadvantage such as: social exclusion and transport poverty [@allenSizingTransportPoverty2019a; @delboscTransportProblemsThat2011; @churchillTransportPovertySubjective2019]], food deserts [@mckeyCrowdsourcedMappingHealthy2020], and energy poverty [@robinsonDoubleEnergyVulnerability2020; @berryInvestigatingFuelPoverty2016; @berryDistributionalEffectsCarbon2019].
@@
 ### Standards and methods of measuring fairness
-With the ways that conceptualisations are frequently applied within the corpus detailed, the following are categories of standards that overlap with at least one conceptualisation (see @fig-fig2 for definitions):
+
+We identify several categories of standards that overlap with at least one of the conceptualisations of fairness discussed above (see @fig-fig2 for definitions):
 
 - Opportunity standards (`r df_standard_mcat |> group_by(std_mcat) |> summarise(total = n()) |> subset(std_mcat == "Opportunity", select = "total") |> mutate(total = total/165) |> pull(total)  |> percent()`)
+
 - Population standards  (`r df_standard_mcat |> group_by(std_mcat) |> summarise(total = n()) |> subset(std_mcat == "Population", select = "total") |> mutate(total = total/165) |> pull(total)  |> percent()`)
+
 - Infrastructure standards (`r df_standard_mcat |> group_by(std_mcat) |> summarise(total = n()) |> subset(std_mcat == "Infrastructure", select = "total") |> mutate(total = total/165) |> pull(total)  |> percent()`)
+
 - Environment+ standards (`r df_standard_mcat |> group_by(std_mcat) |> summarise(total = n()) |> subset(std_mcat == "Environmental +", select = "total") |> mutate(total = total/165) |> pull(total)  |> percent()`)
 
-<!-- Opportunity standards often address horizontal and spatial equity, focusing on travel impedance metrics like speed, distance, or cost.
-For instance, @chenImpactHighspeedRail2017 set a 4-hour high-speed rail travel limit for municipalities to be considered "comfortably connected." Similarly, @yenisettyMeasuringAccessibilityVarious2020 deem a travel distance of less than 1,200 meters sufficient for local transit access.
-@shenMeasuringSpatialAccessibility2020 highlight regions where hospital access exceeds a 1-hour car journey, indicating inequity.
-These standards serve as benchmarks to assess disparities in travel impedance, acting as difference thresholds.
--->
-Papers suggesting "Opportunity" standards often employ quantitative methods to analyze disparities and assess distributional fairness.
-Many stem from travel impedance thresholds based on speed, distance, or cost [@chenImpactHighspeedRail2017; @yenisettyMeasuringAccessibilityVarious2020; @shenMeasuringSpatialAccessibility2020].
+Papers with "Opportunity" standards often employ quantitative methods to analyze disparities and assess distributional fairness.
+Many deal with travel impedance thresholds based on speed, distance, or cost [@chenImpactHighspeedRail2017; @yenisettyMeasuringAccessibilityVarious2020; @shenMeasuringSpatialAccessibility2020].
 Inequality measures like the Gini coefficient and poverty measures are used to empirically define travel impedance thresholds [@vanderveenOperationalizingIndicatorSufficient2020; @tiznadoaitkenRoleAccessibilityPublic2018].
-Further, methods tangential to travel impedance, like limiting transport expenditure to 10% of monthly income [@rivasHowAffordableTransportation2018], or addressing spatial mismatch [@mulleyAreNetworkPlanning2015], pinpointing areas of relative regional inequities are also used.
-Notably, many papers suggesting "Opportunity" standards include cross-dimensional considerations, employing similar methods but tailored to different focal points.
+Further, methods tangential to travel impedance, like limiting transport expenditure to 10% of monthly income [@rivasHowAffordableTransportation2018], addressing spatial mismatch [@mulleyAreNetworkPlanning2015], or pinpointing areas of relative regional inequities are also used.
+Notably, many papers with "Opportunity" standards consider multiple dimensions, employing similar methods but tailored to different focal points.
 For instance @peungnumsaiGridBasedSpatialAnalysis2020 suggest service benchmarks of equal supply and demand of transit, revealing "Horizontal equities" as well.
@@
 
-"Population" standards are often founded on "Well-being" conceptualisations, assessing what constitutes a satisfactory life related to transportation from a variety of socio-demographic and spatial angles.
+"Population" standards are often founded on "Well-being" conceptualisations from a variety of socio-demographic and spatial angles.
 Methods include: establishing thresholds based on questionnaires and comparisons to recommended physical activity levels [@auchinclossDesignBaselineDescription2020; @mccormackRelationshipClusteranalysisDerived2012; @kimNeighborhoodEnvironmentWalkability2016; @timperioSafetyNumbersDoes2015], region-relative comparisons in health outcomes in a spatial unit such as premature mortality rates [@awuorInfluenceEnvironmentalHealth2019], spatial access benchmarks based on population-related characteristics like supermarket access [@murphySupermarketAccessTransport2017] and hospital access @pereiraGeographicAccessCOVID192021], summative per capita benchmarks such as decent living energy consumption levels [@raoDecentLivingEmissions2012], and community-informed spatial boundaries like EJ defined communities [@rowangouldIdentifyingEnvironmentalJustice2016].
-While most of these papers use quantitative or mixed-methods approaches, some employ exclusively qualitative methods to explore themes [@berheAdaptationDissonanceQuality2014].
+While most of these papers use quantitative or mixed-methods approaches, some employ exclusively qualitative methods [@berheAdaptationDissonanceQuality2014].
 
@@
 Some exclusively use qualitative methods to analyze survey data on travel willingness/barriers or conduct interviews on unmet activity needs [@chenExploringTravelCharacteristics2010; @mehdizadehWalkingTimeSchool2017; @nordbakkeTransportUnmetActivity2015].
-<!-- Understanding the reasons for travel that did not occur, whether due to preference or constraints, remains challenging (A.
-Paez and Farber, 2012).-->
-
-"Infrastructure" standards offer another perspective on fairness, commonly grounded in "Rights" conceptualisations.
-Discussions of rights are twice as frequent in this segment of the corpus compared to other conceptualisations.
-These papers most frequently address the rights of non-car users and populations with mobility disabilities.
+
+"Infrastructure" standards offer another perspective on fairness, commonly grounded in "Rights" conceptualisations, which are twice as frequent in this segment of the corpus compared to other conceptualisations.
+These papers most frequently address the rights of non-car users and populations with disabilities.
 Though various methods are applied, infrastructure and environmental audits as well as qualitative approaches are most prominent.
@@
 
-Additionally, "Infrastructure" standards papers sometimes encompass multiple dimensions, moving beyond rights (to the infrastructure) and suggest "Opportunity" and/or "Population" standards.
+Additionally, "Infrastructure" standards papers sometimes encompass multiple dimensions, moving beyond rights (to the infrastructure) to provide "Opportunity" and/or "Population" standards.
 These papers often employ "Vertical", "Horizontal", and "Spatial equity" lenses.
-These multi-dimensional papers often refer to guidelines and propose composite indices.
+These papers often refer to guidelines and propose composite indices.
 For example, [@racheleNeighbourhoodSocioeconomicTransport2017a] integrate various transport network properties to define an indicator supporting walkability and public transport access.
@@
 
-Our corpus minimally focuses on "Environmental+" standards.
-These studies predominantly examine "inequitable externalities", possibly because environmental burdens of transportation are addressed more broadly in other literature (e.g., environmental justice).
-However, these papers included in our corpus present some interesting insights.
+"Environmental+" standards feature less prominently in our corpus, with studies that predominantly examine "inequitable externalities", possibly because the environmental burdens of transportation are addressed more broadly in other literature (e.g., environmental justice).
+However, the papers included in our corpus present some interesting insights.
 They often use traffic-related air and noise pollution, green space, urban design, urban air temperature, health outcomes, and physical activity guidelines to assess transport-related externalities.
-Methods employed are primarily quantitative or mixed-methods, identifying inequalities through Gini coefficients [@fengTradeoffsMobilityEquity2014] and some develop composite indices [@agostfelipInclusiveModelAssessing2021; @mirandaBenchmarkingSustainableUrban2012; @corazzaMethodologyEvidenceCase2020], occasionally incorporating spatial analysis [@jephcoteGeospatialAnalysisNaturally2013; @carrierApplicationThreeMethods2014].
-Commonly, many use established thresholds or health guidelines e.g., WHO guidelines for Active Aging and targets included in the United Nations’ SDG [@agostfelipInclusiveModelAssessing2021] and OECD or WHO standards for traffic noise level, $NO_2$ levels, available green space and $PM_{2.5}$ levels [@kruizeEnvironmentalEquityRole2007; @iungmanImpactUrbanTransport2021; @apparicioCyclingOneMost2021; @khomenkoLiveableCityHealthy2020; @muellerSocioeconomicInequalitiesUrban2018].
+Methods employed are primarily quantitative or mixed-methods, identifying inequalities through Gini coefficients [@fengTradeoffsMobilityEquity2014] or composite indices [@agostfelipInclusiveModelAssessing2021; @mirandaBenchmarkingSustainableUrban2012; @corazzaMethodologyEvidenceCase2020], occasionally incorporating spatial analysis [@jephcoteGeospatialAnalysisNaturally2013; @carrierApplicationThreeMethods2014].
+Many use established thresholds or health guidelines e.g., WHO guidelines for Active Aging and targets included in the United Nations’ SDG [@agostfelipInclusiveModelAssessing2021] and OECD or WHO standards for traffic noise level, $NO_2$ levels, available green space and $PM_{2.5}$ levels [@kruizeEnvironmentalEquityRole2007; @iungmanImpactUrbanTransport2021; @apparicioCyclingOneMost2021; @khomenkoLiveableCityHealthy2020; @muellerSocioeconomicInequalitiesUrban2018].
 These metrics are sometimes criticized for their general applicability.
@@
 This paper makes two contributions to the literature.
-It outlines a conceptual and flexible "Where", "When", "Who", "What" and "How" framework with supporting definitions for approaching questions about transportation justice.
+It outlines a conceptual and flexible "Why", Where", "When", "Who", "What" and "How" (5WH) framework with supporting definitions for approaching questions about transportation justice.
 It also applies the framework in the synthesis and appraisal of 165 academic articles drawn from the transportation justice literature; these studies use equity analysis as instruments to gauge proximity to their context's just situation.
@@
 
-## Call 1: Underpin standards with rigorous concepts of justice {#sect4.1}
+## Call 1: Develop standards based on rigorous concepts of justice {#sect4.1}
 
 The conceptual grounding for standards is often left implicit within the literature: for example, @muellerSocioeconomicInequalitiesUrban2018 suggests the relative risk of mortality from transport-related air pollution should not be higher in deprived groups than the general population.
-While the conceptual justice foundations are not explicitly declared, we infer the standard conceptualizes _wellbeing_ and could be in service of reparative justice.
+While the conceptual justice foundations are not explicitly declared, we infer an egalitarian focus.
+But, what level of relative risk is acceptable for the general population?
+What purpose does mortality risk serve and who benefits from it?
 To move equity analysis outputs towards just transport futures, explicit inclusion of justice rationales (the "Why?") should become more common place.
-We must be clear with our terms – what is equity and for whom?
-
-Standards serve as benchmarks for fairness, yet some standards are seemingly arbitrary in the reviewed literature.
-For example, @caoIndicatorsSocioSpatialTransport2017 proposes 20 ferries per day to avoid social exclusion for inter-island transport planning in the Philippines, though acknowledging the standard should be politically determined.
-It is unclear from the paper however, if selecting a different benchmark such as 10 or 30 ferries would make a difference in any specific object of justice (e.g., accessibility to particular destinations) or if that number is tied to funding or resource constraints.
-As another example, what is the conceptual underpinning of the 15-minute city, where 15 minutes is the standard for travel times.
-Is it sufficientarism?
-Or is it egalitarianism?
+We must be clear with our terms--what is equity and for whom?
+
+Standards are statements of fairness, yet some standards are seemingly arbitrary in the reviewed literature--in other words, set arbitrary justice goals.
+For example, @caoIndicatorsSocioSpatialTransport2017 proposes 20 ferries per day to avoid social exclusion for inter-island transport planning, though acknowledging the standard should be politically determined.
+It is unclear if selecting a different benchmark such as 10 or 30 ferries would make a difference in any specific object of justice (e.g., accessibility to particular destinations) or if that number is tied to funding or resource constraints.
+Another example is the conceptual underpinnings of the 15-minute city, where 15 minutes is the standard for travel times.
+Is this sufficientarism?
+Or egalitarianism?
 The standard can be interpreted in a multitude of ways.
-We argue that the framing used to conceptualize justice must guide the selection of the standard.
-As @martensFair2021 convincingly show, being explicit about "Why" is important.
+We argue that conceptualization of justice must guide the selection of the standard.
+As @martensFair2021 demonstrate, being explicit about "Why" is important.
 Recent work by @karnerAdvancesPitfallsMeasuring2024 provides several concrete examples for developing standards that follow directly from different justice conceptualizations.
 
-For decision-makers, setting standards, measuring inequities and moving towards setting flexible guidelines for standards that are also compatible with community-informed calls for justice is the next step for transportation equity planning.
-If we assume that one function of the academic literature is to recommend standards, researchers must connect compatible equity conceptualisations and standards to justice frameworks.
-Currently, these fundamental connections are sorely needed.
-
-## Call 2: Develop creative methods for systems-thinking approaches to equity {#sect4.2}
-
-On the methodological side, transportation equity research would benefit from wider use of mixed-methods.
+For decision-makers, setting standards, measuring inequities, and developing flexible guidelines for standards that are also compatible with community-informed calls for justice, are the next steps for transportation equity planning.
+If we assume that one function of the academic literature is to recommend standards, researchers must connect compatible conceptualisations and standards to justice frameworks.
+Currently, these fundamental connections are mostly missing.
+
+## Call 2: Develop creative methods for systems-thinking approaches to fairness {#sect4.2}
+
+On the methodological side, transportation justice research would benefit from wider use of mixed-methods.
 Concepts and standards are often discussed from purely qualitative or quantitative approaches.
 This is a missed opportunity to combine the strengths of both approaches, whether by deep diving into some particular experiences or perceptions through qualitative methods or tailoring more meaningful quantitative analysis after qualitative explorations.
-As mixed-method examples, @xieWonCycleRoute2018 find through interviews and go-alongs with women cyclists that the standard Cycling Level of Service (CLS) tools used by engineers to plan cycling infrastructure misses the critical gendered perspective.
+As mixed-method examples, @xieWonCycleRoute2018 find through interviews and go-alongs with women cyclists that the standard Cycling Level of Service (CLS) tools used by engineers to plan cycling infrastructure misses a critical gendered perspective.
 Further, @somenahalliAgingTransportMobility2007 survey older adults to understand their mobility issues, revealing factors that are unseen in standard daily travel surveys.
 
-While disparity analysis is frequently used in the reviewed literature, the standards suggested do not often align with practical application.
+While disparity analysis is frequently used, the resulting standards do not often align with practical applications.
 For example, metrics of accessibility (usually measured with travel impedance cut-offs of between 15 to 60 minutes depending on the destination, population group and mode) are used to show descriptive differences among areas and groups but with scarce implications as to the experience of travelers.
-How low of an accessibility index is too low?
-Relatedly, given sufficient accessibility levels are reached, excessively high accessibility values will result in high inequalities.
+How low must an accessibility index be before it is too low?
+Rellatedly, once sufficient accessibility levels are reached, excessively high accessibility values can result in high inequalities.
 Are inequalities of a certain level an issue?
-This lack of discussion makes translating results from these equity analysis into policy practice that may move conditions towards justice challenging.
+A lack of discussion poses a challenge to translating results from these analyses into policy and practice.
 Creative methods and discussions of quantitative accessibility metrics should be paired with results; they should yield interpretable results.
@@
 Other times, when analysis engages with metrics that may be tied to particular concepts of equity (like the Gini coefficient or Theil index), they fall short in assessing whether the results are good or bad [@mijaresEquityAnalysisUrban2013].
-If a Gini coefficient of 0 means that all people have the same access to public transport stops, what does it mean if the result is 0.2, 0.3, or 0.4?
+For example, a Gini coefficient of 0 would mean that all people have the same access to public transport stops.
+But is this level of access good enough or lacking?
+And what does it mean if the result is 0.2, 0.3, or 0.4?
 Is this good or bad news?
@@
 Are new policies needed to reduce that number to a certain threshold, orienting future interventions?
-These questions usually remain unanswered in the literature despite its importance.
-These measurements can also bring some challenges and pitfalls (as recently summarized by @karnerAdvancesPitfallsMeasuring2024) but are necessary to move equity analysis into application and towards transport justice.
+These questions usually remain unanswered in the literature despite their importance.
+These measurements can also bring some challenges and pitfalls as summarized by @karnerAdvancesPitfallsMeasuring2024 but are necessary for more effective equity analysis.
 
@@
 
-In the review of the literature, we were left wondering specifically - what are the motivations for the use of some destination-types and the choice to not include other destinations types?
-For instance, papers including leisure destinations (e.g., green space, parks, recreation) are infrequently studied and some categories are missing all together.
+In our review of the literature, we were left wondering specifically--what are the motivations for the use of some destinations and the choice to not include others?
+For instance, leisure destinations (e.g., green space, parks, recreation) are infrequently studied and some categories are missing altogether.
 We suspect that since methods used are predominately quantitative, the reliance on commonly used point of interest databases is also high.
-These databases typically include education, health and occasionally include aggregated categories for leisure and community.
-They are quiet on types of 'community' and 'leisure' (are they a community organization, government services, a visit to a family or friend, a care-giving destination).
-Further, they may be missing categories all together like childcare, typically daycare or facilities -- domestic work, mobilities of care, and mobility interdependence.
-These missing categories are underrepresented in the reviewed corpus relative to the presence of work and healthcare destinations.
-
-We can only track what we have collected and compiled; and we know that transport systems' focus is more than just to work or as a source of economic development (though in underdeveloped regions, transport systems as a force of economic development is pronounced based on the concentration of Global South transport-induce economic development studies e.g., high-speed rail [@chenImpactHighspeedRail2017; @monzonEfficiencySpatialEquity2013; @kimImpactsHighspeedRail2015]).
-As such, data availability matters, especially in the context of the Global South and rural geographies as fewer official data sources and public research resources exist relative to urban communities and areas in the Global North.
+These databases typically include education, health and occasionally include aggregated categories for leisure and community, but are less generous in terms of 'community' and 'leisure' points of interest (community organizations, government services, care-giving destinations).
+Further, categories like childcare are often missing, typically daycare or facilities--domestic work, mobilities of care, and mobility interdependence.
+Not coincidentally, these categories are underrepresented compared to work and healthcare destinations.
+
+We can only track what we have collected and compiled; and we know that transport systems are more than just facilitators to work or as a source of economic development (though in underdeveloped regions, transport systems as a force of economic development is pronounced, e.g., high-speed rail [@chenImpactHighspeedRail2017; @monzonEfficiencySpatialEquity2013; @kimImpactsHighspeedRail2015]).
+As such, data availability matters.
+In the context of the Global South and rural geographies fewer official data sources and public research resources exist relative to urban communities and areas in the Global North.
 As well as in the operationalization of emerging conceptual theories in equity analysis, such as sufficientariansm [@vanderveenOperationalizingIndicatorSufficient2020].
@@
 And, who gets to consult and use the data?
-Deciding who is the subject of justice frames the data collection activity; if it's the mobility of those that do domestic work, the classification of who does this work and how it changes over time/space is fundamental.
-How we classify has implications for our understanding of just who is a subject of justice, as illustrated by the history of racial classifications in the US [@leeRacial1993].
-What methods we're using, under what spatio-temporal boundaries we conduct data collection, and who has (and doesn't have) access to the collected data can all impact fulsome conceptualizations of justice.
+Deciding who is the subject of justice frames the data collection activity; if it is the mobility of those that do domestic work, the classification of who does this work and how it changes over time/space is fundamental.
+How we classify mobility and access has implications for our understanding of just who is a subject of justice, as illustrated by the history of racial classifications in the US [@leeRacial1993].
+The methods we use, the spatio-temporal boundaries for data collection, and who has (and does not have) access to the collected data, can all impact how we think about justice.
 In the case of transportation, the issue of data collection/availability as a matter of justice is gaining traction as digitization of data casts a starker light on these questions [@sourbatiSmart2021; @behrendtData2023].
@@
 
-Robust assessments of the implications of equity standards on the experience of the public is still lacking in the literature, but it is essential for equity analysis that translates into just practice.
-While the estimation of the benefits of increased mobility or accessibility, or reducing affordability burdens and transport externalities is commonplace in the literature, these estimations need to be associated with outcomes like life and neighbourhood satisfaction, subjective well-being, mental and physical health, social capital, among others.
-
-Composite measures such as a transport-land-use index proposed by @appleyardTransitCorridorLivability2017 is an example of systems-thinking approach that links findings to quality-of-life proxies; they ground their measure in the principle of _livability_ along corridors of varying levels of estimated transport-land-use integration.
-However, the methods used demonstrate areas of relative high and low livability could go further; they could be tied to absolute goals of integration or livability.
+Robust assessments of the implications of equity standards on lived experiences are still lacking in the literature, but are essential for equity analysis that translates into just practice.
+While the estimation of the benefits of increased mobility or accessibility, or reducing burdens is commonplace in the literature, there is a need to associate them to outcomes, including life and neighbourhood satisfaction, subjective well-being, mental and physical health, and social capital.
+
+Composite measures such as the transport-land-use index proposed by @appleyardTransitCorridorLivability2017 is an example of systems-thinking approach that links findings to quality-of-life proxies; they ground their measure in the principle of _livability_ along corridors of varying levels of estimated transport-land-use integration.
+However, the methods could go further; they could be tied to absolute goals of integration or livability.
 Relatably, @higgsUrbanLiveabilityIndex2019 develops an urban livability index and demonstrates the relationship between the population's use of a certain mode in a neighbourhood and a one unit increase in the index.
@@
 
-We believe there is a need to evaluate equity interventions and policies: track their before, during, and after impacts.
-In this review, only approximately 10% of studies assess specific transport-related policy interventions through an equity lens<!-- double check this number-->.
+There is a need to evaluate equity interventions and policies: track their before, during, and after.
+In this review, barely 10% of studies assess specific transport-related policy interventions through an equity lens.
 Examples include mode-shift from driving to active school travel [@mammenSchoolTravelPlanning2014], transit fare restructures [@hickeyUsingQuantitativeMethods2010] and spatial analysis of Low Traffic neighbourhoods [@aldredEquityNewActive2021].
@@
 
-During the preparation of this work, the first and second authors used ChatGPT exclusively for copy-editing in the section "An appraisal of the lay of the land", aiming to reduce word count while keeping the original writing style and to correct typos.
+During the preparation of this work, the first and second authors used ChatGPT exclusively to reduce word count in the section "An appraisal of the lay of the land" while keeping the original writing style.
 After using this tool, all authors reviewed the manuscript and take full responsibility for the content of the publication.
@@
 ```
+
+:::landscape
 
@@
 ```{r}
-set_flextable_defaults(font.size= 7,
-                       font.family = NULL,
-                       text.align = 'justify')
-
-flextable(dat) |>
-  ftExtra::colformat_md() |>
-  compose(
-    i = 1, j = c(1,2,3,4), part = "header", value = as_paragraph(c("Dimension", "Continent", "Conceptualization", "Standard"))) |>
-  merge_v(j=1) |> #merge 'dimensions' column
-  rotate(part = "body", rotation = "btlr", j=1, i=c(1:5,6:7,8:12, 13:14,15:18)) |>
-  valign(valign="center", part="body", j=1) |>
-  hline( i = c(5,7,12,14), part = "body") |>
-  set_table_properties(layout = "autofit")
-
-```
+dat |>
+  gt() |>
+  tab_options(table.font.size = "xx-small") |>
+  cols_label(X1 = "Dimension",
+             X2 = "Continent",
+             X3 = "Conceptualization",
+             X4 = "Standard") |>
+  tab_caption("Examples of Data Extraction")
+```
+:::

### 2025-02-17 Journal-manuscript/Journal-manuscript.qmd
New definitions; revised manuscript. For Nacho review

@@
 ---
-title: "Searching for standards of fairness in the transportation justice literature"
+title: "Searching for standards of fairness in the transportation literature"
 format: docx
@@
 #       - ref: MAC
-#     orcid: 0000-0003-2402-304X
+#     orcid: 0000-0003-4371-4831
 #     email: soukho@mcmaster.ca
@@
 #       - ref: TORONTO
-#     orcid: 0000-0003-2402-304X
+#     orcid: 0000-0002-7385-2357
 #     email: i.tiznadoaitken@utoronto.ca
@@
 #       - ref: UNC
-#     orcid: 0000-0003-2402-304X
+#     orcid: 0000-0002-8800-2777
 #     email: palmmatt@unc.edu
@@
 #       - ref: TORONTO
-#     orcid: 0000-0003-2402-304X
+#     orcid: 0000-0002-3870-5984
 #     email: steven.farber@utoronto.ca
@@
 #       - ref: MAC
-#     orcid: 0000-0003-2402-304X
+#     orcid: 0000-0001-6912-9919
 #     email: paezha@mcmaster.ca
@@
   # - literature review
-
+abstract: |
+
+keywords: [transportation; equity; justice; standards; transportation planning; fairness; literature review]
 bibliography: bibliography.bib 
@@
 df_mode <- df_mode |> group_by(mode) |> 
-  summarize(count = n()) |> 
+  summarise(count = n()) |> 
   mutate(count = count/165,
@@
 
-df_pop <- df_pop |> group_by(pop) |> summarize(count = n()) |> 
+df_pop <- df_pop |> group_by(pop) |> summarise(count = n()) |> 
   mutate(count = count/165)
@@
 
-NUM_standards_conc <- standards_conc |> group_by(id) |> summarize(count = n())
+NUM_standards_conc <- standards_conc |> group_by(id) |> summarise(count = n())
 NUM_ENTRIES_standards_conc <- NUM_standards_conc$count |> sum()
@@
 #   geom_bar() +
-#   labs(x = "Conceptualizations (summarized)", fill = "Standards (main categories)") +
+#   labs(x = "Conceptualizations (summarised)", fill = "Standards (main categories)") +
 #   theme_classic() + 
@@
   case_cont=unlist(case_cont_count))
-case_cont_count <- case_cont_count |> group_by(case_cont) |> summarize(count = n())
+case_cont_count <- case_cont_count |> group_by(case_cont) |> summarise(count = n())
 
@@
   urban_form=unlist(urban_form_count))
-urban_form_count <- urban_form_count |> group_by(urban_form) |> summarize(count = n())
+urban_form_count <- urban_form_count |> group_by(urban_form) |> summarise(count = n())
 
@@
 
-dest_count <- dest_count |> group_by(dest_count) |> summarize(count = n()) 
+dest_count <- dest_count |> group_by(dest_count) |> summarise(count = n()) 
 ```
@@
 standards_conc <- standards_conc |> 
-  group_by(conc, std_mcat) |> summarize(total = n())
+  group_by(conc, std_mcat) |> summarise(total = n())
 
@@
   conc_count=unlist(conc_count))
-conc_count <- conc_count |> group_by(conc_count) |> summarize(count = n())
+conc_count <- conc_count |> group_by(conc_count) |> summarise(count = n())
 
@@
   stand_count=unlist(stand_count))
-stand_count <- stand_count |> group_by(stand_count) |> summarize(count = n()) 
+stand_count <- stand_count |> group_by(stand_count) |> summarise(count = n()) 
 ```
@@
 
-# Introduction {#sect1}
+# Abstract
+
+This work provides a synthesis of how transportation fairness, justice and equity literature has operationalized standards.
+We first outline a flexible framework for engaging fairness questions and then apply it to collate the review's results.
+We find most articles were published within the last half decade and center urban areas, with a minority of work (40%) exploring Global South case studies.
+Income groups, followed by specific age groups (i.e., elders) and people with disabilities, are most commonly the subjects of justice.
+Transit and pedestrian modes are the most frequently studied mobility tools; with many analyses being multimodal and comparative in nature.
+The benefits and burdens, the "What" of mobility, focuses on movement (i.e., quality of trips taken) or the potential for movement (i.e., accessibility).
+How the reviewed works conceptualizes fairness is varied (e.g., Vertical Equity, Wellbeing, Rights-Based); but is often delineated with a supporting opportunity standard (37%) (i.e., number of parks within a 30-minute travel time) or population standard (36%) (i.e., bottom income quartile), though Infrastructure (i.e., level of service) and Environmental+ (i.e., air pollution) standards are also prominent.
+While the reviewed literature covers plenty of ground, we conclude with calls for action directed at researchers, practitioners, and transportation advocates.
+
+# Introduction
 
@@
 Beyond ethical motivations, tracking objective and perceived inequalities is of interest for governing bodies to respond to popular demands for fairness.
-This is a challenging task: transportation systems are notoriously complex, with benefits and burdens that are diffuse over space and time.
+However, this has proven to be a challenging task: transportation systems are notoriously complex, with benefits and burdens that are diffuse over space and time.
 To compound matters, emerging technologies and service models can swiftly change the balance of benefits and burdens among a population [@guoSystematicOverviewTransportation2020].
-Transportation systems that are engineered to offer higher mobility for people _somewhere_ can simultaneously cut others off from essential opportunities _elsewhere_, as @rajeEngineering2004 and others have poignantly illustrated.
-Over time, the shades of policies past long haunt a region and the planet, as shown by the legacy of urban highways all across the US [@archerWhite2020] and the time horizon for the impacts of climate change [@markolfTransportation2019].
-
-The objective of this work is to scan the state of knowledge in terms of defining and operationalizing fairness in transportation.
-Much research has been devoted to the issues of _measuring_ equity in transportation, including (among many others) [@delboscUsing2011; @martensMeasuring2019; @pritchardNew2022].
+Transportation systems that are engineered to offer higher mobility for people _somewhere_ can simultaneously cut others off from essential opportunities _elsewhere_ [@rajeEngineering2004].
+The shades of policies cast long shadows, as shown by the legacy of U.S.
+urban highways [@archerWhite2020] and impacts of transportation-related climate change [@markolfTransportation2019].
+
+Responding to this topic, a plethora of academic literature has emerged in recent decades.
+For instance, much research has been devoted to the issues of _measuring_ equity in transportation, including (among many others) [@delboscUsing2011; @martensMeasuring2019; @pritchardNew2022].
 Further, there are multiple works that discuss the conceptual and philosophical foundations of equity and fairness in transportation [e.g., @martensTransport2016; @pereiraDistributiveJusticeEquity2017a; @vanoutriveHow2019].
-Previous reviews of equity in planning documents have been tightly scoped to cover accessibility [e.g., @boisjolyHowGetThere2017] or a particular mode of transportation [e.g., cycling, @doranPursuitCyclingEquity2021].
-These efforts are valuable, but in our estimation there remains a gap in terms of understanding how standards for equity are developed and implemented for transportation systems.
-
-To illustrate this gap, we note how @beilerExploring2016 in their exploration of transport equity, cite the following strategies identified by the U.S.
-Department of Transportation to address matters of justice [p.
-287]:
-
-- Reduce adverse human health and environmental effects on minority and low-income populations.
-- Include all potentially affected communities in the transportation decision-making process.
-- Ensure that minority and low-income populations receive equitable benefits.
-
-While commendable, the strategies are too vague, which means it is possible to implement them in multiple ways, to genuinely comply with the spirit of justice, or performatively to deceive it [@mcculloughPerformative2023].
-Relevant questions include: how much should adverse effects be reduced?
-To zero?
-Or to some tolerable level of adversity greater than zero?
-What should that level be?
-What benefits should be distributed?
-What share of the benefits should minority and low-income residents receive for a project to be sufficiently ‘equitable'?
-These questions touch on the development and use of _standards_.
-In this work, we choose to focus exclusively on the academic domain as we know that adoption of equity concepts in planning practice has lagged developments in academic work [@pereiraTransportationEquity2021; @boisjolyHowGetThere2017; @linovskiEquityTransportationPlanning2020; @doranPursuitCyclingEquity2021].
-
-Supporting the creation of (more) just transportation systems involves developing standards to track the production and management of benefits and costs; how they are distributed; what values they implement, as well as by whom [@pereiraDistributiveJusticeEquity2017a; @pereiraTransportationEquity2021].
-In this spirit, our work seeks to make two contributions.
-First, it outlines a conceptual and flexible framework for engaging questions of transportation fairness and equity analysis, based on the questions "Why?", "Where?", "When?", "Who?", "What?", and "How?" (5WH).
-Second, it applies the framework to collate the existing knowledge about equity standards.
-In contrast to previous reviews on measuring inequality in transportation systems, we are concerned with the implicit or explicit standards used to judge whether inequalities are fundamentally “fair” or unacceptable.
-To achieve this we scan the state of academic knowledge in defining standards of fairness in transportation.
-
-# Background {#sect2}
-
-In the introduction the terms "justice", "equity", and "fairness" were used relatively loosely.
-In the following, we describe justice as a target, equity as a set of measurement tools, fairness as a yardstick of justice, and standards as statements of fairness.
-
-**Justice** is an end goal, a desirable state of affairs that we are morally obligated to achieve.
+Previous reviews of equity in planning documents have been tightly scoped to cover accessibility [e.g., @boisjolyHowGetThere2017] or a particular mode of transportation [e.g., cycling in @doranPursuitCyclingEquity2021].
+These efforts are valuable, but in our estimation there remains a gap in terms of understanding _how_ standards for equity are developed and implemented for transportation systems.
+
+The objective of this work is to broadly scan the state of this knowledge.
+Through a distributive justice lens [@pereiraDistributiveJusticeEquity2017a; @pereiraTransportationEquity2021], our work seeks to make two contributions.
+First, it outlines a conceptual and flexible framework for engaging and analysis of transportation fairness questions, based on the questions "Why?", "Where?", "When?", "Who?", "What?", and "How?" (5WH).
+Second, it applies the framework to collate the existing knowledge about fairness standards.
+To achieve this, we scan the state of academic knowledge in defining standards of fairness in transportation.
+In contrast to previous reviews on measuring inequality in transportation systems, this work is concerned with the implicit or explicit standards used to judge whether inequalities are fundamentally “fair” or unacceptable.
+
+# Background
+
+## Defintions
+
+Fairness, equity and justice are closely related and often colloquially used interchangeably.
+In this section, we clarify these concepts definitions to facilitate their precise use throughout the manuscript.
+
+**Fairness** is a subjective and moral assessment of a treatment, outcome, or both, in relation to a given state of affairs.
+Within the transportation domain, the fairness of the distribution of transportation "goods" and "burdens" (e.g., access to opportunities, transportation-related externalities) for individual, groups and society is often at issue [@pereiraTransportationEquity2021].
+As examples: congestion pricing schemes could be unfair if they disproportionately disadvantage lower income groups [@eliassonCongestionPricingFair2016], systems designed to service "mandatory" destinations like employment or commercial retail that are not as relevant to all populations like those intangled in caring-work [@ravensbergenExploratoryAnalysisMobility2023; @hailConceptFairnessRelation2021], or even the application of justice such as Title VI of the U.S.
+Civil Rights Act of 1964 for the purpose of accessibility planning as discussed in @martensFair2021.
+Individuals may feel certain situations are unfair, processes can be seen as unfair, or even outcomes of these processes and situations can be unfair.
+
+There are a multitude of aspects to **fairness** in transportation [@hailConceptFairnessRelation2021], namely because the distribution of transportation "goods" and "burdens" are often ambiguous.
+Individuals' ability to access and benefit from transportation-related outcomes varies continuously, often in ways that are unrecognized (implicitly or explicitly) or inadequately considered by transportation planners and other influential actors in the system.
+To put it briefly, when it comes to the distribution of "goods," fairness can be understood as a _moral_ appraisal concerning the amount and strength of moral claims to those goods [@winteinBeFairClaims2024].
+Hence, in this manuscript, we define **fairness** as the moral evaluation of the rightness or wrongness of a given state of affairs (e.g., the impartiality and consistency of treatment or associated outcomes); fairness serves as a yardstick for justice.
+
+Relatedly, **justice** can be viewed as the formalized goal of fairness, an analytically reasoned _fair_ state of affairs.
 Justice is attained when people "give and receive whatever they are due" [@jaggarPhilosophicalChallengesGlobal2009, pp.
@@
 440].
-Justice is fluid because it depends on the desirability of different states of affairs, which may change between populations, places, and times.
-That said, several forms of justice can be distinguished [@jaggarPhilosophicalChallengesGlobal2009; @pereiraDistributiveJusticeEquity2017a; @karnerTransportationEquityTransportation2020].
+Different scopes and approaches to justice have been develop, as formalizing fairness depends on the desirability of different states of affairs, which depend on populations, places, times, and different scopes.
+For instance, several forms of justice can be distinguished [@jaggarPhilosophicalChallengesGlobal2009; @pereiraDistributiveJusticeEquity2017a; @karnerTransportationEquityTransportation2020]:
 
@@
 - **Reparative (or restorative) justice** focuses on the reparation of caused harm; it centers the needs and voices of victims to restore wrongdoers and the community according to moral values [@tylerRestorativeJusticeProcedural2006; @braithwaite2003principles].
-In planning and policy contexts, reparative justice often involves accountability mechanisms to compensate victims [@safransky2022grammars].
+In planning and policy contexts, reparative justice involves accountability mechanisms (material, powers, rights, processes) to compensate victims [@safransky2022grammars; @williamsWeStepHow2023].
 
@@
 Its main concern is the collection of benefits and burdens of the tangible and intangible products of society by different segments of a population.
-
-It might be argued that all forms of justice have a distributive facet.
-To illustrate, retributive justice is achieved by distributing societal intangibles like "freedom" (e.g., of movement, of association) as benefits, and/or tangible resources as burdens (e.g., fines as a punishment).
-Reparative justice deals with the distribution of benefits and burdens to redress past wrongs, for example by asking those who have benefited from said wrongs, even if unwittingly, to shoulder a bigger fiscal burden to benefit those who are still harmed by them.
-Procedural justice is the distribution of societal benefits (e.g., the right to voice an opinion as a recognized stakeholder in a process) and burdens (e.g., the effort required to develop an educated opinion) when making decisions with collective consequences.
-
-From this perspective, we can speak of the _purposes_ of distributive justice: to mete out retribution _fairly_, to repair past harms, and to ensure that procedures offer _equitable_ opportunities to influence outcomes.
-Accordingly, equity and fairness are the "instruments" of justice, the tools by which society moves towards it.
-
-**Equity**, as conceptualized alongside distributive justice, tends to encompass tools to understand the distribution of benefits and burdens of things among a population, often emphasizing those with the least advantage.
-In the transportation domain, the term is somewhat loaded because it is perceived as stemming from the authority of the state, and meant to assist with decisions about regulating and financing spending [@karnerTransportationEquityTransportation2020].
-We are in agreement with @karnerTransportationEquityTransportation2020 that equity analysis should not be seen as an end in and of itself, but rather as a means to gather information about actual and perceived inequities.
-In this respect, the analytical tradition of equity, at least in transportation planning, means that the relevant models become embedded in a political ecology of estimated truth, and their assumptions and scope must be open and transparent to avoid misuse and even abuse as tools of subjugation.
-
-**Fairness** is somewhat more complicated to define.
-The concept does not have the same history of development as an analytical tool, and can be interpreted in numerous, and possibly discordant ways.
-This is demonstrated by @martensFair2021 in their study of the application of Title VI of the US Civil Rights Act of 1964 in accessibility planning.
-Title VI explicitly talks about the distribution of benefits derived from Federal funding: "[n]o person in the United States shall, on the ground of race, color, or national origin, be excluded from participation in, be denied the benefits of, or be subjected to discrimination under any program or activity receiving Federal financial assistance." However, as @martensFair2021 show, there are wildly different paths to compliance, ranging from the banal (do not _knowingly_ discriminate), to the substantive (compensation for past discrimination within a societal context that recognizes harm was done, i.e., reparative justice).
-What kind of justice does fairness serve in each case?
-It depends on the rationale for seeking justice in the first place.
-Our reading of @martensFair2021 is that fairness is a yardstick that is best deployed _a priori_ than _a posteriori_, for doing the latter risks rationalizing the outcomes rather than driving them.
-
-A **standard** is a way of making concrete statements about fairness.
-Returning to the ambiguities in Title VI, the attainment of justice _depends on the standard_.
-For example, explicit non-discrimination constitutes a very weak standard that takes aim at the actions of agencies instead of the recipients of the benefits.
-Accordingly, any distribution of benefits would be considered fair, as long as the agency does not explicitly and knowingly target or deny benefits to particular groups.
-The standard provides conditions to determine whether a situation is _fair_.
-A similarly weak standard is a Pareto improvement, whereby it is possible to concentrate the benefits as long as no group is worse off compared to the status quo [@tan2016pareto; @xuParetoimprovingPoliciesIdealized2018].
-A policy based on this standard would be fair as long it does no harm.
-<!--A somewhat more strict standard, a Pareto-Plus improvement, stipulates that an intervention is fair when all groups receive at least _some_ (non-trivial) benefits; the size of the benefits for each group is irrelevant.
-In contrast to the notion of "do no harm", such a standard embodies the ideal that no one is denied benefits.
---> A standard based on egalitarian principles (e.g., proportional equity) would weigh the benefits or burdens by population; in fairness, each group would give or receive in proportion to their size [@billsLookingMeanEquity2017;@martensJusticetheoreticApproachDistribution2012].
-In contrast, an affirmative action standard (restorative justice) would be even stricter, requiring the benefits to be distributed in a non-egalitarian way to favor people still harmed by past or present discriminatory practices [@bierbaumMobilityJusticeLinking2021].
-
-## A framework to analyze questions of justice: 5WH {#sect2.2}
-
-Having defined "justice", "fairness", "equity" and "standards", we approach the literature with an analytical apparatus inspired by the framing of @jaggarPhilosophicalChallengesGlobal2009 for philosophical questions of justice.^[Similar questions are found peppered throughout the literature.
+The just distributive aspects of transportation systems is also of focus in this work.
+
+Wile justice (i.e., formalized fairness) is a broad moral concept, **equity**  and **standards** can be understood as the 'instruments' of justice- the tools through which society moves towards a just state of affairs.
+
+**Equity**, as conceptualized alongside distributive justice, tends to encompass tools to understand the distribution of benefits and burdens of things among a population (e.g., disparities), emphasizing those with the least advantage or most disadvantage.
+In the transportation domain, equity analysis often flows from the top: stemming from the authority of the state and meant to assist with decisions about regulating and financing spending [@karnerTransportationEquityTransportation2020].
+As in @karnerTransportationEquityTransportation2020, equity and disparity analysis should not be seen as an end in and of itself, but rather as a means to gather information about actual, observed, and perceived inequities.
+
+And lastly, in this work's context, a **standard** is a concrete statement about fairness; it is a formalized threshold in understanding disparities (i.e., related to equity) as part of the goal of formalized fairness (i.e., justice).
+A standard establishes criteria for how benefits or burdens should be distributed or allocated in a way that aligns with moral or formalized principles.
+For example, a weak standard might be a Pareto improvement, where benefits may be concentrated in certain groups (even those already advantaged), as long as no group is made worse off compared to the existing situation [@tan2016pareto; @xuParetoimprovingPoliciesIdealized2018].
+As another example, a more strict standard could be based on egalitarian principles (e.g., proportional equity) where the benefits or burdens are weighted by population; in the consistent application of this standard, each group would give or receive in proportion to their relevant 'size' [@billsLookingMeanEquity2017;@martensJusticetheoreticApproachDistribution2012].
+In contrast, an affirmative action standard (i.e., could be related to restorative justice) could be an even stricter standard, requiring the benefits to be distributed in a non-egalitarian way to favour people still harmed by past or present discriminatory practices [@bierbaumMobilityJusticeLinking2021].
+
+To summarise, we use **justice** to describe a moral target, **fairness** as an evaluation or yardstick of justice, **equity** as a set of related disparity measurement tools, and **standards** as statements of fairness.
+
+## A framework to analyze questions of justice: 5WH
+
+Having defined justice, fairness, equity and standards, we approach the literature with an analytical apparatus inspired by the framing of @jaggarPhilosophicalChallengesGlobal2009 for philosophical questions of justice.^[Similar questions are found peppered throughout the literature.
 This is done either explicitly, as for example in @karnerTransportationEquityTransportation2020, who ask "of what", "for whom," and "how much" in reference to equity; or implicitly, as in @gosslingUrban2016, who asks of the outputs ("what?") of transportation (exposure, space, access) and "for whom?" (gender, age, ethnicity).].
@@
 Conventionally, the question of **"When?"** refers to the temporal circumstances within which the demands of justice apply.
-When it comes to transportation, questions regarding temporailty are important, as for example: _when_ did the equity analysis take place and under what historical policy context; when is _the right time_ to invest in infrastructure and as a result when to generate a spatial inequality [@rabelloEvaluation2015]; for _how long_ the burdens and benefits can be associated to a specific intervention; or even the _timeline_ of reparative justice to reconcile the shadows of past transportation-related injustices.
+When it comes to transportation, questions regarding temporarily are important, as for example: _when_ did the analysis take place and under what historical policy context; when is _the right time_ to invest in infrastructure and as a result when to generate a spatial inequality [@rabelloEvaluation2015]; for _how long_ the burdens and benefits can be associated to a specific intervention; or even the _timeline_ of reparative justice to reconcile the shadows of past transport-related injustices.
 
@@
 The next question is **"How?"**, and it relates to the allocation of various objects of justice ("what") to various subjects of justice ("who") in various circumstances ("when" and "where").
-Equity standards are a tool for answering this distributive question, by defining what is fair.
-The thresholds can be quantitative (e.g., square meters of green space per capita), or they can be qualitative descriptions (e.g., do not knowingly discriminate), or a mix of the two.
+Fairness standards are an equity tool for answering this distributive question.
+The thresholds can be quantitative (e.g., square meters of green space per capita), qualitative descriptions (e.g., do not knowingly discriminate) or a mix of the two.
 Some examples include: maximum travel distance/cost/time to or from key destinations, levels of maximum exposure to externalities (i.e., noise or air pollution), un/fulfilled needs, and dis/satisfaction with travel.
-To support us when approaching this question we draw from concepts in transport-related social exclusion, transport disadvantage, and/or transport poverty, which are typically based on utilitarian or sufficientarian philosophies.
+To support us when approaching this question, we draw from concepts in transport-related social exclusion, transport disadvantage, and transport poverty, which are typically based on utilitarian or sufficientarian philosophies.
+The "How?" is supported by fairness conceptualisations along with standards.
 
 Above all, convincing answers to the above questions require a supporting rationale: a **"Why?"** [@jaggarPhilosophicalChallengesGlobal2009].
-This is perhaps the most slippery of the questions posed here.
+This is perhaps the most slippery of 5WH.
 Justice is inherently a social construct.
@@
 These contracts can be defined by constitution, but there are often unwritten and possibly contested variants.
-In this way, analyzing the "Why?" in the reviewed literature is not the focus of this review, partly because answers to "Why?" are seldom explicitly stated.
-Instead, our focus is on the standards of fairness that, combined with the use of equity analysis, can help us understand how better to move towards just transportation systems and better formulate answers to "Why?".
-
-## Methods {#sect2.3}
-
-This review examines the academic literature on transportation to identify the extent to which standards for equity are defined and employed.
+In this way, analyzing the "Why?" in the corpus is not the focus of this review, partly because answers to "Why?" are seldom explicitly stated.
+Instead, our focus is on the standards of fairness that, combined with the use of analysis, can help us understand how better to move towards just transportation systems and better formulate answers to "Why?".
+
+## Review methods
+
+This work examines the academic literature on transportation to identify the extent to which standards for equity are defined and employed.
 In this task, we follow the Joanna Briggs Institute (JBI) approach to the conduct of scoping reviews, an approach that builds upon the Arksey and O'Malley (2005) framework [@petersUpdatedMethodologicalGuidance2020].
@@
 The primary research question and the protocol were initially defined by the authors, a group of experts in the field of transportation.
-The initial draft of the protocol was refined from preliminary searches of related reviews [e.g., @zhangLiterature2021; @pereiraDistributiveJusticeEquity2017a; @vecchioAccessibilityCapabilitiesApproach2021], and in consultation with research services librarians.
+The initial draft of the protocol was refined from preliminary searches of related reviews [e.g., @zhangLiterature2021; @pereiraDistributiveJusticeEquity2017a; @vecchioAccessibilityCapabilitiesApproach2021] and in consultation with research services librarians.
+The final protocol was released on OSF [@soukhovEquityJusticeConceptualizations2022], the methods are summarised as follows.
 
@@
 The terms were bundled by means of Boolean operators.
-These stages are summarized as follows (see Appendix section @sec-sect61 for details):
+These stages are summarised as follows (see Appendix section @sec-sect61 for details):
 
@@
 
-The semi-automated nature of the search strategy was overly inclusive, and reduced the risk of omitting relevant material.
-The next stage was to trim the corpus of documents, a task that could no longer be automated.
-The authors working with a group of undergraduate research assistants scanned the documents to assess their relevance based on titles and abstracts.
+The semi-automated nature of the search strategy was overly inclusive, thereby reducing the risk of omitting relevant material.
+The next stage was to trim the corpus; the authors worked with a group of undergraduate research assistants to scan the documents and assess their relevance based on titles and abstracts.
 Two research assistants voted on each document, and a third vote from the authorship team broke ties.
@@
 Next, using a data extraction template and workflow that was pilot-tested with a subset of papers, the authorship team extracted data from the eligible documents using *Covidence* [@covidenceCovidence2023], an online application for literature screening.
-The evidence selection workflow, (PRISMA) flow diagram [@pagePRISMA2020Statement2021], data extraction template, and some sample data extractions can be consulted in Appendix @fig-figA2.
+The evidence selection workflow, flow diagram, data extraction template, and some sample data extractions can be consulted in Appendix @fig-figA2.
 The end result of this scan was a corpus of 165 documents retained for data extraction.
 
-# An appraisal of the lay of the land {#sect3}
-
-This section threads together the trends identified from the data extracted from the corpus (@fig-figA2)<!-- update this cross reference!-->.
-Specifically, the "How" and its application to the objects of justice (the "What"), the subjects of justice (the "Who"), and under which situations ("When" and "Where").
-As an overview, @fig-fig1 shows the prominence of each category under "When?", "Where?", "Who?", and "What?".
-The "How?" category is shown in @fig-fig2.
+# Synthesis of findings
+
+This section threads together the trends identified from the reviewed transportation literature.
+Specifically, a thematic overview of "When?" and "Where?" transportation fairness is considered, "Who?" are the subjects of justice, "What?" could be considered the objects of justice, and lastly "How?" is fairness measured.
+As an overview, @fig-fig1 illustrates the prominence of each thematic category for "When?", "Where?", "Who?", and "What?", while @fig-fig2 presents an overview of the types of fairness conceptualizations and thresholds, addressing the "How?".
 
@@
 #| label: fig-fig1
-#| fig-cap: "The proportion of category for each When, Where, Who and What; papers do often contain more than one topic and those papers are represented multiple times.
-Topic categories were generated upon data extraction."
+#| fig-cap: "The proportion of papers in each 'When', 'Where', 'Who', and 'What' by topics in each category. Topic categories were generated by the authors based on the reviewed literature."
 
@@
 ```
-
 ```{r plot-standards-concs-perc, warning=FALSE}
 #| label: fig-fig2
-#| fig-cap: "The prominence of Hows: the proportion of equity standards (vertical axis) within each type of equity conceptualisation (horizontal axis) category."
+#| fig-cap: "The proportion of papers by topics in the 'How' categories.
+'How' is split into two categories: topics related to fairness conceptualisation (horizontal axis) and topics related to standards (vertical axis).
+Topic categories were generated by the authors based on the reviewed literature."
 
@@
 
-## "When" and "Where" is transportation considered a sphere of justice {#sect3.1}
-
-Within our corpus most papers (`r ((case_cont_count$count[4] + case_cont_count$count[5]+ case_cont_count$count[6])/165) |> percent()`) focus on studies in the Global North, with many studies from North America (particularly USA and Canada), Europe (particularly France, Spain and Northern Europe), Oceania (Australia and New Zealand), and Asia (Japan and Israel).
-Though their subject matter is varied, their spatial context mainly pertains to North America and Europe, and thus more often than not deal with more developed and formal government transport planning apparatuses and technologies e.g., planning for equitable high-speed rail [@monzonEfficiencySpatialEquity2013], autonomous vehicle technology [@eppenbergerOpportunitySharedAutonomous2021], or on the public consultation processes [@reddySafeguardingMinorityCivil2010].
+## "When" and "Where" is transportation fairness considered
+
+Within our corpus, most papers (`r ((case_cont_count$count[4] + case_cont_count$count[5]+ case_cont_count$count[6])/165) |> percent()`) focus on studies in the Global North, with many studies from North America (particularly U.S.
+and Canada), Europe (France, Spain and Northern Europe), Oceania (Australia and New Zealand), and Asia (Japan and Israel).
+Though their subject matter is varied, their spatial context mainly pertains to North America and Europe, and thus more often than not deals with more developed and formal government transport planning apparatuses and technologies e.g., planning for equitable high-speed rail [@monzonEfficiencySpatialEquity2013], autonomous vehicle technology [@eppenbergerOpportunitySharedAutonomous2021], or on the public consultation processes [@reddySafeguardingMinorityCivil2010].
+
+Of note, work included in the reviewed literature does not often explore the historical or temporal aspects of transportation fairness in depth.
+Therefore, in this section we provide a thematic overview of the temporal and spatial contexts associated with the studied case studies and the publication location.
 
@@
 Of all the studies reviewed, `r (urban_form_count$count[3]/sum(urban_form_count$count)) |> percent()` focus on urban and suburban settings and are highly varied in their research aims.
-To give an example, @coxQualitativeStudyAccessibility2020 qualitatively examine cycling as a mode of travel for people with disabilities in a typical mid-size town in the UK.
+To give an example, @coxQualitativeStudyAccessibility2020 qualitatively examine cycling as a mode of travel for people with disabilities in a typical mid-size town in the U.K.
 @ampeImpactChildBike2020, on the other hand, work to identify the lateral clearance that motorists should maintain when passing cyclists with children seats.
@@
 
-The literatures on the Global North and South literature seem to be situated at different stages in the process of addressing transport inequities.
-As formal planning processes in the South operate under greater financial precariety, they rely on more informal processes to address unmet needs.
+We gather that the literature on the Global North and South tend to focus on different aspects of transport inequities, related to the level of their transportation infrastructure.
+Formal planning processes in the South operate under greater financial precariety, they rely on more informal processes to address unmet needs.
 This turns out to be a significant source of equity concerns e.g, informal transit [@friedMeasuringSustainableDevelopment2020], populations living in informal settlements [@sharmaPublicTransitAccessibility2021].
-Lack of resources makes international standards highly relevant e.g., using the WHO noise and air pollution standard [@apparicioCyclingOneMost2021].
-In contrast, these minimum standards may not be relevant in some areas of the Global North e.g., $NO_2$ pollution near roadways in Montreal are consistently below the WHO standard though populations are disproportionately impacted [@carrierApplicationThreeMethods2014].
-Overall, the Global South lags behind the Global North, but this situation presents surprising opportunities for the Global South, to adopt potentially successful enhancements from Global North formal equity planning processes (e.g., indicator creation for disadvantages groups [@cuiSpatialAccessPublic2020]) and to avoid past mistakes (e.g., from entrenched car-centric development [@warrenDevelopingEquitableSustainable2015] to the disproportionate contribution to carbon-intense mobility [@chancelCarbonInequalityKyoto2015].
+International standards are more highly relevant in literature from the Global South e.g., using the WHO noise and air pollution standard [@apparicioCyclingOneMost2021], while not as frequently featured in the Global North literature e.g., in @carrierApplicationThreeMethods2014, disadvantaged populations are disproportionately impacted by higher $NO_2$ pollution near roadways in Montreal but these levels are still consistently below the WHO standard.
+This difference in focus present interesting opportunities for the Global South, to adopt potentially successful enhancements from Global North formal equity planning processes (e.g., indicator creation for disadvantages groups [@cuiSpatialAccessPublic2020]) and to avoid past and on-going mistakes (e.g., from entrenched car-centric development [@warrenDevelopingEquitableSustainable2015] to the disproportionate contribution to carbon-intense mobility [@chancelCarbonInequalityKyoto2015]).
 For the Global North, there is an opportunity to reckon with its own contributions to uneven development globally and environmental impacts as well as adopt relevant aspects from informal planning processes.
 
-## "Who" are the subjects of transportation justice and their mobility tools {#sect3.2}
+## "Who" are the subjects of transportation justice and their mobility tools
 
@@
 
-Turning to the question of "who?" in @fig-fig1, the focus of the literature tends to be on "Income", especially the lowest-income groups who simultaneously face lower mobility and accessibility _and_ higher costs and exposure [@peungnumsaiGridBasedSpatialAnalysis2020; @zhaoUnequableSpatialAccessibility2020; @falavignaAssessingInequalitiesPublic2016].
+Turning to the question of "Who?" in @fig-fig1, the focus of the literature tends to be on "Income", especially the lowest-income groups who simultaneously face lower mobility and accessibility _and_ higher costs and exposure [@peungnumsaiGridBasedSpatialAnalysis2020; @zhaoUnequableSpatialAccessibility2020; @falavignaAssessingInequalitiesPublic2016].
 Evidence suggests that low household income is a significant determinant of transport-related inequities e.g., urban access to public transport [@peungnumsaiGridBasedSpatialAnalysis2020], access to urban employment opportunities [@boisjolyAccessibilityMeasurementsSao2020], and unfavorable rates of environmental noise, air pollution, and green space [@kruizeEnvironmentalEquityRole2007].
 Yet, low-income is not universally associated with lower transport-related benefits for every object of justice.
-For instance, in Sheffield, UK, @mearsUnderstandingSocioeconomicEquity2019 demonstrate that historically working-class (lower-income) neighbourhoods have _more_ access to green space than other neighbourhoods, but of lower quality, likely due to historic urban planning approaches.
-Similarly, @bertrandMeasuringMappingDisparities2008 find that the granularity of the analysis matters, and lower income groups do not always have lower accessibility , something echoed in other spatial and temporal contexts [@fothEquitableTransitExamining2013a; @allenSizingTransportPoverty2019a]
-
-"Age" is the second most common personal trait.
-In this regards, @martinez-jimenezAccessibilityCultureEducation2019 and @arranz-lopezSocialSpatialEquity2019 investigate travel distances/times to various opportunities, demonstrating how age is associated with differences in opportunity access.
+For instance, in Sheffield, U.K., @mearsUnderstandingSocioeconomicEquity2019 demonstrate that historically working-class (lower-income) neighbourhoods have _more_ access to green space than other neighbourhoods, but of lower quality, likely due to historic urban planning approaches.
+Similarly, @bertrandMeasuringMappingDisparities2008 find that the granularity of the analysis matters, and lower income groups do not always have lower accessibility, something echoed in other spatial and temporal contexts [@fothEquitableTransitExamining2013a; @allenSizingTransportPoverty2019a]
+
+"Age" is the second most common "Who?" focus.
+In this regard, @martinez-jimenezAccessibilityCultureEducation2019 and @arranz-lopezSocialSpatialEquity2019 investigate travel distances/times to various opportunities, demonstrating how age is associated with differences in opportunity access.
 School-aged children and older populations are a common focus.
@@
 These indices are generated from government sources or author-informed census data creation methods.
-As an example, the Neighbourhood Equity Index (NEI) is a measure of vulnerability created by the City of Toronto and used in @awuorInfluenceEnvironmentalHealth2019 to examine disparities in premature death.
+As an example, the Neighbourhood Equity Index is a measure of vulnerability created by the City of Toronto and used in @awuorInfluenceEnvironmentalHealth2019 to examine disparities in premature death.
 Other works use national census indicators such as the social and housing deprivation index [@pucciInequalitiesJobrelatedAccessibility2019] or explore transport disadvantage, equity in policy implementation, or transport-related mortality burden by means of census measures (e.g., household poverty) and transport-related accessibility indicators [@aldredEquityNewActive2021; @iungmanImpactUrbanTransport2021; @sunPublicTransportAvailability2021; @scheurerSpatialAccessibilityPublic2017].
-Similarly, Environmental Justice (EJ) indicators have been used in the US literature to identify neighbourhoods that have a higher than average proportion of low-income and non-white populations and evaluate the equity impacts of transportation projects [@rowangouldIdentifyingEnvironmentalJustice2016; @parkTransitParksEnvironmental2021; @reddySafeguardingMinorityCivil2010].
-
-Multi-dimensional considerations are so prevalent in the corpus that "Gender", "Race/ethnicity", "Education", or "Employment" are infrequently studied in isolation.
+Similarly, Environmental Justice (EJ) indicators have been used in the U.S.
+literature to identify neighbourhoods that have a higher than average proportion of low-income and non-white populations and evaluate the equity impacts of transportation projects [@rowangouldIdentifyingEnvironmentalJustice2016; @parkTransitParksEnvironmental2021; @reddySafeguardingMinorityCivil2010].
+
+Multi-dimensional considerations are so prevalent in the reviewed literature that "Gender", "Race/ethnicity", "Education", or "Employment" are infrequently studied in isolation.
 Only a few papers focus exclusively on gendered differences in active transportation [@adlakhaMindGapGender2020, @xieWonCycleRoute2018], race/ethnicity's relationship to green space proximity [@silvaEnvironmentalJusticeAccessibility2018], and culturally-appropriate opportunities [@wangModelingSpatialAccessibility2011].
-Papers that focus *solely* on "Employment" status or "Education" level are completely absent.
-Furthermore, "Other" population groups are also frequently considered in the corpus: this is a catch-all category that includes populations less commonly targeted by research e.g., veterans and access to specific-healthcare needs [@mooneyTravelDistanceBarrier2000], pregnant people and access to maternity services [@vadrevuMeasuringSpatialEquity2016], and youth who live in foster care [@batscheUsingGISEnhance2012].
-Overall, the diversity of population groups considered in the corpus demonstrates the variety of transportation-equity concerns addressed in the literature.
+Papers that focus *solely* on "Employment" status or "Education" level are completely absent in the corpus.
+Furthermore, "Other" population groups are also frequently considered: this is a catch-all category that includes populations less commonly targeted by research e.g., veterans and access to specific-healthcare needs [@mooneyTravelDistanceBarrier2000], pregnant people and access to maternity services [@vadrevuMeasuringSpatialEquity2016], and youth who live in foster care [@batscheUsingGISEnhance2012].
+Overall, the diversity of population groups considered in the corpus demonstrates the variety of transportation-equity concerns addressed.
 
@@
 
-Travel mode, though modifiable, is intertwined with individual identity, posing challenges for equity analysis.
-Here we choose to view the mode of travel primarily as those who use the mode, hence as subjects of justice (i.e., the “who”), but occasionally, this will be enjoined to an object of justice (i.e., a “what” such as inequities in rural transit [@giuffridaAddressingPublicTransport2021]).
-
-Transit characterises the majority of the corpus (@fig-fig1).
-Despite transit being perceived unfavourably by some [e.g., @mellaDrivers2021], it is often viewed as the only and primary mobility option mode for many [@jacquesRescuing2012].
-Papers that focus on transit assess a variety of topics, including food deserts [@mckeyCrowdsourcedMappingHealthy2020] and barrier-free transportation for people who face disabilities [@jimenez-espadaResearchProblemUniversal2021; @liuStatusQuoChallenges2019; @limFacilitatingIndependentCommuting2021; @feeleyValidationParatransitSkills2019].
-We infer that transit is reckoned as a public good and hence a natural object of justice.
-Accordingly, public transportation can be modified to meet the demands of justice, by funding it sufficiently to provide barrier-free transport for most, despite challenges like low densities, fiscal constraints and political will [@markardUnsustainabilities2023].
-
-Transit also plays a central role in multimodal comparisons for transport equity analysis, and often feature in papers dealing with "Walk", "Car", or some other category in @fig-fig1.
+Travel mode, though modifiable, is intertwined with individual identity, presenting challenges for the analysis of disparities.
+In this subsection, we choose to view the mode of travel primarily as those who use the mode, hence as subjects of justice (i.e., the “Who?”).
+However, at times, travel mode will also be considered as an object of justice (i.e., a “What?”, such as disparities in rural transit service).
+
+To begin, transit characterises the majority of the reviewed literature (@fig-fig1), reflecting some intuitive sense.
+Despite transit being perceived unfavourably by some (e.g., @mellaDrivers2021), it is often viewed as the only or primary mobility option mode for many [@jacquesRescuing2012].
+Overall, we can infer that transit is largely considered a public good and, therefore, a natural object of justice.
+As such, public transportation is especially amendable to being modified to meet the demands of justice by ensuring adequate funding to provide barrier-free transport for most, despite challenges such as low densities, fiscal constraints, and political will [@markardUnsustainabilities2023].
+As such, papers that focus on transit assess a variety of related topics, including food deserts [@mckeyCrowdsourcedMappingHealthy2020] and barrier-free transportation for people who face disabilities [@jimenez-espadaResearchProblemUniversal2021; @liuStatusQuoChallenges2019; @limFacilitatingIndependentCommuting2021; @feeleyValidationParatransitSkills2019].
+
+Transit also plays a central role in multimodal comparisons in transport disparities, especially comparing "Walk", "Car", or some other category in @fig-fig1.
 For instance, @brusselAccessAccessibilityCritique2019 compares transit, pedestrian and road network accessibility measurements in the context of SDG 11.2.
-@renneWhatHasAmerica2018 reviews natural disaster evacuation plans, focusing on car-less households and emphasizing transit and pedestrian networks.
+@renneWhatHasAmerica2018 reviews natural disaster evacuation plans, focusing on carless households and emphasizing transit and pedestrian networks.
 A few papers also frame transit as a direct competitor of car travel or use it as a benchmark [@golubUsingPrinciplesJustice2014; @martensJusticetheoreticApproachDistribution2012].
-For instance, @warrenDevelopingEquitableSustainable2015 propose car ownership standards while acknowledging the tension between mobility needs in transit under-served areas and emission reduction goals.
+For instance, @warrenDevelopingEquitableSustainable2015 proposes car ownership standards while acknowledging the tension between mobility needs in transit under-served areas and emission reduction goals.
 However, this framing is not universal, and transit is sometimes seen as a mode to fulfill individual capabilities.
@@
 Pedestrians represent a unique convergence of "what" and "who," utilizing their own bodies for mobility.
-Papers focusing solely on walking often use walkability scores to assess neighbourhood quality [@evansAccessibilityUserNeeds2015], mobility by different demographics [@towneUsingWalkScore2016; @kimNeighborhoodEnvironmentWalkability2016], or urban peripheral regions [@blecicCapabilitywiseWalkabilityEvaluation2021].
+Papers focusing primarily on walking often use walkability scores to assess neighbourhood quality [@evansAccessibilityUserNeeds2015], mobility by different demographics [@towneUsingWalkScore2016; @kimNeighborhoodEnvironmentWalkability2016], or urban peripheral regions [@blecicCapabilitywiseWalkabilityEvaluation2021].
 Some use walkability to gauge public health and urban vitality [@sungResidentialBuiltEnvironment2015; @mccormackRelationshipClusteranalysisDerived2012].
@@
 
-The third most studied mode is "Car".
-Unlike transit and walking, cars are rarely the sole focus of equity research.
-In our corpus they are often critically compared with transit or considered in areas with inadequate transit [@kimmelStructuralBarriersComprehensive2018; @aljoufie2016].
-Upon the rise of automobility, when walking and cycling got in the way of faster travel, transit came to be seen as a social service and sometimes as a hindrance to the full realization of automobility.
-Our assessement of the focus on non-car modes in our corpus is that it represents a reckoning with automobility's legacy and ongoing demands for space, public subsidies, and government supports that underestimate the true cost of cars [@gosslingSocial2019; @timperleyFight2021].
-In addition to comparing "Car" to other modes, some papers reflect on "Other" forms of mobility (@fig-fig1), including wheelchair accessible taxi and/or paratransit services [@marquezFactorsAffectingPersonal2019; @wilkinsonmeyersLiveOrdinaryLife2015], travel on waterways [@vadrevuMeasuringSpatialEquity2016; @caoIndicatorsSocioSpatialTransport2017; @parrySocialVulnerabilityClimatic2018], motorcycle or other micro-mobility [@berryInvestigatingFuelPoverty2016; @savingmothersDidSavingMothers2019; @tiwariDevelopmentPublicTransport2021], or emergency vehicles [@pedigoInvestigationDisparitiesGeographic2010; @patelDeterminingGeographicAreas2007].
+The third most studied mode in the reviewed literature is "Car".
+Unlike transit and walking, these studies rarely feature car as the mode of primary focus.
+Car is often assessed against transit and active modes [e.g., low traffic neighbourhoods @aldredEquityNewActive2021] or in areas with inadequate transit service [@kimmelStructuralBarriersComprehensive2018; @aljoufie2016].
+Under our appraisal, transit and walk modes are of most interest because they represent a reckoning with automobility's legacy and ongoing demands for space, public subsidies and government supports that underestimate their true cost [@gosslingSocial2019; @timperleyFight2021].
+Upon the rise of the automobile, pedestrian spaces were distinguished, non-car modes came to be seen as a hindrance to full automobility, and transit seen as a social service.
+We ascertain that for these reasons, "Car" is often compared against non-car modes and never put forward as the only mode that should be expanded.
+Furthermore, some papers reflect on "Other" forms of mobility (@fig-fig1), including wheelchair accessible taxi and/or paratransit services [@marquezFactorsAffectingPersonal2019; @wilkinsonmeyersLiveOrdinaryLife2015], travel on waterways [@vadrevuMeasuringSpatialEquity2016; @caoIndicatorsSocioSpatialTransport2017; @parrySocialVulnerabilityClimatic2018], motorcycle or other micro-mobility [@berryInvestigatingFuelPoverty2016; @savingmothersDidSavingMothers2019; @tiwariDevelopmentPublicTransport2021], or emergency vehicles [@pedigoInvestigationDisparitiesGeographic2010; @patelDeterminingGeographicAreas2007].
 
@@
 
-### Who is missing in the "who"?
-
-As shown, "who" is the subject of inequities provides ample but not exhaustive coverage of populations.
-But who is missing from the "Who" question?
-There is a need to go beyond the focus on low-income transit riders; positioning intersectionality of socio-economic and mode-use to create community-based definitions of equity and tailored standards are critical.
-How decision makers define an equity-deserving community will impact results e.g., [@rowangouldIdentifyingEnvironmentalJustice2015].
-In this sense, standards need to be sensitive to changing community-based definitions of inequities.
-Are issues of economic-inequity at the root of transport inequities for a specific community?
-Are (dis)abilities?
-Are inequities in the service of transit the focus of a study because it can be improved to address transport inequities and are alternative modes driving these inequities?
-Access to what sorts of opportunities are driving transport inequities?
-How do populations, transit modes, and opportunities sought intersect to define the "who" of inequities?
-A community- informed understanding of inequities and tracking how it changes are needed.
-
-## "What" are the objects of transportation justice {#sect3.3}
-
-The most fundamental benefit of transportation systems is mobility (enabling or impeding movement).
+## "What" could be the objects of transportation justice
+The objects of fairness assessment, the "What", are often motivated by mobility or accessibility.
+A fundamental benefit of transportation systems is mobility (enabling or impeding movement).
 This benefit is sometimes valued by itself but is often seen by the literature as instrumental to achieve an ulterior goal (e.g., activity participation and associated benefits).
@@
 
-Most papers in our corpus take a broad approach, with `r (dest_count$count[8]/165) |> percent()` focusing on "Unspecified" destinations (@fig-fig1).
+Most papers in the reviewed literature take a broad approach, with `r (dest_count$count[8]/165) |> percent()` focusing on "Unspecified" destinations (@fig-fig1).
 They examine various equity dimensions across different transportation modes.
@@
 
-In terms of destinations, "Healthcare" services (`r (dest_count$count[6]/165) |> percent()`) and "Employment" (`r (dest_count$count[5]/165) |> percent()`) are the most common in our corpus (@fig-fig1).
-Papers on healthcare often highlight disparities in services, like @wangModelingSpatialAccessibility2011, who assess access to Mandarin-speaking family physicians for Chinese speakers in Toronto.
+In terms of destinations, "Healthcare" services (`r (dest_count$count[6]/165) |> percent()`) and "Employment" (`r (dest_count$count[5]/165) |> percent()`) are the most common in the reviewed literature (@fig-fig1).
+Papers on healthcare often highlight disparities in services, like @wangModelingSpatialAccessibility2011, who assess access to Mandarin-speaking family physicians for Mainland Chinese immigrants in Toronto.
 Similarly, papers focusing on employment are often aimed at identifying transportation-poor neighbourhoods [@allenSizingTransportPoverty2019a; @churchillTransportPovertySubjective2019].
@@
 Fewer papers cover "Community" destinations (e.g., public service centres, places of community support or worship) (`r (dest_count$count[2]/165) |> percent()`) or "Childcare" (`r (dest_count$count[1]/165) |> percent()`), but they are integral to holistically study activity participation [@albertsRebuildingWomenLivelihoods2016; @smithAccessibilityCapabilityMinimum2012].
-The lack of information about community destinations, especially for children, is noticeable [@desjardinsChildren2022].
-
-## "How" is fairness determined {#sect3.4}
+The lack of information about community destinations in the reviewed literature, especially for children, is noticeable [@desjardinsChildren2022].
+
+## Standards: "How" fairness is measured
 
@@
 
-The conceptual foundations of fairness are not always made explicit by researchers and often had to be inferred from the corpus.
-Some categories emerged from our review and we must note that they are not exhaustive.
-In broad terms, they are (@fig-fig2 details definitions):
+The conceptual foundations of fairness are not always made explicit within the reviewed literature, they often had to be inferred.
+Though not exhaustive, the following categories emerged from our review and in broad terms are (@fig-fig2 details definitions):
 
@@
 
-- Transportation related social exclusion (`r df_conc|> group_by(conc) |> summarise(total = n()) |> subset(conc == "TR exclusion", select = "total") |> mutate(total = total/165) |> pull(total)  |> percent()`)
+- Transport-related social exclusion (`r df_conc|> group_by(conc) |> summarise(total = n()) |> subset(conc == "TR exclusion", select = "total") |> mutate(total = total/165) |> pull(total)  |> percent()`)
 
@@
 
-A slice of the corpus is supported by broader "Rights" conceptualisations of fairness: these papers often focus on equity for people with disabilities or non-car users and associated challenges accessing transport infrastructure [@jimenez-espadaResearchProblemUniversal2021; @daamenAssessingGapPublic2008; @bharathyRevisitingClearFloor2018].
-While many papers are underpinned with the right to the city (the _right_ to participate in the production of urban space [@lefebvreDroitVille1967]) [@adliRightCityApplying2018], others emphasize legal _Rights_ like ADA regulations [@bharathyRevisitingClearFloor2018] or the goal of _access for all_ in land-use transportation master plans [@limFacilitatingIndependentCommuting2021].
-
-In another subset of the corpus, distributions are examined through concepts of "Horizontal" and "Spatial equity", often using quantitative methods to assess distributional disparities without explicit justice rationales.
+A slice of the reviewed literature is supported by broader "Rights" conceptualisations of fairness: these papers often focus on equity for people with disabilities or non-car users and associated challenges accessing transport infrastructure [@jimenez-espadaResearchProblemUniversal2021; @daamenAssessingGapPublic2008; @bharathyRevisitingClearFloor2018].
+While many papers are underpinned with the right to the city (the _right_ to participate in the production of urban space [@lefebvreDroitVille1967; @adliRightCityApplying2018]), others emphasize legal _Rights_ like regulations aligned with The Americans with Disabilities Act (ADA) [@bharathyRevisitingClearFloor2018] or the goal of _access for all_ in land-use transportation master plans [@limFacilitatingIndependentCommuting2021].
+
+In another subset of the reviewed literature, distributions are examined through concepts of "Horizontal" and "Spatial equity", often using quantitative methods to assess distributional disparities without explicit justice rationales.
 Examples include setting travel impedance thresholds [@shenMeasuringSpatialAccessibility2020] and mapping accessibility indices spatially across populations [@monzonEfficiencySpatialEquity2013] or population-groups [@sharmaPublicTransitAccessibility2021].
@@
 Some papers use physical activity guidelines and surveys to understand the effect of active transportation infrastructure [@adlakhaMindGapGender2020; @auchinclossDesignBaselineDescription2020; @mccormackRelationshipClusteranalysisDerived2012].
-Mixed or qualitative methods combined with health-related outcome standards (e.g., travel times for emergency treatment [@savingmothersDidSavingMothers2019], premature mortality [@awuorInfluenceEnvironmentalHealth2019]) lead to more concrete statements of fairness.
+Mixed or qualitative methods combined with health-related outcome standards lead to more concrete statements of fairness e.g., travel times for emergency treatment [@savingmothersDidSavingMothers2019] or premature mortality [@awuorInfluenceEnvironmentalHealth2019].
 
@@
 Here, the objects of justice are often clearly identified and link standards to tangible welfare-informed outcomes.
-They focus on groups from perspectives of disadvantage such as: social exclusion and transport poverty [@allenSizingTransportPoverty2019a; @delboscTransportProblemsThat2011; @churchillTransportPovertySubjective2019]], food deserts [@mckeyCrowdsourcedMappingHealthy2020], and energy poverty [@robinsonDoubleEnergyVulnerability2020; @berryInvestigatingFuelPoverty2016; @berryDistributionalEffectsCarbon2019].
+They focus on groups from perspectives of disadvantage such as: social exclusion and transport poverty [@allenSizingTransportPoverty2019a; @delboscTransportProblemsThat2011; @churchillTransportPovertySubjective2019], food deserts [@mckeyCrowdsourcedMappingHealthy2020], and energy poverty [@robinsonDoubleEnergyVulnerability2020; @berryInvestigatingFuelPoverty2016; @berryDistributionalEffectsCarbon2019].
 
@@
 Notably, many papers with "Opportunity" standards consider multiple dimensions, employing similar methods but tailored to different focal points.
-For instance @peungnumsaiGridBasedSpatialAnalysis2020 suggest service benchmarks of equal supply and demand of transit, revealing "Horizontal equities" as well.
+For instance, @peungnumsaiGridBasedSpatialAnalysis2020 suggest service benchmarks of equal supply and demand of transit, revealing horizontal inequities as well.
 Others conceive the externalities of transportation system as trade-offs and aim to maximize transport-related benefits (i.e., time savings, emissions reductions, congestion reductions, user fares) through optimization/location-allocation methodologies [@fakhrmoosaviIncorporatingTravelTime2021; @zhengAreabasedEquitablePricing2020; @wismadiSpatialPreferenceModelling2014].
@@
 "Population" standards are often founded on "Well-being" conceptualisations from a variety of socio-demographic and spatial angles.
-Methods include: establishing thresholds based on questionnaires and comparisons to recommended physical activity levels [@auchinclossDesignBaselineDescription2020; @mccormackRelationshipClusteranalysisDerived2012; @kimNeighborhoodEnvironmentWalkability2016; @timperioSafetyNumbersDoes2015], region-relative comparisons in health outcomes in a spatial unit such as premature mortality rates [@awuorInfluenceEnvironmentalHealth2019], spatial access benchmarks based on population-related characteristics like supermarket access [@murphySupermarketAccessTransport2017] and hospital access @pereiraGeographicAccessCOVID192021], summative per capita benchmarks such as decent living energy consumption levels [@raoDecentLivingEmissions2012], and community-informed spatial boundaries like EJ defined communities [@rowangouldIdentifyingEnvironmentalJustice2016].
+Methods include: establishing thresholds based on questionnaires and comparisons to recommended physical activity levels [@auchinclossDesignBaselineDescription2020; @mccormackRelationshipClusteranalysisDerived2012; @kimNeighborhoodEnvironmentWalkability2016; @timperioSafetyNumbersDoes2015], region-relative comparisons in health outcomes in a spatial unit such as premature mortality rates [@awuorInfluenceEnvironmentalHealth2019], spatial access benchmarks based on population-related characteristics like supermarket access [@murphySupermarketAccessTransport2017] and hospital access [@pereiraGeographicAccessCOVID192021], summative per capita benchmarks such as decent living energy consumption levels [@raoDecentLivingEmissions2012], and community-informed spatial boundaries like EJ defined communities [@rowangouldIdentifyingEnvironmentalJustice2016].
 While most of these papers use quantitative or mixed-methods approaches, some employ exclusively qualitative methods [@berheAdaptationDissonanceQuality2014].
@@
 
-"Environmental+" standards feature less prominently in our corpus, with studies that predominantly examine "inequitable externalities", possibly because the environmental burdens of transportation are addressed more broadly in other literature (e.g., environmental justice).
+"Environmental+" standards are featured less prominently in the reviewed literature, possibly because the environmental burdens of transportation are addressed more broadly in other literature (e.g., environmental justice).
 However, the papers included in our corpus present some interesting insights.
-They often use traffic-related air and noise pollution, green space, urban design, urban air temperature, health outcomes, and physical activity guidelines to assess transport-related externalities.
+They often use traffic-related air and noise pollution, green space, urban design, urban air temperature, health outcomes, and physical activity guidelines to assess transport-related "inequitable externalities".
 Methods employed are primarily quantitative or mixed-methods, identifying inequalities through Gini coefficients [@fengTradeoffsMobilityEquity2014] or composite indices [@agostfelipInclusiveModelAssessing2021; @mirandaBenchmarkingSustainableUrban2012; @corazzaMethodologyEvidenceCase2020], occasionally incorporating spatial analysis [@jephcoteGeospatialAnalysisNaturally2013; @carrierApplicationThreeMethods2014].
@@
 Nonetheless, their use provides interpretable values for tracking progress, offering comparability across communities, unlike accessibility measures, which vary in methods and assumptions.
-
-# Moving forward: calls for action {#sect4}
-
-This paper makes two contributions to the literature.
-It outlines a conceptual and flexible "Why", Where", "When", "Who", "What" and "How" (5WH) framework with supporting definitions for approaching questions about transportation justice.
-It also applies the framework in the synthesis and appraisal of 165 academic articles drawn from the transportation justice literature; these studies use equity analysis as instruments to gauge proximity to their context's just situation.
-
-In brief, justice is an end objective and standards are definitive statements of fairness.
-As such, they are essential to assess proximity to the objective.
-Context-specific evidence of transport inequities is necessary to acknowledge injustices and inspire creative interventions to advance the cause of justice.
-Armed with appropriate knowledge, policy can be designed to address inequities while remaining adaptable to the evolving demands of justice.
-Flexible frameworks are crucial.
-What is (in)formally considered (in)essential today, could change during the next pandemic or technological revolution.
-Creating adaptable transport justice frameworks that lend themselves to change is required to ensure resiliency to backsliding.
-From the diverse corpus reviewed in this paper, we argue for the necessity of setting standards from a systems approach, considering both _positive_ rights (e.g., a right to have access to sufficient quality of essential services) and _negative_ rights (e.g., mobility of cars must be limited to reduce their impact on air quality, health, safety, and other positive rights).
-
-We conclude with the following five calls to action.
-
-## Call 1: Develop standards based on rigorous concepts of justice {#sect4.1}
+# Moving forward: discussion and calls for action
+
+This work makes two contributions to the literature.
+First, it outlines a conceptual and flexible "Why", Where", "When", "Who", "What" and "How" (5WH) framework with supporting definitions for approaching questions about transportation justice.
+Second, it applies the framework in the synthesis and appraisal of 165 academic articles drawn from the transportation justice literature; these studies use equity analysis as instruments to gauge proximity to their context's just situation.
+We summarise what is in the literature in the preceding section using the 5WH framework as appropriate, the following is a discussion on what may be missing from the literature:
+
+- **"Who?"**: While who is subject of inequities in the literature is extensive, dimensions are still missing.
+There is a need to go beyond the focus on low-income transit riders; positioning intersectional considerations within the creation of community-based equity definitions and tailored standards are critical, as the act of delineating equity-deserving communities impacts results e.g., [@rowangouldIdentifyingEnvironmentalJustice2015].
+Standards should be sensitive to evolving community-based definitions of inequities.
+Are issues of economic inequity at the root of transport inequities for a particular community?
+Are inequities in access provided by certain modes emphasised because access offered by other modes are _relatively_ better?
+Access to what types of opportunities are driving transport inequities?
+How do populations, travel modes, and opportunities sought intersect to define the "who" of inequities?
+A community-informed understanding of inequities and tracking how it changes are needed.
+There is also a need for further intersectional considerations, as clear "Who?"-specific standards are lacking.
+
+- **"What?"** and **"How?"**: The connection of what is the focus of inequity measure to some sort standard.
+Specifically, "What?" are the benefits and burdens of a transportation system that could link accessibility or mobility to some sort of tangible fairness object of justice.
+Could it be some environmental externalities (eg.), or human-health externalities (eg.), or traffic related safety (eg.), or multi-criteria mix (eg.).
+We believe that all forms of these standards, when linked to tangible outcomes can be used to track "fairness" in a region and can be used to track multi-dimensional progress towards justice depending on how robustly these standards are justified.
+Justification is exceptionally important, and we believe that a lot of the literature reviewed may miss this mark.
+However, literature that does not, usually connects to qualitative results and/or operationalizes mixed-methods.
+
+- **"Why?"** is not within the scope of this review.
+Also, **"When?"** and **"Where?"** are only reviewed from a cursory perspective.
+These dimensions were not clear enough in the literature to be explicitly commented on.
+
+Based on the preceding sections of what is in the literature, and what is lacking from our perspective, we conclude with the following five calls to action.
+
+## Call 1: Develop standards based on rigorous concepts of justice
 
@@
 
-## Call 2: Develop creative methods for systems-thinking approaches to fairness {#sect4.2}
+## Call 2: Develop creative methods for systems-thinking approaches to fairness
 
@@
 Concepts and standards are often discussed from purely qualitative or quantitative approaches.
-This is a missed opportunity to combine the strengths of both approaches, whether by deep diving into some particular experiences or perceptions through qualitative methods or tailoring more meaningful quantitative analysis after qualitative explorations.
-As mixed-method examples, @xieWonCycleRoute2018 find through interviews and go-alongs with women cyclists that the standard Cycling Level of Service (CLS) tools used by engineers to plan cycling infrastructure misses a critical gendered perspective.
+This is a missed opportunity to combine the strengths of both approaches, whether by diving deeper into some particular experiences or perceptions through qualitative methods or tailoring more meaningful quantitative analysis after qualitative explorations.
+As mixed-method examples, @xieWonCycleRoute2018 find through interviews and go-alongs with women cyclists that the standard Cycling Level of Service tools used by engineers to plan cycling infrastructure misses a critical gendered perspective.
 Further, @somenahalliAgingTransportMobility2007 survey older adults to understand their mobility issues, revealing factors that are unseen in standard daily travel surveys.
@@
 How low must an accessibility index be before it is too low?
-Rellatedly, once sufficient accessibility levels are reached, excessively high accessibility values can result in high inequalities.
+Relatedly, once sufficient accessibility levels are reached, excessively high accessibility values can result in high inequalities.
 Are inequalities of a certain level an issue?
@@
 These questions usually remain unanswered in the literature despite their importance.
-These measurements can also bring some challenges and pitfalls as summarized by @karnerAdvancesPitfallsMeasuring2024 but are necessary for more effective equity analysis.
-
-## Call 3: Making data available is a matter of justice {#sect4.3}
-
-In our review of the literature, we were left wondering specifically--what are the motivations for the use of some destinations and the choice to not include others?
-For instance, leisure destinations (e.g., green space, parks, recreation) are infrequently studied and some categories are missing altogether.
-We suspect that since methods used are predominately quantitative, the reliance on commonly used point of interest databases is also high.
-These databases typically include education, health and occasionally include aggregated categories for leisure and community, but are less generous in terms of 'community' and 'leisure' points of interest (community organizations, government services, care-giving destinations).
-Further, categories like childcare are often missing, typically daycare or facilities--domestic work, mobilities of care, and mobility interdependence.
-Not coincidentally, these categories are underrepresented compared to work and healthcare destinations.
-
-We can only track what we have collected and compiled; and we know that transport systems are more than just facilitators to work or as a source of economic development (though in underdeveloped regions, transport systems as a force of economic development is pronounced, e.g., high-speed rail [@chenImpactHighspeedRail2017; @monzonEfficiencySpatialEquity2013; @kimImpactsHighspeedRail2015]).
+These measurements can also bring some challenges and pitfalls as summarised by @karnerAdvancesPitfallsMeasuring2024 but are necessary for more effective equity analysis.
+
+## Call 3: Making data available is a matter of justice
+
+In our review of the literature, we were left wondering specifically-- what are the motivations for the inclusion of some destinations and the choice to not include others?
+For instance, leisure (e.g., green space, parks, recreation) and certain care destinations (e.g., community organizations, government services, banking locations, daycares) are infrequently studied or missing altogether.
+We suspect that since the methods used in the reviewed literature are predominately quantitative, the reliance on commonly used point of interest databases is also high.
+These databases typically include education, health and occasionally include aggregated categories for leisure and community destinations, but are less generous for points of interest that are more often associated with understudied mobility patterns.
+
+However, through the reviewed literature, we know that transport systems are more than just facilitators to work or as a source of economic development (though in underdeveloped regions, transport systems as a force of economic development is pronounced, e.g., high-speed rail [@chenImpactHighspeedRail2017; @monzonEfficiencySpatialEquity2013; @kimImpactsHighspeedRail2015]).
 As such, data availability matters.
@@
 
-The calls for and relevant issues of data availability are not new, but they have at least three parts.
+The calls for and relevant issues of data availability are not new, but they have at least three pieces.
 What and who is the subject of justice?
@@
 Deciding who is the subject of justice frames the data collection activity; if it is the mobility of those that do domestic work, the classification of who does this work and how it changes over time/space is fundamental.
-How we classify mobility and access has implications for our understanding of just who is a subject of justice, as illustrated by the history of racial classifications in the US [@leeRacial1993].
+How we classify mobility and access has implications for our understanding of just who is a subject of justice, as illustrated by the history of racial classifications in the U.S.
+[@leeRacial1993].
 The methods we use, the spatio-temporal boundaries for data collection, and who has (and does not have) access to the collected data, can all impact how we think about justice.
@@
 
-## Call 4: Develop more direct and explicit links between standards and lived experiences {#sect4.4}
-
-Robust assessments of the implications of equity standards on lived experiences are still lacking in the literature, but are essential for equity analysis that translates into just practice.
+## Call 4: Develop more direct and explicit links between standards and lived experiences
+
+Robust assessments of the implications of standards on lived experiences are still lacking in the literature, but are essential for equity analysis that translates into just practice.
 While the estimation of the benefits of increased mobility or accessibility, or reducing burdens is commonplace in the literature, there is a need to associate them to outcomes, including life and neighbourhood satisfaction, subjective well-being, mental and physical health, and social capital.
@@
 Composite measures such as the transport-land-use index proposed by @appleyardTransitCorridorLivability2017 is an example of systems-thinking approach that links findings to quality-of-life proxies; they ground their measure in the principle of _livability_ along corridors of varying levels of estimated transport-land-use integration.
-However, the methods could go further; they could be tied to absolute goals of integration or livability.
+However, the methods could go further: they could be tied to absolute goals of integration or livability.
 Relatably, @higgsUrbanLiveabilityIndex2019 develops an urban livability index and demonstrates the relationship between the population's use of a certain mode in a neighbourhood and a one unit increase in the index.
@@
 
-## Call 5: Rigorously evaluate interventions and policies {#sect4.5}
+## Call 5: Rigorously evaluate interventions and policies
 
 There is a need to evaluate equity interventions and policies: track their before, during, and after.
-In this review, barely 10% of studies assess specific transport-related policy interventions through an equity lens.
+In this review, a small proportion of studies assess specific transport-related policy interventions through an equity lens.
 Examples include mode-shift from driving to active school travel [@mammenSchoolTravelPlanning2014], transit fare restructures [@hickeyUsingQuantitativeMethods2010] and spatial analysis of Low Traffic neighbourhoods [@aldredEquityNewActive2021].
@@
 
-# Declarations
-
-During the preparation of this work, the first and second authors used ChatGPT exclusively to reduce word count in the section "An appraisal of the lay of the land" while keeping the original writing style.
-After using this tool, all authors reviewed the manuscript and take full responsibility for the content of the publication.
-
-\newpage
-
 # References
@@
 :::
+
+# Acknowledgements and declarations
+
+This work would not have been possible without the help and inspiration of many individuals not listed as authors.
+In the early stages of this report, we received invaluable help from specialists at the University of Toronto as we learned about various approaches to literature reviews, particularly the Research Services and Liaison Librarians.
+These experts provided guidance in cornering down the topic, identifying guidance material on literature synthesis and relevant databases, and eased us into the world of knowledge synthesis.
+Though our path diverged from a traditional scoping review, the protocol published in September 2022 (found [here](https://osf.io/rsb92)) was followed with only minor changes, with only a reduction in the databases searched due to resource constraints.
+We also wish to acknowledge the many hours of work by undergraduate and graduate research assistants who poured through title and abstract screening to help determine full-text eligibility at the earliest stages of our search.
+The volume of literature initially reviewed was greater than any of us had experienced before.
+For this support, we are deeply thankful to Jan Domalaon, Dancel Gayle, Eryn Maloney, Mahdis Moghadasi, Anika Munir, and Subaita Refaaf.
+
+**Competing interests’ declaration:** The authors declare there are no personal or financial conflicts of interest.
+
+**Funding declaration:** This work was supported by the Mobilizing Justice Partnership and the Canada Graduate Scholarships—Doctoral Program from the Social Sciences and Humanities Research Council of Canada (SSHRC).
+
+**Ethical approval declaration:** Not applicable.
+
+**Participate consent declaration:** Not applicable.
+
+**Competing interests’ declaration:** The authors declare there are no personal or financial conflicts of interest.
+
+**Technology declaration:** During the preparation of this work, the first and second authors used ChatGPT to assist in reducing the word count in an early version of the “An appraisal of the lay of the land” section.
+After using this tool, all authors reviewed the manuscript and take full responsibility for the content of the publication.
 
@@
 
-## Evidence search strategy {#sect5-2}
+## Evidence search strategy {#sec-sect61}
 
@@
 #| label: fig-figA1
-#| fig-cap: "PRISMA flow diagram for the evidence selection process. ES signifies equity standard and EC signifies equity conceptualization."
+#| fig-cap: "PRISMA flow diagram for the evidence selection process. ES signifies standards and EC signifies conceptualisations."
 knitr::include_graphics("figures/Pflow_chart.tiff")
@@
 This inclusion criteria is reflected in the second and third set of topic search terms developed in the search strategy.
-These terms relate to types of equity dimensions (e.g., "accessibility" OR "mobility" or "transport-related air pollution" - see blue text in the @fig-figA2 for the full list) and equity conceptualizations (e.g., "Justice" OR "equity" - see purple text in @fig-figA2 for the full list).
--   **Context**: the included studies should also be limited to publications that include equity standards.
-Context can be more difficult to explicitly search for with key terms so synonyms for 'standards' were added to the query as a four set of topic search terms (e.g., threshold, indicator, criteria - see orange text in @fig-figA1 for full list).
+These terms relate to types of equity dimensions (e.g., "accessibility" OR "mobility" or "transport-related air pollution" - see blue text in the @fig-figA2 for the full list) and fairness conceptualizations (e.g., "Justice" OR "equity" - see purple text in @fig-figA2 for the full list).
+-   **Context**: the included studies should also be limited to publications that include types of standards.
+Context can be more difficult to explicitly search for with key terms so synonyms for 'standards' were added to the query as a four set of topic search terms (e.g., threshold, indicator, criteria - see orange text in @fig-figA2 for full list).
 Additionally, journal article and conference papers, English-language literature from any country, any study design (e.g., quantitative, qualitative, or mixed-method studies, or conceptual frameworks), and any record published within the past 30 years are included (January 1992 to March 2022).
-The time period is selected as the first (to the authors knowledge) peer-reviewed article which operationalized equity standards and equity conceptualization was published in 1996 [@khistyOperationalizingConceptsEquity1996]; we are broadening the search by a few years for completeness.
+The time period is selected as the first (to the authors knowledge) peer-reviewed article which operationalized standards and fairness conceptualization was published in 1996 [@khistyOperationalizingConceptsEquity1996]; we are broadening the search by a few years for completeness.
 English is selected as it is the common language spoken across the authorship team.
@@
 
-## Evidence selection and data extraction {#sect5-3}
+## Evidence selection and data extraction {#sec-sect62}
 
@@
 
-The following steps summarize the evidence selection process:
+The following steps summarise the evidence selection process:
 
@@
 In the last step, a data extraction template for each record was filled by one reviewer (purple box in @fig-figA3).
-The data extraction template was created with the aim of striking a balance between the complexity of categories and the simplicity of summary; information related to "Where?" (the geographical context and sphere of life), "When?" (temporal circumstances for the application of justice), "Who?" (the subject of justice), "What?" (the object of justice), and "How?" (equity standard(s) and concepts) was filled out for each study.
+The data extraction template was created with the aim of striking a balance between the complexity of categories and the simplicity of summary; information related to "Where?" (the geographical context and sphere of life), "When?" (temporal circumstances for the application of justice), "Who?" (the subject of justice), "What?" (the object of justice), and "How?" ( standards and conceptualisations) was filled out for each study.
 The following table contains the template that was input into *Covidence* and used throughout.
@@
                            "Analyses how affordable urban public transportation is in select Latin American and Caribbean countries.
-They look at the estimated average monthly cost of transit trips and average monthly household income and conceptualize **transportation-related** **affordability**, especially for the most economically vulnerable (**vertical equity**).",
+They look at the estimated average monthly cost of transit trips and average monthly household income and conceptualize **transport-related** **affordability**, especially for the most economically vulnerable (**vertical equity**).",
                            "How?: The financial burden of a basket of urban public transportation trips (60 trip fares, representing 30 round-trips per month) should not exceed 10% of household monthly income.",
@@
                            "Where? [@agostfelipInclusiveModelAssessing2021] - Europe (Castellon, Spain)",
-                           "Conceptualizes equity through age-friendly urban spaces that reduce (and eliminate) conditions for **transportation-related social exclusion** for older populations and prioritize those who are economically vulnerable (**vertical equity**).
+                           "Conceptualizes equity through age-friendly urban spaces that reduce (and eliminate) conditions for **transport-related social exclusion** for older populations and prioritize those who are economically vulnerable (**vertical equity**).
 These guidelines are inspired by the SDGs in addition to planning guidelines used national, regional, and local guidelines used in Spain.",
@@
 :::
+
+<!-- OLD ABSTRACT
+
+Transportation systems are technologies essential for social inclusion and activity participation, and therefore tracking inequalities is of interest to many.
+Our work responds by providing a synthesis of the state of the knowledge in the academic transportation fairness, justice, and equity related literature.
+First, we outline a flexible framework for engaging questions of transportation fairness and equity analysis based on the questions "Why?", "Where?", "When?", "Who?", "What?", and "How?" (5WH).
+Second, we apply the framework to collate findings from a broad scoping review of the transportation literature.
+In contrast to previous reviews on measuring inequality in transportation systems, our work is concerned with the implicit or explicit standards used to judge whether inequalities are fundamentally “fair” or unacceptable.
+We find the majority of work is published within the last half decade (55%) and centers the urban context (85%); the minority of work explores Global South case studies (40%) which focus disproportionally on transport financial affordability as well as informal transportation systems and infrastructure development needs relative to studies from the Global North.
+Income groups, followed by specific age groups (i.e., elders) and people with disabilities are most commonly the subjects of justice (the Who) as they experience disporportionate transport-burdens.
+What mobility tools are studied are most often transit or pedestrian modes; with many analyses being multimodal and comparative in nature.
+The benefits and burdens, the "What" of mobility, can be summarised as a focus on movement (i.e., trips taken, quality of trips) or the potential for movement (i.e., accessibility).
+How the reviewed works conceptualise fairness is delineated most often with a cut-off opportunity standard (37%) (i.e., number of parks within a 30-minute travel time) or population standard (36%) (i.e., bottom income quartile), though Infrastructure (i.e., level of service) and Environmental+ (i.e., air pollution) are also prominent.
+How the works conceptualise fairness is varied and associated with what type of standard is proposed, but often applies Vertical Equity, Transport-Related social exclusions, and Wellbeing overall with Population and Infrastructure standards being supported by Rights-based and Inequitable Externalitiy conceptualisations and Opportunity standards with Horizontal equity and Spatial equity.
+We identify ways forward to continue progress within the transportation justice literature by synthesizing five calls for action directed to researchers, practitioners, decision-makers and transportation advocates.
+-->

### 2025-02-17 just-transportation/just-transportation.qmd
New definitions; revised manuscript. For Nacho review

@@
 subtitle: "Searching for standards of fairness"
-format:
-  mj-report-pdf: 
-    keep-tex: true
-doctype: "Report (Draft)"
+# format:
+#   mj-report-pdf: 
+#     keep-tex: true
+# doctype: "Report (Draft)"
 
 # Author information. You can add or delete authors here
-author:
-  - name:
-      given: Anastasia
-      family: Soukhov
-    affiliations: 
-      - name: McMaster University
-        department: School of Earth, Environment and Society
-    email: soukhoa@mcmaster.ca
-  - name:
-      given: Ignacio
-      family: Tiznado-Aitken
-    affiliations: 
-      - name: University of Toronto
-        department: Department of Geography and Planning
-    email: i.tiznadoaitken@utoronto.ca  
-  - name:
-      given: Matthew
-      family: Palm
-    affiliations: 
-      - name: The University of North Carolina at Chapel Hill
-        department: Department of City and Regional Planning
-    email: palmmatt@unc.edu  
-  - name:
-      given: Steven
-      family: Farber
-    affiliations: 
-      - name: University of Toronto
-        department: Department of Geography and Planning
-    email: steven.farber@utoronto.ca  
-  - name:
-      given: Antonio
-      family: Páez
-    affiliations: 
-      - name: McMaster University
-        department: School of Earth, Environment and Society
-    email: paezha@mcmaster.ca
+# author:
+#   - name:
+#       given: Anastasia
+#       family: Soukhov
+#     affiliations: 
+#       - name: McMaster University
+#         department: School of Earth, Environment and Society
+#     email: soukhoa@mcmaster.ca
+#   - name:
+#       given: Ignacio
+#       family: Tiznado-Aitken
+#     affiliations: 
+#       - name: University of Toronto
+#         department: Department of Geography and Planning
+#     email: i.tiznadoaitken@utoronto.ca  
+#   - name:
+#       given: Matthew
+#       family: Palm
+#     affiliations: 
+#       - name: The University of North Carolina at Chapel Hill
+#         department: Department of City and Regional Planning
+#     email: palmmatt@unc.edu  
+#   - name:
+#       given: Steven
+#       family: Farber
+#     affiliations: 
+#       - name: University of Toronto
+#         department: Department of Geography and Planning
+#     email: steven.farber@utoronto.ca  
+#   - name:
+#       given: Antonio
+#       family: Páez
+#     affiliations: 
+#       - name: McMaster University
+#         department: School of Earth, Environment and Society
+#     email: paezha@mcmaster.ca
 
@@
 bibliography: bibliography.bib
+
+format: docx
 ---

### 2025-03-09 Journal-manuscript/Journal-manuscript.qmd
Integrated Nacho's comments

@@
 title: "Searching for standards of fairness in the transportation literature"
-format: docx
 csl: apa7.csl
-#   tandf-pdf:
-#     keep-tex: true
-#   tandf-html: default
-# author:
-#   - name: Anastasia Soukhov
+author:
+   - name: AAA #Anastasia Soukhov
     # affiliations:
@@
     # email: soukho@mcmaster.ca
-#   - name: Ignacio Tiznado-Aitken
+   - name: BBB #Ignacio Tiznado-Aitken
     # affiliations:
@@
     # email: i.tiznadoaitken@utoronto.ca
-#   - name: Matthew Palm
+   - name: CCC #Matthew Palm
     # affiliations:
@@
     # email: palmmatt@unc.edu
-#   - name: Steven Farber
+   - name: DDD #Steven Farber
     # affiliations:
@@
     # email: steven.farber@utoronto.ca
-#   - name: Antonio Páez
+   - name: EEE #Antonio Páez
     # affiliations:
@@
 #     country: USA
-# keywords:
-  # - transportation
-  # - equity
-  # - literature review
 abstract: |
-
-keywords: [transportation; equity; justice; standards; transportation planning; fairness; literature review]
+  This work provides a synthesis of how transportation fairness, justice and equity literature has operationalized standards.
+We first outline a flexible framework for engaging fairness questions and then apply it to collate the literature review's results.
+We find most articles were published within the last half decade and center urban areas, with 40% exploring Global South case studies.
+Income groups, followed by specific age groups (i.e., older adults) and people with disabilities, are most commonly the subjects of justice.
+Transit and pedestrian modes are the most frequently studied mobility tools; with many analyses being multimodal and comparative in nature.
+The benefits and burdens, the "What" of mobility, focuses on movement (i.e., quality of trips taken) or the potential for movement (i.e., accessibility).
+How the reviewed works conceptualize fairness is varied (e.g., Vertical Equity, Wellbeing, Rights-Based); but is often delineated with a supporting opportunity standard (37%) (i.e., number of parks within a 30-minute travel time) or population standard (36%) (i.e., bottom income quartile), though Infrastructure (i.e., level of service) and Environmental+ (i.e., air pollution) standards are also prominent.
+While the reviewed literature is extensive, we conclude with calls to action for researchers, practitioners, and transportation advocates to drive meaningful change.
+These include developing standards based on rigorous conceptualizations, advancing systems-thinking approaches to fairness, ensuring data accessibility as a matter of justice, strengthening the connection between standards and lived experiences, and rigorously evaluating interventions and policies.
+
+format: 
+  #docx
+  elsevier-pdf:
+    keep-tex: true
+    include-in-header:
+      text: |
+        \usepackage{graphicx}
+        \usepackage{unicode-math}
+        \usepackage{hyperref}
+        \def\tightlist{}
+        \usepackage{setspace}
+        \doublespacing
+        \usepackage{lineno}
+        \linenumbers
+    journal:
+      name: Travel Behavior and Society
+      formatting: preprint
+      model: 3p
+      cite-style: authoryear
+keywords: [transportation; equity; justice; fairness; standards; transportation planning; literature review]
 bibliography: bibliography.bib
-filters:
-  - docx-landscape.lua
+
+#filters:
+#  - docx-landscape.lua
 ---
@@
   pop=unlist(df_pop)) |> 
-  mutate(pop = ifelse(pop == "Vulnerable populations (composite)", "Comp. vul", pop))
+  mutate(pop = ifelse(pop == "Vulnerable populations (composite)", "Composite vulnerability", pop))
 
@@
   summarise(count = n())|>
-  mutate(sub_ob_j = "1.\nWhen?",
+  mutate(sub_ob_j = "When?", #1
          count = count/165) |>
@@
                                           sum(case_cont_count[4:6,2], 2))) |>
-  mutate(sub_ob_j = "2.\nWhere?",
+  mutate(sub_ob_j = "Where?", #2
          count = count/165) |>
@@
 # THE WHO'S ARE MODE AND POPULATION GROUP
-df_mode_2 <- df_mode |> mutate(sub_ob_j = "3.b Who?\n(mode)") |> 
+df_mode_2 <- df_mode |> mutate(sub_ob_j = "Who?\n(Mode)") |> #3b
   rename("sub_ob_j_type" = "mode")
 
-df_pop_2 <- df_pop |> mutate(sub_ob_j = "3.a Who?\n(pop group)")  |>
+df_pop_2 <- df_pop |> mutate(sub_ob_j = "Who?\n(Population group)")  |> #3a
   rename("sub_ob_j_type" = "pop") |> select(-c("group"))
@@
 # THE WHAT IS ARE DESTINATION group
-df_dest <- dest_count |> mutate(sub_ob_j = "4.\nWhat?",
+df_dest <- dest_count |> mutate(sub_ob_j = "What?", #4
          count = count/165) |> 
@@
   mutate(
-    sub_ob_j_type = ifelse(sub_ob_j_type=="Unspecified" & sub_ob_j == "3.b Who?\n(mode)", "Unspecified 3.b",sub_ob_j_type),
-    sub_ob_j_type = ifelse(sub_ob_j_type=="Unspecified" & sub_ob_j == "4.\nWhat?", "Unspecified 4.",sub_ob_j_type),
-    sub_ob_j_type = ifelse(sub_ob_j_type=="Employment" & sub_ob_j == "4.\nWhat?", "Employment 4.",sub_ob_j_type),
-    sub_ob_j_type = ifelse(sub_ob_j_type=="Employment" & sub_ob_j == "3.a Who?\n(pop group)", "Employment 3.a",sub_ob_j_type),
-    sub_ob_j_type = ifelse(sub_ob_j_type=="Other" & sub_ob_j == "3.b Who?\n(mode)", "Other 3.b",sub_ob_j_type),
-        sub_ob_j_type = ifelse(sub_ob_j_type=="Other" & sub_ob_j == "3.a Who?\n(pop group)", "Other 3.a",sub_ob_j_type),
+    sub_ob_j_type = ifelse(sub_ob_j_type=="Unspecified" & sub_ob_j == "Who?\n(Mode)", 
+                           "Unspecified (Mode)",sub_ob_j_type),
+    sub_ob_j_type = ifelse(sub_ob_j_type=="Unspecified" & sub_ob_j == "What?", 
+                           "Unspecified (What?)",sub_ob_j_type),
+    sub_ob_j_type = ifelse(sub_ob_j_type=="Employment" & sub_ob_j == "What?", 
+                           "Employment (What?)",sub_ob_j_type),
+    sub_ob_j_type = ifelse(sub_ob_j_type=="Employment" & sub_ob_j == "Who?\n(Population group)",
+                           "Employment (Pop. group)",sub_ob_j_type),
+    sub_ob_j_type = ifelse(sub_ob_j_type=="Other" & sub_ob_j == "Who?\n(Mode)",
+                           "Other (Mode)",sub_ob_j_type),
+        sub_ob_j_type = ifelse(sub_ob_j_type=="Other" & sub_ob_j == "Who?\n(Population group)",
+                               "Other (Pop. group)", sub_ob_j_type),
     sub_ob_j_type = fct_reorder(sub_ob_j_type, count, .fun='median'))
@@
 df_all_sub_ob_just_reordered$ID = seq.int(nrow(df_all_sub_ob_just))
+
+df_all_sub_ob_just_reordered$sub_ob_j_f = factor(df_all_sub_ob_just_reordered$sub_ob_j, levels=c('When?','Where?','Who?\n(Population group)','Who?\n(Mode)','What?'))
 
@@
   geom_bar(stat="identity",  col = "grey", alpha=0.8, size=0.05) +
-  facet_grid(rows=vars(sub_ob_j),
+  facet_grid(rows=vars(sub_ob_j_f),
              scales="free_y",
@@
                       breaks = seq(0,0.8, by=0.2),
-                      limits = c(0,0.65),
+                      limits = c(0,0.67),
                       expand = c(0,0))+
@@
                               "All populations" =  "#F7FBFF",
-                              "Comp. vul" =  "#C6DBEF", 
+                              "Composite vulnerability" =  "#C6DBEF", 
                               "Education level" =  "#6BAED6", 
-                              "Employment 3.a" =  "#9ECAE1", 
+                              "Employment (Pop. group)" =  "#9ECAE1", 
                               "Gender" =  "#DEEBF7", 
                               "Income" =  "#4292C6", 
-                              "Other 3.a" =  "#9ECAE1",
+                              "Other (Pop. group)" =  "#9ECAE1",
                               "Race/Ethnicity" =  "#C6DBEF",
@@
                               "Car" =  "#74C476", 
-                              "Other 3.b" =  "#E5F5E0", 
+                              "Other (Mode)" =  "#E5F5E0", 
                               "Pedestrian" =  "#41AB5D",
                               "Transit" =  "#238B45",
-                              "Unspecified 3.b" =  "#C7E9C0",
+                              "Unspecified (Mode)" =  "#C7E9C0",
                               "Childcare" =  "#BCBDDC", 
@@
                               "Education" =  "#EFEDF5", 
-                              "Employment 4." =  "#9E9AC8", 
+                              "Employment (What?)" =  "#9E9AC8", 
                               "Healthcare" =  "#807DBA", 
@@
                               "Shops" =  "#BCBDDC", 
-                              "Unspecified 4." =  "#6A51A3")) + 
+                              "Unspecified (What?)" =  "#6A51A3")) + 
                       #c(rep(RColorBrewer::brewer.pal(9,"Pastel2"),4))) +
-  geom_text(size = 2.5, position = position_stack(vjust = 1), col = "Black") +
+  geom_text(size = 2.5, position = position_stack(vjust = 1.02), col = "Black") +
   theme_classic() +
@@
 ggsave("figures/df_all_sub_ob_just_reordered_plot.tiff", dpi=300)
-#manually remove the 3.a, 3.b. and 4. from the figure.
 ```
@@
 
-# Abstract
-
-This work provides a synthesis of how transportation fairness, justice and equity literature has operationalized standards.
-We first outline a flexible framework for engaging fairness questions and then apply it to collate the review's results.
-We find most articles were published within the last half decade and center urban areas, with a minority of work (40%) exploring Global South case studies.
-Income groups, followed by specific age groups (i.e., elders) and people with disabilities, are most commonly the subjects of justice.
-Transit and pedestrian modes are the most frequently studied mobility tools; with many analyses being multimodal and comparative in nature.
-The benefits and burdens, the "What" of mobility, focuses on movement (i.e., quality of trips taken) or the potential for movement (i.e., accessibility).
-How the reviewed works conceptualizes fairness is varied (e.g., Vertical Equity, Wellbeing, Rights-Based); but is often delineated with a supporting opportunity standard (37%) (i.e., number of parks within a 30-minute travel time) or population standard (36%) (i.e., bottom income quartile), though Infrastructure (i.e., level of service) and Environmental+ (i.e., air pollution) standards are also prominent.
-While the reviewed literature covers plenty of ground, we conclude with calls for action directed at researchers, practitioners, and transportation advocates.
-
 # Introduction
@@
 Transportation systems are technologies essential for social inclusion and activity participation, and therefore important from an equity perspective [@martensTransport2016; @karnerAdvancesPitfallsMeasuring2024; @vecchioTransportEquityLatin2020].
-Beyond ethical motivations, tracking objective and perceived inequalities is of interest for governing bodies to respond to popular demands for fairness.
+Beyond ethical motivations, tracking objective and perceived inequalities is of interest for governing bodies to respond to popular and needed demands for fairness.
 However, this has proven to be a challenging task: transportation systems are notoriously complex, with benefits and burdens that are diffuse over space and time.
@@
 Responding to this topic, a plethora of academic literature has emerged in recent decades.
-For instance, much research has been devoted to the issues of _measuring_ equity in transportation, including (among many others) [@delboscUsing2011; @martensMeasuring2019; @pritchardNew2022].
+For instance, much research has been devoted to the issues of _measuring_ equity in transportation  [e.g., @delboscUsing2011; @martensMeasuring2019; @pritchardNew2022].
 Further, there are multiple works that discuss the conceptual and philosophical foundations of equity and fairness in transportation [e.g., @martensTransport2016; @pereiraDistributiveJusticeEquity2017a; @vanoutriveHow2019].
 Previous reviews of equity in planning documents have been tightly scoped to cover accessibility [e.g., @boisjolyHowGetThere2017] or a particular mode of transportation [e.g., cycling in @doranPursuitCyclingEquity2021].
-These efforts are valuable, but in our estimation there remains a gap in terms of understanding _how_ standards for equity are developed and implemented for transportation systems.
+While these efforts are valuable, there remains a gap in terms of understanding _how_ standards for equity are developed and implemented for transportation systems.
 
@@
 Through a distributive justice lens [@pereiraDistributiveJusticeEquity2017a; @pereiraTransportationEquity2021], our work seeks to make two contributions.
-First, it outlines a conceptual and flexible framework for engaging and analysis of transportation fairness questions, based on the questions "Why?", "Where?", "When?", "Who?", "What?", and "How?" (5WH).
+First, it outlines a conceptual and flexible framework for engaging and analysing transportation fairness questions, based on the questions "Why?", "Where?", "When?", "Who?", "What?", and "How?" (5WH).
 Second, it applies the framework to collate the existing knowledge about fairness standards.
@@
 
-## Defintions
+## Definitions
 
 Fairness, equity and justice are closely related and often colloquially used interchangeably.
-In this section, we clarify these concepts definitions to facilitate their precise use throughout the manuscript.
+In this section, we clarify these concepts' definitions to facilitate their precise use throughout the manuscript.
 
 **Fairness** is a subjective and moral assessment of a treatment, outcome, or both, in relation to a given state of affairs.
-Within the transportation domain, the fairness of the distribution of transportation "goods" and "burdens" (e.g., access to opportunities, transportation-related externalities) for individual, groups and society is often at issue [@pereiraTransportationEquity2021].
-As examples: congestion pricing schemes could be unfair if they disproportionately disadvantage lower income groups [@eliassonCongestionPricingFair2016], systems designed to service "mandatory" destinations like employment or commercial retail that are not as relevant to all populations like those intangled in caring-work [@ravensbergenExploratoryAnalysisMobility2023; @hailConceptFairnessRelation2021], or even the application of justice such as Title VI of the U.S.
+Within the transportation domain, the fairness of the distribution of transportation "goods" and "burdens" (e.g., access to opportunities, transportation-related externalities) for individuals, groups and society is often at issue [@pereiraTransportationEquity2021].
+For example, congestion pricing schemes could be unfair if they disproportionately disadvantage lower income groups [@eliassonCongestionPricingFair2016], systems designed to service "mandatory" destinations like employment or commercial retail that are not as relevant to all populations like those entangled in caring -work [@ravensbergenExploratoryAnalysisMobility2023; @hailConceptFairnessRelation2021], or even the application of justice such as Title VI of the U.S.
 Civil Rights Act of 1964 for the purpose of accessibility planning as discussed in @martensFair2021.
@@
 To put it briefly, when it comes to the distribution of "goods," fairness can be understood as a _moral_ appraisal concerning the amount and strength of moral claims to those goods [@winteinBeFairClaims2024].
-Hence, in this manuscript, we define **fairness** as the moral evaluation of the rightness or wrongness of a given state of affairs (e.g., the impartiality and consistency of treatment or associated outcomes); fairness serves as a yardstick for justice.
+Hence, in this manuscript, we define **fairness** as the moral evaluation of the rightness or wrongness of a given state of affairs (e.g., the impartiality and consistency of treatment or associated outcomes).
+In other words, fairness serves as a yardstick for justice.
 
@@
 440].
-Different scopes and approaches to justice have been develop, as formalizing fairness depends on the desirability of different states of affairs, which depend on populations, places, times, and different scopes.
+Different scopes and approaches to justice have been developed, as formalizing fairness depends on the desirability of different states of affairs, which depend on populations, places, times, and different scopes.
 For instance, several forms of justice can be distinguished [@jaggarPhilosophicalChallengesGlobal2009; @pereiraDistributiveJusticeEquity2017a; @karnerTransportationEquityTransportation2020]:
@@
 
-Wile justice (i.e., formalized fairness) is a broad moral concept, **equity**  and **standards** can be understood as the 'instruments' of justice- the tools through which society moves towards a just state of affairs.
+While justice (i.e., formalized fairness) is a broad moral concept, **equity**  and **standards** can be understood as the 'instruments' of justice- the tools through which society moves towards a just state of affairs.
 
@@
 As another example, a more strict standard could be based on egalitarian principles (e.g., proportional equity) where the benefits or burdens are weighted by population; in the consistent application of this standard, each group would give or receive in proportion to their relevant 'size' [@billsLookingMeanEquity2017;@martensJusticetheoreticApproachDistribution2012].
-In contrast, an affirmative action standard (i.e., could be related to restorative justice) could be an even stricter standard, requiring the benefits to be distributed in a non-egalitarian way to favour people still harmed by past or present discriminatory practices [@bierbaumMobilityJusticeLinking2021].
-
-To summarise, we use **justice** to describe a moral target, **fairness** as an evaluation or yardstick of justice, **equity** as a set of related disparity measurement tools, and **standards** as statements of fairness.
+In contrast, an affirmative action standard (i.e., those related to restorative justice) could be an even stricter standard, requiring the benefits to be distributed in a non-egalitarian way to favour people still harmed by past or present discriminatory practices [@bierbaumMobilityJusticeLinking2021].
+
+To summarise, in this work we use **justice** to describe a moral target, **fairness** as an evaluation or yardstick of justice, **equity** as a set of related disparity measurement tools, and **standards** as concrete statements of fairness.
 
@@
 
-In the case of transportation the question of **"Where?"** is paramount, as transportation by its very nature concentrates the benefits (e.g., access points to the system are not ubiquitous).
-The burdens, in contrast, are often diffuse: they are incrementally paid, for example by a distributed population in the form of taxes, or by a (possibly different) population in the form of poor health.
+In the case of transportation, the question of **"Where?"** is paramount as transportation by its very nature concentrates the benefits (e.g., access points to the system are not ubiquitous).
+The burdens, in contrast, are often diffuse: they are incrementally paid, for example, by a distributed population in the form of taxes, or by a (possibly different) population in the form of poor health.
 As such, the answer to "Where?" is the definition of the spatial boundaries.
@@
 Conventionally, the question of **"When?"** refers to the temporal circumstances within which the demands of justice apply.
-When it comes to transportation, questions regarding temporarily are important, as for example: _when_ did the analysis take place and under what historical policy context; when is _the right time_ to invest in infrastructure and as a result when to generate a spatial inequality [@rabelloEvaluation2015]; for _how long_ the burdens and benefits can be associated to a specific intervention; or even the _timeline_ of reparative justice to reconcile the shadows of past transport-related injustices.
-
-When asking **"Who?"** we think about the entities that should be regarded as subjects/arbiters of justice.
+When it comes to transportation, questions regarding temporarily are important in different domains: _when_ did the analysis take place and under what historical policy context; when is _the right time_ to invest in infrastructure and as a result when to generate a spatial inequality [@rabelloEvaluation2015]; for _how long_ the burdens and benefits can be associated to a specific intervention; or even the _timeline_ of reparative justice to reconcile the shadows of past transport-related injustices.
+
+When asking **"Who?"**, we think about the entities that should be regarded as subjects/arbiters of justice.
 For tractability, this question is often approached through the filter of population groups, which may include several concurrent traits, including gender identity, ableness, ethnicity, age, caste, and income.
@@
 Although a person is not their mode of transportation, there are large segments of the population who cannot extricate themselves from the mobility tools they use, either because they have driven themselves out of choices [@laveryDriving2013], or have been driven out of choices by factors beyond their control [@jacquesRescuing2012].
-While it is important to avoid conflating the "Who" with the "What", for analytical purposes we need to be mindful of the connection between a person and their mobility tools.
+While it is important to avoid conflating the "Who" with the "What", we need to be mindful of the connection between a person and their mobility tools for analytical purposes.
 
@@
 To understand the distributional implications of transportation systems, it is essential to understand what they _produce_.
-Transportation systems are technologies to improve the rate at which space is traded for time by increasing the speed of movement.
+Transportation systems rely on technologies to improve the rate at which space is traded for time by increasing the speed of movement.
 But as the adage goes, travel is derived demand [@mokhtarianUnderstanding2001].
@@
 Coupled with land use, mobility creates accessibility to opportunities as well as varied burdens.
+For instance, some are direct and paid by the traveler (e.g., travel time, out-of-pocket costs), but many others are indirect and related to network externalities (e.g., exposure to pollution).
 Transportation justice, thus, involves proximate (mobility tools and mobility) and ulterior (accessibility and activity opportunities) objectives.
@@
 The initial draft of the protocol was refined from preliminary searches of related reviews [e.g., @zhangLiterature2021; @pereiraDistributiveJusticeEquity2017a; @vecchioAccessibilityCapabilitiesApproach2021] and in consultation with research services librarians.
-The final protocol was released on OSF [@soukhovEquityJusticeConceptualizations2022], the methods are summarised as follows.
+The final protocol was released on OSF [@soukhovEquityJusticeConceptualizations2022], and the methods are summarised as follows.
 
@@
 The terms were bundled by means of Boolean operators.
-These stages are summarised as follows (see Appendix section @sec-sect61 for details):
+These stages are summarised as follows (see Appendix @sec-sect61 for details):
 
@@
 The semi-automated nature of the search strategy was overly inclusive, thereby reducing the risk of omitting relevant material.
-The next stage was to trim the corpus; the authors worked with a group of undergraduate research assistants to scan the documents and assess their relevance based on titles and abstracts.
+The next stage was to trim the corpus; the authors worked with a group of previously trained undergraduate research assistants to scan the documents and assess their relevance based on titles and abstracts.
 Two research assistants voted on each document, and a third vote from the authorship team broke ties.
@@
 
-The smattering of studies from the Global South are predominately from Asia, notably China, but also India, Thailand, Iran, Philippines, and Indonesia.
+40% of studies from the Global South are predominately from Asia, notably China, but also India, Thailand, Iran, Philippines, and Indonesia.
 The next most common focus within the literature from the Global South is from South America.
@@
 Formal planning processes in the South operate under greater financial precariety, they rely on more informal processes to address unmet needs.
-This turns out to be a significant source of equity concerns e.g, informal transit [@friedMeasuringSustainableDevelopment2020], populations living in informal settlements [@sharmaPublicTransitAccessibility2021].
-International standards are more highly relevant in literature from the Global South e.g., using the WHO noise and air pollution standard [@apparicioCyclingOneMost2021], while not as frequently featured in the Global North literature e.g., in @carrierApplicationThreeMethods2014, disadvantaged populations are disproportionately impacted by higher $NO_2$ pollution near roadways in Montreal but these levels are still consistently below the WHO standard.
-This difference in focus present interesting opportunities for the Global South, to adopt potentially successful enhancements from Global North formal equity planning processes (e.g., indicator creation for disadvantages groups [@cuiSpatialAccessPublic2020]) and to avoid past and on-going mistakes (e.g., from entrenched car-centric development [@warrenDevelopingEquitableSustainable2015] to the disproportionate contribution to carbon-intense mobility [@chancelCarbonInequalityKyoto2015]).
+This turns out to be a significant source of equity concerns, e.g, informal transit [@friedMeasuringSustainableDevelopment2020], and populations living in informal settlements [@sharmaPublicTransitAccessibility2021].
+International standards are more highly relevant in literature from the Global South, such as the WHO noise and air pollution standard [@apparicioCyclingOneMost2021], while not as frequently featured in the Global North literature.
+For example, in @carrierApplicationThreeMethods2014, disadvantaged populations are disproportionately impacted by higher $NO_2$ pollution near roadways in Montreal, but these levels are still consistently below the WHO standard.
+This difference in focus presents interesting opportunities for the Global South, to adopt potentially successful enhancements from Global North formal equity planning processes (e.g., indicator creation for disadvantaged groups [@cuiSpatialAccessPublic2020]) and to avoid past and ongoing mistakes (e.g., from entrenched car-centric development [@warrenDevelopingEquitableSustainable2015] to the disproportionate contribution to carbon-intense mobility [@chancelCarbonInequalityKyoto2015]).
 For the Global North, there is an opportunity to reckon with its own contributions to uneven development globally and environmental impacts as well as adopt relevant aspects from informal planning processes.
@@
 "All Populations" papers make no distinction in population.
-This is done, for instance in @kitaDoesLevelTransportation2020, who investigates disparities in accessibility to food-stores and self-reported capability/frequency of going outdoors.
+This is done, for instance, by @kitaDoesLevelTransportation2020, who investigates disparities in accessibility to food stores and self-reported capability/frequency of going outdoors.
 Often, the implicit or explicit motivation of these papers is access (to necessities and desired destinations) for all.
@@
 Multi-dimensional considerations are so prevalent in the reviewed literature that "Gender", "Race/ethnicity", "Education", or "Employment" are infrequently studied in isolation.
-Only a few papers focus exclusively on gendered differences in active transportation [@adlakhaMindGapGender2020, @xieWonCycleRoute2018], race/ethnicity's relationship to green space proximity [@silvaEnvironmentalJusticeAccessibility2018], and culturally-appropriate opportunities [@wangModelingSpatialAccessibility2011].
+Only a few papers focus exclusively on gendered differences in active transportation [@adlakhaMindGapGender2020; @xieWonCycleRoute2018], race/ethnicity's relationship to green space proximity [@silvaEnvironmentalJusticeAccessibility2018], and culturally-appropriate opportunities [@wangModelingSpatialAccessibility2011].
 Papers that focus *solely* on "Employment" status or "Education" level are completely absent in the corpus.
@@
 We ascertain that for these reasons, "Car" is often compared against non-car modes and never put forward as the only mode that should be expanded.
-Furthermore, some papers reflect on "Other" forms of mobility (@fig-fig1), including wheelchair accessible taxi and/or paratransit services [@marquezFactorsAffectingPersonal2019; @wilkinsonmeyersLiveOrdinaryLife2015], travel on waterways [@vadrevuMeasuringSpatialEquity2016; @caoIndicatorsSocioSpatialTransport2017; @parrySocialVulnerabilityClimatic2018], motorcycle or other micro-mobility [@berryInvestigatingFuelPoverty2016; @savingmothersDidSavingMothers2019; @tiwariDevelopmentPublicTransport2021], or emergency vehicles [@pedigoInvestigationDisparitiesGeographic2010; @patelDeterminingGeographicAreas2007].
+Furthermore, some papers reflect on "Other" forms of mobility (@fig-fig1), including wheelchair-accessible taxi and/or paratransit services [@marquezFactorsAffectingPersonal2019; @wilkinsonmeyersLiveOrdinaryLife2015], travel on waterways [@vadrevuMeasuringSpatialEquity2016; @caoIndicatorsSocioSpatialTransport2017; @parrySocialVulnerabilityClimatic2018], motorcycle or other micro-mobility [@berryInvestigatingFuelPoverty2016; @savingmothersDidSavingMothers2019; @tiwariDevelopmentPublicTransport2021], or emergency vehicles [@pedigoInvestigationDisparitiesGeographic2010; @patelDeterminingGeographicAreas2007].
 
@@
 ## "What" could be the objects of transportation justice
-The objects of fairness assessment, the "What", are often motivated by mobility or accessibility.
+The objects of fairness assessment, the "What", are often motivated by concepts encompassed by either mobility or accessibility.
 A fundamental benefit of transportation systems is mobility (enabling or impeding movement).
@@
 For this reason, although the right to the road (and transportation systems more generally) is important, the literature leans heavily on the ulterior object, namely accessibility to (the ease of reaching) destinations.
+In tandem, themes of existing transportation system burdens associated with mobility and accessibility can be interpreted: for instance, these could be environmental (e.g., potential for GHG emissions reductions relative to the status quo [@bocarejoImpactsFormalizationIntegration2022]), or human-health and traffic-related externalities (e.g., decreased wellbeing due to transport poverty [@churchillTransportPovertySubjective2019] or reduction in potential road-related fatalities [@bocarejoImpactsFormalizationIntegration2022])
 
@@
 
-Most papers in the reviewed literature take a broad approach, with `r (dest_count$count[8]/165) |> percent()` focusing on "Unspecified" destinations (@fig-fig1).
-They examine various equity dimensions across different transportation modes.
-Some papers focus on factors that impact the trip itself like infrastructure quality, safety, and service level [@zheEvaluationSharedUse2008; @prasertsubpakijEvaluatingAccessibilityBangkok2012; @furstBestBadPractices2013 ; @lattmanPerceivedAccessibilityPublic2016].
-Others analyze trips tailored for specific groups like women or people with disabilities [@russellPedallingEquityExploring2021; @wilkinsonmeyersLiveOrdinaryLife2015], often with a consideration of what constitutes 'sufficient' quality of life [@churchillTransportPovertySubjective2019].
-In sum, these papers reveal the multifaceted nature of transportation systems: they serve utilitarian purposes while also shaping user experiences.
-
-### Accessibility
-
-In terms of destinations, "Healthcare" services (`r (dest_count$count[6]/165) |> percent()`) and "Employment" (`r (dest_count$count[5]/165) |> percent()`) are the most common in the reviewed literature (@fig-fig1).
-Papers on healthcare often highlight disparities in services, like @wangModelingSpatialAccessibility2011, who assess access to Mandarin-speaking family physicians for Mainland Chinese immigrants in Toronto.
+Most papers in the reviewed literature take a broad approach, with `r (dest_count$count[8]/165) |> percent()` focusing on "Unspecified" destinations (@fig-fig1), examining various equity dimensions across different transportation modes.
+Many of these papers are 'mobility' focused and examine the impact of the trip itself on the individuals, groups or the environment.
+These benefits and burdens include safety [@zheEvaluationSharedUse2008], infrastructure quality and service levels [@prasertsubpakijEvaluatingAccessibilityBangkok2012; @furstBestBadPractices2013; @lattmanPerceivedAccessibilityPublic2016] or air and pollution levels [@apparicioCyclingOneMost2021].
+Others analyze trips taken by specific groups like women or people with disabilities [@russellPedallingEquityExploring2021; @wilkinsonmeyersLiveOrdinaryLife2015], often with a consideration of what constitutes 'sufficient' quality of life [@churchillTransportPovertySubjective2019].
+In sum, these papers reveal the multifaceted nature of transportation systems: they serve utilitarian purposes related to travel while also shaping user experiences, expressing some aspects that can be summarized as societal and/or planetary benefits and burdens.
+
+### Access to opportunities
+
+Generally, papers that focus on a particular destination can be construed to interpret "accessibility" as the "What?" as the object of fairness, with access _to_ opportunities being the benefit.
+When analyzing access to particular destinations, "Healthcare" services (`r (dest_count$count[6]/165) |> percent()`) and "Employment" (`r (dest_count$count[5]/165) |> percent()`) are the most common in the reviewed literature (@fig-fig1).
+Papers on healthcare often highlight disparities in services, like @wangModelingSpatialAccessibility2011, who assess access to Mandarin speaking family physicians for Mainland Chinese immigrants in Toronto.
 Similarly, papers focusing on employment are often aimed at identifying transportation-poor neighbourhoods [@allenSizingTransportPoverty2019a; @churchillTransportPovertySubjective2019].
@@
 The lack of information about community destinations in the reviewed literature, especially for children, is noticeable [@desjardinsChildren2022].
+In sum, these works that focus on 'accessibility' can be seen to frame access and its related externalities as benefits, while the lack of access (or associated externalities) as burdens.
 
@@
 
-The conceptual foundations of fairness are not always made explicit within the reviewed literature, they often had to be inferred.
+The conceptual foundations of fairness are often not made explicit within the reviewed literature, they have to be inferred.
 Though not exhaustive, the following categories emerged from our review and in broad terms are (@fig-fig2 details definitions):
@@
 Examples include setting travel impedance thresholds [@shenMeasuringSpatialAccessibility2020] and mapping accessibility indices spatially across populations [@monzonEfficiencySpatialEquity2013] or population-groups [@sharmaPublicTransitAccessibility2021].
-These papers may also address traffic-related air and noise pollution or urban temperatures.
+These papers may also address burdens such as traffic-related air and noise pollution or urban temperatures.
 In these papers, equity is theoretically achieved if similar levels are attained for all populations (horizontal equity) or spatial areas (spatial equity).
@@
 Nonetheless, their use provides interpretable values for tracking progress, offering comparability across communities, unlike accessibility measures, which vary in methods and assumptions.
+
+Appendix @sec-sect63 presents detailed examples of various standards, outlined within the 5WH framework, as derived from the examined corpus.
+
 # Moving forward: discussion and calls for action
 
-This work makes two contributions to the literature.
+As previously noted, this work makes two contributions to the literature.
 First, it outlines a conceptual and flexible "Why", Where", "When", "Who", "What" and "How" (5WH) framework with supporting definitions for approaching questions about transportation justice.
-Second, it applies the framework in the synthesis and appraisal of 165 academic articles drawn from the transportation justice literature; these studies use equity analysis as instruments to gauge proximity to their context's just situation.
-We summarise what is in the literature in the preceding section using the 5WH framework as appropriate, the following is a discussion on what may be missing from the literature:
-
-- **"Who?"**: While who is subject of inequities in the literature is extensive, dimensions are still missing.
-There is a need to go beyond the focus on low-income transit riders; positioning intersectional considerations within the creation of community-based equity definitions and tailored standards are critical, as the act of delineating equity-deserving communities impacts results e.g., [@rowangouldIdentifyingEnvironmentalJustice2015].
-Standards should be sensitive to evolving community-based definitions of inequities.
-Are issues of economic inequity at the root of transport inequities for a particular community?
-Are inequities in access provided by certain modes emphasised because access offered by other modes are _relatively_ better?
-Access to what types of opportunities are driving transport inequities?
-How do populations, travel modes, and opportunities sought intersect to define the "who" of inequities?
-A community-informed understanding of inequities and tracking how it changes are needed.
-There is also a need for further intersectional considerations, as clear "Who?"-specific standards are lacking.
-
-- **"What?"** and **"How?"**: The connection of what is the focus of inequity measure to some sort standard.
+Second, it applies the framework in synthesizing and appraising of 165 academic articles drawn from the transportation justice literature; these studies use equity analysis as instruments to gauge proximity to their context's just situation.
+Based on the preceding sections of what is in the literature, and what is lacking from our perspective, we conclude with the following five calls to action.
+Based on the preceding sections of what is in the literature, and what is lacking from our perspective, we conclude with the following five calls to action.
+
+<!-- - **"What?"** and **"How?"**: The connection of what is the focus of inequity measure to some sort standard.
 Specifically, "What?" are the benefits and burdens of a transportation system that could link accessibility or mobility to some sort of tangible fairness object of justice.
@@
 However, literature that does not, usually connects to qualitative results and/or operationalizes mixed-methods.
-
-- **"Why?"** is not within the scope of this review.
+-->
+
+<!-- - **"Why?"** is not within the scope of this review.
 Also, **"When?"** and **"Where?"** are only reviewed from a cursory perspective.
 These dimensions were not clear enough in the literature to be explicitly commented on.
-
-Based on the preceding sections of what is in the literature, and what is lacking from our perspective, we conclude with the following five calls to action.
-
-## Call 1: Develop standards based on rigorous concepts of justice
-
-The conceptual grounding for standards is often left implicit within the literature: for example, @muellerSocioeconomicInequalitiesUrban2018 suggests the relative risk of mortality from transport-related air pollution should not be higher in deprived groups than the general population.
+-->
+
+<!--  -->
+
+## Call 1: Underpin tailored standards based on rigorous concepts of justice
+
+The conceptual grounding for standards is often left implicit within the literature.
+For example, @muellerSocioeconomicInequalitiesUrban2018 suggests the relative risk of mortality from transport-related air pollution should not be higher in deprived groups than the general population.
 While the conceptual justice foundations are not explicitly declared, we infer an egalitarian focus.
@@
 
+Additionally, there is a need to go beyond the focus on low-income transit riders, as clear “Who?”-specific standards are lacking.
+Positioning intersectional considerations within the creation of community-based equity definitions and tailored standards are critical, as the act of delineating equity-deserving communities impacts results (e.g., @rowangouldIdentifyingEnvironmentalJustice2015).
+Standards should be sensitive to evolving community-based definitions of inequities.
+Are issues of economic inequity at the root of transport inequities for a particular community?
+Are inequities in access provided by certain modes emphasized because access offered by other modes are _relatively_ better?
+Access to what types of opportunities are driving transport inequities?
+How do populations, travel modes, and opportunities sought intersect to define the “who” of inequities?
+A community-informed understanding of inequities and tracking how they change are needed.
+
 For decision-makers, setting standards, measuring inequities, and developing flexible guidelines for standards that are also compatible with community-informed calls for justice, are the next steps for transportation equity planning.
@@
 
-On the methodological side, transportation justice research would benefit from wider use of mixed-methods.
+On the methodological side, transportation justice research would benefit from a wider use of mixed-methods.
 Concepts and standards are often discussed from purely qualitative or quantitative approaches.
@@
 These questions usually remain unanswered in the literature despite their importance.
-These measurements can also bring some challenges and pitfalls as summarised by @karnerAdvancesPitfallsMeasuring2024 but are necessary for more effective equity analysis.
+These measurements can also bring some challenges and pitfalls as summarised by @karnerAdvancesPitfallsMeasuring2024 but are necessary for more effective equity analyses.
 
@@
 We suspect that since the methods used in the reviewed literature are predominately quantitative, the reliance on commonly used point of interest databases is also high.
-These databases typically include education, health and occasionally include aggregated categories for leisure and community destinations, but are less generous for points of interest that are more often associated with understudied mobility patterns.
+These databases typically include education, health, and occasionally include aggregated categories for leisure and community destinations, but are less generous for points of interest that are more often associated with understudied mobility patterns.
 
@@
 In the context of the Global South and rural geographies fewer official data sources and public research resources exist relative to urban communities and areas in the Global North.
-As well as in the operationalization of emerging conceptual theories in equity analysis, such as sufficientariansm [@vanderveenOperationalizingIndicatorSufficient2020].
+The operationalization of emerging conceptual theories in equity analysis, such as sufficientariansm [@vanderveenOperationalizingIndicatorSufficient2020], might be impacted if relevant data is not available.
 
@@
 And, who gets to consult and use the data?
-Deciding who is the subject of justice frames the data collection activity; if it is the mobility of those that do domestic work, the classification of who does this work and how it changes over time/space is fundamental.
+Deciding who is the subject of justice frames the data collection activity; if it is the mobility of those who do domestic work, the classification of who does this work and how it changes over time/space is fundamental.
 How we classify mobility and access has implications for our understanding of just who is a subject of justice, as illustrated by the history of racial classifications in the U.S.
@@
 Robust assessments of the implications of standards on lived experiences are still lacking in the literature, but are essential for equity analysis that translates into just practice.
-While the estimation of the benefits of increased mobility or accessibility, or reducing burdens is commonplace in the literature, there is a need to associate them to outcomes, including life and neighbourhood satisfaction, subjective well-being, mental and physical health, and social capital.
+While the estimation of the benefits of increased mobility or accessibility, or reducing burdens is commonplace in the literature, there is a need to associate them with outcomes, including life and neighbourhood satisfaction, subjective well-being, mental and physical health, and social capital.
+The literature must move beyond describing inequities and their contributing factors.
+Instead, it should incorporate causal methods that establish clear cause-and-effect relationships, enabling policymakers to quantify the impact of various policies and infrastructure decisions on advancing justice.
 
@@
 Outcomes of interventions can be compared within and between communities, and cross community comparisons can be created that may expedite the adoption of effective policies that move towards just outcomes; the presence of these synthesis and comparative studies could support brave decision-makers in the application of research into practice.
+Again, causal methods and mixed methods approaches might serve as valuable approaches to reach this in practice.
 
@@
 
-To provide further example of data extraction, a table of the resulting literature looked like:
+## Select reviewed works summarized by the 5WH {#sec-sect63}
+
 ```{r}

### 2025-03-09 Journal-manuscript/Journal-manuscript.tex
Integrated Nacho's comments

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\usepackage{colortbl}
\usepackage{array}
\usepackage{anyfontsize}
\usepackage{multirow}
\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Searching for standards of fairness in the transportation literature},
  pdfkeywords={transportation; equity; justice; standards;
transportation planning; fairness; literature review},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Searching for standards of fairness in the transportation
literature}
\author{}
\date{}

\begin{document}
\maketitle

\section{Abstract}\label{abstract}

This work provides a synthesis of how transportation fairness, justice
and equity literature has operationalized standards. We first outline a
flexible framework for engaging fairness questions and then apply it to
collate the literature review's results. We find most articles were
published within the last half decade and center urban areas, with 40\%
exploring Global South case studies. Income groups, followed by specific
age groups (i.e., older adults) and people with disabilities, are most
commonly the subjects of justice. Transit and pedestrian modes are the
most frequently studied mobility tools; with many analyses being
multimodal and comparative in nature. The benefits and burdens, the
``What'' of mobility, focuses on movement (i.e., quality of trips taken)
or the potential for movement (i.e., accessibility). How the reviewed
works conceptualize fairness is varied (e.g., Vertical Equity,
Wellbeing, Rights-Based); but is often delineated with a supporting
opportunity standard (37\%) (i.e., number of parks within a 30-minute
travel time) or population standard (36\%) (i.e., bottom income
quartile), though Infrastructure (i.e., level of service) and
Environmental+ (i.e., air pollution) standards are also prominent. While
the reviewed literature is extensive, we conclude with calls to action
for researchers, practitioners, and transportation advocates to drive
meaningful change. These include developing standards based on rigorous
conceptualizations, advancing systems-thinking approaches to fairness,
ensuring data accessibility as a matter of justice, strengthening the
connection between standards and lived experiences, and rigorously
evaluating interventions and policies.

Keywords: transportation, equity, justice, fairness, standards, metrics

\section{Introduction}\label{introduction}

Transportation systems are technologies essential for social inclusion
and activity participation, and therefore important from an equity
perspective (Karner et al., 2024; Martens, 2016; Vecchio et al., 2020).
Beyond ethical motivations, tracking objective and perceived
inequalities is of interest for governing bodies to respond to popular
and needed demands for fairness. However, this has proven to be a
challenging task: transportation systems are notoriously complex, with
benefits and burdens that are diffuse over space and time. To compound
matters, emerging technologies and service models can swiftly change the
balance of benefits and burdens among a population (Guo et al., 2020).
Transportation systems that are engineered to offer higher mobility for
people \emph{somewhere} can simultaneously cut others off from essential
opportunities \emph{elsewhere} (Raje, 2004). The shades of policies cast
long shadows, as shown by the legacy of U.S. urban highways (Archer,
2020) and impacts of transportation-related climate change (Markolf et
al., 2019).

Responding to this topic, a plethora of academic literature has emerged
in recent decades. For instance, much research has been devoted to the
issues of \emph{measuring} equity in transportation (e.g., Delbosc \&
Currie, 2011b; Martens et al., 2019; Pritchard et al., 2022). Further,
there are multiple works that discuss the conceptual and philosophical
foundations of equity and fairness in transportation (e.g., Martens,
2016; R. H. M. Pereira et al., 2017; Vanoutrive \& Cooper, 2019).
Previous reviews of equity in planning documents have been tightly
scoped to cover accessibility (e.g., Boisjoly \& El-Geneidy, 2017) or a
particular mode of transportation (e.g., cycling in Doran et al., 2021).
While these efforts are valuable, there remains a gap in terms of
understanding \emph{how} standards for equity are developed and
implemented for transportation systems.

The objective of this work is to broadly scan the state of this
knowledge. Through a distributive justice lens (R. H. M. Pereira et al.,
2017; R. H. M. Pereira \& Karner, 2021), our work seeks to make two
contributions. First, it outlines a conceptual and flexible framework
for engaging and analysing transportation fairness questions, based on
the questions ``Why?'', ``Where?'', ``When?'', ``Who?'', ``What?'', and
``How?'' (5WH). Second, it applies the framework to collate the existing
knowledge about fairness standards. To achieve this, we scan the state
of academic knowledge in defining standards of fairness in
transportation. In contrast to previous reviews on measuring inequality
in transportation systems, this work is concerned with the implicit or
explicit standards used to judge whether inequalities are fundamentally
``fair'' or unacceptable.

\section{Background}\label{background}

\subsection{Definitions}\label{definitions}

Fairness, equity and justice are closely related and often colloquially
used interchangeably. In this section, we clarify these concepts'
definitions to facilitate their precise use throughout the manuscript.

\textbf{Fairness} is a subjective and moral assessment of a treatment,
outcome, or both, in relation to a given state of affairs. Within the
transportation domain, the fairness of the distribution of
transportation ``goods'' and ``burdens'' (e.g., access to opportunities,
transportation-related externalities) for individuals, groups and
society is often at issue (R. H. M. Pereira \& Karner, 2021). For
example, congestion pricing schemes could be unfair if they
disproportionately disadvantage lower income groups (Eliasson, 2016),
systems designed to service ``mandatory'' destinations like employment
or commercial retail that are not as relevant to all populations like
those entangled in caring -work (Hail \& McQuaid, 2021; Ravensbergen et
al., 2023), or even the application of justice such as Title VI of the
U.S. Civil Rights Act of 1964 for the purpose of accessibility planning
as discussed in Martens \& Golub (2021). Individuals may feel certain
situations are unfair, processes can be seen as unfair, or even outcomes
of these processes and situations can be unfair.

There are a multitude of aspects to \textbf{fairness} in transportation
(Hail \& McQuaid, 2021), namely because the distribution of
transportation ``goods'' and ``burdens'' are often ambiguous.
Individuals' ability to access and benefit from transportation-related
outcomes varies continuously, often in ways that are unrecognized
(implicitly or explicitly) or inadequately considered by transportation
planners and other influential actors in the system. To put it briefly,
when it comes to the distribution of ``goods,'' fairness can be
understood as a \emph{moral} appraisal concerning the amount and
strength of moral claims to those goods (Wintein, 2024). Hence, in this
manuscript, we define \textbf{fairness} as the moral evaluation of the
rightness or wrongness of a given state of affairs (e.g., the
impartiality and consistency of treatment or associated outcomes). In
other words, fairness serves as a yardstick for justice.

Relatedly, \textbf{justice} can be viewed as the formalized goal of
fairness, an analytically reasoned \emph{fair} state of affairs. Justice
is attained when people ``give and receive whatever they are due''
(Jaggar, 2009, pp. 1--2), and it ceases to exist when there are persons
or groups that are denied ``access to the opportunities they need to
lead a meaningful and dignified life'' (Karner et al., 2020, p. 440).
Different scopes and approaches to justice have been developed, as
formalizing fairness depends on the desirability of different states of
affairs, which depend on populations, places, times, and different
scopes. For instance, several forms of justice can be distinguished
(Jaggar, 2009; Karner et al., 2020; R. H. M. Pereira et al., 2017):

\begin{itemize}
\item
  \textbf{Retributive justice} is concerned with the proportional
  retribution of wrongdoers relative to legitimate punishers and the
  innocent (Walen, 2023).
\item
  \textbf{Reparative (or restorative) justice} focuses on the reparation
  of caused harm; it centers the needs and voices of victims to restore
  wrongdoers and the community according to moral values (Braithwaite et
  al., 2003; Tyler, 2006). In planning and policy contexts, reparative
  justice involves accountability mechanisms (material, powers, rights,
  processes) to compensate victims (Safransky, 2022; Williams \& Steil,
  2023).
\item
  \textbf{Procedural justice} strives to ensure that the views and
  preferences of all stakeholders are fairly accounted for in the
  decision-making and inter-personal procedures affecting their lives
  and communities (Tyler, 2006).
\item
  \textbf{Distributive justice} is perhaps the most studied form of
  justice in transportation (R. H. M. Pereira et al., 2017) and
  elsewhere (see Jaggar, 2009, p. 2). Its main concern is the collection
  of benefits and burdens of the tangible and intangible products of
  society by different segments of a population. The just distributive
  aspects of transportation systems is also of focus in this work.
\end{itemize}

While justice (i.e., formalized fairness) is a broad moral concept,
\textbf{equity} and \textbf{standards} can be understood as the
`instruments' of justice- the tools through which society moves towards
a just state of affairs.

\textbf{Equity}, as conceptualized alongside distributive justice, tends
to encompass tools to understand the distribution of benefits and
burdens of things among a population (e.g., disparities), emphasizing
those with the least advantage or most disadvantage. In the
transportation domain, equity analysis often flows from the top:
stemming from the authority of the state and meant to assist with
decisions about regulating and financing spending (Karner et al., 2020).
As in Karner et al. (2020), equity and disparity analysis should not be
seen as an end in and of itself, but rather as a means to gather
information about actual, observed, and perceived inequities.

And lastly, in this work's context, a \textbf{standard} is a concrete
statement about fairness; it is a formalized threshold in understanding
disparities (i.e., related to equity) as part of the goal of formalized
fairness (i.e., justice). A standard establishes criteria for how
benefits or burdens should be distributed or allocated in a way that
aligns with moral or formalized principles. For example, a weak standard
might be a Pareto improvement, where benefits may be concentrated in
certain groups (even those already advantaged), as long as no group is
made worse off compared to the existing situation (Tan et al., 2016;
S.-X. Xu et al., 2018). As another example, a more strict standard could
be based on egalitarian principles (e.g., proportional equity) where the
benefits or burdens are weighted by population; in the consistent
application of this standard, each group would give or receive in
proportion to their relevant `size' (Bills \& Walker, 2017; Martens et
al., 2012). In contrast, an affirmative action standard (i.e., those
related to restorative justice) could be an even stricter standard,
requiring the benefits to be distributed in a non-egalitarian way to
favour people still harmed by past or present discriminatory practices
(Bierbaum et al., 2021).

To summarise, in this work we use \textbf{justice} to describe a moral
target, \textbf{fairness} as an evaluation or yardstick of justice,
\textbf{equity} as a set of related disparity measurement tools, and
\textbf{standards} as concrete statements of fairness.

\subsection{A framework to analyze questions of justice:
5WH}\label{a-framework-to-analyze-questions-of-justice-5wh}

Having defined justice, fairness, equity and standards, we approach the
literature with an analytical apparatus inspired by the framing of
Jaggar (2009) for philosophical questions of justice.\footnote{Similar
  questions are found peppered throughout the literature. This is done
  either explicitly, as for example in Karner et al. (2020), who ask
  ``of what'', ``for whom,'' and ``how much'' in reference to equity; or
  implicitly, as in Gössling (2016), who asks of the outputs (``what?'')
  of transportation (exposure, space, access) and ``for whom?'' (gender,
  age, ethnicity).}. According to Jaggar (2009), Western philosophy has
approached the issue of justice by asking ``Why?'', ``Where?'',
``When?'', ``Who?'', ``What?'', and ``How?'' (5WH), applying them to a
particular domain or sphere of life relevant to justice.

In the case of transportation, the question of \textbf{``Where?''} is
paramount as transportation by its very nature concentrates the benefits
(e.g., access points to the system are not ubiquitous). The burdens, in
contrast, are often diffuse: they are incrementally paid, for example,
by a distributed population in the form of taxes, or by a (possibly
different) population in the form of poor health. As such, the answer to
``Where?'' is the definition of the spatial boundaries.

Conventionally, the question of \textbf{``When?''} refers to the
temporal circumstances within which the demands of justice apply. When
it comes to transportation, questions regarding temporarily are
important in different domains: \emph{when} did the analysis take place
and under what historical policy context; when is \emph{the right time}
to invest in infrastructure and as a result when to generate a spatial
inequality (Rabello Quadros \& Nassi, 2015); for \emph{how long} the
burdens and benefits can be associated to a specific intervention; or
even the \emph{timeline} of reparative justice to reconcile the shadows
of past transport-related injustices.

When asking \textbf{``Who?''}, we think about the entities that should
be regarded as subjects/arbiters of justice. For tractability, this
question is often approached through the filter of population groups,
which may include several concurrent traits, including gender identity,
ableness, ethnicity, age, caste, and income. Often, it is appropriate to
consider the intersections between traits, given differences in a
person's lived experiences. A complication in the case of transportation
is that disentangling the ``Who'' from their mobility tools is not
always straightforward. Although a person is not their mode of
transportation, there are large segments of the population who cannot
extricate themselves from the mobility tools they use, either because
they have driven themselves out of choices (Lavery et al., 2013), or
have been driven out of choices by factors beyond their control (Jacques
et al., 2012). While it is important to avoid conflating the ``Who''
with the ``What'', we need to be mindful of the connection between a
person and their mobility tools for analytical purposes.

\textbf{``What?''} refers to which entities should be regarded as
objects of inequities, meaning which categories of things should be
distributed in a just manner. To understand the distributional
implications of transportation systems, it is essential to understand
what they \emph{produce}. Transportation systems rely on technologies to
improve the rate at which space is traded for time by increasing the
speed of movement. But as the adage goes, travel is derived demand
(Mokhtarian et al., 2001). For this reason, we cannot stop at
considering only mobility but must consider its ulterior goal--reaching
destinations. Coupled with land use, mobility creates accessibility to
opportunities as well as varied burdens. For instance, some are direct
and paid by the traveler (e.g., travel time, out-of-pocket costs), but
many others are indirect and related to network externalities (e.g.,
exposure to pollution). Transportation justice, thus, involves proximate
(mobility tools and mobility) and ulterior (accessibility and activity
opportunities) objectives.

The next question is \textbf{``How?''}, and it relates to the allocation
of various objects of justice (``what'') to various subjects of justice
(``who'') in various circumstances (``when'' and ``where''). Fairness
standards are an equity tool for answering this distributive question.
The thresholds can be quantitative (e.g., square meters of green space
per capita), qualitative descriptions (e.g., do not knowingly
discriminate) or a mix of the two. Some examples include: maximum travel
distance/cost/time to or from key destinations, levels of maximum
exposure to externalities (i.e., noise or air pollution), un/fulfilled
needs, and dis/satisfaction with travel. To support us when approaching
this question, we draw from concepts in transport-related social
exclusion, transport disadvantage, and transport poverty, which are
typically based on utilitarian or sufficientarian philosophies. The
``How?'' is supported by fairness conceptualisations along with
standards.

Above all, convincing answers to the above questions require a
supporting rationale: a \textbf{``Why?''} (Jaggar, 2009). This is
perhaps the most slippery of 5WH. Justice is inherently a social
construct. Asking \textbf{Why?} amounts to asking what sort of social
contracts regulate human interactions, that is, the self-imposed rules
that result from our collective will to believe. These contracts can be
defined by constitution, but there are often unwritten and possibly
contested variants. In this way, analyzing the ``Why?'' in the corpus is
not the focus of this review, partly because answers to ``Why?'' are
seldom explicitly stated. Instead, our focus is on the standards of
fairness that, combined with the use of analysis, can help us understand
how better to move towards just transportation systems and better
formulate answers to ``Why?''.

\subsection{Review methods}\label{review-methods}

This work examines the academic literature on transportation to identify
the extent to which standards for equity are defined and employed. In
this task, we follow the Joanna Briggs Institute (JBI) approach to the
conduct of scoping reviews, an approach that builds upon the Arksey and
O'Malley (2005) framework (Peters et al., 2020). The review is also
guided by the Preferred Reporting Items for Systematic Reviews and
Meta-Analyses, particularly the extension for scoping reviews
(PRISMA-ScR) which is consistent with the JBI approach (Tricco et al.,
2018).

The primary research question and the protocol were initially defined by
the authors, a group of experts in the field of transportation. The
initial draft of the protocol was refined from preliminary searches of
related reviews (R. H. M. Pereira et al., 2017; Vecchio \& Martens,
2021; e.g., Zhang \& Zhao, 2021) and in consultation with research
services librarians. The final protocol was released on OSF (Soukhov et
al., 2022), and the methods are summarised as follows.

The search strategy was developed iteratively using \textbf{inclusion}
and \textbf{exclusion} criteria (Peters et al., 2020). For the inclusion
criteria, the mnemonic PCC (population, concept, and context) was
adopted. The search strategy was refined by adding search terms. The
terms were bundled by means of Boolean operators. These stages are
summarised as follows (see Appendix section Section~\ref{sec-sect61} for
details):

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  An initial limited search of Web of Science (WoS) Core Collection
  (containing journals, conference proceedings, and books published all
  over the world) was undertaken to identify key documents. Separate
  searches using the terms `transportation' and `equity' were generated.
  From these searches, we examined the text contained in the titles and
  abstracts, the index terms, and subject heading searches when
  available. As we developed a clearer outline of the literature, we
  refined the terms used for the search. This took the form:
  (``Transport'' OR ``Transit'' OR ``Car*'' OR ``Walk'' OR
  ``Bike''\ldots{}\textbf{1}) AND (``Equity'' OR ``Justice'' OR
  ``Fair''\ldots{}\textbf{2}), where \textbf{1} and \textbf{2} signify
  additional terms relating to `transportation' and `equity',
  respectively.
\item
  Upon inspection of the preliminary search results and after achieving
  a consensus among the authors, the set of search terms related to
  `equity' was expanded into three sets of terms. The first describes
  theories and concepts of equity, the second describes the object of
  justice (i.e., the ``what'' in our analytical framework), and the
  third describes terms referring to standards (i.e., the ``how'').
  These sets of terms were augmented iteratively. The final search query
  took the following general form: (``Transport'' OR ``Transit'' OR
  ``Car*'' OR ``Walk'' OR ``Bike''\ldots{}\textbf{1}) AND (``Equity'' OR
  ``Justice'' OR ``Equity'' OR ``Fair''\ldots{}\textbf{2}) AND
  (``Accessibility'' OR ``Mobility'' OR \ldots{}\textbf{3}) AND
  (``Standard'' OR ``Threshold'' OR \ldots{}\textbf{4}) where
  \textbf{1},\textbf{2},\textbf{3}, and \textbf{4} signify additional
  terms included in the sets combined with ``OR'' operators.
\end{enumerate}

After testing the search strategy on WoS Core Collection, we applied it
to an augmented list of databases. The databases used were: WoS General
Collection-Science Citation Index Expanded, WoS Social Sciences Citation
Index, and Transportation Research International Documentation (TRID).
The definitive version of the search was completed and exported by the
lead author on March 21st, 2021. The number of documents identified in
this was was 6,382.

The semi-automated nature of the search strategy was overly inclusive,
thereby reducing the risk of omitting relevant material. The next stage
was to trim the corpus; the authors worked with a group of previously
trained undergraduate research assistants to scan the documents and
assess their relevance based on titles and abstracts. Two research
assistants voted on each document, and a third vote from the authorship
team broke ties. After this step, 1,710 documents were assessed based on
full-text, again with each being voted on by two research assistants and
an authorship team member tie-breaker. Next, using a data extraction
template and workflow that was pilot-tested with a subset of papers, the
authorship team extracted data from the eligible documents using
\emph{Covidence} (Covidence, 2023), an online application for literature
screening. The evidence selection workflow, flow diagram, data
extraction template, and some sample data extractions can be consulted
in Appendix Figure~\ref{fig-figA2}. The end result of this scan was a
corpus of 165 documents retained for data extraction.

\section{Synthesis of findings}\label{synthesis-of-findings}

This section threads together the trends identified from the reviewed
transportation literature. Specifically, a thematic overview of
``When?'' and ``Where?'' transportation fairness is considered, ``Who?''
are the subjects of justice, ``What?'' could be considered the objects
of justice, and lastly ``How?'' is fairness measured. As an overview,
Figure~\ref{fig-fig1} illustrates the prominence of each thematic
category for ``When?'', ``Where?'', ``Who?'', and ``What?'', while
Figure~\ref{fig-fig2} presents an overview of the types of fairness
conceptualizations and thresholds, addressing the ``How?''.

\begin{figure}

\centering{

\includegraphics{figures/df_all_sub_ob_just_reordered_plot.tiff}

}

\caption{\label{fig-fig1}The proportion of papers in each `When',
`Where', `Who', and `What' by topics in each category. Topic categories
were generated by the authors based on the reviewed literature.}

\end{figure}%

\begin{figure}

\centering{

\includegraphics{figures/standards_conc_percentages_plot_and_table.tiff}

}

\caption{\label{fig-fig2}The proportion of papers by topics in the `How'
categories. `How' is split into two categories: topics related to
fairness conceptualisation (horizontal axis) and topics related to
standards (vertical axis). Topic categories were generated by the
authors based on the reviewed literature.}

\end{figure}%

\subsection{``When'' and ``Where'' is transportation fairness
considered}\label{when-and-where-is-transportation-fairness-considered}

Within our corpus, most papers (60\%) focus on studies in the Global
North, with many studies from North America (particularly U.S. and
Canada), Europe (France, Spain and Northern Europe), Oceania (Australia
and New Zealand), and Asia (Japan and Israel). Though their subject
matter is varied, their spatial context mainly pertains to North America
and Europe, and thus more often than not deals with more developed and
formal government transport planning apparatuses and technologies e.g.,
planning for equitable high-speed rail (Monzon et al., 2013), autonomous
vehicle technology (Eppenberger \& Richter, 2021), or on the public
consultation processes (Reddy et al., 2010).

Of note, work included in the reviewed literature does not often explore
the historical or temporal aspects of transportation fairness in depth.
Therefore, in this section we provide a thematic overview of the
temporal and spatial contexts associated with the studied case studies
and the publication location.

40\% of studies from the Global South are predominately from Asia,
notably China, but also India, Thailand, Iran, Philippines, and
Indonesia. The next most common focus within the literature from the
Global South is from South America. Many of these studies mention a
systemic absence of evidence relevant to the region (Vecchio et al.,
2020). Despite a growing recognition of the interconnections between
transport development, social exclusion, and poverty (Benevenuto \&
Caulfield, 2020), a number of studies underscore ongoing neglect of the
social dimension of transport during the planning stage (Benevenuto \&
Caulfield, 2020; Boisjoly et al., 2020). Many studies also point at
affordability as one of the main mobility barriers in the region
(Falavigna \& Hernandez, 2016; Rivas et al., 2018), while some highlight
multi-dimensional concerns such as public transport accessibility and
quality of walking environments that contribute to mobility inequalities
(Tiznado-Aitken et al., 2018). Studies pertaining to Africa are even
less numerous.

A shared characteristic among studies from Africa and South America is a
scarcity of official transport data (Fried et al., 2020). These studies
also incorporate the use of informal transportation options and tensions
in developing road network infrastructure (which tends to support car
dependency) over meeting the mobility/accessibility needs of citizens
more equitably and sustainably (Thondoo et al., 2020). To address these
challenges, researchers compile databases based on open and
geo-referenced data, calculate objective and/or subjective measures
(Berhe et al., 2014), and focus on advancing transport justice for low
to medium income countries by aligning their goals with external policy
guidelines such as the Sustainability Development Goals (SDG),
particularly those related to universal accessibility (Fried et al.,
2020).

Of all the studies reviewed, 85\% focus on urban and suburban settings
and are highly varied in their research aims. To give an example, Cox \&
Bartle (2020) qualitatively examine cycling as a mode of travel for
people with disabilities in a typical mid-size town in the U.K. Ampe et
al. (2020), on the other hand, work to identify the lateral clearance
that motorists should maintain when passing cyclists with children
seats. The remainder of the studies focus on rural regions (14\%), or
remote regions, such as those that rely on inter-island ferry trips in
the Philippines (Cao \& Stanley, 2017). Similarly, Parry et al. (2018)
studied remote communities in the Amazonian region, and suggest that
``increasing accessibility through road building would be maladaptive,
exposing marginalized people to further harm and exacerbating climatic
change by driving deforestation'' (pp.~125).

Overall, studies from the Global South often have some key differences
in focus compared to the Global North:

\begin{itemize}
\item
  affordability as a barrier at the user- or policy maker- level is more
  often the motivation in work from the Global South;
\item
  the expression of greater tensions in investing in new transportation
  infrastructure, such as roads in rural/under-developed areas, compared
  to prioritization of non-car modes. Studies centered in the Global
  North often focus on the later; and
\item
  more significant data availability limitations and reliance on
  `informally' collected data.
\end{itemize}

These differences center on the direct economic outcomes tied to
transport infrastructure. Often, work from the Global South does not
engage as intimately with emerging mobility technologies. As well, the
informal aspects of transport planning are more overbearing. Clearly,
countries in the Global South still struggle with the consequences of
past colonialism, which has left them more reliant on primary sector
exports (lower efficiency, lower national GDP) under growing global
financial markets, and with more fragile democracies. Due to lower data
availability, reliance on crowd-sourced or `informally' collected data,
and more extreme needs for `sufficient' transport, analysis of
transportation inequities is often cast along economic lines in the
Global South.

We gather that the literature on the Global North and South tend to
focus on different aspects of transport inequities, related to the level
of their transportation infrastructure. Formal planning processes in the
South operate under greater financial precariety, they rely on more
informal processes to address unmet needs. This turns out to be a
significant source of equity concerns, e.g, informal transit (Fried et
al., 2020), and populations living in informal settlements (Sharma \&
Patil, 2021). International standards are more highly relevant in
literature from the Global South, such as the WHO noise and air
pollution standard (Apparicio et al., 2021), while not as frequently
featured in the Global North literature. For example, in Carrier et al.
(2014), disadvantaged populations are disproportionately impacted by
higher \(NO_2\) pollution near roadways in Montreal, but these levels
are still consistently below the WHO standard. This difference in focus
presents interesting opportunities for the Global South, to adopt
potentially successful enhancements from Global North formal equity
planning processes (e.g., indicator creation for disadvantaged groups
(Cui et al., 2020)) and to avoid past and ongoing mistakes (e.g., from
entrenched car-centric development (Warren et al., 2015) to the
disproportionate contribution to carbon-intense mobility (Chancel \&
Piketty, 2015)). For the Global North, there is an opportunity to reckon
with its own contributions to uneven development globally and
environmental impacts as well as adopt relevant aspects from informal
planning processes.

\subsection{``Who'' are the subjects of transportation justice and their
mobility
tools}\label{who-are-the-subjects-of-transportation-justice-and-their-mobility-tools}

\subsubsection{Population groups}\label{population-groups}

Turning to the question of ``Who?'' in Figure~\ref{fig-fig1}, the focus
of the literature tends to be on ``Income'', especially the
lowest-income groups who simultaneously face lower mobility and
accessibility \emph{and} higher costs and exposure (Falavigna \&
Hernandez, 2016; Peungnumsai et al., 2020; Zhao et al., 2020). Evidence
suggests that low household income is a significant determinant of
transport-related inequities e.g., urban access to public transport
(Peungnumsai et al., 2020), access to urban employment opportunities
(Boisjoly et al., 2020), and unfavorable rates of environmental noise,
air pollution, and green space (Kruize et al., 2007). Yet, low-income is
not universally associated with lower transport-related benefits for
every object of justice. For instance, in Sheffield, U.K., Mears et al.
(2019) demonstrate that historically working-class (lower-income)
neighbourhoods have \emph{more} access to green space than other
neighbourhoods, but of lower quality, likely due to historic urban
planning approaches. Similarly, Bertrand et al. (2008) find that the
granularity of the analysis matters, and lower income groups do not
always have lower accessibility, something echoed in other spatial and
temporal contexts (Allen \& Farber, 2019; Foth et al., 2013)

``Age'' is the second most common ``Who?'' focus. In this regard,
Martinez-Jimenez \& Salinas-Perez (2019) and Arranz-Lopez et al. (2019)
investigate travel distances/times to various opportunities,
demonstrating how age is associated with differences in opportunity
access. School-aged children and older populations are a common focus.
For children, analysis of wellbeing (Laszkiewicz \& Sikorska, 2020),
safety and access to schools (Corazza et al., 2020; Sharma \& Patil,
2022), and promoting active travel (Mackie, 2009; Mehdizadeh et al.,
2017) are common-place. Papers that focus on older adults similarly
investigate transport-related wellbeing (Y. Chen et al., 2020), access
to age-specific destinations (Cheng et al., 2019), and options to reduce
unmet travel needs (Nordbakke \& Schwanen, 2015).

Many papers focus on intersecting characteristics. As an exception, we
classify some studies as focusing on ``(Dis)abilities'' or ``All
Populations''. Studies with a ``(Dis)abilities'' focus assess travel
capabilities, namely through physical accessibility and universal design
guidelines (Chiscano, 2021; Orellana et al., 2020; J. Park et al.,
2017). ``All Populations'' papers make no distinction in population.
This is done, for instance, by Kita et al. (2020), who investigates
disparities in accessibility to food stores and self-reported
capability/frequency of going outdoors. Often, the implicit or explicit
motivation of these papers is access (to necessities and desired
destinations) for all.

A large proportion of papers apply composite vulnerability indices that
combine several individual traits like low income, unemployment, and/or
immigrant status. These indices are generated from government sources or
author-informed census data creation methods. As an example, the
Neighbourhood Equity Index is a measure of vulnerability created by the
City of Toronto and used in Awuor \& Melles (2019) to examine
disparities in premature death. Other works use national census
indicators such as the social and housing deprivation index (Pucci et
al., 2019) or explore transport disadvantage, equity in policy
implementation, or transport-related mortality burden by means of census
measures (e.g., household poverty) and transport-related accessibility
indicators (Aldred et al., 2021; Iungman et al., 2021; Scheurer et al.,
2017; Sun \& Thakuriah, 2021). Similarly, Environmental Justice (EJ)
indicators have been used in the U.S. literature to identify
neighbourhoods that have a higher than average proportion of low-income
and non-white populations and evaluate the equity impacts of
transportation projects (K. Park et al., 2021; Reddy et al., 2010;
Rowangould et al., 2016).

Multi-dimensional considerations are so prevalent in the reviewed
literature that ``Gender'', ``Race/ethnicity'', ``Education'', or
``Employment'' are infrequently studied in isolation. Only a few papers
focus exclusively on gendered differences in active transportation
(Adlakha \& Parra, 2020; Xie \& Spinney, 2018), race/ethnicity's
relationship to green space proximity (Silva et al., 2018), and
culturally-appropriate opportunities (Wang \& Roisman, 2011). Papers
that focus \emph{solely} on ``Employment'' status or ``Education'' level
are completely absent in the corpus. Furthermore, ``Other'' population
groups are also frequently considered: this is a catch-all category that
includes populations less commonly targeted by research e.g., veterans
and access to specific-healthcare needs (Mooney et al., 2000), pregnant
people and access to maternity services (Vadrevu \& Kanjilal, 2016), and
youth who live in foster care (Batsche \& Reader, 2012). Overall, the
diversity of population groups considered in the corpus demonstrates the
variety of transportation-equity concerns addressed.

\subsubsection{Modes (mobility tools)}\label{modes-mobility-tools}

Travel mode, though modifiable, is intertwined with individual identity,
presenting challenges for the analysis of disparities. In this
subsection, we choose to view the mode of travel primarily as those who
use the mode, hence as subjects of justice (i.e., the ``Who?'').
However, at times, travel mode will also be considered as an object of
justice (i.e., a ``What?'', such as disparities in rural transit
service).

To begin, transit characterises the majority of the reviewed literature
(Figure~\ref{fig-fig1}), reflecting some intuitive sense. Despite
transit being perceived unfavourably by some (e.g., Mella Lira \& Paez
(2021)), it is often viewed as the only or primary mobility option mode
for many (Jacques et al., 2012). Overall, we can infer that transit is
largely considered a public good and, therefore, a natural object of
justice. As such, public transportation is especially amendable to being
modified to meet the demands of justice by ensuring adequate funding to
provide barrier-free transport for most, despite challenges such as low
densities, fiscal constraints, and political will (Markard et al.,
2023). As such, papers that focus on transit assess a variety of related
topics, including food deserts (McKey et al., 2020) and barrier-free
transportation for people who face disabilities (Feeley, 2019;
Jiménez-Espada \& González-Escobar, 2021; Lim et al., 2021; Liu et al.,
2019).

Transit also plays a central role in multimodal comparisons in transport
disparities, especially comparing ``Walk'', ``Car'', or some other
category in Figure~\ref{fig-fig1}. For instance, Brussel et al. (2019)
compares transit, pedestrian and road network accessibility measurements
in the context of SDG 11.2. Renne \& Mayorga (2018) reviews natural
disaster evacuation plans, focusing on carless households and
emphasizing transit and pedestrian networks. A few papers also frame
transit as a direct competitor of car travel or use it as a benchmark
(Golub \& Martens, 2014; Martens et al., 2012). For instance, Warren et
al. (2015) proposes car ownership standards while acknowledging the
tension between mobility needs in transit under-served areas and
emission reduction goals. However, this framing is not universal, and
transit is sometimes seen as a mode to fulfill individual capabilities.
As an example, Smith et al. (2012) explores perspectives about transport
needs and costs to achieve perceived sufficient living standards for
those living in rural areas. Notably, papers vary in the importance they
place on climate urgency, with some focusing more on satisfying
\emph{all} sufficient individual needs while planning for less
car-dependent cities in the future.

After transit, pedestrian travel (``Walk'') is the second most studied
object of justice. Pedestrians represent a unique convergence of
``what'' and ``who,'' utilizing their own bodies for mobility. Papers
focusing primarily on walking often use walkability scores to assess
neighbourhood quality (Evans, 2015), mobility by different demographics
(H. Kim et al., 2016; Towne et al., 2016), or urban peripheral regions
(Blecic et al., 2021). Some use walkability to gauge public health and
urban vitality (McCormack et al., 2012; Sung \& Lee, 2015). Papers with
a pedestrian focus also often see walking as a bridge to connect
multiple modes: they discuss `walkability' as part of active
transportation, which focuses on both walking, bicycle and/or transit.
Concepts discussed include how active transport contributes to
children's physical activity levels (Mammen et al., 2014), walkability
as an alternative to car predominance (Bertrand et al., 2008) or tension
that exists between modes, creating unsafe conditions for walking
(Ferenchak \& Marshall, 2019; Siu, 2019).

The third most studied mode in the reviewed literature is ``Car''.
Unlike transit and walking, these studies rarely feature car as the mode
of primary focus. Car is often assessed against transit and active modes
(e.g., low traffic neighbourhoods Aldred et al., 2021) or in areas with
inadequate transit service (Aljoufie, 2016; Kimmel et al., 2018). Under
our appraisal, transit and walk modes are of most interest because they
represent a reckoning with automobility's legacy and ongoing demands for
space, public subsidies and government supports that underestimate their
true cost (Gössling et al., 2019; Timperley, 2021). Upon the rise of the
automobile, pedestrian spaces were distinguished, non-car modes came to
be seen as a hindrance to full automobility, and transit seen as a
social service. We ascertain that for these reasons, ``Car'' is often
compared against non-car modes and never put forward as the only mode
that should be expanded. Furthermore, some papers reflect on ``Other''
forms of mobility (Figure~\ref{fig-fig1}), including
wheelchair-accessible taxi and/or paratransit services (Marquez et al.,
2019; Wilkinson-Meyers et al., 2015), travel on waterways (Cao \&
Stanley, 2017; Parry et al., 2018; Vadrevu \& Kanjilal, 2016),
motorcycle or other micro-mobility (Berry et al., 2016; Schmitz et al.,
2019; Tiwari \& Phillip, 2021), or emergency vehicles (Patel et al.,
2007; Pedigo \& Odoi, 2010).

Papers that pay no particular attention to any mode are ``Unspecified'':
as examples, a focus on road infrastructure or road network distances
(Mishra et al., 2014; Wismadi et al., 2014), travel needs generally
(Benevenuto \& Caulfield, 2020; Titheridge et al., 2008), realized
travel (Abasolo et al., 2001), or externalities of realized travel
(Iungman et al., 2021).

\subsection{``What'' could be the objects of transportation
justice}\label{what-could-be-the-objects-of-transportation-justice}

The objects of fairness assessment, the ``What'', are often motivated by
concepts encompassed by either mobility or accessibility. A fundamental
benefit of transportation systems is mobility (enabling or impeding
movement). This benefit is sometimes valued by itself but is often seen
by the literature as instrumental to achieve an ulterior goal (e.g.,
activity participation and associated benefits). For example, although
vehicle kilometers traveled (VKT) is sometimes seen as a useful policy
instrument (Zhao \& Li, 2021), travelling more is not necessarily a sign
of advantage when accessibility is low, and short trips may actually be
a sign of advantage (K. Park et al., 2021). For this reason, although
the right to the road (and transportation systems more generally) is
important, the literature leans heavily on the ulterior object, namely
accessibility to (the ease of reaching) destinations. In tandem, themes
of existing transportation system burdens associated with mobility and
accessibility can be interpreted: for instance, these could be
environmental (e.g., potential for GHG emissions reductions relative to
the status quo (Bocarejo \& Urrego, 2022)), or human-health and
traffic-related externalities (e.g., decreased wellbeing due to
transport poverty (Churchill \& Smyth, 2019) or reduction in potential
road-related fatalities (Bocarejo \& Urrego, 2022))

\subsubsection{Mobility}\label{mobility}

Most papers in the reviewed literature take a broad approach, with 47\%
focusing on ``Unspecified'' destinations (Figure~\ref{fig-fig1}),
examining various equity dimensions across different transportation
modes. Many of these papers are `mobility' focused and examine the
impact of the trip itself on the individuals, groups or the environment.
These benefits and burdens include safety (Zhe et al., 2008),
infrastructure quality and service levels (Fürst \& Vogelauer, 2013;
Lattman et al., 2016; Prasertsubpakij \& Nitivattananon, 2012) or air
and pollution levels (Apparicio et al., 2021). Others analyze trips
taken by specific groups like women or people with disabilities (Russell
et al., 2021; Wilkinson-Meyers et al., 2015), often with a consideration
of what constitutes `sufficient' quality of life (Churchill \& Smyth,
2019). In sum, these papers reveal the multifaceted nature of
transportation systems: they serve utilitarian purposes related to
travel while also shaping user experiences, expressing some aspects that
can be summarized as societal and/or planetary benefits and burdens.

\subsubsection{Access to opportunities}\label{access-to-opportunities}

Generally, papers that focus on a particular destination can be
construed to interpret ``accessibility'' as the ``What?'' as the object
of fairness, with access \emph{to} opportunities being the benefit. When
analyzing access to particular destinations, ``Healthcare'' services
(18\%) and ``Employment'' (25\%) are the most common in the reviewed
literature (Figure~\ref{fig-fig1}). Papers on healthcare often highlight
disparities in services, like Wang \& Roisman (2011), who assess access
to Mandarin speaking family physicians for Mainland Chinese immigrants
in Toronto. Similarly, papers focusing on employment are often aimed at
identifying transportation-poor neighbourhoods (Allen \& Farber, 2019;
Churchill \& Smyth, 2019). Employment is frequently used as a proxy for
overall accessibility since it is the most common trip purpose and
employment is usually co-located with destinations like shops,
recreation, and other services. These studies typically use travel
surveys, census data, and point-of-interest databases, and benefit from
well-developed institutional data. This especially holds in the Global
North where these data are more readily available.

Other destinations have received less attention despite serving
essential needs. ``Shopping'' destinations (19\%) often aim to identify
food deserts (Choi \& Suzuki, 2013; Jiao et al., 2012; D. Kim \& Park,
2020; McKey et al., 2020). ``Education''-related papers (18\%) explore
children's active transportation to school (Larkins et al., 2011) or
universal design (Larkins et al., 2011). Places of ``Leisure'' (18\%)
prompt questions about their spatial distribution (M. Xu et al., 2017)
and accessibility (Mavoa et al., 2015). Fewer papers cover ``Community''
destinations (e.g., public service centres, places of community support
or worship) (10\%) or ``Childcare'' (5\%), but they are integral to
holistically study activity participation (Alberts et al., 2016; Smith
et al., 2012). The lack of information about community destinations in
the reviewed literature, especially for children, is noticeable
(Desjardins et al., 2022). In sum, these works that focus on
`accessibility' can be seen to frame access and its related
externalities as benefits, while the lack of access (or associated
externalities) as burdens.

\subsection{Standards: ``How'' fairness is
measured}\label{standards-how-fairness-is-measured}

\subsubsection{Conceptualisations of
fairness}\label{conceptualisations-of-fairness}

The conceptual foundations of fairness are often not made explicit
within the reviewed literature, they have to be inferred. Though not
exhaustive, the following categories emerged from our review and in
broad terms are (Figure~\ref{fig-fig2} details definitions):

\begin{itemize}
\item
  Vertical equity (27\%)
\item
  Transport-related social exclusion (27\%)
\item
  Well-being (27\%)
\item
  Spatial equity (26\%)
\item
  Horizontal equity (17\%)
\item
  Inequitable externalities (17\%)
\item
  Rights (14\%)
\item
  Emerging theories (5\%)
\item
  Utilitarian (1\%)
\end{itemize}

A slice of the reviewed literature is supported by broader ``Rights''
conceptualisations of fairness: these papers often focus on equity for
people with disabilities or non-car users and associated challenges
accessing transport infrastructure (Bharathy \& D'Souza, 2018; Daamen et
al., 2008; Jiménez-Espada \& González-Escobar, 2021). While many papers
are underpinned with the right to the city (the \emph{right} to
participate in the production of urban space (Adli \& Donovan, 2018;
Lefebvre, 1967)), others emphasize legal \emph{Rights} like regulations
aligned with The Americans with Disabilities Act (ADA) (Bharathy \&
D'Souza, 2018) or the goal of \emph{access for all} in land-use
transportation master plans (Lim et al., 2021).

In another subset of the reviewed literature, distributions are examined
through concepts of ``Horizontal'' and ``Spatial equity'', often using
quantitative methods to assess distributional disparities without
explicit justice rationales. Examples include setting travel impedance
thresholds (Shen et al., 2020) and mapping accessibility indices
spatially across populations (Monzon et al., 2013) or population-groups
(Sharma \& Patil, 2021). These papers may also address burdens such as
traffic-related air and noise pollution or urban temperatures. In these
papers, equity is theoretically achieved if similar levels are attained
for all populations (horizontal equity) or spatial areas (spatial
equity). This egalitarian perspective rarely delves into minimum or
maximum levels associated with harm or need satisfaction.

Papers that center ``Well-being'' assess what constitutes a satisfactory
life in relation to transportation; mixed-methods are often used and, in
contrast to the last set, the objects of injustice are often identified.
Some papers use physical activity guidelines and surveys to understand
the effect of active transportation infrastructure (Adlakha \& Parra,
2020; Auchincloss et al., 2020; McCormack et al., 2012). Mixed or
qualitative methods combined with health-related outcome standards lead
to more concrete statements of fairness e.g., travel times for emergency
treatment (Schmitz et al., 2019) or premature mortality (Awuor \&
Melles, 2019).

Another research branch, often quantitative with some qualitative or
mixed-methods studies, focuses on ``Transport-related social
exclusion'', ``Vertical equity'', and/or
``Sufficientarian/capabilities''. Here, the objects of justice are often
clearly identified and link standards to tangible welfare-informed
outcomes. They focus on groups from perspectives of disadvantage such
as: social exclusion and transport poverty (Allen \& Farber, 2019;
Churchill \& Smyth, 2019; Delbosc \& Currie, 2011a), food deserts (McKey
et al., 2020), and energy poverty (Berry et al., 2016; Berry, 2019;
Robinson \& Mattioli, 2020).

\subsubsection{Standards and methods of measuring
fairness}\label{standards-and-methods-of-measuring-fairness}

We identify several categories of standards that overlap with at least
one of the conceptualisations of fairness discussed above (see
Figure~\ref{fig-fig2} for definitions):

\begin{itemize}
\item
  Opportunity standards (66\%)
\item
  Population standards (64\%)
\item
  Infrastructure standards (41\%)
\item
  Environment+ standards (7\%)
\end{itemize}

Papers with ``Opportunity'' standards often employ quantitative methods
to analyze disparities and assess distributional fairness. Many deal
with travel impedance thresholds based on speed, distance, or cost (Z.
Chen \& Haynes, 2017; Shen et al., 2020; Yenisetty \& Bahadure, 2020).
Inequality measures like the Gini coefficient and poverty measures are
used to empirically define travel impedance thresholds (Tiznado-Aitken
et al., 2018; van der Veen et al., 2020). Further, methods tangential to
travel impedance, like limiting transport expenditure to 10\% of monthly
income (Rivas et al., 2018), addressing spatial mismatch (Mulley et al.,
2015), or pinpointing areas of relative regional inequities are also
used. Notably, many papers with ``Opportunity'' standards consider
multiple dimensions, employing similar methods but tailored to different
focal points. For instance, Peungnumsai et al. (2020) suggest service
benchmarks of equal supply and demand of transit, revealing horizontal
inequities as well. Others conceive the externalities of transportation
system as trade-offs and aim to maximize transport-related benefits
(i.e., time savings, emissions reductions, congestion reductions, user
fares) through optimization/location-allocation methodologies
(Fakhrmoosavi et al., 2021; Wismadi et al., 2014; Zheng \& Geroliminis,
2020).

``Population'' standards are often founded on ``Well-being''
conceptualisations from a variety of socio-demographic and spatial
angles. Methods include: establishing thresholds based on questionnaires
and comparisons to recommended physical activity levels (Auchincloss et
al., 2020; H. Kim et al., 2016; McCormack et al., 2012; Timperio et al.,
2015), region-relative comparisons in health outcomes in a spatial unit
such as premature mortality rates (Awuor \& Melles, 2019), spatial
access benchmarks based on population-related characteristics like
supermarket access (Murphy et al., 2017) and hospital access (R. Pereira
et al., 2021), summative per capita benchmarks such as decent living
energy consumption levels (Rao \& Baer, 2012), and community-informed
spatial boundaries like EJ defined communities (Rowangould et al.,
2016). While most of these papers use quantitative or mixed-methods
approaches, some employ exclusively qualitative methods (Berhe et al.,
2014).

Papers that feature both ``Population'' and ``Opportunity'' standards
are often founded on ``Vertical equity'', ``Well-being'', and/or
``Transport-related social exclusion'' conceptualisations. They feature
mixed-methods, with questionnaires and qualitative approaches for
``population'' standards and quantitative methods like accessibility
indices for ``opportunity'' standards. Census data and household
estimates within specific travel distances or times to key destinations
identify social exclusionary situations (W.-H. Chen, 2010; Daniels \&
Mulley, 2011; Sharma \& Patil, 2021; Sun \& Thakuriah, 2021), linkages
between transport disadvantages (Delbosc \& Currie, 2011a), areas
experiencing transport poverty (Allen \& Farber, 2019; Churchill \&
Smyth, 2019), food deserts (McKey et al., 2020), or transport-related
energy poverty (Berry et al., 2016; Berry, 2019; Robinson \& Mattioli,
2020). They employ various methods, such as clustering techniques (Mohri
et al., 2021). Some exclusively use qualitative methods to analyze
survey data on travel willingness/barriers or conduct interviews on
unmet activity needs (W.-H. Chen, 2010; Mehdizadeh et al., 2017;
Nordbakke \& Schwanen, 2015).

``Infrastructure'' standards offer another perspective on fairness,
commonly grounded in ``Rights'' conceptualisations, which are twice as
frequent in this segment of the corpus compared to other
conceptualisations. These papers most frequently address the rights of
non-car users and populations with disabilities. Though various methods
are applied, infrastructure and environmental audits as well as
qualitative approaches are most prominent. Infrastructure audits compare
existing infrastructure against universal design best practices
(Jiménez-Espada \& González-Escobar, 2021; Larkins et al., 2011; Odeck
et al., 2010; Perez-delHoyo et al., 2021) or investigate elements
correlating with mode use by specific population groups (Moniruzzaman \&
Paez, 2016). Qualitative methods include interviews/surveys on perceived
access (Desjardins et al., 2021; Fürst \& Vogelauer, 2013; Lim et al.,
2021; Marquez et al., 2019; Mateo-Babiano et al., 2017; J. Park et al.,
2017; Stjernborg, 2019; Velho et al., 2016) and assessment of standards
under best-practice criteria (Bharathy \& D'Souza, 2018; Daamen et al.,
2008; Velho et al., 2016).

Additionally, ``Infrastructure'' standards papers sometimes encompass
multiple dimensions, moving beyond rights (to the infrastructure) to
provide ``Opportunity'' and/or ``Population'' standards. These papers
often employ ``Vertical'', ``Horizontal'', and ``Spatial equity''
lenses. These papers often refer to guidelines and propose composite
indices. For example, (Rachele et al., 2017) integrate various transport
network properties to define an indicator supporting walkability and
public transport access. Others evaluate infrastructure quality (M. Xu
et al., 2017), accident severity (Appleyard et al., 2017; Benevenuto \&
Caulfield, 2020), and user-groups, especially disadvantaged ones with
respect to vertical equity of multi-criteria indicators (Prasertsubpakij
\& Nitivattananon, 2012). Some focus explicitly on affordability and
barriers, proposing infrastructure enhancements for better inclusivity,
especially for the most disadvantaged (Basu \& Alves, 2019; Song et al.,
2019; Welch, 2013), grounded in transport-related social exclusion (Kent
\& Karner, 2019) or capabilities approaches (Smith et al., 2012).

``Environmental+'' standards are featured less prominently in the
reviewed literature, possibly because the environmental burdens of
transportation are addressed more broadly in other literature (e.g.,
environmental justice). However, the papers included in our corpus
present some interesting insights. They often use traffic-related air
and noise pollution, green space, urban design, urban air temperature,
health outcomes, and physical activity guidelines to assess
transport-related ``inequitable externalities''. Methods employed are
primarily quantitative or mixed-methods, identifying inequalities
through Gini coefficients (Feng \& Timmermans, 2014) or composite
indices (Agost-Felip et al., 2021; Corazza et al., 2020; Miranda \& da
Silva, 2012), occasionally incorporating spatial analysis (Carrier et
al., 2014; Jephcote \& Chen, 2013). Many use established thresholds or
health guidelines e.g., WHO guidelines for Active Aging and targets
included in the United Nations' SDG (Agost-Felip et al., 2021) and OECD
or WHO standards for traffic noise level, \(NO_2\) levels, available
green space and \(PM_{2.5}\) levels (Apparicio et al., 2021; Iungman et
al., 2021; Khomenko et al., 2020; Kruize et al., 2007; Mueller et al.,
2018). These metrics are sometimes criticized for their general
applicability. Nonetheless, their use provides interpretable values for
tracking progress, offering comparability across communities, unlike
accessibility measures, which vary in methods and assumptions.

Appendix Section~\ref{sec-sect63} presents detailed examples of various
standards, outlined within the 5WH framework, as derived from the
examined corpus.

\section{Moving forward: discussion and calls for
action}\label{moving-forward-discussion-and-calls-for-action}

As previously noted, this work makes two contributions to the
literature. First, it outlines a conceptual and flexible ``Why'',
Where'', ``When'', ``Who'', ``What'' and ``How'' (5WH) framework with
supporting definitions for approaching questions about transportation
justice. Second, it applies the framework in synthesizing and appraising
of 165 academic articles drawn from the transportation justice
literature; these studies use equity analysis as instruments to gauge
proximity to their context's just situation. Based on the preceding
sections of what is in the literature, and what is lacking from our
perspective, we conclude with the following five calls to action. Based
on the preceding sections of what is in the literature, and what is
lacking from our perspective, we conclude with the following five calls
to action.

\subsection{Call 1: Underpin tailored standards based on rigorous
concepts of
justice}\label{call-1-underpin-tailored-standards-based-on-rigorous-concepts-of-justice}

The conceptual grounding for standards is often left implicit within the
literature. For example, Mueller et al. (2018) suggests the relative
risk of mortality from transport-related air pollution should not be
higher in deprived groups than the general population. While the
conceptual justice foundations are not explicitly declared, we infer an
egalitarian focus. But, what level of relative risk is acceptable for
the general population? What purpose does mortality risk serve and who
benefits from it? To move equity analysis outputs towards just transport
futures, explicit inclusion of justice rationales (the ``Why?'') should
become more common place. We must be clear with our terms--what is
equity and for whom?

Standards are statements of fairness, yet some standards are seemingly
arbitrary in the reviewed literature--in other words, set arbitrary
justice goals. For example, Cao \& Stanley (2017) proposes 20 ferries
per day to avoid social exclusion for inter-island transport planning,
though acknowledging the standard should be politically determined. It
is unclear if selecting a different benchmark such as 10 or 30 ferries
would make a difference in any specific object of justice (e.g.,
accessibility to particular destinations) or if that number is tied to
funding or resource constraints. Another example is the conceptual
underpinnings of the 15-minute city, where 15 minutes is the standard
for travel times. Is this sufficientarism? Or egalitarianism? The
standard can be interpreted in a multitude of ways. We argue that
conceptualization of justice must guide the selection of the standard.
As Martens \& Golub (2021) demonstrate, being explicit about ``Why'' is
important. Recent work by Karner et al. (2024) provides several concrete
examples for developing standards that follow directly from different
justice conceptualizations.

Additionally, there is a need to go beyond the focus on low-income
transit riders, as clear ``Who?''-specific standards are lacking.
Positioning intersectional considerations within the creation of
community-based equity definitions and tailored standards are critical,
as the act of delineating equity-deserving communities impacts results
(e.g., Rowangould et al. (2015)). Standards should be sensitive to
evolving community-based definitions of inequities. Are issues of
economic inequity at the root of transport inequities for a particular
community? Are inequities in access provided by certain modes emphasized
because access offered by other modes are \emph{relatively} better?
Access to what types of opportunities are driving transport inequities?
How do populations, travel modes, and opportunities sought intersect to
define the ``who'' of inequities? A community-informed understanding of
inequities and tracking how they change are needed.

For decision-makers, setting standards, measuring inequities, and
developing flexible guidelines for standards that are also compatible
with community-informed calls for justice, are the next steps for
transportation equity planning. If we assume that one function of the
academic literature is to recommend standards, researchers must connect
compatible conceptualisations and standards to justice frameworks.
Currently, these fundamental connections are mostly missing.

\subsection{Call 2: Develop creative methods for systems-thinking
approaches to
fairness}\label{call-2-develop-creative-methods-for-systems-thinking-approaches-to-fairness}

On the methodological side, transportation justice research would
benefit from a wider use of mixed-methods. Concepts and standards are
often discussed from purely qualitative or quantitative approaches. This
is a missed opportunity to combine the strengths of both approaches,
whether by diving deeper into some particular experiences or perceptions
through qualitative methods or tailoring more meaningful quantitative
analysis after qualitative explorations. As mixed-method examples, Xie
\& Spinney (2018) find through interviews and go-alongs with women
cyclists that the standard Cycling Level of Service tools used by
engineers to plan cycling infrastructure misses a critical gendered
perspective. Further, Somenahalli \& Taylor (2007) survey older adults
to understand their mobility issues, revealing factors that are unseen
in standard daily travel surveys.

While disparity analysis is frequently used, the resulting standards do
not often align with practical applications. For example, metrics of
accessibility (usually measured with travel impedance cut-offs of
between 15 to 60 minutes depending on the destination, population group
and mode) are used to show descriptive differences among areas and
groups but with scarce implications as to the experience of travelers.
How low must an accessibility index be before it is too low? Relatedly,
once sufficient accessibility levels are reached, excessively high
accessibility values can result in high inequalities. Are inequalities
of a certain level an issue? A lack of discussion poses a challenge to
translating results from these analyses into policy and practice.
Creative methods and discussions of quantitative accessibility metrics
should be paired with results; they should yield interpretable results.
The explicit discussion of minimum \emph{and} maximum values in the
distribution of the object of justice (the ``What?''), as applicable, is
critical.

Other times, when analysis engages with metrics that may be tied to
particular concepts of equity (like the Gini coefficient or Theil
index), they fall short in assessing whether the results are good or bad
(Mijares et al., 2013). For example, a Gini coefficient of 0 would mean
that all people have the same access to public transport stops. But is
this level of access good enough or lacking? And what does it mean if
the result is 0.2, 0.3, or 0.4? Is this good or bad news? How should a
decision-maker interpret this? Are new policies needed to reduce that
number to a certain threshold, orienting future interventions? These
questions usually remain unanswered in the literature despite their
importance. These measurements can also bring some challenges and
pitfalls as summarised by Karner et al. (2024) but are necessary for
more effective equity analyses.

\subsection{Call 3: Making data available is a matter of
justice}\label{call-3-making-data-available-is-a-matter-of-justice}

In our review of the literature, we were left wondering specifically--
what are the motivations for the inclusion of some destinations and the
choice to not include others? For instance, leisure (e.g., green space,
parks, recreation) and certain care destinations (e.g., community
organizations, government services, banking locations, daycares) are
infrequently studied or missing altogether. We suspect that since the
methods used in the reviewed literature are predominately quantitative,
the reliance on commonly used point of interest databases is also high.
These databases typically include education, health, and occasionally
include aggregated categories for leisure and community destinations,
but are less generous for points of interest that are more often
associated with understudied mobility patterns.

However, through the reviewed literature, we know that transport systems
are more than just facilitators to work or as a source of economic
development (though in underdeveloped regions, transport systems as a
force of economic development is pronounced, e.g., high-speed rail (Z.
Chen \& Haynes, 2017; H. Kim \& Sultana, 2015; Monzon et al., 2013)). As
such, data availability matters. In the context of the Global South and
rural geographies fewer official data sources and public research
resources exist relative to urban communities and areas in the Global
North. The operationalization of emerging conceptual theories in equity
analysis, such as sufficientariansm (van der Veen et al., 2020), might
be impacted if relevant data is not available.

The calls for and relevant issues of data availability are not new, but
they have at least three pieces. What and who is the subject of justice?
When/Where/How is it measured? And, who gets to consult and use the
data? Deciding who is the subject of justice frames the data collection
activity; if it is the mobility of those who do domestic work, the
classification of who does this work and how it changes over time/space
is fundamental. How we classify mobility and access has implications for
our understanding of just who is a subject of justice, as illustrated by
the history of racial classifications in the U.S. (Lee, 1993). The
methods we use, the spatio-temporal boundaries for data collection, and
who has (and does not have) access to the collected data, can all impact
how we think about justice. In the case of transportation, the issue of
data collection/availability as a matter of justice is gaining traction
as digitization of data casts a starker light on these questions
(Behrendt \& Sheller, 2023; Sourbati \& Behrendt, 2021).

\subsection{Call 4: Develop more direct and explicit links between
standards and lived
experiences}\label{call-4-develop-more-direct-and-explicit-links-between-standards-and-lived-experiences}

Robust assessments of the implications of standards on lived experiences
are still lacking in the literature, but are essential for equity
analysis that translates into just practice. While the estimation of the
benefits of increased mobility or accessibility, or reducing burdens is
commonplace in the literature, there is a need to associate them with
outcomes, including life and neighbourhood satisfaction, subjective
well-being, mental and physical health, and social capital. The
literature must move beyond describing inequities and their contributing
factors. Instead, it should incorporate causal methods that establish
clear cause-and-effect relationships, enabling policymakers to quantify
the impact of various policies and infrastructure decisions on advancing
justice.

Composite measures such as the transport-land-use index proposed by
Appleyard et al. (2017) is an example of systems-thinking approach that
links findings to quality-of-life proxies; they ground their measure in
the principle of \emph{livability} along corridors of varying levels of
estimated transport-land-use integration. However, the methods could go
further: they could be tied to absolute goals of integration or
livability. Relatably, Higgs et al. (2019) develops an urban livability
index and demonstrates the relationship between the population's use of
a certain mode in a neighbourhood and a one unit increase in the index.
However, are there absolute minimums or maximums for the index or the
mode-choice goals that should not be crossed?

More explicit discussion of the boundaries in the distribution of the
object of justice (the ``What?'') alongside these creative methods are
needed. These links may be used to track progress towards justice across
time and space, a critical point for practitioners.

\subsection{Call 5: Rigorously evaluate interventions and
policies}\label{call-5-rigorously-evaluate-interventions-and-policies}

There is a need to evaluate equity interventions and policies: track
their before, during, and after. In this review, a small proportion of
studies assess specific transport-related policy interventions through
an equity lens. Examples include mode-shift from driving to active
school travel (Mammen et al., 2014), transit fare restructures (Hickey
et al., 2010) and spatial analysis of Low Traffic neighbourhoods (Aldred
et al., 2021). The assessment of interventions and tracking associated
outcomes can be thought of as a key step towards transport justice.
Similarly, for how long the burdens and benefits can still be associated
to a specific transportation intervention remains a critical avenue to
explore.

Outcomes of interventions can be compared within and between
communities, and cross community comparisons can be created that may
expedite the adoption of effective policies that move towards just
outcomes; the presence of these synthesis and comparative studies could
support brave decision-makers in the application of research into
practice. Again, causal methods and mixed methods approaches might serve
as valuable approaches to reach this in practice.

\section{References}\label{references}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-abasoloEquityUtilizationAccess2001a}
Abasolo, I., Manning, R., \& Jones, A. M. (2001). Equity in utilization
of and access to public-sector {GPs} in {Spain}. \emph{Applied
Economics}, \emph{33}(3), 349--364.
\url{https://doi.org/10.1080/00036840122511}

\bibitem[\citeproctext]{ref-adlakhaMindGapGender2020}
Adlakha, D., \& Parra, D. (2020). Mind the gap: {Gender} differences in
walkability, transportation and physical activity in urban {India}.
\emph{Journal of Transport \& Health}, \emph{18}.
\url{https://doi.org/10.1016/j.jth.2020.100875}

\bibitem[\citeproctext]{ref-adliRightCityApplying2018}
Adli, S., \& Donovan, S. (2018). Right to the city: {Applying} justice
tests to public transport investments. \emph{Transport Policy},
\emph{66}, 56--65. \url{https://doi.org/10.1016/j.tranpol.2018.03.005}

\bibitem[\citeproctext]{ref-agostfelipInclusiveModelAssessing2021}
Agost-Felip, R., Rua, M., \& Kouidmi, F. (2021). An {Inclusive Model}
for {Assessing Age-Friendly Urban Environments} in {Vulnerable Areas}.
\emph{Sustainability}, \emph{13}(15).
\url{https://doi.org/10.3390/su13158352}

\bibitem[\citeproctext]{ref-albertsRebuildingWomenLivelihoods2016}
Alberts, A., Pfeffer, K., \& Baud, I. (2016). Rebuilding women's
livelihoods strategies at the city fringe: {Agency}, spatial practices,
and access to transportation from {Semmencherry}, {Chennai}.
\emph{Journal of Transport Geography}, \emph{55}, 142--151.
\url{https://doi.org/10.1016/j.jtrangeo.2015.11.004}

\bibitem[\citeproctext]{ref-aldredEquityNewActive2021}
Aldred, R., Verlinghieri, E., Sharkey, M., Itova, I., \& Goodman, A.
(2021). Equity in new active travel infrastructure: {A} spatial analysis
of {London}'s new {Low Traffic Neighbourhoods}. \emph{Journal of
Transport Geography}, \emph{96}.
\url{https://doi.org/10.1016/j.jtrangeo.2021.103194}

\bibitem[\citeproctext]{ref-aljoufie2016}
Aljoufie, M. (2016). \emph{URBAN PLANNING AND ARCHITECTURAL DESIGN FOR
SUSTAINABLE DEVELOPMENT (UPADSD)} (F. Naselli, F. Pollice, \& M. Amer,
Eds.; Vol. 216, pp. 535--544).
\url{https://doi.org/10.1016/j.sbspro.2015.12.013}

\bibitem[\citeproctext]{ref-allenSizingTransportPoverty2019a}
Allen, J., \& Farber, S. (2019). Sizing up transport poverty: {A}
national scale accounting of low-income households suffering from
inaccessibility in {Canada}, and what to do about it. \emph{Transport
Policy}, \emph{74}, 214--223.
\url{https://doi.org/10.1016/j.tranpol.2018.11.018}

\bibitem[\citeproctext]{ref-ampeImpactChildBike2020}
Ampe, T., de Geus, B., Walker, I., Serrien, B., Truyen, B., Durlet, H.,
\& Meeusen, R. (2020). The impact of a child bike seat and trailer on
the objective overtaking behaviour of motorized vehicles passing
cyclists. \emph{TRANSPORTATION RESEARCH PART F-TRAFFIC PSYCHOLOGY AND
BEHAVIOUR}, \emph{75}, 55--65.
\url{https://doi.org/10.1016/j.trf.2020.09.014}

\bibitem[\citeproctext]{ref-apparicioCyclingOneMost2021}
Apparicio, P., Gelb, J., Jarry, V., \& Lesage-Mann, E. (2021). Cycling
in one of the most polluted cities in the world: {Exposure} to noise and
air pollution and potential adverse health impacts in {Delhi}.
\emph{International Journal of Health Geographics}, \emph{20}(1).
\url{https://doi.org/10.1186/s12942-021-00272-2}

\bibitem[\citeproctext]{ref-appleyardTransitCorridorLivability2017}
Appleyard, B., Ferrell, C., \& Taecker, M. (2017). Transit {Corridor
Livability Realizing} the {Potential} of {Transportation} and {Land Use
Integration}. \emph{Transportation Research Record}, \emph{2671},
20--30. \url{https://doi.org/10.3141/2671-03}

\bibitem[\citeproctext]{ref-archerWhite2020}
Archer, D. N. (2020). "White men's roads through black men's homes":
Advancing racial equity through highway reconstruction. \emph{Vand. L.
Rev.}, \emph{73}, 1259.

\bibitem[\citeproctext]{ref-arranz-lopezSocialSpatialEquity2019}
Arranz-Lopez, A., Soria-Lara, J., \& Pueyo-Campos, A. (2019). Social and
spatial equity effects of non-motorised accessibility to retail.
\emph{CITIES}, \emph{86}, 71--82.
\url{https://doi.org/10.1016/j.cities.2018.12.012}

\bibitem[\citeproctext]{ref-auchinclossDesignBaselineDescription2020}
Auchincloss, A., Michael, Y., Fuller, D., Li, S., Niamatullah, S.,
Fillmore, C., Setubal, C., \& Bettigole, C. (2020). Design and baseline
description of a cohort of bikeshare users in the city of
{Philadelphia}. \emph{Journal of Transport \& Health}, \emph{16}.
\url{https://doi.org/10.1016/j.jth.2020.100836}

\bibitem[\citeproctext]{ref-awuorInfluenceEnvironmentalHealth2019}
Awuor, L., \& Melles, S. (2019). The influence of environmental and
health indicators on premature mortality: {An} empirical analysis of the
{City} of {Toronto}'s 140 neighborhoods. \emph{Health \& Place},
\emph{58}. \url{https://doi.org/10.1016/j.healthplace.2019.102155}

\bibitem[\citeproctext]{ref-basuPracticalFrameworkBenchmarking2019}
Basu, R., \& Alves, B. (2019). Practical {Framework} for {Benchmarking}
and {Impact Evaluation} of {Public Transportation Infrastructure}:
{Case} of {Belo Horizonte}, {Brazil}. \emph{Transportation Research
Record}, \emph{2673}(3), 711--721.
\url{https://doi.org/10.1177/0361198119835528}

\bibitem[\citeproctext]{ref-batscheUsingGISEnhance2012}
Batsche, C., \& Reader, S. (2012). Using {GIS} to enhance programs
serving emancipated youth leaving foster care. \emph{EVALUATION AND
PROGRAM PLANNING}, \emph{35}(1), 25--33.
\url{https://doi.org/10.1016/j.evalprogplan.2011.06.003}

\bibitem[\citeproctext]{ref-behrendtData2023}
Behrendt, F., \& Sheller, M. (2023). Mobility data justice {[}Journal
Article{]}. \emph{Mobilities}, 1--19.
\url{https://doi.org/10.1080/17450101.2023.2200148}

\bibitem[\citeproctext]{ref-benevenutoExaminingTransportNeeds2020}
Benevenuto, R., \& Caulfield, B. (2020). Examining transport needs in
the global south using a screening framework. \emph{Journal of Transport
Geography}, \emph{88}.
\url{https://doi.org/10.1016/j.jtrangeo.2020.102845}

\bibitem[\citeproctext]{ref-berheAdaptationDissonanceQuality2014}
Berhe, R., Martinez, J., \& Verplanke, J. (2014). Adaptation and
{Dissonance} in {Quality} of {Life}: {A Case Study} in {Mekelle},
{Ethiopia}. \emph{Social Indicators Research}, \emph{118}(2), 535--554.
\url{https://doi.org/10.1007/s11205-013-0448-y}

\bibitem[\citeproctext]{ref-berryDistributionalEffectsCarbon2019}
Berry, A. (2019). The distributional effects of a carbon tax and its
impact on fuel poverty: {A} microsimulation study in the {French}
context. \emph{Energy Policy}, \emph{124}, 81--94.
\url{https://doi.org/10.1016/j.enpol.2018.09.021}

\bibitem[\citeproctext]{ref-berryInvestigatingFuelPoverty2016}
Berry, A., Jouffe, Y., Coulombel, N., \& Guivarch, C. (2016).
Investigating fuel poverty in the transport sector: {Toward} a composite
indicator of vulnerability. \emph{Energy Research \& Social Science},
\emph{18}, 7--20. \url{https://doi.org/10.1016/j.erss.2016.02.001}

\bibitem[\citeproctext]{ref-bertrandMeasuringMappingDisparities2008}
Bertrand, L., Therien, F., \& Cloutier, M. (2008). Measuring and mapping
disparities in access to fresh fruits and vegetables in {Montreal}.
\emph{CANADIAN JOURNAL OF PUBLIC HEALTH-REVUE CANADIENNE DE SANTE
PUBLIQUE}, \emph{99}(1), 6--11. \url{https://doi.org/10.1007/BF03403732}

\bibitem[\citeproctext]{ref-bharathyRevisitingClearFloor2018}
Bharathy, A., \& D'Souza, C. (2018). Revisiting {Clear Floor Area
Requirements} for {Wheeled Mobility Device Users} in {Public
Transportation}. \emph{Transportation Research Record: Journal of the
Transportation Research Board}, \emph{2672}(8), 675--685.
\url{https://doi.org/10.1177/0361198118787082}

\bibitem[\citeproctext]{ref-bierbaumMobilityJusticeLinking2021}
Bierbaum, A. H., Karner, A., \& Barajas, J. M. (2021). Toward mobility
justice: Linking transportation and education equity in the context of
school choice. \emph{Journal of the American Planning Association},
\emph{87}(2), 197--210.
\url{https://doi.org/10.1080/01944363.2020.1803104}

\bibitem[\citeproctext]{ref-billsLookingMeanEquity2017}
Bills, T. S., \& Walker, J. L. (2017). Looking beyond the mean for
equity analysis: Examining distributional impacts of transportation
improvements. \emph{Transport Policy}, \emph{54}, 61--69.
\url{https://doi.org/10.1016/j.tranpol.2016.08.003}

\bibitem[\citeproctext]{ref-blecicCapabilitywiseWalkabilityEvaluation2021}
Blecic, I., Cecchini, A., Congiu, T., Fancello, G., Talu, V., \&
Trunfio, G. (2021). Capability-wise walkability evaluation as an
indicator of urban peripherality. \emph{ENVIRONMENT AND PLANNING B-URBAN
ANALYTICS AND CITY SCIENCE}, \emph{48}(4), 895--911.
\url{https://doi.org/10.1177/2399808320908294}

\bibitem[\citeproctext]{ref-bocarejoImpactsFormalizationIntegration2022}
Bocarejo, J. P., \& Urrego, L. F. (2022). The impacts of formalization
and integration of public transport in social equity: {The} case of
{Bogota}. \emph{Research in Transportation Business \& Management},
\emph{42}. \url{https://doi.org/10.1016/j.rtbm.2020.100560}

\bibitem[\citeproctext]{ref-boisjolyHowGetThere2017}
Boisjoly, G., \& El-Geneidy, A. M. (2017). How to get there? {A}
critical assessment of accessibility objectives and indicators in
metropolitan transportation plans. \emph{Transport Policy}, \emph{55},
38--50. \url{https://doi.org/10.1016/j.tranpol.2016.12.011}

\bibitem[\citeproctext]{ref-boisjolyAccessibilityMeasurementsSao2020}
Boisjoly, G., Serra, B., Oliveira, G. T., \& El-Geneidy, A. (2020).
Accessibility measurements in {Sao Paulo}, {Rio} de {Janeiro},
{Curitiba} and {Recife}, {Brazil}. \emph{Journal of Transport
Geography}, \emph{82}.
\url{https://doi.org/10.1016/j.jtrangeo.2019.102551}

\bibitem[\citeproctext]{ref-braithwaite2003principles}
Braithwaite, J. et al. (2003). Principles of restorative justice.
\emph{Restorative Justice and Criminal Justice: Competing or
Reconcilable Paradigms}, 1--20.

\bibitem[\citeproctext]{ref-brusselAccessAccessibilityCritique2019}
Brussel, M., Zuidgeest, A., Pfeffer, K., \& van Maarseveen, M. (2019).
Access or {Accessibility}? {A Critique} of the {Urban Transport SDG
Indicator}. \emph{ISPRS International Journal of Geo-Information},
\emph{8}(2). \url{https://doi.org/10.3390/ijgi8020067}

\bibitem[\citeproctext]{ref-caoIndicatorsSocioSpatialTransport2017}
Cao, D., \& Stanley, J. (2017). Indicators of {Socio-Spatial Transport
Disadvantage} for {Inter-Island Transport Planning} in {Rural Philippine
Communities}. \emph{Social Inclusion}, \emph{5}(4), 116--131.
\url{https://doi.org/10.17645/si.v5i4.1098}

\bibitem[\citeproctext]{ref-carrierApplicationThreeMethods2014}
Carrier, M., Apparicio, P., Seguin, A., \& Crouse, D. (2014). The
application of three methods to measure the statistical association
between different social groups and the concentration of air pollutants
in {Montreal}: {A} case of environmental equity. \emph{Transportation
Research Part D: Transport and Environment}, \emph{30}, 38--52.
\url{https://doi.org/10.1016/j.trd.2014.05.001}

\bibitem[\citeproctext]{ref-chancelCarbonInequalityKyoto2015}
Chancel, L., \& Piketty, T. (2015). \emph{Carbon and inequality: From
kyoto to paris}.

\bibitem[\citeproctext]{ref-chenExploringTravelCharacteristics2010}
Chen, W.-H. (2010). Exploring travel characteristics and factors
affecting the degree of willingness of seniors in taiwan to use an
alternative service bus. \emph{{TRANSPORTATION} {RESEARCH} {RECORD}},
\emph{2182}, 71--78. \url{https://doi.org/10.3141/2182-10}

\bibitem[\citeproctext]{ref-chenSpatialAnalysisFramework2020}
Chen, Y., Bouferguene, A., Shirgaokar, M., \& Al-Hussein, M. (2020).
Spatial {Analysis Framework} for {Age-Restricted Communities Integrating
Spatial Distribution} and {Accessibility Evaluation}. \emph{JOURNAL OF
URBAN PLANNING AND DEVELOPMENT}, \emph{146}(1).
\url{https://doi.org/10.1061/(ASCE)UP.1943-5444.0000537}

\bibitem[\citeproctext]{ref-chenImpactHighspeedRail2017}
Chen, Z., \& Haynes, K. (2017). Impact of high-speed rail on regional
economic disparity in {China}. \emph{Journal of Transport Geography},
\emph{65}, 80--91. \url{https://doi.org/10.1016/j.jtrangeo.2017.08.003}

\bibitem[\citeproctext]{ref-chengInvestigatingWalkingAccessibility2019}
Cheng, L., Caset, F., De Vos, J., Derudder, B., \& Witlox, F. (2019).
Investigating walking accessibility to recreational amenities for
elderly people in {Nanjing}, {China}. \emph{Transportation Research Part
D: Transport and Environment}, \emph{76}, 85--99.
\url{https://doi.org/10.1016/j.trd.2019.09.019}

\bibitem[\citeproctext]{ref-chiscanoImprovingDesignUrban2021}
Chiscano, M. (2021). Improving the design of urban transport experience
with people with disabilities. \emph{Research in Transportation Business
and Management}, \emph{41}.
\url{https://doi.org/10.1016/j.rtbm.2020.100596}

\bibitem[\citeproctext]{ref-choiFoodDesertsActivity2013}
Choi, Y., \& Suzuki, T. (2013). Food deserts, activity patterns, \&
social exclusion: {The} case of {Tokyo}, {Japan}. \emph{Applied
Geography}, \emph{43}, 87--98.
\url{https://doi.org/10.1016/j.apgeog.2013.05.009}

\bibitem[\citeproctext]{ref-churchillTransportPovertySubjective2019}
Churchill, S., \& Smyth, R. (2019). Transport poverty and subjective
wellbeing. \emph{Transportation Research Part A: Policy and Practice},
\emph{124}, 40--54. \url{https://doi.org/10.1016/j.tra.2019.03.004}

\bibitem[\citeproctext]{ref-corazzaMethodologyEvidenceCase2020}
Corazza, M. V., D'Alessandro, D., Di Mascio, P., \& Moretti, L. (2020).
Methodology and evidence from a case study in {Rome} to increase
pedestrian safety along home-to-school routes. \emph{Journal of Traffic
and Transportation Engineering (English Edition)}, \emph{7}(5), pp
715--727. \url{https://doi.org/10.1016/j.jtte.2020.03.003}

\bibitem[\citeproctext]{ref-covidenceCovidence2023}
Covidence. (2023). \emph{Covidence} {[}Computer software{]}.
\url{https://www.covidence.org/}

\bibitem[\citeproctext]{ref-coxQualitativeStudyAccessibility2020}
Cox, B., \& Bartle, C. (2020). A qualitative study of the accessibility
of a typical {UK} town cycle network to disabled cyclists. \emph{Journal
of Transport \& Health}, \emph{19}.
\url{https://doi.org/10.1016/j.jth.2020.100954}

\bibitem[\citeproctext]{ref-cuiSpatialAccessPublic2020}
Cui, B., Boisjoly, G., Wasfi, R., Orpana, H., Manaugh, K., Buliung, R.,
Kestens, Y., \& El-Geneidy, A. (2020). Spatial {Access} by {Public
Transport} and {Likelihood} of {Healthcare Consultations} at
{Hospitals}. \emph{Transportation Research Record: Journal of the
Transportation Research Board}, \emph{2674}(12), 188--198.
\url{https://doi.org/10.1177/0361198120952793}

\bibitem[\citeproctext]{ref-daamenAssessingGapPublic2008}
Daamen, W., de Boer, E., \& de Kloe, R. (2008). Assessing the {Gap
Between Public Transport Vehicles} and {Platforms} as a {Barrier} for
the {Disabled Use} of {Laboratory Experiments}. \emph{Transportation
Research Record: Journal of the Transportation Research Board},
\emph{2072}, 131--138. \url{https://doi.org/10.3141/2072-14}

\bibitem[\citeproctext]{ref-danielsProposalAccessibilityPlanning2011}
Daniels, R., \& Mulley, C. (2011). \emph{A proposal for accessibility
planning in {NSW}: Research and policy issues.} 16p.
\url{https://trid.trb.org/view/1105622}

\bibitem[\citeproctext]{ref-delboscTransportProblemsThat2011}
Delbosc, A., \& Currie, G. (2011a). Transport problems that matter -
social and psychological links to transport disadvantage. \emph{Journal
of Transport Geography}, \emph{19}(1), 170--178.
\url{https://doi.org/10.1016/j.jtrangeo.2010.01.003}

\bibitem[\citeproctext]{ref-delboscUsing2011}
Delbosc, A., \& Currie, G. (2011b). Using lorenz curves to assess public
transport equity {[}Journal Article{]}. \emph{Journal of Transport
Geography}, \emph{19}(6), 1252--1259.
\url{https://doi.org/10.1016/j.jtrangeo.2011.02.008}

\bibitem[\citeproctext]{ref-desjardinsGoing2021}
Desjardins, E., Apatu, E., Razavi, S. D., Higgins, C. D., Scott, D. M.,
\& Páez, A. (2021). {``Going through a little bit of growing pains''}: A
qualitative study of the factors that influence the route choice of
regular bicyclists in a developing cycling city {[}Journal Article{]}.
\emph{Transportation Research Part F: Traffic Psychology and Behaviour},
\emph{81}, 431--444.
https://doi.org/\url{https://doi.org/10.1016/j.trf.2021.06.005}

\bibitem[\citeproctext]{ref-desjardinsChildren2022}
Desjardins, E., Tavakoli, Z., Páez, A., \& Waygood, E. O. (2022).
\emph{Children's access to non-school destinations by active or
independent travel: A scoping review} (Electronic Article 19; Vol. 19).
\url{https://doi.org/10.3390/ijerph191912345}

\bibitem[\citeproctext]{ref-doranPursuitCyclingEquity2021}
Doran, A., El-Geneidy, A., \& Manaugh, K. (2021). The pursuit of cycling
equity: {A} review of {Canadian} transport plans. \emph{Journal of
Transport Geography}, \emph{90}, 102927.
\url{https://doi.org/10.1016/j.jtrangeo.2020.102927}

\bibitem[\citeproctext]{ref-eliassonCongestionPricingFair2016}
Eliasson, J. (2016). Is congestion pricing fair? Consumer and citizen
perspectives on equity effects. \emph{Transport Policy}, \emph{52},
1--15. \url{https://doi.org/10.1016/j.tranpol.2016.06.009}

\bibitem[\citeproctext]{ref-eppenbergerOpportunitySharedAutonomous2021}
Eppenberger, N., \& Richter, M. (2021). The opportunity of shared
autonomous vehicles to improve spatial equity in accessibility and
socio-economic developments in {European} urban areas. \emph{European
Transport Research Review}, \emph{13}(1).
\url{https://doi.org/10.1186/s12544-021-00484-4}

\bibitem[\citeproctext]{ref-evansAccessibilityUserNeeds2015}
Evans, G. (2015). Accessibility and user needs: Pedestrian mobility and
urban design in the {UK}. \emph{PROCEEDINGS OF THE INSTITUTION OF CIVIL
ENGINEERS-MUNICIPAL ENGINEER}, \emph{168}(1), 32--44.
\url{https://doi.org/10.1680/muen.14.00012}

\bibitem[\citeproctext]{ref-fakhrmoosaviIncorporatingTravelTime2021}
Fakhrmoosavi, F., Zockaie, A., \& Abdelghany, K. (2021). Incorporating
{Travel Time Reliability} in {Equitable Congestion Pricing Schemes} for
{Heterogeneous Users} and {Bimodal Networks}. \emph{Transportation
Research Record: Journal of the Transportation Research Board},
\emph{2675}(11), 754--768.
\url{https://doi.org/10.1177/03611981211019737}

\bibitem[\citeproctext]{ref-falavignaAssessingInequalitiesPublic2016}
Falavigna, C., \& Hernandez, D. (2016). Assessing inequalities on public
transport affordability in two latin {American} cities: {Montevideo}
({Uruguay}) and {Cordoba} ({Argentina}). \emph{Transport Policy},
\emph{45}, 145--155. \url{https://doi.org/10.1016/j.tranpol.2015.09.011}

\bibitem[\citeproctext]{ref-feeleyValidationParatransitSkills2019}
Feeley, C. (2019). Validation of the {Paratransit Skills Assessment} for
{Paratransit Travel} and {Mobility} of {Adults} on the {Autism
Spectrum}. \emph{Transportation Research Record: Journal of the
Transportation Research Board}, \emph{2673}(5), 759--769.
\url{https://doi.org/10.1177/0361198119839342}

\bibitem[\citeproctext]{ref-fengTradeoffsMobilityEquity2014}
Feng, T., \& Timmermans, H. J. P. (2014). Trade-offs between mobility
and equity maximization under environmental capacity constraints: {A}
case study of an integrated multi-objective model. \emph{Transportation
Research Part C: Emerging Technologies}, \emph{43, Part 3}, pp 267--279.
\url{https://doi.org/10.1016/j.trc.2014.03.012}

\bibitem[\citeproctext]{ref-ferenchakEquityAnalysisProactively2019}
Ferenchak, N., \& Marshall, W. (2019). Equity {Analysis} of
{Proactively-} vs. {Reactively-Identified Traffic Safety Issues}.
\emph{Transportation Research Record: Journal of the Transportation
Research Board}, \emph{2673}(7), 596--606.
\url{https://doi.org/10.1177/0361198119841296}

\bibitem[\citeproctext]{ref-fothEquitableTransitExamining2013a}
Foth, N., Manaugh, K., \& El-Geneidy, A. (2013). Towards equitable
transit: Examining transit accessibility and social need in {Toronto},
{Canada}, 1996-2006. \emph{Journal of Transport Geography}, \emph{29},
1--10. \url{https://doi.org/10.1016/j.jtrangeo.2012.12.008}

\bibitem[\citeproctext]{ref-friedMeasuringSustainableDevelopment2020}
Fried, T., Tun, T., Klopp, J., \& Welle, B. (2020). Measuring the
{Sustainable Development Goal} ({SDG}) {Transport Target} and
{Accessibility} of {Nairobi}'s {Matatus}. \emph{Transportation Research
Record: Journal of the Transportation Research Board}, \emph{2674}(5),
196--207. \url{https://doi.org/10.1177/0361198120914620}

\bibitem[\citeproctext]{ref-furstBestBadPractices2013}
Fürst, E., \& Vogelauer, C. (2013). \emph{Best and bad practices in
public transport: Approaches to a barrier-free design for the visually
and hearing impaired}. 29p.
\url{https://aetransport.org/past-etc-papers/conference-papers-2013https://trid.trb.org/view/1330058}

\bibitem[\citeproctext]{ref-golubUsingPrinciplesJustice2014}
Golub, A., \& Martens, K. (2014). Using principles of justice to assess
the modal equity of regional transportation plans. \emph{Journal of
Transport Geography}, \emph{41}, 10--20.
\url{https://doi.org/10.1016/j.jtrangeo.2014.07.014}

\bibitem[\citeproctext]{ref-gosslingUrban2016}
Gössling, S. (2016). Urban transport justice {[}Journal Article{]}.
\emph{Journal of Transport Geography}, \emph{54}, 1--9.
https://doi.org/\url{https://doi.org/10.1016/j.jtrangeo.2016.05.002}

\bibitem[\citeproctext]{ref-gosslingSocial2019}
Gössling, S., Choi, A., Dekker, K., \& Metzler, D. (2019). The social
cost of automobility, cycling and walking in the european union
{[}Journal Article{]}. \emph{Ecological Economics}, \emph{158}, 65--74.
https://doi.org/\url{https://doi.org/10.1016/j.ecolecon.2018.12.016}

\bibitem[\citeproctext]{ref-guoSystematicOverviewTransportation2020}
Guo, Y., Chen, Z., Stuart, A., Li, X., \& Zhang, Y. (2020). A systematic
overview of transportation equity in terms of accessibility, traffic
emissions, and safety outcomes: {From} conventional to emerging
technologies. \emph{Transportation Research Interdisciplinary
Perspectives}, \emph{4}, 100091.
\url{https://doi.org/10.1016/j.trip.2020.100091}

\bibitem[\citeproctext]{ref-hailConceptFairnessRelation2021}
Hail, Y., \& McQuaid, R. (2021). The concept of fairness in relation to
women transport users. \emph{Sustainability}, \emph{13}(5), 2919.
\url{https://doi.org/10.3390/su13052919}

\bibitem[\citeproctext]{ref-hickeyUsingQuantitativeMethods2010}
Hickey, R., Lu, A., \& Reddy, A. (2010). Using {Quantitative Methods} in
{Equity} and {Demographic Analysis} to {Inform Transit Fare
Restructuring Decisions}. \emph{Transportation Research Record: Journal
of the Transportation Research Board}, \emph{2144}, 80--92.
\url{https://doi.org/10.3141/2144-10}

\bibitem[\citeproctext]{ref-higgsUrbanLiveabilityIndex2019}
Higgs, C., Badland, H., Simons, K., Knibbs, L., \& Giles-Corti, B.
(2019). The {Urban Liveability Index}: Developing a policy-relevant
urban liveability composite measure and evaluating associations with
transport mode choice. \emph{International Journal of Health
Geographics}, \emph{18}. \url{https://doi.org/10.1186/s12942-019-0178-8}

\bibitem[\citeproctext]{ref-iungmanImpactUrbanTransport2021}
Iungman, T., Khomenko, S., Nieuwenhuijsen, M., Barboza, E., Ambros, A.,
Padilla, C., \& Mueller, N. (2021). The impact of urban and transport
planning on health: {Assessment} of the attributable mortality burden in
{Madrid} and {Barcelona} and its distribution by socioeconomic status.
\emph{ENVIRONMENTAL RESEARCH}, \emph{196}.
\url{https://doi.org/10.1016/j.envres.2021.110988}

\bibitem[\citeproctext]{ref-jacquesRescuing2012}
Jacques, C., Manaugh, K., \& El-Geneidy, A. (2012). Rescuing the captive
{[}mode{]} user: An alternative approach to transport market
segmentation {[}Journal Article{]}. \emph{Transportation}, 1--21.
\url{https://doi.org/10.1007/s11116-012-9437-2}

\bibitem[\citeproctext]{ref-jaggarPhilosophicalChallengesGlobal2009}
Jaggar, A. M. (2009). The philosophical challenges of global gender
justice. \emph{Philosophical Topics}, \emph{37}(2), 1+.
\href{https://link.gale.com/apps/doc/A284016231/AONE?u=ocul_mcmaster&sid=bookmark-AONE&xid=390bfcb0}{link.gale.com/apps/doc/A284016231/AONE?u=ocul\_mcmaster\&sid=bookmark-AONE\&xid=390bfcb0}

\bibitem[\citeproctext]{ref-jephcoteGeospatialAnalysisNaturally2013}
Jephcote, C., \& Chen, H. (2013). Geospatial analysis of naturally
occurring boundaries in road-transport emissions and children's
respiratory health across a demographically diverse cityscape.
\emph{SOCIAL SCIENCE \& MEDICINE}, \emph{82}, 87--99.
\url{https://doi.org/10.1016/j.socscimed.2013.01.030}

\bibitem[\citeproctext]{ref-jiaoHowIdentifyFood2012}
Jiao, J., Moudon, A., Ulmer, J., Hurvitz, P., \& Drewnowski, A. (2012).
How to {Identify Food Deserts}: {Measuring Physical} and {Economic
Access} to {Supermarkets} in {King County}, {Washington}. \emph{AMERICAN
JOURNAL OF PUBLIC HEALTH}, \emph{102}(10), E32--E39.
\url{https://doi.org/10.2105/AJPH.2012.300675}

\bibitem[\citeproctext]{ref-jimenez-espadaResearchProblemUniversal2021}
Jiménez-Espada, M., \& González-Escobar, R. (2021). \emph{Research on
the problem of universal accessibility in urban public transport. {Case}
study: The city of {Cáceres}.} \emph{58}, pp 21--28.
\url{https://doi.org/10.1016/j.trpro.2021.11.004}

\bibitem[\citeproctext]{ref-karnerTransportationEquityTransportation2020}
Karner, A., London, J., Rowangould, D., \& Manaugh, K. (2020). From
{Transportation Equity} to {Transportation Justice}: {Within},
{Through}, and {Beyond} the {State}. \emph{Journal of Planning
Literature}, \emph{35}(4), 440--459.
\url{https://doi.org/10.1177/0885412220927691}

\bibitem[\citeproctext]{ref-karnerAdvancesPitfallsMeasuring2024}
Karner, A., Pereira, R. H. M., \& Farber, S. (2024). Advances and
pitfalls in measuring transportation equity. \emph{Transportation}.
\url{https://doi.org/10.1007/s11116-023-10460-7}

\bibitem[\citeproctext]{ref-kentPrioritizingLowstressEquitable2019}
Kent, M., \& Karner, A. (2019). Prioritizing low-stress and equitable
bicycle networks using neighborhood-based accessibility measures.
\emph{International Journal of Sustainable Transportation},
\emph{13}(2), 100--110.
\url{https://doi.org/10.1080/15568318.2018.1443177}

\bibitem[\citeproctext]{ref-khistyOperationalizingConceptsEquity1996}
Khisty, C. J. (1996). Operationalizing {Concepts} of {Equity} for
{Public Project Investments}. \emph{Transportation Research Record:
Journal of the Transportation Research Board}, \emph{1559}(1), 94--99.
\url{https://doi.org/10.1177/0361198196155900112}

\bibitem[\citeproctext]{ref-khomenkoLiveableCityHealthy2020}
Khomenko, S., Nieuwenhuijsen, M., Ambros, A., Wegener, S., \& Mueller,
N. (2020). Is a liveable city a healthy city? {Health} impacts of urban
and transport planning in {Vienna}, {Austria}. \emph{ENVIRONMENTAL
RESEARCH}, \emph{183}.
\url{https://doi.org/10.1016/j.envres.2020.109238}

\bibitem[\citeproctext]{ref-kimAssessingSocialSpatial2020}
Kim, D., \& Park, J. (2020). Assessing {Social} and {Spatial Equity} of
{Neighborhood Retail} and {Service Access} in {Seoul}, {South Korea}.
\emph{Sustainability}, \emph{12}(20).
\url{https://doi.org/10.3390/su12208537}

\bibitem[\citeproctext]{ref-kimNeighborhoodEnvironmentWalkability2016}
Kim, H., Choi, Y., Ma, J., Hyung, K., Miyashita, M., \& Lee, S. (2016).
The {Neighborhood Environment Walkability Scale} for the {Republic} of
{Korea}: {Reliability} and {Relationship} with {Walking}. \emph{Iranian
Journal of Public Health}, \emph{45}(11), 1427--1435.

\bibitem[\citeproctext]{ref-kimImpactsHighspeedRail2015}
Kim, H., \& Sultana, S. (2015). The impacts of high-speed rail
extensions on accessibility and spatial equity changes in south korea
from 2004 to 2018. \emph{Journal of Transport Geography}, \emph{45},
48--61. \url{https://doi.org/10.1016/j.jtrangeo.2015.04.007}

\bibitem[\citeproctext]{ref-kimmelStructuralBarriersComprehensive2018}
Kimmel, A., Masiano, S., Bono, R., Martin, E., Belgrave, F., Adimora,
A., Dahman, B., Galadima, H., \& Sabik, L. (2018). Structural barriers
to comprehensive, coordinated {HIV} care: Geographic accessibility in
the {US South}. \emph{AIDS Care}, \emph{30}(11), 1459--1468.
\url{https://doi.org/10.1080/09540121.2018.1476656}

\bibitem[\citeproctext]{ref-kitaDoesLevelTransportation2020}
Kita, H., Yotsutsuji, H., Komoda, S., \& Yasunaga, K. (2020). \emph{Does
the level of transportation service affect the disparity of activity
opportunities between localities?} \emph{48}, pp 2527--2536.
\url{https://doi.org/10.1016/j.trpro.2020.08.257}

\bibitem[\citeproctext]{ref-kruizeEnvironmentalEquityRole2007}
Kruize, H., Driessen, P., Glasbergen, P., \& van Egmond, K. (2007).
Environmental equity and the role of public policy: {Experiences} in the
rijnmond region. \emph{ENVIRONMENTAL MANAGEMENT}, \emph{40}(4),
578--595. \url{https://doi.org/10.1007/s00267-005-0378-9}

\bibitem[\citeproctext]{ref-larkinsAccessibleTransportationBuilt2011}
Larkins, K., Dunning, A., \& Ridout, J. (2011). Accessible
{Transportation} and the {Built Environment} on {College Campuses}.
\emph{Transportation Research Record: Journal of the Transportation
Research Board}, \emph{2218}, 88--97.
\url{https://doi.org/10.3141/2218-10}

\bibitem[\citeproctext]{ref-laszkiewiczChildrenGreenWalk2020}
Laszkiewicz, E., \& Sikorska, D. (2020). Children's green walk to
school: {An} evaluation of welfare-related disparities in the visibility
of greenery among children. \emph{ENVIRONMENTAL SCIENCE \& POLICY},
\emph{110}, 1--13. \url{https://doi.org/10.1016/j.envsci.2020.05.009}

\bibitem[\citeproctext]{ref-lattmanPerceivedAccessibilityPublic2016}
Lattman, K., Friman, M., \& Olsson, L. (2016). Perceived {Accessibility}
of {Public Transport} as a {Potential Indicator} of {Social Inclusion}.
\emph{Social Inclusion}, \emph{4}(3), 36--45.
\url{https://doi.org/10.17645/si.v4i3.481}

\bibitem[\citeproctext]{ref-laveryDriving2013}
Lavery, T. A., Paez, A., \& Kanaroglou, P. S. (2013). Driving out of
choices: An investigation of transport modality in a university sample
{[}Journal Article{]}. \emph{Transportation Research Part A-Policy and
Practice}, \emph{57}, 37--46.
\url{https://doi.org/10.1016/j.tra.2013.09.010}

\bibitem[\citeproctext]{ref-leeRacial1993}
Lee, S. M. (1993). Racial classifications in the US census: 1890--1990
{[}Journal Article{]}. \emph{Ethnic and Racial Studies}, \emph{16}(1),
75--94. \url{https://doi.org/10.1080/01419870.1993.9993773}

\bibitem[\citeproctext]{ref-lefebvreDroitVille1967}
Lefebvre, H. (1967). Le droit à la ville. \emph{L Homme et la société},
\emph{6}(1), 29--35. \url{https://doi.org/10.3406/homso.1967.1063}

\bibitem[\citeproctext]{ref-limFacilitatingIndependentCommuting2021}
Lim, P., Kong, P., Cornet, H., \& Frenkler, F. (2021). Facilitating
independent commuting among individuals with autism-{A} design study in
{Singapore}. \emph{Journal of Transport \& Health}, \emph{21}.
\url{https://doi.org/10.1016/j.jth.2021.101022}

\bibitem[\citeproctext]{ref-liuStatusQuoChallenges2019}
Liu, X., Chen, X., Gao, C., \& American Society of Civil Engineers.
(2019). \emph{The {Status Quo}, {Challenges}, and {Policy
Recommendation} of {Transport Barrier-Free Environment Development} in
{China}}. pp 5351--5363. \url{https://doi.org/10.1061/9780784482292.461}

\bibitem[\citeproctext]{ref-mackieOvercomingBarriersCycling2009}
Mackie, H. (2009). \emph{Overcoming barriers to cycling to school: A key
to improving transport system performance}. \emph{32}, 11p (session
Thurs 2A).
\url{http://atrf.info/papers/2009/2009_Mackie.pdfhttps://trid.trb.org/view/1149648}

\bibitem[\citeproctext]{ref-mammenSchoolTravelPlanning2014}
Mammen, G., Stone, M., Buliung, R., \& Faulkner, G. (2014). School
travel planning in {Canada}: {Identifying} child, family, and
school-level characteristics associated with travel mode shift from
driving to active school travel. \emph{Journal of Transport \& Health},
\emph{1}(4), 288--294. \url{https://doi.org/10.1016/j.jth.2014.09.004}

\bibitem[\citeproctext]{ref-markardUnsustainabilities2023}
Markard, J., Wells, P., Yap, X.-S., \& Lente, H. van. (2023).
Unsustainabilities: A study on SUVs and space tourism and a research
agenda for transition studies {[}Journal Article{]}. \emph{Energy
Research \& Social Science}, \emph{106}, 103302.
https://doi.org/\url{https://doi.org/10.1016/j.erss.2023.103302}

\bibitem[\citeproctext]{ref-markolfTransportation2019}
Markolf, S. A., Hoehne, C., Fraser, A., Chester, M. V., \& Underwood, B.
S. (2019). Transportation resilience to climate change and extreme
weather events -- beyond risk and robustness {[}Journal Article{]}.
\emph{Transport Policy}, \emph{74}, 174--186.
https://doi.org/\url{https://doi.org/10.1016/j.tranpol.2018.11.003}

\bibitem[\citeproctext]{ref-marquezFactorsAffectingPersonal2019}
Marquez, L., Poveda, J., \& Vega, L. (2019). Factors affecting personal
autonomy and perceived accessibility of people with mobility impairments
in an urban transportation choice context. \emph{Journal of Transport \&
Health}, \emph{14}. \url{https://doi.org/10.1016/j.jth.2019.100583}

\bibitem[\citeproctext]{ref-martensTransport2016}
Martens, K. (2016). \emph{Transport justice: Designing fair
transportation systems}. Routledge.

\bibitem[\citeproctext]{ref-martensMeasuring2019}
Martens, K., Bastiaanssen, J., \& Lucas, K. (2019). Measuring transport
equity: Key components, framings and metrics {[}Book Section{]}. In K.
Lucas, K. Martens, F. Di Ciommo, \& A. Dupont-Kieffer (Eds.),
\emph{Measuring transport equity} (pp. 13--36). Elsevier.
https://doi.org/\url{https://doi.org/10.1016/B978-0-12-814818-1.00002-0}

\bibitem[\citeproctext]{ref-martensFair2021}
Martens, K., \& Golub, A. (2021). A fair distribution of accessibility:
Interpreting civil rights regulations for regional transportation plans
{[}Journal Article{]}. \emph{Journal of Planning Education and
Research}, \emph{41}(4), 425--444.
\url{https://doi.org/10.1177/0739456x18791014}

\bibitem[\citeproctext]{ref-martensJusticetheoreticApproachDistribution2012}
Martens, K., Golub, A., \& Robinson, G. (2012). A justice-theoretic
approach to the distribution of transportation benefits: {Implications}
for transportation planning practice in the {United States}.
\emph{Transportation Research Part A: Policy and Practice},
\emph{46}(4), 684--695. \url{https://doi.org/10.1016/j.tra.2012.01.004}

\bibitem[\citeproctext]{ref-martinez-jimenezAccessibilityCultureEducation2019}
Martinez-Jimenez, E., \& Salinas-Perez, J. (2019). Accessibility to
culture and education. {Educative} city of {Cordoba} ({Spain}).
\emph{Journal of Maps}, \emph{15}(1), 39--45.
\url{https://doi.org/10.1080/17445647.2019.1575776}

\bibitem[\citeproctext]{ref-mateo-babianoBicycleSharingAsia2017}
Mateo-Babiano, I., Kumar, S., \& Mejia, A. (2017). \emph{Bicycle
{Sharing} in {Asia}: {A Stakeholder Perception} and {Possible Futures}}.
\emph{25}, pp 4970--4982.
\url{https://doi.org/10.1016/j.trpro.2017.05.375}

\bibitem[\citeproctext]{ref-mavoaAreaLevelDisparitiesPublic2015}
Mavoa, S., Koohsari, M., Badland, H., Davern, M., Feng, X., Astell-Burt,
T., \& Giles-Corti, B. (2015). Area-{Level Disparities} of {Public Open
Space}: {A Geographic Information Systems Analysis} in {Metropolitan
Melbourne}. \emph{Urban Policy and Research}, \emph{33}(3), 306--323.
\url{https://doi.org/10.1080/08111146.2014.974747}

\bibitem[\citeproctext]{ref-mccormackRelationshipClusteranalysisDerived2012}
McCormack, G., Friedenreich, C., Sandalack, B., Giles-Corti, B.,
Doyle-Baker, P., \& Shiell, A. (2012). The relationship between
cluster-analysis derived walkability and local recreational and
transportation walking among {Canadian} adults. \emph{Health \& Place},
\emph{18}(5), 1079--1087.
\url{https://doi.org/10.1016/j.healthplace.2012.04.014}

\bibitem[\citeproctext]{ref-mckeyCrowdsourcedMappingHealthy2020}
McKey, T., Kim, D., \& Seo, S. (2020). Crowdsourced {Mapping} for
{Healthy Food Accessibility} in {Dallas}, {Texas}: {A Feasibility
Study}. \emph{FRONTIERS IN PUBLIC HEALTH}, \emph{8}.
\url{https://doi.org/10.3389/fpubh.2020.00071}

\bibitem[\citeproctext]{ref-mearsUnderstandingSocioeconomicEquity2019}
Mears, M., Brindley, P., Maheswaran, R., \& Jorgensen, A. (2019).
Understanding the socioeconomic equity of publicly accessible greenspace
distribution: {The} example of {Sheffield}, {UK}. \emph{GEOFORUM},
\emph{103}, 126--137.
\url{https://doi.org/10.1016/j.geoforum.2019.04.016}

\bibitem[\citeproctext]{ref-mehdizadehWalkingTimeSchool2017}
Mehdizadeh, M., Mamdoohi, A., \& Nordfjaern, T. (2017). Walking time to
school, children's active school travel and their related factors.
\emph{Journal of Transport \& Health}, \emph{6}, 313--326.
\url{https://doi.org/10.1016/j.jth.2017.01.012}

\bibitem[\citeproctext]{ref-mellaDrivers2021}
Mella Lira, B., \& Paez, A. (2021). Do drivers dream of walking? An
investigation of travel mode dissonance from the perspective of
affective values {[}Journal Article{]}. \emph{Journal of Transport \&
Health}, \emph{20}, 101015.
https://doi.org/\url{https://doi.org/10.1016/j.jth.2021.101015}

\bibitem[\citeproctext]{ref-mijaresEquityAnalysisUrban2013}
Mijares, A. C., Suzuki, M., \& Yai, T. (2013). Equity analysis of urban
rail fare policy and passenger overload delay: An international
comparison and the case of metro manila {MRT}-3. \emph{Journal of the
Eastern Asia Society for Transportation Studies}, \emph{10}, pp 45--65.
\url{https://doi.org/10.11175/easts.10.45}

\bibitem[\citeproctext]{ref-mirandaBenchmarkingSustainableUrban2012}
Miranda, H., \& da Silva, A. (2012). Benchmarking sustainable urban
mobility: {The} case of {Curitiba}, {Brazil}. \emph{Transport Policy},
\emph{21}, 141--151. \url{https://doi.org/10.1016/j.tranpol.2012.03.009}

\bibitem[\citeproctext]{ref-mishraEIABasedComparative2014}
Mishra, R. K., Shukla, A., Parida, M., \& Rangnekar, S. (2014). {EIA
Based Comparative Urban Traffic Noise Analysis Between Operational} and
{Under Construction Phase Public Transport Corridor}.
\emph{International Journal for Traffic and Transport Engineering},
\emph{4}(3), pp 352--362.
\url{https://doi.org/10.7708/ijtte.2014.4(3).08}

\bibitem[\citeproctext]{ref-mohriClusteringMethodMeasuring2021}
Mohri, S., Mortazavi, S., \& Nassir, N. (2021). A clustering method for
measuring accessibility and equity in public transportation service:
{Case} study of {Melbourne}. \emph{Sustainable Cities and Society},
\emph{74}. \url{https://doi.org/10.1016/j.scs.2021.103241}

\bibitem[\citeproctext]{ref-mokhtarianUnderstanding2001}
Mokhtarian, P. L., Salomon, I., \& Redmond, L. S. (2001). Understanding
the demand for travel: It's not purely 'derived' {[}Journal Article{]}.
\emph{Innovation}, \emph{14}(4).

\bibitem[\citeproctext]{ref-moniruzzamanInvestigation2016}
Moniruzzaman, M., \& Paez, A. (2016). An investigation of the attributes
of walkable environments from the perspective of seniors in montreal
{[}Journal Article{]}. \emph{Journal of Transport Geography}, \emph{51},
85--96. \url{https://doi.org/10.1016/j.jtrangeo.2015.12.001}

\bibitem[\citeproctext]{ref-monzonEfficiencySpatialEquity2013}
Monzon, A., Ortega, E., \& Lopez, E. (2013). Efficiency and spatial
equity impacts of high-speed rail extensions in urban areas.
\emph{CITIES}, \emph{30}, 18--30.
\url{https://doi.org/10.1016/j.cities.2011.11.002}

\bibitem[\citeproctext]{ref-mooneyTravelDistanceBarrier2000}
Mooney, C., Zwanziger, J., Phibbs, C., \& Schmitt, S. (2000). Is travel
distance a barrier to veterans' use of {VA} hospitals for medical
surgical care? \emph{SOCIAL SCIENCE \& MEDICINE}, \emph{50}(12),
1743--1755. \url{https://doi.org/10.1016/S0277-9536(99)00414-1}

\bibitem[\citeproctext]{ref-muellerSocioeconomicInequalitiesUrban2018}
Mueller, N., Rojas-Rueda, D., Khreis, H., Cirach, M., Mila, C.,
Espinosa, A., Foraster, M., McEachan, R., Kelly, B., Wright, J., \&
Nieuwenhuijsen, M. (2018). Socioeconomic inequalities in urban and
transport planning related exposures and mortality: {A} health impact
assessment study for {Bradford}, {UK}. \emph{ENVIRONMENT INTERNATIONAL},
\emph{121}, 931--941. \url{https://doi.org/10.1016/j.envint.2018.10.017}

\bibitem[\citeproctext]{ref-mulleyAreNetworkPlanning2015}
Mulley, C., Ma, L., Clifton, G. T., \& Tanner, M. (2015). \emph{Are
network planning guidelines based on equal access equitable?} 18p.
\url{http://atrf.info/papers/2015/index.aspxhttps://trid.trb.org/view/1395093}

\bibitem[\citeproctext]{ref-murphySupermarketAccessTransport2017}
Murphy, M., Koohsari, M., Badland, H., \& Giles-Corti, B. (2017).
Supermarket access, transport mode and {BMI}: The potential for urban
design and planning policy across socio-economic areas. \emph{PUBLIC
HEALTH NUTRITION}, \emph{20}(18), 3304--3315.
\url{https://doi.org/10.1017/S1368980017002336}

\bibitem[\citeproctext]{ref-nordbakkeTransportUnmetActivity2015}
Nordbakke, S., \& Schwanen, T. (2015). Transport, unmet activity needs
and wellbeing in later life: Exploring the links. \emph{TRANSPORTATION},
\emph{42}(6), 1129--1151.
\url{https://doi.org/10.1007/s11116-014-9558-x}

\bibitem[\citeproctext]{ref-odeckEconomicAppraisalUniversal2010}
Odeck, J., Hagen, T., \& Fearnley, N. (2010). Economic {Appraisal} of
{Universal Design} in {Transport}: {Experiences From Norway}.
\emph{Research in Transportation Economics}, \emph{29}(1), pp 304--311.
\url{http://www.sciencedirect.com/science/article/B8JHM-5119FS6-2/2/3212b0f3260bbd5899dbf18cc4b3cf0ehttps://trid.trb.org/view/981277}

\bibitem[\citeproctext]{ref-orellanaWalkRollMapping2020}
Orellana, D., Bustos, M., Marin-Palacios, M., Cabrera-Jara, N., \&
Hermida, M. (2020). Walk'n'roll: {Mapping} street-level accessibility
for different mobility conditions in {Cuenca}, {Ecuador}. \emph{Journal
of Transport \& Health}, \emph{16}.
\url{https://doi.org/10.1016/j.jth.2020.100821}

\bibitem[\citeproctext]{ref-pagePRISMA2020Statement2021}
Page, M. J., McKenzie, J. E., Bossuyt, P. M., Boutron, I., Hoffmann, T.
C., Mulrow, C. D., Shamseer, L., Tetzlaff, J. M., Akl, E. A., Brennan,
S. E., Chou, R., Glanville, J., Grimshaw, J. M., Hróbjartsson, A., Lalu,
M. M., Li, T., Loder, E. W., Mayo-Wilson, E., McDonald, S., \ldots{}
Moher, D. (2021). The {PRISMA} 2020 statement: An updated guideline for
reporting systematic reviews. \emph{BMJ}, n71.
\url{https://doi.org/10.1136/bmj.n71}

\bibitem[\citeproctext]{ref-parkJourneyVisuallyImpaired2017}
Park, J., BAMFORD, J., Byun, H., \& Chowdhury, S. (2017). \emph{Journey
by visually impaired public transport users: {Barriers} and
consequences}. 6p.
\url{https://atrf.info/papers/2017/index.aspxhttps://trid.trb.org/view/1596698}

\bibitem[\citeproctext]{ref-parkTransitParksEnvironmental2021}
Park, K., Rigolon, A., Choi, D., Lyons, T., \& Brewer, S. (2021).
Transit to parks: {An} environmental justice study of transit access to
large parks in the {US West}. \emph{Urban Forestry \& Urban Greening},
\emph{60}. \url{https://doi.org/10.1016/j.ufug.2021.127055}

\bibitem[\citeproctext]{ref-parrySocialVulnerabilityClimatic2018}
Parry, L., Davies, G., Almeida, O., Frausin, G., de Moraes, A., Rivero,
S., Filizola, N., \& Torres, P. (2018). Social {Vulnerability} to
{Climatic Shocks Is Shaped} by {Urban Accessibility}. \emph{Annals of
the American Association of Geographers}, \emph{108}(1), 125--143.
\url{https://doi.org/10.1080/24694452.2017.1325726}

\bibitem[\citeproctext]{ref-patelDeterminingGeographicAreas2007}
Patel, A., Waters, N., \& Ghali, W. (2007). Determining geographic areas
and populations with timely access to cardiac catheterization facilities
for acute myocardial infarction care in {Alberta}, {Canada}.
\emph{International Journal of Health Geographics}, \emph{6}.
\url{https://doi.org/10.1186/1476-072X-6-47}

\bibitem[\citeproctext]{ref-pedigoInvestigationDisparitiesGeographic2010}
Pedigo, A., \& Odoi, A. (2010). Investigation of {Disparities} in
{Geographic Accessibility} to {Emergency Stroke} and {Myocardial
Infarction Care} in {East Tennessee Using Geographic Information
Systems} and {Network Analysis}. \emph{ANNALS OF EPIDEMIOLOGY},
\emph{20}(12), 924--930.
\url{https://doi.org/10.1016/j.annepidem.2010.06.013}

\bibitem[\citeproctext]{ref-pereiraTransportationEquity2021}
Pereira, R. H. M., \& Karner, A. (2021). Transportation {Equity}. In
\emph{International {Encyclopedia} of {Transportation}} (pp. 271--277).
{Elsevier}. \url{https://doi.org/10.1016/B978-0-08-102671-7.10053-3}

\bibitem[\citeproctext]{ref-pereiraDistributiveJusticeEquity2017a}
Pereira, R. H. M., Schwanen, T., \& Banister, D. (2017). Distributive
justice and equity in transportation. \emph{Transport Reviews},
\emph{37}(2), 170--191.
\url{https://doi.org/10.1080/01441647.2016.1257660}

\bibitem[\citeproctext]{ref-pereiraGeographicAccessCOVID192021}
Pereira, R., Braga, C., Servo, L., Serra, B., Amaral, P., Gouveia, N.,
\& Paez, A. (2021). Geographic access to {COVID-19} healthcare in
{Brazil} using a balanced float catchment area approach. \emph{SOCIAL
SCIENCE \& MEDICINE}, \emph{273}.
\url{https://doi.org/10.1016/j.socscimed.2021.113773}

\bibitem[\citeproctext]{ref-perezdelhoyoParticipatoryManagementImprove2021}
Perez-delHoyo, R., Andujar-Montoya, M., Mora, H., Gilart-Iglesias, V.,
\& Molla-Sirvent, R. (2021). Participatory {Management} to {Improve
Accessibility} in {Consolidated Urban Environments}.
\emph{Sustainability}, \emph{13}(15).
\url{https://doi.org/10.3390/su13158323}

\bibitem[\citeproctext]{ref-petersUpdatedMethodologicalGuidance2020}
Peters, M. D. J., Marnie, C., Tricco, A. C., Pollock, D., Munn, Z.,
Alexander, L., McInerney, P., Godfrey, C. M., \& Khalil, H. (2020).
Updated methodological guidance for the conduct of scoping reviews.
\emph{JBI Evidence Synthesis}, \emph{18}(10), 2119--2126.
\url{https://doi.org/10.11124/JBIES-20-00167}

\bibitem[\citeproctext]{ref-peungnumsaiGridBasedSpatialAnalysis2020}
Peungnumsai, A., Miyazaki, H., Witayangkurn, A., \& Kim, S. (2020). A
{Grid-Based Spatial Analysis} for {Detecting Supply-Demand Gaps} of
{Public Transports}: {A Case Study} of the {Bangkok Metropolitan
Region}. \emph{Sustainability}, \emph{12}(24).
\url{https://doi.org/10.3390/su122410382}

\bibitem[\citeproctext]{ref-prasertsubpakijEvaluatingAccessibilityBangkok2012}
Prasertsubpakij, D., \& Nitivattananon, V. (2012). Evaluating
{Accessibility} to {Bangkok Metro Systems Using Multi-Dimensional
Criteria} across {User Groups}. \emph{IATSS Research}, \emph{36}(1), pp
56--65.
\url{http://www.sciencedirect.com/science/article/pii/S0386111212000040https://trid.trb.org/view/1148211}

\bibitem[\citeproctext]{ref-pritchardNew2022}
Pritchard, J. P., Zanchetta, A., \& Martens, K. (2022). A new index to
assess the situation of subgroups, with an application to public
transport disadvantage in US metropolitan areas {[}Journal Article{]}.
\emph{Transportation Research Part A: Policy and Practice}, \emph{166},
86--100. https://doi.org/\url{https://doi.org/10.1016/j.tra.2022.10.002}

\bibitem[\citeproctext]{ref-pucciInequalitiesJobrelatedAccessibility2019}
Pucci, P., Vecchio, G., Bocchimuzzi, L., \& Lanza, G. (2019).
Inequalities in job-related accessibility: {Testing} an evaluative
approach and its policy relevance in {Buenos Aires}. \emph{Applied
Geography}, \emph{107}, 1--11.
\url{https://doi.org/10.1016/j.apgeog.2019.04.002}

\bibitem[\citeproctext]{ref-rabelloEvaluation2015}
Rabello Quadros, S. G., \& Nassi, C. D. (2015). An evaluation on the
criteria to prioritize transportation infrastructure investments in
brazil {[}Journal Article{]}. \emph{Transport Policy}, \emph{40}, 8--16.
https://doi.org/\url{https://doi.org/10.1016/j.tranpol.2015.02.002}

\bibitem[\citeproctext]{ref-racheleNeighbourhoodSocioeconomicTransport2017a}
Rachele, J., Learnihan, V., Badland, H., Mavoa, S., Turrell, G., \&
Giles-Corti, B. (2017). Neighbourhood socioeconomic and transport
disadvantage: {The} potential to reduce social inequities in health
through transport. \emph{Journal of Transport \& Health}, \emph{7},
256--263. \url{https://doi.org/10.1016/j.jth.2017.09.002}

\bibitem[\citeproctext]{ref-rajeEngineering2004}
Raje, F. (2004). Engineering social exclusion? Poor transport links and
severance {[}Journal Article{]}. \emph{Proceedings of the Institution of
Civil Engineers-Municipal Engineer}, \emph{157}(4), 267--273.
\href{https://ISI:000226631000008\%0AC:/Papers/Proceedings\%20of\%20the\%20Institute\%20of\%20Civil\%20Engineers/Proceedings\%20of\%20the\%20Institute\%20of\%20Civil\%20Engineers\%20(2004)\%20157\%20(4)\%20267-273.pdf}{ISI:000226631000008
C:/Papers/Proceedings of the Institute of Civil Engineers/Proceedings of
the Institute of Civil Engineers (2004) 157 (4) 267-273.pdf}

\bibitem[\citeproctext]{ref-raoDecentLivingEmissions2012}
Rao, N., \& Baer, P. (2012). "{Decent Living}" {Emissions}: {A
Conceptual Framework}. \emph{Sustainability}, \emph{4}(4), 656--681.
\url{https://doi.org/10.3390/su4040656}

\bibitem[\citeproctext]{ref-ravensbergenExploratoryAnalysisMobility2023}
Ravensbergen, L., Fournier, J., \& El-Geneidy, A. (2023). Exploratory
analysis of mobility of care in montreal, canada. \emph{Transportation
Research Record}, \emph{2677}(1), 1499--1509.
\url{https://doi.org/10.1177/03611981221105070}

\bibitem[\citeproctext]{ref-reddySafeguardingMinorityCivil2010}
Reddy, A., Chennadu, T., \& Lu, A. (2010). Safeguarding {Minority Civil
Rights} and {Environmental Justice} in {Service Delivery} and
{Reductions Case Study} of {New York City Transit Authority Title VI
Program}. \emph{Transportation Research Record: Journal of the
Transportation Research Board}, \emph{2163}, 45--56.
\url{https://doi.org/10.3141/2163-05}

\bibitem[\citeproctext]{ref-renneWhatHasAmerica2018}
Renne, J. L., \& Mayorga, E. (2018). \emph{What {Has America Learned
Since Hurricane Katrina}? {Evaluating Evacuation Plans} for {Carless}
and {Vulnerable Populations} in 50 {Large Cities Across} the {United
States}}. 13p. \url{https://trid.trb.org/view/1495593}

\bibitem[\citeproctext]{ref-rivasHowAffordableTransportation2018}
Rivas, M. E., Serebrisky, T., \& Suárez-Alemán, A. (2018). \emph{How
affordable is transportation in latin america and the caribbean?} 15p.
\url{https://annualmeeting.mytrb.org/OnlineProgram/Details/15652https://trid.trb.org/view/1759304}

\bibitem[\citeproctext]{ref-robinsonDoubleEnergyVulnerability2020}
Robinson, C., \& Mattioli, G. (2020). Double energy vulnerability:
{Spatial} intersections of domestic and transport energy poverty in
{England}. \emph{Energy Research \& Social Science}, \emph{70}.
\url{https://doi.org/10.1016/j.erss.2020.101699}

\bibitem[\citeproctext]{ref-rowangouldIdentifyingEnvironmentalJustice2015}
Rowangould, D., Karner, A., \& London, J. (2015). \emph{Identifying
{Environmental Justice Communities} for {Transportation Analysis}}. 11p.
\url{https://trid.trb.org/view/1339508}

\bibitem[\citeproctext]{ref-rowangouldIdentifyingEnvironmentalJustice2016}
Rowangould, D., Karner, A., \& London, J. (2016). Identifying
environmental justice communities for transportation analysis.
\emph{Transportation Research Part A: Policy and Practice}, \emph{88},
151--162. \url{https://doi.org/10.1016/j.tra.2016.04.002}

\bibitem[\citeproctext]{ref-russellPedallingEquityExploring2021}
Russell, M., Davies, C., Wild, K., \& Shaw, C. (2021). Pedalling towards
equity: {Exploring} women's cycling in a {New Zealand} city.
\emph{Journal of Transport Geography}, \emph{91}.
\url{https://doi.org/10.1016/j.jtrangeo.2021.102987}

\bibitem[\citeproctext]{ref-safransky2022grammars}
Safransky, S. (2022). Grammars of reckoning: Redressing racial regimes
of property. \emph{Environment and Planning D: Society and Space},
\emph{40}(2), 292--305.

\bibitem[\citeproctext]{ref-scheurerSpatialAccessibilityPublic2017}
Scheurer, J., Curtis, C., \& McLeod, S. (2017). Spatial accessibility of
public transport in {Australian} cities: {Does} it relieve or entrench
social and economic inequality? \emph{Journal of Transport and Land
Use}, \emph{10}(1), 911--930.
\url{https://doi.org/10.5198/jtlu.2017.1097}

\bibitem[\citeproctext]{ref-savingmothersDidSavingMothers2019}
Schmitz, M. M., Serbanescu, F., Kamara, V., Kraft, J. M., Cunningham,
M., Opio, G., Komakech, P., Conlon, C. M., \& Goodwin, M. M. (2019). Did
saving mothers, giving life expand timely access to lifesaving care in
uganda? A spatial district-level analysis of travel time to emergency
obstetric and newborn care. \emph{Global Health: Science and Practice},
\emph{7}, S151--S167. \url{https://doi.org/10.9745/GHSP-D-18-00366}

\bibitem[\citeproctext]{ref-sharmaPublicTransitAccessibility2021}
Sharma, G., \& Patil, G. (2021). Public transit accessibility approach
to understand the equity for public healthcare services: {A} case study
of {Greater Mumbai}. \emph{Journal of Transport Geography}, \emph{94}.
\url{https://doi.org/10.1016/j.jtrangeo.2021.103123}

\bibitem[\citeproctext]{ref-sharmaSpatialSocialInequities2022}
Sharma, G., \& Patil, G. (2022). Spatial and social inequities for
educational services accessibility - {A} case study for schools in
{Greater Mumbai}. \emph{CITIES}, \emph{122}.
\url{https://doi.org/10.1016/j.cities.2021.103543}

\bibitem[\citeproctext]{ref-shenMeasuringSpatialAccessibility2020}
Shen, C., Zhou, Z., Lai, S., Lu, L., Dong, W., Su, M., Zhang, J., Wang,
X., Deng, Q., Chen, Y., \& Chen, X. (2020). Measuring spatial
accessibility and within-province disparities in accessibility to county
hospitals in {Shaanxi Province} of {Western China} based on web mapping
navigation data. \emph{International Journal for Equity in Health},
\emph{19}(1). \url{https://doi.org/10.1186/s12939-020-01217-0}

\bibitem[\citeproctext]{ref-silvaEnvironmentalJusticeAccessibility2018}
Silva, C., Viegas, I., Panagopoulos, T., \& Bell, S. (2018).
Environmental {Justice} in {Accessibility} to {Green Infrastructure} in
{Two European Cities}. \emph{Land}, \emph{7}(4).
\url{https://doi.org/10.3390/land7040134}

\bibitem[\citeproctext]{ref-siuAssessmentPhysicalEnvironment2019}
Siu, B. (2019). Assessment of physical environment factors for mobility
of older adults: {A} case study in {Hong Kong}. \emph{Research in
Transportation Business and Management}, \emph{30}.
\url{https://doi.org/10.1016/j.rtbm.2019.100370}

\bibitem[\citeproctext]{ref-smithAccessibilityCapabilityMinimum2012}
Smith, N., Hirsch, D., \& Davis, A. (2012). Accessibility and
capability: The minimum transport needs and costs of rural households.
\emph{Journal of Transport Geography}, \emph{21}, 93--101.
\url{https://doi.org/10.1016/j.jtrangeo.2012.01.004}

\bibitem[\citeproctext]{ref-somenahalliAgingTransportMobility2007}
Somenahalli, S. V., \& Taylor, M. A. (2007). Aging and transport:
Mobility issues: A case study for adelaide. \emph{{STATE} {OF}
{AUSTRALIAN} {CITIES} {NATIONAL} {CONFERENCE}, 2007, {ADELAIDE}, {SOUTH}
{AUSTRALIA}}, 11P. \url{https://trid.trb.org/view/868838}

\bibitem[\citeproctext]{ref-songWomenWheelsGender2019}
Song, L., Kirschen, M., \& Taylor, J. (2019). Women on wheels: {Gender}
and cycling in {Solo}, {Indonesia}. \emph{SINGAPORE JOURNAL OF TROPICAL
GEOGRAPHY}, \emph{40}(1), 140--157.
\url{https://doi.org/10.1111/sjtg.12257}

\bibitem[\citeproctext]{ref-soukhovEquityJusticeConceptualizations2022}
Soukhov, A., Aitken, I. T., Moghadasi, M., Palm, M., Farber, S., \&
Paez, A. (2022). \emph{Equity and justice conceptualizations, dimensions
and standards for urban transportation planning: A scoping review
protocol}. \url{https://doi.org/10.17605/OSF.IO/RSB92}

\bibitem[\citeproctext]{ref-sourbatiSmart2021}
Sourbati, M., \& Behrendt, F. (2021). Smart mobility, age and data
justice {[}Journal Article{]}. \emph{New Media \&Amp; Society},
\emph{23}(6), 1398--1414. \url{https://doi.org/10.1177/1461444820902682}

\bibitem[\citeproctext]{ref-stjernborgAccessibilityAllPublic2019}
Stjernborg, V. (2019). Accessibility for {All} in {Public Transport} and
the {Overlooked} ({Social}) {Dimension-A Case Study} of {Stockholm}.
\emph{Sustainability}, \emph{11}(18).
\url{https://doi.org/10.3390/su11184902}

\bibitem[\citeproctext]{ref-sunPublicTransportAvailability2021}
Sun, Y., \& Thakuriah, P. (2021). Public transport availability
inequalities and transport poverty risk across {England}.
\emph{ENVIRONMENT AND PLANNING B-URBAN ANALYTICS AND CITY SCIENCE},
\emph{48}(9), 2775--2789. \url{https://doi.org/10.1177/2399808321991536}

\bibitem[\citeproctext]{ref-sungResidentialBuiltEnvironment2015}
Sung, H., \& Lee, S. (2015). Residential built environment and walking
activity: {Empirical} evidence of {Jane Jacobs}' urban vitality.
\emph{Transportation Research Part D: Transport and Environment},
\emph{41}, 318--329. \url{https://doi.org/10.1016/j.trd.2015.09.009}

\bibitem[\citeproctext]{ref-tan2016pareto}
Tan, Z., Yang, H., Tan, W., \& Li, Z. (2016). Pareto-improving
transportation network design and ownership regimes.
\emph{Transportation Research Part B: Methodological}, \emph{91},
292--309.

\bibitem[\citeproctext]{ref-thondooSmallCitiesBig2020}
Thondoo, M., Marquet, O., Marquez, S., \& Nieuwenhuijsen, M. (2020).
Small cities, big needs: {Urban} transport planning in cities of
developing countries. \emph{Journal of Transport \& Health}, \emph{19}.
\url{https://doi.org/10.1016/j.jth.2020.100944}

\bibitem[\citeproctext]{ref-timperioSafetyNumbersDoes2015}
Timperio, A., Veitch, J., \& Carver, A. (2015). Safety in numbers:
{Does} perceived safety mediate associations between the neighborhood
social environment and physical activity among women living in
disadvantaged neighborhoods? \emph{PREVENTIVE MEDICINE}, \emph{74},
49--54. \url{https://doi.org/10.1016/j.ypmed.2015.02.012}

\bibitem[\citeproctext]{ref-timperleyFight2021}
Timperley, J. (2021). The fight to end fossil-fuel subsidies {[}Journal
Article{]}. \emph{Nature}, \emph{598}, 403--405.

\bibitem[\citeproctext]{ref-titheridgeSocialExclusionAccessibility2008}
Titheridge, H., Solomon, J., \& Accessibility and User Needs in
Transport for Sustainable Urban Environments Consortium (AUNT-SUE).
(2008). \emph{Social exclusion, accessibility and lone parents}. 14p.
\url{http://www.sortclearinghouse.info/cgi/viewcontent.cgi?article=1232&context=researchhttps://trid.trb.org/view/1153041}

\bibitem[\citeproctext]{ref-tiwariDevelopmentPublicTransport2021}
Tiwari, G., \& Phillip, C. (2021). Development of public transport
systems in small cities: {A} roadmap for achieving sustainable
development goal indicator 11.2. \emph{IATSS Research}, \emph{45}(1), pp
31--38. \url{https://doi.org/10.1016/j.iatssr.2021.02.002}

\bibitem[\citeproctext]{ref-tiznadoaitkenRoleAccessibilityPublic2018}
Tiznado-Aitken, I., Munoz, J., \& Hurtubia, R. (2018). The {Role} of
{Accessibility} to {Public Transport} and {Quality} of {Walking
Environment} on {Urban Equity}: {The Case} of {Santiago} de {Chile}.
\emph{Transportation Research Record}, \emph{2672}(35), 129--138.
\url{https://doi.org/10.1177/0361198118782036}

\bibitem[\citeproctext]{ref-towneUsingWalkScore2016}
Towne, S., Won, J., Lee, S., Ory, M., Forjuoh, S., Wang, S., \& Lee, C.
(2016). Using {Walk Score} ({TM}) and {Neighborhood Perceptions} to
{Assess Walking Among Middle-Aged} and {Older Adults}. \emph{JOURNAL OF
COMMUNITY HEALTH}, \emph{41}(5), 977--988.
\url{https://doi.org/10.1007/s10900-016-0180-z}

\bibitem[\citeproctext]{ref-triccoPRISMAExtensionScoping2018}
Tricco, A. C., Lillie, E., Zarin, W., O'Brien, K. K., Colquhoun, H.,
Levac, D., Moher, D., Peters, M. D. J., Horsley, T., Weeks, L., Hempel,
S., Akl, E. A., Chang, C., McGowan, J., Stewart, L., Hartling, L.,
Aldcroft, A., Wilson, M. G., Garritty, C., \ldots{} Straus, S. E.
(2018). {PRISMA Extension} for {Scoping Reviews} ({PRISMA-ScR}):
{Checklist} and {Explanation}. \emph{Annals of Internal Medicine},
\emph{169}(7), 467--473. \url{https://doi.org/10.7326/M18-0850}

\bibitem[\citeproctext]{ref-tylerRestorativeJusticeProcedural2006}
Tyler, T. R. (2006). Restorative justice and procedural justice: Dealing
with rule breaking. \emph{Journal of Social Issues}, \emph{62}(2),
307--326. \url{https://doi.org/10.1111/j.1540-4560.2006.00452.x}

\bibitem[\citeproctext]{ref-vadrevuMeasuringSpatialEquity2016}
Vadrevu, L., \& Kanjilal, B. (2016). Measuring spatial equity and access
to maternal health services using enhanced two step floating catchment
area method ({E2SFCA}) - a case study of the {Indian Sundarbans}.
\emph{International Journal for Equity in Health}, \emph{15}.
\url{https://doi.org/10.1186/s12939-016-0376-y}

\bibitem[\citeproctext]{ref-vanderveenOperationalizingIndicatorSufficient2020}
van der Veen, A. S., Annema, J. A., Martens, K., van Arem, B., \& de
Almeida Correia, G. H. (2020). Operationalizing an {Indicator} of
{Sufficient Accessibility} - a {Case Study} for {The City} of
{Rotterdam}. \emph{Case Studies on Transport Policy}, \emph{8}(4), pp
1360--1370. \url{https://doi.org/10.1016/j.cstp.2020.09.007}

\bibitem[\citeproctext]{ref-vanoutriveHow2019}
Vanoutrive, T., \& Cooper, E. (2019). How just is transportation justice
theory? The issues of paternalism and production {[}Journal Article{]}.
\emph{Transportation Research Part A: Policy and Practice}, \emph{122},
112--119.
https://doi.org/\url{https://doi.org/10.1016/j.tra.2019.02.009}

\bibitem[\citeproctext]{ref-vecchioAccessibilityCapabilitiesApproach2021}
Vecchio, G., \& Martens, K. (2021). Accessibility and the {Capabilities
Approach}: A review of the literature and proposal for conceptual
advancements. \emph{Transport Reviews}, 1--22.
\url{https://doi.org/10.1080/01441647.2021.1931551}

\bibitem[\citeproctext]{ref-vecchioTransportEquityLatin2020}
Vecchio, G., Tiznado-Aitken, I., \& Hurtubia, R. (2020). Transport and
equity in {Latin America}: A critical review of socially oriented
accessibility assessments. \emph{Transport Reviews}, \emph{40}(3),
354--381. \url{https://doi.org/10.1080/01441647.2020.1711828}

\bibitem[\citeproctext]{ref-velhoEffectTransportAccessibility2016}
Velho, R., Holloway, C., Symonds, A., \& Balmer, B. (2016). The {Effect}
of {Transport Accessibility} on the {Social Inclusion} of {Wheelchair
Users}: {A Mixed Method Analysis}. \emph{Social Inclusion}, \emph{4}(3),
24--35. \url{https://doi.org/10.17645/si.v4i3.484}

\bibitem[\citeproctext]{ref-sep-justice-retributive}
Walen, A. (2023). {Retributive Justice}. In E. N. Zalta \& U. Nodelman
(Eds.), \emph{The {Stanford} encyclopedia of philosophy} ({F}all 2023).
\url{https://plato.stanford.edu/archives/fall2023/entries/justice-retributive/};
Metaphysics Research Lab, Stanford University.

\bibitem[\citeproctext]{ref-wangModelingSpatialAccessibility2011}
Wang, L., \& Roisman, D. (2011). Modeling {Spatial Accessibility} of
{Immigrants} to {Culturally Diverse Family Physicians}.
\emph{PROFESSIONAL GEOGRAPHER}, \emph{63}(1), 73--91.
\url{https://doi.org/10.1080/00330124.2010.510087}

\bibitem[\citeproctext]{ref-warrenDevelopingEquitableSustainable2015}
Warren, J., Morris, E., Enoch, M., Magdaleno, I., Arias, Z., \& Guanche,
J. (2015). Developing an equitable and sustainable mobility strategy for
{Havana}. \emph{CITIES}, \emph{45}, 133--141.
\url{https://doi.org/10.1016/j.cities.2015.02.007}

\bibitem[\citeproctext]{ref-welchEquityTransportDistribution2013}
Welch, T. F. (2013). Equity in transport: {The} distribution of transit
access and connectivity among affordable housing units. \emph{Transport
Policy}, \emph{30}, 283--293.
\url{https://doi.org/10.1016/j.tranpol.2013.09.020}

\bibitem[\citeproctext]{ref-wilkinsonmeyersLiveOrdinaryLife2015}
Wilkinson-Meyers, L., Brown, P., McNeill, R., Reeve, J., Patston, P., \&
Baker, R. (2015). To live an ordinary life: Resource needs and
additional costs for people with a physical impairment. \emph{Disability
\& Society}, \emph{30}(7), 976--990.
\url{https://doi.org/10.1080/09687599.2015.1061479}

\bibitem[\citeproctext]{ref-williamsWeStepHow2023}
Williams, R., \& Steil, J. (2023). {``The past we step into and how we
repair it''}: A normative framework for reparative planning.
\emph{Journal of the American Planning Association}, \emph{89}(4),
580--591. \url{https://doi.org/10.1080/01944363.2022.2154247}

\bibitem[\citeproctext]{ref-winteinBeFairClaims2024}
Wintein, S. (2024). To be fair: Claims have amounts and strengths.
\emph{Social Choice and Welfare}, \emph{62}(3), 443--464.
\url{https://doi.org/10.1007/s00355-023-01494-y}

\bibitem[\citeproctext]{ref-wismadiSpatialPreferenceModelling2014}
Wismadi, A., Zuidgeest, M., Brussel, M., \& van Maarseveen, M. (2014).
Spatial {Preference Modelling} for equitable infrastructure provision:
An application of {Sen}'s {Capability Approach}. \emph{Journal of
Geographical Systems}, \emph{16}(1), 19--48.
\url{https://doi.org/10.1007/s10109-013-0185-4}

\bibitem[\citeproctext]{ref-xieWonCycleRoute2018}
Xie, L., \& Spinney, J. (2018). "{I} won't cycle on a route like this;
{I} don't think {I} fully understood what isolation meant": {A} critical
evaluation of the safety principles in {Cycling Level} of {Service}
({CLoS}) tools from a gender perspective. \emph{Travel Behaviour and
Society}, \emph{13}, 197--213.
\url{https://doi.org/10.1016/j.tbs.2018.07.002}

\bibitem[\citeproctext]{ref-xuSocialInequalitiesPark2017}
Xu, M., Xin, J., Su, S., Weng, M., \& Cai, Z. (2017). Social
inequalities of park accessibility in {Shenzhen}, {China}: {The} role of
park quality, transport modes, and hierarchical socioeconomic
characteristics. \emph{Journal of Transport Geography}, \emph{62},
38--50. \url{https://doi.org/10.1016/j.jtrangeo.2017.05.010}

\bibitem[\citeproctext]{ref-xuParetoimprovingPoliciesIdealized2018}
Xu, S.-X., Liu, R., Liu, T.-L., \& Huang, H.-J. (2018). Pareto-improving
policies for an idealized two-zone city served by two congestible modes.
\emph{Transportation Research Part B: Methodological}, \emph{117},
876--891. \url{https://doi.org/10.1016/j.trb.2017.08.010}

\bibitem[\citeproctext]{ref-yenisettyMeasuringAccessibilityVarious2020}
Yenisetty, P., \& Bahadure, P. (2020). Measuring {Accessibility} to
{Various ASFs} from {Public Transit} using {Spatial Distance Measures}
in {Indian Cities}. \emph{ISPRS International Journal of
Geo-Information}, \emph{9}(7). \url{https://doi.org/10.3390/ijgi9070446}

\bibitem[\citeproctext]{ref-zhangLiterature2021}
Zhang, M., \& Zhao, P. (2021). Literature review on urban transport
equity in transitional china: From empirical studies to universal
knowledge {[}Journal Article{]}. \emph{Journal of Transport Geography},
\emph{96}, 103177.
https://doi.org/\url{https://doi.org/10.1016/j.jtrangeo.2021.103177}

\bibitem[\citeproctext]{ref-zhaoRethinking2021}
Zhao, P., \& Li, P. (2021). Rethinking the determinants of vehicle
kilometers traveled (VKT) in an auto-dependent city: Transport policies,
socioeconomic factors and the built environment {[}Journal Article{]}.
\emph{Transportation Planning and Technology}, \emph{44}(3), 273--302.
\url{https://doi.org/10.1080/03081060.2021.1883228}

\bibitem[\citeproctext]{ref-zhaoUnequableSpatialAccessibility2020}
Zhao, P., Li, S., \& Liu, D. (2020). Unequable spatial accessibility to
hospitals in developing megacities: {New} evidence from {Beijing}.
\emph{Health \& Place}, \emph{65}.
\url{https://doi.org/10.1016/j.healthplace.2020.102406}

\bibitem[\citeproctext]{ref-zheEvaluationSharedUse2008}
Zhe, P., Yamanaka, H., Kakihara, K., \& WIT Press. (2008).
\emph{Evaluation of {Shared Use} of {Bicycles} and {Pedestrians} in
{Japan}}. pp 47--56. \url{https://trid.trb.org/view/873583}

\bibitem[\citeproctext]{ref-zhengAreabasedEquitablePricing2020}
Zheng, N., \& Geroliminis, N. (2020). Area-based equitable pricing
strategies for multimodal urban networks with heterogeneous users.
\emph{Transportation Research Part A: Policy and Practice}, \emph{136},
357--374. \url{https://doi.org/10.1016/j.tra.2020.04.009}

\end{CSLReferences}

\section{Acknowledgements and
declarations}\label{acknowledgements-and-declarations}

This work would not have been possible without the help and inspiration
of many individuals not listed as authors. In the early stages of this
report, we received invaluable help from specialists at the University
of Toronto as we learned about various approaches to literature reviews,
particularly the Research Services and Liaison Librarians. These experts
provided guidance in cornering down the topic, identifying guidance
material on literature synthesis and relevant databases, and eased us
into the world of knowledge synthesis. Though our path diverged from a
traditional scoping review, the protocol published in September 2022
(found \href{https://osf.io/rsb92}{here}) was followed with only minor
changes, with only a reduction in the databases searched due to resource
constraints. We also wish to acknowledge the many hours of work by
undergraduate and graduate research assistants who poured through title
and abstract screening to help determine full-text eligibility at the
earliest stages of our search. The volume of literature initially
reviewed was greater than any of us had experienced before. For this
support, we are deeply thankful to Jan Domalaon, Dancel Gayle, Eryn
Maloney, Mahdis Moghadasi, Anika Munir, and Subaita Refaaf.

\textbf{Competing interests' declaration:} The authors declare there are
no personal or financial conflicts of interest.

\textbf{Funding declaration:} This work was supported by the Mobilizing
Justice Partnership and the Canada Graduate Scholarships---Doctoral
Program from the Social Sciences and Humanities Research Council of
Canada (SSHRC).

\textbf{Ethical approval declaration:} Not applicable.

\textbf{Participate consent declaration:} Not applicable.

\textbf{Competing interests' declaration:} The authors declare there are
no personal or financial conflicts of interest.

\textbf{Technology declaration:} During the preparation of this work,
the first and second authors used ChatGPT to assist in reducing the word
count in an early version of the ``An appraisal of the lay of the land''
section. After using this tool, all authors reviewed the manuscript and
take full responsibility for the content of the publication.

\section{Appendix}\label{sec-sect6}

\subsection{Evidence search strategy}\label{sec-sect61}

\begin{figure}

\centering{

\includegraphics{figures/Pflow_chart.tiff}

}

\caption{\label{fig-figA1}PRISMA flow diagram for the evidence selection
process. ES signifies standards and EC signifies conceptualisations.}

\end{figure}%

The evidence selection process is also represented using the Preferred
Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) flow
diagram (Page et al., 2021) in Figure~\ref{fig-figA1}. Notably, two
rounds of exclusion occurred during the assessment for full-text
eligibility. 1710 studies entered step 2, 1223 were excluded and the
remaining 487 papers entered step 3. The data extraction template used
by the reviewers (authorship team) in step 3 revealed that, as expected,
inclusion was initially too generous, and some papers were not
sufficiently relevant, because of a lack of content on standards and/or
conceptual/theoretical elements. In this fashion, 322 papers were
further excluded and data extraction was completed to give a final
corpus of 165 papers. A summary of the reasons for exclusion of the 1545
papers (between steps 2 and 3) are included in Figure~\ref{fig-figA1}.

\begin{figure}

\centering{

\includegraphics{figures/Search-query.png}

}

\caption{\label{fig-figA2}The search query. TS = topic search (keywords,
abstract, title). TASCA = subject categories. Green text area
transportation system related terms, blue text are equity dimension
related terms, purple text are equity/justice conceptualization related
terms, and orange text are standards related terms. Hits corresponds the
number of papers that the search yielded and was retained into the
evidence selection process.}

\end{figure}%

Definitions of the population-concept context (PCC) used in the creation
of the inclusion and exclusion criteria for the search strategy.

\begin{itemize}
\tightlist
\item
  \textbf{Population}: the focus of the included studies should be on
  individuals, groups, communities, or entire regional areas that are
  impacted by passenger transportation infrastructure and systems (i.e.,
  all modes and flows) from the perspective of equity (i.e., fair
  distribution, production, and re-production of burdens and benefits).
  This criteria is reflected in the creation of the first set of topic
  search terms that relate to transportation modes (e.g., ``walking'' OR
  ``cycling'' OR ``transit'' - see green text in Figure~\ref{fig-figA2}
  for the full list).
\item
  \textbf{Concept}: the included studies should also include equity
  dimensions and conceptualizes equity as discussed in the previous
  section. This inclusion criteria is reflected in the second and third
  set of topic search terms developed in the search strategy. These
  terms relate to types of equity dimensions (e.g., ``accessibility'' OR
  ``mobility'' or ``transport-related air pollution'' - see blue text in
  the Figure~\ref{fig-figA2} for the full list) and fairness
  conceptualizations (e.g., ``Justice'' OR ``equity'' - see purple text
  in Figure~\ref{fig-figA2} for the full list).
\item
  \textbf{Context}: the included studies should also be limited to
  publications that include types of standards. Context can be more
  difficult to explicitly search for with key terms so synonyms for
  `standards' were added to the query as a four set of topic search
  terms (e.g., threshold, indicator, criteria - see orange text in
  Figure~\ref{fig-figA2} for full list). Additionally, journal article
  and conference papers, English-language literature from any country,
  any study design (e.g., quantitative, qualitative, or mixed-method
  studies, or conceptual frameworks), and any record published within
  the past 30 years are included (January 1992 to March 2022). The time
  period is selected as the first (to the authors knowledge)
  peer-reviewed article which operationalized standards and fairness
  conceptualization was published in 1996 (Khisty, 1996); we are
  broadening the search by a few years for completeness. English is
  selected as it is the common language spoken across the authorship
  team. Furthermore, papers that explicitly fall within the
  Transportation or related topic/category is included in the query
  (e.g., ``Transportation'', ``Social Sciences'', ``Geography'', ``Civil
  Engineering'', ``Philosophy'' - see the Figure~\ref{fig-figA2} for
  full query).
\end{itemize}

The \textbf{exclusion criteria} for the search are papers that are not
within the inclusion criteria. Specifically:

\begin{itemize}
\tightlist
\item
  Literature published before January 1992.
\item
  Papers which do not include transportation equity dimensions.
\item
  Grey literature, as concepts contained within are frequently published
  in a more developed form in journals.
\end{itemize}

\subsection{Evidence selection and data extraction}\label{sec-sect62}

\begin{figure}

\centering{

\includegraphics{figures/Methods_framework.png}

}

\caption{\label{fig-figA3}Evidence selection workflow. Step 1 (orange)
is title and abstract screening, step 2 (green) is full-text review, and
step 3 (purple) is data extraction.}

\end{figure}%

The following steps summarise the evidence selection process:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  The first step (orange box in Figure~\ref{fig-figA3}) included
  screening all titles and abstracts of papers on whether they included
  transportation equity as defined by the PCC. Each paper was screened
  by two independent reviewers who then voted for inclusion, exclusion,
  or uncertain inclusion. All uncertain papers, conflicting papers, and
  papers missing abstracts were reviewed by a third person for inclusion
  or exclusion.
\item
  The second step (green box in Figure~\ref{fig-figA3}) included
  scanning all full-text papers which passed step 1. These papers were
  reviewed to determine if they included a relevant ``how'', i.e., an
  standard and/or relevant theoretical or conceptual discussion. At this
  stage, papers were evaluated again by two independent reviewers who
  voted for inclusion or exclusion. If an article was voted to be
  excluded, it was tagged with one of five possible reasons for
  exclusion, namely (1) no standards included; (2) no relevant
  conceptual elements included; (3) no standard and no conceptual
  elements included; (4) send back -- QA issue; or (5) other.
  Discrepancies were resolved by a third reviewer.
\item
  In the last step, a data extraction template for each record was
  filled by one reviewer (purple box in Figure~\ref{fig-figA3}). The
  data extraction template was created with the aim of striking a
  balance between the complexity of categories and the simplicity of
  summary; information related to ``Where?'' (the geographical context
  and sphere of life), ``When?'' (temporal circumstances for the
  application of justice), ``Who?'' (the subject of justice), ``What?''
  (the object of justice), and ``How?'' ( standards and
  conceptualisations) was filled out for each study. The following table
  contains the template that was input into \emph{Covidence} and used
  throughout.
\end{enumerate}

Data extraction for each document that passed through all three steps
was then extracted using this template:

\begin{figure}

\centering{

\includegraphics{figures/Data-extract-template.png}

}

\caption{\label{fig-figA4}The data extraction template with associated
defintions.}

\end{figure}%

\subsection{Select reviewed works summarized by the
5WH}\label{sec-sect63}

\begin{table}
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}llll}
\toprule
Dimension & Continent & Conceptualization & Standard \\ 
\midrule\addlinespace[2.5pt]
What?: **mobility and accessibilty** & Where? [@rivasHowAffordableTransportation2018] - South America (select cities) & Analyses how affordable urban public transportation is in select Latin American and Caribbean countries. They look at the estimated average monthly cost of transit trips and average monthly household income and conceptualize **transport-related** **affordability**, especially for the most economically vulnerable (**vertical equity**). & How?: The financial burden of a basket of urban public transportation trips (60 trip fares, representing 30 round-trips per month) should not exceed 10\% of household monthly income. \\ 
What?: **mobility and accessibilty** & Where?: [@bharathyRevisitingClearFloor2018] - North America (USA - National) & This study designed a web-based tool and took a representative sample of wheeled mobility device (WhMD) users anthropometry measurements to determine if the minimum standard suggested by the ADA is sufficient. We understand this conceptualization as a type of **Rights** conceptualization that WhMD should have minimum clear floor space (as described the guidelines in line with the American Disabilities Act) to access bus shelters, bus stop pads, and transit terminals. & How?: The clear floor area for wheelchairs: 760 mm (30 in.) wide by 1220 mm (48 in.) in length as described by the ADA standards. Of note, this minimum clear floor area is insufficient for a variety of the WhMD users. \\ 
What?: **mobility and accessibilty** & [@ryanWhatAreWe2021] - Europe (Stockholm, Gothenburg and Malmo cities in Sweden) & Investigates what the literature and planning process is missing when we measure accessibility by comparing objective and self-reported accounts of accessibility among older people. This paper conceptualizes accessibility as from the position of the **capabilities approach** and **vertical equity** (particularly acknowledging that older people have capabilities that differ from the general population). & How?: Specifically for older populations (aged 65+), the following travel distances are suggested as equitable trip lengths to grocery stores per mode: Walking: less than or equal to 1500m, Combined transit and walking (less than or equal to 1000m (walking element)), Combined car and walking: less than or equal to 1000m less than or equal to 1000m (walking element)), Bicycle: less than or equal to 3000m in addition to travel time threshold of less than 15 mins. \\ 
What?: **mobility and accessibilty** & Where?: [@wismadiSpatialPreferenceModelling2014] - Asia (Yogyakarta, Indonesia) & Explores the equitable provision of transport infrastructure provision: an application of Sen's **capability approach**. Conceptualizes equity through Sen's capability approach and spatial equity. & How?: Areas below the relative poverty line (of its neighbours) can only be located transport resources (i.e., measure in person*kms that can be travelled at car speed, i.e., mobility) based on the following 2 benchmarks (they can be considered, together as the floor/minmum access): 1) Global: standard deviation (SD) distance to mean should be minimized. 2) Local: priority to minimise the differences with its neighbourhood \\ 
What?: **mobility and accessibilty** & Where?: [@zhengAreabasedEquitablePricing2020] - North America & This paper conceptualizies equity in the multimodal network (transit, car) being fair toll-pricing across differences in populatins value of time (VOT). VOT is determined based on household income, with lower income households having lower VOT and thus deserving of lower tolls (vertical equity). From this perspective, a utilitarian perspective that seeks to minimize multimodal traffic congestion through introducing toll-pricing based on VOT is implemented. & How?: suggest that a toll-pricing scheme based on individuals travel value-of-time (lower income people have a lower VOT) is equitable. \\ 
What?: **environmental pollution** & Where?: [@carrierApplicationThreeMethods2014] - North America (Montreal, Canada)  & This work examines the statistical association between different social groups and the concentration of air pollutants. They frame their work from the perspective of environmental equity. We interpret the conceptualizations to be along the lines of **inequitable externalities**, **spatial** and **vertical equity** - transport-related air pollution is a product of road transport and it impacts the air of residents in unequal spatial ways. The paper then frames this impact as unfair, particularly from the perspective of disproportionately disadvantaged residents & How?: The literature suggests that the health implications from the transport-related air pollution from major roadways is most acute at residential distance locations of 200 m or less. Residential locations should not be located within this distance threshold from the perspective of human health. **Environmental+** and **Population standards**: Uses the WHO NO² threshold as a point of comparison (annual concentrations of NO² should not exceed 40 μg/m-3). They argue that even through no neighbourhood, even those disproportionately low income, exceed the WHO limit in this case study, they still suggest that air pollution should not be disproportionately impacting disadvantaged neighbourhoods. It can be interpreted that they use the WHO threshold as a minimum threshold and suggest that air pollution levels should not be impacting disadvantaged populations disproportionately ( a relative population standard) \\ 
What?: **environmental pollution** & Where?: [@jephcoteGeospatialAnalysisNaturally2013] - Europe (Leicester, UK) & Geospatial analysis of naturally occurring boundaries in road-transport emissions and childrens respiratory health across a demographically diverse cityscape. Emperically identifies at what distance away from major roadways children are most impacted by transport-related pollution. This is framed in the perspective of children’s **well-being**. Children are at most risk for acute respiratory distress from elevated levels of air pollution, and as such planning should consider this point of public health. & How?: Finds that children (most vulnerable to air pollution - related to motoized traffic) are most impacted by air pollution within 283 m of a road way. This should be the distance threshold that schools and other childrens facilities are located. \\ 
What?: **health impacts** & Where? [@adlakhaMindGapGender2020] - Asia (Chennai, India) & From the perspective of disparity in gendered physical activity, this paper focuses on women's cycling as both transport and exercise. They advocate for all people achieving physical activity thresholds (**horizontal equity**) but prioritize women and especially women in neighbourhoods with low-walkability and socio-economic status (**vertical equity**). & How?: All people should get 150 min of moderate activity a week or 75 min of vigorous physical activity per week. \\ 
What?: **health impacts** & Where? [@savingmothersDidSavingMothers2019] - Africa (Select urban and rural regions in Uganda) & The **well-being** of mothers, this paper examines the timely access to emergency obsteric and newborn care for child-bearing aged women in Uganda. & How?: 2 hours to the nearest facility with surgical capacity with anesthesia services - this threshold is determined through the onset of bleeding to death if a women with obstetric hemorrhage does not receive adequate treatment). \\ 
What?: **health impacts** & Where?: [@iungmanImpactUrbanTransport2021] - Europe (Madrid and Barcelona, Spain) & They use environmental pollution guidelines, but from the position of health. They investigate the impact of urban and transport planning on attributable mortality burden in Madrid and Barcelona and its distribution by socioeconomic status . Pre-mature mortality is linked to the exposure to pollution and motorized vehicles (**inequitable externalities**). These externalities should not be impacting people disproportionately (**vertical equity**) and should be even across space (**spatial equity**). & How?: All minimum thresholds, if exceeded this is inequitable: NO² concentration 40 μg/m³; PM 2.5 concentration 10 μg/m³; Noise 53dB for average 24 hours; Living with 300 m crow-flies distance from at least .5 hectares of greenspace; and a Change of air temperature of at least 1 ⁰C.  \\ 
What?: **health impacts** & Where?: [@mehdizadehWalkingTimeSchool2017] - Asia (Rasht, Iran) & From the perspective of children’s **well-being**, assesses the walking time to school. They frame walking to school as health-related. & How?: perceived walking time to school for students aged 7-9 yrs is 10 mins, and the longer the PWTS the less likely they were to use an active mode to travel to school. \\ 
What?: **health impacts** & Where? [@murphySupermarketAccessTransport2017] - Oceania (Melbourne, Australia) & Assesses the relationship between supermarket access and transport mode used, the body mass index (BMI) of the mode-user (**wellbeing**) and the equity in access distribution by income (**vertical equity**). & How?: all households should be sufficiently active (greater than 150 min and at least 5 sessions) and households should be within 1 km euclidean distance to supermarket (80-90\% of the dwellings should meet this). Planners should prioritize socially disadvantaged areas to meeting these standards first. \\ 
What?: **transport-related safety** & Where?: [@ferenchakEquityAnalysisProactively2019] - North America (Denver, USA) & Operationalizes and compares an equity analysis of proactively- and reactively-identified traffic safety issues from the perspective of **Spatial equity**, **Vertical equity** and **Inequitable exposure to externalities**. & How?: standards are suggested for both reactive and proactive analysis. First, the lower the number of collisions on the road with pedestrians/cyclists (i.e., reactive safety analysis), the better. No/minimal inequalities for general population vs. equity seeking groups (high proportion of POC and/or low income in tract). Second, the lower the perceived safety, the better (i.e., if travel to school by ped. or bike is unsafe due to traffic conditions). No/minimal inequalities for general population vs. equity seeking groups (high proportion of POC and/or low income in tract). \\ 
What?: **transport-related safety** & Where?: [@zheEvaluationSharedUse2008] - Asia (Tokyo, Takamatsu, and Tokushima) & Evaluates the observed safety of  shared use pedestrian and bicycle paths from the perspective of **well-being**. & How?: the study suggests that the safety threshold for bicycles and pedestrians to coexist on shared infrastructure is less than 0.5 pedestrians/minute per metre of sidewalk (width) and less than 3.0 cyclists/minute per metre of sidewalk (width). The standard for pedestrian/bicycle share use in terms of hourly traffic volume is less than 26 pedestrians / hour and 108 cyclists / hour for 2m wide sidewalks. \\ 
What?: **mobility/accessibility and health impacts** & Where?: [@aldertonWhatMeaningUrban2019] - Asia (Bangkok, Thailand) – **Mobility/ accessibility** and **health** & Establishes short-, medium-, and long-term goals for the city in collaboration with technical leaders within the municipal government for the perspective of **well-being** (urban livability): the standards included in this table relate directly to transportation systems. Indicators are inspired by the Sustainable Development Goals (SDGs) as well other global planning standards. & How?: 1) Green space: \% of residents living < 400 m from public open space, a large park (> 1.5ha), and/or local park, 2) transit access: \% of residents living < 400 m of a local bus stop and <800 m of train station, 3) Facilities: \% of residents living < 400 m of a community centre. The following **Infrastructure standard** is suggested: Canal water quality - dissolved oxygen content of equal to or less than 2.0 mL/L \\ 
What?: **Mobility/ accessibility** and **health impacts** & Where?: [@berheAdaptationDissonanceQuality2014] - Africa (Mekelle, Ethiopia) & Examines adaption and dissonance in the quality of life (QoL) of residents. QoL is conceptualized along the lines of **well-being** and aspects of QoL directly tie into transport systems. They conduct a qualitative QoL survey of residents on the topic of three QoL domains: housing quality, access to important destinations, and affordability. They also measure quantitative indicators associated with these domains. We assume the equity goal for this paper is that subjective and objective QoL measures should not be mismatched: as discussed by the authors of this study, subjective QoL is higher than objective QoL the participant is experiencing adaption and in the reverse scenario the participant is experience dissonance. & How: 1 \& 2) Access to primary or secondary education facility, percentage of households living within 1 km or 2km (walking distance), respectively from a primary school or secondary school. 3) Access to health facility, percentage of households within 40 min walking time from a health facility. 4) Access to public transport, percentage of households within a distance of 500 m from a mini-bus stop. **Population standards**: 1) Adequate family income, percentage of households earning more than the official poverty line. 2) Subjective QoL is constructed based on the households level of satisfaction for each of the eight indicators using a six point Likert-scale (1=very satisfied to 6=very dissatisfied). \\ 
What?: **mobility/accessibility, health impacts, and safety** & Where? [@agostfelipInclusiveModelAssessing2021] - Europe (Castellon, Spain) & Conceptualizes equity through age-friendly urban spaces that reduce (and eliminate) conditions for **transport-related social exclusion** for older populations and prioritize those who are economically vulnerable (**vertical equity**). These guidelines are inspired by the SDGs in addition to planning guidelines used national, regional, and local guidelines used in Spain. & How: 1) Access to facilities needed for old age health. Minimum distance thresholds from the geometric center of neighbourhood are suggested: at least: 1000 m from health facilities (600 m or less is preferred), elderly-specific care facilities and shops should be 600 m (300 m or less is preferred). **Population standards**: 1) Certain neighbourhoods should be prioritized above others. From this papers focus on age-friendly urban environment, they suggest that if the neighbourhood has an average old age indicator (i.e., greater than 64 years, and/or greater than 79 years, and/or aging ratio of persons aged greater than 64 relative to 15 to 64 age) should be prioritized. 2) Economic vulnerable and non-civically engaged neighbourhoods should also be prioritized. If the neighbourhood has a lower percentage of civic associations within the neighbourhood than average, and/or household income, and/or a higher than average interventions for dependency and/or social subsidies, they should be priorized. **Infrastructure standards **: 1) Green space: should be at least 10 m2 per inhabitant in the neighbourhood, greater than 15 m2 per inhab. is the goal. 2) As related to sidewalk infrastructure at least 50\% of all sidewalks (preferably 75\% or greater) should: have a width of 1.5m or larger, ramps should have a grade of 8\% or less, be well maintained (free from deficiencies), be paved for pedestrian use, and cover transit stops. 3) Lighting is critical for traffic-safety and a sense of safety overall. As such, at least 50\% roads should: have a min. of 35 lux (road traffic) and 20 lux (pedestrian streets), and adapted traffic lights. 4) Buildings should be age-friendly. As a proxy for the quality of residential living space quality, at least 50\% of residential buildings in a neighbourhood should be built within the last 50 years (preferably 75\% or more). In terms of physical access into the buildings, at least 10\% should have elevators and accessible entrances (preferably 25\% or more). **Environment + standards **: 1) Noise at the street level should be less than 55 dB and 45 dB (but preferably less than 50 dB and 40 dB) in the daytime and nighttime, respectively. \\ 
What?: **mobility/accessibility and safety** & Where?: [@mateobabianoPedestrianNeedsMatter2016] - Asian (Manila, Philippine) & The perception of pedestrians’ walking environments should be sufficient across 6 themes. Equity is conceptualized around **spatial equity** (equally fair walking environments for all locations) and **rights** (the right to mobility/accessibility for pedestrians) & How?: perceived pedestrian perception on protection, ease, equitable access, mobility, identity, and enjoyment must be met. \\ 
\bottomrule
\end{tabular*}
\end{table}

\end{document}

### 2025-03-11 paper/paper.qmd
'paper' - submitted to journal.

--- 2025-03-09 Journal-manuscript/Journal-manuscript.qmd
@@
 ---
-title: "Searching for standards of fairness in the transportation literature"
-csl: apa7.csl
+title: Searching for standards of fairness in the transportation literature
 author:
   - name: AAA #Anastasia Soukhov
+    #orcid: 0000-0003-4371-4831
+    # email: soukhoa@example.com
     # affiliations: 
     #     - ref: MAC
-    # orcid: 0000-0003-4371-4831
-    # email: soukho@mcmaster.ca
+    #       name: McMaster University
+    #       department: School of Earth, Environment and Society
+    #       address: 1280 Main Street West
+    #       city: Hamilton, ON
+    #       country: Canada
+    #       postal-code: L8S 4L8
+    # attributes:
+    #     corresponding: true
   - name: BBB #Ignacio Tiznado-Aitken
+    # orcid: 0000-0002-7385-2357
+    # email: i.tiznadoaitken@utoronto.ca
     # affiliations:
     #     - ref: TORONTO
-    # orcid: 0000-0002-7385-2357
-    # email: i.tiznadoaitken@utoronto.ca
-   - name: CCC #Matthew Palm
-    # affiliations:
-    #   - ref: UNC
-    # orcid: 0000-0002-8800-2777
-    # email: palmmatt@unc.edu
-   - name: DDD #Steven Farber
-    # affiliations:
-    #   - ref: TORONTO
-    # orcid: 0000-0002-3870-5984
-    # email: steven.farber@utoronto.ca
-   - name: EEE #Antonio Páez
-    # affiliations:
-    #   - ref: MAC
-    # orcid: 0000-0001-6912-9919
-    # email: paezha@mcmaster.ca
-# 
-# affiliations:
-#   - id: MAC
-#     name: McMaster University
-#     department: School of Earth, Environment and Society
-#     city: Hamilton, ON
-#     country: Canada
-#   - id: TORONTO
     #       name: University of Toronto Scarborough
@@
     #       country: Canada
-#   - id: UNC
+    #       postal-code: M1C 1A4
+  # - name: Matthew Palm
+  #   # orcid: 0000-0002-8800-2777
+  #   email: palmmatt@unc.edu
+  #   affiliations:
+  #       - ref: UNC
   #         name: The University of North Carolina at Chapel Hill
@@
   #         country: USA
+  #         postal-code: 27599
+  # - name: Steven Farber
+  #   #orcid: 0000-0002-3870-5984
+  #   email: steven.farber@utoronto.ca
+  #   affiliations:
+  #       - ref: TORONTO
+  #         name: University of Toronto Scarborough
+  #         department: Department of Geography and Planning
+  #         city: Toronto, ON
+  #         country: Canada
+  #         postal-code: M1C 1A4
+  # - name: Antonio Páez
+  #   #orcid: 0000-0001-6912-9919
+  #   email: paezha@mcmaster.ca
+  #   affiliations: 
+  #       - ref: MAC
+  #         name: McMaster University
+  #         department: School of Earth, Environment and Society
+  #         address: 1280 Main Street West
+  #         city: Hamilton, ON
+  #         country: Canada
+  #         postal-code: L8S 4L8
 abstract: |
@@
 
+keywords: [transportation; equity; justice; fairness; standards; transportation planning; literature review]
+date: last-modified
+bibliography: bibliography.bib
 format:
-  #docx
   elsevier-pdf:
-    keep-tex: true
+    #keep-tex: true
     include-in-header:
       text: |
-        \usepackage{graphicx}
-        \usepackage{unicode-math}
-        \usepackage{hyperref}
-        \def\tightlist{}
-        \usepackage{setspace}
-        \doublespacing
         \usepackage{lineno}
@@
     journal:
-      name: Travel Behavior and Society
+      #name: Travel Behavior and Society
       formatting: preprint
-      model: 3p
+      # model: 3p # Don't set a model with preprint
       cite-style: authoryear
-keywords: [transportation; equity; justice; fairness; standards; transportation planning; literature review]
-bibliography: bibliography.bib
-
-#filters:
-#  - docx-landscape.lua
 ---
@@
 df_all_sub_ob_just_reordered_plot
-ggsave("figures/df_all_sub_ob_just_reordered_plot.tiff", dpi=300)
+ggsave("figures/df_all_sub_ob_just_reordered_plot.tiff", dpi=300) #fig1
 ```
@@
 standards_conc_percentages_plot_and_table
-ggsave("figures/standards_conc_percentages_plot_and_table.tiff", dpi=300)
+ggsave("figures/standards_conc_percentages_plot_and_table.tiff", dpi=300) #fig2
 ```
@@
 The initial draft of the protocol was refined from preliminary searches of related reviews [e.g., @zhangLiterature2021; @pereiraDistributiveJusticeEquity2017a; @vecchioAccessibilityCapabilitiesApproach2021] and in consultation with research services librarians.
-The final protocol was released on OSF [@soukhovEquityJusticeConceptualizations2022], and the methods are summarised as follows.
+The final protocol was released on OSF [redacted for double-blind review] <!--[@soukhovEquityJusticeConceptualizations2022]-->, and the methods are summarised as follows.
 
@@
 #| label: fig-fig1
-#| fig-cap: "The proportion of papers in each 'When', 'Where', 'Who', and 'What' by topics in each category. Topic categories were generated by the authors based on the reviewed literature."
-
-knitr::include_graphics("figures/df_all_sub_ob_just_reordered_plot.tiff")
-```
+#| fig-cap: "The proportion of papers in each When, Where, Who, and What by topics in each category. Topic categories were generated by the authors based on the reviewed literature."
+
+knitr::include_graphics("figures/Fig1.png")
+```
+
 ```{r plot-standards-concs-perc, warning=FALSE}
 #| label: fig-fig2
-#| fig-cap: "The proportion of papers by topics in the 'How' categories.
-'How' is split into two categories: topics related to fairness conceptualisation (horizontal axis) and topics related to standards (vertical axis).
+#| fig-cap: "The proportion of papers by topics in the How categories.
+How is split into two categories: topics related to fairness conceptualisation (horizontal axis) and topics related to standards (vertical axis).
 Topic categories were generated by the authors based on the reviewed literature."
 
-knitr::include_graphics("figures/standards_conc_percentages_plot_and_table.tiff")
+knitr::include_graphics("figures/Fig2.png")
 ```
@@
 
+<!--
 # Acknowledgements and declarations
@@
 
-**Technology declaration:** During the preparation of this work, the first and second authors used ChatGPT to assist in reducing the word count in an early version of the “An appraisal of the lay of the land” section.
+**Technology declaration:** During the preparation of this work, the first and second authors used ChatGPT to assist in reducing the word count in an early version of the “Synthesis of Findings” section.
 After using this tool, all authors reviewed the manuscript and take full responsibility for the content of the publication.
+-->
 
 # Appendix {#sec-sect6}
-
-<!-- ## Why? A framework extension to analyse questions of justice {#sec-sect61} -->
-
-<!-- To give an example drawing from the Canadian Charter of Rights and Freedoms, a number of rights and fundamental freedoms [including the liberty to move freely; see @chartopediaArt72023] are recognized to exist in Canada "without discrimination by reason of race, national origin, colour, religion or sex".
-Notice that this declaration includes several individual traits that define the subjects of justice, but does not consider age as one aspect of the person.
-Does this mean that they apply universally irrespective of age?
-Surely not, since no reasonable person would consider a toddler's demands for freedom of association as absolute.
-So, when do these rights fully apply in the life of a person?
-In 2017, Adrian Crook of Vancouver, B.C., was warned by the province's Ministry of Children and Family Development that his kids could not be out of home in the community, alone or with other kids the same age, without supervision.
-The Ministry's argument was that children riding the bus unsupervised compromised the parent's ability to provide care and placed them at risk [@brendHere2017].
-Crook argued that the goal of teaching his kids (then aged 7, 8, 9 and 11) to ride transit to school was to raise self-reliant children [@stueckYoung2017].
-While both sides discovered to everyone's surprise that no clear rules existed regarding children's independent transit use [@brendHere2017], statistically there is evidence that travel by car is riskier than travel by bus [@morencyTraveling2018].
-If the public good is to reduce risk, shouldn't children be banned from riding in cars?
-The Canadian legal system concerns itself with fundamental freedoms and rights, one of which is the liberty to move freely.
-This provides an answer (of many) as to "why" we would even consider children as subjects of justice.
-But the devil is in the details, and answering the other questions discussed above is essential to pin the devil by the tail.
-Are children under the age of 10 subjects of justice (i.e., are they a "who") and who decides if they are (i.e., the courts or the parents in this case)?
-If so, is the ability to use transit unsupervised an object of justice that should be justly distributed (i.e., the "what") and under what circumstances (e.g, to school, to a social event; during the day, or late at night?).
-"When" (e.g., for all children going forward, for only 5 years) and "where" (e.g., only in the suburbs, across the country) do these questions apply?
-Eventually, in 2020, a court of appeals ruled in Crook's favor, allowing his children to continue to ride the bus, thus providing much needed clarity with respect to the use of transit by children who desire and are encouraged to exercise independent mobility [@stueckCourt2020].
-In actuality, the courts _created_ a standard of equity: starting at a certain age, children are objects of justice from the perspective of freedom of movement, and they are due the benefits of independent mobility by transit.
--->
-
-<!-- Another perspective on the answer to "Why?" are cultural _norms_ that provide a rationale.
-In this case, the liberty of movement for all---including children---reigns supreme.
-But a competing concern is that the outside is dangerous, and it is so in part due to vehicular traffic [@morencyTraveling2018], and in part because of potential violence from others.
-Parents may have over-pronounced 'stranger danger' based on neighbourhood and socio-economic characteristics of neighbourhoods [@francisLlBeDriving2017] and children themselves may fear potential violence based on real threats of victimization [@painParanoidParentingRematerializing2006].
-Whose voice, what perception of fear, and actions we take to intervene define and reproduce cultural norms, all play into the rationale behind the question "Why?".
--->
 
@@
 #| fig-cap: "PRISMA flow diagram for the evidence selection process. ES signifies standards and EC signifies conceptualisations."
-knitr::include_graphics("figures/Pflow_chart.tiff")
+knitr::include_graphics("figures/Fig3_A.png")
 ```
@@
 Hits corresponds the number of papers that the search yielded and was retained into the evidence selection process."
-knitr::include_graphics("figures/Search-query.png")
+knitr::include_graphics("figures/Fig4_A.png")
 ```
@@
 #| fig-cap: "Evidence selection workflow. Step 1 (orange) is title and abstract screening, step 2 (green) is full-text review, and step 3 (purple) is data extraction."
-knitr::include_graphics("figures/Methods_framework.png")
+knitr::include_graphics("figures/Fig5_A.png")
 ```
@@
 
-:::landscape
-
-## Select reviewed works summarized by the 5WH {#sec-sect63}
+<!--:::landscape-->
+\newpage
+
+## Select reviewed works summarised using the 5WH {#sec-sect63}
 
 ```{r}
-dat <- data.frame(matrix(c("What?: **mobility and accessibilty**",
-                           "Where? [@rivasHowAffordableTransportation2018] - South America (select cities)",
+#1. What dimension?. 2) when and where?, 3) How is it conceptualised wrt who/what? 4. How is the standard set? 
+dat <- data.frame(matrix(c("**Mobility and accessibility**",#What?: 
+                           "[@rivasHowAffordableTransportation2018] - South America (select cities)", #Where? 
                            "Analyses how affordable urban public transportation is in select Latin American and Caribbean countries.
 They look at the estimated average monthly cost of transit trips and average monthly household income and conceptualize **transport-related** **affordability**, especially for the most economically vulnerable (**vertical equity**).",
-                           "How?: The financial burden of a basket of urban public transportation trips (60 trip fares, representing 30 round-trips per month) should not exceed 10% of household monthly income.",
-                           "What?: **mobility and accessibilty**",
-                           "Where?: [@bharathyRevisitingClearFloor2018] - North America (USA - National)",
+                           " The financial burden of a basket of urban public transportation trips (60 trip fares, representing 30 round-trips per month) should not exceed 10% of household monthly income.", #How?:
+                           "**Mobility and accessibility**", #What?: 
+                           "[@bharathyRevisitingClearFloor2018] - North America (USA - National)", #Where?: 
                            "This study designed a web-based tool and took a representative sample of wheeled mobility device (WhMD) users anthropometry measurements to determine if the minimum standard suggested by the ADA is sufficient.
 We understand this conceptualization as a type of **Rights** conceptualization that WhMD should have minimum clear floor space (as described the guidelines in line with the American Disabilities Act) to access bus shelters, bus stop pads, and transit terminals.",
-                           "How?: The clear floor area for wheelchairs: 760 mm (30 in.) wide by 1220 mm (48 in.) in length as described by the ADA standards.
-Of note, this minimum clear floor area is insufficient for a variety of the WhMD users.",
-                           "What?: **mobility and accessibilty**",
+                           "The clear floor area for wheelchairs: 760 mm (30 in.) wide by 1220 mm (48 in.) in length as described by the ADA standards.
+Of note, this minimum clear floor area is insufficient for a variety of the WhMD users.",#How?:
+                           "**Mobility and accessibility**",#What?: 
                            "[@ryanWhatAreWe2021] - Europe (Stockholm, Gothenburg and Malmo cities in Sweden)",
@@
 This paper conceptualizes accessibility as from the position of the **capabilities approach** and **vertical equity** (particularly acknowledging that older people have capabilities that differ from the general population).",
-                           "How?: Specifically for older populations (aged 65+), the following travel distances are suggested as equitable trip lengths to grocery stores per mode: Walking: less than or equal to 1500m, Combined transit and walking (less than or equal to 1000m (walking element)), Combined car and walking: less than or equal to 1000m less than or equal to 1000m (walking element)), Bicycle: less than or equal to 3000m in addition to travel time threshold of less than 15 mins.",
-                           "What?: **mobility and accessibilty**",
-                           "Where?: [@wismadiSpatialPreferenceModelling2014] - Asia (Yogyakarta, Indonesia)",
+                           "Specifically for older populations (aged 65+), the following travel distances are suggested as equitable trip lengths to grocery stores per mode: Walking: less than or equal to 1500m, Combined transit and walking (less than or equal to 1000m (walking element)), Combined car and walking: less than or equal to 1000m less than or equal to 1000m (walking element)), Bicycle: less than or equal to 3000m in addition to travel time threshold of less than 15 mins.", #How?:
+                           "**Mobility and accessibility**",#What?: 
+                           "[@wismadiSpatialPreferenceModelling2014] - Asia (Yogyakarta, Indonesia)", #Where?: 
                            "Explores the equitable provision of transport infrastructure provision: an application of Sen's **capability approach**.
 Conceptualizes equity through Sen's capability approach and spatial equity.",
-                           "How?: Areas below the relative poverty line (of its neighbours) can only be located transport resources (i.e., measure in person*kms that can be travelled at car speed, i.e., mobility) based on the following 2 benchmarks (they can be considered, together as the floor/minmum access): 1) Global: standard deviation (SD) distance to mean should be minimized.
-2) Local: priority to minimise the differences with its neighbourhood",
-                           "What?: **mobility and accessibilty**",
-                           "Where?: [@zhengAreabasedEquitablePricing2020] - North America",
-                           "This paper conceptualizies equity in the multimodal network (transit, car) being fair toll-pricing across differences in populatins value of time (VOT).
+                           "Areas below the relative poverty line (of its neighbours) can only be located transport resources (i.e., measure in person*kms that can be travelled at car speed, i.e., mobility) based on the following 2 benchmarks (they can be considered, together as the floor/minimum access): 1) Global: standard deviation (SD) distance to mean should be minimized.
+2) Local: priority to minimise the differences with its neighbourhood", #How?:
+                           "**Mobility and accessibility**", #What?: 
+                           "[@zhengAreabasedEquitablePricing2020] - North America", #Where?: 
+                           "This paper conceptualizes equity in the multimodal network (transit, car) being fair toll-pricing across differences in populations value of time (VOT).
 VOT is determined based on household income, with lower income households having lower VOT and thus deserving of lower tolls (vertical equity).
 From this perspective, a utilitarian perspective that seeks to minimize multimodal traffic congestion through introducing toll-pricing based on VOT is implemented.",
-                           "How?: suggest that a toll-pricing scheme based on individuals travel value-of-time (lower income people have a lower VOT) is equitable.",
-
-                           "What?: **environmental pollution**",
-                           "Where?: [@carrierApplicationThreeMethods2014] - North America (Montreal, Canada) ",
+                           "Suggests that a toll-pricing scheme based on individuals travel value-of-time (lower income people have a lower VOT) is equitable.", #How?:
+
+                           "**Environmental pollution**", #What?: 
+                           "[@carrierApplicationThreeMethods2014] - North America (Montreal, Canada) ", #Where?:
                            "This work examines the statistical association between different social groups and the concentration of air pollutants.
@@
 The paper then frames this impact as unfair, particularly from the perspective of disproportionately disadvantaged residents",
-                           "How?: The literature suggests that the health implications from the transport-related air pollution from major roadways is most acute at residential distance locations of 200 m or less.
+                           "The literature suggests that the health implications from the transport-related air pollution from major roadways is most acute at residential distance locations of 200 m or less.
 Residential locations should not be located within this distance threshold from the perspective of human health.
@@
 They argue that even through no neighbourhood, even those disproportionately low income, exceed the WHO limit in this case study, they still suggest that air pollution should not be disproportionately impacting disadvantaged neighbourhoods.
-It can be interpreted that they use the WHO threshold as a minimum threshold and suggest that air pollution levels should not be impacting disadvantaged populations disproportionately ( a relative population standard)",
-                           "What?: **environmental pollution**",
-                           "Where?: [@jephcoteGeospatialAnalysisNaturally2013] - Europe (Leicester, UK)",
-                           "Geospatial analysis of naturally occurring boundaries in road-transport emissions and childrens respiratory health across a demographically diverse cityscape.
+It can be interpreted that they use the WHO threshold as a minimum threshold and suggest that air pollution levels should not be impacting disadvantaged populations disproportionately ( a relative population standard)", #How?:
+                           "**Environmental pollution**", #what?
+                           "[@jephcoteGeospatialAnalysisNaturally2013] - Europe (Leicester, UK)", #Where?: 
+                           "Geospatial analysis of naturally occurring boundaries in road-transport emissions and children's respiratory health across a demographically diverse cityscape.
 Emperically identifies at what distance away from major roadways children are most impacted by transport-related pollution.
@@
 Children are at most risk for acute respiratory distress from elevated levels of air pollution, and as such planning should consider this point of public health.",
-                           "How?: Finds that children (most vulnerable to air pollution - related to motoized traffic) are most impacted by air pollution within 283 m of a road way.
-This should be the distance threshold that schools and other childrens facilities are located.",
-                           "What?: **health impacts**",
-                           "Where? [@adlakhaMindGapGender2020] - Asia (Chennai, India)",
+                           "Finds that children (most vulnerable to air pollution - related to motorized traffic) are most impacted by air pollution within 283 m of a road way.
+This should be the distance threshold that schools and other children's facilities are located.", #How?:
+                           "**Health impacts**", #What?: 
+                           "[@adlakhaMindGapGender2020] - Asia (Chennai, India)", #Where? 
                            "From the perspective of disparity in gendered physical activity, this paper focuses on women's cycling as both transport and exercise.
 They advocate for all people achieving physical activity thresholds (**horizontal equity**) but prioritize women and especially women in neighbourhoods with low-walkability and socio-economic status (**vertical equity**).",
-                           "How?: All people should get 150 min of moderate activity a week or 75 min of vigorous physical activity per week.",
-                           "What?: **health impacts**",
-                           "Where? [@savingmothersDidSavingMothers2019] - Africa (Select urban and rural regions in Uganda)",
-                           "The **well-being** of mothers, this paper examines the timely access to emergency obsteric and newborn care for child-bearing aged women in Uganda.",
+                           "All people should get 150 min of moderate activity a week or 75 min of vigorous physical activity per week.", #How?: 
+                           "**Health impacts**", #How?: 
+                           "[@savingmothersDidSavingMothers2019] - Africa (Select urban and rural regions in Uganda)", #Where? 
+                           "The **well-being** of mothers, this paper examines the timely access to emergency obstetric and newborn care for child-bearing aged women in Uganda.",
                            "How?: 2 hours to the nearest facility with surgical capacity with anesthesia services - this threshold is determined through the onset of bleeding to death if a women with obstetric hemorrhage does not receive adequate treatment).",
-                           "What?: **health impacts**",
-                           "Where?: [@iungmanImpactUrbanTransport2021] - Europe (Madrid and Barcelona, Spain)",
+                           "**Health impacts**", #What?: 
+                           "[@iungmanImpactUrbanTransport2021] - Europe (Madrid and Barcelona, Spain)",
                            "They use environmental pollution guidelines, but from the position of health.
@@
 Pre-mature mortality is linked to the exposure to pollution and motorized vehicles (**inequitable externalities**).
-These externalities should not be impacting people disproportionately (**vertical equity**) and should be even across space (**spatial equity**).",
-                           "How?: All minimum thresholds, if exceeded this is inequitable: NO² concentration 40 μg/m³; PM 2.5 concentration 10 μg/m³; Noise 53dB for average 24 hours; Living with 300 m crow-flies distance from at least .5 hectares of greenspace; and a Change of air temperature of at least 1 ⁰C.
-",
-                           "What?: **health impacts**",
-                           "Where?: [@mehdizadehWalkingTimeSchool2017] - Asia (Rasht, Iran)",
+These externalities should not be impacting people disproportionately (**vertical equity**) and should be even across space (**spatial equity**).", #Where?:
+                           "All minimum thresholds, if exceeded this is inequitable: NO² concentration 40 μg/m³; PM 2.5 concentration 10 μg/m³; Noise 53dB for average 24 hours; Living with 300 m crow-flies distance from at least .5 hectares of greenspace; and a Change of air temperature of at least 1 ⁰C.
+", #How?:
+                           "**Health impacts**", #What?: 
+                           "[@mehdizadehWalkingTimeSchool2017] - Asia (Rasht, Iran)", #Where?: 
                            "From the perspective of children’s **well-being**, assesses the walking time to school. They frame walking to school as health-related.",
-                           "How?: perceived walking time to school for students aged 7-9 yrs is 10 mins, and the longer the PWTS the less likely they were to use an active mode to travel to school.",
-                           "What?: **health impacts**",
-                           "Where? [@murphySupermarketAccessTransport2017] - Oceania (Melbourne, Australia)",
+                           "Perceived walking time to school for students aged 7-9 yrs is 10 mins, and the longer the PWTS the less likely they were to use an active mode to travel to school.",#How?: 
+                           "**Health impacts**", #What?: 
+                           "[@murphySupermarketAccessTransport2017] - Oceania (Melbourne, Australia)", #Where? 
                            "Assesses the relationship between supermarket access and transport mode used, the body mass index (BMI) of the mode-user (**wellbeing**) and the equity in access distribution by income (**vertical equity**).",
-                           "How?: all households should be sufficiently active (greater than 150 min and at least 5 sessions) and households should be within 1 km euclidean distance to supermarket (80-90% of the dwellings should meet this).
-Planners should prioritize socially disadvantaged areas to meeting these standards first.",
-
-                           "What?: **transport-related safety**",
-                           "Where?: [@ferenchakEquityAnalysisProactively2019] - North America (Denver, USA)",
+                           "All households should be sufficiently active (greater than 150 min and at least 5 sessions) and households should be within 1 km euclidean distance to supermarket (80-90% of the dwellings should meet this).
+Planners should prioritize socially disadvantaged areas to meeting these standards first.",#How?:
+
+                           "**Transport-related safety**", #What?: 
+                           "[@ferenchakEquityAnalysisProactively2019] - North America (Denver, USA)", #Where?: 
                            "Operationalizes and compares an equity analysis of proactively- and reactively-identified traffic safety issues from the perspective of **Spatial equity**, **Vertical equity** and **Inequitable exposure to externalities**.",
-                           "How?: standards are suggested for both reactive and proactive analysis.
+                           "Standards are suggested for both reactive and proactive analysis.
 First, the lower the number of collisions on the road with pedestrians/cyclists (i.e., reactive safety analysis), the better.
@@
 No/minimal inequalities for general population vs.
-equity seeking groups (high proportion of POC and/or low income in tract).",
-                           "What?: **transport-related safety**",
-                           "Where?: [@zheEvaluationSharedUse2008] - Asia (Tokyo, Takamatsu, and Tokushima)",
+equity seeking groups (high proportion of POC and/or low income in tract).",#How?:
+                           "**Transport-related safety**", #What?: 
+                           "[@zheEvaluationSharedUse2008] - Asia (Tokyo, Takamatsu, and Tokushima)", #Where?: 
                            "Evaluates the observed safety of  shared use pedestrian and bicycle paths from the perspective of **well-being**.",
-                           "How?: the study suggests that the safety threshold for bicycles and pedestrians to coexist on shared infrastructure is less than 0.5 pedestrians/minute per metre of sidewalk (width) and less than 3.0 cyclists/minute per metre of sidewalk (width).
-The standard for pedestrian/bicycle share use in terms of hourly traffic volume is less than 26 pedestrians / hour and 108 cyclists / hour for 2m wide sidewalks.",
-
-                           "What?: **mobility/accessibility and health impacts**",
-                           "Where?: [@aldertonWhatMeaningUrban2019] - Asia (Bangkok, Thailand) – **Mobility/ accessibility** and **health**",
+                           "The study suggests that the safety threshold for bicycles and pedestrians to coexist on shared infrastructure is less than 0.5 pedestrians/minute per metre of sidewalk (width) and less than 3.0 cyclists/minute per metre of sidewalk (width).
+The standard for pedestrian/bicycle share use in terms of hourly traffic volume is less than 26 pedestrians / hour and 108 cyclists / hour for 2m wide sidewalks.", #How?:
+
+                           "**Mobility, accessibility and health impacts**", #What?: 
+                           "[@aldertonWhatMeaningUrban2019] - Asia (Bangkok, Thailand) – **Mobility/ accessibility** and **health**", #Where?: 
                            "Establishes short-, medium-, and long-term goals for the city in collaboration with technical leaders within the municipal government for the perspective of **well-being** (urban livability): the standards included in this table relate directly to transportation systems.
 Indicators are inspired by the Sustainable Development Goals (SDGs) as well other global planning standards.",
-                           "How?: 1) Green space: % of residents living < 400 m from public open space, a large park (> 1.5ha), and/or local park, 2) transit access: % of residents living < 400 m of a local bus stop and <800 m of train station, 3) Facilities: % of residents living < 400 m of a community centre.
-The following **Infrastructure standard** is suggested: Canal water quality - dissolved oxygen content of equal to or less than 2.0 mL/L",
-                           "What?: **Mobility/ accessibility** and **health impacts**",
-                           "Where?: [@berheAdaptationDissonanceQuality2014] - Africa (Mekelle, Ethiopia)",
+                           "1) Green space: % of residents living < 400 m from public open space, a large park (> 1.5ha), and/or local park, 2) transit access: % of residents living < 400 m of a local bus stop and <800 m of train station, 3) Facilities: % of residents living < 400 m of a community centre.
+The following **Infrastructure standard** is suggested: Canal water quality - dissolved oxygen content of equal to or less than 2.0 mL/L", #How?:
+                           "**Mobility/ accessibility** and **health impacts**", #What?: 
+                           "[@berheAdaptationDissonanceQuality2014] - Africa (Mekelle, Ethiopia)", #Where?: 
                          "Examines adaption and dissonance in the quality of life (QoL) of residents.
@@
 We assume the equity goal for this paper is that subjective and objective QoL measures should not be mismatched: as discussed by the authors of this study, subjective QoL is higher than objective QoL the participant is experiencing adaption and in the reverse scenario the participant is experience dissonance.",
-                           "How: 1 & 2) Access to primary or secondary education facility, percentage of households living within 1 km or 2km (walking distance), respectively from a primary school or secondary school.
+                           "1 & 2) Access to primary or secondary education facility, percentage of households living within 1 km or 2km (walking distance), respectively from a primary school or secondary school.
 3) Access to health facility, percentage of households within 40 min walking time from a health facility.
@@
 **Population standards**: 1) Adequate family income, percentage of households earning more than the official poverty line.
-2) Subjective QoL is constructed based on the households level of satisfaction for each of the eight indicators using a six point Likert-scale (1=very satisfied to 6=very dissatisfied).",
-                           "What?: **mobility/accessibility, health impacts, and safety**",
-                           "Where? [@agostfelipInclusiveModelAssessing2021] - Europe (Castellon, Spain)",
+2) Subjective QoL is constructed based on the households level of satisfaction for each of the eight indicators using a six point Likert-scale (1=very satisfied to 6=very dissatisfied).", #How:
+                           "**Mobility, accessibility, health impacts, and safety**", #What?: 
+                           "[@agostfelipInclusiveModelAssessing2021] - Europe (Castellon, Spain)",#Where? 
                            "Conceptualizes equity through age-friendly urban spaces that reduce (and eliminate) conditions for **transport-related social exclusion** for older populations and prioritize those who are economically vulnerable (**vertical equity**).
 These guidelines are inspired by the SDGs in addition to planning guidelines used national, regional, and local guidelines used in Spain.",
-                           "How: 1) Access to facilities needed for old age health.
+                           "1) Access to facilities needed for old age health.
 Minimum distance thresholds from the geometric center of neighbourhood are suggested: at least: 1000 m from health facilities (600 m or less is preferred), elderly-specific care facilities and shops should be 600 m (300 m or less is preferred).
 **Population standards**: 1) Certain neighbourhoods should be prioritized above others.
-From this papers focus on age-friendly urban environment, they suggest that if the neighbourhood has an average old age indicator (i.e., greater than 64 years, and/or greater than 79 years, and/or aging ratio of persons aged greater than 64 relative to 15 to 64 age) should be prioritized.
+From this paper's focus on age-friendly urban environment, they suggest that if the neighbourhood has an average old age indicator (i.e., greater than 64 years, and/or greater than 79 years, and/or aging ratio of persons aged greater than 64 relative to 15 to 64 age) should be prioritized.
 2) Economic vulnerable and non-civically engaged neighbourhoods should also be prioritized.
-If the neighbourhood has a lower percentage of civic associations within the neighbourhood than average, and/or household income, and/or a higher than average interventions for dependency and/or social subsidies, they should be priorized.
+If the neighbourhood has a lower percentage of civic associations within the neighbourhood than average, and/or household income, and/or a higher than average interventions for dependency and/or social subsidies, they should be prioritized.
 **Infrastructure standards **: 1) Green space: should be at least 10 m2 per inhabitant in the neighbourhood, greater than 15 m2 per inhab.
@@
 In terms of physical access into the buildings, at least 10% should have elevators and accessible entrances (preferably 25% or more).
-**Environment + standards **: 1) Noise at the street level should be less than 55 dB and 45 dB (but preferably less than 50 dB and 40 dB) in the daytime and nighttime, respectively.",
-                           "What?: **mobility/accessibility and safety**",
-                           "Where?: [@mateobabianoPedestrianNeedsMatter2016] - Asian (Manila, Philippine)",
+**Environment + standards **: 1) Noise at the street level should be less than 55 dB and 45 dB (but preferably less than 50 dB and 40 dB) in the daytime and nighttime, respectively.",#How:
+                           "**Mobility, accessibility and safety**",#What?: 
+                           "[@mateobabianoPedestrianNeedsMatter2016] - Asian (Manila, Philippine)",#Where?: 
                            "The perception of pedestrians’ walking environments should be sufficient across 6 themes.
 Equity is conceptualized around **spatial equity** (equally fair walking environments for all locations) and **rights** (the right to mobility/accessibility for pedestrians)",
-                           "How?: perceived pedestrian perception on protection, ease, equitable access, mobility, identity, and enjoyment must be met."
+                           "Perceived pedestrian perception on protection, ease, equitable access, mobility, identity, and enjoyment must be met."#How?: 
                            ), ncol = 4, byrow = T))
+
+# dat |>
+#   gt() |>
+#   tab_options(table.font.size = "xx-small") |>
+#   cols_label(X1 = "Dimension (What?)",
+#              X2 = "Continent (Where?),
+#              X3 = "Conceptualization",
+#              X4 = "Standard") |>
+#   tab_caption("Examples of Data Extraction") |>
+#   fmt_markdown(columns = everything()) |>
+#   tab_options(table.width = px(400))
+
+colnames(dat) <- c("What dimensions?", "When and where?", "How is fairness conceptualised wrt Who and What?","How is the standard defined?")
 ```
@@
 ```{r}
-dat |>
-  gt() |>
-  tab_options(table.font.size = "xx-small") |>
-  cols_label(X1 = "Dimension",
-             X2 = "Continent",
-             X3 = "Conceptualization",
-             X4 = "Standard") |>
-  tab_caption("Examples of Data Extraction")
-```
-:::
-
-<!-- OLD ABSTRACT
-
-Transportation systems are technologies essential for social inclusion and activity participation, and therefore tracking inequalities is of interest to many.
-Our work responds by providing a synthesis of the state of the knowledge in the academic transportation fairness, justice, and equity related literature.
-First, we outline a flexible framework for engaging questions of transportation fairness and equity analysis based on the questions "Why?", "Where?", "When?", "Who?", "What?", and "How?" (5WH).
-Second, we apply the framework to collate findings from a broad scoping review of the transportation literature.
-In contrast to previous reviews on measuring inequality in transportation systems, our work is concerned with the implicit or explicit standards used to judge whether inequalities are fundamentally “fair” or unacceptable.
-We find the majority of work is published within the last half decade (55%) and centers the urban context (85%); the minority of work explores Global South case studies (40%) which focus disproportionally on transport financial affordability as well as informal transportation systems and infrastructure development needs relative to studies from the Global North.
-Income groups, followed by specific age groups (i.e., elders) and people with disabilities are most commonly the subjects of justice (the Who) as they experience disporportionate transport-burdens.
-What mobility tools are studied are most often transit or pedestrian modes; with many analyses being multimodal and comparative in nature.
-The benefits and burdens, the "What" of mobility, can be summarised as a focus on movement (i.e., trips taken, quality of trips) or the potential for movement (i.e., accessibility).
-How the reviewed works conceptualise fairness is delineated most often with a cut-off opportunity standard (37%) (i.e., number of parks within a 30-minute travel time) or population standard (36%) (i.e., bottom income quartile), though Infrastructure (i.e., level of service) and Environmental+ (i.e., air pollution) are also prominent.
-How the works conceptualise fairness is varied and associated with what type of standard is proposed, but often applies Vertical Equity, Transport-Related social exclusions, and Wellbeing overall with Population and Infrastructure standards being supported by Rights-based and Inequitable Externalitiy conceptualisations and Opportunity standards with Horizontal equity and Spatial equity.
-We identify ways forward to continue progress within the transportation justice literature by synthesizing five calls for action directed to researchers, practitioners, decision-makers and transportation advocates.
--->
+#| label: tbl-tabA63
+#tbl-cap: "Sample extracted articles."
+ft <- dat |>
+flextable() |>
+colformat_md() |>
+width(j = c(1,2), width = 2, unit = "cm")|>
+width(j = c(3,4), width = 6, unit = "cm") |>
+fontsize(size = 8, part = "all")
+ft
+```
+
+<!-- :::-->

### 2025-03-11 paper/paper_old_submittedtoreviews.qmd
'paper' - submitted to journal.

--- 2024-05-15 Journal-manuscript/Journal-manuscript.qmd
@@
   # - literature review
-
+abstract: | 
+  | Our review provides a synthesis of how transportation fairness, justice and equity literature has operationalized standards.
+We first outline a flexible framework for engaging fairness questions and then apply it to collate the review's results.
+We find most articles were published within the last half decade and center urban areas, with a minority of work (40%) exploring Global South case studies.
+Income groups, followed by specific age groups (i.e., elders) and people with disabilities, are most commonly the subjects of justice.
+Transit and pedestrian modes are the most frequently studied mobility tools; with many analyses being multimodal and comparative in nature.
+The benefits and burdens, the "What" of mobility, focuses on movement (i.e., quality of trips taken) or the potential for movement (i.e., accessibility).
+How the reviewed works conceptualizes fairness is varied (e.g., Vertical Equity, Wellbeing, Rights-Based); but is often delineated with a supporting opportunity standard (37%) (i.e., number of parks within a 30-minute travel time) or population standard (36%) (i.e., bottom income quartile), though Infrastructure (i.e., level of service) and Environmental+ (i.e., air pollution) standards are also prominent.
+While the reviewed literature covers plenty of ground, we conclude with calls for action directed at researchers, practitioners, and transportation advocates.
+
+keywords: [transportation; equity; justice; standards; transportation planning; fairness; literature review]
 bibliography: bibliography.bib 
@@
 
-# Introduction {#sect1}
+# Introduction
 
@@
 Transportation systems that are engineered to offer higher mobility for people _somewhere_ can simultaneously cut others off from essential opportunities _elsewhere_, as @rajeEngineering2004 and others have poignantly illustrated.
-Over time, the shades of policies past long haunt a region and the planet, as shown by the legacy of urban highways all across the US [@archerWhite2020] and the time horizon for the impacts of climate change [@markolfTransportation2019].
+Over time, the shades of policies past long haunt a region and the planet, as shown by the legacy of urban highways all across the U.S.
+[@archerWhite2020] and the time horizon for the impacts of climate change [@markolfTransportation2019].
 
@@
 Further, there are multiple works that discuss the conceptual and philosophical foundations of equity and fairness in transportation [e.g., @martensTransport2016; @pereiraDistributiveJusticeEquity2017a; @vanoutriveHow2019].
-Previous reviews of equity in planning documents have been tightly scoped to cover accessibility [e.g., @boisjolyHowGetThere2017] or a particular mode of transportation [e.g., cycling, @doranPursuitCyclingEquity2021].
+Previous reviews of equity in planning documents have been tightly scoped to cover accessibility [e.g., @boisjolyHowGetThere2017] or a particular mode of transportation [e.g., cycling in @doranPursuitCyclingEquity2021].
 These efforts are valuable, but in our estimation there remains a gap in terms of understanding how standards for equity are developed and implemented for transportation systems.
@@
 Second, it applies the framework to collate the existing knowledge about equity standards.
+To achieve this, we scan the state of academic knowledge in defining standards of fairness in transportation.
 In contrast to previous reviews on measuring inequality in transportation systems, we are concerned with the implicit or explicit standards used to judge whether inequalities are fundamentally “fair” or unacceptable.
-To achieve this we scan the state of academic knowledge in defining standards of fairness in transportation.
-
-# Background {#sect2}
-
-In the introduction the terms "justice", "equity", and "fairness" were used relatively loosely.
+
+# Background
+
+In the introduction, the terms "justice", "equity", and "fairness" were used relatively loosely.
 In the following, we describe justice as a target, equity as a set of measurement tools, fairness as a yardstick of justice, and standards as statements of fairness.
@@
 To illustrate, retributive justice is achieved by distributing societal intangibles like "freedom" (e.g., of movement, of association) as benefits, and/or tangible resources as burdens (e.g., fines as a punishment).
-Reparative justice deals with the distribution of benefits and burdens to redress past wrongs, for example by asking those who have benefited from said wrongs, even if unwittingly, to shoulder a bigger fiscal burden to benefit those who are still harmed by them.
+Reparative justice deals with the distribution of benefits and burdens to redress past wrongs, for example by asking those who have benefited from said wrongs, even if unwittingly, to shoulder a bigger fiscal burden or lesser power in planning processes, to benefit those who are still harmed by them.
 Procedural justice is the distribution of societal benefits (e.g., the right to voice an opinion as a recognized stakeholder in a process) and burdens (e.g., the effort required to develop an educated opinion) when making decisions with collective consequences.
@@
 The concept does not have the same history of development as an analytical tool, and can be interpreted in numerous, and possibly discordant ways.
-This is demonstrated by @martensFair2021 in their study of the application of Title VI of the US Civil Rights Act of 1964 in accessibility planning.
-Title VI explicitly talks about the distribution of benefits derived from Federal funding: "[n]o person in the United States shall, on the ground of race, color, or national origin, be excluded from participation in, be denied the benefits of, or be subjected to discrimination under any program or activity receiving Federal financial assistance." However, as @martensFair2021 show, there are wildly different paths to compliance, ranging from the banal (do not _knowingly_ discriminate), to the substantive (compensation for past discrimination within a societal context that recognizes harm was done, i.e., reparative justice).
+This is demonstrated by @martensFair2021 in their study of the application of Title VI of the U.S.
+Civil Rights Act of 1964 in accessibility planning.
+Title VI explicitly talks about the distribution of benefits derived from Federal funding: "[n]o person in the United States shall, on the ground of race, color, or national origin, be excluded from participation in, be denied the benefits of, or be subjected to discrimination under any program or activity receiving Federal financial assistance." However, as @martensFair2021 show, there are wildly different paths to compliance, ranging from the banal (do not _knowingly_ discriminate), to the substantive (compensation for past discrimination within a societal context that recognizes harm was done i.e., an aspect of reparative justice [@williamsWeStepHow2023]).
 What kind of justice does fairness serve in each case?
@@
 A policy based on this standard would be fair as long it does no harm.
-<!--A somewhat more strict standard, a Pareto-Plus improvement, stipulates that an intervention is fair when all groups receive at least _some_ (non-trivial) benefits; the size of the benefits for each group is irrelevant.
-In contrast to the notion of "do no harm", such a standard embodies the ideal that no one is denied benefits.
---> A standard based on egalitarian principles (e.g., proportional equity) would weigh the benefits or burdens by population; in fairness, each group would give or receive in proportion to their size [@billsLookingMeanEquity2017;@martensJusticetheoreticApproachDistribution2012].
+A standard based on egalitarian principles (e.g., proportional equity) would weigh the benefits or burdens by population; in fairness, each group would give or receive in proportion to their size [@billsLookingMeanEquity2017;@martensJusticetheoreticApproachDistribution2012].
 In contrast, an affirmative action standard (restorative justice) would be even stricter, requiring the benefits to be distributed in a non-egalitarian way to favor people still harmed by past or present discriminatory practices [@bierbaumMobilityJusticeLinking2021].
 
-## A framework to analyze questions of justice: 5WH {#sect2.2}
+## A framework to analyze questions of justice: 5WH
 
@@
 Conventionally, the question of **"When?"** refers to the temporal circumstances within which the demands of justice apply.
-When it comes to transportation, questions regarding temporailty are important, as for example: _when_ did the equity analysis take place and under what historical policy context; when is _the right time_ to invest in infrastructure and as a result when to generate a spatial inequality [@rabelloEvaluation2015]; for _how long_ the burdens and benefits can be associated to a specific intervention; or even the _timeline_ of reparative justice to reconcile the shadows of past transportation-related injustices.
+When it comes to transportation, questions regarding temporailty are important, as for example: _when_ did the equity analysis take place and under what historical policy context; when is _the right time_ to invest in infrastructure and as a result when to generate a spatial inequality [@rabelloEvaluation2015]; for _how long_ the burdens and benefits can be associated to a specific intervention; or even the _timeline_ of reparative justice to reconcile the shadows of past transport-related injustices.
 
@@
 Equity standards are a tool for answering this distributive question, by defining what is fair.
-The thresholds can be quantitative (e.g., square meters of green space per capita), or they can be qualitative descriptions (e.g., do not knowingly discriminate), or a mix of the two.
+The thresholds can be quantitative (e.g., square meters of green space per capita), qualitative descriptions (e.g., do not knowingly discriminate) or a mix of the two.
 Some examples include: maximum travel distance/cost/time to or from key destinations, levels of maximum exposure to externalities (i.e., noise or air pollution), un/fulfilled needs, and dis/satisfaction with travel.
-To support us when approaching this question we draw from concepts in transport-related social exclusion, transport disadvantage, and/or transport poverty, which are typically based on utilitarian or sufficientarian philosophies.
+To support us when approaching this question, we draw from concepts in transport-related social exclusion, transport disadvantage, and transport poverty, which are typically based on utilitarian or sufficientarian philosophies.
 
 Above all, convincing answers to the above questions require a supporting rationale: a **"Why?"** [@jaggarPhilosophicalChallengesGlobal2009].
-This is perhaps the most slippery of the questions posed here.
+This is perhaps the most slippery of 5WH.
 Justice is inherently a social construct.
@@
 
-## Methods {#sect2.3}
+## Methods
 
@@
 
-The semi-automated nature of the search strategy was overly inclusive, and reduced the risk of omitting relevant material.
+The semi-automated nature of the search strategy was overly inclusive, thereby reducing the risk of omitting relevant material.
 The next stage was to trim the corpus of documents, a task that could no longer be automated.
@@
 Next, using a data extraction template and workflow that was pilot-tested with a subset of papers, the authorship team extracted data from the eligible documents using *Covidence* [@covidenceCovidence2023], an online application for literature screening.
-The evidence selection workflow, (PRISMA) flow diagram [@pagePRISMA2020Statement2021], data extraction template, and some sample data extractions can be consulted in Appendix @fig-figA2.
+The evidence selection workflow, flow diagram, data extraction template, and some sample data extractions can be consulted in Appendix @fig-figA2.
 The end result of this scan was a corpus of 165 documents retained for data extraction.
 
-# An appraisal of the lay of the land {#sect3}
-
-This section threads together the trends identified from the data extracted from the corpus (@fig-figA2)<!-- update this cross reference!-->.
+# An appraisal of the lay of the land
+
+This section threads together the trends identified from the corpus (@fig-figA2)<!-- update this cross reference!-->.
 Specifically, the "How" and its application to the objects of justice (the "What"), the subjects of justice (the "Who"), and under which situations ("When" and "Where").
@@
 #| label: fig-fig1
-#| fig-cap: "The proportion of category for each When, Where, Who and What; papers do often contain more than one topic and those papers are represented multiple times.
-Topic categories were generated upon data extraction."
+#| fig-cap: "The proportion of category for the 5Ws; papers do often contain more than one topic and those papers are represented multiple times. Topic categories were generated upon data extraction."
 
@@
 
-## "When" and "Where" is transportation considered a sphere of justice {#sect3.1}
-
-Within our corpus most papers (`r ((case_cont_count$count[4] + case_cont_count$count[5]+ case_cont_count$count[6])/165) |> percent()`) focus on studies in the Global North, with many studies from North America (particularly USA and Canada), Europe (particularly France, Spain and Northern Europe), Oceania (Australia and New Zealand), and Asia (Japan and Israel).
-Though their subject matter is varied, their spatial context mainly pertains to North America and Europe, and thus more often than not deal with more developed and formal government transport planning apparatuses and technologies e.g., planning for equitable high-speed rail [@monzonEfficiencySpatialEquity2013], autonomous vehicle technology [@eppenbergerOpportunitySharedAutonomous2021], or on the public consultation processes [@reddySafeguardingMinorityCivil2010].
+## "When" and "Where" is transportation considered a sphere of justice
+
+Within our corpus most papers (`r ((case_cont_count$count[4] + case_cont_count$count[5]+ case_cont_count$count[6])/165) |> percent()`) focus on studies in the Global North, with many studies from North America (particularly U.S.
+and Canada), Europe (France, Spain and Northern Europe), Oceania (Australia and New Zealand), and Asia (Japan and Israel).
+Though their subject matter is varied, their spatial context mainly pertains to North America and Europe, and thus more often than not deals with more developed and formal government transport planning apparatuses and technologies e.g., planning for equitable high-speed rail [@monzonEfficiencySpatialEquity2013], autonomous vehicle technology [@eppenbergerOpportunitySharedAutonomous2021], or on the public consultation processes [@reddySafeguardingMinorityCivil2010].
 
@@
 
-The literatures on the Global North and South literature seem to be situated at different stages in the process of addressing transport inequities.
+The literatures on the Global North and South seem to be situated at different stages in the process of addressing transport inequities.
 As formal planning processes in the South operate under greater financial precariety, they rely on more informal processes to address unmet needs.
@@
 In contrast, these minimum standards may not be relevant in some areas of the Global North e.g., $NO_2$ pollution near roadways in Montreal are consistently below the WHO standard though populations are disproportionately impacted [@carrierApplicationThreeMethods2014].
-Overall, the Global South lags behind the Global North, but this situation presents surprising opportunities for the Global South, to adopt potentially successful enhancements from Global North formal equity planning processes (e.g., indicator creation for disadvantages groups [@cuiSpatialAccessPublic2020]) and to avoid past mistakes (e.g., from entrenched car-centric development [@warrenDevelopingEquitableSustainable2015] to the disproportionate contribution to carbon-intense mobility [@chancelCarbonInequalityKyoto2015].
+Overall, the Global South lags behind the Global North, but this situation presents surprising opportunities for the Global South, to adopt potentially successful enhancements from Global North formal equity planning processes (e.g., indicator creation for disadvantages groups [@cuiSpatialAccessPublic2020]) and to avoid past mistakes (e.g., from entrenched car-centric development [@warrenDevelopingEquitableSustainable2015] to the disproportionate contribution to carbon-intense mobility [@chancelCarbonInequalityKyoto2015]).
 For the Global North, there is an opportunity to reckon with its own contributions to uneven development globally and environmental impacts as well as adopt relevant aspects from informal planning processes.
 
-## "Who" are the subjects of transportation justice and their mobility tools {#sect3.2}
+## "Who" are the subjects of transportation justice and their mobility tools
 
@@
 Yet, low-income is not universally associated with lower transport-related benefits for every object of justice.
-For instance, in Sheffield, UK, @mearsUnderstandingSocioeconomicEquity2019 demonstrate that historically working-class (lower-income) neighbourhoods have _more_ access to green space than other neighbourhoods, but of lower quality, likely due to historic urban planning approaches.
-Similarly, @bertrandMeasuringMappingDisparities2008 find that the granularity of the analysis matters, and lower income groups do not always have lower accessibility , something echoed in other spatial and temporal contexts [@fothEquitableTransitExamining2013a; @allenSizingTransportPoverty2019a]
+For instance, in Sheffield, U.K., @mearsUnderstandingSocioeconomicEquity2019 demonstrate that historically working-class (lower-income) neighbourhoods have _more_ access to green space than other neighbourhoods, but of lower quality, likely due to historic urban planning approaches.
+Similarly, @bertrandMeasuringMappingDisparities2008 find that the granularity of the analysis matters, and lower income groups do not always have lower accessibility, something echoed in other spatial and temporal contexts [@fothEquitableTransitExamining2013a; @allenSizingTransportPoverty2019a]
 
 "Age" is the second most common personal trait.
-In this regards, @martinez-jimenezAccessibilityCultureEducation2019 and @arranz-lopezSocialSpatialEquity2019 investigate travel distances/times to various opportunities, demonstrating how age is associated with differences in opportunity access.
+In this regard, @martinez-jimenezAccessibilityCultureEducation2019 and @arranz-lopezSocialSpatialEquity2019 investigate travel distances/times to various opportunities, demonstrating how age is associated with differences in opportunity access.
 School-aged children and older populations are a common focus.
@@
 These indices are generated from government sources or author-informed census data creation methods.
-As an example, the Neighbourhood Equity Index (NEI) is a measure of vulnerability created by the City of Toronto and used in @awuorInfluenceEnvironmentalHealth2019 to examine disparities in premature death.
+As an example, the Neighbourhood Equity Index is a measure of vulnerability created by the City of Toronto and used in @awuorInfluenceEnvironmentalHealth2019 to examine disparities in premature death.
 Other works use national census indicators such as the social and housing deprivation index [@pucciInequalitiesJobrelatedAccessibility2019] or explore transport disadvantage, equity in policy implementation, or transport-related mortality burden by means of census measures (e.g., household poverty) and transport-related accessibility indicators [@aldredEquityNewActive2021; @iungmanImpactUrbanTransport2021; @sunPublicTransportAvailability2021; @scheurerSpatialAccessibilityPublic2017].
-Similarly, Environmental Justice (EJ) indicators have been used in the US literature to identify neighbourhoods that have a higher than average proportion of low-income and non-white populations and evaluate the equity impacts of transportation projects [@rowangouldIdentifyingEnvironmentalJustice2016; @parkTransitParksEnvironmental2021; @reddySafeguardingMinorityCivil2010].
+Similarly, Environmental Justice (EJ) indicators have been used in the U.S.
+literature to identify neighbourhoods that have a higher than average proportion of low-income and non-white populations and evaluate the equity impacts of transportation projects [@rowangouldIdentifyingEnvironmentalJustice2016; @parkTransitParksEnvironmental2021; @reddySafeguardingMinorityCivil2010].
 
@@
 Only a few papers focus exclusively on gendered differences in active transportation [@adlakhaMindGapGender2020, @xieWonCycleRoute2018], race/ethnicity's relationship to green space proximity [@silvaEnvironmentalJusticeAccessibility2018], and culturally-appropriate opportunities [@wangModelingSpatialAccessibility2011].
-Papers that focus *solely* on "Employment" status or "Education" level are completely absent.
-Furthermore, "Other" population groups are also frequently considered in the corpus: this is a catch-all category that includes populations less commonly targeted by research e.g., veterans and access to specific-healthcare needs [@mooneyTravelDistanceBarrier2000], pregnant people and access to maternity services [@vadrevuMeasuringSpatialEquity2016], and youth who live in foster care [@batscheUsingGISEnhance2012].
+Papers that focus *solely* on "Employment" status or "Education" level are completely absent in the corpus.
+Furthermore, "Other" population groups are also frequently considered: this is a catch-all category that includes populations less commonly targeted by research e.g., veterans and access to specific-healthcare needs [@mooneyTravelDistanceBarrier2000], pregnant people and access to maternity services [@vadrevuMeasuringSpatialEquity2016], and youth who live in foster care [@batscheUsingGISEnhance2012].
 Overall, the diversity of population groups considered in the corpus demonstrates the variety of transportation-equity concerns addressed in the literature.
@@
 
-Transit characterises the majority of the corpus (@fig-fig1).
-Despite transit being perceived unfavourably by some [e.g., @mellaDrivers2021], it is often viewed as the only and primary mobility option mode for many [@jacquesRescuing2012].
+Transit characterises the majority of the corpus (@fig-fig1) with some intuitive sense.
+Despite transit being perceived unfavourably by some [e.g., @mellaDrivers2021], it is often viewed as the only or primary mobility option mode for many [@jacquesRescuing2012].
 Papers that focus on transit assess a variety of topics, including food deserts [@mckeyCrowdsourcedMappingHealthy2020] and barrier-free transportation for people who face disabilities [@jimenez-espadaResearchProblemUniversal2021; @liuStatusQuoChallenges2019; @limFacilitatingIndependentCommuting2021; @feeleyValidationParatransitSkills2019].
@@
 
-Transit also plays a central role in multimodal comparisons for transport equity analysis, and often feature in papers dealing with "Walk", "Car", or some other category in @fig-fig1.
+Transit also plays a central role in multimodal comparisons for transport equity analysis, and often featured in papers dealing with "Walk", "Car", or some other category in @fig-fig1.
 For instance, @brusselAccessAccessibilityCritique2019 compares transit, pedestrian and road network accessibility measurements in the context of SDG 11.2.
-@renneWhatHasAmerica2018 reviews natural disaster evacuation plans, focusing on car-less households and emphasizing transit and pedestrian networks.
+@renneWhatHasAmerica2018 reviews natural disaster evacuation plans, focusing on carless households and emphasizing transit and pedestrian networks.
 A few papers also frame transit as a direct competitor of car travel or use it as a benchmark [@golubUsingPrinciplesJustice2014; @martensJusticetheoreticApproachDistribution2012].
-For instance, @warrenDevelopingEquitableSustainable2015 propose car ownership standards while acknowledging the tension between mobility needs in transit under-served areas and emission reduction goals.
+For instance, @warrenDevelopingEquitableSustainable2015 proposes car ownership standards while acknowledging the tension between mobility needs in transit under-served areas and emission reduction goals.
 However, this framing is not universal, and transit is sometimes seen as a mode to fulfill individual capabilities.
@@
 
-The third most studied mode is "Car".
-Unlike transit and walking, cars are rarely the sole focus of equity research.
-In our corpus they are often critically compared with transit or considered in areas with inadequate transit [@kimmelStructuralBarriersComprehensive2018; @aljoufie2016].
-Upon the rise of automobility, when walking and cycling got in the way of faster travel, transit came to be seen as a social service and sometimes as a hindrance to the full realization of automobility.
-Our assessement of the focus on non-car modes in our corpus is that it represents a reckoning with automobility's legacy and ongoing demands for space, public subsidies, and government supports that underestimate the true cost of cars [@gosslingSocial2019; @timperleyFight2021].
-In addition to comparing "Car" to other modes, some papers reflect on "Other" forms of mobility (@fig-fig1), including wheelchair accessible taxi and/or paratransit services [@marquezFactorsAffectingPersonal2019; @wilkinsonmeyersLiveOrdinaryLife2015], travel on waterways [@vadrevuMeasuringSpatialEquity2016; @caoIndicatorsSocioSpatialTransport2017; @parrySocialVulnerabilityClimatic2018], motorcycle or other micro-mobility [@berryInvestigatingFuelPoverty2016; @savingmothersDidSavingMothers2019; @tiwariDevelopmentPublicTransport2021], or emergency vehicles [@pedigoInvestigationDisparitiesGeographic2010; @patelDeterminingGeographicAreas2007].
+The third most studied mode in our corpus is "Car".
+Unlike transit and walking, these studies rarely feature car as the mode of primary focus.
+Car is often assessed against transit and active modes [e.g., low traffic neighbourhoods @aldredEquityNewActive2021] or in areas with inadequate transit service [@kimmelStructuralBarriersComprehensive2018; @aljoufie2016].
+Under our appraisal, transit and walk modes are of most interest because they represent a reckoning with automobility's legacy and ongoing demands for space, public subsidies and government supports that underestimate their true cost [@gosslingSocial2019; @timperleyFight2021].
+Upon the rise of the automobile, pedestrian spaces were distinguished, non-car modes came to be seen as a hindrance to full automobility, and transit seen as a social service.
+We ascertain that for these reasons, "Car" is often compared against non-car modes and never put forward as the only mode that should be expanded.
+Furthermore, some papers reflect on "Other" forms of mobility (@fig-fig1), including wheelchair accessible taxi and/or paratransit services [@marquezFactorsAffectingPersonal2019; @wilkinsonmeyersLiveOrdinaryLife2015], travel on waterways [@vadrevuMeasuringSpatialEquity2016; @caoIndicatorsSocioSpatialTransport2017; @parrySocialVulnerabilityClimatic2018], motorcycle or other micro-mobility [@berryInvestigatingFuelPoverty2016; @savingmothersDidSavingMothers2019; @tiwariDevelopmentPublicTransport2021], or emergency vehicles [@pedigoInvestigationDisparitiesGeographic2010; @patelDeterminingGeographicAreas2007].
 
@@
 
-As shown, "who" is the subject of inequities provides ample but not exhaustive coverage of populations.
-But who is missing from the "Who" question?
-There is a need to go beyond the focus on low-income transit riders; positioning intersectionality of socio-economic and mode-use to create community-based definitions of equity and tailored standards are critical.
-How decision makers define an equity-deserving community will impact results e.g., [@rowangouldIdentifyingEnvironmentalJustice2015].
-In this sense, standards need to be sensitive to changing community-based definitions of inequities.
+As shown, "Who?" is the subject of inequities provides ample but not exhaustive coverage of populations.
+But who is missing from this question?
+There is a need to go beyond the focus on low-income transit riders; positioning intersectional considerations within the creation of community-based equity definitions and tailored standards are critical, as the act of delineating equity-deserving communities impacts results e.g., [@rowangouldIdentifyingEnvironmentalJustice2015].
+Standards should be sensitive to changing community-based definitions of inequities.
 Are issues of economic-inequity at the root of transport inequities for a specific community?
-Are (dis)abilities?
-Are inequities in the service of transit the focus of a study because it can be improved to address transport inequities and are alternative modes driving these inequities?
+Are inequities in the access provided by certain modes the focus in part because access offered by other modes are _relatively_ better?
 Access to what sorts of opportunities are driving transport inequities?
-How do populations, transit modes, and opportunities sought intersect to define the "who" of inequities?
+How do populations, travel modes, and opportunities sought intersect to define the "who" of inequities?
 A community- informed understanding of inequities and tracking how it changes are needed.
 
-## "What" are the objects of transportation justice {#sect3.3}
+## "What" are the objects of transportation justice
 
@@
 In terms of destinations, "Healthcare" services (`r (dest_count$count[6]/165) |> percent()`) and "Employment" (`r (dest_count$count[5]/165) |> percent()`) are the most common in our corpus (@fig-fig1).
-Papers on healthcare often highlight disparities in services, like @wangModelingSpatialAccessibility2011, who assess access to Mandarin-speaking family physicians for Chinese speakers in Toronto.
+Papers on healthcare often highlight disparities in services, like @wangModelingSpatialAccessibility2011, who assess access to Mandarin-speaking family physicians for Mainland Chinese immigrants in Toronto.
 Similarly, papers focusing on employment are often aimed at identifying transportation-poor neighbourhoods [@allenSizingTransportPoverty2019a; @churchillTransportPovertySubjective2019].
@@
 Fewer papers cover "Community" destinations (e.g., public service centres, places of community support or worship) (`r (dest_count$count[2]/165) |> percent()`) or "Childcare" (`r (dest_count$count[1]/165) |> percent()`), but they are integral to holistically study activity participation [@albertsRebuildingWomenLivelihoods2016; @smithAccessibilityCapabilityMinimum2012].
-The lack of information about community destinations, especially for children, is noticeable [@desjardinsChildren2022].
-
-## "How" is fairness determined {#sect3.4}
+The lack of information about community destinations in the corpus, especially for children, is noticeable [@desjardinsChildren2022].
+
+## "How" is fairness determined
 
@@
 
-The conceptual foundations of fairness are not always made explicit by researchers and often had to be inferred from the corpus.
-Some categories emerged from our review and we must note that they are not exhaustive.
-In broad terms, they are (@fig-fig2 details definitions):
+The conceptual foundations of fairness are not always made explicit within the corpus, they often had to be inferred.
+Though not exhaustive, the following categories emerged from our review and in broad terms are (@fig-fig2 details definitions):
 
@@
 A slice of the corpus is supported by broader "Rights" conceptualisations of fairness: these papers often focus on equity for people with disabilities or non-car users and associated challenges accessing transport infrastructure [@jimenez-espadaResearchProblemUniversal2021; @daamenAssessingGapPublic2008; @bharathyRevisitingClearFloor2018].
-While many papers are underpinned with the right to the city (the _right_ to participate in the production of urban space [@lefebvreDroitVille1967]) [@adliRightCityApplying2018], others emphasize legal _Rights_ like ADA regulations [@bharathyRevisitingClearFloor2018] or the goal of _access for all_ in land-use transportation master plans [@limFacilitatingIndependentCommuting2021].
+While many papers are underpinned with the right to the city (the _right_ to participate in the production of urban space [@lefebvreDroitVille1967; @adliRightCityApplying2018]), others emphasize legal _Rights_ like regulations aligned with The Americans with Disabilities Act (ADA) [@bharathyRevisitingClearFloor2018] or the goal of _access for all_ in land-use transportation master plans [@limFacilitatingIndependentCommuting2021].
 
@@
 
-"Environmental+" standards feature less prominently in our corpus, with studies that predominantly examine "inequitable externalities", possibly because the environmental burdens of transportation are addressed more broadly in other literature (e.g., environmental justice).
+"Environmental+" standards are featured less prominently in our corpus, possibly because the environmental burdens of transportation are addressed more broadly in other literature (e.g., environmental justice).
 However, the papers included in our corpus present some interesting insights.
-They often use traffic-related air and noise pollution, green space, urban design, urban air temperature, health outcomes, and physical activity guidelines to assess transport-related externalities.
+They often use traffic-related air and noise pollution, green space, urban design, urban air temperature, health outcomes, and physical activity guidelines to assess transport-related "inequitable externalities".
 Methods employed are primarily quantitative or mixed-methods, identifying inequalities through Gini coefficients [@fengTradeoffsMobilityEquity2014] or composite indices [@agostfelipInclusiveModelAssessing2021; @mirandaBenchmarkingSustainableUrban2012; @corazzaMethodologyEvidenceCase2020], occasionally incorporating spatial analysis [@jephcoteGeospatialAnalysisNaturally2013; @carrierApplicationThreeMethods2014].
@@
 
-# Moving forward: calls for action {#sect4}
-
-This paper makes two contributions to the literature.
-It outlines a conceptual and flexible "Why", Where", "When", "Who", "What" and "How" (5WH) framework with supporting definitions for approaching questions about transportation justice.
-It also applies the framework in the synthesis and appraisal of 165 academic articles drawn from the transportation justice literature; these studies use equity analysis as instruments to gauge proximity to their context's just situation.
+# Moving forward: calls for action
+
+This work makes two contributions to the literature.
+First, it outlines a conceptual and flexible "Why", Where", "When", "Who", "What" and "How" (5WH) framework with supporting definitions for approaching questions about transportation justice.
+Second, it applies the framework in the synthesis and appraisal of 165 academic articles drawn from the transportation justice literature; these studies use equity analysis as instruments to gauge proximity to their context's just situation.
 
@@
 Armed with appropriate knowledge, policy can be designed to address inequities while remaining adaptable to the evolving demands of justice.
+
 Flexible frameworks are crucial.
@@
 Creating adaptable transport justice frameworks that lend themselves to change is required to ensure resiliency to backsliding.
-From the diverse corpus reviewed in this paper, we argue for the necessity of setting standards from a systems approach, considering both _positive_ rights (e.g., a right to have access to sufficient quality of essential services) and _negative_ rights (e.g., mobility of cars must be limited to reduce their impact on air quality, health, safety, and other positive rights).
+From the diverse corpus reviewed in this work, we argue for the necessity of setting standards from a systems approach, considering both _positive_ rights (e.g., a right to have access to sufficient quality of essential services) and _negative_ rights (e.g., mobility of cars must be limited to reduce their impact on air quality, health, safety, and other positive rights).
 
@@
 
-## Call 1: Develop standards based on rigorous concepts of justice {#sect4.1}
+## Call 1: Develop standards based on rigorous concepts of justice
 
@@
 
-## Call 3: Making data available is a matter of justice {#sect4.3}
-
-In our review of the literature, we were left wondering specifically--what are the motivations for the use of some destinations and the choice to not include others?
-For instance, leisure destinations (e.g., green space, parks, recreation) are infrequently studied and some categories are missing altogether.
-We suspect that since methods used are predominately quantitative, the reliance on commonly used point of interest databases is also high.
-These databases typically include education, health and occasionally include aggregated categories for leisure and community, but are less generous in terms of 'community' and 'leisure' points of interest (community organizations, government services, care-giving destinations).
-Further, categories like childcare are often missing, typically daycare or facilities--domestic work, mobilities of care, and mobility interdependence.
-Not coincidentally, these categories are underrepresented compared to work and healthcare destinations.
-
-We can only track what we have collected and compiled; and we know that transport systems are more than just facilitators to work or as a source of economic development (though in underdeveloped regions, transport systems as a force of economic development is pronounced, e.g., high-speed rail [@chenImpactHighspeedRail2017; @monzonEfficiencySpatialEquity2013; @kimImpactsHighspeedRail2015]).
+## Call 3: Making data available is a matter of justice
+
+In our review of the literature, we were left wondering specifically-- what are the motivations for the inclusion of some destinations and the choice to not include others?
+For instance, leisure (e.g., green space, parks, recreation) and certain care destinations (e.g., community organizations, government services, banking locations, daycares) are infrequently studied or missing altogether.
+We suspect that since the methods used in our corpus are predominately quantitative, the reliance on commonly used point of interest databases is also high.
+These databases typically include education, health and occasionally include aggregated categories for leisure and community destinations, but are less generous for points of interest that are more often associated with understudied mobility patterns.
+
+However, through our corpus we know that transport systems are more than just facilitators to work or as a source of economic development (though in underdeveloped regions, transport systems as a force of economic development is pronounced, e.g., high-speed rail [@chenImpactHighspeedRail2017; @monzonEfficiencySpatialEquity2013; @kimImpactsHighspeedRail2015]).
 As such, data availability matters.
@@
 
-## Call 4: Develop more direct and explicit links between standards and lived experiences {#sect4.4}
+## Call 4: Develop more direct and explicit links between standards and lived experiences
 
@@
 
-# Declarations
-
-During the preparation of this work, the first and second authors used ChatGPT exclusively to reduce word count in the section "An appraisal of the lay of the land" while keeping the original writing style.
+# Acknowledgements and declarations
+
+This work would not have been possible without the help and inspiration of many individuals not listed as authors.
+In the early stages of this report, we received invaluable help from specialists at the University of Toronto as we learned about various approaches to literature reviews, particularly the Research Services and Liaison Librarians.
+These experts provided guidance in cornering down the topic, identifying guidance material on literature synthesis and relevant databases, and eased us into the world of knowledge synthesis.
+Though our path diverged from a traditional scoping review, the protocol published in September 2022 (found [here](https://osf.io/rsb92)) was followed with only minor changes, with only a reduction in the databases searched due to resource constraints.
+We also wish to acknowledge the many hours of work by undergraduate and graduate research assistants who poured through title and abstract screening to help determine full-text eligibility at the earliest stages of our search.
+The volume of literature initially reviewed was greater than any of us had experienced before.
+For this support, we are deeply thankful to Jan Domalaon, Dancel Gayle, Eryn Maloney, Mahdis Moghadasi, Anika Munir, and Subaita Refaaf.
+
+**Competing interests’ declaration:** The authors declare there are no personal or financial conflicts of interest.
+
+**Funding declaration:** This work was supported by the Mobilizing Justice Partnership and the Canada Graduate Scholarships—Doctoral Program from the Social Sciences and Humanities Research Council of Canada (SSHRC).
+
+**Ethical approval declaration:** Not applicable.
+
+**Participate consent declaration:** Not applicable.
+
+**Competing interests’ declaration:** The authors declare there are no personal or financial conflicts of interest.
+
+**Technology declaration:** During the preparation of this work, the first and second authors used ChatGPT to assist in reducing the word count in an early version of the “An appraisal of the lay of the land” section.
 After using this tool, all authors reviewed the manuscript and take full responsibility for the content of the publication.
-
-\newpage
 
@@
 -   **Context**: the included studies should also be limited to publications that include equity standards.
-Context can be more difficult to explicitly search for with key terms so synonyms for 'standards' were added to the query as a four set of topic search terms (e.g., threshold, indicator, criteria - see orange text in @fig-figA1 for full list).
+Context can be more difficult to explicitly search for with key terms so synonyms for 'standards' were added to the query as a four set of topic search terms (e.g., threshold, indicator, criteria - see orange text in @fig-figA2 for full list).
 Additionally, journal article and conference papers, English-language literature from any country, any study design (e.g., quantitative, qualitative, or mixed-method studies, or conceptual frameworks), and any record published within the past 30 years are included (January 1992 to March 2022).

### 2025-03-11 paper/paper.qmd
docx and coverletter, submitted!

@@
 author:
-  - name: AAA #Anastasia Soukhov
+  - name: Anastasia Soukhov
     #orcid: 0000-0003-4371-4831
-    # email: soukhoa@example.com
-    # affiliations: 
-    #     - ref: MAC
-    #       name: McMaster University
-    #       department: School of Earth, Environment and Society
-    #       address: 1280 Main Street West
-    #       city: Hamilton, ON
-    #       country: Canada
-    #       postal-code: L8S 4L8
-    # attributes:
-    #     corresponding: true
-  - name: BBB #Ignacio Tiznado-Aitken
-    # orcid: 0000-0002-7385-2357
-    # email: i.tiznadoaitken@utoronto.ca
-    # affiliations:
-    #     - ref: TORONTO
-    #       name: University of Toronto Scarborough
-    #       department: Department of Geography and Planning
-    #       city: Toronto, ON
-    #       country: Canada
-    #       postal-code: M1C 1A4
-  # - name: Matthew Palm
-  #   # orcid: 0000-0002-8800-2777
-  #   email: palmmatt@unc.edu
-  #   affiliations:
-  #       - ref: UNC
-  #         name: The University of North Carolina at Chapel Hill
-  #         department: Department of City and Regional Planning
-  #         city: Chapel Hill, NC
-  #         country: USA
-  #         postal-code: 27599
-  # - name: Steven Farber
-  #   #orcid: 0000-0002-3870-5984
-  #   email: steven.farber@utoronto.ca
-  #   affiliations:
-  #       - ref: TORONTO
-  #         name: University of Toronto Scarborough
-  #         department: Department of Geography and Planning
-  #         city: Toronto, ON
-  #         country: Canada
-  #         postal-code: M1C 1A4
-  # - name: Antonio Páez
-  #   #orcid: 0000-0001-6912-9919
-  #   email: paezha@mcmaster.ca
-  #   affiliations: 
-  #       - ref: MAC
-  #         name: McMaster University
-  #         department: School of Earth, Environment and Society
-  #         address: 1280 Main Street West
-  #         city: Hamilton, ON
-  #         country: Canada
-  #         postal-code: L8S 4L8
+    email: soukhoa@example.com
+    affiliations:
+        - ref: MAC
+          name: McMaster University
+          department: School of Earth, Environment and Society
+          address: 1280 Main Street West
+          city: Hamilton, ON
+          country: Canada
+          postal-code: L8S 4L8
+    attributes:
+        corresponding: true
+  - name: Ignacio Tiznado-Aitken
+  #orcid: 0000-0002-7385-2357
+    email: i.tiznadoaitken@utoronto.ca
+    affiliations:
+      - ref: TORONTO
+        name: University of Toronto Scarborough
+        department: Department of Geography and Planning
+        city: Toronto, ON
+        country: Canada
+        postal-code: M1C 1A4
+  - name: Matthew Palm
+    #orcid: 0000-0002-8800-2777
+    email: palmmatt@unc.edu
+    affiliations:
+        - ref: UNC
+          name: The University of North Carolina at Chapel Hill
+          department: Department of City and Regional Planning
+          city: Chapel Hill, NC
+          country: USA
+          postal-code: 27599
+  - name: Steven Farber
+    #orcid: 0000-0002-3870-5984
+    email: steven.farber@utoronto.ca
+    affiliations:
+        - ref: TORONTO
+          name: University of Toronto Scarborough
+          department: Department of Geography and Planning
+          city: Toronto, ON
+          country: Canada
+          postal-code: M1C 1A4
+  - name: Antonio Páez
+    #orcid: 0000-0001-6912-9919
+    email: paezha@mcmaster.ca
+    affiliations:
+        - ref: MAC
+          name: McMaster University
+          department: School of Earth, Environment and Society
+          address: 1280 Main Street West
+          city: Hamilton, ON
+          country: Canada
+          postal-code: L8S 4L8
 abstract: |
@@
 keywords: [transportation; equity; justice; fairness; standards; transportation planning; literature review]
-date: last-modified
+#date: last-modified
 bibliography: bibliography.bib
-format:
-  elsevier-pdf:
-    #keep-tex: true
-    include-in-header:
-      text: |
-        \usepackage{lineno}
-        \linenumbers
-    journal:
-      #name: Travel Behavior and Society
-      formatting: preprint
-      # model: 3p # Don't set a model with preprint
-      cite-style: authoryear
+format: docx
+  # nature-pdf: 
+  #   journal: "default"
+  #   keep-tex: true
+  #   include-in-header:
+  #     text: |
+  #       \usepackage{lineno}
+  #       \linenumbers
+  #   # journal:
+  #   #   # model: 3p # Don't set a model with preprint
+  #   #   cite-style: authoryear
 ---
